### 2020-03-02 README.Rmd
update readme to a Rmd file, updated docs to be cleaner (saw odd error - examples don't get the geoms allowed under ggtern still...)

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# timeternR

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/skgallagher/timeternR.svg?branch=master)](https://travis-ci.org/skgallagher/timeternR)
[![codecov](https://codecov.io/gh/skgallagher/timeternR/branch/master/graph/badge.svg)](https://codecov.io/gh/skgallagher/timeternR)
<!-- badges: end -->

The goal of timeternR is to develop functional based visualization and statistics that focus on use in ternary plots.

## Installation

You can install the developmental version of timeternR from github using:

``` r
# install.packages("devtools")
devtools::install_github("skgallagher/timeternR")
```

```{r echo = F, warning = F, message = F}
devtools::load_all()
```
## Data

   Currently there are 13 data sets in which to explore the different `timeternR` functions.  These correspond to different sources of raw data along, agent vs. aggregate format, and emphases of different processes such as vital dynamics (e.g. birth and).

   Details of each can be found with `?{data}`.

  - `agent_sims` this is 50 x 3 x 188 array where entry (i,j,k) looks at the ith simulation, the jth statistic, and the kth agent.  The statistics are (inititial state (0/1/2), SMax, IMax).

  - `agent_sims_tidy` ("Simulations of Measles outbreaks for Hagelloch, Germany, 1861 (Tidy agent format)"): tidy version of `agent_sims` (9400 x 5), each row corresponds to an individual for a single simulation, and contains additional information on the individual's initial state, and SMax, IMax.

  - `EpiModel_agg_bd` Example output from the `EpiModel package` for an ICM with birth and death rates.  The population $N$ is non-constant.

  - `EpiModel_det` output from the `EpiModel` package for a deterministic model (aggregate)

  - `EpiModel_icm` Output from the `EpiModel` package for a stochastic ICM (aggregate)

  - `hagelloch_agents` -- One row is a "sufficient" statistic for each agent's infection.  Each agent's infection is uniquely identified by an initial state, max time before infection (or T), and max time before recovery (or T).  For the states, 0 = S, 1 = I, 2 = R.

  - `hagelloch_aug_births` Like Hagelloch raw but augmented with births (it already had deaths).  Five fake people have been added who join the population over time

  - `hagelloch_raw` -- One row is an agent.  This is imported from the `surveillance` R package and the variable descriptions are found [here](https://rdrr.io/rforge/surveillance/man/hagelloch.html) where it is originally labeled `hagelloch.df`.  We have renamed it here to help distinguish it from the other data sets we derive from it.

  - `hagelloch_raw2` Like Hagelloch raw but with additional, fake people with NA values for `tI` and `tR`

  - `hagelloch_sir`  -- One row is a state of $(t, s_t, i_t, r_t)$ where $s_t + i_t + r_t = N$ for $t = 0, \dots, T=94$

  - `pomp_arr` Example SIR simulation output of class `array` from the `pomp` packagee

  - `pomp_df` Example SIR simulation of class `data.frame` from the `pomp` package

  - `pomp_pomp` Example SIR simulation output of class `pomp` from the `pomp` package

  - `pomp_sir` Example output from the `pomp` package.

## Example

```{r message = F, warning = F}
library(tidyverse)
library(ggtern)
library(timeternR)
```
```{r}
hagelloch_raw %>%
   dplyr::filter(SEX %in% c("male", "female")) %>%
   ggplot(aes(y = tI, z = tR, color = SEX)) +
       geom_aggregate() + 
       coord_tern() +
       labs(x = "S", y = "I", z = "R",
            color = "Gender")
```

### Simulate SIR data

```{r}
n_sims <- 100
n_time_steps <- 100
beta <- .1
gamma <- .03
init_SIR <- c(950, 50, 0)
output_format <- "data.frame"

out <- simulate_SIR_agents(n_sims = n_sims,
                           n_time_steps = n_time_steps,
                           beta = beta, gamma = gamma,
                           init_SIR = init_SIR,
                           output_format = output_format)

df_groups <- out %>% dplyr::group_by(sim) %>%
    agents_to_aggregate_SIR

df_groups %>% ggplot(aes(x = S, y = I, z = R, group = sim)) +
    geom_line(alpha = .1) +
    coord_tern()
```

## Notes:
1. For writing code that works with `tidyverse` 1.0 vs `tidyverse` <= 0.8.3 see:
https://tidyr.tidyverse.org/articles/in-packages.html, and for doing both with Travis CI (using a matrix), see https://github.com/tidyverse/design/issues/95. See Wiki for more comments on this.
    a. Also https://principles.tidyverse.org/changes-multivers.html has been being updated to give better approaches as well.
2. For writing your own `geom`s and `stat`s that works with `ggtern` (which are generally restricted), the following 2 stack-exchange articles helped use do so with ease:
    a. https://stackoverflow.com/questions/57175114/ternary-plot-scaling-opacity-across-groups
    b. https://stackoverflow.com/questions/31589479/how-to-fix-no-item-called-packagepkg-on-the-search-list-without-using-libra#comment51172891_31589479
    c. we've also leveraged ideas from https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`)

## Contributors

- Shannon Gallagher ([`skgallagher`](https://github.com/skgallagher))
- Benjamin LeRoy ([`benjaminleroy`](https://github.com/benjaminleroy))

### 2020-03-02 README.Rmd
changes to function names and examples relative to pkg-down website

@@
     b. https://stackoverflow.com/questions/31589479/how-to-fix-no-item-called-packagepkg-on-the-search-list-without-using-libra#comment51172891_31589479
-    c. we've also leveraged ideas from https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`)
+    c.
+we've also leveraged ideas from https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`).
+For some reason - when documenting for `pkg-down` website, we need to do `library(ggtern); timeternR:::update_approved_layers()`
 

### 2020-03-02 R/aaa.R

@@ -1,3 +1,12 @@
+# List of new stats
+.newstat = c(sir_aggregate        = "StatSirAggregate",
+             conf_band_kde        = "StatConfBandKDE",
+             conf_band_delta_ball = "StatConfBandDeltaBall",
+             conf_band_spherical  = "StatConfBandSpherical",
+             conf_band_chull      = "StatConfBandConvexHull")
+.newgeom = c(sir       = "GeomSirAggregate",
+             conf_band = "GeomConfBand")
+
 #' import hidden function / variable from other package
 #'
 #' see ?`:::`
@@ -6,7 +15,7 @@
 #' @param name hidden item (function / variable)
 #' @return hidden object or function
 #' @export
-importsHiddenFrom <- function(pkg, name){
+imports_hidden_from <- function(pkg, name){
   pkg <- as.character(substitute(pkg))
   name <- as.character(substitute(name))
   get(name, envir = asNamespace(pkg), inherits = FALSE)
@@ -22,31 +31,19 @@ importsHiddenFrom <- function(pkg, name){
 #'
 #' @import ggtern
 #'
-#' @examples
-#' .newstat = c(sir_raw       = "StatSirRaw",
-#'              sir_fortified = "StatSirFortified")
-#' .newgeom = NULL
-#' #update_approved_layers(stat_name = .newstat, geom_name = .newgeom)
-update_approved_layers <- function(stat_names = NULL, geom_names = NULL){
+update_approved_layers <- function(stat_names = .newstat,
+                                   geom_names = .newgeom){
   requireNamespace("ggtern", quietly = TRUE)
-  approvestatupdate <- c(importsHiddenFrom("ggtern",".approvedstat"),
+  approvestatupdate <- c(imports_hidden_from("ggtern",".approvedstat"),
                          stat_names)
   utils::assignInNamespace(".approvedstat", approvestatupdate,
                            pos = "package:ggtern")
 
-  approvegeomupdate <- c(importsHiddenFrom("ggtern",".approvedgeom"),
+  approvegeomupdate <- c(imports_hidden_from("ggtern",".approvedgeom"),
                          geom_names)
   utils::assignInNamespace(".approvedgeom", approvegeomupdate,
                            pos = "package:ggtern")
 
 }
 
-# List of new stats
-.newstat = c(sir_aggregate        = "StatSirAggregate",
-             conf_band_kde        = "StatConfBandKDE",
[…]

### 2020-03-02 R/bands.R

@@ -405,9 +405,9 @@ StatConfBandSpherical <- ggplot2::ggproto("StatConfBandSpherical",
 
         x_dim <- 2
 
-        a <- data2d %>% group_by(t) %>%
-          nest() %>%
-          mutate(n = purrr::map(data, function(df){nrow(df)}),
+        a <- data2d %>% dplyr::group_by(t) %>%
+          tidyr::nest() %>%
+          dplyr::mutate(n = purrr::map(data, function(df){nrow(df)}),
                  mean = purrr::map(data,function(df){df %>%
                      dplyr::select(x,y) %>%
                      sapply(mean)}),
@@ -415,11 +415,11 @@ StatConfBandSpherical <- ggplot2::ggproto("StatConfBandSpherical",
                      dplyr::select(x,y) %>%
                      cov})
           ) %>%
-          mutate(
+          dplyr::mutate(
             bound = purrr::map(n, function(n) {
               return((x_dim * (n-1)) / (n - x_dim) *
                        pf(q = alpha_level, df1 = x_dim, df2 = n - x_dim))})) %>%
-          mutate(inside_func = purrr::pmap(list(bound, mean, Sigma),
+          dplyr::mutate(inside_func = purrr::pmap(list(bound, mean, Sigma),
                   function(bound, mean, Sigma) {
                     check_inside_elipsoid_func(Sigma, mean, bound,
                                                suppress_warning = TRUE)}))
@@ -635,8 +635,9 @@ stat_confidence_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #' @export
 #' @examples
 #' library(ggplot2)
-#' library(ggtern)
 #' library(dplyr)
+#' library(ggtern); timeternR:::update_approved_layers()
+#' #                ^ this doesn't generally need to be done
 #'
 #' vis_data <- timeternR::pomp_df %>%
 #'   rename(x = "S", y = "I", z = "R") %>%

### 2020-03-02 R/stat-aggregate.R

@@ -96,7 +96,8 @@ StatSirAggregate <- ggplot2::ggproto("StatSirAggregate", ggplot2::Stat,
 #'@examples
 #' library(ggplot2)
 #' library(dplyr)
-#' library(ggtern)
+#' library(ggtern); timeternR:::update_approved_layers()
+#' #                ^ this generally need not be done
 #'
 #'
 #' # geom_aggregate
@@ -117,7 +118,7 @@ StatSirAggregate <- ggplot2::ggproto("StatSirAggregate", ggplot2::Stat,
 #' timeternR::hagelloch_raw %>% dplyr::filter(SEX %in% c("male", "female")) %>%
 #'   ggplot(., aes(y = tI, z = tR, color = SEX)) +
 #'   stat_aggregate(geom = "path") +
-#'   # note geom = "path" is the default (as is data_type = "raw")
+#'   # note geom = "path" is the default
 #'   coord_tern() +
 #'   labs(x = "S", y = "I", z = "R",
 #'        color = "Gender")

### 2020-03-02 R/timeternR.R
readme and pkgdown updates

@@ -1,0 +1,39 @@
+#' timeternR: Ternary plot visualization for SIR curves (UPDATE)
+#'
+#' The goal of timeternR is to develop functional based visualization and
+#' statistics that focus on use in ternary plots. (UPDATE)
+#'
+#' This package attempts to:
+#' \itemize{
+#'   \item \strong{Comparisons Model and Epidemic Comparisons}: We provide easy
+#'   ways to compare multiple epidemics, epidemics vs models, or models verse
+#'   models with ternary plots and confidence analysis
+#'     \itemize{
+#'       \item Ternary plots have been shown to help show time-independent like
+#'       comparing R_0
+#'       \item Added confidence band structure allows for examining variability
+#'       within simulations from models cleanly (and without worrying about time
+#'       variation)
+#'         \itemize{
+#'           \item Provides ways to moves beyond pointwise examination to
+#'           uniform containment
+#'         }
+#'     }
+#'   \item \strong{Agent (->/ vs) Aggregate}: Provide easy way to compare agents
+#'   based data/modeling vs aggregation
+#'     \itemize{
+#'       \item Our tools provide easy ways to explore different grouping of
+#'       agents and understanding how these groups different through aggregation
+#'       visualizations (see Shannon’s thesis for rational of quick access to
+#'       grouping abilities)
+#'    }
+#'   \item\strong{ggplot2/ggtern and tidyverse}: tools exist in the rapidly
+#'   expanding tidyverse/ggplot2 paradigm, allowing for quick deployment and
+#'   less cost in learning new assessment tools
+#'  }
+#' @import ggtern
+#' @importFrom rlang .data
+#'
+#' @docType package
+#' @name timeternR
+NULL

### 2020-03-04 README.Rmd
readme and pkgdown updates
documention update for check_ordered() and expand_info() along with README and vignette update for hagolloch data and pkgdown::build_site() + devtools::document() only

@@
 
-# timeternR
+# timeternR <img src="man/figures/logo.png" align="right" alt="" width="120" />
 
@@
 ```
+The following example comes from a Measles outbreak in Hagelloch, Germany in 1861. We have data on each child (agent) in the town.
 ```{r}
 hagelloch_raw %>%
-   dplyr::filter(SEX %in% c("male", "female")) %>%
-   ggplot(aes(y = tI, z = tR, color = SEX)) +
+  ggplot(aes(y = tI, z = tR)) +
+    geom_aggregate() + 
+    coord_tern() +
+    labs(x = "S", y = "I", z = "R", title = "Town Analysis")
+```
+
+Previous work has suggested that the class (`CL`) the student was in effected how the experienced the outbreak. The below figure shows differences in the outbreak relative to this grouping.
+```{r}
+hagelloch_raw %>% 
+  rename(`school grade` = CL) %>%
+  group_by(`school grade`) %>%
+  summarize(`number of students` = n())
+
+hagelloch_raw %>%
+   ggplot(aes(y = tI, z = tR, color = CL)) +
        geom_aggregate() + 
@@
        labs(x = "S", y = "I", z = "R",
-            color = "Gender")
+            color = "School Grade",
+            title = "Town analysis by grade") 
 ```

### 2020-03-04 R/fortify-aggregate-data-frame.R

@@ -3,11 +3,11 @@
 
 
 
-#' Takes in data from the R pomp package  where the output is a data frame and puts it in SIR format for timeternR
+#' Takes in data from the R pomp package where the output is a data frame and puts it in SIR format for timeternR
 #'
 #' @param ext_data Output from a pomp simulation where the output is a data frame, \code{pomp::simulate()}
 #' @param states vector of state names
-#' @param package_source optional package name 
+#' @param package_source optional package name
 #' @details The default variables that are retained are SIR, but can be modified with the \code{states} argument.  If code{states = NULL}, we will attempt to find all single letter names in POMP and output those.
 #' @return data frame with the following columns
 #' \describe{
@@ -27,6 +27,6 @@ fortify_aggregate_ext.data.frame <- function(ext_data,
         return(out)
     }
     stop("We currently do not support fortifying an object of class 'data.frame' if 'package_source' is not 'pomp'")
- 
+
 }
 

### 2020-03-06 README.Rmd
small updates to readme

@@
 
-# timeternR <img src="man/figures/logo.png" align="right" alt="" width="120" />
+# EpiCompare <img src="man/figures/logo.png" align="right" alt="" width="120" />
 
@@
 
-```{r}
+```{r eval = F}
 n_sims <- 100

### 2020-03-06 R/fortify-aggregate-epimodel-dcm.R
merging correction to check_ordered and updates to small elements in fortify_agg functions to pass checks.

@@ -1,15 +1,18 @@
-
-#'  Take external aggregate data and put it in a format used in this package
+#' Take external aggregate data and put it in a format used in this package
 #'
 #' @param data output from external source package.  See details
 #' @param states names of states we want aggregate totals of at each time
-#' @param package_source optional argument to include the package from which the output is derived from, which helps with the fortify function when outputs are of generic classes such as list or data.frame
+#' @param package_source optional argument to include the package from which the
+#'   output is derived from, which helps with the fortify function when outputs
+#'   are of generic classes such as list or data.frame
 #' @return a data frame with the following columns
 #' \describe{
 #' \item{t}{time}
 #' \item{Xk}{columns X0, ..., X_K. which are numeric}
 #' }
-#' @details This function converts external data sources (we currently support output from the EpiModel and pomp R packages), which is already aggregated and puts it in a format that can be used by our exploring functions.
+#' @details This function converts external data sources (we currently support
+#'   output from the EpiModel and pomp R packages), which is already aggregated
+#'   and puts it in a format that can be used by our exploring functions.
 #' @export
 #' @examples
 #' ## For dcm

### 2020-03-06 R/fortify-aggregate-epimodel-icm.R

@@ -1,11 +1,11 @@
 ## March 3, 2020
-
-#' Comments for Ben and Shannon
-#' 
-#' This is the start of a reformat of fortify_epimodel and fortify_pomp into one function
-#' "fortify_aggregate".  This will take in aggregate level data from external sources (currently pomp and
-#' epimodel) and output, if possible, output a data.frame  with the columns t, X0, .., XK where Xk are the
-#' state names and the values in Xk are the numbers in state k at time t.
+# Comments for Ben and Shannon
+#
+# This is the start of a reformat of fortify_epimodel and fortify_pomp into one
+# function "fortify_aggregate".  This will take in aggregate level data from
+# external sources (currently pomp and epimodel) and output, if possible,
+# output a data.frame  with the columns t, X0, .., XK where Xk are the state
+# names and the values in Xk are the numbers in state k at time t.
 
 
 
@@ -15,13 +15,17 @@
 #'
 #' @param data output from external source package.  See details
 #' @param states names of states we want aggregate totals of at each time
-#' @param package_source optional argument to include the package from which the output is derived from, which helps with the fortify function when outputs are of generic classes such as list or data.frame
+#' @param package_source optional argument to include the package from which the
+#'   output is derived from, which helps with the fortify function when outputs
+#'   are of generic classes such as list or data.frame
 #' @return a data frame with the following columns
 #' \describe{
 #' \item{t}{time}
 #' \item{Xk}{columns X0, ..., X_K. which are numeric}
 #' }
-#' @details This function converts external data sources (we currently support output from the EpiModel and pomp R packages), which is already aggregated and puts it in a format that can be used by our exploring functions.
+#' @details This function converts external data sources (we currently support
+#'   output from the EpiModel and pomp R packages), which is already aggregated
+#'   and puts it in a format that can be used by our exploring functions.
 #' @export
 #' @examples
 #' ## For icm
@@ -65,7 +69,7 @@ fortify_aggregate.icm <- function(data,
     agg_df$t <- t
     agg_df <- agg_df %>%
         dplyr::select(.data$t, dplyr::everything())
-    
+
     return(agg_df)
 }
 
@@ -96,12 +100,12 @@ extract_icm_cols <- function(nm, ii, epi){
          df <- tidyr::pivot_longer(as.data.frame(mat),
                                    cols = tidyr::everything(),
                                    names_to = "sim",
-                                   values_to = "X") %>%
[…]

### 2020-03-06 R/fortify-aggregate-pomp-list.R

@@ -1,13 +1,13 @@
-
-
-
-
-#' Takes in data from the R pomp package  where the output is a data frame and puts it in SIR format for timeternR
+#' Takes in data from the R pomp package  where the output is a data frame and
+#' puts it in SIR format for timeternR
 #'
-#' @param data Output from a pomp simulation where the output is a data frame, \code{pomp::simulate()}
+#' @param data Output from a pomp simulation where the output is a data frame,
+#'   \code{pomp::simulate()}
 #' @param states vector of state names
-#' @param package_source optional package name 
-#' @details The default variables that are retained are SIR, but can be modified with the \code{states} argument.  If code{states = NULL}, we will attempt to find all single letter names in POMP and output those.
+#' @param package_source optional package name
+#' @details The default variables that are retained are SIR, but can be modified
+#'   with the \code{states} argument.  If code{states = NULL}, we will attempt
+#'   to find all single letter names in POMP and output those.
 #' @return data frame with the following columns
 #' \describe{
 #' \item{t}{the time}

### 2020-03-06 R/fortify-aggregate-pomp-pomp-df.R

@@ -1,12 +1,14 @@
-
-
-
-#' Takes in data from the R pomp package  where the output is a data frame and puts it in SIR format for timeternR
+#' Takes in data from the R pomp package  where the output is a data frame and
+#' puts it in SIR format for timeternR
 #'
-#' @param data Output from a pomp simulation where the output is a data frame, \code{pomp::simulate()}, and we have added the name pomp_df to the class names.
+#' @param data Output from a pomp simulation where the output is a data frame,
+#'   \code{pomp::simulate()}, and we have added the name pomp_df to the class
+#'   names.
 #' @param states vector of state names
-#' @param package_source optional package name 
-#' @details The default variables that are retained are SIR, but can be modified with the \code{states} argument.  If code{states = NULL}, we will attempt to find all single letter names in POMP and output those.
+#' @param package_source optional package name
+#' @details The default variables that are retained are SIR, but can be modified
+#'   with the \code{states} argument.  If code{states = NULL}, we will attempt
+#'   to find all single letter names in POMP and output those.
 #' @return data frame with the following columns
 #' \describe{
 #' \item{t}{the time}
@@ -33,7 +35,7 @@ fortify_aggregate.pomp_df <- function(data,
         dplyr::select(dplyr::one_of(c("t", "sim", states))) %>%
         dplyr::mutate(sim = factor(.data$sim, ordered = FALSE)) %>%
         dplyr::arrange(dplyr::desc(-as.numeric(.data$sim)))
-    
+
     colnames(out)[-c(1:2)] <- paste0("X", 0:(ncol(out)-3))
 
 
@@ -42,5 +44,5 @@ fortify_aggregate.pomp_df <- function(data,
     return(out)
 
 
- 
+
 }

### 2020-03-06 R/fortify-aggregate-pomp-pompList.R

@@ -1,12 +1,13 @@
-
-
-
-#' Takes in data from the R pomp package  where the output is a data frame and puts it in SIR format for timeternR
+#' Takes in data from the R pomp package  where the output is a data frame and
+#' puts it in SIR format for timeternR
 #'
-#' @param data Output from a pomp simulation where the output is a data frame, \code{pomp::simulate()}
+#' @param data Output from a pomp simulation where the output is a data frame,
+#'   \code{pomp::simulate()}
 #' @param states vector of state names
-#' @param package_source optional package name 
-#' @details The default variables that are retained are SIR, but can be modified with the \code{states} argument.  If code{states = NULL}, we will attempt to find all single letter names in POMP and output those.
+#' @param package_source optional package name
+#' @details The default variables that are retained are SIR, but can be modified
+#'   with the \code{states} argument.  If code{states = NULL}, we will attempt
+#'   to find all single letter names in POMP and output those.
 #' @return data frame with the following columns
 #' \describe{
 #' \item{t}{the time}
@@ -24,9 +25,9 @@ fortify_aggregate.pompList <- function(data,
                                            package_source = NULL){
     pomp_output <- data
     df <- as.data.frame(pomp_output)
-    class(df) <- c("pomp_df", class(df)) 
+    class(df) <- c("pomp_df", class(df))
     out <- fortify_aggregate.pomp_df(df, states = states)
 
- 
+
 }
 

### 2020-03-06 R/fortify-aggregate.R

@@ -1,31 +1,32 @@
 ## March 3, 2020
+# Comments for Ben and Shannon
+#
+# This is the start of a reformat of fortify_epimodel and fortify_pomp into one
+# function "fortify_aggregate".  This will take in aggregate level data from
+# external sources (currently pomp and epimodel) and output, if possible,
+# output a data.frame  with the columns t, X0, .., XK where Xk are the state
+# names and the values in Xk are the numbers in state k at time t.
 
-#' Comments for Ben and Shannon
-#' 
-#' This is the start of a reformat of fortify_epimodel and fortify_pomp into one function
-#' "fortify_aggregate".  This will take in aggregate level data from external sources (currently pomp and
-#' epimodel) and output, if possible, output a data.frame  with the columns t, X0, .., XK where Xk are the
-#' state names and the values in Xk are the numbers in state k at time t.
-#'
-#'
-#'
-#' 
 
 
 #'  Take external aggregate data and put it in a format used in this package
 #'
 #' @param data output from external source package.  See details
 #' @param states names of states we want aggregate totals of at each time
-#' @param package_source optional argument to include the package from which the output is derived from, which helps with the fortify function when outputs are of generic classes such as list or data.frame
+#' @param package_source optional argument to include the package from which the
+#'   output is derived from, which helps with the fortify function when outputs
+#'   are of generic classes such as list or data.frame
 #' @return a data frame with the following columns
 #' \describe{
 #' \item{t}{time}
 #' \item{Xk}{columns X0, ..., X_K. which are numeric}
 #' }
-#' @details This function converts external data sources (we currently support output from the EpiModel and pomp R packages), which is already aggregated and puts it in a format that can be used by our exploring functions.
+#' @details This function converts external data sources (we currently support
+#'   output from the EpiModel and pomp R packages), which is already aggregated
+#'   and puts it in a format that can be used by our exploring functions.
 #' @export
 #' @examples
-#' ## TODO 
+#' ## TODO
 fortify_aggregate <- function(data,
                               states = NULL,
                               package_source = NULL){

### 2020-03-06 R/sir-array-to-df.R

@@ -9,7 +9,10 @@
 
 #' Convert simulation output from simulate_SIR_agents to data frame format
 #'
-#' @param arr array of dimension \eqn{\# agents \times \# sims \times 3} where agent ijk is the ith agent in simulation j, and looking at the kth column where 1 is the initial state, 1 is the time of infection, and 2 is the time of recovery
+#' @param arr array of dimension \eqn{\# agents \times \# sims \times 3} where
+#'   agent ijk is the ith agent in simulation j, and looking at the kth column
+#'   where 1 is the initial state, 1 is the time of infection, and 2 is the time
+#'   of recovery
 #' @return data.frame of dimension (n_agents x n_sims) x 4 where the columns are
 #' \describe{
 #' \item{agent_id}{unique ID}
@@ -44,7 +47,7 @@ sim_arr_to_df <- function(arr){
                                                   0, .data$tI),
                                       tR = ifelse(.data$init_state == 2,
                                                   0, .data$tR),
-                                      sim = factor(sim)) %>%
+                                      sim = factor(.data$sim)) %>%
         dplyr::select(.data$agent_id, .data$sim,
                       .data$tI,
                       .data$tR)

### 2020-03-10 README.Rmd
minor clean up of readme (more to check travis)

@@
 ## Notes:
-1. For writing code that works with `tidyverse` 1.0 vs `tidyverse` <= 0.8.3 see:
-https://tidyr.tidyverse.org/articles/in-packages.html, and for doing both with Travis CI (using a matrix), see https://github.com/tidyverse/design/issues/95. See Wiki for more comments on this.
-    a. Also https://principles.tidyverse.org/changes-multivers.html has been being updated to give better approaches as well.
+1.
+For writing code that works with `tidyverse` 1.0 vs `tidyverse` <= 0.8.3.
+We followed ideas found in [tidyr: in-packages](https://tidyr.tidyverse.org/articles/in-packages.html), for the code, and - when working with Travis CI (using a matrix for multiple builds) - we leverage ideas in [tidyverse travis on github](https://github.com/tidyverse/design/issues/95) and [tidyverse principles](https://principles.tidyverse.org/changes-multivers.html).
 2. For writing your own `geom`s and `stat`s that works with `ggtern` (which are generally restricted), the following 2 stack-exchange articles helped use do so with ease:
-    a. https://stackoverflow.com/questions/57175114/ternary-plot-scaling-opacity-across-groups
-    b. https://stackoverflow.com/questions/31589479/how-to-fix-no-item-called-packagepkg-on-the-search-list-without-using-libra#comment51172891_31589479
+    a. [stack-exchange: personal code idea](https://stackoverflow.com/questions/57175114/ternary-plot-scaling-opacity-across-groups)
+    b.
+[stack-exchange: being able to access ggtern's element write away](https://stackoverflow.com/questions/31589479/how-to-fix-no-item-called-packagepkg-on-the-search-list-without-using-libra#comment51172891_31589479)
     c.
-we've also leveraged ideas from https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`).
+Finally, we've also leveraged ideas from [R-devel: on avoiding problems with `:::`](https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html) in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`).
 For some reason - when documenting for `pkg-down` website, we need to do `library(ggtern); timeternR:::update_approved_layers()`

### 2020-03-10 README.Rmd
readme correct for everything and higher dpi (120 >> 72). minor correct for change of internal function name

@@
   fig.path = "man/figures/README-",
-  out.width = "100%"
+  out.width = "100%",
+  dpi = 120
 )
@@
 
-```{r eval = F}
+```{r}
 n_sims <- 100
@@
 init_SIR <- c(950, 50, 0)
-output_format <- "data.frame"
 
@@
                            beta = beta, gamma = gamma,
-                           init_SIR = init_SIR,
-                           output_format = output_format)
+                           init_SIR = init_SIR)
 
 df_groups <- out %>% dplyr::group_by(sim) %>%
-    agents_to_aggregate_SIR
+    agents_to_aggregate(states = c(tI, tR)) %>%
+    rename(S = X0, I = X1, R = X2)
 

### 2020-03-10 R/bands.R

@@ -317,7 +317,7 @@ StatConfBandDeltaBall <- ggplot2::ggproto("StatConfBandDeltaBall",
     data2d_list <- split(x = data2d, f = data2d$sim_group)
     xy_position <- which(names(data2d_list[[1]]) %in% c("x","y"))
 
-    dist_mat <- dist_matrix_innersq(data2d_list,
+    dist_mat <- dist_matrix_innersq_direction(data2d_list,
                                     position = xy_position,
                                     verbose = FALSE)
     data_deep_points <- depth_curves_to_points(data2d_list,
@@ -489,7 +489,7 @@ StatConfBandConvexHull <- ggplot2::ggproto("StatConfBandConvexHull",
         data2d_list <- split(x = data2d, f = data2d$sim_group)
         xy_position <- which(names(data2d_list[[1]]) %in% c("x","y"))
 
-        dist_mat <- dist_matrix_innersq(data2d_list,
+        dist_mat <- dist_matrix_innersq_direction(data2d_list,
                                         position = xy_position,
                                         verbose = FALSE)
         data_deep_points <- depth_curves_to_points(

### 2020-03-11 README.Rmd
containment and band creation outside ternary visual. convex hull has problems - need to project to simplex space (which is annoying - especially given that we haven't been thinking about that project for the delta ball approach).

@@
   out.width = "100%",
-  dpi = 120
+  dpi = 300
 )
@@
 Finally, we've also leveraged ideas from [R-devel: on avoiding problems with `:::`](https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html) in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`).
-For some reason - when documenting for `pkg-down` website, we need to do `library(ggtern); timeternR:::update_approved_layers()`
+For some reason - when documenting for `pkgdown` website, we need to do `library(ggtern); timeternR:::update_approved_layers()`
 

### 2020-03-11 R/containment-and-band-creation.R

@@ -1,0 +1,351 @@
+# These functions will need to be move around (but probably should think about
+# if there are redundant functions now)
+
+
+#' Compress filaments to filaments with the same number of points (equally
+#' linearly space compared to original filament definition)
+#'
+#' @param grouped_df grouped_df data.frame object (assumed rows per filament are
+#'   ordered) - grouped per each filament
+#' @param data_columns columns (currently strings) of data.frame that relate to
+#'   the filament's coordinates in euclidean space.
+#' @param number_points integer number of points for each filament to be
+#'   compressed to
+#'
+#' @return updated grouped_df with new rows so that each filament has the same
+#'   number of points.
+#' @export
+#'
+#' @examples
+#' library(dplyr)
+#' t13compression <- timeternR::pomp_sir %>%
+#'   arrange(time) %>% # just to be safe
+#'   select(-H, -cases, -time) %>%
+#'   filter(.id <= 5) %>%
+#'   group_by(.id) %>%
+#'   filament_compression()
+#'
+#' t9compression <- timeternR::pomp_sir %>%
+#'   filter(.id <= 5) %>%
+#'   group_by(.id) %>%
+#'   filament_compression(data_columns = c("S","I","R"), number_points = 9)
+filament_compression <- function(grouped_df, data_columns = NULL,
+                                 number_points = 13){
+  if (is.null(data_columns)){ # use all columns except the grouped columns
+    data_columns <- names(grouped_df)
+    group_columns <- names(attr(grouped_df, "groups"))[names(attr(grouped_df, "groups")) != ".rows"]
+    data_columns <- data_columns[!(data_columns %in% group_columns)]
+  }
+
+  if (tidyr_new_interface()){
+    compression_df <- grouped_df %>%
+      tidyr::nest() %>%
+      dplyr::mutate(data =
+                      purrr::map(.data$data,
+                                 function(df) {equa_dist_points_direction(
+                                   df[, data_columns],num_splits = number_points)})) %>%
+      tidyr::unnest(cols = c(.data$data))
+  } else {
+    compression_df <- grouped_df %>%
+      tidyr::nest() %>%
+      dplyr::mutate(data =
+                      purrr::map(.data$data,
+                                 function(df) {equa_dist_points_direction(
+                                   df[, data_columns],num_splits = number_points)})) %>%
[…]

### 2020-03-11 tests/testthat/test-delta-ball.R

@@ -36,7 +36,6 @@ test_that("delta_structure tests", {
   testthat::expect_true(all(line_info))
 })
 
-
 test_that("get_delta basic tests", {
   # static
   data <- data.frame(x = c(0,1), y = c(0,1))

### 2020-03-17 README.Rmd
attempt to correct visualization error (should pass tests and "visually" look correct). Still not correct...

@@
 
-df_groups %>% ggplot(aes(x = S, y = I, z = R, group = sim)) +
-    geom_line(alpha = .1) +
+df_groups %>% ggplot() +
+    geom_prediction_band(aes(x = S, y = I, z = R, sim_group = as.numeric(sim)),
+                         alpha = .2, fill = "blue", color = "blue") +
+    geom_line(aes(x = S, y = I, z = R, group = sim), alpha = .1) +
     coord_tern()

### 2020-03-17 R/aaa.R

@@ -1,11 +1,11 @@
 # List of new stats
 .newstat = c(sir_aggregate        = "StatSirAggregate",
-             conf_band_kde        = "StatConfBandKDE",
-             conf_band_delta_ball = "StatConfBandDeltaBall",
-             conf_band_spherical  = "StatConfBandSpherical",
-             conf_band_chull      = "StatConfBandConvexHull")
+             conf_band_kde        = "StatPredBandKDE",
+             conf_band_delta_ball = "StatPredBandDeltaBall",
+             conf_band_spherical  = "StatPredBandSpherical",
+             conf_band_chull      = "StatPredBandConvexHull")
 .newgeom = c(sir       = "GeomSirAggregate",
-             conf_band = "GeomConfBand")
+             conf_band = "GeomPredBand")
 
 #' import hidden function / variable from other package
 #'

### 2020-03-17 R/geom-aggregate.R

@@ -29,5 +29,4 @@ geom_aggregate <- function(mapping = NULL, data = NULL,
 #' @usage NULL
 #' @export
 GeomSirAggregate <- ggplot2::ggproto("GeomSirAggregate",
-                                     ggplot2::GeomPath,
-                                     handle_na = function(data, params){data})
+                                     ggplot2::GeomPath)

### 2020-03-17 R/kde.R

@@ -110,8 +110,8 @@ fit_kde_object = function(df, h_band = NULL, position = 1:2,
 #' }
 #' @export
 extract_contour <- function(kde_obj, alpha) {
-  # contour lines actually does 1-alpha
-  alpha <- round((1-alpha)*100)
+  # contour lines actually does alpha
+  alpha <- round(alpha*100)
 
   cont_level <- paste0(as.character(alpha), "%")
   cont <- with(kde_obj, contourLines(x = eval.points[[1]],y = eval.points[[2]],

### 2020-03-17 R/stat-aggregate.R

@@ -4,25 +4,35 @@
 #' @format NULL
 #' @usage NULL
 #' @export
-StatSirAggregate <- ggplot2::ggproto("StatSirAggregate", ggplot2::Stat,
-                                     setup_data = function(data, params){
-                                       data  # This is so ggplot does NOT remove NA rows
-                                     },
-                                     compute_group = function(data, scales){
-                                       # saving panel and group info
-                                       info_inner <- data[, c("PANEL", "group")] %>%
-                                         sapply(unique)
+StatSirAggregate <- ggplot2::ggproto("StatSirAggregate",
+                      ggplot2::Stat,
+                      setup_data = function(data, params){
+                        # This is so ggplot does NOT remove NA rows
+                         inner_data <- data
+                         inner_data$missing_y <- data$y
+                         inner_data$missing_z <- data$z
+                         inner_data$y <- 100
+                         inner_data$z <- 100
 
-                                       out <- agents_to_aggregate(data,
-                                                                  states = c(.data$y,.data$z))
+                         return(inner_data)
+                       },
+                      compute_group = function(data, scales){
+                         # saving panel and group info
+                         info_inner <- data[, c("PANEL", "group")] %>%
+                           sapply(unique)
 
-                                       out <- out %>% dplyr::mutate(PANEL = info_inner[1],
-                                                                    group = info_inner[2])
-                                       names(out)[names(out) %in% c("X0","X1","X2")] <-
-                                         c("x", "y", "z")
-                                       return(out)
-                                     },
-                                     required_aes = c("y", "z"))
+                         out <- agents_to_aggregate(data,
+                                   states = c(.data$missing_y,
+                                              .data$missing_z))
+
+                         out <- out %>% dplyr::mutate(PANEL = info_inner[1],
+                                                      piece = info_inner[2],
+                                                      group = info_inner[2])
+                         names(out)[names(out) %in% c("X0","X1","X2")] <-
+                           c("x", "y", "z")
+                         return(out)
+                       },
+                      required_aes = c("y", "z"))
 
 
 #'aggregate SIR path visuals from agent data

### 2020-03-17 tests/testthat/test-agent-to-aggregate.R

@@ -693,4 +693,96 @@ test_that("agents_to_aggregate.group_df -string passes basic checks", {
                          sir_group_1 %>% ungroup %>% select(t, X0, X1, X2))
 })
 
+test_that("agents_to_aggregate.group_df, max_t = NA basic checks", {
+  suppressWarnings(
+    suppressPackageStartupMessages(
+      library(dplyr, quietly = TRUE)
+    ))
+  # similar test as in the example string
+  new_raw <- hagelloch_raw %>%
+    mutate(AGE2 = as.numeric(cut(AGE,3)))
+  agents_g <- new_raw %>% group_by(AGE2)
 
+  max_time_each <- agents_g %>% summarize(nmax = ceiling(max(tR)))
+
+  max_time <- max_time_each %>% pull(nmax) %>% max()
+
+  sir_group <- agents_to_aggregate(agents_g, states = c(tI, tR),
+                                   min_max_time = c(0, NA))
+
+  # should all share max value if min_max_time[2] = NA
+  testthat::expect_equal(sum((sir_group %>% summarize(maxt = max(t)) %>%
+                           pull(maxt)) == max_time), 3)
+
+  for (age2_value in 1:3){
+    agents <- agents_g %>%
+      filter(AGE2 == age2_value) %>% ungroup()
+    sir_group_sub_id <- agents_to_aggregate(agents, states = c(tI, tR),
+                                            min_max_time = c(0, NA))
+    sir_group_sub_g <- sir_group %>% filter(AGE2 == age2_value)
+
+    individual_max_t <- max_time_each$nmax[max_time_each$AGE2 == age2_value]
+    testthat::expect_equal(sir_group_sub_id %>% pull(t) %>% max,
+                           individual_max_t)
+
+    testthat::expect_equal(sir_group_sub_id,
+                           sir_group_sub_g %>% ungroup %>%
+                             select(t, X0, X1, X2) %>%
+                             filter(t <= individual_max_t))
+
+    if (individual_max_t < max_time){
+      unique_rows_after_max <- sir_group_sub_g %>% ungroup %>%
+        filter(t > individual_max_t) %>%
+        select(X0, X1, X2)  %>% distinct() %>% nrow()
+      testthat::expect_equal(unique_rows_after_max, 1)
+    }
+  }
+
+})
+
+test_that("agents_to_aggregate.group_df, max_t = NA basic checks, strings", {
+  suppressWarnings(
[…]

### 2020-03-17 tests/testthat/test-containment-and-band-creation.R

@@ -331,7 +331,7 @@ test_that("filament_distance_depth correct depth, tidified",{
 })
 
 test_that(paste0("test grab_top_depth_filaments (.remove_group = both",
-          "no data_columns)"), {
+          ", no data_columns)"), {
   top_filaments <- timeternR::pomp_df %>% group_by(.id) %>%
     filter(.id <= 10) %>% select(-time, -H, -cases) %>%
     grab_top_depth_filaments(alpha_level = .5,
@@ -432,6 +432,58 @@ test_that(paste0("test grab_top_depth_filaments (.remove_group = both",
                              alpha_level = 1))
 
 })
+
+
+test_that(paste0("test grab_top_depth_filaments (.remove_group = both",
+                 "all column options), error when none returned."), {
+                   # no column names
+                   testthat::expect_error(
+                     top_filaments <- timeternR::pomp_df %>%
+                       group_by(.id) %>%
+                       filter(.id <= 10) %>% select(-time, -H, -cases) %>%
+                       grab_top_depth_filaments(alpha_level = 1,
+                                                .remove_group = FALSE))
+
+                   testthat::expect_error(
+                     top_filaments_points <- timeternR::pomp_df %>%
+                       group_by(.id) %>%
+                       filter(.id <= 10) %>% select(-time, -H, -cases) %>%
+                       grab_top_depth_filaments(alpha_level = 1))
+
+                   # string column names
+                   testthat::expect_error(
+                     top_filaments <- timeternR::pomp_df %>%
+                       group_by(.id) %>%
+                       filter(.id <= 10) %>% select(-time, -H, -cases) %>%
+                       grab_top_depth_filaments(alpha_level = 1,
+                                                .remove_group = FALSE,
+                                                data_columns = c("S","I","R")
+                                                ))
+
+                   testthat::expect_error(
+                     top_filaments_points <- timeternR::pomp_df %>%
+                       group_by(.id) %>%
+                       filter(.id <= 10) %>% select(-time, -H, -cases) %>%
+                       grab_top_depth_filaments(alpha_level = 1,
+                                                data_columns = c("S","I","R")))
+                   # promise column names
+                   testthat::expect_error(
+                     top_filaments <- timeternR::pomp_df %>%
+                       group_by(.id) %>%
+                       filter(.id <= 10) %>% select(-time, -H, -cases) %>%
+                       grab_top_depth_filaments(alpha_level = 1,
+                                                .remove_group = FALSE,
[…]

### 2020-03-17 tests/testthat/test-geom-aggregate.R

@@ -41,3 +41,23 @@ test_that("check geom_aggregate for raw works correctly with groups", {
   testthat::expect_true(all(data_vis[,c("x", "y", "z")] >= 0))
 
 })
+
+test_that("check geom_aggregate for raw works correctly with NAs", {
+  library(ggplot2)
+  set.seed(1)
+  pretend_actual_data <- timeternR::simulate_SIR_agents(
+    n_sims = 1,
+    n_time_steps = 1000,
+    beta = .0099, gamma = .0029,
+    init_SIR = c(950, 50, 0))
+
+  # just making sure it has NA values...
+  testthat::expect_true(sum(is.na(pretend_actual_data[, c("tI", "tR")])) > 0)
+
+  geom_agg_vis_with_na <- pretend_actual_data %>%
+    ggplot() +
+    geom_aggregate(aes(y = tI, z = tR)) +
+    coord_tern()
+  # tells us below if there are removals (which there shouldn't be...)
+  testthat::expect_silent(layer_data(geom_agg_vis_with_na))
+})

### 2020-03-17 tests/testthat/test-kde.R

@@ -58,10 +58,8 @@ test_that("test extract_contour - basic static example", {
   cont_2 <- extract_contour(kde_object, .2)
 
   # note that kde objects store the values relative to an increasing value
-  expect_equal(cont_05[[1]]$level, as.numeric(kde_object$cont["95%"]))
-  expect_equal(cont_1[[1]]$level, as.numeric(kde_object$cont["90%"]))
-  expect_equal(cont_2[[1]]$level, as.numeric(kde_object$cont["80%"]))
-
-
+  expect_equal(cont_05[[1]]$level, as.numeric(kde_object$cont["5%"]))
+  expect_equal(cont_1[[1]]$level, as.numeric(kde_object$cont["10%"]))
+  expect_equal(cont_2[[1]]$level, as.numeric(kde_object$cont["20%"]))
 })
 

### 2020-03-17 README.Rmd
Changed every instance of timeternR to EpiCompare

@@
 <!-- badges: start -->
-[![Travis build status](https://travis-ci.org/skgallagher/timeternR.svg?branch=master)](https://travis-ci.org/skgallagher/timeternR)
-[![codecov](https://codecov.io/gh/skgallagher/timeternR/branch/master/graph/badge.svg)](https://codecov.io/gh/skgallagher/timeternR)
+[![Travis build status](https://travis-ci.org/skgallagher/EpiCompare.svg?branch=master)](https://travis-ci.org/skgallagher/EpiCompare)
+[![codecov](https://codecov.io/gh/skgallagher/EpiCompare/branch/master/graph/badge.svg)](https://codecov.io/gh/skgallagher/EpiCompare)
 <!-- badges: end -->
 
-The goal of timeternR is to develop functional based visualization and statistics that focus on use in ternary plots.
+The goal of EpiCompare is to develop functional based visualization and statistics that focus on use in ternary plots.
 
@@
 
-You can install the developmental version of timeternR from github using:
+You can install the developmental version of EpiCompare from github using:
 
@@
 # install.packages("devtools")
-devtools::install_github("skgallagher/timeternR")
+devtools::install_github("skgallagher/EpiCompare")
 ```
@@
 
-   Currently there are 13 data sets in which to explore the different `timeternR` functions.
+   Currently there are 13 data sets in which to explore the different `EpiCompare` functions.
 These correspond to different sources of raw data along, agent vs.
@@
 library(ggtern)
-library(timeternR)
+library(EpiCompare)
 ```
@@
 Finally, we've also leveraged ideas from [R-devel: on avoiding problems with `:::`](https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html) in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`).
-For some reason - when documenting for `pkgdown` website, we need to do `library(ggtern); timeternR:::update_approved_layers()`
+For some reason - when documenting for `pkgdown` website, we need to do `library(ggtern); EpiCompare:::update_approved_layers()`
 

### 2020-03-17 R/bands.R

@@ -743,17 +743,17 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #' @examples
 #' library(ggplot2)
 #' library(dplyr)
-#' library(ggtern); timeternR:::update_approved_layers()
+#' library(ggtern); EpiCompare:::update_approved_layers()
 #' #                ^ this doesn't generally need to be done
 #'
-#' vis_data <- timeternR::pomp_df %>%
+#' vis_data <- EpiCompare::pomp_df %>%
 #'   rename(x = "S", y = "I", z = "R") %>%
 #'   ggplot(aes(x = x, y =y, z = z, group = .id)) +
 #'   geom_path(alpha = .03) +
 #'   coord_tern() +
 #'   labs(title = "Actually data paths")
 #'
-#' vis_spherical <- timeternR::pomp_df %>%
+#' vis_spherical <- EpiCompare::pomp_df %>%
 #'   rename(x = "S", y = "I", z = "R", t = "time") %>%
 #'   ggplot(aes(x = x, y = y, z = z, t = t)) +
 #'   geom_prediction_band(pb_type = "spherical_ball",
@@ -762,7 +762,7 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #'   coord_tern() +
 #'   labs(title = "Spherical CB")
 #'
-#' vis_delta_ball <- timeternR::pomp_df %>%
+#' vis_delta_ball <- EpiCompare::pomp_df %>%
 #'   rename(x = "S", y = "I", z = "R") %>%
 #'   mutate(.id = as.numeric(.id)) %>%
 #'   ggplot(aes(x = x, y = y, z = z, sim_group = .id)) +
@@ -772,7 +772,7 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #'   coord_tern() +
 #'   labs(title = "Delta-ball CB")
 #'
-#' vis_kde <- timeternR::pomp_df %>%
+#' vis_kde <- EpiCompare::pomp_df %>%
 #'   rename(x = "S", y = "I", z = "R") %>%
 #'   mutate(.id = as.numeric(.id)) %>%
 #'   ggplot(aes(x = x, y = y, z = z, sim_group = .id)) +
@@ -782,7 +782,7 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
[…]

### 2020-03-17 R/containment-and-band-creation.R

@@ -20,14 +20,14 @@
 #'
 #' @examples
 #' library(dplyr)
-#' t13compression <- timeternR::pomp_sir %>%
+#' t13compression <- EpiCompare::pomp_sir %>%
 #'   arrange(time) %>% # just to be safe
 #'   select(-H, -cases, -time) %>%
 #'   filter(.id <= 5) %>%
 #'   group_by(.id) %>%
 #'   filament_compression()
 #'
-#' t9compression <- timeternR::pomp_sir %>%
+#' t9compression <- EpiCompare::pomp_sir %>%
 #'   filter(.id <= 5) %>%
 #'   group_by(.id) %>%
 #'   filament_compression(data_columns = c("S","I","R"), number_points = 9)
@@ -163,13 +163,13 @@ filament_distance_depth <- function(grouped_df,
 #'
 #' @examples
 #' library(dplyr)
-#' top_filaments <- timeternR::pomp_df %>% group_by(.id) %>%
+#' top_filaments <- EpiCompare::pomp_df %>% group_by(.id) %>%
 #'   filter(.id <= 10) %>%
 #'   grab_top_depth_filaments(data_columns =c("S","I","R"),
 #'                            alpha_level = .5)
 #'
 #' # note that the below doesn't contain all filaments - as some have 0 depth
-#' all_but_extreme_filaments <- timeternR::pomp_df %>% group_by(.id) %>%
+#' all_but_extreme_filaments <- EpiCompare::pomp_df %>% group_by(.id) %>%
 #'   filter(.id <= 10) %>%
 #'   grab_top_depth_filaments(data_columns = c("S","I","R"),
 #'                            alpha_level = 0)
@@ -252,7 +252,7 @@ grab_top_depth_filaments <- function(grouped_df, data_columns = NULL,
 #'
 #' @examples
 #' library(dplyr)
-#' delta_ball_cb <- timeternR::pomp_df %>%
+#' delta_ball_cb <- EpiCompare::pomp_df %>%
 #'   filter(.id <= 10) %>%
 #'   arrange(time) %>% # just to be safe
 #'   select(-time, -H, -cases) %>%
@@ -323,7 +323,7 @@ create_delta_ball_structure <- function(data_points, data_columns = NULL,
 #' library(dplyr)
 #' # this will look like a 2d data frame (but this is becuase pomp_df info is
 #' # on the 3d simplex)
[…]

### 2020-03-17 R/fortify-aggregate-data-frame.R

@@ -1,5 +1,5 @@
 #' Takes in data from the R pomp package where the output is a data frame and
-#' puts it in SIR format for timeternR
+#' puts it in SIR format for EpiCompare
 #'
 #' @param data Output from a pomp simulation where the output is a data frame,
 #'   \code{pomp::simulate()}

### 2020-03-17 R/fortify-aggregate-pomp-pomp-df.R

@@ -1,5 +1,5 @@
 #' Takes in data from the R pomp package  where the output is a data frame and
-#' puts it in SIR format for timeternR
+#' puts it in SIR format for EpiCompare
 #'
 #' @param data Output from a pomp simulation where the output is a data frame,
 #'   \code{pomp::simulate()}, and we have added the name pomp_df to the class

### 2020-03-17 R/timeternR.R

@@ -1,6 +1,6 @@
-#' timeternR: Ternary plot visualization for SIR curves (UPDATE)
+#' EpiCompare: Ternary plot visualization for SIR curves (UPDATE)
 #'
-#' The goal of timeternR is to develop functional based visualization and
+#' The goal of EpiCompare is to develop functional based visualization and
 #' statistics that focus on use in ternary plots. (UPDATE)
 #'
 #' This package attempts to:
@@ -35,5 +35,5 @@
 #' @importFrom rlang .data
 #'
 #' @docType package
-#' @name timeternR
+#' @name EpiCompare
 NULL

### 2020-03-17 tests/testthat/test-geom-aggregate.R

@@ -26,7 +26,7 @@ test_that("check geom_aggregate for raw works correctly with groups", {
                         label = paste("*S should always be decreasing*:",
                                       "all(diff(sir_out$S) <= -1)"))
   # multiple groups:
-  vis <- timeternR::hagelloch_raw %>%
+  vis <- EpiCompare::hagelloch_raw %>%
     dplyr::filter(SEX %in% c("male", "female")) %>%
     ggplot(., aes(y = tI, z = tR, color = SEX)) +
     geom_path(stat = StatSirAggregate) + ggtern::coord_tern() +
@@ -45,7 +45,7 @@ test_that("check geom_aggregate for raw works correctly with groups", {
 test_that("check geom_aggregate for raw works correctly with NAs", {
   library(ggplot2)
   set.seed(1)
-  pretend_actual_data <- timeternR::simulate_SIR_agents(
+  pretend_actual_data <- EpiCompare::simulate_SIR_agents(
     n_sims = 1,
     n_time_steps = 1000,
     beta = .0099, gamma = .0029,

### 2020-03-17 README.Rmd
pkgdown website should be working

@@
 
-# EpiCompare <img src="man/figures/logo.png" align="right" alt="" width="120" />
+# EpiCompare <a href = "https://skgallagher.github.io/EpiCompare/"><img src="man/figures/logo.png" align="right" alt="" width="120" /></a>
 
@@
 The goal of EpiCompare is to develop functional based visualization and statistics that focus on use in ternary plots.
+
+## Documentation
+
+Please see our website at [https://skgallagher.github.io/EpiCompare/](https://skgallagher.github.io/EpiCompare/).
 

### 2020-03-17 README.Rmd
correction to bands to get the correct visual display (also testing). vignette now takes less time to build

--- 2020-03-17 README.Rmd
@@
 <!-- badges: start -->
-[![Travis build status](https://travis-ci.org/skgallagher/timeternR.svg?branch=master)](https://travis-ci.org/skgallagher/timeternR)
-[![codecov](https://codecov.io/gh/skgallagher/timeternR/branch/master/graph/badge.svg)](https://codecov.io/gh/skgallagher/timeternR)
+[![Travis build status](https://travis-ci.org/skgallagher/EpiCompare.svg?branch=master)](https://travis-ci.org/skgallagher/EpiCompare)
+[![codecov](https://codecov.io/gh/skgallagher/EpiCompare/branch/master/graph/badge.svg)](https://codecov.io/gh/skgallagher/EpiCompare)
 <!-- badges: end -->
@@
 For some reason - when documenting for `pkgdown` website, we need to do `library(ggtern); timeternR:::update_approved_layers()`
+3.
+`geom_prediction_band` required not just `compute_group` but `compute_layer` - there is very little documentation on how to approach this correctly.
+Basically - there are problems when the `compute_group` wants to make multiple `pieces`/`groups` - and it is similar to the problem that if you do something like `aes(color= var1, group =var2)` you may actually want to do `aes(color= var1, group =paste(var1, var2)`, if there are the same `var2` values across different `var1` values but they don't mean they should be grouped together.
 

### 2020-03-17 R/bands.R

@@ -282,24 +282,34 @@ get_grid_elipsoid_containment <- function(inside_func_list,
 #' stat object for use in kde based stat_prediction_band and
 #' geom_prediction_band
 #' @export
-StatPredBandKDE <- ggplot2::ggproto("StatPredBandKDE",
-                                    ggplot2::Stat,
-                                    compute_group =
-  function(data, scales, params,
-           pb_type = NULL,
-           #^ needed to match same format as stat/geom_prediction_band
-           grid_size = rep(100,2),
-           conf_level = .9, over_delta = NULL){
 
+StatPredBandKDE <- ggplot2::ggproto("StatPredBandKDE",
+  ggplot2::Stat,
+  compute_layer = function(self, data, params, layout){
+    # first run the regular layer calculation to infer densities
+    data <- ggplot2::ggproto_parent(ggplot2::Stat,self = self)$compute_layer(data, params, layout)
+
+    # required piece and group to be cleaned up
+    data_cleaned_up <- data %>% mutate(piece_old = .data$piece,
+                                       group_old = .data$group,
+                                       piece = as.numeric(
+                                         factor(paste(.data$PANEL,
+                                                      .data$piece_old,
+                                                      .data$group_old,
+                                                      sep = "*"))),
+                                       group = piece)
+
+    return(data_cleaned_up)
+  },
+  compute_group = function(data, scales, params,
+                           pb_type = NULL,
+                           #^ needed to match same format as stat/geom_prediction_band
+                           grid_size = rep(100,2),
+                           conf_level = .9, over_delta = NULL){
     assertthat::assert_that(!is.factor(data$sim_group),
                             msg = paste("'sim_group' cannot be a factor"))
 
-    info_inner <- data[, c("PANEL", "group")] %>%
-      sapply(unique)
-
-    if (info_inner[1] == 1 & info_inner[2] <= 1){
-      .number_contours <<- 1
-    }
+    info_inner <- data[, c("PANEL", "group")] %>% sapply(unique)
 
     data <- data %>% mutate(sim_group = factor(sim_group))
 
[…]

### 2020-03-17 R/fortify-aggregate-epimodel-icm.R
fixed merges

@@ -35,7 +35,7 @@ fortify_aggregate.icm <- function(data,
                               states = NULL,
                               package_source = NULL){
 
-
+  #  states <- enquo(states)
 
     ## Grab the state names if not specified
     if(!is.null(states)){

### 2020-03-18 R/agent-to-aggregate.R
minor updates to code a2a and bands

@@ -816,7 +816,8 @@ agents_to_aggregate.grouped_df <- function(agents,
       dplyr::select(-.data$data) %>%
       tidyr::unnest(cols = c(.data$update)) # only change
   } else {
-    # old
+    # old tidyr
+
     out <- agents %>% tidyr::nest() %>%
       dplyr::mutate(update = purrr::map(.data$data, agents_to_aggregate,
                                         states = !!!dplyr::enquos(states),
@@ -825,6 +826,19 @@ agents_to_aggregate.grouped_df <- function(agents,
                                         min_max_time = min_max_time_all)) %>%
       dplyr::select(-.data$data) %>%
       tidyr::unnest(.drop = FALSE)
+
+    # making sure output is also grouped_df
+    group_columns <- names(attr(agents, "groups"))[names(attr(agents, "groups")) != ".rows"]
+
+    if (length(group_columns) == 1) {
+      group_sym <- dplyr::sym(group_columns)
+      out <- out %>% dplyr::group_by(!!group_sym)
+    }
+    if (length(group_columns) > 1) {
+      group_sym <- dplyr::syms(group_columns)
+      out <- out %>% dplyr::group_by(!!!group_sym)
+    }
+
   }
 
   return(out)

### 2020-03-18 R/bands.R

@@ -290,7 +290,7 @@ StatPredBandKDE <- ggplot2::ggproto("StatPredBandKDE",
     data <- ggplot2::ggproto_parent(ggplot2::Stat,self = self)$compute_layer(data, params, layout)
 
     # required piece and group to be cleaned up
-    data_cleaned_up <- data %>% mutate(piece_old = .data$piece,
+    data_cleaned_up <- data %>% dplyr::mutate(piece_old = .data$piece,
                                        group_old = .data$group,
                                        piece = as.numeric(
                                          factor(paste(.data$PANEL,
@@ -731,10 +731,10 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #'
 #' \itemize{
 #' \item \code{delta_ball}: relative to all the points in the top
-#' \code{conf_level} curves, we find the minimum delta such all of these
-#' points are contained in at least 1 ball around another point with radius delta.
-#' This can be mathematically expressed as: \eqn{\delta = \max_{i} \min_{j} d(x_i, x_j)}.
-#' Then we take the union of delta-balls surround all the points as the
+#' \code{conf_level} curves, we find the minimum delta such all of these points
+#' are contained in at least 1 ball around another point with radius delta. This
+#' can be mathematically expressed as: \eqn{\delta = \max_{i} \min_{j} d(x_i,
+#' x_j)}. Then we take the union of delta-balls surround all the points as the
 #' prediction band.
 #'
 #' \item \code{convex_hull}: with to all the points in the top

### 2020-03-18 R/containment-and-band-creation.R

@@ -154,7 +154,7 @@ filament_distance_depth <- function(grouped_df,
 #' @param filament_depth_function function to calculate depth relative to the
 #'   filaments in the \code{grouped_df}. Will take in \code{data_columns}
 #'   parameter as well
-#' @param alpha_level proportion of filaments to keep
+#' @param conf_level proportion of filaments to keep
 #' @param .remove_group boolean (default true). Should the output keep the
 #' grouping columns or just havd data points?
 #'
@@ -166,16 +166,16 @@ filament_distance_depth <- function(grouped_df,
 #' top_filaments <- timeternR::pomp_df %>% group_by(.id) %>%
 #'   filter(.id <= 10) %>%
 #'   grab_top_depth_filaments(data_columns =c("S","I","R"),
-#'                            alpha_level = .5)
+#'                            conf_level = .5)
 #'
 #' # note that the below doesn't contain all filaments - as some have 0 depth
 #' all_but_extreme_filaments <- timeternR::pomp_df %>% group_by(.id) %>%
 #'   filter(.id <= 10) %>%
 #'   grab_top_depth_filaments(data_columns = c("S","I","R"),
-#'                            alpha_level = 0)
+#'                            conf_level = 1)
 grab_top_depth_filaments <- function(grouped_df, data_columns = NULL,
                                      filament_depth_function = filament_distance_depth,
-                                     alpha_level = .95,
+                                     conf_level = .95,
                                      .remove_group = TRUE){
   #quos
   data_columns_q <- dplyr::enquos(data_columns)
@@ -187,10 +187,12 @@ grab_top_depth_filaments <- function(grouped_df, data_columns = NULL,
   depth_vector <- grouped_df %>%
     filament_depth_function(data_columns = data_columns)
 
-  num_top <- sum(depth_vector > stats::quantile(depth_vector, probs = alpha_level))
+  num_top <- sum(depth_vector > stats::quantile(depth_vector,
+                                                probs = 1 - conf_level))
+  # ^conf_level = 1 - alpha
 
   if (num_top == 0){
-    stop("selected 'alpha_level' returns 0 filaments")
+    stop("selected 'conf_level' returns 0 filaments")
   }
 
   if (tidyr_new_interface()){
@@ -198,7 +200,9 @@ grab_top_depth_filaments <- function(grouped_df, data_columns = NULL,
[…]

### 2020-03-18 README.Rmd

--- 2020-03-17 README.Rmd
@@
 ```{r message = F, warning = F}
-library(tidyverse)
+library(tidyr)
+library(ggplot2)
+library(dplyr)
 library(ggtern)
@@
 [stack-exchange: being able to access ggtern's element write away](https://stackoverflow.com/questions/31589479/how-to-fix-no-item-called-packagepkg-on-the-search-list-without-using-libra#comment51172891_31589479)
+
     c.
 Finally, we've also leveraged ideas from [R-devel: on avoiding problems with `:::`](https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html) in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`).
-For some reason - when documenting for `pkgdown` website, we need to do `library(ggtern); EpiCompare:::update_approved_layers()`
+For some reason - when documenting for `pkgdown` website, we need to do `library(ggtern); timeternR:::update_approved_layers()`
+3.
+`geom_prediction_band` required not just `compute_group` but `compute_layer` - there is very little documentation on how to approach this correctly.
+Basically - there are problems when the `compute_group` wants to make multiple `pieces`/`groups` - and it is similar to the problem that if you do something like `aes(color= var1, group =var2)` you may actually want to do `aes(color= var1, group =paste(var1, var2)`, if there are the same `var2` values across different `var1` values but they don't mean they should be grouped together.
 

### 2020-03-18 R/bands.R
merging

@@ -772,7 +772,7 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #'
 
 #' # for speed purposes
-#' smaller_pomp_df <- timeternR::pomp_df %>% filter(.id < 10)
+#' smaller_pomp_df <- EpiCompare::pomp_df %>% filter(.id < 10)
 #'
 #' vis_data <- smaller_pomp_df %>%
 #'  rename(x = "S", y = "I", z = "R") %>%

### 2020-03-18 R/fortify-aggregate-epimodel-icm.R

@@ -35,10 +35,15 @@ fortify_aggregate.icm <- function(data,
                               states = NULL,
                               package_source = NULL){
 
-  #  states <- enquo(states)
+
+    ## Tidifying code
+    state_cols <- dplyr::enquos(states)
+    states <- unname(tidyselect::vars_select(names(data$epi),
+                                             !!!state_cols))
+
 
     ## Grab the state names if not specified
-    if(!is.null(states)){
+    if(length(states) == 0){
         states <- get_epimodel_icm_states(data)
     }
 

### 2020-03-18 R/fortify-aggregate-pomp-pomp-df.R

@@ -24,7 +24,17 @@
 fortify_aggregate.pomp_df <- function(data,
                                            states = c("S", "I", "R"),
                                           package_source = NULL){
-    if(is.null(states)){
+
+    browser()
+    ## pull out state names
+    if(class(states!= "quosure")){
+        state_cols <- dplyr::enquos(states)
+    }
+    states <- unname(tidyselect::vars_select(colnames(data),
+                                             !!!state_cols))
+
+
+    if(length(states) == 0){
         states <- grep("^[A-Z]{1}$", colnames(data),
                        value = TRUE)
     }

### 2020-03-18 tests/testthat/test-fortify-aggregate-list.R

@@ -6,5 +6,8 @@ test_that("fortify_aggregate.list", {
     ##
     out <- fortify_aggregate(my_list, states = c("S", "I", "R"), package_source= "pomp")
     expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
+    ##
+    out2 <- fortify_aggregate(my_list, states = c(S, I, R), package_source= "pomp")
+    expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
 
 })

### 2020-03-18 tests/testthat/test-fortify-aggregate-pomp-list.R

@@ -7,5 +7,9 @@ test_that("fortify_aggregate.pomp_list", {
     ##
     out <- fortify_aggregate(my_list, states = c("S", "I", "R"))
     expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
+    ##
+    out2 <- fortify_aggregate(my_list, states = c(S, I, R))
+    expect_equal(colnames(out2), c("t", "sim", "X0", "X1", "X2"))
+    expect_equal(out, out2)
 
 })

### 2020-03-18 tests/testthat/test-fortify-aggregate-pomp-pomp-df.R

@@ -2,14 +2,20 @@ test_that("fortify_aggregate.pomp_df()", {
 
     pomp_output <- pomp_pomp
     df <- as.data.frame(pomp_output)
-    class(df) <- c("pomp_df", class(df)) 
+    class(df) <- c("pomp_df", class(df))
     out <- fortify_aggregate.pomp_df(df, states = NULL)
     expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2", "X3"))
 ###
     pomp_output <- pomp_pomp
     df <- as.data.frame(pomp_output)
-    class(df) <- c("pomp_df", class(df)) 
+    class(df) <- c("pomp_df", class(df))
     out <- fortify_aggregate.pomp_df(df, states = c("S", "I", "R"))
     expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
-
+    ##
+    pomp_output <- pomp_pomp
+    df <- as.data.frame(pomp_output)
+    class(df) <- c("pomp_df", class(df))
+    out2 <- fortify_aggregate.pomp_df(df, states = c(S, I, R))
+    expect_equal(colnames(out2), c("t", "sim", "X0", "X1", "X2"))
+    expect_equal(out, out2)
 })

### 2020-03-18 tests/testthat/test-fortify-aggregate-pomp-pompList.R

@@ -5,5 +5,8 @@ test_that("fortify_aggregate.pompList()", {
     ##
     out <- fortify_aggregate(pomp_pomp, states = c("S", "I", "R"))
     expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
+    ##
+    out2 <- fortify_aggregate(pomp_pomp, states = c(S, I, R))
+    expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
 
 })

### 2020-03-18 README.Rmd
merging
added docs

@@
 Finally, we've also leveraged ideas from [R-devel: on avoiding problems with `:::`](https://stat.ethz.ch/pipermail/r-devel/2013-August/067210.html) in `R/aaa.R` to overcome messages from CRAN relative to this hack (using `:::`).
-For some reason - when documenting for `pkgdown` website, we need to do `library(ggtern); timeternR:::update_approved_layers()`
+For some reason - when documenting for `pkgdown` website, we need to do `library(ggtern); EpiCompare:::update_approved_layers()`
 3.
@@
 Basically - there are problems when the `compute_group` wants to make multiple `pieces`/`groups` - and it is similar to the problem that if you do something like `aes(color= var1, group =var2)` you may actually want to do `aes(color= var1, group =paste(var1, var2)`, if there are the same `var2` values across different `var1` values but they don't mean they should be grouped together.
+
+4.
+I (`skgallagher`) had problems with tidy evaluation with `fortify_aggregate`, specifically when one method called another method.
+To get around this, I used a hack to introducing a new argument which would only ever be `NULL` or an object of class `quosure`.
+Thus the outer methods that called inner methods passed quosures.
+See code in `fortify-aggregate-list.R` e.g..
 

### 2020-03-18 R/fortify-aggregate-epimodel-dcm.R

@@ -2,6 +2,7 @@
 #'
 #' @param data output from external source package.  See details
 #' @param states names of states we want aggregate totals of at each time
+#' @param enquo_states vector of quoted names
 #' @param package_source optional argument to include the package from which the
 #'   output is derived from, which helps with the fortify function when outputs
 #'   are of generic classes such as list or data.frame
@@ -21,6 +22,7 @@
 #' @export
 fortify_aggregate.dcm <- function(data,
                                   states = NULL,
+                                  enquo_states = NULL,
                                   package_source = NULL){
 
     out <- fortify_aggregate.icm(data)

### 2020-03-18 R/fortify-aggregate-epimodel-icm.R

@@ -15,6 +15,7 @@
 #'
 #' @param data output from external source package.  See details
 #' @param states names of states we want aggregate totals of at each time
+#' @param enquo_states vector of quoted names
 #' @param package_source optional argument to include the package from which the
 #'   output is derived from, which helps with the fortify function when outputs
 #'   are of generic classes such as list or data.frame
@@ -33,6 +34,7 @@
 #' head(out)
 fortify_aggregate.icm <- function(data,
                               states = NULL,
+                              enquo_states = NULL,
                               package_source = NULL){
 
 

### 2020-03-18 R/fortify-aggregate-list.R

@@ -4,6 +4,7 @@
 #' @param data Output from a pomp simulation where the output is a data frame,
 #'   \code{pomp::simulate()}
 #' @param states vector of state names
+#' @param enquo_states vector of quoted names
 #' @param package_source optional package name
 #' @details The default variables that are retained are SIR, but can be modified
 #'   with the \code{states} argument.  If code{states = NULL}, we will attempt
@@ -17,15 +18,14 @@
 #' @export
 fortify_aggregate.list <- function(data,
                                            states = NULL,
+                                   enquo_states = NULL,
                                            package_source = NULL){
 
 
-    if(!is.null(states)){
-        states <- dplyr::enquos(states)
-    }
+
     if(!is.null(package_source) & package_source == "pomp"){
         out <- fortify_aggregate.pomp_list(data,
-                                               states = states)
+                                               enquo_states = dplyr::enquo(states))
         return(out)
     }
     stop(paste("We currently do not support fortifying an object of class",

### 2020-03-18 R/fortify-aggregate-pomp-list.R

@@ -4,6 +4,7 @@
 #' @param data Output from a pomp simulation where the output is a data frame,
 #'   \code{pomp::simulate()}
 #' @param states vector of state names
+#' @param enquo_states either is null or a quosure
 #' @param package_source optional package name
 #' @details The default variables that are retained are SIR, but can be modified
 #'   with the \code{states} argument.  If code{states = NULL}, we will attempt
@@ -16,14 +17,20 @@
 #' }
 fortify_aggregate.pomp_list <- function(data,
                                            states = c("S", "I", "R"),
+                                        enquo_states = NULL,
                                            package_source = NULL){
 
+    nms <- dimnames(data$states)
+    nm_vars <- nms$variable
     ## pull out state names
-    browser()
-    nms <- dimnames(data[[1]])
-    if(class(states) != "quosures")
-    state_cols <- dplyr::enquos(states)
-    states <- unname(tidyselect::vars_select(nms$variable,
+  #  browser()
+    if(!is.null(enquo_states)){
+        state_cols <- enquo_states
+    } else{
+        state_cols <- dplyr::enquo(states)
+    }
+
+    states <- unname(tidyselect::vars_select(nm_vars,
                                              !!!state_cols))
 
     if(length(states) == 0){

### 2020-03-18 R/fortify-aggregate-pomp-pompList.R

@@ -4,6 +4,7 @@
 #' @param data Output from a pomp simulation where the output is a data frame,
 #'   \code{pomp::simulate()}
 #' @param states vector of state names
+#' @param enquo_states vector of quoted names
 #' @param package_source optional package name
 #' @details The default variables that are retained are SIR, but can be modified
 #'   with the \code{states} argument.  If code{states = NULL}, we will attempt
@@ -22,10 +23,11 @@
 #' unique(rowSums(out[, 3:5]))
 fortify_aggregate.pompList <- function(data,
                                            states = NULL,
+                                          enquo_states = NULL,
                                            package_source = NULL){
     pomp_output <- data
     df <- as.data.frame(pomp_output)
-    out <- fortify_aggregate.pomp_df(df, states = states)
+    out <- fortify_aggregate.pomp_df(df, enquo_states = dplyr::enquo(states))
 
 
 }

### 2020-03-18 R/fortify-aggregate.R

@@ -29,6 +29,7 @@
 #' ## TODO
 fortify_aggregate <- function(data,
                               states = NULL,
+                              enquo_states = NULL,
                               package_source = NULL){
 
     UseMethod("fortify_aggregate")

### 2020-03-18 tests/testthat/test-fortify-aggregate-data-frame.R

@@ -6,5 +6,9 @@ test_that("fortify_aggregate.data.frame", {
     ##
     out <- fortify_aggregate(df, states = c("S", "I", "R"), package_source= "pomp")
     expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
+    ##
+    out2 <- fortify_aggregate(df, states = c("S", "I", "R"), package_source= "pomp")
+    expect_equal(colnames(out2), c("t", "sim", "X0", "X1", "X2"))
+    expect_equal(out, out2)
 
 })

### 2020-03-18 tests/testthat/test-fortify-aggregate-list.R

@@ -9,5 +9,6 @@ test_that("fortify_aggregate.list", {
     ##
     out2 <- fortify_aggregate(my_list, states = c(S, I, R), package_source= "pomp")
     expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
+    expect_equal(out, out2)
 
 })

### 2020-03-19 README.Rmd
minor tweeks, and commentary on Hausdorff distance

@@
 
-## Documentation
-
-Please see our website at [https://skgallagher.github.io/EpiCompare/](https://skgallagher.github.io/EpiCompare/).
-
 ## Installation
@@
 
+Description of data including in this package can be found in the data section of the [reference](https://skgallagher.github.io/EpiCompare/reference/index.html#section-data) page of the documentation website.
+
+```{r eval = F, echo = F}
    Currently there are 13 data sets in which to explore the different `EpiCompare` functions.
@@
   - `pomp_sir` Example output from the `pomp` package.
-
+```
 ## Example
@@
 
-## Notes:
+## Package Creation Notes:
 1.

### 2020-03-19 README.Rmd
updates to README from philosophy

@@
 
-The goal of EpiCompare is to develop functional based visualization and statistics that focus on use in ternary plots.
+The goal of EpiCompare is to provide the epidemiology community with easy-to-use tools to encourage comparing and assessing epidemics and epidemiology models in a time-free manner.
+All tools attempt to adhere to tidyverse/ggplot2 style to enhance easy of use.
+
+Time free analysis allows for stronger comparison of epidemics and model based simulations avoiding different scaling and shifts of time that mess up time-based comparisons.
+
+To achieve this goal, the package contains:
+
+  - **Visualization tools** to visualize SIR epidemics and simulations from SIR models in a time-free manner using `ggtern`’s ternary plots and prediction bands.
+For agent-based SIR models we also provide visualization tools to let the user easily explore how different characteristics of the agents relate to different experiences in the epidemic.
+  - General **comparison tools** to compare epidemics and epidemic models that have higher numbers of states (again in a time-free manner), allowing for the user to examine the differences between models through simulations, and if an epidemic is similar to a model through simulations and prediction bands.
+  - **Conversion tools** to: 
+    - Convert and then compare models from *standard epidemic packages* like `EpiModels`, `pomp`, as well as internal agent-based models, and epidemics in a common framework. 
+    - Convert *agent-based information into aggregate* to compare in the aggregate framework described above.
 

### 2020-03-31 R/agent-to-aggregate.R
fixed tibble bug, still having issue with vignette with NA

@@ -73,7 +73,7 @@ check_min_max_time <- function(min_max_time){
 #' checks if states within data frame are ordered as inputted (<=)
 #'
 #' This function assesses whether agents's time they enter each state is
-#' correctly order. The assumption is that these would be ordered in increasing
+#' correctly ordered. The assumption is that these would be ordered in increasing
 #' values, with the allowance of \code{NA} values not effecting this decision.
 #'
 #' If this important assumption is violated this function either raises an error
@@ -109,6 +109,9 @@ check_min_max_time <- function(min_max_time){
 #' @export
 check_ordered <- function(df, states, assert_error = TRUE){
 
+  if(c("tbl_df") %in% class(df)){
+    df <- as.data.frame(df)
+  }
   df_select <- df[, states]
   K <- length(states)
 

### 2020-03-31 R/fortify-aggregate-epimodel-inner.R

@@ -49,11 +49,16 @@ fortify_aggregate.epimodel_inner <- function(data,
        (data$control$dt != 1)){
         sprintf("Warning!  dt != 1 so the times",
                 "in the output data frame will be",
-                "rescaled to evenly spaced, integer values") 
+                "rescaled to evenly spaced, integer values")
     }
+
     ## Extract time steps
     mat <- data$epi[[1]]
-    t <- rep(1:length(data$control$timesteps), ncol(mat))
+    if("timesteps" %in% names(data$control)){
+        t <- rep(1:length(data$control$timesteps), ncol(mat))
+    } else{
+        t <- rep(1:data$control$nsteps, ncol(mat))
+    }
 
 
     ## Put it all together

### 2020-03-31 tests/testthat/test-agent-to-aggregate.R
still having an issue with limits

@@ -786,3 +786,28 @@ test_that("agents_to_aggregate.group_df, max_t = NA basic checks, strings", {
   }
 
 })
+
+
+
+test_that("NAs for geom_aggregate", {
+
+  library(ggplot2)
+  df <- data.frame(agent_id = factor(1:5),
+                   sim = factor(1),
+                   tI = c(446, NA, 196, 465, NA),
+                   tR = c(464, NA, 425, 476, NA))
+
+  g <- df %>% ggplot() +
+    geom_aggregate(aes(y = tI, z = tR),
+                   color = "blue") +
+    coord_tern()
+
+
+  expect_silent(print(g))
+
+
+})
+
+
+
+

### 2020-04-01 tests/testthat/test-agent-to-aggregate.R
changed test to is.ggplot instead of expect_silent.  can't figure out why that one produced warnings in check but not devtools::test().  Everything seems to be working though?

@@ -791,7 +791,6 @@ test_that("agents_to_aggregate.group_df, max_t = NA basic checks, strings", {
 
 test_that("NAs for geom_aggregate", {
 
-  library(ggplot2)
   df <- data.frame(agent_id = factor(1:5),
                    sim = factor(1),
                    tI = c(446, NA, 196, 465, NA),
@@ -802,8 +801,9 @@ test_that("NAs for geom_aggregate", {
                    color = "blue") +
     coord_tern()
 
+  print(g)
 
-  expect_silent(print(g))
+  expect_true(is.ggplot(g))
 
 
 })

### 2020-04-01 R/fortify-aggregate-epimodel-inner.R
Made an EpiModel fortification vignette.  Also corrected a couple issues with fortify_aggregate

@@ -52,12 +52,13 @@ fortify_aggregate.epimodel_inner <- function(data,
                 "rescaled to evenly spaced, integer values")
     }
 
+    
     ## Extract time steps
     mat <- data$epi[[1]]
     if("timesteps" %in% names(data$control)){
-        t <- rep(1:length(data$control$timesteps), ncol(mat))
+        t <- rep(1:length(data$control$timesteps), each = ncol(mat))
     } else{
-        t <- rep(1:data$control$nsteps, ncol(mat))
+        t <- rep(1:data$control$nsteps, each = ncol(mat))
     }
 
 

### 2020-04-02 README.Rmd
trying to fix loss of documentation and functions when overriding from github - pkgdown doesn't work that well

@@
 
-# EpiCompare <a href = "https://skgallagher.github.io/EpiCompare/"><img src="man/figures/logo.png" align="right" alt="" width="120" /></a>
+# EpiCompare <a href = "https://skgallagher.github.io/EpiCompare/"><img src="man/figures/logo.png" align="right" alt="EpiCompare" width="120" /></a>
 
@@
 
-The goal of EpiCompare is to provide the epidemiology community with easy-to-use tools to encourage comparing and assessing epidemics and epidemiology models in a time-free manner.
-All tools attempt to adhere to tidyverse/ggplot2 style to enhance easy of use.
+The goal of `EpiCompare` is to provide the epidemiology community with easy-to-use tools to encourage comparing and assessing epidemics and epidemiology models in a time-free manner.
+All tools attempt to adhere to `tidyverse`/`ggplot2` style to enhance easy of use.
 

### 2020-04-02 R/EpiCompare.R

@@ -1,36 +1,39 @@
-#' EpiCompare: Ternary plot visualization for SIR curves (UPDATE)
+#' EpiCompare: 
+#' 
+#' The goal of \code{EpiCompare} is to provide the epidemiology community with
+#' easy-to-use tools to encourage comparing and assessing epidemics and
+#' epidemiology models in a time-free manner. All tools attempt to adhere to
+#' \code{\link[tidyverse]{tidyverse}}/\code{\link[ggplot2]{ggplot2}} style to
+#' enhance easy of use.
 #'
-#' The goal of EpiCompare is to develop functional based visualization and
-#' statistics that focus on use in ternary plots. (UPDATE)
+#' Time free analysis allows for stronger comparison of epidemics and model
+#' based simulations avoiding different scaling and shifts of time that mess up
+#' time-based comparisons.
 #'
-#' This package attempts to:
+#' To achieve this goal, the package contains:
 #' \itemize{
-#'   \item \strong{Comparisons Model and Epidemic Comparisons}: We provide easy
-#'   ways to compare multiple epidemics, epidemics vs models, or models verse
-#'   models with ternary plots and confidence analysis
-#'     \itemize{
-#'       \item Ternary plots have been shown to help show time-independent like
-#'       comparing R_0
-#'       \item Added confidence band structure allows for examining variability
-#'       within simulations from models cleanly (and without worrying about time
-#'       variation)
-#'         \itemize{
-#'           \item Provides ways to moves beyond pointwise examination to
-#'           uniform containment
-#'         }
-#'     }
-#'   \item \strong{Agent (->/ vs) Aggregate}: Provide easy way to compare agents
-#'   based data/modeling vs aggregation
-#'     \itemize{
-#'       \item Our tools provide easy ways to explore different grouping of
-#'       agents and understanding how these groups different through aggregation
-#'       visualizations (see Shannon’s thesis for rational of quick access to
-#'       grouping abilities)
-#'    }
-#'   \item\strong{ggplot2/ggtern and tidyverse}: tools exist in the rapidly
-#'   expanding tidyverse/ggplot2 paradigm, allowing for quick deployment and
-#'   less cost in learning new assessment tools
-#'  }
+#' \item \strong{Visualization tools} to visualize SIR epidemics and simulations
+#' from SIR models in a time-free manner using \code{\link[ggtern]{ggtern}}’s
+#' ternary plots and prediction bands. For agent-based SIR models we also
[…]

### 2020-04-02 README.Rmd
updated documentation along with updates with local depth and depth vignette

@@
 `geom_prediction_band` required not just `compute_group` but `compute_layer` - there is very little documentation on how to approach this correctly.
-Basically - there are problems when the `compute_group` wants to make multiple `pieces`/`groups` - and it is similar to the problem that if you do something like `aes(color= var1, group =var2)` you may actually want to do `aes(color= var1, group =paste(var1, var2)`, if there are the same `var2` values across different `var1` values but they don't mean they should be grouped together.
-
-4.
-I (`skgallagher`) had problems with tidy evaluation with `fortify_aggregate`, specifically when one method called another method.
-To get around this, I used a hack to introducing a new argument which would only ever be `NULL` or an object of class `quosure`.
-Thus the outer methods that called inner methods passed quosures.
-See code in `fortify-aggregate-list.R` e.g..
+Basically - there are problems when the `compute_group` wants to make multiple `pieces`/`groups` - and it is similar to the problem that if you do something like `aes(color = var1, group = var2)` you may actually want to do `aes(color = var1, group = paste(var1, var2))`, if there are the same `var2` values across different `var1` values but they don't mean they should be grouped together.
 

### 2020-04-02 R/distance-depth.R

@@ -87,8 +87,8 @@ depth_curves_to_points.grouped_df <- function(x, alpha, dist_mat,
 #' Specifically we use Geenens & Nieto-Reyes's global distance-based depth
 #' defined as:
 #'
-#' \eqn{DD(x, \hat{P}) = 1/(n choose 2) * sum_{i!=j} I(d(X_1, X_2) > max(d(X_1,x),
-#' d(X_2,x)))}
+#' \eqn{DD(x, \hat{P}) = 1/(n choose 2) \cdot \sum_{i!=j} I(d(X_i, X_j) >
+#' max(d(X_i,x), d(X_j,x)))}
 #'
 #' @details
 #' This function (renamed as \code{depth_function}) is shared with
@@ -147,5 +147,91 @@ distance_depth_function <- function(dist_mat){
                                 ])
   }
 
+  return(depth)
+}
+
+
+
+#' Localized version of Geenens & Nieto-Reyes functional distance-based depth
+#'
+#' Calculates a localized distance-based depth vector using a distance matrix.
+#' Specifically we alter Geenens & Nieto-Reyes's global distance-based depth to
+#' a local depth similar to Agostinelli & Romanazzi (2011) localized approach,
+#' and define our local distance-based depth as:
+#' 
+#' \eqn{LDD(x, \hat{P}, \tau) = 1/(|S| choose 2) \cdot \sum_{i!=j, i,j \in S}
+#' I(d(X_i, X_j) > max(d(X_i,x), d(X_j,x)))}
+#'
+#' where \eqn{S = {i : D(X_i,x) < \tau}}.
+#'
+#' Note that if \eqn{|S| = 1}, then we say that \code{LDD(x) = 0}.
+#'
+#' @details
+#' This function (renamed as \code{depth_function}) is shared with
+#' \pkg{TCpredictionbands} on github:
+#' \href{https://github.com/Mr8ND/TC-prediction-bands/tree/master/TCpredictionbands}{TCpredictionbands}.
+#' 
+#' @param dist_mat a n x n square positive symmetric matrix
+#' @param tau localizing parameter (default is Inf)
+#'
+#' @return depth vector length n with depth values associated with indices in
+#'   \code{dist_mat}
+#' @export
+#'
+#' @examples
+#' dist_mat <- matrix(c(0,   1, 1.5,
+#'                      1,   0, 2,
+#'                      1.5, 2, 0   ),
[…]

### 2020-04-02 tests/testthat/test-distance-depth.R

@@ -66,3 +66,37 @@ test_that("test for distance_depth_function", {
   testthat::expect_error(distance_depth_function(dist_mat_not_pos))
 
 })
+
+test_that("test for local_distance_depth_function", {
+  dist_mat <- matrix(c(0,   1, 1.5,
+                       1,   0, 2,
+                       1.5, 2, 0   ),
+                     nrow = 3,
+                     byrow = TRUE)
+  
+  
+  dd_vec <- local_distance_depth_function(dist_mat) # c(1,0,0)
+  
+  ldd_vec1 <- local_distance_depth_function(dist_mat, tau = 2) # c(1,0,0)
+  ldd_vec2 <- local_distance_depth_function(dist_mat, tau = 1.5) # c(1,0,0)
+  ldd_vec3 <- local_distance_depth_function(dist_mat, tau = 1) # c(0,0,0)
+  ldd_vec4 <- local_distance_depth_function(dist_mat, tau = .1) # c(0,0,0)
+  
+  testthat::expect_equal(dd_vec, c(1,0,0)) # same as global depth
+  testthat::expect_equal(ldd_vec1, c(1,0,0)) # same as global depth
+  testthat::expect_equal(ldd_vec2, c(1,0,0)) # works relative to S constraint
+  testthat::expect_equal(ldd_vec3, c(0,0,0)) # works relative to S constraint (only 1 in each box)
+  testthat::expect_equal(ldd_vec4, c(0,0,0)) # works relative to S constraint (only 0 in each box)
+  
+  dist_mat_not_sym <- matrix(c(0,   1, 0,
+                               1,   0, 2,
+                               1.5, 2, 0   ))
+  dist_mat_not_pos <- matrix(c(0,   -1, 1.5,
+                               -1,   0, 2,
+                               1.5, 2, 0   ))
+  
+  testthat::expect_error(local_distance_depth_function(dist_mat_not_sym))
+  testthat::expect_error(local_distance_depth_function(dist_mat_not_pos))
+  
+})
+

### 2020-04-02 jss/jss-article-rnw/article.Rnw
jss article format in Rsweave style is included. Mostly to allow for more easily making into a vignette as well.

\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{thumbpdf,lmodern}

%% another package (only for this demo article)
\usepackage{framed}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
@

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Shannon K. Gallagher\\
        Biostatistics Research Branch \\ 
        National Institute of Allergy and Infectious Diseases
   \AND Benjamin LeRoy\\
        Department of Statistics \& Data Science\\ 
        Carnegie Mellon University}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{A Short Demo Article: Regression Models for Count Data in \proglang{R}}
\Plaintitle{A Short Demo Article: Regression Models for Count Data in R}
\Shorttitle{A Short Demo Article in \proglang{R}}

%% - \Abstract{} almost as usual
\Abstract{
  This short article illustrates how to write a manuscript for the
  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
  it is recommended to keep the {\LaTeX} code as simple as possible,
  i.e., avoid inclusion of packages/commands that are not necessary.
  For outlining the typical structure of a JSS article some brief text snippets
  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
  discussing count data regression in \proglang{R}. Editorial comments and
  instructions are marked by vertical bars.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{JSS, style guide, comma-separated, not capitalized, \proglang{R}}
\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Achim Zeileis\\
  Journal of Statistical Software\\
  \emph{and}\\
  Department of Statistics\\
  Faculty of Economics and Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstr.~15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Achim.Zeileis@R-project.org}\\
  URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
}

\begin{document}
\SweaveOpts{concordance=TRUE}

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section[Introduction: Count data regression in R]{Introduction: Count data regression in \proglang{R}} \label{sec:intro}

\begin{leftbar}
The introduction is in principle ``as usual''. However, it should usually embed
both the implemented \emph{methods} and the \emph{software} into the respective
relevant literature. For the latter both competing and complementary software
should be discussed (within the same software environment and beyond), bringing
out relative (dis)advantages. All software mentioned should be properly
\verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
\textsc{Bib}{\TeX}.)

For writing about software JSS requires authors to use the markup
\verb|\proglang{}| (programming languages and large programmable systems),
\verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands,
arguments, etc.). If there is such markup in (sub)section titles (as above), a
plain text version has to be provided in the {\LaTeX} command as well. Below we
also illustrate how abbrevations should be introduced and citation commands can
be employed. See the {\LaTeX} code for more details.
\end{leftbar}

Modeling count variables is a common task in economics and the social sciences.
The classical Poisson regression model for count data is often of limited use in
these disciplines because empirical count data sets typically exhibit
overdispersion and/or an excess number of zeros. The former issue can be
addressed by extending  the plain Poisson regression model in various
directions: e.g., using sandwich covariances or estimating an additional
dispersion parameter (in a so-called quasi-Poisson model). Another more formal
way is to use a negative binomial (NB) regression. All of these models belong to
the family of generalized linear models (GLMs). However, although these models
typically can capture overdispersion rather well, they are in many applications
not sufficient for  modeling excess zeros. Since \cite{Mullahy:1986} there is
increased interest in zero-augmented models that address this issue by a second
model component capturing zero counts. An overview of count data models in
econometrics, including  hurdle and zero-inflated models, is provided in
\cite{Cameron+Trivedi:2013}.

In \proglang{R} \citep{R}, GLMs are provided by the model fitting functions
\fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the \pkg{MASS} package
\citep[][Chapter~7.4]{Venables+Ripley:2002} along with associated methods for
diagnostics and inference. The manuscript that this document is based on
\citep{Zeileis+Kleiber+Jackman:2008} then introduced hurdle and zero-inflated
count models in the functions \fct{hurdle} and \fct{zeroinfl} in the \pkg{pscl}
package \citep{Jackman:2015}. Of course, much more software could be discussed
here, including (but not limited to) generalized additive models for count data
as available in the \proglang{R} packages \pkg{mgcv} \cite{Wood:2006}, 
\pkg{gamlss} \citep{Stasinopoulos+Rigby:2007}, or \pkg{VGAM} \citep{Yee:2009}.

%% -- Manuscript ---------------------------------------------------------------

%% - In principle "as usual" again.
%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
%%   avoid empty lines before and after the equation (which would signal a new
%%   paragraph.
%% - When describing longer chunks of code that are _not_ meant for execution
%%   (e.g., a function synopsis or list of arguments), the environment {Code}
%%   is recommended. Alternatively, a plain {verbatim} can also be used.
%%   (For executed code see the next section.)

\section{Models and software} \label{sec:models}

The basic Poisson regression model for count data is a special case of the GLM
framework \cite{McCullagh+Nelder:1989}. It describes the dependence of a count
response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution
$y_i \sim \mathrm{Pois}(\mu_i)$. The dependence of the conditional mean
$\E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a
log link and a linear predictor
%
\begin{equation} \label{eq:mean}
\log(\mu_i) \quad = \quad x_i^\top \beta,
\end{equation}
%
where the regression coefficients $\beta$ are estimated by maximum likelihood
(ML) using the iterative weighted least squares (IWLS) algorithm.

\begin{leftbar}
Note that around the \verb|{equation}| above there should be no spaces (avoided
in the {\LaTeX} code by \verb|%| lines) so that ``normal'' spacing is used and
not a new paragraph started.
\end{leftbar}

\proglang{R} provides a very flexible implementation of the general GLM
framework in the function \fct{glm} \citep{Chambers+Hastie:1992} in the
\pkg{stats} package. Its most important arguments are
\begin{Code}
glm(formula, data, subset, na.action, weights, offset,
  family = gaussian, start = NULL, control = glm.control(...),
  model = TRUE, y = TRUE, x = FALSE, ...)
\end{Code}
where \code{formula} plus \code{data} is the now standard way of specifying
regression relationships in \proglang{R}/\proglang{S} introduced in
\cite{Chambers+Hastie:1992}. The remaining arguments in the first line
(\code{subset}, \code{na.action}, \code{weights}, and \code{offset}) are also
standard  for setting up formula-based regression models in
\proglang{R}/\proglang{S}. The arguments in the second line control aspects
specific to GLMs while the arguments in the last line specify which components
are returned in the fitted model object (of class \class{glm} which inherits
from \class{lm}). For further arguments to \fct{glm} (including alternative
specifications of starting values) see \code{?glm}. For estimating a Poisson
model \code{family = poisson} has to be specified.

\begin{leftbar}
As the synopsis above is a code listing that is not meant to be executed,
one can use either the dedicated \verb|{Code}| environment or a simple
\verb|{verbatim}| environment for this. Again, spaces before and after should be
avoided.

Finally, there might be a reference to a \verb|{table}| such as
Table~\ref{tab:overview}. Usually, these are placed at the top of the page
(\verb|[t!]|), centered (\verb|\centering|), with a caption below the table,
column headers and captions in sentence style, and if possible avoiding vertical
lines.
\end{leftbar}

\begin{table}[t!]
\centering
\begin{tabular}{lllp{7.4cm}}
\hline
Type           & Distribution & Method   & Description \\ \hline
GLM            & Poisson      & ML       & Poisson regression: classical GLM,
                                           estimated by maximum likelihood (ML) \\
               &              & Quasi    & ``Quasi-Poisson regression'':
                                           same mean function, estimated by
                                           quasi-ML (QML) or equivalently
                                           generalized estimating equations (GEE),
                                           inference adjustment via estimated
                                           dispersion parameter \\
               &              & Adjusted & ``Adjusted Poisson regression'':
                                           same mean function, estimated by
                                           QML/GEE, inference adjustment via
                                           sandwich covariances\\
               & NB           & ML       & NB regression: extended GLM,
                                           estimated by ML including additional
                                           shape parameter \\ \hline
Zero-augmented & Poisson      & ML       & Zero-inflated Poisson (ZIP),
                                           hurdle Poisson \\
               & NB           & ML       & Zero-inflated NB (ZINB),
                                           hurdle NB \\ \hline
\end{tabular}
\caption{\label{tab:overview} Overview of various count regression models. The
table is usually placed at the top of the page (\texttt{[t!]}), centered
(\texttt{centering}), has a caption below the table, column headers and captions
are in sentence style, and if possible vertical lines should be avoided.}
\end{table}

%% -- Illustrations ------------------------------------------------------------

%% - Virtually all JSS manuscripts list source code along with the generated
%%   output. The style files provide dedicated environments for this.
%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
%%   or knitr using the render_sweave() hook - are used (without the need to
%%   load Sweave.sty).
%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
%% - The code input should use "the usual" command prompt in the respective
%%   software system.
%% - For R code, the prompt "R> " should be used with "+  " as the
%%   continuation prompt.
%% - Comments within the code chunks should be avoided - these should be made
%%   within the regular LaTeX text.

\section{Illustrations} \label{sec:illustrations}

For a simple illustration of basic Poisson and NB count regression the
\code{quine} data from the \pkg{MASS} package is used. This provides the number
of \code{Days} that children were absent from school in Australia in a
particular year, along with several covariates that can be employed as regressors.
The data can be loaded by
%
<<data>>=
data("quine", package = "MASS")
@
%
and a basic frequency distribution of the response variable is displayed in
Figure~\ref{fig:quine}.

\begin{leftbar}
For code input and output, the style files provide dedicated environments.
Either the ``agnostic'' \verb|{CodeInput}| and \verb|{CodeOutput}| can be used
or, equivalently, the environments \verb|{Sinput}| and \verb|{Soutput}| as
produced by \fct{Sweave} or \pkg{knitr} when using the \code{render_sweave()}
hook. Please make sure that all code is properly spaced, e.g., using
\code{y = a + b * x} and \emph{not} \code{y=a+b*x}. Moreover, code input should
use ``the usual'' command prompt in the respective software system. For
\proglang{R} code, the prompt \code{"R> "} should be used with \code{"+  "} as
the continuation prompt. Generally, comments within the code chunks should be
avoided -- and made in the regular {\LaTeX} text instead. Finally, empty lines
before and after code input/output should be avoided (see above).
\end{leftbar}

\begin{figure}[t!]
\centering
<<visualization, echo=FALSE, fig=TRUE, height=5.2, width=7>>=
par(mar = c(4, 4, 1, 1))
plot(table(quine$Days), xlab = "Days", ylab = "Frequency", axes = FALSE)
axis(2)
axis(1, at = 0:16 * 5, labels = FALSE)
axis(1, at = 0:8 * 10)
box()
@
\caption{\label{fig:quine} Frequency distribution for number of days absent
from school.}
\end{figure}

As a first model for the \code{quine} data, we fit the basic Poisson regression
model. (Note that JSS prefers when the second line of code is indented by two
spaces.)
%
<<poisson>>=
m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine,
  family = poisson)
@
%
To account for potential overdispersion we also consider a negative binomial
GLM.
%
<<negbin>>=
library("MASS")
m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
@
%
In a comparison with the BIC the latter model is clearly preferred.
%
<<comparison>>=
BIC(m_pois, m_nbin)
@
%
Hence, the full summary of that model is shown below.
%
<<summary>>=
summary(m_nbin)
@

%% -- Summary/conclusions/discussion -------------------------------------------

\section{Summary and discussion} \label{sec:summary}

\begin{leftbar}
As usual \dots
\end{leftbar}

%% -- Optional special unnumbered sections -------------------------------------

\section*{Computational details}

\begin{leftbar}
If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualizations,
maps, tables, \dots) that are not cited in the main text can be credited here.
\end{leftbar}

The results in this paper were obtained using
\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the
\pkg{MASS}~\Sexpr{packageVersion("MASS")} package. \proglang{R} itself
and all packages used are available from the Comprehensive
\proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/}.

\section*{Acknowledgments}

\begin{leftbar}
All acknowledgments (note the AE spelling) should be collected in this
unnumbered section before the references. It may contain the usual information
about funding and feedback from colleagues/reviewers/etc. Furthermore,
information such as relative contributions of the authors may be added here
(if any).
\end{leftbar}

%% -- Bibliography -------------------------------------------------------------
%% - References need to be provided in a .bib BibTeX database.
%% - All references should be made with \cite, \citet, \citep, \citealp etc.
%%   (and never hard-coded). See the FAQ for details.
%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
%% - Titles in the .bib should be in title case.
%% - DOIs should be included where available.

\bibliography{refs}

%% -- Appendix (if any) --------------------------------------------------------
%% - After the bibliography with page break.
%% - With proper section titles and _not_ just "Appendix".

\newpage

\begin{appendix}

\section{More technical details} \label{app:technical}

\begin{leftbar}
Appendices can be included after the bibliography (with a page break). Each
section within the appendix should have a proper section title (rather than
just \emph{Appendix}).

For more technical style details, please check out JSS's style FAQ at
\url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
which includes the following topics:
\begin{itemize}
  \item Title vs.\ sentence case.
  \item Graphics formatting.
  \item Naming conventions.
  \item Turning JSS manuscripts into \proglang{R} package vignettes.
  \item Trouble shooting.
  \item Many other potentially helpful details\dots
\end{itemize}
\end{leftbar}

\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}

\begin{leftbar}
References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
\verb|\citealp| etc.\ (and never hard-coded). This commands yield different
formats of author-year citations and allow to include additional details (e.g.,
pages, chapters, \dots) in brackets. In case you are not familiar with these
commands see the JSS style FAQ for details.

Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially
when acquiring the entries automatically from mixed online sources. However,
it is important that informations are complete and presented in a consistent
style to avoid confusions. JSS requires the following format.
\begin{itemize}
  \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
    be used in the references.
  \item Titles should be in title case.
  \item Journal titles should not be abbreviated and in title case.
  \item DOIs should be included where available.
  \item Software should be properly cited as well. For \proglang{R} packages
    \code{citation("pkgname")} typically provides a good starting point.
\end{itemize}
\end{leftbar}

\end{appendix}

%% -----------------------------------------------------------------------------

\end{document}

### 2020-04-02 jss/jss-article-rnw/article.tex
jss article format in Rsweave style is included. Mostly to allow for more easily making into a vignette as well.

--- 2020-04-02 jss/jss-article-rnw/article.Rnw
@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
-library("MASS")
-@
 
@@
 %% - \AND starts a new line, \And does not.
-\author{Shannon K. Gallagher\\
-        Biostatistics Research Branch \\ 
-        National Institute of Allergy and Infectious Diseases
-   \AND Benjamin LeRoy\\
-        Department of Statistics \& Data Science\\ 
-        Carnegie Mellon University}
+\author{Shannon K. Gallagher\\Biostatistics Research Branch \\ National Institute of Allergy and Infectious Diseases
+   \AND Benjamin LeRoy\\Department of Statistics \& Data Science\\ Carnegie Mellon University}
 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{article-concordance}
 
@@
 %
-<<data>>=
-data("quine", package = "MASS")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> data("quine", package = "MASS")
+\end{Sinput}
+\end{Schunk}
 %
@@
 \centering
-<<visualization, echo=FALSE, fig=TRUE, height=5.2, width=7>>=
-par(mar = c(4, 4, 1, 1))
-plot(table(quine$Days), xlab = "Days", ylab = "Frequency", axes = FALSE)
-axis(2)
-axis(1, at = 0:16 * 5, labels = FALSE)
-axis(1, at = 0:8 * 10)
-box()
-@
+\includegraphics{article-visualization}
 \caption{\label{fig:quine} Frequency distribution for number of days absent
@@
 %
-<<poisson>>=
-m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine,
-  family = poisson)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine,
++    family = poisson)
+\end{Sinput}
+\end{Schunk}
 %
@@
 %
-<<negbin>>=
-library("MASS")
-m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> library("MASS")
+R> m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
+\end{Sinput}
+\end{Schunk}
 %
@@
 %
-<<comparison>>=
-BIC(m_pois, m_nbin)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> BIC(m_pois, m_nbin)
+\end{Sinput}
+\begin{Soutput}
+       df      BIC
+m_pois 18 2046.851
+m_nbin 19 1157.235
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<summary>>=
-summary(m_nbin)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> summary(m_nbin)
+\end{Sinput}
+\begin{Soutput}
+Call:
+glm.nb(formula = Days ~ (Eth + Sex + Age + Lrn)^2, data = quine, 
+    init.theta = 1.60364105, link = log)
+
+Deviance Residuals: 
+    Min       1Q   Median       3Q      Max  
+-3.0857  -0.8306  -0.2620   0.4282   2.0898
+
+Coefficients: (1 not defined because of singularities)
+            Estimate Std. Error z value Pr(>|z|)    
+(Intercept)  3.00155    0.33709   8.904  < 2e-16 ***
+EthN        -0.24591    0.39135  -0.628  0.52977    
+SexM        -0.77181    0.38021  -2.030  0.04236 *  
+AgeF1       -0.02546    0.41615  -0.061  0.95121    
+AgeF2       -0.54884    0.54393  -1.009  0.31296    
+AgeF3       -0.25735    0.40558  -0.635  0.52574    
+LrnSL        0.38919    0.48421   0.804  0.42153    
+EthN:SexM    0.36240    0.29430   1.231  0.21818    
+EthN:AgeF1  -0.70000    0.43646  -1.604  0.10876    
+EthN:AgeF2  -1.23283    0.42962  -2.870  0.00411 ** 
+EthN:AgeF3   0.04721    0.44883   0.105  0.91622    
+EthN:LrnSL   0.06847    0.34040   0.201  0.84059    
+SexM:AgeF1   0.02257    0.47360   0.048  0.96198    
+SexM:AgeF2   1.55330    0.51325   3.026  0.00247 ** 
+SexM:AgeF3   1.25227    0.45539   2.750  0.00596 ** 
+SexM:LrnSL   0.07187    0.40805   0.176  0.86019    
+AgeF1:LrnSL -0.43101    0.47948  -0.899  0.36870    
+AgeF2:LrnSL  0.52074    0.48567   1.072  0.28363    
+AgeF3:LrnSL       NA         NA      NA       NA    
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for Negative Binomial(1.6036) family taken to be 1)
+
+    Null deviance: 235.23  on 145  degrees of freedom
+Residual deviance: 167.53  on 128  degrees of freedom
+AIC: 1100.5
+
+Number of Fisher Scoring iterations: 1
+
+              Theta:  1.604 
+          Std. Err.:  0.214
+
+ 2 x log-likelihood:  -1062.546 
+\end{Soutput}
+\end{Schunk}
 
@@
 The results in this paper were obtained using
-\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the
-\pkg{MASS}~\Sexpr{packageVersion("MASS")} package. \proglang{R} itself
+\proglang{R}~3.6.1 with the
+\pkg{MASS}~7.3.51.4 package. \proglang{R} itself
 and all packages used are available from the Comprehensive

### 2020-04-05 jss/jss-article-rnw/article.Rnw
jss minor updates

@@
 options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
-library("MASS")
+devtools::load_all()
 @
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{A Short Demo Article: Regression Models for Count Data in \proglang{R}}
-\Plaintitle{A Short Demo Article: Regression Models for Count Data in R}
-\Shorttitle{A Short Demo Article in \proglang{R}}
+\title{EpiCompare: A tidy \proglang{R} package for comparing and assessing epidemics and epidimology models}
+\Plaintitle{EpiCompare: A tidy R package for comparing and assessing epidemics and epidimology models}
+\Shorttitle{EpiCompare: an \proglang{R} package for comparing and assessing epidemics and models}
 
@@
 
-\bibliography{refs}
+\bibliography{EpiCompare}
 

### 2020-04-05 jss/jss-article-rnw/article.tex
jss minor updates

@@
 %% - \AND starts a new line, \And does not.
-\author{Shannon K. Gallagher\\Biostatistics Research Branch \\ National Institute of Allergy and Infectious Diseases
-   \AND Benjamin LeRoy\\Department of Statistics \& Data Science\\ Carnegie Mellon University}
+\author{Shannon K. Gallagher\\
+        Biostatistics Research Branch \\ 
+        National Institute of Allergy and Infectious Diseases
+   \AND Benjamin LeRoy\\
+        Department of Statistics \& Data Science\\ 
+        Carnegie Mellon University}
 

### 2020-04-15 jss/jss-article-rnw/article.Rnw
remove concordance from being produced
first attempt on introduction ~ben

@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\SweaveOpts{concordance=FALSE}
 
@@
 %%   "generalized linear models" below).
+
+\section[Ben's Introduction]{Ben's Introduction} \label{sec:ben_intro}
+
+The recent (and currently on-going) COVID-19 global epidemic has galvenized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area \cite{FILL IN}.
+Infectious disease models models attempt, typically, to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{biggerstaff2016}; or 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \cite{roberts2004}.
+The COVID-19 epidemic highlights how both goals are very important, and that preventing and mitigating the future spread of infectious disease which inflicts enormous burdens on the world: millions of lives lost and billions of dollars spent yearly, is so important.
+
+The current epidemic has also highlighted that these models are just one piece of the overall analysis pipeline.
+John Hopkin's \_\_\_ Dashboard and county, state and national tabulation of rudimentary numerical summaries emphasis the need to strong data to assess the spread of infectious disease (multiple \cite{FILL IN}?).
+Stilly, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individal / agent level, network structure or just aggregate flows (review paper \cite{FILL IN}?).
+All placing individuals / proportions of the populations into different states (e.g.
+suspectible, exposed, infected, recovered, etc.).
+With all these models, review and comparison papers in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center helps the individual practioner decide the correct approach.
+
+Within this paradigm / pipeline of exploring the data, modeling the epidemics and assessing different models, we present our \proglang{R} package \pkg{EpiCompare}.
+Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models using \textit{time-free analysis tools}.
+
+More specifically, \pkg{EpiCompare} provides epidemiologists a toolkit containing:
+\begin{itemize}
+  \item \textbf{Visualization tools} to visualize three-state epidemics and simulations from three-state\footnote{The most common three-state model that will be examined is the SIR model (with the three states: Susceptible, Infected and Recovered), and will be discussed in more detail in Section \ref{}.} models in a time-free manner using ternary plots and prediction bands.
+For agent-based three-state models we also provide visualization tools to let the user easily explore how different characteristics of the agents relate to different experiences in the epidemic.
+  \item General \textbf{comparison tools} to compare epidemics and epidemic models that have higher numbers of states (again in a time-free manner), allowing for the user to examine the differences between models through simulations, and if an epidemic is similar to a model through simulations and prediction bands.
+  \item \textbf{Conversion tools} to:
+    \begin{itemize}
+      \item Convert and then compare models from \textit{standard \proglang{R} epidemic packages} as well as internal agent-based models, and epidemics in a common framework.
+      \item Convert \textit{agent-based information into aggregate} to compare in the aggregate framework described above.
+Installation
+    \end{itemize}
+\end{itemize}
+
+This paper is broken up into the following sections, section 2 contains a literature review, section 3 contains discuss of the time-free analysis, and section 4 presents the package...
 

### 2020-04-21 jss/jss-article-rnw/article-woven.tex
added orig_t column to EpiModels.  Also write a start of a vignette for visualizing covid-19 data.

--- 2020-04-15 jss/jss-article-rnw/article.Rnw
@@
+\documentclass{article}
+
+\begin{document}
+
+\end{document}
 \documentclass[article]{jss}
@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
-devtools::load_all()
-@
 
@@
 \begin{document}
-\SweaveOpts{concordance=FALSE}
 
@@
 \centering
-<<visualization, echo=FALSE, fig=TRUE, height=5.2, width=7>>=
-par(mar = c(4, 4, 1, 1))
-plot(table(quine$Days), xlab = "Days", ylab = "Frequency", axes = FALSE)
-axis(2)
-axis(1, at = 0:16 * 5, labels = FALSE)
-axis(1, at = 0:8 * 10)
-box()
-@
+\includegraphics{article-woven-visualization}
 \caption{\label{fig:quine} Frequency distribution for number of days absent
@@
 The results in this paper were obtained using
-\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the
-\pkg{MASS}~\Sexpr{packageVersion("MASS")} package. \proglang{R} itself
+\proglang{R}~3.6.3 with the
+\pkg{MASS}~7.3.51.5 package. \proglang{R} itself
 and all packages used are available from the Comprehensive

### 2020-04-21 R/fortify-aggregate-epimodel-dcm.R

@@ -24,7 +24,7 @@ fortify_aggregate.dcm <- function(data,
                                   package_source = NULL){
 
     out <- fortify_aggregate.epimodel_inner(data,
-                                 states = dplyr::enquo(states))
+                                            states = dplyr::enquo(states))
     return(out)
 
 }

### 2020-04-21 tests/testthat/test-fortify-aggregate-epimodel-inner.R

@@ -1,23 +1,23 @@
 test_that("fortify_aggregate.epimodel_inner", {
 
     out <- fortify_aggregate(EpiModel_icm)
-    expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
-    expect_true(all(rowSums(out[, 3:5]) == 1000))
+    expect_equal(colnames(out), c("t", "orig_t", "sim", "X0", "X1", "X2"))
+    expect_true(all(rowSums(out[, 4:6]) == 1000))
 
     ############3
     out2 <- fortify_aggregate(EpiModel_agg_bd)
-    expect_equal(colnames(out2), c("t", "sim", "X0", "X1", "X2"))
+    expect_equal(colnames(out2), c("t", "orig_t", "sim", "X0", "X1", "X2"))
 
     ########
     out3 <- fortify_aggregate.icm(EpiModel_det)
-    expect_equal(colnames(out3), c("t", "sim", "X0", "X1", "X2"))
-    expect_equal(rowSums(out3[, 3:5]),  rep(1000.0, 300))
+    expect_equal(colnames(out3), c("t", "orig_t", "sim", "X0", "X1", "X2"))
+    expect_equal(rowSums(out3[, 4:6]),  rep(1000.0, 300))
     #############
     #############
     out <- fortify_aggregate(EpiModel_icm, states = c("s.num", "i.num"))
-    expect_equal(colnames(out), c("t", "sim", "X0", "X1"))
+    expect_equal(colnames(out), c("t", "orig_t", "sim", "X0", "X1"))
     out2 <- fortify_aggregate(EpiModel_icm, states = c(s.num, i.num))
-    expect_equal(colnames(out2), c("t", "sim", "X0", "X1"))
+    expect_equal(colnames(out2), c("t", "orig_t", "sim", "X0", "X1"))
     expect_equal(out, out2)
 
 })

### 2020-04-21 tests/testthat/test-fortify-aggregate.R

@@ -1,17 +1,17 @@
 test_that("fortify_aggregate", {
     ## ICM
     out <- fortify_aggregate(EpiModel_icm)
-    expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
+    expect_equal(colnames(out), c("t", "orig_t", "sim", "X0", "X1", "X2"))
     ## DCM
     out3 <- fortify_aggregate(EpiModel_det)
-    expect_equal(colnames(out3), c("t", "sim", "X0", "X1", "X2"))
-    expect_equal(rowSums(out3[, 3:5]),  rep(1000.0, 300))
+    expect_equal(colnames(out3), c("t", "orig_t", "sim", "X0", "X1", "X2"))
+    expect_equal(rowSums(out3[, 4:6]),  rep(1000.0, 300))
     ## POMP (POMP_LIST)
     out <- fortify_aggregate(pomp_pomp, states = NULL)
     expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2", "X3"))
     ##
     out <- fortify_aggregate(pomp_pomp, states = c("S", "I", "R"))
-    expect_equal(colnames(out), c("t", "sim", "X0", "X1", "X2"))
+    expect_equal(colnames(out), c("t",  "sim", "X0", "X1", "X2"))
     ## POMP
 
 })

### 2020-04-21 tests/testthat/test-impute-recovery-counts.R

@@ -1,0 +1,48 @@
+## SKG
+## April 16, 2020
+## Tests for chain binomial SIR imputation
+
+test_that("cases_to_SIR", {
+    df <- data.frame(t = 0:4,
+                     confirmed = c(0, 1, 3, 9, 9),
+                     N = 10)
+    out <- cases_to_SIR(data = df,
+                        par = 1)
+    exp_S <- c(10, 9, 7, 1, 1)
+    exp_I <- c(0, 1, 2, 6, 0)
+    exp_R <- c(0, 0, 1, 3, 9)
+    expect_equal(out$X0, exp_S)
+    expect_equal(out$X1, exp_I)
+    expect_equal(out$X2, exp_R)
+    expect_equal(out$X2 + out$X1, out$confirmed)
+    expect_true(all(out$X1 >= 0))
+    expect_true(all(out$X2 >= 0))
+    expect_true(all((out$X2 - dplyr::lag(out$X2) >= 0) |
+                    is.na(dplyr::lag(out$X2))))
+    ## ########################################
+    df <- data.frame(t = 0:4,
+                     confirmed = c(0, 1, 3, 9, 9),
+                     N = 10)
+    out <- cases_to_SIR(data = df,
+                        par = .8)
+    expect_equal(out$X2 + out$X1, out$confirmed)
+    expect_true(all(out$X1 >= 0))
+    expect_true(all(out$X2 >= 0))
+    expect_true(all((out$X2 - dplyr::lag(out$X2) >= 0) |
+                    is.na(dplyr::lag(out$X2))))
+    ## GROUPED DF ##########
+     df1 <- data.frame(t = 0:4,
+                     confirmed = c(0, 1, 3, 9, 9),
+                     N = 10,
+                     group = 1)
+    df2 <- data.frame(t = 0:4,
+                      confirmed = c(0, 3, 3, 4, 4),
+                      N = 9, group = 2)
+    df <- dplyr::bind_rows(df1, df2) %>% dplyr::group_by(group)
+    out <- cases_to_SIR(df, par = 1)
+    expect_equal(out$X2 + out$X1, out$confirmed)
+    expect_true(all(out$X1 >= 0))
+    expect_true(all(out$X2 >= 0))
+    expect_equal(out$X0 + out$X1 + out$X2, out$N)
+    
+})

### 2020-04-21 jss/jss-article-rnw/article.Rnw
added orig_t column to EpiModels.  Also write a start of a vignette for visualizing covid-19 data.
Added template for mini sections in .Rnw form

--- 2020-04-02 jss/jss-article-rnw/article.Rnw
@@
 \usepackage{thumbpdf,lmodern}
+\usepackage{import}
 
@@
 %% - \AND starts a new line, \And does not.
-\author{Shannon K. Gallagher\\
-        Biostatistics Research Branch \\ 
-        National Institute of Allergy and Infectious Diseases
-   \AND Benjamin LeRoy\\
-        Department of Statistics \& Data Science\\ 
-        Carnegie Mellon University}
+\author{Shannon K. Gallagher\\ National Institute of\\ Allergy and Infectious Diseases
+   \And Benjamin P. Leroy\\Carnegie Mellon University}
+\Plainauthor{Shannon K. Gallagher, Benjamin P. Leroy}
 
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{A Short Demo Article: Regression Models for Count Data in \proglang{R}}
-\Plaintitle{A Short Demo Article: Regression Models for Count Data in R}
-\Shorttitle{A Short Demo Article in \proglang{R}}
+\title{\proglang{EpiCompare}: Working title}
+\Plaintitle{EpiCompare}
+\Shorttitle{\prolang{EpiCompare}}
 
@@
 \begin{document}
-\SweaveOpts{concordance=TRUE}
 
-%% -- Introduction -------------------------------------------------------------
+\SweaveInput{chapters/ch1-intro.Rnw}
 
-%% - In principle "as usual".
-%% - But should typically have some discussion of both _software_ and _methods_.
-%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
-%% - If such markup is in (sub)section titles, a plain text version has to be
-%%   added as well.
-%% - All software mentioned should be properly \cite-d.
-%% - All abbreviations should be introduced.
-%% - Unless the expansions of abbreviations are proper names (like "Journal
-%%   of Statistical Software" above) they should be in sentence case (like
-%%   "generalized linear models" below).
+\section{Purpose and Details}
+\SweaveInput{chapters/ch2-vis.Rnw}
+\SweaveInput{chapters/ch3-compare.Rnw}
+\SweaveInput{chapters/ch4-convert.Rnw}
 
-\section[Introduction: Count data regression in R]{Introduction: Count data regression in \proglang{R}} \label{sec:intro}
+\section{In practice}
+\SweaveInput{chapters/ch5-ex1.Rnw}
+\SweaveInput{chapters/ch6-ex2.Rnw}
 
-\begin{leftbar}
-The introduction is in principle ``as usual''. However, it should usually embed
-both the implemented \emph{methods} and the \emph{software} into the respective
-relevant literature. For the latter both competing and complementary software
-should be discussed (within the same software environment and beyond), bringing
-out relative (dis)advantages. All software mentioned should be properly
-\verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
-\textsc{Bib}{\TeX}.)
+\SweaveInput{chapters/ch7-discussion.Rnw}
 
-For writing about software JSS requires authors to use the markup
-\verb|\proglang{}| (programming languages and large programmable systems),
-\verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands,
-arguments, etc.). If there is such markup in (sub)section titles (as above), a
-plain text version has to be provided in the {\LaTeX} command as well. Below we
-also illustrate how abbrevations should be introduced and citation commands can
-be employed. See the {\LaTeX} code for more details.
-\end{leftbar}
+\section{references}
 
-Modeling count variables is a common task in economics and the social sciences.
-The classical Poisson regression model for count data is often of limited use in
-these disciplines because empirical count data sets typically exhibit
-overdispersion and/or an excess number of zeros. The former issue can be
-addressed by extending  the plain Poisson regression model in various
-directions: e.g., using sandwich covariances or estimating an additional
-dispersion parameter (in a so-called quasi-Poisson model). Another more formal
-way is to use a negative binomial (NB) regression. All of these models belong to
-the family of generalized linear models (GLMs). However, although these models
-typically can capture overdispersion rather well, they are in many applications
-not sufficient for  modeling excess zeros. Since \cite{Mullahy:1986} there is
-increased interest in zero-augmented models that address this issue by a second
-model component capturing zero counts. An overview of count data models in
-econometrics, including  hurdle and zero-inflated models, is provided in
-\cite{Cameron+Trivedi:2013}.
+\bibliography{EpiCompare}
 
-In \proglang{R} \citep{R}, GLMs are provided by the model fitting functions
-\fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the \pkg{MASS} package
-\citep[][Chapter~7.4]{Venables+Ripley:2002} along with associated methods for
-diagnostics and inference. The manuscript that this document is based on
-\citep{Zeileis+Kleiber+Jackman:2008} then introduced hurdle and zero-inflated
-count models in the functions \fct{hurdle} and \fct{zeroinfl} in the \pkg{pscl}
-package \citep{Jackman:2015}. Of course, much more software could be discussed
-here, including (but not limited to) generalized additive models for count data
-as available in the \proglang{R} packages \pkg{mgcv} \cite{Wood:2006}, 
-\pkg{gamlss} \citep{Stasinopoulos+Rigby:2007}, or \pkg{VGAM} \citep{Yee:2009}.
-
-%% -- Manuscript ---------------------------------------------------------------
-
-%% - In principle "as usual" again.
-%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
-%%   avoid empty lines before and after the equation (which would signal a new
-%%   paragraph.
-%% - When describing longer chunks of code that are _not_ meant for execution
-%%   (e.g., a function synopsis or list of arguments), the environment {Code}
-%%   is recommended. Alternatively, a plain {verbatim} can also be used.
-%%   (For executed code see the next section.)
-
-\section{Models and software} \label{sec:models}
-
-The basic Poisson regression model for count data is a special case of the GLM
-framework \cite{McCullagh+Nelder:1989}. It describes the dependence of a count
-response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution
-$y_i \sim \mathrm{Pois}(\mu_i)$. The dependence of the conditional mean
-$\E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a
-log link and a linear predictor
-%
-\begin{equation} \label{eq:mean}
-\log(\mu_i) \quad = \quad x_i^\top \beta,
-\end{equation}
-%
-where the regression coefficients $\beta$ are estimated by maximum likelihood
-(ML) using the iterative weighted least squares (IWLS) algorithm.
-
-\begin{leftbar}
-Note that around the \verb|{equation}| above there should be no spaces (avoided
-in the {\LaTeX} code by \verb|%| lines) so that ``normal'' spacing is used and
-not a new paragraph started.
-\end{leftbar}
-
-\proglang{R} provides a very flexible implementation of the general GLM
-framework in the function \fct{glm} \citep{Chambers+Hastie:1992} in the
-\pkg{stats} package. Its most important arguments are
-\begin{Code}
-glm(formula, data, subset, na.action, weights, offset,
-  family = gaussian, start = NULL, control = glm.control(...),
-  model = TRUE, y = TRUE, x = FALSE, ...)
-\end{Code}
-where \code{formula} plus \code{data} is the now standard way of specifying
-regression relationships in \proglang{R}/\proglang{S} introduced in
-\cite{Chambers+Hastie:1992}. The remaining arguments in the first line
-(\code{subset}, \code{na.action}, \code{weights}, and \code{offset}) are also
-standard  for setting up formula-based regression models in
-\proglang{R}/\proglang{S}. The arguments in the second line control aspects
-specific to GLMs while the arguments in the last line specify which components
-are returned in the fitted model object (of class \class{glm} which inherits
-from \class{lm}). For further arguments to \fct{glm} (including alternative
-specifications of starting values) see \code{?glm}. For estimating a Poisson
-model \code{family = poisson} has to be specified.
-
-\begin{leftbar}
-As the synopsis above is a code listing that is not meant to be executed,
-one can use either the dedicated \verb|{Code}| environment or a simple
-\verb|{verbatim}| environment for this. Again, spaces before and after should be
-avoided.
-
-Finally, there might be a reference to a \verb|{table}| such as
-Table~\ref{tab:overview}. Usually, these are placed at the top of the page
-(\verb|[t!]|), centered (\verb|\centering|), with a caption below the table,
-column headers and captions in sentence style, and if possible avoiding vertical
-lines.
-\end{leftbar}
-
-\begin{table}[t!]
-\centering
-\begin{tabular}{lllp{7.4cm}}
-\hline
-Type           & Distribution & Method   & Description \\ \hline
-GLM            & Poisson      & ML       & Poisson regression: classical GLM,
-                                           estimated by maximum likelihood (ML) \\
-               &              & Quasi    & ``Quasi-Poisson regression'':
-                                           same mean function, estimated by
-                                           quasi-ML (QML) or equivalently
-                                           generalized estimating equations (GEE),
-                                           inference adjustment via estimated
-                                           dispersion parameter \\
-               &              & Adjusted & ``Adjusted Poisson regression'':
-                                           same mean function, estimated by
-                                           QML/GEE, inference adjustment via
-                                           sandwich covariances\\
-               & NB           & ML       & NB regression: extended GLM,
-                                           estimated by ML including additional
-                                           shape parameter \\ \hline
-Zero-augmented & Poisson      & ML       & Zero-inflated Poisson (ZIP),
-                                           hurdle Poisson \\
-               & NB           & ML       & Zero-inflated NB (ZINB),
-                                           hurdle NB \\ \hline
-\end{tabular}
-\caption{\label{tab:overview} Overview of various count regression models. The
-table is usually placed at the top of the page (\texttt{[t!]}), centered
-(\texttt{centering}), has a caption below the table, column headers and captions
-are in sentence style, and if possible vertical lines should be avoided.}
-\end{table}
-
-%% -- Illustrations ------------------------------------------------------------
-
-%% - Virtually all JSS manuscripts list source code along with the generated
-%%   output. The style files provide dedicated environments for this.
-%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
-%%   or knitr using the render_sweave() hook - are used (without the need to
-%%   load Sweave.sty).
-%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
-%% - The code input should use "the usual" command prompt in the respective
-%%   software system.
-%% - For R code, the prompt "R> " should be used with "+  " as the
-%%   continuation prompt.
-%% - Comments within the code chunks should be avoided - these should be made
-%%   within the regular LaTeX text.
-
-\section{Illustrations} \label{sec:illustrations}
-
-For a simple illustration of basic Poisson and NB count regression the
-\code{quine} data from the \pkg{MASS} package is used. This provides the number
-of \code{Days} that children were absent from school in Australia in a
-particular year, along with several covariates that can be employed as regressors.
-The data can be loaded by
-%
-<<data>>=
-data("quine", package = "MASS")
-@
-%
-and a basic frequency distribution of the response variable is displayed in
-Figure~\ref{fig:quine}.
-
-\begin{leftbar}
-For code input and output, the style files provide dedicated environments.
-Either the ``agnostic'' \verb|{CodeInput}| and \verb|{CodeOutput}| can be used
-or, equivalently, the environments \verb|{Sinput}| and \verb|{Soutput}| as
-produced by \fct{Sweave} or \pkg{knitr} when using the \code{render_sweave()}
-hook. Please make sure that all code is properly spaced, e.g., using
-\code{y = a + b * x} and \emph{not} \code{y=a+b*x}. Moreover, code input should
-use ``the usual'' command prompt in the respective software system. For
-\proglang{R} code, the prompt \code{"R> "} should be used with \code{"+  "} as
-the continuation prompt. Generally, comments within the code chunks should be
-avoided -- and made in the regular {\LaTeX} text instead. Finally, empty lines
-before and after code input/output should be avoided (see above).
-\end{leftbar}
-
-\begin{figure}[t!]
-\centering
-<<visualization, echo=FALSE, fig=TRUE, height=5.2, width=7>>=
-par(mar = c(4, 4, 1, 1))
-plot(table(quine$Days), xlab = "Days", ylab = "Frequency", axes = FALSE)
-axis(2)
-axis(1, at = 0:16 * 5, labels = FALSE)
-axis(1, at = 0:8 * 10)
-box()
-@
-\caption{\label{fig:quine} Frequency distribution for number of days absent
-from school.}
-\end{figure}
-
-As a first model for the \code{quine} data, we fit the basic Poisson regression
-model. (Note that JSS prefers when the second line of code is indented by two
-spaces.)
-%
-<<poisson>>=
-m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine,
-  family = poisson)
-@
-%
-To account for potential overdispersion we also consider a negative binomial
-GLM.
-%
-<<negbin>>=
-library("MASS")
-m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
-@
-%
-In a comparison with the BIC the latter model is clearly preferred.
-%
-<<comparison>>=
-BIC(m_pois, m_nbin)
-@
-%
-Hence, the full summary of that model is shown below.
-%
-<<summary>>=
-summary(m_nbin)
-@
-
-%% -- Summary/conclusions/discussion -------------------------------------------
-
-\section{Summary and discussion} \label{sec:summary}
-
-\begin{leftbar}
-As usual \dots
-\end{leftbar}
-
-%% -- Optional special unnumbered sections -------------------------------------
-
-\section*{Computational details}
-
-\begin{leftbar}
-If necessary or useful, information about certain computational details
-such as version numbers, operating systems, or compilers could be included
-in an unnumbered section. Also, auxiliary packages (say, for visualizations,
-maps, tables, \dots) that are not cited in the main text can be credited here.
-\end{leftbar}
-
-The results in this paper were obtained using
-\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the
-\pkg{MASS}~\Sexpr{packageVersion("MASS")} package. \proglang{R} itself
-and all packages used are available from the Comprehensive
-\proglang{R} Archive Network (CRAN) at
-\url{https://CRAN.R-project.org/}.
-
-\section*{Acknowledgments}
-
-\begin{leftbar}
-All acknowledgments (note the AE spelling) should be collected in this
-unnumbered section before the references. It may contain the usual information
-about funding and feedback from colleagues/reviewers/etc. Furthermore,
-information such as relative contributions of the authors may be added here
-(if any).
-\end{leftbar}
-
-%% -- Bibliography -------------------------------------------------------------
-%% - References need to be provided in a .bib BibTeX database.
-%% - All references should be made with \cite, \citet, \citep, \citealp etc.
-%%   (and never hard-coded). See the FAQ for details.
-%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
-%% - Titles in the .bib should be in title case.
-%% - DOIs should be included where available.
-
-\bibliography{refs}
-
-%% -- Appendix (if any) --------------------------------------------------------
-%% - After the bibliography with page break.
-%% - With proper section titles and _not_ just "Appendix".
-
-\newpage
-
-\begin{appendix}
-
-\section{More technical details} \label{app:technical}
-
-\begin{leftbar}
-Appendices can be included after the bibliography (with a page break). Each
-section within the appendix should have a proper section title (rather than
-just \emph{Appendix}).
-
-For more technical style details, please check out JSS's style FAQ at
-\url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
-which includes the following topics:
-\begin{itemize}
-  \item Title vs.\ sentence case.
-  \item Graphics formatting.
-  \item Naming conventions.
-  \item Turning JSS manuscripts into \proglang{R} package vignettes.
-  \item Trouble shooting.
-  \item Many other potentially helpful details\dots
-\end{itemize}
-\end{leftbar}
-
-\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
-
-\begin{leftbar}
-References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
-references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
-\verb|\citealp| etc.\ (and never hard-coded). This commands yield different
-formats of author-year citations and allow to include additional details (e.g.,
-pages, chapters, \dots) in brackets. In case you are not familiar with these
-commands see the JSS style FAQ for details.
-
-Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially
-when acquiring the entries automatically from mixed online sources. However,
-it is important that informations are complete and presented in a consistent
-style to avoid confusions. JSS requires the following format.
-\begin{itemize}
-  \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
-    be used in the references.
-  \item Titles should be in title case.
-  \item Journal titles should not be abbreviated and in title case.
-  \item DOIs should be included where available.
-  \item Software should be properly cited as well. For \proglang{R} packages
-    \code{citation("pkgname")} typically provides a good starting point.
-\end{itemize}
-\end{leftbar}
-
-\end{appendix}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2020-04-21 jss/jss-article-rnw/chapters/ch1-intro.Rnw
Added template for mini sections in .Rnw form

--- 2020-04-15 jss/jss-article-rnw/article.Rnw
@@
-\documentclass[article]{jss}
-
-%% -- LaTeX packages and custom commands ---------------------------------------
-
-%% recommended packages
-\usepackage{thumbpdf,lmodern}
-
-%% another package (only for this demo article)
-\usepackage{framed}
-
-%% new custom commands
-\newcommand{\class}[1]{`\code{#1}'}
-\newcommand{\fct}[1]{\code{#1()}}
-
-%% For Sweave-based articles about R packages:
-%% need no \usepackage{Sweave}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
-devtools::load_all()
-@
-
-%% -- Article metainformation (author, title, ...) -----------------------------
-
-%% - \author{} with primary affiliation
-%% - \Plainauthor{} without affiliations
-%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
-%% - \AND starts a new line, \And does not.
-\author{Shannon K. Gallagher\\
-        Biostatistics Research Branch \\ 
-        National Institute of Allergy and Infectious Diseases
-   \AND Benjamin LeRoy\\
-        Department of Statistics \& Data Science\\ 
-        Carnegie Mellon University}
-
-%% - \title{} in title case
-%% - \Plaintitle{} without LaTeX markup (if any)
-%% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{EpiCompare: A tidy \proglang{R} package for comparing and assessing epidemics and epidimology models}
-\Plaintitle{EpiCompare: A tidy R package for comparing and assessing epidemics and epidimology models}
-\Shorttitle{EpiCompare: an \proglang{R} package for comparing and assessing epidemics and models}
-
-%% - \Abstract{} almost as usual
-\Abstract{
-  This short article illustrates how to write a manuscript for the
-  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
-  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
-  it is recommended to keep the {\LaTeX} code as simple as possible,
-  i.e., avoid inclusion of packages/commands that are not necessary.
-  For outlining the typical structure of a JSS article some brief text snippets
-  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
-  discussing count data regression in \proglang{R}. Editorial comments and
-  instructions are marked by vertical bars.
-}
-
-%% - \Keywords{} with LaTeX markup, at least one required
-%% - \Plainkeywords{} without LaTeX markup (if necessary)
-%% - Should be comma-separated and in sentence case.
-\Keywords{JSS, style guide, comma-separated, not capitalized, \proglang{R}}
-\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}
-
-%% - \Address{} of at least one author
-%% - May contain multiple affiliations for each author
-%%   (in extra lines, separated by \emph{and}\\).
-%% - May contain multiple authors for the same affiliation
-%%   (in the same first line, separated by comma).
-\Address{
-  Achim Zeileis\\
-  Journal of Statistical Software\\
-  \emph{and}\\
-  Department of Statistics\\
-  Faculty of Economics and Statistics\\
-  Universit\"at Innsbruck\\
-  Universit\"atsstr.~15\\
-  6020 Innsbruck, Austria\\
-  E-mail: \email{Achim.Zeileis@R-project.org}\\
-  URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
-}
-
-\begin{document}
-\SweaveOpts{concordance=FALSE}
-
-%% -- Introduction -------------------------------------------------------------
-
-%% - In principle "as usual".
-%% - But should typically have some discussion of both _software_ and _methods_.
-%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
-%% - If such markup is in (sub)section titles, a plain text version has to be
-%%   added as well.
-%% - All software mentioned should be properly \cite-d.
-%% - All abbreviations should be introduced.
-%% - Unless the expansions of abbreviations are proper names (like "Journal
-%%   of Statistical Software" above) they should be in sentence case (like
-%%   "generalized linear models" below).
+% !Rnw root = ../article.Rnw
 
@@
 This paper is broken up into the following sections, section 2 contains a literature review, section 3 contains discuss of the time-free analysis, and section 4 presents the package...
-
-\section[Introduction: Count data regression in R]{Introduction: Count data regression in \proglang{R}} \label{sec:intro}
-
-\begin{leftbar}
-The introduction is in principle ``as usual''. However, it should usually embed
-both the implemented \emph{methods} and the \emph{software} into the respective
-relevant literature. For the latter both competing and complementary software
-should be discussed (within the same software environment and beyond), bringing
-out relative (dis)advantages. All software mentioned should be properly
-\verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
-\textsc{Bib}{\TeX}.)
-
-For writing about software JSS requires authors to use the markup
-\verb|\proglang{}| (programming languages and large programmable systems),
-\verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands,
-arguments, etc.). If there is such markup in (sub)section titles (as above), a
-plain text version has to be provided in the {\LaTeX} command as well. Below we
-also illustrate how abbrevations should be introduced and citation commands can
-be employed. See the {\LaTeX} code for more details.
-\end{leftbar}
-
-Modeling count variables is a common task in economics and the social sciences.
-The classical Poisson regression model for count data is often of limited use in
-these disciplines because empirical count data sets typically exhibit
-overdispersion and/or an excess number of zeros. The former issue can be
-addressed by extending  the plain Poisson regression model in various
-directions: e.g., using sandwich covariances or estimating an additional
-dispersion parameter (in a so-called quasi-Poisson model). Another more formal
-way is to use a negative binomial (NB) regression. All of these models belong to
-the family of generalized linear models (GLMs). However, although these models
-typically can capture overdispersion rather well, they are in many applications
-not sufficient for  modeling excess zeros. Since \cite{Mullahy:1986} there is
-increased interest in zero-augmented models that address this issue by a second
-model component capturing zero counts. An overview of count data models in
-econometrics, including  hurdle and zero-inflated models, is provided in
-\cite{Cameron+Trivedi:2013}.
-
-In \proglang{R} \citep{R}, GLMs are provided by the model fitting functions
-\fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the \pkg{MASS} package
-\citep[][Chapter~7.4]{Venables+Ripley:2002} along with associated methods for
-diagnostics and inference. The manuscript that this document is based on
-\citep{Zeileis+Kleiber+Jackman:2008} then introduced hurdle and zero-inflated
-count models in the functions \fct{hurdle} and \fct{zeroinfl} in the \pkg{pscl}
-package \citep{Jackman:2015}. Of course, much more software could be discussed
-here, including (but not limited to) generalized additive models for count data
-as available in the \proglang{R} packages \pkg{mgcv} \cite{Wood:2006}, 
-\pkg{gamlss} \citep{Stasinopoulos+Rigby:2007}, or \pkg{VGAM} \citep{Yee:2009}.
-
-%% -- Manuscript ---------------------------------------------------------------
-
-%% - In principle "as usual" again.
-%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
-%%   avoid empty lines before and after the equation (which would signal a new
-%%   paragraph.
-%% - When describing longer chunks of code that are _not_ meant for execution
-%%   (e.g., a function synopsis or list of arguments), the environment {Code}
-%%   is recommended. Alternatively, a plain {verbatim} can also be used.
-%%   (For executed code see the next section.)
-
-\section{Models and software} \label{sec:models}
-
-The basic Poisson regression model for count data is a special case of the GLM
-framework \cite{McCullagh+Nelder:1989}. It describes the dependence of a count
-response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution
-$y_i \sim \mathrm{Pois}(\mu_i)$. The dependence of the conditional mean
-$\E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a
-log link and a linear predictor
-%
-\begin{equation} \label{eq:mean}
-\log(\mu_i) \quad = \quad x_i^\top \beta,
-\end{equation}
-%
-where the regression coefficients $\beta$ are estimated by maximum likelihood
-(ML) using the iterative weighted least squares (IWLS) algorithm.
-
-\begin{leftbar}
-Note that around the \verb|{equation}| above there should be no spaces (avoided
-in the {\LaTeX} code by \verb|%| lines) so that ``normal'' spacing is used and
-not a new paragraph started.
-\end{leftbar}
-
-\proglang{R} provides a very flexible implementation of the general GLM
-framework in the function \fct{glm} \citep{Chambers+Hastie:1992} in the
-\pkg{stats} package. Its most important arguments are
-\begin{Code}
-glm(formula, data, subset, na.action, weights, offset,
-  family = gaussian, start = NULL, control = glm.control(...),
-  model = TRUE, y = TRUE, x = FALSE, ...)
-\end{Code}
-where \code{formula} plus \code{data} is the now standard way of specifying
-regression relationships in \proglang{R}/\proglang{S} introduced in
-\cite{Chambers+Hastie:1992}. The remaining arguments in the first line
-(\code{subset}, \code{na.action}, \code{weights}, and \code{offset}) are also
-standard  for setting up formula-based regression models in
-\proglang{R}/\proglang{S}. The arguments in the second line control aspects
-specific to GLMs while the arguments in the last line specify which components
-are returned in the fitted model object (of class \class{glm} which inherits
-from \class{lm}). For further arguments to \fct{glm} (including alternative
-specifications of starting values) see \code{?glm}. For estimating a Poisson
-model \code{family = poisson} has to be specified.
-
-\begin{leftbar}
-As the synopsis above is a code listing that is not meant to be executed,
-one can use either the dedicated \verb|{Code}| environment or a simple
-\verb|{verbatim}| environment for this. Again, spaces before and after should be
-avoided.
-
-Finally, there might be a reference to a \verb|{table}| such as
-Table~\ref{tab:overview}. Usually, these are placed at the top of the page
-(\verb|[t!]|), centered (\verb|\centering|), with a caption below the table,
-column headers and captions in sentence style, and if possible avoiding vertical
-lines.
-\end{leftbar}
-
-\begin{table}[t!]
-\centering
-\begin{tabular}{lllp{7.4cm}}
-\hline
-Type           & Distribution & Method   & Description \\ \hline
-GLM            & Poisson      & ML       & Poisson regression: classical GLM,
-                                           estimated by maximum likelihood (ML) \\
-               &              & Quasi    & ``Quasi-Poisson regression'':
-                                           same mean function, estimated by
-                                           quasi-ML (QML) or equivalently
-                                           generalized estimating equations (GEE),
-                                           inference adjustment via estimated
-                                           dispersion parameter \\
-               &              & Adjusted & ``Adjusted Poisson regression'':
-                                           same mean function, estimated by
-                                           QML/GEE, inference adjustment via
-                                           sandwich covariances\\
-               & NB           & ML       & NB regression: extended GLM,
-                                           estimated by ML including additional
-                                           shape parameter \\ \hline
-Zero-augmented & Poisson      & ML       & Zero-inflated Poisson (ZIP),
-                                           hurdle Poisson \\
-               & NB           & ML       & Zero-inflated NB (ZINB),
-                                           hurdle NB \\ \hline
-\end{tabular}
-\caption{\label{tab:overview} Overview of various count regression models. The
-table is usually placed at the top of the page (\texttt{[t!]}), centered
-(\texttt{centering}), has a caption below the table, column headers and captions
-are in sentence style, and if possible vertical lines should be avoided.}
-\end{table}
-
-%% -- Illustrations ------------------------------------------------------------
-
-%% - Virtually all JSS manuscripts list source code along with the generated
-%%   output. The style files provide dedicated environments for this.
-%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
-%%   or knitr using the render_sweave() hook - are used (without the need to
-%%   load Sweave.sty).
-%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
-%% - The code input should use "the usual" command prompt in the respective
-%%   software system.
-%% - For R code, the prompt "R> " should be used with "+  " as the
-%%   continuation prompt.
-%% - Comments within the code chunks should be avoided - these should be made
-%%   within the regular LaTeX text.
-
-\section{Illustrations} \label{sec:illustrations}
-
-For a simple illustration of basic Poisson and NB count regression the
-\code{quine} data from the \pkg{MASS} package is used. This provides the number
-of \code{Days} that children were absent from school in Australia in a
-particular year, along with several covariates that can be employed as regressors.
-The data can be loaded by
-%
-<<data>>=
-data("quine", package = "MASS")
-@
-%
-and a basic frequency distribution of the response variable is displayed in
-Figure~\ref{fig:quine}.
-
-\begin{leftbar}
-For code input and output, the style files provide dedicated environments.
-Either the ``agnostic'' \verb|{CodeInput}| and \verb|{CodeOutput}| can be used
-or, equivalently, the environments \verb|{Sinput}| and \verb|{Soutput}| as
-produced by \fct{Sweave} or \pkg{knitr} when using the \code{render_sweave()}
-hook. Please make sure that all code is properly spaced, e.g., using
-\code{y = a + b * x} and \emph{not} \code{y=a+b*x}. Moreover, code input should
-use ``the usual'' command prompt in the respective software system. For
-\proglang{R} code, the prompt \code{"R> "} should be used with \code{"+  "} as
-the continuation prompt. Generally, comments within the code chunks should be
-avoided -- and made in the regular {\LaTeX} text instead. Finally, empty lines
-before and after code input/output should be avoided (see above).
-\end{leftbar}
-
-\begin{figure}[t!]
-\centering
-<<visualization, echo=FALSE, fig=TRUE, height=5.2, width=7>>=
-par(mar = c(4, 4, 1, 1))
-plot(table(quine$Days), xlab = "Days", ylab = "Frequency", axes = FALSE)
-axis(2)
-axis(1, at = 0:16 * 5, labels = FALSE)
-axis(1, at = 0:8 * 10)
-box()
-@
-\caption{\label{fig:quine} Frequency distribution for number of days absent
-from school.}
-\end{figure}
-
-As a first model for the \code{quine} data, we fit the basic Poisson regression
-model. (Note that JSS prefers when the second line of code is indented by two
-spaces.)
-%
-<<poisson>>=
-m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine,
-  family = poisson)
-@
-%
-To account for potential overdispersion we also consider a negative binomial
-GLM.
-%
-<<negbin>>=
-library("MASS")
-m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
-@
-%
-In a comparison with the BIC the latter model is clearly preferred.
-%
-<<comparison>>=
-BIC(m_pois, m_nbin)
-@
-%
-Hence, the full summary of that model is shown below.
-%
-<<summary>>=
-summary(m_nbin)
-@
-
-%% -- Summary/conclusions/discussion -------------------------------------------
-
-\section{Summary and discussion} \label{sec:summary}
-
-\begin{leftbar}
-As usual \dots
-\end{leftbar}
-
-%% -- Optional special unnumbered sections -------------------------------------
-
-\section*{Computational details}
-
-\begin{leftbar}
-If necessary or useful, information about certain computational details
-such as version numbers, operating systems, or compilers could be included
-in an unnumbered section. Also, auxiliary packages (say, for visualizations,
-maps, tables, \dots) that are not cited in the main text can be credited here.
-\end{leftbar}
-
-The results in this paper were obtained using
-\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} with the
-\pkg{MASS}~\Sexpr{packageVersion("MASS")} package. \proglang{R} itself
-and all packages used are available from the Comprehensive
-\proglang{R} Archive Network (CRAN) at
-\url{https://CRAN.R-project.org/}.
-
-\section*{Acknowledgments}
-
-\begin{leftbar}
-All acknowledgments (note the AE spelling) should be collected in this
-unnumbered section before the references. It may contain the usual information
-about funding and feedback from colleagues/reviewers/etc. Furthermore,
-information such as relative contributions of the authors may be added here
-(if any).
-\end{leftbar}
-
-%% -- Bibliography -------------------------------------------------------------
-%% - References need to be provided in a .bib BibTeX database.
-%% - All references should be made with \cite, \citet, \citep, \citealp etc.
-%%   (and never hard-coded). See the FAQ for details.
-%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
-%% - Titles in the .bib should be in title case.
-%% - DOIs should be included where available.
-
-\bibliography{EpiCompare}
-
-%% -- Appendix (if any) --------------------------------------------------------
-%% - After the bibliography with page break.
-%% - With proper section titles and _not_ just "Appendix".
-
-\newpage
-
-\begin{appendix}
-
-\section{More technical details} \label{app:technical}
-
-\begin{leftbar}
-Appendices can be included after the bibliography (with a page break). Each
-section within the appendix should have a proper section title (rather than
-just \emph{Appendix}).
-
-For more technical style details, please check out JSS's style FAQ at
-\url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
-which includes the following topics:
-\begin{itemize}
-  \item Title vs.\ sentence case.
-  \item Graphics formatting.
-  \item Naming conventions.
-  \item Turning JSS manuscripts into \proglang{R} package vignettes.
-  \item Trouble shooting.
-  \item Many other potentially helpful details\dots
-\end{itemize}
-\end{leftbar}
-
-\section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
-
-\begin{leftbar}
-References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
-references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
-\verb|\citealp| etc.\ (and never hard-coded). This commands yield different
-formats of author-year citations and allow to include additional details (e.g.,
-pages, chapters, \dots) in brackets. In case you are not familiar with these
-commands see the JSS style FAQ for details.
-
-Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially
-when acquiring the entries automatically from mixed online sources. However,
-it is important that informations are complete and presented in a consistent
-style to avoid confusions. JSS requires the following format.
-\begin{itemize}
-  \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
-    be used in the references.
-  \item Titles should be in title case.
-  \item Journal titles should not be abbreviated and in title case.
-  \item DOIs should be included where available.
-  \item Software should be properly cited as well. For \proglang{R} packages
-    \code{citation("pkgname")} typically provides a good starting point.
-\end{itemize}
-\end{leftbar}
-
-\end{appendix}
-
-%% -----------------------------------------------------------------------------
-
-\end{document}

### 2020-04-21 jss/jss-article-rnw/chapters/ch2-vis.Rnw
Added template for mini sections in .Rnw form

% !Rnw root = ../article.Rnw

\subsection[Visualize]{Visualize} \label{sec:purpose-vis}

### 2020-04-21 jss/jss-article-rnw/chapters/ch5-ex1.Rnw
Added template for mini sections in .Rnw form

% !Rnw root = ../article.Rnw

\subsection[Case Study 1]{Case Study 1} \label{sec:ex1}

EXAMPLE 1

### 2020-04-22 jss/jss-article-rnw/article.Rnw
Added R0 section. yay

@@
 \newcommand{\fct}[1]{\code{#1()}}
+
+\usepackage{amsmath}
+\usepackage{amsthm}
+\newtheorem{theorem}{Theorem}
 
@@
 \Abstract{
-  This short article illustrates how to write a manuscript for the
-  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
-  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
-  it is recommended to keep the {\LaTeX} code as simple as possible,
-  i.e., avoid inclusion of packages/commands that are not necessary.
-  For outlining the typical structure of a JSS article some brief text snippets
-  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
-  discussing count data regression in \proglang{R}. Editorial comments and
-  instructions are marked by vertical bars.
+We do something
 }
@@
 \Address{
-  Achim Zeileis\\
-  Journal of Statistical Software\\
-  \emph{and}\\
-  Department of Statistics\\
-  Faculty of Economics and Statistics\\
-  Universit\"at Innsbruck\\
-  Universit\"atsstr.~15\\
-  6020 Innsbruck, Austria\\
-  E-mail: \email{Achim.Zeileis@R-project.org}\\
-  URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
+  Shanon K. Gallagher\\
+  Biostatistics Research Branch\\
+  Division of Clinical Research\\
+  National Institute of Allergy and Infectious Diseases\\
+  5601 Fishers Lane\\
+  Rockville, MD 20852\\
+  E-mail: \email{shannon.gallagher@nih.gov}
 }
@@
 
+\SweaveInput{chapters/app1-proofs.Rnw}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-04-22 jss/jss-article-rnw/chapters/app1-proofs.Rnw
Added R0 section. yay

% !Rnw root = ../article.Rnw

\section[Proof of Theorem]{Proof of Theorem} \label{app:proof}

\begin{proof}
\cite{harko2014} provide an analytical solution for the Kermack and McKendrick equations (Eq. \eqref{eq:sir-ode}) by reparameterizing the ODEs so that $\mathcal{S}(u) = S(t)$, $\mathcal{I}(u) = S(t)$, and $\mathcal{R}(u) = R(t)$ for $0< u_T < 1$ with
\begin{align}\label{eq:harko-odes}
\mathcal{S}(u) &= S(0)u\\
\mathcal{I}(u) &= N - R(0) + NR_0^{-1}\log u - S(0)u \nonumber\\
\mathcal{R}(u) &= R(0) - NR_0^{-1} \log u, \nonumber
\end{align}
and $u$ and t are related by the following integral,
\begin{align*}
    t &= \int_{u}^1 \frac{N}{\beta \tau (N - R(0) + R_{0}^{-1} \log \tau - S(0)\tau)}d\tau \\
    &= \int_{u}^1 \frac{1}{\beta f(S(0), R(0), N, R_0, \tau)} d \tau\\
    &= \int_{u}^1 \frac{1}{\beta f(\tau)} d\tau,
\end{align*}
where we have made the denominator of the integral a function of $N$, the initial values, $R_0$, and $\tau$, which we further condense to $f(\tau)$ for brevity.
Then for a given $t$ we want to find $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Or equivalently, for a fixed $u$ want to find $v$ such that  $\mathcal{S}_1(u) = \mathcal{S}_2(v)$ and then the corresponding $t$ and $s$ are given by
\begin{align*}
    t & = \int_{u}^1 \frac{1}{\beta_1 f(\tau)} d\tau \\
    s & = \int_{v}^1 \frac{1}{\beta_2 f(\tau)} d\tau.
\end{align*}
Note that since the equations in Eq. \eqref{eq:harko-odes} are functions of the initial values and $R_0$, then $u = v$. We then can find a relation for $s$,
    \begin{align*}
    s & = \int_{u}^1 \frac{1}{\beta_2 f(\tau)} d\tau  \\
    & = \int_{u}^1 \frac{1}{a\beta_1 f(\tau)} d\tau \\ 
    &= \frac{1}{a}\int_{u}^1 \frac{1}{\beta_1 f(\tau)} d\tau \\
    &= \frac{1}{a}t.
\end{align*}
\end{proof}

### 2020-04-22 jss/jss-article-rnw/chapters/ch1-intro.Rnw
Added R0 section. yay

@@
 
-The recent (and currently on-going) COVID-19 global epidemic has galvenized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area \cite{FILL IN}.
+The recent (and currently on-going) COVID-19 global epidemic has galvanized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area of infectious disease epidemiology.
 Infectious disease models models attempt, typically, to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{biggerstaff2016}; or 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \cite{roberts2004}.
-The COVID-19 epidemic highlights how both goals are very important, and that preventing and mitigating the future spread of infectious disease which inflicts enormous burdens on the world: millions of lives lost and billions of dollars spent yearly, is so important.
+A third use of these models is to 3) forecast or project scenarios under pre-specified parameters (i.e.
+'what if's) \citep{}.
+The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.
+Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.
+Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks.
 
-The current epidemic has also highlighted that these models are just one piece of the overall analysis pipeline.
-John Hopkin's \_\_\_ Dashboard and county, state and national tabulation of rudimentary numerical summaries emphasis the need to strong data to assess the spread of infectious disease (multiple \cite{FILL IN}?).
+The current epidemic of COVID-19 also highlights that these models are just one piece of the overall analysis pipeline.
+John Hopkin's \_\_\_ Dashboard and county, state and national tabulation of reported cases and deaths emphasize the need to strong data to assess the spread of infectious disease (multiple \cite{FILL IN}?).
 Stilly, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individal / agent level, network structure or just aggregate flows (review paper \cite{FILL IN}?).
@@
 
+The goal of \pkg{EpicCompare} is to suplement and enhance current infectious disease modeling pipelines as well as to encourage comparisons across these pipelines.
 More specifically, \pkg{EpiCompare} provides epidemiologists a toolkit containing:

### 2020-04-22 jss/jss-article-rnw/chapters/ch2-vis.Rnw
Added R0 section. yay

@@
 \subsection[Visualize]{Visualize} \label{sec:purpose-vis}
+
+One issue that the COVID-19 outbreak has brought to the forefront is the importance of figures and graphs to inform, support, and even persuade lay people of technical and scientific results \cite{}.
+Quantities and trends shown in these visualizations include exponential growth curves, S-curves, peak day and percent of infectious indviduals, ICU capacity, and deaths over time and uncertainty of such estimates.
+
+\pkg{EpiCompare} supports traditional visualizations (e.g.
+\# in condition vs.
+time) and also introduces a novel visualization of epidemics via time free analysis.
+In particular, we use ternary plots that show Susceptible-Infectious-Recovered data/models (SIR).
+The advantages of these ternary plots are 1) a standard way to compare epidemics with different outbreak times and time scales, 2) a novel interpretation of $R_0$, the initial reproduction number under some circumstances, and  3) clear visualization of joint SIR uncertainty via conformal bands.
+
+Ternary plots are sometimes used in chemistry \citep{} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
+Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.
+The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{}), and also happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.
+
+Ternary plots are time-free in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g.
+days vs.
+years) or of SIR data of the same disease in different areas (our first case study in Sec.
+\ref{sec:ex1} focuses on this).
+
+Ternary plots also, under certain conditions, have a special relationship with $R_0$, the initial reproduction number, which has been called the ``most important quantity in epidemiology'' \citep{}.
+The reproduction is itself a time-free concept and is the average number of secondary infections primary infection will generate when the primary infection is added to a completely susceptible population.
+To see the importance of $R_0$, one need only read the newspaper \citep{} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{}.
+
+In regards to traditional $X$ vs.
+time plots, $R_0$ is difficult to visualize.
+For example, as seen in Fig.
+\ref{fig:r0-trad}, it is possible that two models have the same value of $R_0$ but there is no way of knowing that from looking at the graph.
+
+It is possible to identify, at a glance, which epidemic paths have the same value of $R_0$ via ternary plot visualizations, under some circumstances.
+To show, this we first need to introduce the classic, deterministic Kermack and McKendrick SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
+The movement of individuals among the states is given by the following differential equations (Eq.
+\eqref{eq:sir-ode}) where $\beta$ is the average infection rate and $\gamma$ is the average recovery rate,
+\begin{align}\label{eq:sir-ode}
+      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
+      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
+      R^\prime(t) &= \gamma I(t) \nonumber.
+  \end{align}
+
+If we have two SIR models that follow the Kermack and McKendrick SIR model, have the same percentage of individuals in the initial states, and the same value of $R_0$ then the two models will create exactly overlapping paths in a ternary plot.
+
+More formally, let two Kermack and McKendrick (see Eq.
+\eqref{eq:sir-ode}) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
+Assume both models have initial values $(S(0), I(0), R(0))$.
+Let $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
+Equivalently, $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$ for some $a > 0$.
+
+\begin{theorem}\label{thm:sir-scale}
+Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+\end{theorem}
+
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{} and is shown in detail in Appendix \ref{app:proof}.
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.

### 2020-04-22 jss/jss-article-rnw/chapters/ch5-ex1.Rnw
Added R0 section. yay

@@
 EXAMPLE 1
+
+%%% Move to example?  Think I will
+As a motivating example for time free analysis visualizations, we would like to compare the outbreak of COVID-19 across different states.
+Typically, the raw data must be adjusted (or stochastically imputed) in order for \textit{fair} comparisons to be seen.
+A fairly common example of this is when visualizatios show deaths or infections by percent of the area's population.
+Another issue is that outbreaks differ temporally:  the first state (CA?) with confirmed COVID cases occurred on January X while the final state (WV?) occurred on March Y.
+One approach to adjust for the different outbreak times is to show the number of cumulative cases versus the time since the first reported outbreak in the state.
+However, this can result in unexpected features such as shown in Figure \ref{fig:covid-states-time-adjusted}, where Massachusetts has an early reported case but does not have a sustained outbreak until long after that first case.
+Instead of adjusting temporally to the first case, we could instead adjust to when an threshold of cases (raw or percentage) is determined but how this threshold is determined is unclear.
+Time free analysis

### 2020-04-26 jss/jss-article-rnw/chapters/ch1-intro.Rnw
updates to introduction and bibtech

@@
 The recent (and currently on-going) COVID-19 global epidemic has galvanized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area of infectious disease epidemiology.
-Infectious disease models models attempt, typically, to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{biggerstaff2016}; or 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \cite{roberts2004}.
-A third use of these models is to 3) forecast or project scenarios under pre-specified parameters (i.e.
-'what if's) \citep{}.
+Infectious disease models models typlically attempt to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{Biggerstaff2016}, 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \citep{Neal2004}, or 3) forecast or project scenarios under pre-specified parameters (i.e.
+'what if's) such as \citep{}.
 The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.

### 2020-04-27 jss/jss-article-rnw/chapters/ch1-intro.Rnw
minor updates to introduction and shortnames update

@@
 The recent (and currently on-going) COVID-19 global epidemic has galvanized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area of infectious disease epidemiology.
-Infectious disease models models typlically attempt to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{Biggerstaff2016}, 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \citep{Neal2004}, or 3) forecast or project scenarios under pre-specified parameters (i.e.
-'what if's) such as \citep{}.
-The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.
+\ben{\sout{Infectious disease models typically attempt to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{Biggerstaff2016}, 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \citep{Neal2004}, or 3) forecast or project scenarios under pre-specified parameters (i.e.
+'what if's) such as \citep{}.} Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g.
+flue prediction][]{Biggerstaff2016}, 2) analyse past and current epidemics to increase scientific knowledge \citep[e.g.
+historical measle outbreaks][]{Neal2004}, and 3) forcast or project epidemic scenarios under pre-specificed parameters \citep[e.g.
+...][]{cite}.} The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.
 Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.
@@
 The current epidemic of COVID-19 also highlights that these models are just one piece of the overall analysis pipeline.
-John Hopkin's \_\_\_ Dashboard and county, state and national tabulation of reported cases and deaths emphasize the need to strong data to assess the spread of infectious disease (multiple \cite{FILL IN}?).
-Stilly, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individal / agent level, network structure or just aggregate flows (review paper \cite{FILL IN}?).
+\ben{\sout{John Hopkin's \_\_\_ Dashboard and county, state and national tabulation of reported cases and deaths emphasize the need to strong data to assess the spread of infectious disease} University based resources like John Hopkin's \_ and governemnt numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and viaualization can be a important first step in the process (multiple \cite{FILL IN}?).} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individal / agent level, network structure or just aggregate flows (review paper \cite{FILL IN}?).
 All placing individuals / proportions of the populations into different states (e.g.

### 2020-04-30 jss/jss-article-rnw/chapters/ch1-intro.Rnw
update of article to correctly do bibtex. Note: must "compile" from article.tex to get references in document

@@
 \ben{\sout{Infectious disease models typically attempt to 1) predict the spread of current and future epidemics such as the Center of Disease Control and Prevention's (CDCP) yearly ``Predict the Influenza Season Challenge'' \citep{Biggerstaff2016}, 2) learn about past and current epidemics to increase scientific knowledge about the mechanisms behind the spead of disease such as the effect of seasonality in the measles model of \citep{Neal2004}, or 3) forecast or project scenarios under pre-specified parameters (i.e.
-'what if's) such as \citep{}.} Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g.
+'what if's) such as \citep{FILL IN}.} Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g.
 flue prediction][]{Biggerstaff2016}, 2) analyse past and current epidemics to increase scientific knowledge \citep[e.g.
 historical measle outbreaks][]{Neal2004}, and 3) forcast or project epidemic scenarios under pre-specificed parameters \citep[e.g.
-...][]{cite}.} The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.
+...][]{FILL IN}.} The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.
 Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.
@@
 \begin{itemize}
-  \item \textbf{Visualization tools} to visualize three-state epidemics and simulations from three-state\footnote{The most common three-state model that will be examined is the SIR model (with the three states: Susceptible, Infected and Recovered), and will be discussed in more detail in Section \ref{}.} models in a time-free manner using ternary plots and prediction bands.
+  \item \textbf{Visualization tools} to visualize three-state epidemics and simulations from three-state\footnote{The most common three-state model that will be examined is the SIR model (with the three states: Susceptible, Infected and Recovered), and will be discussed in more detail in Section \ref{FILL IN}.} models in a time-free manner using ternary plots and prediction bands.
 For agent-based three-state models we also provide visualization tools to let the user easily explore how different characteristics of the agents relate to different experiences in the epidemic.

### 2020-04-30 jss/jss-article-rnw/chapters/ch3-time-free.Rnw
literature review option and time-free motivation and presention section included

% !Rnw root = ../article.Rnw

Although epidemics are by definition a process that evolves over time, empidemics, including the recent COVID-19, need to oftern be compared in a less time focused, or even time-free way to understand the processes at play. Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as inteligently capture the natural connections between the proportion of individuals in these states. Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-free tools and in this section we present benefits of the time-free analysis, 1) motivation for time-free analysis through $R_0$ (the initial reproduction number), 2) time-free visualization tools for 3-state models (e.g. SIR models), and 3) potential for similar analysis for models with more states (e.g. SEIR models).

\subsection{Motivation through $R_0$}

$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology''. To see the importance of $R_0$, one need only read the newspaper \citep{FILL IN} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{FILL IN}. $R_O$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is most commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models. The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILL IN}).

In regards to traditional $X$ vs. time plots, $R_0$ is difficult to visualize. For example, in a SIR simulation with two models have the same value of $R_0$, one can see in Fig. \ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph.

<<echo = F>>=
# \@Shannon: Make this figure reproducible here plz
# Also --- my vote is that we do this with points and lines? ~ Ben
@

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Bivariate view of \# in each state vs. time. Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.}
    \label{fig:different-scales-standard}
\end{figure}

\subsection{Ternary and time-free visualizations for $R_0$}

It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig. \ref{fig:different-scales}  have the same value of $R_0$ via ``time-free`` visualizations, under some circumstances. Specifically, for SIR model's we propose using ternary to examine an epidemic's trajectory in a more time-free manner.

Ternary plots are sometimes used in chemistry \citep{} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).  Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.  The SIR framework happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.

Ternary plots are time-free in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g. days vs. years) or of SIR data of the same disease in different areas (our first case study in Sec. \ref{sec:ex1} focuses on this).

To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.  The movement of individuals among the states is given by the following differential equations (Eq. \eqref{eq:sir-ode}) where $\beta$ is the average infection rate and $\gamma$ is the average recovery rate,
\begin{align}\label{eq:sir-ode}
      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
      R^\prime(t) &= \gamma I(t) \nonumber.
  \end{align}

If we have two SIR models that follow the Kermack and McKendrick SIR model, have the same percentage of individuals in the initial states, and the same value of $R_0$ then the two models will create exactly overlapping paths in a ternary plot.

More formally, let two Kermack and McKendrick (see Eq. \eqref{eq:sir-ode}) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$. Assume both models have initial values $(S(0), I(0), R(0))$.  Let $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.  Equivalently, $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$ for some $a > 0$.

\begin{theorem}\label{thm:sir-scale}
Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Appendix \ref{app:proof}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model. Taking the sample simulations from Fig. \ref{fig:different-scales-standard}, Fig. \ref{fig:different-scales-tern} presents these models in a ternary plot.

<<echo = F>>=
# \@Shannon: Make this figure reproducible here plz
# Also --- my vote is that we do this with points and lines? ~ Ben
@

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Ternary view of \# in each state.  Both this Fig. and Fig. \ref{fig:different-scales-standard} display the same two sets of data.  Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.  While there are obvious are differences in Fig.\ref{fig:different0-scale-standard}, the data sets look quite similar in the ternary view.}
    \label{fig:different-scales-tern}
\end{figure}

In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-free space, which is presented in more details in Section \ref{sec:time-free-in-simplex}.

\subsection{Simplexes beyond ternary plots and time-free comparison tools} \label{sec:time-free-in-simplex}

\ben{Maybe should include something like "beyond ternary plots as a tool for time-free analysis..."}

The basis of time-free analysis in ternary plots is extended to higher dimensional models (i.e. models with more states) in \pkg{EpiCompare} as well. The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions. This same scaling for larger models (i.e. with more states) can be done onto different simplexes. In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a 1 dimension less space through the simplexical structure of the data.

\ben{I think this would be a good place to actually present the geometric tools we'd like to use for comparison of different models... - @Shannon - thoughts?}

### 2020-04-30 jss/jss-article-rnw/chapters/ch4-orientation.Rnw
more outline provided - "Orientation" section proposed

% !Rnw root = ../article.Rnw

\ben{This potentially could similar to that done in EpiModel's \citep{Jenness2018} orientation section, but I'm not sure - it might also depend upon which journal we aim to publish in. I also expect that this section will mirror how we present the pipeline in the Literature review.}

### 2020-04-30 jss/jss-article-rnw/chapters/ch3-time-free.Rnw
added comments relative to the outline of the paper

@@
 \subsection{Ternary and time-free visualizations for $R_0$}
+
+\ben{My original notes for this section note: ``Ternary plots (pitch from Shannon’s thesis) - being useful to assess.
+`Peak shows corresponds to the same event in the time analysis' (maybe?)''.
+I don't necessarily think this section perfectly does that - so I'm open to having it include that or just deciding this is better.}
 

### 2020-05-02 jss/jss-article-rnw/chapters/ch3-time-free.Rnw
added 2 new articles and 1 citation to document.

@@
 $R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology''.
-To see the importance of $R_0$, one need only read the newspaper \citep{FILL IN} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{FILL IN}.
+To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{FILL IN}.
 $R_O$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is most commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.

### 2020-05-02 jss/jss-article-rnw/chapters/ch3-time-free.Rnw
breaking up chapter 4

--- 2020-04-30 jss/jss-article-rnw/chapters/ch3-time-free.Rnw
@@
 
-\subsection{Motivation through $R_0$}
-
-$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology''.
-To see the importance of $R_0$, one need only read the newspaper \citep{FILL IN} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{FILL IN}.
-$R_O$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is most commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
-The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILL IN}).
-
-In regards to traditional $X$ vs.
-time plots, $R_0$ is difficult to visualize.
-For example, in a SIR simulation with two models have the same value of $R_0$, one can see in Fig.
-\ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph.
-
-<<echo = F>>=
-# \@Shannon: Make this figure reproducible here plz
-# Also --- my vote is that we do this with points and lines? ~ Ben
-@
-
-\begin{figure}[!ht]
-%% Code to reproduce is in inst/paper_figs/different-time-scales.R
-    \centering
-    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
-    \caption{Bivariate view of \# in each state vs. time. Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.}
-    \label{fig:different-scales-standard}
-\end{figure}
-
-\subsection{Ternary and time-free visualizations for $R_0$}
-
-It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig.
-\ref{fig:different-scales}  have the same value of $R_0$ via ``time-free`` visualizations, under some circumstances.
-Specifically, for SIR model's we propose using ternary to examine an epidemic's trajectory in a more time-free manner.
-
-Ternary plots are sometimes used in chemistry \citep{} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
-Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.
-The SIR framework happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.
-
-Ternary plots are time-free in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g.
-days vs.
-years) or of SIR data of the same disease in different areas (our first case study in Sec.
-\ref{sec:ex1} focuses on this).
-
-To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
-The movement of individuals among the states is given by the following differential equations (Eq.
-\eqref{eq:sir-ode}) where $\beta$ is the average infection rate and $\gamma$ is the average recovery rate,
-\begin{align}\label{eq:sir-ode}
-      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
-      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
-      R^\prime(t) &= \gamma I(t) \nonumber.
-  \end{align}
-
-If we have two SIR models that follow the Kermack and McKendrick SIR model, have the same percentage of individuals in the initial states, and the same value of $R_0$ then the two models will create exactly overlapping paths in a ternary plot.
-
-More formally, let two Kermack and McKendrick (see Eq.
-\eqref{eq:sir-ode}) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
-Assume both models have initial values $(S(0), I(0), R(0))$.
-Let $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
-Equivalently, $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$ for some $a > 0$.
-
-\begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
-\end{theorem}
-
-The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Appendix \ref{app:proof}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.
-Taking the sample simulations from Fig.
-\ref{fig:different-scales-standard}, Fig.
-\ref{fig:different-scales-tern} presents these models in a ternary plot.
-
-<<echo = F>>=
-# \@Shannon: Make this figure reproducible here plz
-# Also --- my vote is that we do this with points and lines? ~ Ben
-@
-
-\begin{figure}[!ht]
-%% Code to reproduce is in inst/paper_figs/different-time-scales.R
-    \centering
-    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
-    \caption{Ternary view of \# in each state.
-Both this Fig.
-and Fig.
-\ref{fig:different-scales-standard} display the same two sets of data.
-Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
-While there are obvious are differences in Fig.\ref{fig:different0-scale-standard}, the data sets look quite similar in the ternary view.}
-    \label{fig:different-scales-tern}
-\end{figure}
-
-In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-free space, which is presented in more details in Section \ref{sec:time-free-in-simplex}.
-
-\subsection{Simplexes beyond ternary plots and time-free comparison tools} \label{sec:time-free-in-simplex}
-
-\ben{Maybe should include something like "beyond ternary plots as a tool for time-free analysis..."}
-
-The basis of time-free analysis in ternary plots is extended to higher dimensional models (i.e.
-models with more states) in \pkg{EpiCompare} as well.
-The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions.
-This same scaling for larger models (i.e.
-with more states) can be done onto different simplexes.
-In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a 1 dimension less space through the simplexical structure of the data.
-
-\ben{I think this would be a good place to actually present the geometric tools we'd like to use for comparison of different models... - @Shannon - thoughts?}
+\SweaveInput{ch3.1-r0.Rnw}
+\SweaveInput{ch3.2-ternary.Rnw}
+\SweaveInput{ch3.3-simplex-and-beyond.Rnw}

### 2020-05-02 jss/jss-article-rnw/chapters/ch3.1-r0.Rnw
breaking up chapter 4

% !Rnw root = ../article.Rnw

\subsection{Motivation through $R_0$}

$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology''. To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{FILL IN}. $R_O$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is most commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models. The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILL IN}).

In regards to traditional $X$ vs. time plots, $R_0$ is difficult to visualize. For example, in a SIR simulation with two models have the same value of $R_0$, one can see in Fig. \ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph.

<<echo = F>>=
# \@Shannon: Make this figure reproducible here plz
# Also --- my vote is that we do this with points and lines? ~ Ben
@

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Bivariate view of \# in each state vs. time. Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.}
    \label{fig:different-scales-standard}
\end{figure}

### 2020-05-02 jss/jss-article-rnw/chapters/ch3.2-ternary.Rnw
breaking up chapter 4

--- 2020-04-21 jss/jss-article-rnw/chapters/ch2-vis.Rnw
@@
 
-\subsection[Visualize]{Visualize} \label{sec:purpose-vis}
+\subsection{Ternary and time-free visualizations for $R_0$}
+
+\ben{My original notes for this section note: ``Ternary plots (pitch from Shannon’s thesis) - being useful to assess.
+`Peak shows corresponds to the same event in the time analysis' (maybe?)''.
+I don't necessarily think this section perfectly does that - so I'm open to having it include that or just deciding this is better.}
+
+It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig.
+\ref{fig:different-scales}  have the same value of $R_0$ via ``time-free`` visualizations, under some circumstances.
+Specifically, for SIR model's we propose using ternary to examine an epidemic's trajectory in a more time-free manner.
+
+Ternary plots are sometimes used in chemistry \citep{} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
+Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.
+The SIR framework happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.
+
+Ternary plots are time-free in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g.
+days vs.
+years) or of SIR data of the same disease in different areas (our first case study in Sec.
+\ref{sec:ex1} focuses on this).
+
+To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
+The movement of individuals among the states is given by the following differential equations (Eq.
+\eqref{eq:sir-ode}) where $\beta$ is the average infection rate and $\gamma$ is the average recovery rate,
+\begin{align}\label{eq:sir-ode}
+      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
+      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
+      R^\prime(t) &= \gamma I(t) \nonumber.
+  \end{align}
+
+If we have two SIR models that follow the Kermack and McKendrick SIR model, have the same percentage of individuals in the initial states, and the same value of $R_0$ then the two models will create exactly overlapping paths in a ternary plot.
+
+More formally, let two Kermack and McKendrick (see Eq.
+\eqref{eq:sir-ode}) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
+Assume both models have initial values $(S(0), I(0), R(0))$.
+Let $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
+Equivalently, $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$ for some $a > 0$.
+
+\begin{theorem}\label{thm:sir-scale}
+Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+\end{theorem}
+
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Appendix \ref{app:proof}.
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.
+Taking the sample simulations from Fig.
+\ref{fig:different-scales-standard}, Fig.
+\ref{fig:different-scales-tern} presents these models in a ternary plot.
+
+<<echo = F>>=
+# \@Shannon: Make this figure reproducible here plz
+# Also --- my vote is that we do this with points and lines? ~ Ben
+@
+
+\begin{figure}[!ht]
+%% Code to reproduce is in inst/paper_figs/different-time-scales.R
+    \centering
+    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
+    \caption{Ternary view of \# in each state.
+Both this Fig.
+and Fig.
+\ref{fig:different-scales-standard} display the same two sets of data.
+Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
+While there are obvious are differences in Fig.\ref{fig:different0-scale-standard}, the data sets look quite similar in the ternary view.}
+    \label{fig:different-scales-tern}
+\end{figure}
+
+In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-free space, which is presented in more details in Section \ref{sec:time-free-in-simplex}.

### 2020-05-02 jss/jss-article-rnw/chapters/ch3.3-simplex-and-beyond.Rnw
breaking up chapter 4

% !Rnw root = ../article.Rnw

\subsection{Simplexes beyond ternary plots and time-free comparison tools} \label{sec:time-free-in-simplex}

\ben{Maybe should include something like "beyond ternary plots as a tool for time-free analysis..."}

The basis of time-free analysis in ternary plots is extended to higher dimensional models (i.e. models with more states) in \pkg{EpiCompare} as well. The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions. This same scaling for larger models (i.e. with more states) can be done onto different simplexes. In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a 1 dimension less space through the simplexical structure of the data.

\ben{I think this would be a good place to actually present the geometric tools we'd like to use for comparison of different models... - @Shannon - thoughts?}

Tools to do comparisons beyond visualizations focus on created prediction bands \citep{Vovk2005)

### 2020-05-06 R/theme-sir.R
Added a small theme with arrows

@@ -1,0 +1,14 @@
+## SKG
+## ADDING SIR THEME
+## May 6, 2020
+
+
+#' Add a portable theme for SIR ternary plots
+#' 
+#' @return theme to add to ggtern plots
+#' @export
+theme_sir <- function(){
+  ggplot2::theme_bw() +
+    ggtern::theme_hideticks() +
+    ggtern::theme_showarrows() 
+}

### 2020-05-07 README.Rmd
minor documention updates and correction to S3 for R4.0.0

@@
     coord_tern() +
-    labs(x = "S", y = "I", z = "R", title = "Town Analysis")
+    labs(x = "S", y = "I", z = "R", title = "Town Analysis") +
+    theme_sir()
 ```
@@
             color = "School Grade",
-            title = "Town analysis by grade") 
+            title = "Town analysis by grade") +
+       theme_sir()
 ```
@@
     geom_line(aes(x = S, y = I, z = R, group = sim), alpha = .1) +
-    coord_tern()
+    coord_tern() +
+    theme_sir()
 ```
@@
 Basically - there are problems when the `compute_group` wants to make multiple `pieces`/`groups` - and it is similar to the problem that if you do something like `aes(color = var1, group = var2)` you may actually want to do `aes(color = var1, group = paste(var1, var2))`, if there are the same `var2` values across different `var1` values but they don't mean they should be grouped together.
+4.
+Now that `R` have come out with version >= 4.0.0, we now need to call `.S3method("method", "class")` to define the connection for `S3` methods (e.g.
+`method.class` function), which we have for the `contained` function.
 

### 2020-05-07 R/aaa.R

@@ -47,3 +47,12 @@ update_approved_layers <- function(stat_names = .newstat,
 }
 
 update_approved_layers()
+
+
+#' logic to check if R is >= 4
+#'
+#' @return logical value (boolean)
+#' @export
+r_new_interface <- function() {
+  as.numeric(R.version$major) >= 4
+}

### 2020-05-07 R/containment-and-band-creation.R

@@ -436,6 +436,7 @@ contained <- function(x, y, .lower_simplex_project = TRUE){
   UseMethod("contained")
 }
 
+
 #' @rdname contained
 contained.delta_ball_structure <- function(x, y,
                                            .lower_simplex_project = TRUE){
@@ -451,6 +452,9 @@ contained.delta_ball_structure <- function(x, y,
   interior_dist <- dist_mat %>% apply(2, min) %>% max
 
   return(interior_dist <= delta)
+}
+if (r_new_interface()){
+  .S3method("contained", "delta_ball_structure")
 }
 
 #' @rdname contained
@@ -471,7 +475,9 @@ contained.convex_hull_structure <- function(x, y,
   }
   return(all(contained_vec))
 }
-
+if (r_new_interface()){
+  .S3method("contained", "convex_hull_structure")
+}
 
 #' Create linear map to move simplex with p vertices to p-1 dimensional space
 #'

### 2020-05-07 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw

% !Rnw root = ../article.Rnw

\section[SKG tries an Intro]{SKG Intro} \label{sec:skg_intro}

COVID-19 has changed the course of scientific advancement as evidenced by the plethora of studies, modeling tools, and forecasts made publicly available on arXiv and medArXiv as well as peer-reviewed journals.  As a consequence of the large number of models and results, it is more difficult to parse through the models and results, which are often nuanced and specific to a certain region of the disease analysis pipeline.  By disease analysis pipeline, we refer to the process of how data is collected, processed and explored, statistically modeled, analyzed, and finally diseminated to others.

We introduce the \texttt{R} package \pkg{EpiCompare} which is used to supplement and enhance (as opposed to supplant and compete with) existing infectious disease analysis software at every step in the analysis pipeline.  We offer a suite of functions and vignettes to aid with visualization, comparison, and conversion of existing tools.  The goal of \pkg{EpiCompare} is a streamlined process to standardize output from a variety of epidemic models, and when possible to compare disease data and disease models to one another through standard comparison tools (e.g. prevalence vs. time plots) along with novel contributions through what we call time-free analysis.

More specifically, \pkg{EpiCompare} complements and enhances the disease analysis pipeline in the following areas:

\begin{itemize}
\item \textbf{Data harmonization and Exploratory Data Analysis (EDA)}.  \pkg{EpiCompare} contains \pkg{tidyverse}-style  functions to harmonize output from popular infectious disease software so they can be simply compared to one another.  Additionally, we provide tools to aggregate and compare three standard formats of infectious disease data.

\item \textbf{Data analysis and comparison}. \pkg{EpiCompare} complements existing conventional data visualizations (e.g. prevalence vs. time) through time-free analysis, which we introduce via the well-known epidemiological concept of $R_0$, the infectious reproduction number.

Additionally, we support statistical comparison of different models via conformal bands (e.g. prediction bands), which can be used to determine differences among models or sub-groups within models.

\end{itemize}

The paper is arranged as follows.

### 2020-05-11 jss/jss-article-rnw/chapters/ch1-intro-v2-woven.tex
Made an intro I'm pretty happy with?

% !Rnw root = ../article.Rnw

\section[SKG tries an Intro]{SKG Intro} \label{sec:skg_intro}

As scientific resources have increased through the years, the number of models to analyze a single phenomenon has also increased.  This is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \citep{}.  As both the quantity of data and models increase, the `compare and assess' step of the data analysis pipeline of infectious disease also increases in importance.

In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases as a) data pre-processing, b) exploratory data analysis (EDA), c) modeling and simulating, d) post-processing, and e) comparison and assessment, where each previous part influences the next.  We introduce \pkg{EpiCompare}, a \texttt{R} package used to complement and enhance the available tools for every part of the pipeline, with special emphasis on tools a) to harmonize input and output data and b) to compare multiple infectious disease data sets or infectious disease models to one another.

The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create \textit{fair} comparisons among models developed for outbreaks and outbreak data.  To reach this goal, \pkg{EpiCompare}

\begin{itemize}
  \item adds \textbf{harmonization} tools for a) standardizing outputs from common infectious disease model software  and b) converting individual/agent level data into a format that is comparable to aggregate models
\end{itemize}

The paper is arranged as follows.

### 2020-05-11 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw
Made an intro I'm pretty happy with?

@@
 
-\section[SKG tries an Intro]{SKG Intro} \label{sec:skg_intro}
+\section[Introduction]{Introduction} \label{sec:skg_intro}
 
-COVID-19 has changed the course of scientific advancement as evidenced by the plethora of studies, modeling tools, and forecasts made publicly available on arXiv and medArXiv as well as peer-reviewed journals.
-As a consequence of the large number of models and results, it is more difficult to parse through the models and results, which are often nuanced and specific to a certain region of the disease analysis pipeline.
-By disease analysis pipeline, we refer to the process of how data is collected, processed and explored, statistically modeled, analyzed, and finally diseminated to others.
+As the availability of scientific resources has increased through the decades, the number of models to analyze a particular disease outbreak has also increased.
+This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \citep{}.
+As both the quantity of data and models increase, so does the need for standard and fair comparisons of models, as according to the data analysis pipeline of infectious diseases.
 
-We introduce the \texttt{R} package \pkg{EpiCompare} which is used to supplement and enhance (as opposed to supplant and compete with) existing infectious disease analysis software at every step in the analysis pipeline.
-We offer a suite of functions and vignettes to aid with visualization, comparison, and conversion of existing tools.
-The goal of \pkg{EpiCompare} is a streamlined process to standardize output from a variety of epidemic models, and when possible to compare disease data and disease models to one another through standard comparison tools (e.g.
-prevalence vs.
-time plots) along with novel contributions through what we call time-free analysis.
+We introduce \pkg{EpiCompare}, a \texttt{R} package used to complement and enhance the available tools for every part of the infectious disease analysis pipeline, with special emphasis on tools a) to harmonize input and output data and b) to compare multiple infectious disease data sets or infectious disease models to one another.
+  In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment, where each previous part of the pipeline influences the next.
 
-More specifically, \pkg{EpiCompare} complements and enhances the disease analysis pipeline in the following areas:
-
-\begin{itemize}
-\item \textbf{Data harmonization and Exploratory Data Analysis (EDA)}.
-\pkg{EpiCompare} contains \pkg{tidyverse}-style  functions to harmonize output from popular infectious disease software so they can be simply compared to one another.
-Additionally, we provide tools to aggregate and compare three standard formats of infectious disease data.
-
-\item \textbf{Data analysis and comparison}.
-\pkg{EpiCompare} complements existing conventional data visualizations (e.g.
-prevalence vs.
-time) through time-free analysis, which we introduce via the well-known epidemiological concept of $R_0$, the infectious reproduction number.
-
-Additionally, we support statistical comparison of different models via conformal bands (e.g. prediction bands), which can be used to determine differences among models or sub-groups within models.
-
-\end{itemize}
+The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
+To this end, \pkg{EpiCompare}
+\begin{enumerate}
+  \ite Standardizes outputs from common infectious disease model software as a form of data pre-processing
+  \item Complements existing software for data visualization and EDA
+  \item Provides support for modeling and simulation
+  \item Harmonizes data and model formats to be used in conjunction with one another
+  \item Assesses models through time free analysis and conformal prediction bands and comparisons.
+\end{enumerate}
+Although \pkg{EpiCompare} aids in every part of the pipeline, our tools especially emphasize improvements for pipeline parts (1), (4), and (5) which can be summarized as data input and output harmonization and contributions to fairly comparing models and data to one another.
 
 The paper is arranged as follows.
+In Section \ref{sec:lit}, we review some of the prior work in the infectious disease data analysis pipeline and describe how \pkg{EpiCompare} improves this process.
+In Section \ref{sec:data-harmony}, we describe how we use \pkg{tidyverse}-style tools to harmonize data for work in a usable format.
+In Section \ref{sec:assess}, we discuss our novel contributions of comparing and assessing outbreak data and models.
+In Section \ref{sec:vignette}, we provide a full pipeline analysis of a measles outbreak, including code.
+Finally, in Section \ref{sec:discuss}, we summarize our contributions and discuss future work.

### 2020-05-12 R/data.R
Added a SIRS data set

@@ -1,3 +1,24 @@
+#' Example of SIRS model generated  from the \code{pomp} package.
+#' 
+#' @description  This is a deterministic SIRS model where 
+#' the infection parameter \eqn{\beta = 0.2} and the recovery from infection parameter is
+#' parameter is \eqn{\gamma = 0.1}.  The recovery to susceptible parameter is \eqn{\eta = .05}.  The initial number of susceptible is \eqn{S(0) = 990}
+#' and the initial number of infectious is \eqn{I(0)=10}.  There are 100 steps.
+#' So in summary, \eqn{S^\prime = -\beta SI / N + \eta R}, \eqn{I^\prime} = \beta S I /N -\gamma I}
+#' and \eqn{R^\prime = \gamma I - \eta R}.
+#' @format The object is a data frame.
+#' @examples 
+#' library(ggplot2)
+#' library(tidyr)
+#' head(sirs_data)
+#' ggplot(data = sirs_data %>%
+#'          tidyr::pivot_longer(-c(time, .id)),
+#'           aes(x = time, y = value,
+#'                                                 group = name,
+#'                                                  col = name)) + geom_line()
+"sirs_data"
+
+
 #' Example SIS model  of class `dcm` from the \code{EpiModel} package.
 #' 
 #' @description  This is a deterministic SIS model where 

### 2020-05-12 data-raw/pomp_data-sirs.R

@@ -1,0 +1,45 @@
+library(pomp)
+
+## From the pomp example from: https://kingaa.github.io/short-course/parest/odes.html#exercise-sirs-model
+## Makes a seasonal SIR model
+
+## Modified for a SIRS model like here: https://www.idmod.org/docs/typhoid/model-sir.html#sirs-model
+## This is deterministic
+
+sirs.ode <- Csnippet("
+  DS = -Beta*S*I/N + eta * R;
+  DI = Beta*S*I/N-gamma*I;
+  DR = gamma*I - eta * R;
+")
+
+init2 <- Csnippet("
+  S = S_0;
+  I = I_0;
+  R = N-S_0-I_0;
+")
+
+sirs <- pomp(data=data.frame(time=seq(0,100, by = 1),cases=NA),
+     times="time",t0= 0,
+     skeleton=vectorfield(sirs.ode),
+     rinit=init2,
+     statenames=c("S","I","R"),
+     paramnames=c("Beta","gamma","eta","S_0","I_0","N")
+)
+
+params <- c(Beta = .2, gamma = .1, eta = .05,
+            S_0 = 990, I_0 = 10, N = 1000)
+
+sirs_data <- trajectory(sirs, params = params, format = "d")
+
+library(ggplot2)
+library(tidyr)
+ggplot(data = sirs_data %>%
+         tidyr::pivot_longer(-c(time, .id)), aes(x = time, y = value,
+                                                 group = name,
+                                                 col = name)) + geom_line()
+
+ 
+usethis::use_data(sirs_data, overwrite = TRUE)
+
+
+

### 2020-05-14 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw
Added a SIRS data set
ben's first edits to chapter 1 intro (version 3)

@@
 As the availability of scientific resources has increased through the decades, the number of models to analyze a particular disease outbreak has also increased.
-This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \citep{}.
+This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \ben{\citep[e.g.][]{Sanche2020}}, \ben{and is well documented in many literature reviews in epidemology \citep[e.g.][]{Canzani2015}}.
 As both the quantity of data and models increase, so does the need for standard and fair comparisons of models, as according to the data analysis pipeline of infectious diseases.
 
-We introduce \pkg{EpiCompare}, a \texttt{R} package used to complement and enhance the available tools for every part of the infectious disease analysis pipeline, with special emphasis on tools a) to harmonize input and output data and b) to compare multiple infectious disease data sets or infectious disease models to one another.
-  In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment, where each previous part of the pipeline influences the next.
+We introduce \pkg{EpiCompare}, a \texttt{R} package used to complement and enhance the available tools for every part of the infectious disease analysis pipeline, with special emphasis on tools \ben{\sout{a)}}  to compare multiple infectious disease data sets or infectious disease models to one another  and \ben{\sout{b)}} to harmonize input and output data\ben{\footnote{\ben{I rearranged these 2 since we might want to push the new a) more?}}}.
+  In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next.
+\ben{In Figure \ref{fig:pipeline} we also highlight where \pkg{EpiCompare} can aid in this pipeline.\footnote{\ben{I just feel like we want to cycle back to emphasis the package - but I'm open.}}}
+
+\begin{figure}[!ht]
+%% Code to reproduce is in inst/paper_figs/different-time-scales.R
+    \centering
+    \includegraphics[width = 1\textwidth]{images/pipeline1.png}
+    \caption{\ben{Temporary Figure...}}
+    \label{fig:pipeline}
+\end{figure}
 
 The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
-To this end, \pkg{EpiCompare}
+\ben\{pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-free tools (motivated in more detail in Section \ref{sec:time-free}).} \ben{More broadly,\sout{To this end,}} \pkg{EpiCompare} \ben{aids the data analysis pipeline for infecetious diseases by}
 \begin{enumerate}
-  \ite Standardizes outputs from common infectious disease model software as a form of data pre-processing
-  \item Complements existing software for data visualization and EDA
-  \item Provides support for modeling and simulation
-  \item Harmonizes data and model formats to be used in conjunction with one another
-  \item Assesses models through time free analysis and conformal prediction bands and comparisons.
+  \item Standardiz\ben{\sout{es}ing} outputs from common \ben{\proglang{R}-based} infectious disease model software \ben{\sout{as a form of data pre-processing} and common data storage types}
+  \item \ben{\sout{Complements existing software for data visualization and EDA} Expanding upon existing data visualization and EDA tools}
+  \item Provides support for modeling and simulation\ben{\footnote{\ben{What are you trying to say here?}}}
+  \item \ben{\sout{Harmonizes data and model formats to be used in conjunction with one another}\footnote{\ben{I think this is the same as the first bullet?}}}
+  \item \ben{Aiding in comparison and assessment of \sout{Assesses}} models through time free analysis and conformal prediction bands and comparisons.
 \end{enumerate}
@@
 The paper is arranged as follows.
-In Section \ref{sec:lit}, we review some of the prior work in the infectious disease data analysis pipeline and describe how \pkg{EpiCompare} improves this process.
+In Section \ref{sec:lit}, we review some of the prior work in the infectious disease data analysis pipeline and describe how \pkg{EpiCompare} improves this process\ben{\footnote{\ben{Currently we break this into sections 2 and 3 - discuss?}}}.
 In Section \ref{sec:data-harmony}, we describe how we use \pkg{tidyverse}-style tools to harmonize data for work in a usable format.

### 2020-05-14 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw
similar change to Rnw

@@
 The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
-\ben\{pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-free tools (motivated in more detail in Section \ref{sec:time-free}).} \ben{More broadly,\sout{To this end,}} \pkg{EpiCompare} \ben{aids the data analysis pipeline for infecetious diseases by}
+\ben
+{\pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-free tools (motivated in more detail in Section \ref{sec:time-free}).} \ben{More broadly,\sout{To this end,}} \pkg{EpiCompare} \ben{aids the data analysis pipeline for infecetious diseases by}
 \begin{enumerate}

### 2020-05-14 jss/jss-article-rnw/chapters/ch3.1-r0.Rnw
clean up of later section

@@
 
-$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology''.
-To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{FILL IN}.
+$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[\ben{are you ok being cited here?}][]{Gallagher2020}.
+To see the importance of $R_0$, one need only read the newspaper \citep{Fisher} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep[\ben{at shannon: is this an ok reference or do you want a better one?}][]{Aronson2020}.
 $R_O$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is most commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
@@
     \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
-    \caption{Bivariate view of \# in each state vs. time. Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.}
+    \caption{Bivariate view of \# in each state vs.
+time.
+Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\ben{We probably need to define $\beta/\gamma$.}}
     \label{fig:different-scales-standard}

### 2020-05-14 jss/jss-article-rnw/chapters/ch3.2-ternary.Rnw
clean up of later section

@@
 
-\subsection{Ternary and time-free visualizations for $R_0$}
+\subsection{Ternary and time-free visualizations for $R_0$} \label{sec:ternary}
 
@@
 It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig.
-\ref{fig:different-scales}  have the same value of $R_0$ via ``time-free`` visualizations, under some circumstances.
-Specifically, for SIR model's we propose using ternary to examine an epidemic's trajectory in a more time-free manner.
+\ref{fig:different-scales-tern}  have the same value of $R_0$ via ``time-free'' visualizations, under some circumstances.
+Specifically, for SIR models we propose using ternary to examine an epidemic's trajectory in a more time-free manner.
 
-Ternary plots are sometimes used in chemistry \citep{} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
+Ternary plots are sometimes used in chemistry \citep[\ben{shannon: is this a good reference?
+- was in your thesis.}][]{Gillespie1976} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
 Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.
@@
 
-To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
+To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick \cite{CITE THIS?} SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
 The movement of individuals among the states is given by the following differential equations (Eq.
@@
 Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
-While there are obvious are differences in Fig.\ref{fig:different0-scale-standard}, the data sets look quite similar in the ternary view.}
+\ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different0-scale-standard}, the data sets look quite similar in the ternary view.}
     \label{fig:different-scales-tern}

### 2020-05-14 jss/jss-article-rnw/chapters/ch3.3-simplex-and-beyond.Rnw
clean up of later section

@@
 
-\ben{Maybe should include something like "beyond ternary plots as a tool for time-free analysis..."}
-
-The basis of time-free analysis in ternary plots is extended to higher dimensional models (i.e.
-models with more states) in \pkg{EpiCompare} as well.
-The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions.
+\ben{\sout{Maybe should include something like "beyond ternary plots as a tool for time-free analysis..."}}
+\ben{Beyond ternary plots, and the associated SIR models, we can extend the time-free approach described in Section \ref{sec:ternary} to higher dimensions (i.e.
+models with more states).\sout{The basis of time-free analysis in ternary plots is extended to higher dimensional models (i.e.
+models with more states) in \pkg{EpiCompare} as well.}} The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions.
 This same scaling for larger models (i.e.
@@
 
-Tools to do comparisons beyond visualizations focus on created prediction bands \citep{Vovk2005)
+\ben{Should introduce 
+\begin{enumerate}
+\item thinking about epidemics as functions in high dimensional space
+\item containment of functions
+\item comparison of sets with Hausdorff distance 
+\end{enumerate}
+}

### 2020-05-15 jss/jss-article-rnw/chapters/ch3-time-free.Rnw

@@
 
-Although epidemics are by definition a process that evolves over time, empidemics, including the recent COVID-19, need to oftern be compared in a less time focused, or even time-free way to understand the processes at play.
-Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as inteligently capture the natural connections between the proportion of individuals in these states.
-Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-free tools and in this section we present benefits of the time-free analysis, 1) motivation for time-free analysis through $R_0$ (the initial reproduction number), 2) time-free visualization tools for 3-state models (e.g.
+Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a less time focused, or even time-free way to understand the processes at play.
+Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
+Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-free approaches and in this section we present benefits of the time-free analysis, with 1) motivation for time-free analysis through $R_0$ (the initial reproduction number), 2) time-free visualization tools for 3-state models (e.g.
 SIR models), and 3) potential for similar analysis for models with more states (e.g.

### 2020-05-19 jss/jss-article-rnw/chapters/ch3.3-simplex-and-beyond.Rnw
update of simplex and model comparison in motivation section - this is very rough update.

@@
 models with more states).\sout{The basis of time-free analysis in ternary plots is extended to higher dimensional models (i.e.
-models with more states) in \pkg{EpiCompare} as well.}} The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions.
+models with more states) in \pkg{EpiCompare} as well.}} The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling).
 This same scaling for larger models (i.e.
@@
 
-\ben{I think this would be a good place to actually present the geometric tools we'd like to use for comparison of different models... - @Shannon - thoughts?}
-
-\ben{Should introduce 
-\begin{enumerate}
-\item thinking about epidemics as functions in high dimensional space
-\item containment of functions
-\item comparison of sets with Hausdorff distance 
-\end{enumerate}
-}
+\ben{In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes.
+To aid in the comparison of models and epidemics we provide multiple tools.
+The first of which is, through multiple simulations under specific model parameters we can generate multiple time-series under the same assumptions.
+In \pkg{EpiCompare} we provide ways (leveraging conformal inference \cite{Vovk2005}) to create prediction regions for a true epidemic under the model assumptions.
+These regions require representing multi-dimensional structures for functions to completely contain epidemics.
+We extend off of papers like \citet{Dalmasso2019a} and \citet{LeRoy2020} to create these bands.
+These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets.
+We recommend using the Hausdorff distance to compare such sets as it captures how much bigger the sets would have to expand to cover each other, and is defined mathematically as
+\[
+d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
+\]}

### 2020-05-19 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw
clean up of paper from May 18th discussion

--- 2020-05-14 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw
@@
 As the availability of scientific resources has increased through the decades, the number of models to analyze a particular disease outbreak has also increased.
-This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \ben{\citep[e.g.][]{Sanche2020}}, \ben{and is well documented in many literature reviews in epidemology \citep[e.g.][]{Canzani2015}}.
+This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \ben{\citep[e.g.][]{Sanche2020a}}, \ben{and is well documented in many literature reviews in epidemology \citep[e.g.][]{Canzani2015}}.
 As both the quantity of data and models increase, so does the need for standard and fair comparisons of models, as according to the data analysis pipeline of infectious diseases.
 
-We introduce \pkg{EpiCompare}, a \texttt{R} package used to complement and enhance the available tools for every part of the infectious disease analysis pipeline, with special emphasis on tools \ben{\sout{a)}}  to compare multiple infectious disease data sets or infectious disease models to one another  and \ben{\sout{b)}} to harmonize input and output data\ben{\footnote{\ben{I rearranged these 2 since we might want to push the new a) more?}}}.
+We introduce \pkg{EpiCompare}, a \texttt{R} package used to complement and enhance the available tools for every part of the infectious disease analysis pipeline, with special emphasis on tools to compare multiple infectious disease data sets or infectious disease models to one another  and to harmonize input and output data.
   In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next.
-\ben{In Figure \ref{fig:pipeline} we also highlight where \pkg{EpiCompare} can aid in this pipeline.\footnote{\ben{I just feel like we want to cycle back to emphasis the package - but I'm open.}}}
+In Figure \ref{fig:pipeline} we also highlight where \pkg{EpiCompare} can aid in this pipeline.
 
@@
 The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
-\ben\{pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-free tools (motivated in more detail in Section \ref{sec:time-free}).} \ben{More broadly,\sout{To this end,}} \pkg{EpiCompare} \ben{aids the data analysis pipeline for infecetious diseases by}
+\pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-free tools (motivated in more detail in Section \ref{sec:time-free}).More broadly, \pkg{EpiCompare} aids the data analysis pipeline for infecetious diseases by
 \begin{enumerate}
-  \item Standardiz\ben{\sout{es}ing} outputs from common \ben{\proglang{R}-based} infectious disease model software \ben{\sout{as a form of data pre-processing} and common data storage types}
-  \item \ben{\sout{Complements existing software for data visualization and EDA} Expanding upon existing data visualization and EDA tools}
-  \item Provides support for modeling and simulation\ben{\footnote{\ben{What are you trying to say here?}}}
-  \item \ben{\sout{Harmonizes data and model formats to be used in conjunction with one another}\footnote{\ben{I think this is the same as the first bullet?}}}
-  \item \ben{Aiding in comparison and assessment of \sout{Assesses}} models through time free analysis and conformal prediction bands and comparisons.
+  \item Standardizing outputs from common \proglang{R}-based infectious disease model software and common data storage types
+  \item Expanding upon existing data visualization and EDA tools
+  \item \ben{\sout{Provides support for modeling and simulation} Providing a interface for simple ODE-style agent based modeling.}
+  \item Aiding in comparison and assessment of models through time free analysis and conformal prediction bands and comparisons.
 \end{enumerate}

### 2020-05-19 jss/jss-article-rnw/chapters/ch3.1-r0.Rnw
clean up of paper from May 18th discussion

@@
 $R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[\ben{are you ok being cited here?}][]{Gallagher2020}.
-To see the importance of $R_0$, one need only read the newspaper \citep{Fisher} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep[\ben{at shannon: is this an ok reference or do you want a better one?}][]{Aronson2020}.
-$R_O$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is most commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
+To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}.
+$R_0$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
 The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILL IN}).

### 2020-05-19 jss/jss-article-rnw/chapters/ch3.2-ternary.Rnw
clean up of paper from May 18th discussion

@@
 Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
-\ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different0-scale-standard}, the data sets look quite similar in the ternary view.}
+\ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view.}
     \label{fig:different-scales-tern}

### 2020-05-19 jss/jss-article-rnw/chapters/ch3.3-simplex-and-beyond.Rnw
clean up of paper from May 18th discussion

@@
 with more states) can be done onto different simplexes.
-In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a 1 dimension less space through the simplexical structure of the data.
+In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a one-dimension less space through the simplexical structure of the data.
 
 \ben{In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes.
-To aid in the comparison of models and epidemics we provide multiple tools.
-The first of which is, through multiple simulations under specific model parameters we can generate multiple time-series under the same assumptions.
-In \pkg{EpiCompare} we provide ways (leveraging conformal inference \cite{Vovk2005}) to create prediction regions for a true epidemic under the model assumptions.
-These regions require representing multi-dimensional structures for functions to completely contain epidemics.
+Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics.
+The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit.
+In \pkg{EpiCompare} we provide ways (leveraging conformal inference \cite{Vovk2005}) to create prediction regions for a true epidemic under the model assumptions using these simulations.
+These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects.
 We extend off of papers like \citet{Dalmasso2019a} and \citet{LeRoy2020} to create these bands.

### 2020-05-26 R/transition-matrix-to-function.R
Added a simulate_agents vignette and made a sirs_data data.

@@ -9,14 +9,18 @@
 #' @param K number of states
 #' @param n_pars number of parameters
 #' @return a list of length \eqn{K^2} where entry K(i-1) + j for i, j <= K is a function.  See details.
-#' @details This function dynamically creates a list of functions that turn the character transition formulas provided by the user in D_mat into \eqn{K^2} functions which take in arguments X0, ..., X{K-1)}, par1, par2, ..., par{n_pars}, and N and output the numeric transition rate of an individual currently in state \eqn{i} to move to state \eqn{j}.  
+#' @details This function dynamically creates a list of functions that turn the
+#'  character transition formulas provided by the user in D_mat into \eqn{K^2}
+#'   functions which take in arguments X0, ..., X{K-1)}, par1, par2, ..., par{n_pars},
+#'  N, and t and output the numeric transition rate of an individual
+#'   currently in state \eqn{i} to move to state \eqn{j}.  
 trans_mat_to_D_fxn <- function(D_mat, K,
                                n_pars){
     sym_list <- vector(mode = "list", length = K^2)
-    my_args <- vector(mode = "list", length = K + n_pars + 1)
+    my_args <- vector(mode = "list", length = K + n_pars + 2)
     names(my_args) <- c(paste0("X", 0:(K-1)),
                         paste0("par", 1:n_pars),
-                        "N")
+                        "N", "t")
     D_vec <- as.vector(t(D_mat))
     ## Go across then down
     for(ii in 1:K^2){

### 2020-05-26 data-raw/epicompare-sirs.R

@@ -1,0 +1,62 @@
+library(tidyverse)
+library(gridExtra)
+devtools::load_all()
+
+
+## SIRS
+
+## This is the SIRS representation
+trans_mat <- matrix(c("X0 * (1 - (X1 + X4) * par1 / N)",
+                      "X0 * (X1 + X4)  * par1 / N", 
+                      "0",  "0", "0", "0", ## ROW 1 DONE
+                      "0", "X1 * (1 - par2)", "par2 * X1", "0", "0", "0",  ## ROW 2 DONE
+                      "0", "0", "X2 * (1 - par3)", "X2 * par3", "0", "0", ## ROW 3 DONE
+                      "0", "0",
+                      "0", "X3 * (1 - (X1 + X4) * par3 / N)",
+                      "X3 * (X1 + X4) * par4 / N", "0", ## ROW 4 DONE
+                      "0", "0", "0",
+                      "0", "X4 * (1 - par5)", "par5 * X4", ## ROW 5 DONE
+                      "0", "0", "0", "0", "0", "X5"
+                      ), byrow = TRUE, nrow = 6)
+rownames(trans_mat) <- c("S1", "I1", "R1", "S2", "I2", "R2")
+colnames(trans_mat) <- c("S1", "I1", "R1", "S2", "I2", "R2")
+init_vals <- c(950, 50, 0, 100, 50, 0)
+par_vals <- c(par1 = .5, par2 = .4,  #beta1, gamma1
+              par3 = .08, # transition from R1 to S2
+              par4 = .08, par5 = .06) #beta2, gamma2
+max_T <- 300
+n_sims <- 25
+
+
+
+
+## Warning:  will take about a minute.  Set `verbose = TRUE` to see progress
+set.seed(2020)
+abm <- simulate_agents(trans_mat,
+                       init_vals,
+                       par_vals,
+                       max_T,
+                       n_sims,
+                       verbose = FALSE,
+                       out_format = "wide")
+
+agg_model <- abm %>% group_by(sim) %>%
+  agents_to_aggregate(states = c(I1, R1, S2, I2, R2))
+colnames(agg_model) <- c("sim", "t",
+                         "S1", "I1", "R1",
+                         "S2", "I2", "R2")
+
+
+g0 <- ggplot(data = agg_model, aes(x = t, y = S1 + S2, group = sim )) +
+  geom_line()
[…]

### 2020-05-26 data-raw/pomp_data-sirs.R

@@ -7,6 +7,8 @@ library(pomp)
 ## Modified for a SIRS model like here: https://www.idmod.org/docs/typhoid/model-sir.html#sirs-model
 ## This is deterministic
 
+## NO LONGER USED.
+
 sirs.ode <- Csnippet("
   DS = -Beta*S*I/N + eta * R;
   DI = Beta*S*I/N-gamma*I;

### 2020-06-07 R/fortify-aggregate-epimodel-icm.R
Fixed dplyr 1.0.0 break due to upgrade.  Also made a Hagelloch EDA vignette

@@ -61,6 +61,7 @@ get_epimodel_icm_states <- function(data){
 #' @param epi list of matrcies from icm output
 #' @return a tibble
 extract_icm_cols <- function(nm, ii, epi){
+
     mat <- epi[[nm]]
     nm <- paste0("X", ii)
     if(tidyr_new_interface()){

### 2020-06-07 R/fortify-aggregate-epimodel-inner.R

@@ -22,6 +22,7 @@
 fortify_aggregate.epimodel_inner <- function(data,
                                              states,
                                              package_source = NULL){
+    
 
     states <- unname(tidyselect::vars_select(names(data$epi),
                                              !!states))
@@ -66,7 +67,14 @@ fortify_aggregate.epimodel_inner <- function(data,
 
     ## Put it all together
     agg_df <-  state_list %>%
-        dplyr::bind_cols() %>%
+        dplyr::bind_cols()
+
+    if(!("sim" %in% colnames(agg_df))){  ## fix for dplyr 1.0.0 breaking change to bind_cols()
+      
+        agg_df <- suppressMessages(agg_df %>% dplyr::rename(sim = "sim...1"))
+    }
+
+    agg_df <- agg_df %>% 
         dplyr::select(.data$sim, dplyr::starts_with("X")) %>%
         dplyr::mutate(sim = factor(.data$sim))
     agg_df$t <- t

### 2020-06-15 R/EpiCompare.R
agents_to_aggregate allow user to specify if you don't need all time points.

@@ -34,7 +34,6 @@
 #' aggregate framework described above.
 #' }
 #' }
-#' @import ggtern
 #' @importFrom rlang .data
 #'
 #' @docType package

### 2020-06-15 tests/testthat/test-agent-to-aggregate.R

@@ -178,6 +178,49 @@ test_that("expanding_info", {
                                                state = as.numeric(state)))
 })
 
+test_that("distinct_time", {
+  # something interesting
+  df <- data.frame(t = 1:100, x = rep(c(1,1,2,2,3,3),
+                                      length = 100),
+                   y = rep(c(1,1,1,2,2,2,3,3,3),
+                           length = 100))
+  
+  df_filtered <- distinct_time(df, 1)
+  testthat::expect_equal(dim(df_filtered), c(67,3))                     
+  testthat::expect_equal(df_filtered, 
+                         df[!(1:100 %in% cumsum(rep(c(2,4), length = 33))),])
+  
+  df_all <- distinct_time(df, NULL)
+  testthat::expect_equal(df_all, df)
+  
+  df_x <- distinct_time(df, c(1,3))
+  testthat::expect_equal(df_x, df[1:100 %% 2 == 1,])
+  
+  df_y <- distinct_time(df, c(1,2))
+  testthat::expect_equal(df_y, df[1:100 %% 3 == 1,])
+  
+  # nothing interesting
+  df <- data.frame(t =1:100, x = rep(1:3, length = 100),
+                   y = rep(1:4, length = 100))
+  
+  df_filtered <- distinct_time(df, 1)
+  testthat::expect_equal(df_filtered, df)
+  
+  df_all <- distinct_time(df, NULL)
+  testthat::expect_equal(df_all, df)
+  
+  df_x <- distinct_time(df, c(1,3))
+  testthat::expect_equal(df_x, df)
+  
+  df_y <- distinct_time(df, c(1,2))
+  testthat::expect_equal(df_y, df)
+  
+  
+  df_single <- distinct_time(df, 1:3)
+  testthat::expect_equal(nrow(df_single), 1)
+  testthat::expect_equal(df_single, df[1,])
+})
+
 test_that("agents_to_aggregate.data.frame, data_example_s example",{
   data_example_s <- data.frame(group1 = c(.1,0),
                                group2 = c(2.1,2),
@@ -857,5 +900,46 @@ test_that("agents_to_aggregate.group_df, max_t = NA basic checks, strings", {
       testthat::expect_equal(unique_rows_after_max, 1)
     }
   }
+
+})
+
+test_that("agents_to_aggregate + integer_time_expansion = FALSE -- more basic",{
+  data_example_s <- data.frame(group1 = c(.1,0),
+                               group2 = c(2.1,2),
+                               group3 = c(3.1,3))
+  
[…]

### 2020-09-08 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw
minor updates to paper article

@@
 As the availability of scientific resources has increased through the decades, the number of models to analyze a particular disease outbreak has also increased.
-This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \ben{\citep[e.g.][]{Sanche2020a}}, \ben{and is well documented in many literature reviews in epidemology \citep[e.g.][]{Canzani2015}}.
-As both the quantity of data and models increase, so does the need for standard and fair comparisons of models, as according to the data analysis pipeline of infectious diseases.
+This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \citep[e.g.][]{Sanche2020a}, and is well documented in many literature reviews in epidemology \citep[e.g.][]{Canzani2015}.
+As both the quantity of data and models increase, so does the need for standard and fair comparisons of models, as according to the data analysis pipeline of infectious diseases.\footnote{\ben{Is our tool more a "after the fact" analysis than some works?}}
 
@@
   In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next.
-In Figure \ref{fig:pipeline} we also highlight where \pkg{EpiCompare} can aid in this pipeline.
+In Figure \ref{fig:pipeline} we also highlight where \pkg{EpiCompare} can aid in this pipeline\footnote{\ben{We haven't done thee part of highlighting what EpiCompare can do...}}.
 
@@
 The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
-\pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-free tools (motivated in more detail in Section \ref{sec:time-free}).More broadly, \pkg{EpiCompare} aids the data analysis pipeline for infecetious diseases by
+\pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-invariant tools (motivated in more detail in Section \ref{sec:time-invariant}).More broadly, \pkg{EpiCompare} aids the data analysis pipeline for infecetious diseases by
 \begin{enumerate}
   \item Standardizing outputs from common \proglang{R}-based infectious disease model software and common data storage types
-  \item Expanding upon existing data visualization and EDA tools
-  \item \ben{\sout{Provides support for modeling and simulation} Providing a interface for simple ODE-style agent based modeling.}
-  \item Aiding in comparison and assessment of models through time free analysis and conformal prediction bands and comparisons.
+  \item Expanding upon existing data visualization and EDA tools (especially in 3 state models like SIR models).
+  \item Providing a interface for simple ODE-style agent based modeling.
+  \item Aiding in comparison and assessment of models through time invariant analysis and conformal prediction bands and comparisons.
 \end{enumerate}
-Although \pkg{EpiCompare} aids in every part of the pipeline, our tools especially emphasize improvements for pipeline parts (1), (4), and (5) which can be summarized as data input and output harmonization and contributions to fairly comparing models and data to one another.
+Although \pkg{EpiCompare} aids in every part of the pipeline, our tools especially emphasize improvements for pipeline parts (1) and (4) which can be summarized as data input and output harmonization and contributions to fairly comparing models and data to one another.
 
-The paper is arranged as follows.
-In Section \ref{sec:lit}, we review some of the prior work in the infectious disease data analysis pipeline and describe how \pkg{EpiCompare} improves this process\ben{\footnote{\ben{Currently we break this into sections 2 and 3 - discuss?}}}.
+The paper is arranged as follows.\footnote{\ben{Looks like this whole paragraph will need a rewrite down the road.}}  In Section \ref{sec:lit}, we review some of the prior work in the infectious disease data analysis pipeline and describe how \pkg{EpiCompare} improves this process.
 In Section \ref{sec:data-harmony}, we describe how we use \pkg{tidyverse}-style tools to harmonize data for work in a usable format.

### 2020-09-08 jss/jss-article-rnw/chapters/ch3-time-free.Rnw
minor updates to paper article

@@
 
-Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a less time focused, or even time-free way to understand the processes at play.
+Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play.
 Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
-Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-free approaches and in this section we present benefits of the time-free analysis, with 1) motivation for time-free analysis through $R_0$ (the initial reproduction number), 2) time-free visualization tools for 3-state models (e.g.
+Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches and in this section we present benefits of the time-invariant analysis, with 1) motivation for time-invariant analysis through $R_0$ (the initial reproduction number), 2) time-invariant visualization tools for 3-state models (e.g.
 SIR models), and 3) potential for similar analysis for models with more states (e.g.

### 2020-09-08 jss/jss-article-rnw/chapters/ch3.1-r0.Rnw
minor updates to paper article

@@
 
-$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[\ben{are you ok being cited here?}][]{Gallagher2020}.
+$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[][]{Gallagher2020}.
 To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}.
-$R_0$ is also, maybe, the most famous \textit{time-free} numerical summary of an epidemic, and is commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
+$R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, and is commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
 The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILL IN}).
@@
 For example, in a SIR simulation with two models have the same value of $R_0$, one can see in Fig.
-\ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph.
+\ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph (this figure will be commented on more below).
 
@@
 time.
-Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\ben{We probably need to define $\beta/\gamma$.}}
+Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\ben{We probably need to define $\beta/\gamma$.}
+\ben{NEW: recommendation - make this plot actually have 3 examples (the current one), one where we have an affine transformation of the time scale, and one with the halleolof data.}}
     \label{fig:different-scales-standard}

### 2020-09-08 jss/jss-article-rnw/chapters/ch3.2-ternary.Rnw
minor updates to paper article

@@
 
-\subsection{Ternary and time-free visualizations for $R_0$} \label{sec:ternary}
-
-\ben{My original notes for this section note: ``Ternary plots (pitch from Shannon’s thesis) - being useful to assess.
-`Peak shows corresponds to the same event in the time analysis' (maybe?)''.
-I don't necessarily think this section perfectly does that - so I'm open to having it include that or just deciding this is better.}
+\subsection{Ternary and time-invariant visualizations for $R_0$} \label{sec:ternary}
 
 It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig.
-\ref{fig:different-scales-tern}  have the same value of $R_0$ via ``time-free'' visualizations, under some circumstances.
-Specifically, for SIR models we propose using ternary to examine an epidemic's trajectory in a more time-free manner.
+\ref{fig:different-scales-tern}  have the same value of $R_0$ via ``time-invariant'' visualizations, under some circumstances.
+Specifically, for SIR models we propose using ternary to examine an epidemic's trajectory in a more time-invariant manner.
 
-Ternary plots are sometimes used in chemistry \citep[\ben{shannon: is this a good reference?
-- was in your thesis.}][]{Gillespie1976} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
+Ternary plots are sometimes used in chemistry \citep[][]{Gillespie1976} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
 Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.
@@
 
-Ternary plots are time-free in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g.
+Ternary plots are time-invariant in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g.
 days vs.
@@
   \end{align}
+
+\ben{SUGGESTION: define $R_0$ mathematically here (though referred to in figure 2 - which still needs to be defined there).}
 
@@
 Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
-\ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view.}
+\ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view.
+\ben{I might suggest using opacity to show time scale here?}}
     \label{fig:different-scales-tern}
@@
 
-In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-free space, which is presented in more details in Section \ref{sec:time-free-in-simplex}.
+In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-invariant space, which is presented in more details in Section \ref{sec:time-invariant-in-simplex}.

### 2020-09-08 jss/jss-article-rnw/chapters/ch3.3-simplex-and-beyond.Rnw
minor updates to paper article

@@
 
-\subsection{Simplexes beyond ternary plots and time-free comparison tools} \label{sec:time-free-in-simplex}
+\subsection{Simplexes beyond ternary plots and time-invariant comparison tools} \label{sec:time-invariant-in-simplex}
 
-\ben{\sout{Maybe should include something like "beyond ternary plots as a tool for time-free analysis..."}}
-\ben{Beyond ternary plots, and the associated SIR models, we can extend the time-free approach described in Section \ref{sec:ternary} to higher dimensions (i.e.
-models with more states).\sout{The basis of time-free analysis in ternary plots is extended to higher dimensional models (i.e.
-models with more states) in \pkg{EpiCompare} as well.}} The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling).
+Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:ternary} to higher dimensions (i.e.
+models with more states).
+The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling).
 This same scaling for larger models (i.e.
@@
 
-\ben{In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes.
+In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes.
 Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics.
 The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit.
-In \pkg{EpiCompare} we provide ways (leveraging conformal inference \cite{Vovk2005}) to create prediction regions for a true epidemic under the model assumptions using these simulations.
+In \pkg{EpiCompare} we provide ways to create prediction regions for a true epidemic under the model assumptions using these simulations.
 These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects.
-We extend off of papers like \citet{Dalmasso2019a} and \citet{LeRoy2020} to create these bands.
+We extend off of papers like \citet{Dalmasso2019a} to create these bands.
 These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets.
@@
 d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
-\]}
+\]

### 2020-09-08 jss/jss-article-rnw/chapters/ch4-orientation.Rnw
added comments relative to the outline of the paper
minor updates to paper article

@@
 I also expect that this section will mirror how we present the pipeline in the Literature review.}
+
+---
+
+\ben{We also called this section ``Presenting EpiCompare'', not sure if we want to do that verse ``Package Orrientation'', but we'll figure it out.}
+
+---
+
+\ben{Also, note that the original sections proposed for this section and the Literature Review were:
+\begin{enumerate}
+\item Data collection
+\item Exploratory analysis
+\item ``Modelling'' $\rightarrow$ simulations
+\item Assessment and comparisons
+\end{enumerate}
+}
+
+---
+
+\ben{Maybe "drop" the "In practice" section and just use Hagollogh's full pipeline in the presentation?}

### 2020-09-23 dev/flu_analysis.Rmd
flu exploration in development

---
title: "Examining the Flu through wILI"
author: "Benjamin LeRoy"
output: 
  html_document:
    number_sections: true
---

```{r include = F, warning = F, message = F}
library(tidyverse)
library(stringr)
library(ISOweek)
```
# Introduction

The common flu, although different from year to year often exhibits common patterns from year to year. We use [CMU's Delphi](https://delphi.cmu.edu/) group's API to collect data from 2000 to 2018 flu seasons in the United States. We will specifically focus on a weekly "wILI" value, which captures a **w**eighted percent of **i**nfluenza-**l**ike **i**llness cases seen in the hospital in that given week.

# Pulling in the Data

## API Logistics

[CMU's Delphi](https://delphi.cmu.edu/) group's API can be sourced by running the following command.

```{r}
source(paste0("https://raw.githubusercontent.com/",
              "cmu-delphi/delphi-epidata/main/src/client/delphi_epidata.R"))
```

This file provides a lists of functions in a more `python`ic way - all functions are avaiable in the `Epidata` list, and can be accessed through the notation `Epidata$function_name`.

## Which data to pull?

We aim to examine flu seasons from 2001 to 2018. Although the API can pull data from as far back as 2001. Note that influenza death estimation (which we are not looking at), change in 2010^[CDC: https://www.cdc.gov/flu/about/burden/faq.htm]. Additionally, flu seasons often are examined October of the previous year to August of the year.

```{r}
flu_data_raw <- Epidata$fluview(list('nat'), Epidata$range(200010, 
                                                           201809))
```

Delphi has documentation on the outcome of the [`fluview`](https://cmu-delphi.github.io/delphi-epidata/api/fluview.html) function (and other functions),
on their [website](https://cmu-delphi.github.io/delphi-epidata/api/README.html). Regions can be found [here](https://github.com/cmu-delphi/delphi-epidata/blob/main/src/acquisition/fluview/fluview_locations.py) and is related to [this](https://gis.cdc.gov/grasp/flu2/GetPhase02InitApp?appVersion=Public).

We can extract the desired information from the second item in the list (under the `epidata` name), into a data.frame with the following function `extract_df`. We've folded the code as it's messy and not that informative.

<details><summary> **`extract_df` code:** </summary>

```{r}
#' extract data frame from delphi's epidata "fluview" function
#'
#' @param epidata "epidata" output as desired by the fluview function (see https://cmu-delphi.github.io/delphi-epidata/api/fluview.html)
#'
#' returns data.frame with the same information
extract_df <- function(epidata){
  # basic structure of data frame -----------------
  nr <- length(epidata)
  nc <- length(epidata[[1]])
  colnames <- names(epidata[[1]])

  # creating an empty data frame ------------------
  ## character and numeric columns
  c_idx <- which(sapply(epidata[[1]], class) %in% c("character", "factor"))
  n_idx <- (1:nc)[!(1:nc %in% c_idx)]

  data <- data.frame(matrix(nrow = nr, ncol = 0))
  for (col_idx in 1:nc){
    if (col_idx %in% c_idx){
      data <- cbind(data, data.frame(matrix("c", nrow = nr, ncol = 1)))
    } else {
      data <- cbind(data, data.frame(matrix(0, nrow = nr, ncol = 1)))
    }
  }
  names(data) <- colnames

  for (row_idx in 1:nr){
    inner <- epidata[[row_idx]]
    inner[sapply(inner, function(x) length(x) == 0L)] <- NA

    # breaking into parts
    data[row_idx,c_idx] <- unlist(inner[c_idx])
    data[row_idx,n_idx] <- unlist(inner[n_idx])
  }

  return(data)
}

```

</details>

We create a data.frame for the flu data (percent of new cases) and clean it up a bit to emphasis the week of the year we're looking at.

```{r echo= F}
flu_data <- extract_df(flu_data_raw$epidata)
# convert epiweek to day
flu_data <- flu_data %>% rowwise() %>%
  mutate(epiweek_year = str_sub(epiweek, start = 1L, end = 4L),
         epiweek_week = str_sub(epiweek, start = 5L, end = 6L),
         epiweek2 = paste0(epiweek_year, "-W", epiweek_week,"-1"),
         epiweek_date = ISOweek2date(epiweek2))
```

## Data Example

Below is an example of the data.

```{r echo = F}
flu_data %>% head() %>% DT::datatable(options = list(scrollX = TRUE))
```

## wILI visualization

```{r season-df_vis, echo = F}
start_year = 1999
end_year = 2017
flu_season_info_df <- data.frame(year = start_year:end_year) %>%
  mutate(start_date = paste0(as.character(year), "-10-01"),
         end_date = paste0(as.character(year+1), "-08-30"),
         even_year = c( "odd", "even")[(year %% 2) + 1]) %>%
  mutate(start_date = as.Date(start_date, format= "%Y-%m-%d"),
         end_date = as.Date(end_date, format= "%Y-%m-%d"))
```

```{r eval = F, echo = F}
# visual
flu_data %>%
ggplot() +
  geom_line(aes(x = epiweek_date, y = wili)) +
  labs(x = "time",
       y = "wILI percentage",
       title = "wILI percentage across multiple seasons") +
  geom_rect(data = flu_season_info_df %>% filter(even_year == "even"), 
            mapping = aes(xmin = start_date, xmax = end_date, ymin = 0, ymax = 8),
            alpha = .1) +
  theme_minimal()
```

```{r echo = F, message = F, warning = F, fig.align="center"}
library(dygraphs)

periods_df <- flu_season_info_df %>% filter(even_year == "even")

add_shades <- function(x, periods_df, ...) {
  for( r_idx in 1:nrow(periods_df) ) {
    x <- dyShading(x, from = periods_df[r_idx, "start_date"] , 
                   to = periods_df[r_idx, "end_date"], ... )
  }
  x
}

flu_xts <- xts::xts(x = flu_data$wili,
           order.by = flu_data$epiweek_date) 
names(flu_xts) <- "wILI percentage"
flu_xts %>% 
  dygraph(data = ., main = "wILI percentage across multiple seasons", 
        ylab = "wILI percentage", width = "100%") %>%
  dyRangeSelector() %>% 
  add_shades(periods_df = periods_df)
```

## wILI visualization (per season)

```{r eval = T, echo = F, message = F, warning = F}

season_index <- rep(0, nrow(flu_data))
for (start_date in flu_season_info_df$start_date){
  season_index <- season_index + 1*(flu_data$epiweek_date > start_date)
}
flu_data$season_index <- season_index
flu_data$after_start <- flu_data$epiweek_date - flu_season_info_df$start_date[flu_data$season_index]
flu_data$season_year <- flu_season_info_df$year[season_index]

# visual
by_season_vis <- flu_data %>%
  mutate(season_year = factor(season_year)) %>%
ggplot() +
  geom_line(aes(x = after_start, y = wili, color = season_year)) +
  labs(x = "Days after start of season",
       y = "wILI percentage",
       title = "wILI percentage across multiple seasons \n (2008 & 2009 odd season)",
       color = "") +
  theme_minimal()

by_season_vis %>% plotly::ggplotly()
```

# EpiCompare with the Data

The set of code below is messing but let's talk about a few things.

```{r}
df <- flu_data %>% tibble::rownames_to_column()

#df

# usa population (by month) from https://fred.stlouisfed.org/series/POPTHM
pop <- read.csv("POPTHM.csv") %>%
  mutate(MONTH = lubridate::month(DATE),
         YEAR = lubridate::year(DATE))

df <- df %>% mutate(epiweek_month = lubridate::month(epiweek_date))

df_plus <- df %>% mutate(epiweek_year = as.numeric(epiweek_year)) %>% 
  left_join(pop, by = c("epiweek_month" = "MONTH", "epiweek_year" = "YEAR")) %>%
  mutate(scaled_num_ili = POPTHM * wili/100)

```

```{r message = F, warning = F}
# needs to be updated to do tidyverse also what is par?
# and what is confirmed (cumulated relative to what?)
# negative values for some output...
devtools::load_all(path = "../")
#?EpiCompare::cases_to_SIR
```

```{r}
# change to days (to get 3 day flu)

# expand?
df2 <- df_plus %>% mutate(n_wili = round(POPTHM * wili / 100),
                          rowname = as.numeric(rowname)*7,
                          d_wili = wili/7) %>% 
  select(d_wili, epiweek_date)

stepwise_fill_in <- function(df2, step = 7){
  # fill in with zeros 
  df_new <- data.frame(df2[1,])

  for (r_idx in 1:nrow(df2)){
    inner_df <- data.frame(d_wili = df2$d_wili[r_idx],
                           epiweek_date = df2$epiweek_date[r_idx] + 0:(step-1))

    df_new <- rbind(df_new, inner_df)
  }

  df_new <- df_new[-1,]

  return(df_new)
}

df3 <- stepwise_fill_in(df2)
df3 %>% head()

df4 <- df3 %>% mutate(month = lubridate::month(epiweek_date),
               year = lubridate::year(epiweek_date)) %>% 
  left_join(pop, by = c("month" = "MONTH", "year" = "YEAR")) %>% 
  mutate(n_wili = round(POPTHM * d_wili / 100))

new <- df4 %>% tibble::rownames_to_column() %>%
  rename(t = rowname, N = POPTHM, current = n_wili) %>%
  mutate(confirmed = cumsum(current)) %>%
  EpiCompare::cases_to_SIR(data = ., par = c("gamma" = 1/3)) %>% #unit relative to days (3 days)
  select(t, X0, X1, X2)

new %>% mutate(t = as.numeric(t)) %>% 
  pivot_longer(cols = c(X0, X1, X2)) %>% 
  filter(name == "X1") %>% 
  ggplot2::ggplot() + ggplot2::geom_line(ggplot2::aes(x = t, y = value, color = name)) +
  ggplot2::theme_minimal()

new2 <- new %>%
  mutate(N = X0 + X1 + X2,
         confirmed = X2)

new2_R_to_S <- new2 %>% select(t, confirmed, N) %>%
  cases_to_SIR(par = c("gamma" = 1/60)) # 3 months to be suspectible again (to the flu)
new2_R_to_S %>% head

new$X3 <- new2_R_to_S$X2

new_SIRS <- new
new_SIRS <- new_SIRS %>% mutate(
  X0 = X0 + X3) %>%
  dplyr::select(t, X0, X1, X2) %>%
  dplyr::rename(S = X0, I = X1, R = X2)

new_SIRS %>% 
ggplot(aes(x = S, y = I, z = R)) +
  geom_path() +
  coord_tern() +
  theme_minimal()

new_SIRS %>% pivot_longer(cols = c(S, I, R)) %>%
  ggplot() + geom_line(aes(x = as.numeric(t), 
                           y = value, group = name, color = name)) +
  theme_minimal()

new %>% pivot_longer(cols = c(X0, X1, X2)) %>%
  filter(name == "X1") %>%
  ggplot() + geom_line(aes(x = as.numeric(t), 
                           y = value, group = name, color = name)) +
  theme_minimal()
```

```{r}
# change to days (to get 3 day flu)

# expand?
df2 <- df_plus %>% mutate(n_wili = round(POPTHM * wili / 100),
                          rowname = as.numeric(rowname)*7,
                          d_wili = wili/7) %>% 
  select(d_wili, epiweek_date, season_year)

stepwise_fill_in <- function(df2, step = 7){
  # fill in with zeros 
  df_new <- data.frame(df2[1,])

  for (r_idx in 1:nrow(df2)){
    inner_df <- data.frame(d_wili = df2$d_wili[r_idx],
                           epiweek_date = df2$epiweek_date[r_idx] + 0:(step-1),
                           season_year = df2$season_year[r_idx])

    df_new <- rbind(df_new, inner_df)
  }

  df_new <- df_new[-1,]

  return(df_new)
}

df3 <- df2 %>% group_by(season_year) %>% stepwise_fill_in()

df4 <- df3 %>% mutate(month = lubridate::month(epiweek_date),
               year = lubridate::year(epiweek_date)) %>% 
  left_join(pop, by = c("month" = "MONTH", "year" = "YEAR")) %>% 
  mutate(n_wili = round(POPTHM * d_wili / 100))

new <- df4 %>% tibble::rownames_to_column() %>%
  rename(t = rowname, N = POPTHM, current = n_wili) %>% 
  group_by(season_year) %>% 
  mutate(confirmed = cumsum(current)) %>%
  EpiCompare::cases_to_SIR(data = ., par = c("gamma" = 1/3)) %>% #unit relative to days (3 days)
  select(t, X0, X1, X2)

new_other <- df4 %>% tibble::rownames_to_column() %>%
  rename(t = rowname, N = POPTHM, current = n_wili) %>%
  mutate(confirmed = cumsum(current)) %>% 
  EpiCompare::cases_to_SIR(data = ., par = c("gamma" = 1/3)) %>% #unit relative to days (3 days)
  select(t, X0, X1, X2)
```
```{r}
new %>% mutate(t = as.numeric(t)) %>% 
  pivot_longer(cols = c(X0, X1, X2)) %>% 
  filter(name == "X1") %>% 
  ggplot2::ggplot() + ggplot2::geom_line(ggplot2::aes(x = t, y = value, color = name)) +
  ggplot2::theme_minimal()

new_other %>% mutate(t = as.numeric(t)) %>% 
  pivot_longer(cols = c(X0, X1, X2)) %>% 
  filter(name == "X1") %>% 
  ggplot2::ggplot() + ggplot2::geom_line(ggplot2::aes(x = t, y = value, color = name)) +
  ggplot2::theme_minimal()
```

```{r}
new2 <- new %>%
  mutate(N = X0 + X1 + X2,
         confirmed = X2)

new2_R_to_S <- new2 %>% select(t, confirmed, N) %>%
  cases_to_SIR(par = c("gamma" = 1/60)) # 3 months to be suspectible again (to the flu)
new2_R_to_S %>% head

new$X3 <- new2_R_to_S$X2

new_SIRS <- new
new_SIRS <- new_SIRS %>% mutate(
  X0 = X0 + X3) %>%
  dplyr::select(t, X0, X1, X2) %>%
  dplyr::rename(S = X0, I = X1, R = X2)

df_lims = data.frame(S = c(1,.4,.4), I = c(0,.6,0), R = c(0,0,.6))

new_SIRS %>% 
ggplot(aes(x = S, y = I, z = R, 
           color = factor(season_year))) +
  geom_path() +
  coord_tern() +
  theme_minimal() +
  tern_limits(T=max(df_lims$I), 
              L=max(df_lims$S), 
              R=max(df_lims$R))

```

```{r}
plotly::ggplotly(new_SIRS %>% pivot_longer(cols = c(S, I, R)) %>%
  ggplot() + geom_line(aes(x = as.numeric(t), 
                           y = value, group = name, color = name)) +
  theme_minimal() )
```

```{r}
new %>% pivot_longer(cols = c(X0, X1, X2)) %>%
  filter(name == "X1") %>%
  ggplot() + geom_line(aes(x = as.numeric(t), 
                           y = value, group = name, color = name)) +
  theme_minimal()

new_other %>% pivot_longer(cols = c(X0, X1, X2)) %>%
  filter(name == "X1") %>%
  ggplot() + geom_line(aes(x = as.numeric(t), y = value, group = name, color = name)) +
  theme_minimal() 
```

how to connect wILI to the total US (https://www.cdc.gov/flu/about/burden/how-cdc-estimates.htm)

### 2020-09-28 R/aaa.R
ggplot2 and ggtern over-wrote with Epicompare:::print.ggplot

@@ -56,3 +56,28 @@ update_approved_layers()
 r_new_interface <- function() {
   as.numeric(R.version$major) >= 4
 }
+
+
+#' Explicitly draw plot (\code{ggtern} and \code{ggplot2} compatible)
+#'
+#' Makes sure both \code{\link[ggtern:::print.ggplot]{ggtern}} and 
+#' \code{\link[ggplot2:::print.ggplot]{ggplot2}} objects 
+#' display correctly.
+#'
+#' @param x plot to display
+#' @param newpage draw new (empty) page first?
+#' @param vp viewport to draw plot in
+#' @param ... other arguments not used by this method
+#'
+#' @return Invisibly returns the result of \code{ggplot_build()}, which is a 
+#' list with components that contain the plot itself, the data, information 
+#' about the scales, panels etc.
+#' @export
+print.ggplot <- function(x, newpage = is.null(vp), vp = NULL, ...){
+  if(inherits(x$coordinates, "CoordTern")){
+    ggtern:::print.ggplot(x, newpage = newpage, vp = vp, ...)
+    
+  } else {
+    ggplot2:::print.ggplot(x, newpage = newpage, vp = vp, ...)
+  }
+}

### 2020-10-01 README.Rmd

@@
 `method.class` function), which we have for the `contained` function.
+5.
+**Does 2 wrongs make a right?** As of 9/23 `ggtern` had an [issue](https://bitbucket.org/nicholasehamilton/ggtern/issues/13/ggtern-breaks-ggplots-plot) that it messed with `ggplot2`'s legends when loaded (it over-wrote the `print.ggplot` and other functions).
+To "correct" this, we've over-wrote `ggtern`'s `print.ggplot` to correct this problem when not producing ternary plots (code in `aaa.R`).
 

### 2020-10-01 R/bands.R
wacky error with geom_predict when binding different data frames together

@@ -297,7 +297,7 @@ StatPredBandKDE <- ggplot2::ggproto("StatPredBandKDE",
                                                       .data$piece_old,
                                                       .data$group_old,
                                                       sep = "*"))),
-                                       group = piece)
+                                       group = .data$piece)
 
     return(data_cleaned_up)
   },

### 2020-10-02 R/aaa.R
updates to bands (missing some dplyr::)

@@ -60,9 +60,9 @@ r_new_interface <- function() {
 
 #' Explicitly draw plot (\code{ggtern} and \code{ggplot2} compatible)
 #'
-#' Makes sure both \code{\link[ggtern:::print.ggplot]{ggtern}} and 
-#' \code{\link[ggplot2:::print.ggplot]{ggplot2}} objects 
-#' display correctly.
+#' Makes sure both \code{\link[ggtern]{ggtern}} and 
+#' \code{\link[ggplot2]{ggplot2}} objects 
+#' display correctly with the \code{print.ggplot} function.
 #'
 #' @param x plot to display
 #' @param newpage draw new (empty) page first?

### 2020-10-07 R/bands.R
prediction band allows for non-constant length of simulations for delta_ball and convex_hull

@@ -276,6 +276,41 @@ get_grid_elipsoid_containment <- function(inside_func_list,
 }
 
 
+#' checks the provided dist_params to make sure they are what is expected 
+#'
+#' @param dist_params named list of parameters
+#' @param data data for visual 
+#'
+#' @return error or updated dist_params
+check_dist_params <- function(dist_params, data){
+  if (dist_params[["dist_approach"]] == "auto"){
+    dist_params[["dist_approach"]] <- "equa_dist"
+  }
+  if (dist_params[["dist_approach"]] == "equa_dist" & 
+      dist_params[["num_steps"]] == "auto"){
+    dist_params[["num_steps"]] <- 20
+  }
+  if (dist_params[["dist_approach"]] == "equa_dist"){
+    x <- dist_params[["num_steps"]]
+    tol <- 2*.Machine$double.eps
+    assertthat::assert_that(min(abs(c(x %% 1, x %% 1-1))) < tol,
+                            msg = paste("'dist_params$num_steps' must
+                                          be an positive integer or string 
+                                          'auto'."))
+  }
+  
+  if (dist_params[["dist_approach"]] == "temporal"){
+    counts <- data %>% as.data.frame() %>% 
+      dplyr::group_by(.data$sim_group) %>%
+      dplyr::summarize(total = dplyr::n())
+    assertthat::assert_that(length(unique(counts$total)) == 1,
+                            msg = paste("if dist_params$dist_approach is 
+                                          'temporal', then every path must
+                                          have the same number of points."))
+  }
+  return(dist_params)
+}
+
 
 # stats and geoms -------------------------
 
@@ -287,9 +322,9 @@ StatPredBandKDE <- ggplot2::ggproto("StatPredBandKDE",
   ggplot2::Stat,
   compute_layer = function(self, data, params, layout){
     # first run the regular layer calculation to infer densities
-    #browser()
+
     data <- ggplot2::ggproto_parent(ggplot2::Stat,self = self)$compute_layer(data, params, layout)
-    #browser()
+
     # required piece and group to be cleaned up
     data_cleaned_up <- data %>% dplyr::mutate(piece_old = .data$piece,
                                        group_old = .data$group,
@@ -306,7 +341,8 @@ StatPredBandKDE <- ggplot2::ggproto("StatPredBandKDE",
                            pb_type = NULL,
                            #^ needed to match same format as stat/geom_prediction_band
                            grid_size = rep(100,2),
[…]

### 2020-10-07 tests/testthat/test-agent-to-aggregate.R

@@ -789,18 +789,19 @@ test_that("agents_to_aggregate.group_df passes basic checks", {
   # similar test as in the example string
   max_time <- 100
   new_raw <- hagelloch_raw %>%
-    mutate(AGE2 = as.numeric(cut(AGE,3)))
-  agents_g <- new_raw %>% group_by(AGE2)
+    dplyr::mutate(AGE2 = as.numeric(cut(AGE,3)))
+  agents_g <- new_raw %>% dplyr::group_by(AGE2)
 
   sir_group <- agents_to_aggregate(agents_g, states = c(tI, tR),
                                    min_max_time = c(0, max_time))
   agents <- agents_g %>%
-    filter(AGE2 == 1) %>% ungroup()
+    dplyr::filter(AGE2 == 1) %>% dplyr::ungroup()
   sir_group1 <- agents_to_aggregate(agents, states = c(tI, tR),
                                    min_max_time = c(0, max_time))
-  sir_group_1 <- sir_group %>% filter(AGE2 == 1)
+  sir_group_1 <- sir_group %>% dplyr::filter(AGE2 == 1)
   testthat::expect_equal(sir_group1,
-                         sir_group_1 %>% ungroup %>% select(t, X0, X1, X2))
+                         dplyr::select(dplyr::ungroup(sir_group_1),
+                                       t, X0, X1, X2))
 })
 
 test_that("agents_to_aggregate.group_df -string passes basic checks", {
@@ -811,18 +812,19 @@ test_that("agents_to_aggregate.group_df -string passes basic checks", {
   # similar test as in the example string
   max_time <- 100
   new_raw <- hagelloch_raw %>%
-    mutate(AGE2 = as.numeric(cut(AGE,3)))
-  agents_g <- new_raw %>% group_by(AGE2)
+    dplyr::mutate(AGE2 = as.numeric(cut(AGE,3)))
+  agents_g <- new_raw %>% dplyr::group_by(AGE2)
 
   sir_group <- agents_to_aggregate(agents_g, states = c("tI", "tR"),
                                    min_max_time = c(0, max_time))
   agents <- agents_g %>%
-    filter(AGE2 == 1) %>% ungroup()
+    dplyr::filter(AGE2 == 1) %>% dplyr::ungroup()
   sir_group1 <- agents_to_aggregate(agents, states = c("tI", "tR"),
                                     min_max_time = c(0, max_time))
-  sir_group_1 <- sir_group %>% filter(AGE2 == 1)
[…]

### 2020-10-08 dev/flu_analysis.Rmd

@@
 flu_season_info_df <- data.frame(year = start_year:end_year) %>%
-  mutate(start_date = paste0(as.character(year), "-10-01"),
-         end_date = paste0(as.character(year+1), "-08-30"),
+  mutate(start_date = paste0(as.character(year), "-09-16"), # splitting september equally
+         end_date = paste0(as.character(year+1), "-09-15"),
          even_year = c( "odd", "even")[(year %% 2) + 1]) %>%

### 2020-10-15 tests/testthat/test-simulate-agents-multinomial-updater.R
Added a draft grouped SIR model vignette.

@@ -2,7 +2,6 @@
 ## May 15, 2020
 
 
-
 test_that("multinomial_updater", {
     ## This is the SIR representation
     D_mat <- matrix(c("X0 * (1 - X1 * par1 / N)", "X0 * X1  * par1 / N", "0",

### 2020-11-05 jss/jss-article-rnw/chapters/ch3.1-r0.Rnw
updated outline

@@
 $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, and is commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
-The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILL IN}).
+The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILLIN}).
 

### 2020-11-05 jss/jss-article-rnw/chapters/ch3.2-ternary.Rnw
updated outline

@@
 
-To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick \cite{CITE THIS?} SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
+To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick \cite{CITETHIS?} SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
 The movement of individuals among the states is given by the following differential equations (Eq.

### 2020-11-05 jss/jss-article-rnw/chapters/ch1-intro-v2.Rnw

@@
 
-As the availability of scientific resources has increased through the decades, the number of models to analyze a particular disease outbreak has also increased.
+The availability of scientific resources has increased through the decades, the number of models to analyze a particular disease outbreak has also increased.
 This phenomenon is evidenced by the appearance of dozens of publicly available estimates of the reproduction number $R_0$ for COVID-19 \citep[e.g.][]{Sanche2020a}, and is well documented in many literature reviews in epidemology \citep[e.g.][]{Canzani2015}.
@@
 The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
-\pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-invariant tools (motivated in more detail in Section \ref{sec:time-invariant}).More broadly, \pkg{EpiCompare} aids the data analysis pipeline for infecetious diseases by
+\pkg{EpiCompare}'s major unique contribution to the existing approaches is a set of time-invariant tools (motivated in more detail in Section \ref{sec:time-invariant}).
+More broadly, \pkg{EpiCompare} aids the data analysis pipeline for infectious diseases by
 \begin{enumerate}

### 2020-11-05 jss/jss-article-rnw/chapters/outline.Rmd

---
title: "Basic simulations with EpiCompare"
output: html_document
bibliography: our-bib.bib
vignette: >
  %\VignetteIndexEntry{Basic simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE)
devtools::load_all()
library(EpiCompare)
library(tidyr)
library(dplyr)
library(kableExtra)
theme_set(theme_bw() + theme(axis.title = element_text()))
```

## Overview

In this vignette, we explore agent-based models with EpiCompare.  Specifically, we

1. Define an agent-based model (ABM).

2. Briefly explain the relationship between ABMs and aggregate models.

3. Demonstrate how EpiCompare can be used to create and visualize basic agent-based models.

## What is an agent-based model?

## ABMs vs. aggregate models

There are numerous sophisticated software to create stochastic aggregate models (e.g. [@pomp2010] or [@jeness2018-epimodel]).  Over the past decade, the number of ABMs has also increased (e.g. FRED, VBI).  In our experience, ABMs are less accessible to the typical scientist because of privacy concerns or computational concerns.

One thing we can do with ABMs is to mimic an aggregate model.  While a compartment model would only look at group-level information with respect to an epidemic (e.g. the number of susceptible, infectious, and recovered at each time step), an agent-based model would allow us to track any and all agents over time.

## Using EpiCompare with basic ABMs

EpiCompare creates stochastic ABMs that are based on the ODEs used to describe transitions among epidemic states.  Note that the functions in EpiCompare are based on **discrete** time-steps or difference equations as opposed to ODEs.

Recall that a SIR model without birth and death in a population of size $N$ and initial population $(S(0), I(0), R(0))$ is given by the following set of equations:

$$
\frac{dS}{dt} = -\beta \frac{SI}{N}
$$

$$
\frac{dI}{dt} = \beta \frac{SI}{N} - \gamma I
$$

$$
\frac{dR}{dt} = \gamma I.
$$

We could instead use a transition matrix (cite) to describe the **non-negative** number of individuals moving from state $i$ to state $j$ (either the instantaneous rate of change or the difference from one time step to the next).

For example, the SIR ODEs describe $K=3$ states and so our transition matrix is

$$
D = \left (\begin{array}{ccc}
S\left (1 - \beta \frac{I}{N} \right ) & S\beta \frac{I}{N} & 0\\
0 & I(1 - \gamma) & I\gamma\\
0 & 0 & R
\end{array} \right ),
$$
where entry (1,2) corresponds to the movement of individuals in state S to state I.

Once we have $D$, we can create a stochastic analog of the model by deriving a transition probability matrix $P$ where entry $P_{ij} = \frac{D_{ij} }{\sum_{k=1}^K D_{ik}}$ is the normalized probability of moving from the current state (the row) to the next state (the column).

The stochastic updates in the ABM are then given through Multinomial draws conditioned on an agent's current state at time $t$.  For agent $n$ currently in state $i$, the agent's next state is given by a Categorical draw which is dependent on the transition matrix $D$,
$$
A_n(t+1) |D(t) \sim \mathrm{Categorical} \left (1, P_i(t) \right ).
$$

This procedure can be done simply with EpiCompare with the following code:

```{r}
set.seed(2020)
## This is the SIR representation
trans_mat <- matrix(c("X0 * (1 - X1 * par1 / N)", "X0 * X1  * par1 / N", "0",
                  "0", "X1 * (1 - par2)", "par2 * X1",
                  "0", "0", "X2"), byrow = TRUE, nrow = 3)
rownames(trans_mat) <- c("S", "I", "R")
init_vals <- c(950, 50, 0)
par_vals <- c(par1 = .1, par2 = .03)
max_T <- 100
n_sims <- 10

abm <- simulate_agents(trans_mat,
                       init_vals,
                       par_vals,
                       max_T,
                       n_sims,
                       verbose = FALSE)

head(abm) %>%
  kable() %>% kable_styling(position = "center")
```

The main difference in the output of `simulate_agents` and functions from EpiModel and pomp is that we can determine exactly when each agent moves from one state as opposed to the aggregate total in each state.

We can compare our ABM to aggregate models with `agents_to_aggregate`.

```{r}
agg_model <- abm %>% dplyr::group_by(sim) %>%
  agents_to_aggregate(states = c(I, R)) 
head(agg_model) %>%
      kable() %>% kable_styling(position = "center")

ggplot(agg_model, aes(x = t, y = X1, group= sim, col = factor(sim))) +
         geom_line() + 
  labs(x = "time", y = "Number of infectious", col = "Sim.")

```

The strength of `simulate_agents` is that it can simulate an outbreak for **any** valid input of $D$.

## A cyclical SIR model (SIRS)

In this model, a recovered individual does not acquire lifelong immunity.  The ODEs we use are

$$
\frac{dS}{dt} = -\beta (t) \frac{SI}{N} + \eta R
$$

$$
\frac{dI}{dt} = \beta (t) \frac{SI}{N} - \gamma I
$$

$$
\frac{dR}{dt} = \gamma I - \eta R.
$$
where $\beta(t) = b + sin(t)$, which allows for a cyclical infection rate.

```{r}
## SIRS

## This is the SIRS representation
trans_mat <- matrix(c("X0 * (1 - X1 * (par1  +  sin(t)) / N)", "X0 * X1  * (par1 + sin(t)) / N", "0",
                  "0", "X1 * (1 - par2)", "par2 * X1",
                  "X2 * par3", "0", "X2 * (1 - par3)"), byrow = TRUE, nrow = 3)
rownames(trans_mat) <- c("S", "I", "R")
colnames(trans_mat) <- c("S", "I", "R")
init_vals <- c(950, 50, 0)
par_vals <- c(par1 = .5, par2 = .25, par3 = .05)
max_T <- 100
n_sims <- 100

## Print the transition matrix
kable(trans_mat) %>% kable_styling(positio = "center")

```

```{r}

## Warning:  will take about a minute.  Set `verbose = TRUE` to see progress
set.seed(2020)
abm <- simulate_agents(trans_mat,
                       init_vals,
                       par_vals,
                       max_T,
                       n_sims,
                       verbose = FALSE,
                       out_format = "long")

head(abm) %>%
  kable() %>% kable_styling(position = "center")
```

In the above table, it is clear that the agent can become susceptible again.

## References

### 2020-11-05 jss/jss-article-rnw/chapters/outline.Rnw

% !Rnw root = ../article.Rnw

\begin{enumerate}

\item Intro

\begin{enumerate}

\item Lots of epidemic \textit{infectious disease} models.  How do we handle them all

\item Pipeline of analysis

\item past work - many things under the wider umbrella of compartment models / multi state models (where one state is susceptible and one state is infectious)

\begin{itemize}
\item surveillance  -- one of the first major infectious disease R packages ED (kinda like it's own whole pipeline?)

\item pomp, epinet -- standardized frameworks of large classes of infectious disease models - modeling and simulation

\item MIDAS has done some work to catalogue different tools and models post- processing and comparison

\item Bring in tidyverse - standardization (plz throw spew a bone) pre-process, EDA and post processing

\item Patient vs. aggregate data (data pre-processing and EDA)

\item conformal bands and visualization for model assessment (skree plots, your hurricanes)

\end{itemize}
\end{enumerate}

\item EpiCompare qualities

\begin{enumerate}

\item standardizing outputs

\item TERNARY PLOTS

\item simple ode-style (also tidy-style) abms

\item time variant and analysis and conformal prediction bands

\end{enumerate}

\item Tour of EpiCompare

\begin{enumerate}
\item Highlights of Hagelloch
\item EDA by class
\item Simulation
\item Comparison with EpiNet model
\end{enumerate}

\end{enumerate}

### 2020-11-09 README.Rmd
first clean draft of depth-impact-on-prediction vignette.

@@
 To "correct" this, we've over-wrote `ggtern`'s `print.ggplot` to correct this problem when not producing ternary plots (code in `aaa.R`).
+6.
+Useful Rstudio shortcuts for `Roxygen2`: (a) create `Roxygen2` comments template with `option` + `command` + `shift` + `R` (b) In `Roxygen2` comments do `control` + `shift` + `/` to format relative to 80 char limit.
+7.[stack overflow](https://stackoverflow.com/questions/11285496/r-cmd-check-warning-functions-methods-with-usage-in-documentation-object-bu) post on how to pass `check` for `rownames<-.tidy_dist_mat`.
 

### 2020-11-09 tests/testthat/test-agent-to-aggregate.R

@@ -1,6 +1,7 @@
 context("tests for agents_to_aggregate")
 
-library(tidyverse)
+library(dplyr)
+library(tidyr)
 
 test_that("checking agents_to_aggregate works for SI model",{
   

### 2020-11-19 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Rmd version of paper

---
documentclass: jss
author:
  - name: Shannon K. Gallagher
    affiliation: |
      | Biostatistics Research Branch
      | National Institute of Allergy 
      | and Infectious Diseases
    # use this syntax to add text on several lines
    address: |
      | 5603 Fishers Lane
      | Rockville, MD 20852
    email: \email{shannon.gallagher@nih.gov}
    url: http://skgallagher.github.io
  - name: Benjamin Leroy
    affiliation: |
      | Dept. of Statistics
      | Carnegie Mellon University
    address: |
      | 5000 Forbes Ave.
      | Pittsburgh, PA 15213
    email: \email{bpleroy@andrew.cmu.edu}
    url:  https://benjaminleroy.github.io/

title:
  formatted: "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
  # If you use tex in the formatted title, also supply version without
  plain:     "A Capitalized Title: Something about a Package foo"
  # For running headers, if needed
  short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: EpiCompare.bib  #Also not included in template by default. 
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
options(kableExtra.latex.load_packages = FALSE)

knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      fig.align = "c",
                      echo=TRUE, warning=FALSE, message=FALSE)
```

\section{Introduction}\label{sec:intro}

This is the Section \ref{sec:intro}. This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.

## Code formatting

In general, don't use Markdown, but use the more precise LaTeX commands instead:

* \proglang{Java}
* \pkg{plyr}

One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.

# \proglang{R} code {short-title="R code" #r-code}

Can be inserted in regular R markdown blocks.

hags hags hags \cite{Neal2004}

```{r}
x <- 1:10
x
```

## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}

* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.

In this section, we highlight a number of the functionalities available in \proglang{EpiCompare}.  These functionalities include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline REF\ref{}. We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner.

## Data and EDA

We analyze an outbreak of measles in the town of Hagelloch, Germany from 1861-1862, a data set organized by \cite{pfeilsticker1863}.  The data was later made visible by \cite{oesterle1992} and made available in an \proglang{R} by \cite{surveillance2017}.  The Hagelloch data includes a rich set of features including household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.  Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.

```{r}
devtools::load_all()
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
library(kableExtra)
```

```{r hagelloch-subset-view}
hagelloch_raw %>% select(PN, NAME, AGE, SEX, HN, PRO, ERU, IFTO) %>%
  head(5) %>% kable(format = "latex", booktabs = TRUE, caption = "Cool table") %>%
  kable_styling(position = "center")
```

\begin{itemize}
\item thing
\end{itemize}

### 2020-11-19 jss/EpiCompare-paper/EpiCompare-paper.tex
Rmd version of paper

% To remove when JSS.cls is fixed
% https://github.com/rstudio/rticles/issues/329
\RequirePackage[2020/02/02]{latexrelease}
\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Shannon K. Gallagher\\Biostatistics Research Branch\\
National Institute of Allergy\\
and Infectious Diseases \And Benjamin Leroy\\Dept. of Statistics\\
Carnegie Mellon University
}
\title{Time invariant analysis of epidemics with \pkg{EpiCompare}}

\Plainauthor{Shannon K. Gallagher, Benjamin Leroy}
\Plaintitle{A Capitalized Title: Something about a Package foo}
\Shorttitle{\pkg{foo}: A Capitalized Title}

\Abstract{
The abstract of the article.
}

\Keywords{keywords, not capitalized, \proglang{Java}}
\Plainkeywords{keywords, not capitalized, Java}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Shannon K. Gallagher\\
    Biostatistics Research Branch\\
National Institute of Allergy\\
and Infectious Diseases\\
    5603 Fishers Lane\\
Rockville, MD 20852\\
  E-mail: \email{shannon.gallagher@nih.gov}\\
  URL: \url{http://skgallagher.github.io}\\~\\
      Benjamin Leroy\\
    Dept. of Statistics\\
Carnegie Mellon University\\
    5000 Forbes Ave.\\
Pittsburgh, PA 15213\\
  E-mail: \email{bpleroy@andrew.cmu.edu}\\
  URL: \url{https://benjaminleroy.github.io/}\\~\\
  }

% Pandoc citation processing

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}

\usepackage{amsmath}

\begin{document}

\section{Introduction}\label{sec:intro}

This is the Section \ref{sec:intro}. This template demonstrates some of
the basic LaTeX that you need to know to create a JSS article.

\hypertarget{code-formatting}{%
\subsection{Code formatting}\label{code-formatting}}

In general, don't use Markdown, but use the more precise LaTeX commands
instead:

\begin{itemize}
\item
  \proglang{Java}
\item
  \pkg{plyr}
\end{itemize}

One exception is inline code, which can be written inside a pair of
backticks (i.e., using the Markdown syntax).

If you want to use LaTeX commands in headers, you need to provide a
\texttt{short-title} attribute. You can also provide a custom identifier
if necessary. See the header of Section \ref{r-code} for example.

\section[R code]{\proglang{R} code}\label{r-code}

Can be inserted in regular R markdown blocks.

hags hags hags \cite{Neal2004}

\begin{CodeChunk}
\begin{CodeInput}
R> x <- 1:10
R> x
\end{CodeInput}
\begin{CodeOutput}
 [1]  1  2  3  4  5  6  7  8  9 10
\end{CodeOutput}
\end{CodeChunk}

\subsection[Features specific to rticles]{Features specific to
\pkg{rticles}}\label{features-specific-to}

\begin{itemize}
\tightlist
\item
  Adding short titles to section headers is a feature specific to
  \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is
  currently not supported by Pandoc and we will update this template if
  \href{https://github.com/jgm/pandoc/issues/4409}{it is officially
  supported in the future}.
\item
  Using the \texttt{\textbackslash{}AND} syntax in the \texttt{author}
  field to add authors on a new line. This is a specific to the
  \texttt{rticles::jss\_article} format.
\end{itemize}

In this section, we highlight a number of the functionalities available
in \proglang{EpiCompare}. These functionalities include data cleaning,
visualization, simulation, and comparison, in accordance with the data
analysis pipeline REF\ref{}. We show a full data analysis from beginning
to end that can be accomplished in a streamlined and standardized
manner.

\hypertarget{data-and-eda}{%
\subsection{Data and EDA}\label{data-and-eda}}

We analyze an outbreak of measles in the town of Hagelloch, Germany from
1861-1862, a data set organized by \cite{pfeilsticker1863}. The data was
later made visible by \cite{oesterle1992} and made available in an
\proglang{R} by \cite{surveillance2017}. The Hagelloch data includes a
rich set of features including household members, school level,
household locations, date of first symptoms (prodromes), date of measles
rash, and even the alleged infector. Because of these rich features,
this data set has been an ideal testing ground methodology in infectious
disease epidemiology and is used in work by
\cite{Neal2004,britton2011,groendyke2012,becker2016}.

\begin{CodeChunk}
\begin{CodeInput}
R> devtools::load_all()
R> library(ggplot2)
R> library(tidyr)
R> library(dplyr)
R> library(knitr)
R> library(kableExtra)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> hagelloch_raw %>% select(PN, NAME, AGE, SEX, HN, PRO, ERU, IFTO) %>%
+   head(5) %>% kable(format = "latex", booktabs = TRUE, caption = "Cool table") %>%
+   kable_styling(position = "center")
\end{CodeInput}
\begin{table}

\caption{\label{tab:hagelloch-subset-view}Cool table}
\centering
\begin{tabular}[t]{rlrlrllr}
\toprule
PN & NAME & AGE & SEX & HN & PRO & ERU & IFTO\\
\midrule
1 & Mueller & 7 & female & 61 & 1861-11-21 & 1861-11-25 & 45\\
2 & Mueller & 6 & female & 61 & 1861-11-23 & 1861-11-27 & 45\\
3 & Mueller & 4 & female & 61 & 1861-11-28 & 1861-12-02 & 172\\
4 & Seibold & 13 & male & 62 & 1861-11-27 & 1861-11-28 & 180\\
5 & Motzer & 8 & female & 63 & 1861-11-22 & 1861-11-27 & 45\\
\bottomrule
\end{tabular}
\end{table}

\end{CodeChunk}

\begin{itemize}
\item thing
\end{itemize}

\bibliography{EpiCompare.bib}

\end{document}

### 2020-11-20 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Finally getting words on paper

@@
     citation_package: "natbib"  #All my citations use biblatex, not natbib. 
+classoption: shortnames
 biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
@@
 knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
-                      fig.align = "c",
-                      echo=TRUE, warning=FALSE, message=FALSE)
+                      fig.align = "center",
+                      echo=TRUE, warning=FALSE, message=FALSE,
+                      fig.pos = "H")
+
 ```
@@
 
-## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}
-
-* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter).
-This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
-* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.
+# A tour of \pkg{EpiCompare} {short-title="Tour" #sec:tour}
 
@@
 
-## Data and EDA
+\subsection{Data and exploratory analysis}
 
@@
 The Hagelloch data includes a rich set of features including household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.
-Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.
+A subset of the data is shown in Table \ref{tab:hags-people}   Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.
 
-```{r}
+After loading in the libraries for this analysis, we examine .  Within it, we
+
+```{r echo = FALSE}
 devtools::load_all()
-library(ggplot2)
 library(tidyr)
@@
 library(kableExtra)
+library(ggplot2)
+
+theme_set(theme_bw(base_size = 20))
 ```
 
-```{r hagelloch-subset-view}
-hagelloch_raw %>% select(PN, NAME, AGE, SEX, HN, PRO, ERU, IFTO) %>%
-  head(5) %>% kable(format = "latex", booktabs = TRUE, caption = "Cool table") %>%
-  kable_styling(position = "center")
+```{r hagelloch-subset-view, echo = FALSE}
+hagelloch_raw %>% select(PN, HN, NAME, AGE, SEX, CL,  PRO, ERU, IFTO) %>%
+  filter(PN %in% paste(c(1:5, 45))) %>% kable(format = "latex", booktabs = TRUE, caption = "Subset of Hagelloch infection data.
+Features include the person ID, household ID (HH ID), age, sex, class level (Pre-K/1st/2nd), date of first symptoms, date of the appearance of the measles rash, and the alleged infector ID of the individual.",
+                                              linesep = "",
+                    col.names = c("ID","HH ID", "Name", "Age", "Sex", 
+                                  "Class", "Symp. Start", "Rash Date", "Infector ID"),
+                    label = "hags-people") %>%
+  kable_styling(position = "center", latex_options = "hold_position")
+```
+
+With \pkg{EpiCompare}, we can easily obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate}.
+The function \code{agents_to_aggregate} is a key component of \pkg{EpiCompare}, allowing the user to easily switch from an individual-level (i.e.
+an agent) view of a disease to an aggregate level.
+For example, the below code shows how we can convert the agent data to a cumulative incidence of the measles rash, in order to see how the disease spread through the population over time.
+We can then compare the cumulative incidence of the rash to the cumulative incidence of the prodromes, i.e.
+the initial symptoms.
+We do this with the below code, and a part of the cumulative incidence data output are shown in Table \ref{tab:cif-rash}.
+The argument \code{integer_time_expansion} indicates whether we should include all time points in the recorded range of the data or only when there is a change in the incidence.
+
+```{r }
+cif_rash  <- hagelloch_raw %>%
+  mutate(time_of_rash = as.numeric(ERU - min(PRO, na.rm = TRUE))) %>%
+  agents_to_aggregate(states = time_of_rash,
+                      integer_time_expansion = FALSE) %>%
+  mutate(type = "Rash")
+```
+
+```{r echo = FALSE}
+cif_rash %>% 
+  dplyr::select(-type) %>%
+  head(5) %>% kable(booktabs = TRUE,
+                               caption = "Turning the individual-level information from the Hagelloch data to an aggregate view of the cumulative incidence of the measles rash in the population over time.",
+                               label = "cif-rash",
+                              col.names = c("Time", "# Susceptible", "# Total rash appearances") ) %>%
+  kable_styling(position = "center", 
+                latex_options = "hold_position"
+                      )
+```
+
+One question of interest is the duration between initial onset of prodromes or symptoms and the appearance of the measles rash.
+Since \code{agent_to_aggregate} outputs a tidy-style data frame, it is a simple task to plot the two sets of incidence curves on the same graph (Fig.
+\ref{fig:cifs}).
+
+```{r}
+cif_prodromes <- hagelloch_raw %>%
+  mutate(time_of_PRO = as.numeric(PRO - min(PRO, na.rm = TRUE))) %>%
+  agents_to_aggregate(states = time_of_PRO,
+                      integer_time_expansion = FALSE) %>%
+  mutate(type = "Pro")
+```
+
+```{r fig.cap = "\\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.
+We see that there is approximately a a constant lag between the two curves."}
+plot_df <- bind_rows(cif_rash, cif_prodromes)
+
+ggplot(data = plot_df,
+       aes(x = t, y = X1, col = type)) + 
+  geom_step() + 
+  labs(title = "Cumulative incidence of measles appearance",
+       x = "Time (days relative to first prodrome appearance)",
+       y = "Cumulative incidence of event") + 
+  coord_cartesian(xlim = c(0, 55)) +
+  scale_color_manual(values = c("blue", "red"))
+
+```
+
+```{r}
+with(hagelloch_raw, sum(PRO > ERU, na.rm = TRUE))
+
+```
+
+The real power of \code{agents_to_aggregate()} lies in its ability to aggregate over any number of pre-specified states.
+For example, the Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the time of infection and recovery, respectively of each individual.
+We can then plot the SIR values through a time-invariant lens via a ternary plot using regular \pkg{ggplot2} functions or with our custom \code{geom}, \code{geom_aggregate}, which takes the original agent data as input  (\ref{fig:hag-tern-raw}.
+
+```{r fig.cap = "\\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.
+We see there are two peaks of infection (the vertical axis)."}
+sir <- hagelloch_raw %>%
+  agents_to_aggregate(states = c(tI, tR),
+                      min_max_time = c(0, 55)) %>%
+  rename(time = t, S = X0, I = X1, R = X2)
+
+ggplot(sir, aes(x = S, y = I, z = R))+
+  coord_tern() +
+  geom_path() +
+  labs(x = "S", y = "I", z = "R",
+       title = "Time invariant view of Hagelloch measles outbreak") + 
+  theme_sir()
+
+```
+
+Moreover, we can look at the outbreaks of the disease by group within \code{agent_to_aggregate()} or \code{geom_aggregate()}, which allows us to examine differences among the groups of individuals.
+For example, we show the time invariant outbreak by class level in Figure \ref{fig:tern-class-data}.
+Immediately, we see that time invariant infection curve is very different for the pre-school group compared to the 1st class.
+In the 1st class, we see about 95\% of the class become infected before anyone recovers, which is indicative of a super-spreading event.
+This suspicion is further confirmed in that `r max(table(hagelloch_raw$IFTO[hagelloch_raw$CL == "1st class"]))` of the `r nrow(hagelloch_raw[hagelloch_raw$CL == "1st class",])` have been reportedly infected by the same individual.
+
+```{r fig.cap = "\\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.
+The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined."}
+hagelloch_raw %>%
+  ggplot(aes(y = tI, z = tR, color = CL)) +
+  geom_aggregate(size = 2) + coord_tern() +
+  labs(x = "S", y = "I", z = "R",
+       color = "Class") +
+  scale_color_brewer(palette = "Dark2") +
+  facet_wrap(~CL)
 ```

### 2020-11-20 jss/EpiCompare-paper/EpiCompare-paper.tex
Finally getting words on paper

@@
 \documentclass[
-]{jss}
+  shortnames]{jss}
 
@@
 
-\subsection[Features specific to rticles]{Features specific to
-\pkg{rticles}}\label{features-specific-to}
-
-\begin{itemize}
-\tightlist
-\item
-  Adding short titles to section headers is a feature specific to
-  \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is
-  currently not supported by Pandoc and we will update this template if
-  \href{https://github.com/jgm/pandoc/issues/4409}{it is officially
-  supported in the future}.
-\item
-  Using the \texttt{\textbackslash{}AND} syntax in the \texttt{author}
-  field to add authors on a new line. This is a specific to the
-  \texttt{rticles::jss\_article} format.
-\end{itemize}
+\section[Tour]{A tour of \pkg{EpiCompare}}\label{sec:tour}
 
@@
 
-\hypertarget{data-and-eda}{%
-\subsection{Data and EDA}\label{data-and-eda}}
+\subsection{Data and exploratory analysis}
 
@@
 household locations, date of first symptoms (prodromes), date of measles
-rash, and even the alleged infector. Because of these rich features,
-this data set has been an ideal testing ground methodology in infectious
-disease epidemiology and is used in work by
+rash, and even the alleged infector. A subset of the data is shown in
+Table \ref{tab:hags-people} Because of these rich features, this data
+set has been an ideal testing ground methodology in infectious disease
+epidemiology and is used in work by
 \cite{Neal2004,britton2011,groendyke2012,becker2016}.
+
+After loading in the libraries for this analysis, we examine . Within
+it, we
+
+\begin{CodeChunk}
+\begin{table}[!h]
+
+\caption{\label{tab:hags-people}Subset of Hagelloch infection data.
+Features include the person ID, household ID (HH ID), age, sex, class level (Pre-K/1st/2nd), date of first symptoms, date of the appearance of the measles rash, and the alleged infector ID of the individual.}
+\centering
+\begin{tabular}[t]{rrlrllllr}
+\toprule
+ID & HH ID & Name & Age & Sex & Class & Symp. Start & Rash Date & Infector ID\\
+\midrule
+1 & 61 & Mueller & 7 & female & 1st class & 1861-11-21 & 1861-11-25 & 45\\
+2 & 61 & Mueller & 6 & female & 1st class & 1861-11-23 & 1861-11-27 & 45\\
+3 & 61 & Mueller & 4 & female & preschool & 1861-11-28 & 1861-12-02 & 172\\
+4 & 62 & Seibold & 13 & male & 2nd class & 1861-11-27 & 1861-11-28 & 180\\
+5 & 63 & Motzer & 8 & female & 1st class & 1861-11-22 & 1861-11-27 & 45\\
+45 & 51 & Goehring & 7 & male & 1st class & 1861-11-11 & 1861-11-13 & 184\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
+With \pkg{EpiCompare}, we can easily obtain the empirical cumulative
+incidence function with respect to the measles rash appearance (variable
+\code{ERU}) with the following tidy-style function,
+\code{agents_to_aggregate}. The function \code{agents_to_aggregate} is a
+key component of \pkg{EpiCompare}, allowing the user to easily switch
+from an individual-level (i.e.~an agent) view of a disease to an
+aggregate level. For example, the below code shows how we can convert
+the agent data to a cumulative incidence of the measles rash, in order
+to see how the disease spread through the population over time. We can
+then compare the cumulative incidence of the rash to the cumulative
+incidence of the prodromes, i.e.~the initial symptoms. We do this with
+the below code, and a part of the cumulative incidence data output are
+shown in Table \ref{tab:cif-rash}. The argument
+\code{integer_time_expansion} indicates whether we should include all
+time points in the recorded range of the data or only when there is a
+change in the incidence.
 
@@
 \begin{CodeInput}
-R> devtools::load_all()
-R> library(ggplot2)
-R> library(tidyr)
-R> library(dplyr)
-R> library(knitr)
-R> library(kableExtra)
+R> cif_rash  <- hagelloch_raw %>%
++   mutate(time_of_rash = as.numeric(ERU - min(PRO, na.rm = TRUE))) %>%
++   agents_to_aggregate(states = time_of_rash,
++                       integer_time_expansion = FALSE) %>%
++   mutate(type = "Rash")
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}[!h]
+
+\caption{\label{tab:cif-rash}Turning the individual-level information from the Hagelloch data to an aggregate view of the cumulative incidence of the measles rash in the population over time.}
+\centering
+\begin{tabular}[t]{rrr}
+\toprule
+Time & \# Susceptible & \# Total rash appearances\\
+\midrule
+0 & 188 & 0\\
+4 & 187 & 1\\
+7 & 186 & 2\\
+9 & 185 & 3\\
+12 & 183 & 5\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
+One question of interest is the duration between initial onset of
+prodromes or symptoms and the appearance of the measles rash. Since
+\code{agent_to_aggregate} outputs a tidy-style data frame, it is a
+simple task to plot the two sets of incidence curves on the same graph
+(Fig. \ref{fig:cifs}).
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> cif_prodromes <- hagelloch_raw %>%
++   mutate(time_of_PRO = as.numeric(PRO - min(PRO, na.rm = TRUE))) %>%
++   agents_to_aggregate(states = time_of_PRO,
++                       integer_time_expansion = FALSE) %>%
++   mutate(type = "Pro")
 \end{CodeInput}
@@
 \begin{CodeInput}
-R> hagelloch_raw %>% select(PN, NAME, AGE, SEX, HN, PRO, ERU, IFTO) %>%
-+   head(5) %>% kable(format = "latex", booktabs = TRUE, caption = "Cool table") %>%
-+   kable_styling(position = "center")
+R> plot_df <- bind_rows(cif_rash, cif_prodromes)
+R> 
+R> ggplot(data = plot_df,
++        aes(x = t, y = X1, col = type)) + 
++   geom_step() + 
++   labs(title = "Cumulative incidence of measles appearance",
++        x = "Time (days relative to first prodrome appearance)",
++        y = "Cumulative incidence of event") + 
++   coord_cartesian(xlim = c(0, 55)) +
++   scale_color_manual(values = c("blue", "red"))
 \end{CodeInput}
-\begin{table}
+\begin{figure}[H]
 
-\caption{\label{tab:hagelloch-subset-view}Cool table}
-\centering
-\begin{tabular}[t]{rlrlrllr}
-\toprule
-PN & NAME & AGE & SEX & HN & PRO & ERU & IFTO\\
-\midrule
-1 & Mueller & 7 & female & 61 & 1861-11-21 & 1861-11-25 & 45\\
-2 & Mueller & 6 & female & 61 & 1861-11-23 & 1861-11-27 & 45\\
-3 & Mueller & 4 & female & 61 & 1861-11-28 & 1861-12-02 & 172\\
-4 & Seibold & 13 & male & 62 & 1861-11-27 & 1861-11-28 & 180\\
-5 & Motzer & 8 & female & 63 & 1861-11-22 & 1861-11-27 & 45\\
-\bottomrule
-\end{tabular}
-\end{table}
+{\centering \includegraphics{Figs/unnamed-chunk-6-1}
 
+}
+
+\caption{\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.
+We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-6}
+\end{figure}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> with(hagelloch_raw, sum(PRO > ERU, na.rm = TRUE))
+\end{CodeInput}
+\begin{CodeOutput}
+[1] 0
+\end{CodeOutput}
+\end{CodeChunk}
+
+The real power of \code{agents_to_aggregate()} lies in its ability to
+aggregate over any number of pre-specified states. For example, the
+Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the
+time of infection and recovery, respectively of each individual. We can
+then plot the SIR values through a time-invariant lens via a ternary
+plot using regular \pkg{ggplot2} functions or with our custom
+\code{geom}, \code{geom_aggregate}, which takes the original agent data
+as input (\ref{fig:hag-tern-raw}.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> sir <- hagelloch_raw %>%
++   agents_to_aggregate(states = c(tI, tR),
++                       min_max_time = c(0, 55)) %>%
++   rename(time = t, S = X0, I = X1, R = X2)
+R> 
+R> 
+R> ggplot(sir, aes(x = S, y = I, z = R))+
++   coord_tern() +
++   geom_path() +
++   labs(x = "S", y = "I", z = "R",
++        title = "Time invariant view of Hagelloch measles outbreak") + 
++   theme_sir()
+\end{CodeInput}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-8-1}
+
+}
+
+\caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.
+We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-8}
+\end{figure}
+\end{CodeChunk}
+
+Moreover, we can look at the outbreaks of the disease by group within
+\code{agent_to_aggregate()} or \code{geom_aggregate()}, which allows us
+to examine differences among the groups of individuals. For example, we
+show the time invariant outbreak by class level in Figure
+\ref{fig:tern-class-data}. Immediately, we see that time invariant
+infection curve is very different for the pre-school group compared to
+the 1st class. In the 1st class, we see about 95\% of the class become
+infected before anyone recovers, which is indicative of a
+super-spreading event. This suspicion is further confirmed in that 26 of
+the 30 have been reportedly infected by the same individual.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> hagelloch_raw %>%
++   ggplot(aes(y = tI, z = tR, color = CL)) +
++   geom_aggregate(size = 2) + coord_tern() +
++   labs(x = "S", y = "I", z = "R",
++        color = "Class") +
++   scale_color_brewer(palette = "Dark2") +
++   facet_wrap(~CL)
+\end{CodeInput}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-9-1}
+
+}
+
+\caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.
+The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-9}
+\end{figure}
 \end{CodeChunk}

### 2020-11-20 jss/jss-article-rnw/article.Rnw
updates to introduction and bibtech
minor updates to introduction and shortnames update
update of article to correctly do bibtex. Note: must "compile" from article.tex to get references in document
literature review option and time-free motivation and presention section included
more outline provided - "Orientation" section proposed
breaking up chapter 4
Made an intro I'm pretty happy with?
update of simplex and model comparison in motivation section - this is very rough update.
clean up of paper from May 18th discussion
minor updates to paper article
Finally getting words on paper

@@
-\documentclass[article]{jss}
+\documentclass[article,shortnames]{jss} 
+%^note: Ben changed this to get shortnames for references ("shortnames")
 
@@
 
-%% another package (only for this demo article)
-\usepackage{framed}
-
 %% new custom commands
@@
 \newcommand{\fct}[1]{\code{#1()}}
+%\newcommand{\sup}{\mathop{\mathrm{sup}}}
+%\newcommand{\inf}{\mathop{\mathrm{inf}}}
 
-\usepackage{amsmath}
-\usepackage{amsthm}
-\newtheorem{theorem}{Theorem}
+\usepackage{amsfonts,amsmath,amssymb,amsthm}
+%\newtheorem{theorem}{Theorem}
 
-%% For Sweave-based articles about R packages:
-%% need no \usepackage{Sweave}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
-<<preliminaries, echo=FALSE, results=hide>>=
-options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
-library("MASS")
-@
+% note making commands
+\newcommand{\ben}{\textcolor{blue}}
+\newcommand{\skg}{\textcolor{red}}
+
+% useful packages
+\usepackage[normalem]{ulem} % for \sout
 
@@
 \author{Shannon K. Gallagher\\ National Institute of\\ Allergy and Infectious Diseases
-   \And Benjamin P. Leroy\\Carnegie Mellon University}
-\Plainauthor{Shannon K. Gallagher, Benjamin P. Leroy}
+   \And Benjamin LeRoy\\Statistics and Data Science Department\\Carnegie Mellon University}
+\Plainauthor{Shannon K. Gallagher, Benjamin LeRoy}
 
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
+% EpiCompare: A tidy R package for comparing and assessing epidemics and epidimology models
 \title{\proglang{EpiCompare}: Working title}
@@
 \Abstract{
-We do something
+We do something, possibly useful
 }
@@
 \begin{document}
+\SweaveOpts{concordance=FALSE}
 
-\SweaveInput{chapters/ch1-intro.Rnw}
+\section{Outline}
 
-\section{Purpose and Details}
-\SweaveInput{chapters/ch2-vis.Rnw}
-\SweaveInput{chapters/ch3-compare.Rnw}
-\SweaveInput{chapters/ch4-convert.Rnw}
+\SweaveInput{chapters/outline.Rnw}
 
-\section{In practice}
-\SweaveInput{chapters/ch5-ex1.Rnw}
-\SweaveInput{chapters/ch6-ex2.Rnw}
+\SweaveInput{chapters/ch1-intro-v2.Rnw}
+
+%\SweaveInput{chapters/ch1-intro.Rnw}
+
+\section{Literature review} \label{sec:lit}
+\SweaveInput{chapters/ch2-lit-review.Rnw}
+
+\section{Reasons and tools for time-invariant analysis} \label{sec:time-invariant}
+\SweaveInput{chapters/ch3-time-free.Rnw}
+
+%\SweaveInput{chapters/ch2-vis.Rnw} % this file still exists
+%\SweaveInput{chapters/ch3-compare.Rnw}
+%\SweaveInput{chapters/ch4-convert.Rnw}
+
+\section{Code Orientation}
+\SweaveInput{chapters/ch4-orientation.Rnw}
+
+\section{A tour of EpiCompare}
+\SweaveInput{chapters/hagelloch-ex.Rnw}
 
@@
 
-\section{references}
+\bibliography{EpiCompare}
 
-\bibliography{EpiCompare}
+\appendix
 

### 2020-11-20 jss/jss-article-rnw/chapters/hagelloch-ex.Rnw
Finally getting words on paper

--- 2020-11-19 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
----
-documentclass: jss
-author:
-  - name: Shannon K. Gallagher
-    affiliation: |
-      | Biostatistics Research Branch
-      | National Institute of Allergy 
-      | and Infectious Diseases
-    # use this syntax to add text on several lines
-    address: |
-      | 5603 Fishers Lane
-      | Rockville, MD 20852
-    email: \email{shannon.gallagher@nih.gov}
-    url: http://skgallagher.github.io
-  - name: Benjamin Leroy
-    affiliation: |
-      | Dept. of Statistics
-      | Carnegie Mellon University
-    address: |
-      | 5000 Forbes Ave.
-      | Pittsburgh, PA 15213
-    email: \email{bpleroy@andrew.cmu.edu}
-    url:  https://benjaminleroy.github.io/
-
-title:
-  formatted: "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
-  # If you use tex in the formatted title, also supply version without
-  plain:     "A Capitalized Title: Something about a Package foo"
-  # For running headers, if needed
-  short:     "\\pkg{foo}: A Capitalized Title"
-abstract: >
-  The abstract of the article.
-keywords:
-  # at least one keyword must be supplied
-  formatted: [keywords, not capitalized, "\\proglang{Java}"]
-  plain:     [keywords, not capitalized, Java]
-preamble: >
-  \usepackage{amsmath}
-output: 
-  rticles::jss_article:
-    number_sections: TRUE     #added argument option 
-    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
-biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
-bibliography: EpiCompare.bib  #Also not included in template by default. 
-header-includes:
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
----
-
-```{r, setup, include=FALSE}
-options(prompt = 'R> ', continue = '+ ')
-options(kableExtra.latex.load_packages = FALSE)
-
-knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
-                      fig.align = "c",
-                      echo=TRUE, warning=FALSE, message=FALSE)
-```
-
-\section{Introduction}\label{sec:intro}
-
-This is the Section \ref{sec:intro}. This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.
-
-## Code formatting
-
-In general, don't use Markdown, but use the more precise LaTeX commands instead:
-
-* \proglang{Java}
-* \pkg{plyr}
-
-One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).
-
-If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.
-
-# \proglang{R} code {short-title="R code" #r-code}
-
-Can be inserted in regular R markdown blocks.
-
-hags hags hags \cite{Neal2004}
-
-```{r}
-x <- 1:10
-x
-```
-
-## Features specific to \pkg{rticles} {short-title="Features specific to rticles"}
-
-* Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter).
-This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
-* Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format.
+% !Rnw root = ../article.Rnw
 
@@
 
-## Data and EDA
-
+\subsection{Data and EDA}
 We analyze an outbreak of measles in the town of Hagelloch, Germany from 1861-1862, a data set organized by \cite{pfeilsticker1863}.
@@
 
-```{r}
-devtools::load_all()
+<<libraries, echo = FALSE>>=
 library(ggplot2)
+library(dplyr)
+library(broom)
 library(tidyr)
-library(dplyr)
 library(knitr)
 library(kableExtra)
-```
+library(RColorBrewer)
+devtools::load_all()
+@
 
-```{r hagelloch-subset-view}
+The Hagelloch data is available in \proglang{EpiCompare} with \proglang{data(hagelloch\_raw)}.  We can see a subset of the data frame with
+
+<<hagelloch-subset-view, results = asis>>=
 hagelloch_raw %>% select(PN, NAME, AGE, SEX, HN, PRO, ERU, IFTO) %>%
-  head(5) %>% kable(format = "latex", booktabs = TRUE, caption = "Cool table") %>%
-  kable_styling(position = "center")
-```
+  head(5) %>% kable(format = "latex", booktabs = TRUE, align = 'c')
+@
 
-\begin{itemize}
-\item thing
-\end{itemize}
+\subsection{Modeling}
+
+\subsection{Comparison}

### 2020-11-21 jss/EpiCompare-paper/EpiCompare-paper.Rmd
paper writing

@@
   # If you use tex in the formatted title, also supply version without
-  plain:     "A Capitalized Title: Something about a Package foo"
+  plain:     "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
   # For running headers, if needed
-  short:     "\\pkg{foo}: A Capitalized Title"
+  short:     "\\pkg{EpiCompare}"
 abstract: >
@@
 
-After loading in the libraries for this analysis, we examine .  Within it, we
-
 ```{r echo = FALSE}
@@
 
-theme_set(theme_bw(base_size = 20))
+theme_set(theme_bw(base_size = 24))
 ```
@@
 For example, the Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the time of infection and recovery, respectively of each individual.
-We can then plot the SIR values through a time-invariant lens via a ternary plot using regular \pkg{ggplot2} functions or with our custom \code{geom}, \code{geom_aggregate}, which takes the original agent data as input  (\ref{fig:hag-tern-raw}.
+We can then plot the SIR values through a time-invariant lens  using \pkg{ggplot2} functions or with our custom \code{geom}, \code{geom_aggregate}, which takes the original agent data as input  (Fig.
+\ref{fig:hag-tern-raw}).
 
@@
 We see there are two peaks of infection (the vertical axis)."}
-sir <- hagelloch_raw %>%
+hagelloch_sir <- hagelloch_raw %>%
   agents_to_aggregate(states = c(tI, tR),
@@
 
-ggplot(sir, aes(x = S, y = I, z = R))+
+ggplot(hagelloch_sir, aes(x = S, y = I, z = R))+
   coord_tern() +
@@
        title = "Time invariant view of Hagelloch measles outbreak") + 
-  theme_sir()
+  theme_sir(base_size = 24)
 
@@
 
-Moreover, we can look at the outbreaks of the disease by group within \code{agent_to_aggregate()} or \code{geom_aggregate()}, which allows us to examine differences among the groups of individuals.
+Moreover, we can look at the outbreaks of the disease by group within \code{agent_to_aggregate()} or \code{geom_aggregate()}.
+This allows us to examine differences among the different groups of individuals.
 For example, we show the time invariant outbreak by class level in Figure \ref{fig:tern-class-data}.
-Immediately, we see that time invariant infection curve is very different for the pre-school group compared to the 1st class.
-In the 1st class, we see about 95\% of the class become infected before anyone recovers, which is indicative of a super-spreading event.
-This suspicion is further confirmed in that `r max(table(hagelloch_raw$IFTO[hagelloch_raw$CL == "1st class"]))` of the `r nrow(hagelloch_raw[hagelloch_raw$CL == "1st class",])` have been reportedly infected by the same individual.
+Immediately, we see that time invariant infection curve is different for the pre-school class compared to the 1st class.
+In the 1st class, we see about 95\% of the class become infected and less than 10\% of them having recovered, which is indicative of a super-spreading event.
+This suspicion is further confirmed in that `r max(table(hagelloch_raw$IFTO[hagelloch_raw$CL == "1st class"]))` of the `r nrow(hagelloch_raw[hagelloch_raw$CL == "1st class",])` 1st class students have been reportedly infected by the same individual.
 
@@
 
+Along with multiple epidemic states, the function \code{agents_to_aggregate} can also be extended to populations with vital dynamics (e.g.
+birth and death) and examples of this are shown in the package vignette.
+In summary, \code{agents_to_aggregate()} is a multi-purpose workhorse that be leveraged to convert individual level records into aggregate information that may be more useful for some forms of epidemic modeling such as compartment modeling.
+
+Up to this point, we have used \pkg{EpiCompare} in the context of observed data.
+We also want to compare statistical models, and \pkg{EpiCompare} aids in that process via a simple but dynamic individual-level data generator, conversion tools for popular epidemic model packages, and model assessments.
+We demonstrate an example here.
+
+We first try to model the Hagelloch data with an SIR model (see Eq.
+\ref{eq:sir}).
+In our vignette, we show how to fit a stochastic model via maximum likelihood and simulate from the model with those best fit parameters.
+Our function \code{simulate_agents()} generates individual level data according to discrete multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.
+For example
+
+```{r}
+trans_mat <- matrix(c("X0 * (1 - X1 * par1 / N)", "X0 * X1  * par1 / N", "0",
+                  "0", "X1 * (1 - par2)", "par2 * X1",
+                  "0", "0", "X2"), byrow = TRUE, nrow = 3)
+```
+
+```{r cache = TRUE}
+set.seed(2020)
+
+best_params <- c("beta" = .36, "gamma" = .13)
+## This is the SIR representation
+
+rownames(trans_mat) <- c("S", "I", "R")
+init_vals <- c(187, 1, 0)
+par_vals <- c(par1 = best_params[1], par2 = best_params[2])
+max_T <- 55
+n_sims <- 100
+
+agents <- simulate_agents(trans_mat,
+                       init_vals,
+                       par_vals,
+                       max_T,
+                       n_sims,
+                       verbose = FALSE)
+
+```
+```{r}
+agg_model <- agents %>% group_by(sim) %>%
+  agents_to_aggregate(states = c(I, R)) %>%
+  mutate(Type = "Simple SIR")
+```
+The result of our simulation is the object \code{agents} which is a `r nrow(agents)` $\times$ `r ncol(agents)` which details the time of entry into the $S$, $I$, and $R$ states for a given simulation.
+Before we examine the results of this simple SIR model, we will also examine another, more sophisticated SIR model, this time from the package \pkg{EpiModel}.
+Briefly, this model fits a contact network, where the class of the student is a covariate and then simulates a SIR-epidemic on that network.
+
+```{r cache = TRUE, results = 'hide'}
+library(EpiModel)
+## WARNING:  Will take a minute or two
+
+set.seed(42)
+nw <- network.initialize(n = 188, directed = FALSE)
+nw <- set.vertex.attribute(nw, "group", rep(0:2, each = 90, 30, 68))
+formation <- ~edges + nodematch("group") + concurrent
+target.stats <- c(200, 300, 200)
+coef.diss <- dissolution_coefs(dissolution = ~offset(edges),  duration = 5)
+est1 <- netest(nw, formation, target.stats, coef.diss, edapprox = TRUE)
+
+param <- param.net(inf.prob = 0.1, act.rate = 5, rec.rate = 0.1)
+status.vector <- c(rep(0, 90), rep(0, 30), rep(0, 67), 1)
+status.vector <- ifelse(status.vector == 1, "i", "s")
+init <- init.net(status.vector = status.vector)
+control <- control.net(type = "SIR", nsteps = 55,
+                       nsims = 100, epi.by = "group")
+epimodel_sir <- netsim(est1, param, init, control)
+
+```
+
+The output of this model is \code{epimodel_sir}, an object of class `r class(epimodel_sir)`, which contains a plethora of modeling information.
+We provide the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output and transform it into a tidy-style data frame.
+
+```{r}
+fortified_net <- fortify_aggregate(epimodel_sir, 
+                                   states = c("s.num", "i.num", "r.num")) %>%
+  mutate(Type = "EpiModel SIR",
+         sim = as.numeric(gsub("sim", "", sim)))
+```
+
+We can then analyze the results of the two models side by side as time-invariant epidemic curves.
+The results are shown in Figure \ref{fig:hag-simple-sir}, where a 90\% prediction band is estimated from the delta ball method for each of the two models.
+For the Simple SIR model, we see that the data generally covers the data fairly well but clearly misses the second peak of infection.
+We also see that the prediction band is very large, covering up a large area of the ternary plot.
+On the other hand, for the \pkg{EpiModel} model, we see that the prediction band covers the data quite well and takes up less area.
+
+```{r cache = TRUE}
+
+both_models <- bind_rows(agg_model, fortified_net)
+
+g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0),
+         aes(x = X0, y = X1, z = X2,
+              sim_group = sim, fill = Type),
+         alpha = .5,
+         conf_level = .90) 
+```
+
+```{r , fig.cap = "\\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models."}
+g +   geom_path(data = both_models %>% filter(t !=0),
+            aes(x = X0, y = X1, z = X2, group = paste(Type, sim)),
+            alpha = .3, col = "gray40") + 
+    coord_tern() + theme_sir(base_size = 24) +
+  geom_point(data = hagelloch_sir,
+             aes(x = S, y = I, z =R), col = "black") +
+  labs(title = "Simple SIR model",
+       subtitle = "90% Prediction band and original data",
+       x = "S", y = "I", z = "R") +
+  scale_fill_manual(values = c("#006677", "#AA6600")) + 
+  facet_wrap(~Type) +
+  theme(legend.position = "bottom")
+
+```
+However, both models are not a good fit to the filamental path as opposed to the individual points in $(S, I, R)$-space.
+What we mean by that is that both models predict epidemic paths that generally have a single defined peak of infection whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
+In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the EpiModel network model is a better fit than the Simple SIR model, and 2) the fit is only good at the individual point level as opposed to the epidemic path level.
+
+HERE BEN IS GOING TO LOOK AT THE EPIMODEL OUTPUT DIFFERENT PREDICTION BAND THINGS.
+
 \begin{itemize}

### 2020-11-21 jss/EpiCompare-paper/EpiCompare-paper.tex
paper writing

@@
 \Plainauthor{Shannon K. Gallagher, Benjamin Leroy}
-\Plaintitle{A Capitalized Title: Something about a Package foo}
-\Shorttitle{\pkg{foo}: A Capitalized Title}
+\Plaintitle{Time invariant analysis of epidemics with \pkg{EpiCompare}}
+\Shorttitle{\pkg{EpiCompare}}
 
@@
 \cite{Neal2004,britton2011,groendyke2012,becker2016}.
-
-After loading in the libraries for this analysis, we examine . Within
-it, we
 
@@
 time of infection and recovery, respectively of each individual. We can
-then plot the SIR values through a time-invariant lens via a ternary
-plot using regular \pkg{ggplot2} functions or with our custom
-\code{geom}, \code{geom_aggregate}, which takes the original agent data
-as input (\ref{fig:hag-tern-raw}.
+then plot the SIR values through a time-invariant lens using
+\pkg{ggplot2} functions or with our custom \code{geom},
+\code{geom_aggregate}, which takes the original agent data as input
+(Fig. \ref{fig:hag-tern-raw}).
 
@@
 \begin{CodeInput}
-R> sir <- hagelloch_raw %>%
+R> hagelloch_sir <- hagelloch_raw %>%
 +   agents_to_aggregate(states = c(tI, tR),
@@
 R> 
-R> ggplot(sir, aes(x = S, y = I, z = R))+
+R> ggplot(hagelloch_sir, aes(x = S, y = I, z = R))+
 +   coord_tern() +
@@
 +        title = "Time invariant view of Hagelloch measles outbreak") + 
-+   theme_sir()
++   theme_sir(base_size = 24)
 \end{CodeInput}
@@
 Moreover, we can look at the outbreaks of the disease by group within
-\code{agent_to_aggregate()} or \code{geom_aggregate()}, which allows us
-to examine differences among the groups of individuals. For example, we
-show the time invariant outbreak by class level in Figure
+\code{agent_to_aggregate()} or \code{geom_aggregate()}. This allows us
+to examine differences among the different groups of individuals. For
+example, we show the time invariant outbreak by class level in Figure
 \ref{fig:tern-class-data}. Immediately, we see that time invariant
-infection curve is very different for the pre-school group compared to
-the 1st class. In the 1st class, we see about 95\% of the class become
-infected before anyone recovers, which is indicative of a
-super-spreading event. This suspicion is further confirmed in that 26 of
-the 30 have been reportedly infected by the same individual.
+infection curve is different for the pre-school class compared to the
+1st class. In the 1st class, we see about 95\% of the class become
+infected and less than 10\% of them having recovered, which is
+indicative of a super-spreading event. This suspicion is further
+confirmed in that 26 of the 30 1st class students have been reportedly
+infected by the same individual.
 
@@
 
+Along with multiple epidemic states, the function
+\code{agents_to_aggregate} can also be extended to populations with
+vital dynamics (e.g.~birth and death) and examples of this are shown in
+the package vignette. In summary, \code{agents_to_aggregate()} is a
+multi-purpose workhorse that be leveraged to convert individual level
+records into aggregate information that may be more useful for some
+forms of epidemic modeling such as compartment modeling.
+
+Up to this point, we have used \pkg{EpiCompare} in the context of
+observed data. We also want to compare statistical models, and
+\pkg{EpiCompare} aids in that process via a simple but dynamic
+individual-level data generator, conversion tools for popular epidemic
+model packages, and model assessments. We demonstrate an example here.
+
+We first try to model the Hagelloch data with an SIR model (see Eq.
+\ref{eq:sir}). In our vignette, we show how to fit a stochastic model
+via maximum likelihood and simulate from the model with those best fit
+parameters. Our function \code{simulate_agents()} generates individual
+level data according to discrete multinomial draws, which depend on the
+number of individuals in each state at the previous time step and a
+matrix of transition probabilities. For example
+
+\begin{CodeChunk}
+\begin{CodeOutput}
+  S                          I                     R          
+S "X0 * (1 - X1 * par1 / N)" "X0 * X1  * par1 / N" "0"        
+I "0"                        "X1 * (1 - par2)"     "par2 * X1"
+R "0"                        "0"                   "X2"       
+\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> set.seed(2020)
+R> 
+R> best_params <- c("beta" = .36, "gamma" = .13)
+R> ## This is the SIR representation
+R> 
+R> rownames(trans_mat) <- c("S", "I", "R")
+R> init_vals <- c(187, 1, 0)
+R> par_vals <- c(par1 = best_params[1], par2 = best_params[2])
+R> max_T <- 55
+R> n_sims <- 100
+R> 
+R> agents <- simulate_agents(trans_mat,
++                        init_vals,
++                        par_vals,
++                        max_T,
++                        n_sims,
++                        verbose = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> agg_model <- agents %>% group_by(sim) %>%
++   agents_to_aggregate(states = c(I, R)) %>%
++   mutate(Type = "Simple SIR")
+\end{CodeInput}
+\end{CodeChunk}
+
+The result of our simulation is the object \code{agents} which is a
+18800 \(\times\) 5 which details the time of entry into the \(S\),
+\(I\), and \(R\) states for a given simulation. Before we examine the
+results of this simple SIR model, we will also examine another, more
+sophisticated SIR model, this time from the package \pkg{EpiModel}.
+Briefly, this model fits a contact network, where the class of the
+student is a covariate and then simulates a SIR-epidemic on that
+network.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library(EpiModel)
+R> ## WARNING:  Will take a minute or two
+R> 
+R> set.seed(42)
+R> nw <- network.initialize(n = 188, directed = FALSE)
+R> nw <- set.vertex.attribute(nw, "group", rep(0:2, each = 90, 30, 68))
+R> formation <- ~edges + nodematch("group") + concurrent
+R> target.stats <- c(200, 300, 200)
+R> coef.diss <- dissolution_coefs(dissolution = ~offset(edges),  duration = 5)
+R> est1 <- netest(nw, formation, target.stats, coef.diss, edapprox = TRUE)
+R> 
+R> param <- param.net(inf.prob = 0.1, act.rate = 5, rec.rate = 0.1)
+R> status.vector <- c(rep(0, 90), rep(0, 30), rep(0, 67), 1)
+R> status.vector <- ifelse(status.vector == 1, "i", "s")
+R> init <- init.net(status.vector = status.vector)
+R> control <- control.net(type = "SIR", nsteps = 55,
++                        nsims = 100, epi.by = "group")
+R> epimodel_sir <- netsim(est1, param, init, control)
+\end{CodeInput}
+\end{CodeChunk}
+
+The output of this model is \code{epimodel_sir}, an object of class
+netsim, which contains a plethora of modeling information. We provide
+the function \code{fortify_aggregate()}, which can take objects from
+specialized classes of modeling output and transform it into a
+tidy-style data frame.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fortified_net <- fortify_aggregate(epimodel_sir, states = c("s.num", "i.num", "r.num")) %>%
++   mutate(Type = "EpiModel SIR",
++          sim = as.numeric(gsub("sim", "", sim)))
+\end{CodeInput}
+\end{CodeChunk}
+
+We can then analyze the results of the two models side by side as
+time-invariant epidemic curves. The results are shown in Figure
+\ref{fig:hag-simple-sir}, where a 90\% prediction band is estimated from
+the delta ball method for each of the two models. For the Simple SIR
+model, we see that the data generally covers the data fairly well but
+clearly misses the second peak of infection. We also see that the
+prediction band is very large, covering up a large area of the ternary
+plot. On the other hand, for the \pkg{EpiModel} model, we see that the
+prediction band covers the data quite well and takes up less area.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> both_models <- bind_rows(agg_model, fortified_net)
+R> 
+R> 
+R> g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0),
++          aes(x = X0, y = X1, z = X2,
++               sim_group = sim, fill = Type),
++          alpha = .5,
++          conf_level = .90) 
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> g +   geom_path(data = both_models %>% filter(t !=0),
++             aes(x = X0, y = X1, z = X2, group = paste(Type, sim)),
++             alpha = .3, col = "gray40") + 
++     coord_tern() + theme_sir(base_size = 24) +
++   geom_point(data = hagelloch_sir,
++              aes(x = S, y = I, z =R), col = "black") +
++   labs(title = "Simple SIR model",
++        subtitle = "90% Prediction band and original data",
++        x = "S", y = "I", z = "R") +
++   scale_fill_manual(values = c("#006677", "#AA6600")) + 
++   facet_wrap(~Type) +
++   theme(legend.position = "bottom")
+\end{CodeInput}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-17-1}
+
+}
+
+\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-17}
+\end{figure}
+\end{CodeChunk}
+
+However, both models are not a good fit to the filamental path as
+opposed to the individual points in \((S, I, R)\)-space. What we mean by
+that is that both models predict epidemic paths that generally have a
+single defined peak of infection whereas the data certainly looks like
+it has two distinct peaks, likely caused by our assumed super-spreader
+event. In conclusion, \pkg{EpiCompare} makes it clear that, at a glance,
+1) the EpiModel network model is a better fit than the Simple SIR model,
+and 2) the fit is only good at the individual point level as opposed to
+the epidemic path level.
+
+HERE BEN IS GOING TO LOOK AT THE EPIMODEL OUTPUT DIFFERENT PREDICTION
+BAND THINGS.
+
 \begin{itemize}

### 2020-12-10 jss/EpiCompare-paper/EpiCompare-paper.Rmd
first attempt at paper introduction

@@
     url: http://skgallagher.github.io
-  - name: Benjamin Leroy
+  - name: Benjamin LeRoy
     affiliation: |
-      | Dept. of Statistics
+      | Dept. of Statistics & Data Science
       | Carnegie Mellon University
@@
   # If you use tex in the formatted title, also supply version without
-  plain:     "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
+  plain:     "Time invariant analysis of epidemics with EpiCompare"
   # For running headers, if needed
@@
 abstract: >
-  The abstract of the article.
+  We present \pkg{EpiCompare}, an \proglang{R} package that suppliments and enhance current infectious disease modeling analysis pipelines as well as to encourage comparisons across these pipelines.
+A major contribution of this work is the set of novel \textit{time-invariate} tools for model and epidemic comparisons - including time-invariate prediction bands.
+\pkg{EpiCompare} encorporates \proglang{R}'s \textit{tidy} coding style to aid it rapid and easy use.
+This paper provides an overview of both the tools in and intuition behind \pkg{EpiCompare} and a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
 keywords:
@@
 
-\section{Introduction}\label{sec:intro}
+# Introduction {short-title="Intro" #sec:intro}
+
+The recent (and currently on-going) COVID-19 global epidemic has galvanized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area of infectious disease epidemiology.
+Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g.
+flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g.
+historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.
+...][]{}.
+The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.
+Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.
+Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks.
+
+The current epidemic of COVID-19 also highlights that these models are just one piece of the overall analysis pipeline.
+University based resources like John Hopkin's and government numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and visualization can be a important first step in the process (multiple \cite{}?).} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individual / agent level, network structure or just aggregate flows (review paper \cite{}?).
+All placing individuals / proportions of the populations into different states (e.g.
+suspectible, exposed, infected, recovered, etc.).
+With all these models, review and comparison papers in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center helps the individual practitioner decide the correct approach.
+
+At the same time, analysis packages often only address a portion of the analysis pipeline.
+Modeling tools often don't provide easy ways to compare and assess their models on new data.
+Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways.
+To fill these gaps, we present our our \proglang{R} package \pkg{EpiCompare}.
+Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models.
+In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next.
+\pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width = 1\textwidth]{images/pipeline1.png}
+    \caption{An idealized epidemiological data analysis pipeline.}
+    \label{fig:pipeline}
+\end{figure}
+
+This package also emphasizes the value of approach epidemics in a \textit{time-invariant} way, not constrained to things like a specific time scale, or initial date of the first infection.
+Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play.
+Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
+Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches.
+
+This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
+
+SHANNON: I THINK JUST INCLUDING LOTS OF LITERATURE IN INTRO + ETC. INSTEAD OF HAVING A LITERATURE REVIEW IS BETTER.
+
+# Motivation and tools for time-invariant analysis {short-title="Time-invariant" #sec:time-invariant}
+
+SHANNON: I FEEL LIKE WE CAN JUST CLEAN UP TO ORIGINAL SECTION AND PUT IT BACK IN HERE
+
+# Package tool overview {short-title="Package overview" #sec:overview}
+
+## RMD Code formatting info
 
 This is the Section \ref{sec:intro}. This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.
-
-## Code formatting
 
@@
 
-# \proglang{R} code {short-title="R code" #r-code}
+## RMD \proglang{R} code {short-title="R code" #r-code}
 

### 2020-12-10 jss/EpiCompare-paper/EpiCompare-paper.tex
first attempt at paper introduction

@@
-% To remove when JSS.cls is fixed
-% https://github.com/rstudio/rticles/issues/329
-\RequirePackage[2020/02/02]{latexrelease}
 \documentclass[
@@
 National Institute of Allergy\\
-and Infectious Diseases \And Benjamin Leroy\\Dept. of Statistics\\
+and Infectious Diseases \And Benjamin LeRoy\\Dept. of Statistics \& Data Science\\
 Carnegie Mellon University
@@
 
-\Plainauthor{Shannon K. Gallagher, Benjamin Leroy}
-\Plaintitle{Time invariant analysis of epidemics with \pkg{EpiCompare}}
+\Plainauthor{Shannon K. Gallagher, Benjamin LeRoy}
+\Plaintitle{Time invariant analysis of epidemics with EpiCompare}
 \Shorttitle{\pkg{EpiCompare}}
@@
 \Abstract{
-The abstract of the article.
+We present \pkg{EpiCompare}, an \proglang{R} package that suppliments
+and enhance current infectious disease modeling analysis pipelines as
+well as to encourage comparisons across these pipelines. A major
+contribution of this work is the set of novel \textit{time-invariate}
+tools for model and epidemic comparisons - including time-invariate
+prediction bands. \pkg{EpiCompare} encorporates \proglang{R}'s
+\textit{tidy} coding style to aid it rapid and easy use. This paper
+provides an overview of both the tools in and intuition behind
+\pkg{EpiCompare} and a thorough demonstrating of the tools through a
+detailed example of a full data analysis pipeline.
 }
@@
   URL: \url{http://skgallagher.github.io}\\~\\
-      Benjamin Leroy\\
-    Dept. of Statistics\\
+      Benjamin LeRoy\\
+    Dept. of Statistics \& Data Science\\
   Carnegie Mellon University\\
@@
   }
-
-% Pandoc citation processing
 
@@
 
-\section{Introduction}\label{sec:intro}
+\section[Intro]{Introduction}\label{sec:intro}
+
+The recent (and currently on-going) COVID-19 global epidemic has
+galvanized public interest in understanding more about infectious
+disease modeling highlighted the usefulness of research in the area of
+infectious disease epidemiology. Infectious disease models typically
+attempt to do one or more of the following: 1) predict the spread of
+current and future epidemics
+\citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
+current epidemics to increase scientific knowledge
+\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
+project epidemic scenarios under pre-specified parameters
+\citep[e.g. ...][]{}. The COVID-19 epidemic highlights how all three
+goals are important both separately and taken as a whole. Infectious
+diseases inflict enormous burdens on the world: millions of lives lost
+and trillions of dollars spent yearly. Correctly analyzing and
+addressing these issues aids in prevention and mitigation of future
+outbreaks.
+
+The current epidemic of COVID-19 also highlights that these models are
+just one piece of the overall analysis pipeline. University based
+resources like John Hopkin's and government numerical dashboards (across
+all levels of government) during the COVID-19 epidemic remind us that
+descriptive statistics and visualization can be a important first step
+in the process (multiple \cite{}?).\} Still, rightly so, a large amount
+of theoretical work goes into modeling epidemics, with different models
+focusing at the individual / agent level, network structure or just
+aggregate flows (review paper \cite{}?). All placing individuals /
+proportions of the populations into different states (e.g.~suspectible,
+exposed, infected, recovered, etc.). With all these models, review and
+comparison papers in the literature and through MIDAS (Models of
+Infectious Disease Agent Study) Control Center helps the individual
+practitioner decide the correct approach.
+
+At the same time, analysis packages often only address a portion of the
+analysis pipeline. Modeling tools often don't provide easy ways to
+compare and assess their models on new data. Moreover, exploring and
+modeling epidemics require transforming and \textit{tidying} data in
+different ways. To fill these gaps, we present our our \proglang{R}
+package \pkg{EpiCompare}. Our package's primary focus is to aid and
+advance research in the area of comparison and assessment of epidemic \&
+epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the
+data analysis pipeline of infectious diseases as 1) data pre-processing,
+2) exploratory data analysis (EDA), 3) modeling and simulating, 4)
+post-processing, and 5) comparison and assessment; where each previous
+part of the pipeline influences the next. \pkg{EpiCompare} provides
+tools to aids practitioners in all areas of this pipeline.
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width = 1\textwidth]{images/pipeline1.png}
+    \caption{An idealized epidemiological data analysis pipeline.}
+    \label{fig:pipeline}
+\end{figure}
+
+This package also emphasizes the value of approach epidemics in a
+\textit{time-invariant} way, not constrained to things like a specific
+time scale, or initial date of the first infection. Although epidemics
+are by definition a process that evolves over time, epidemics, including
+the recent COVID-19, often need to be compared in a time-invariant way
+to understand the processes at play. Additionally, many tools to examine
+the quantity of the population in each state along the infection process
+(for example: quantity of suspectible vs infected vs recovered
+individuals) don't always as intelligently capture the natural
+connections between the proportion of individuals in these states. Tools
+in \pkg{EpiCompare} attempt to give the user the ability to extend their
+toolkit to evaluate epidemics to also include time-invariant approaches.
+
+This paper is broken up into the following sections, section
+\ref{sec:time-invariant} motivates and showcases tools of time-invariant
+analysis, section \ref{sec:overview} presents an outline of how
+\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
+section \ref{sec:tour} provides a thorough demonstrating of the tools
+through a detailed example of a full data analysis pipeline.
+
+SHANNON: I THINK JUST INCLUDING LOTS OF LITERATURE IN INTRO + ETC.
+INSTEAD OF HAVING A LITERATURE REVIEW IS BETTER.
+
+\section[Time-invariant]{Motivation and tools for time-invariant
+analysis}\label{sec:time-invariant}
+
+SHANNON: I FEEL LIKE WE CAN JUST CLEAN UP TO ORIGINAL SECTION AND PUT IT
+BACK IN HERE
+
+\section[Package overview]{Package tool overview}\label{sec:overview}
+
+\hypertarget{rmd-code-formatting-info}{%
+\subsection{RMD Code formatting info}\label{rmd-code-formatting-info}}
 
@@
 the basic LaTeX that you need to know to create a JSS article.
-
-\hypertarget{code-formatting}{%
-\subsection{Code formatting}\label{code-formatting}}
 
@@
 
-\section[R code]{\proglang{R} code}\label{r-code}
+\subsection[R code]{RMD \proglang{R} code}\label{r-code}
 
@@
 \begin{CodeChunk}
-\begin{CodeOutput}
-  S                          I                     R          
-S "X0 * (1 - X1 * par1 / N)" "X0 * X1  * par1 / N" "0"        
-I "0"                        "X1 * (1 - par2)"     "par2 * X1"
-R "0"                        "0"                   "X2"       
-\end{CodeOutput}
+\begin{CodeInput}
+R> trans_mat <- matrix(c("X0 * (1 - X1 * par1 / N)", "X0 * X1  * par1 / N", "0",
++                   "0", "X1 * (1 - par2)", "par2 * X1",
++                   "0", "0", "X2"), byrow = TRUE, nrow = 3)
+\end{CodeInput}
 \end{CodeChunk}
@@
 \begin{CodeInput}
-R> fortified_net <- fortify_aggregate(epimodel_sir, states = c("s.num", "i.num", "r.num")) %>%
+R> fortified_net <- fortify_aggregate(epimodel_sir, 
++                                    states = c("s.num", "i.num", "r.num")) %>%
 +   mutate(Type = "EpiModel SIR",
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-17-1}
+{\centering \includegraphics{Figs/unnamed-chunk-16-1}
 
@@
 
-\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-17}
+\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-16}
 \end{figure}

### 2020-12-14 jss/EpiCompare-paper/EpiCompare-paper.Rmd
comments on draft

@@
   - \usepackage{float}
+  - \usepackage{xcolor}
 ---
@@
 
+\newcommand{\shannon}[1]{\textcolor{orange}{#1}}
+\newcommand{\ben}[1]{\textcolor{violet}{#1}}
+
+\newtheorem{theorem}{Theorem}
+
 # Introduction {short-title="Intro" #sec:intro}
 
-The recent (and currently on-going) COVID-19 global epidemic has galvanized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area of infectious disease epidemiology.
+The recent (and currently on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology.
 Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g.
@@
 ...][]{}.
-The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.
+The COVID-19 pandemic highlights how all three goals are important both separately and taken as a whole.
 Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.
 Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks.
+\shannon{really like this paragraph}
 
-The current epidemic of COVID-19 also highlights that these models are just one piece of the overall analysis pipeline.
-University based resources like John Hopkin's and government numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and visualization can be a important first step in the process (multiple \cite{}?).} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individual / agent level, network structure or just aggregate flows (review paper \cite{}?).
+The current epidemic of COVID-19 also highlights that infectious disease  models are only one piece of the overall analysis pipeline.
+University based resources like John Hopkin's and government numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and visualization can be a important first step in the process (multiple \cite{}?).} \shannon{add comment about NYT or something} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individual / agent level, network structure or just aggregate flows (review paper \cite{}?).
 All placing individuals / proportions of the populations into different states (e.g.
@@
 With all these models, review and comparison papers in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center helps the individual practitioner decide the correct approach.
+\shannon{along with expertise from healthcare professionals...}
 
@@
 Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches.
+The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
+\shannon{flesh out time invariance introduction to better explain to non-experts what we mean.
+perhaps mention R0}
 
@@
 # Motivation and tools for time-invariant analysis {short-title="Time-invariant" #sec:time-invariant}
-
-SHANNON: I FEEL LIKE WE CAN JUST CLEAN UP TO ORIGINAL SECTION AND PUT IT BACK IN HERE
 
@@
 The Hagelloch data includes a rich set of features including household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.
-A subset of the data is shown in Table \ref{tab:hags-people}   Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.
+A subset of the data is shown in Table \ref{tab:hags-people}.
+Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.
 
@@
 
-```{r}
+```{r echo = FALSE, results = 'hide'}
 with(hagelloch_raw, sum(PRO > ERU, na.rm = TRUE))
@@
 For example, the Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the time of infection and recovery, respectively of each individual.
-We can then plot the SIR values through a time-invariant lens  using \pkg{ggplot2} functions or with our custom \code{geom}, \code{geom_aggregate}, which takes the original agent data as input  (Fig.
-\ref{fig:hag-tern-raw}).
+We can then plot the SIR values through a time-invariant lens using \pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
+\ref{fig:hag-tern-raw}) or with our custom \code{geom}, \code{geom_aggregate}, which takes the raw agent data as input.
 
@@
 birth and death) and examples of this are shown in the package vignette.
-In summary, \code{agents_to_aggregate()} is a multi-purpose workhorse that be leveraged to convert individual level records into aggregate information that may be more useful for some forms of epidemic modeling such as compartment modeling.
+In summary, \code{agents_to_aggregate()} is a multi-purpose workhorse that may be leveraged to convert individual level records into aggregate information that may be more useful for some forms of epidemic modeling such as compartment modeling.
 
@@
 Our function \code{simulate_agents()} generates individual level data according to discrete multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.
-For example
+For example, the below code generates 100 simulations of an outbreak of a diseease with one initial infector in a population of $n= 188$ individuals.
 
@@
 Before we examine the results of this simple SIR model, we will also examine another, more sophisticated SIR model, this time from the package \pkg{EpiModel}.
-Briefly, this model fits a contact network, where the class of the student is a covariate and then simulates a SIR-epidemic on that network.
+Briefly, this model first fits a contact network to the set of indivduals, where the class of the student is a covariate.
+The model then simulates a SIR-epidemic on that network.
 

### 2020-12-14 jss/EpiCompare-paper/EpiCompare-paper.tex

@@
+% To remove when JSS.cls is fixed
+% https://github.com/rstudio/rticles/issues/329
+\RequirePackage[2020/02/02]{latexrelease}
 \documentclass[
@@
 National Institute of Allergy\\
-and Infectious Diseases \And Benjamin LeRoy\\Dept. of Statistics \& Data Science\\
+and Infectious Diseases \And Benjamin LeRoy\\Dept. of Statistics \& Data
+Science\\
 Carnegie Mellon University
@@
 
+% Pandoc citation processing
+
 % Pandoc header
@@
 \usepackage{float}
+\usepackage{xcolor}
 
@@
 
+\newcommand{\shannon}[1]{\textcolor{orange}{#1}}
+\newcommand{\ben}[1]{\textcolor{violet}{#1}}
+
+\newtheorem{theorem}{Theorem}
+
 \section[Intro]{Introduction}\label{sec:intro}
 
-The recent (and currently on-going) COVID-19 global epidemic has
+The recent (and currently on-going) COVID-19 global pandemic has
 galvanized public interest in understanding more about infectious
-disease modeling highlighted the usefulness of research in the area of
-infectious disease epidemiology. Infectious disease models typically
-attempt to do one or more of the following: 1) predict the spread of
-current and future epidemics
+disease modeling and has highlighted the usefulness of research in the
+area of infectious disease epidemiology. Infectious disease models
+typically attempt to do one or more of the following: 1) predict the
+spread of current and future epidemics
 \citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
@@
 project epidemic scenarios under pre-specified parameters
-\citep[e.g. ...][]{}. The COVID-19 epidemic highlights how all three
+\citep[e.g. ...][]{}. The COVID-19 pandemic highlights how all three
 goals are important both separately and taken as a whole. Infectious
@@
 addressing these issues aids in prevention and mitigation of future
-outbreaks.
+outbreaks. \textcolor{orange}{really like this paragraph}
 
-The current epidemic of COVID-19 also highlights that these models are
-just one piece of the overall analysis pipeline. University based
-resources like John Hopkin's and government numerical dashboards (across
-all levels of government) during the COVID-19 epidemic remind us that
-descriptive statistics and visualization can be a important first step
-in the process (multiple \cite{}?).\} Still, rightly so, a large amount
-of theoretical work goes into modeling epidemics, with different models
-focusing at the individual / agent level, network structure or just
-aggregate flows (review paper \cite{}?). All placing individuals /
+The current epidemic of COVID-19 also highlights that infectious disease
+models are only one piece of the overall analysis pipeline. University
+based resources like John Hopkin's and government numerical dashboards
+(across all levels of government) during the COVID-19 epidemic remind us
+that descriptive statistics and visualization can be a important first
+step in the process (multiple \cite{}?).\} Still, rightly so, a large
+amount of theoretical work goes into modeling epidemics, with different
+models focusing at the individual / agent level, network structure or
+just aggregate flows (review paper \cite{}?). All placing individuals /
 proportions of the populations into different states (e.g.~suspectible,
@@
 practitioner decide the correct approach.
+\textcolor{orange}{along with expertise from healthcare professionals...}
 
@@
 toolkit to evaluate epidemics to also include time-invariant approaches.
+The goal of \pkg{EpiCompare} is not to supplant existing infectious
+disease modeling tools and software but, rather, is a concerted effort
+to create standard and fair comparisons among models developed for
+disease outbreaks and outbreak data.
 
@@
 
-SHANNON: I FEEL LIKE WE CAN JUST CLEAN UP TO ORIGINAL SECTION AND PUT IT
-BACK IN HERE
+\hypertarget{visualize}{%
+\subsection{Visualize}\label{visualize}}
+
+One issue that the COVID-19 outbreak has brought to the forefront is the
+importance of figures and graphs to inform, support, and even persuade
+lay people of technical and scientific results \cite{}. Quantities and
+trends shown in these visualizations include exponential growth curves,
+S-curves, peak day and percent of infectious indviduals, ICU capacity,
+and deaths over time and uncertainty of such estimates.
+
+\pkg{EpiCompare} supports traditional visualizations (e.g.~\# in
+condition vs.~time) and also introduces a novel visualization of
+epidemics via time free analysis. In particular, we use ternary plots
+that show Susceptible-Infectious-Recovered data/models (SIR). The
+advantages of these ternary plots are 1) a standard way to compare
+epidemics with different outbreak times and time scales, 2) a novel
+interpretation of \(R_0\), the initial reproduction number under some
+circumstances, and 3) clear visualization of joint SIR uncertainty via
+conformal bands.
+
+Ternary plots are sometimes used in chemistry \citep{} but only rarely
+seen in the field of infectious disease epidemiology (to our knowledge,
+only seen in \cite{}). Ternary plots illustrate the relationship of
+constrained 3D data, namely of points
+\((a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]\) with
+\(a_i+ b_i + c_i =1\) and as such, are situational. The SIR framework,
+composed of data and statistical models where individuals pass from
+Susceptible to Infectious to Recovered states is a common, if not the
+most common, modeling framework in infectious disease epidemiology (seen
+in examples of recent published works in \cite{}), and also happens to
+be of the situational format required for ternary plots where
+\((S(t), I(t), R(t))\) are the number of susceptible, infectious, and
+recovered individuals in each state at time \(t\) and
+\(S(t) + I(t) + R(t) = N(t)\) where \(N(t)\) is the population size at
+time \(t\).
+
+Ternary plots are time-free in the sense that the temporal scale does
+not explicitly apear in the visualization and as a consequence, allow
+for direct comparisons of outbreaks on different time scales (e.g.~days
+vs.~years) or of SIR data of the same disease in different areas (our
+first case study in Sec. \ref{sec:ex1} focuses on this).
+
+Ternary plots also, under certain conditions, have a special
+relationship with \(R_0\), the initial reproduction number, which has
+been called the ``most important quantity in epidemiology'\,' \citep{}.
+The reproduction is itself a time-free concept and is the average number
+of secondary infections primary infection will generate when the primary
+infection is added to a completely susceptible population. To see the
+importance of \(R_0\), one need only read the newspaper \citep{} or look
+at the length of the table of estimated \(R_0\) quantities for COVID-19
+\citep{}.
+
+In regards to traditional \(X\) vs.~time plots, \(R_0\) is difficult to
+visualize. For example, as seen in Fig. \ref{fig:r0-trad}, it is
+possible that two models have the same value of \(R_0\) but there is no
+way of knowing that from looking at the graph.
+
+It is possible to identify, at a glance, which epidemic paths have the
+same value of \(R_0\) via ternary plot visualizations, under some
+circumstances. To show, this we first need to introduce the classic,
+deterministic Kermack and McKendrick SIR model where the initial number
+in each state \((S(0), I(0), R(0))\) are known and \(N\) the population
+size is constant. The movement of individuals among the states is given
+by the following differential equations (Eq. \eqref{eq:sir-ode}) where
+\(\beta\) is the average infection rate and \(\gamma\) is the average
+recovery rate, \begin{align}\label{eq:sir-ode}
+      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
+      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
+      R^\prime(t) &= \gamma I(t) \nonumber.
+  \end{align}
+
+If we have two SIR models that follow the Kermack and McKendrick SIR
+model, have the same percentage of individuals in the initial states,
+and the same value of \(R_0\) then the two models will create exactly
+overlapping paths in a ternary plot.
+
+More formally, let two Kermack and McKendrick (see Eq.
+\eqref{eq:sir-ode}) SIR models be denoted \((S_1(t), I_1(t), R_1(t))\)
+and \((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume
+both models have initial values \((S(0), I(0), R(0))\). Let
+\(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
+\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
+rate, respectively, for SIR model \(i=1, 2\). Equivalently,
+\(\beta_2 = a \beta_1\) if and only if \(\gamma_2 = a \gamma_1\) for
+some \(a > 0\).
+
+\begin{theorem}\label{thm:sir-scale}
+Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+\end{theorem}
+
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
+result from \cite{} and is shown in detail in Appendix \ref{app:proof}.
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR
+models that have the same initial percent of individuals in each state
+and \(R_0\) then for every point on the epidemic path of the first SIR
+model is also a point on the epidemic path of the second SIR model.
 
@@
 rash, and even the alleged infector. A subset of the data is shown in
-Table \ref{tab:hags-people} Because of these rich features, this data
+Table \ref{tab:hags-people}. Because of these rich features, this data
 set has been an ideal testing ground methodology in infectious disease
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> with(hagelloch_raw, sum(PRO > ERU, na.rm = TRUE))
-\end{CodeInput}
-\begin{CodeOutput}
-[1] 0
-\end{CodeOutput}
-\end{CodeChunk}
-
 The real power of \code{agents_to_aggregate()} lies in its ability to
@@
 then plot the SIR values through a time-invariant lens using
-\pkg{ggplot2} functions or with our custom \code{geom},
-\code{geom_aggregate}, which takes the original agent data as input
-(Fig. \ref{fig:hag-tern-raw}).
+\pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
+\ref{fig:hag-tern-raw}) or with our custom \code{geom},
+\code{geom_aggregate}, which takes the raw agent data as input.
 
@@
 the package vignette. In summary, \code{agents_to_aggregate()} is a
-multi-purpose workhorse that be leveraged to convert individual level
-records into aggregate information that may be more useful for some
-forms of epidemic modeling such as compartment modeling.
+multi-purpose workhorse that may be leveraged to convert individual
+level records into aggregate information that may be more useful for
+some forms of epidemic modeling such as compartment modeling.
 
@@
 number of individuals in each state at the previous time step and a
-matrix of transition probabilities. For example
+matrix of transition probabilities. For example, the below code
+generates 100 simulations of an outbreak of a diseease with one initial
+infector in a population of \(n= 188\) individuals.
 
@@
 sophisticated SIR model, this time from the package \pkg{EpiModel}.
-Briefly, this model fits a contact network, where the class of the
-student is a covariate and then simulates a SIR-epidemic on that
-network.
+Briefly, this model first fits a contact network to the set of
+indivduals, where the class of the student is a covariate. The model
+then simulates a SIR-epidemic on that network.
 

### 2020-12-14 README.Rmd
tidy_dist_mat and calculation of new observation's depth / psuedo-densities relative to another set of observations (part 3): check passing

@@
 Useful Rstudio shortcuts for `Roxygen2`: (a) create `Roxygen2` comments template with `option` + `command` + `shift` + `R` (b) In `Roxygen2` comments do `control` + `shift` + `/` to format relative to 80 char limit.
-7.[stack overflow](https://stackoverflow.com/questions/11285496/r-cmd-check-warning-functions-methods-with-usage-in-documentation-object-bu) post on how to pass `check` for `rownames<-.tidy_dist_mat`.
+7.
+[stack overflow](https://stackoverflow.com/questions/11285496/r-cmd-check-warning-functions-methods-with-usage-in-documentation-object-bu) post on how to pass `check` for `rownames<-.tidy_dist_mat`.
 

### 2020-12-14 R/distance-psuedo-density.R

@@ -3,6 +3,12 @@
 #' DESCRIBE MORE
 #'
 #' @param x a n x n square positive symmetric matrix or a \code{tidy_dist_mat}
+#' @param x_new a n_new x n matrix or tidy_dist_mat where the rows correspond to
+#' new observations, the columns correspond to points in \code{x} (if \code{x} 
+#' and \code{x_new} are matrices then they need to be corrected ordered). If 
+#' this value is not \code{NULL} (default is \code{NULL}) then the 
+#' psuedo-density vector will be calculated for these observations relative to 
+#' observations defined with \code{x} and \code{x_new}'s columns.
 #' @param sigma scaling parameter. Can either by a standard numerical value or a
 #' string as a percentage (e.g. "20\%")
 #' @param df_out indicates if one should return a data.frame our a vector,

### 2020-12-14 R/tidy-dist-mat.R

@@ -269,6 +269,7 @@ format.tidy_dist_mat <- function(x, ...){
 #' @param x tidy_dist_mat object
 #' @param ... (like \code{digits = 6}) number of significant digits to display
 #'   (uses \code{signif})
+#' @param n maximum number of rows or columns of x to print
 #'
 #' @export
 #'
@@ -299,11 +300,13 @@ print.tidy_dist_mat <- function(x, ..., n = NULL){
 #' convert \code{tidy_dist_mat} to \code{matrix}
 #'
 #' @param x \code{tidy_dist_mat} object
+#' @param ... additional arguments to be passed to or from methods. (currently
+#' doesn't do anything for this.)
 #'
 #' @return matrix representation (without column information)
 #' @export
 #'
-as.matrix.tidy_dist_mat <- function(x){
+as.matrix.tidy_dist_mat <- function(x, ...){
   inner <- unclass(x)
   attr(inner, "rownames_df") <- NULL
   attr(inner, "colnames_df") <- NULL
@@ -361,12 +364,16 @@ process_df_index <- function(x, index, margin = 1){
     j <- process_df_index(x, j, margin = 2)
   }
   
-  
-  
-  
   x_mat <- as.matrix(x)[i,j,drop = F]
-  rownames_new <- rownames(x) %>% tibble::tibble() %>% .[i,] %>% data.frame
-  colnames_new <- colnames(x) %>% tibble::tibble() %>% .[j,] %>% data.frame
+  
+  perserve_rownames_class <- class(rownames(x))
+  perserve_colnames_class <- class(colnames(x))
+  
+  rownames_new <- tibble::tibble(rownames(x))[i,]
+  class(rownames_new) <- perserve_rownames_class
+  colnames_new <- tibble::tibble(colnames(x))[j,] 
+  class(colnames_new) <- perserve_colnames_class
+  
   
   tidy_dist_mat(x_mat, 
                 rownames_df = rownames_new,
@@ -419,7 +426,7 @@ is.not_df <- function(x){
 
 #' find the index of a \code{tidy_dist_mat} relative to a data.frame 
 #'
-#' probably more of an internal function
+#' probably more of an internal function...
 #'
 #' @param x \code{tidy_dist_mat} object
 #' @param index data.frame index associated with \code{x}'s colnames or
@@ -428,7 +435,7 @@ is.not_df <- function(x){
 #' @param margin scalar representing if \code{index} is related to rows or 
[…]

### 2020-12-15 jss/EpiCompare-paper/EpiCompare-paper.Rmd
tidy_dist_mat and calculation of new observation's depth / psuedo-densities relative to another set of observations (part 4): pkgdown creation

@@
 # Motivation and tools for time-invariant analysis {short-title="Time-invariant" #sec:time-invariant}
+
+Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play.
+Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
+Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches and in this section we present benefits of the time-invariant analysis, with 1) motivation for time-invariant analysis through $R_0$ (the initial reproduction number), 2) time-invariant visualization tools for 3-state models (e.g.
+SIR models), and 3) potential for similar analysis for models with more states (e.g.
+SEIR models).
+\ben{Will need to rewrite this section "introduction".}
+
+## Motivation through $R_0$ {short-title="r0" #r0}
+
+$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[][]{Gallagher2020}.
+To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}.
+$R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, and is commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
+The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILLIN}).
+
+In regards to traditional $X$ vs.
+time plots, $R_0$ is difficult to visualize.
+For example, in a SIR simulation with two models have the same value of $R_0$, one can see in Fig.
+\ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph (this figure will be commented on more below).
+
+\begin{figure}[!ht]
+%% Code to reproduce is in inst/paper_figs/different-time-scales.R
+    \centering
+    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
+    \caption{Bivariate view of \# in each state vs.
+time.
+Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\ben{We probably need to define $\beta/\gamma$.}
+\ben{NEW: recommendation - make this plot actually have 3 examples (the current one), one where we have an affine transformation of the time scale, and one with the halleolof data.} \ben{Also these figures should be reproducible.}}
+    \label{fig:different-scales-standard}
+\end{figure}
+
+## Ternary and time-invariant visualizations for $R_0$ {short-title="ternary visualizations" #sec:ternary}
+
+It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig.
+\ref{fig:different-scales-tern}  have the same value of $R_0$ via ``time-invariant'' visualizations, under some circumstances.
+Specifically, for SIR models we propose using ternary to examine an epidemic's trajectory in a more time-invariant manner.
+
+Ternary plots are sometimes used in chemistry \citep[][]{Gillespie1976} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
+Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.
+The SIR framework happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.
+
+Ternary plots are time-invariant in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g.
+days vs.
+years) or of SIR data of the same disease in different areas (our first case study in Sec.
+\ref{sec:ex1} focuses on this).
+
+To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick \cite{CITETHIS?} SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
+The movement of individuals among the states is given by the following differential equations (Eq.
+\eqref{eq:sir-ode}) where $\beta$ is the average infection rate and $\gamma$ is the average recovery rate,
+\begin{align}\label{eq:sir-ode}
+      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
+      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
+      R^\prime(t) &= \gamma I(t) \nonumber.
+  \end{align}
+
+\ben{SUGGESTION: define $R_0$ mathematically here (though referred to in figure 2 - which still needs to be defined there).}
+
+If we have two SIR models that follow the Kermack and McKendrick SIR model, have the same percentage of individuals in the initial states, and the same value of $R_0$ then the two models will create exactly overlapping paths in a ternary plot.
+
+More formally, let two Kermack and McKendrick (see Eq.
+\eqref{eq:sir-ode}) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
+Assume both models have initial values $(S(0), I(0), R(0))$.
+Let $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
+Equivalently, $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$ for some $a > 0$.
+
+\begin{theorem}\label{thm:sir-scale}
+Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+\end{theorem}
+
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Appendix \ref{app:proof}.
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.
+Taking the sample simulations from Fig.
+\ref{fig:different-scales-standard}, Fig.
+\ref{fig:different-scales-tern} presents these models in a ternary plot.
+
+\begin{figure}[!ht]
+%% Code to reproduce is in inst/paper_figs/different-time-scales.R
+    \centering
+    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
+    \caption{Ternary view of \# in each state.
+Both this Fig.
+and Fig.
+\ref{fig:different-scales-standard} display the same two sets of data.
+Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
+\ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view.
+\ben{I might suggest using opacity to show time scale here?} \ben{Make this reproducible too?}}
+    \label{fig:different-scales-tern}
+\end{figure}
+
+In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-invariant space, which is presented in more details in Section \ref{sec:time-invariant-in-simplex}.
+
+## Simplexes beyond ternary plots and time-invariant comparison tools {short-title="Simplexes beyond ternary plots" #sec:time-invariant-in-simplex}
+
+Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:ternary} to higher dimensions (i.e.
+models with more states).
+The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling).
+This same scaling for larger models (i.e.
+with more states) can be done onto different simplexes.
+In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a one-dimension less space through the simplexical structure of the data.
+
+In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes.
+Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics.
+The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit.
+In \pkg{EpiCompare} we provide ways to create prediction regions for a true epidemic under the model assumptions using these simulations.
+These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects.
+We extend off of papers like \citet{Dalmasso2019a} to create these bands.
+These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets.
+We recommend using the Hausdorff distance to compare such sets as it captures how much bigger the sets would have to expand to cover each other, and is defined mathematically as
+\[
+d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
+\]
 

### 2020-12-15 jss/EpiCompare-paper/EpiCompare-paper.tex
tidy_dist_mat and calculation of new observation's depth / psuedo-densities relative to another set of observations (part 4): pkgdown creation

@@
 National Institute of Allergy\\
-and Infectious Diseases \And Benjamin LeRoy\\Dept. of Statistics \& Data
-Science\\
+and Infectious Diseases \And Benjamin LeRoy\\Dept. of Statistics \& Data Science\\
 Carnegie Mellon University
@@
 
-% Pandoc citation processing
-
 % Pandoc header
@@
 that descriptive statistics and visualization can be a important first
-step in the process (multiple \cite{}?).\} Still, rightly so, a large
-amount of theoretical work goes into modeling epidemics, with different
-models focusing at the individual / agent level, network structure or
-just aggregate flows (review paper \cite{}?). All placing individuals /
-proportions of the populations into different states (e.g.~suspectible,
-exposed, infected, recovered, etc.). With all these models, review and
-comparison papers in the literature and through MIDAS (Models of
-Infectious Disease Agent Study) Control Center helps the individual
-practitioner decide the correct approach.
+step in the process (multiple \cite{}?).\}
+\textcolor{orange}{add comment about NYT or something} Still, rightly
+so, a large amount of theoretical work goes into modeling epidemics,
+with different models focusing at the individual / agent level, network
+structure or just aggregate flows (review paper \cite{}?). All placing
+individuals / proportions of the populations into different states
+(e.g.~suspectible, exposed, infected, recovered, etc.). With all these
+models, review and comparison papers in the literature and through MIDAS
+(Models of Infectious Disease Agent Study) Control Center helps the
+individual practitioner decide the correct approach.
 \textcolor{orange}{along with expertise from healthcare professionals...}
@@
 disease outbreaks and outbreak data.
+\textcolor{orange}{flesh out time invariance introduction to better explain to non-experts what we mean.  perhaps mention R0}
 
@@
 
-\hypertarget{visualize}{%
-\subsection{Visualize}\label{visualize}}
+Although epidemics are by definition a process that evolves over time,
+epidemics, including the recent COVID-19, often need to be compared in a
+time-invariant way to understand the processes at play. Additionally,
+many tools to examine the quantity of the population in each state along
+the infection process (for example: quantity of suspectible vs infected
+vs recovered individuals) don't always as intelligently capture the
+natural connections between the proportion of individuals in these
+states. Tools in \pkg{EpiCompare} attempt to give the user the ability
+to extend their toolkit to evaluate epidemics to also include
+time-invariant approaches and in this section we present benefits of the
+time-invariant analysis, with 1) motivation for time-invariant analysis
+through \(R_0\) (the initial reproduction number), 2) time-invariant
+visualization tools for 3-state models (e.g.~SIR models), and 3)
+potential for similar analysis for models with more states (e.g.~SEIR
+models).
+\textcolor{violet}{Will need to rewrite this section "introduction".}
 
-One issue that the COVID-19 outbreak has brought to the forefront is the
-importance of figures and graphs to inform, support, and even persuade
-lay people of technical and scientific results \cite{}. Quantities and
-trends shown in these visualizations include exponential growth curves,
-S-curves, peak day and percent of infectious indviduals, ICU capacity,
-and deaths over time and uncertainty of such estimates.
+\subsection[r0]{Motivation through \(R_0\)}\label{r0}
 
-\pkg{EpiCompare} supports traditional visualizations (e.g.~\# in
-condition vs.~time) and also introduces a novel visualization of
-epidemics via time free analysis. In particular, we use ternary plots
-that show Susceptible-Infectious-Recovered data/models (SIR). The
-advantages of these ternary plots are 1) a standard way to compare
-epidemics with different outbreak times and time scales, 2) a novel
-interpretation of \(R_0\), the initial reproduction number under some
-circumstances, and 3) clear visualization of joint SIR uncertainty via
-conformal bands.
-
-Ternary plots are sometimes used in chemistry \citep{} but only rarely
-seen in the field of infectious disease epidemiology (to our knowledge,
-only seen in \cite{}). Ternary plots illustrate the relationship of
-constrained 3D data, namely of points
-\((a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]\) with
-\(a_i+ b_i + c_i =1\) and as such, are situational. The SIR framework,
+\(R_0\), the initial reproduction number, has been called the ``most
+important quantity in epdiemiology'' \citep[][]{Gallagher2020}. To see
+the importance of \(R_0\), one need only read the newspaper
+\citep{Fisher2020} or look at the length of the table of estimated
+\(R_0\) quantities for COVID-19 \citep{Aronson2020}. \(R_0\) is also,
+maybe, the most famous \textit{time-invariant} numerical summary of an
+epidemic, and is commonly associated with the
+Susceptible-Infectious-Recovered (SIR) data/models. The SIR framework,
 composed of data and statistical models where individuals pass from
@@
 most common, modeling framework in infectious disease epidemiology (seen
-in examples of recent published works in \cite{}), and also happens to
-be of the situational format required for ternary plots where
+in examples of recent published works in \cite{FILLIN}).
+
+In regards to traditional \(X\) vs.~time plots, \(R_0\) is difficult to
+visualize. For example, in a SIR simulation with two models have the
+same value of \(R_0\), one can see in Fig.
+\ref{fig:different-scales-standard} that there is no way of knowing that
+from looking at the graph (this figure will be commented on more below).
+
+\begin{figure}[!ht]
+%% Code to reproduce is in inst/paper_figs/different-time-scales.R
+    \centering
+    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
+    \caption{Bivariate view of \# in each state vs.
+time.
+Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\textcolor{violet}{We probably need to define $\beta/\gamma$.}
+\textcolor{violet}{NEW: recommendation - make this plot actually have 3 examples (the current one), one where we have an affine transformation of the time scale, and one with the halleolof data.} \textcolor{violet}{Also these figures should be reproducible.}}
+    \label{fig:different-scales-standard}
+\end{figure}
+
+\subsection[ternary visualizations]{Ternary and time-invariant
+visualizations for \(R_0\)}\label{sec:ternary}
+
+It is possible to identify, at a glance, which SIR epidemic paths, like
+those in Fig. \ref{fig:different-scales-tern} have the same value of
+\(R_0\) via ``time-invariant'' visualizations, under some circumstances.
+Specifically, for SIR models we propose using ternary to examine an
+epidemic's trajectory in a more time-invariant manner.
+
+Ternary plots are sometimes used in chemistry \citep[][]{Gillespie1976}
+but only rarely seen in the field of infectious disease epidemiology (to
+our knowledge, only seen in \cite{}). Ternary plots illustrate the
+relationship of constrained 3D data, namely of points
+\((a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]\) with
+\(a_i+ b_i + c_i =1\) and as such, are situational. The SIR framework
+happens to be of the situational format required for ternary plots where
 \((S(t), I(t), R(t))\) are the number of susceptible, infectious, and
@@
 
-Ternary plots are time-free in the sense that the temporal scale does
-not explicitly apear in the visualization and as a consequence, allow
-for direct comparisons of outbreaks on different time scales (e.g.~days
-vs.~years) or of SIR data of the same disease in different areas (our
-first case study in Sec. \ref{sec:ex1} focuses on this).
+Ternary plots are time-invariant in the sense that the temporal scale
+does not explicitly apear in the visualization and as a consequence,
+allow for direct comparisons of outbreaks on different time scales
+(e.g.~days vs.~years) or of SIR data of the same disease in different
+areas (our first case study in Sec. \ref{sec:ex1} focuses on this).
 
-Ternary plots also, under certain conditions, have a special
-relationship with \(R_0\), the initial reproduction number, which has
-been called the ``most important quantity in epidemiology'\,' \citep{}.
-The reproduction is itself a time-free concept and is the average number
-of secondary infections primary infection will generate when the primary
-infection is added to a completely susceptible population. To see the
-importance of \(R_0\), one need only read the newspaper \citep{} or look
-at the length of the table of estimated \(R_0\) quantities for COVID-19
-\citep{}.
-
-In regards to traditional \(X\) vs.~time plots, \(R_0\) is difficult to
-visualize. For example, as seen in Fig. \ref{fig:r0-trad}, it is
-possible that two models have the same value of \(R_0\) but there is no
-way of knowing that from looking at the graph.
-
-It is possible to identify, at a glance, which epidemic paths have the
-same value of \(R_0\) via ternary plot visualizations, under some
-circumstances. To show, this we first need to introduce the classic,
-deterministic Kermack and McKendrick SIR model where the initial number
-in each state \((S(0), I(0), R(0))\) are known and \(N\) the population
-size is constant. The movement of individuals among the states is given
-by the following differential equations (Eq. \eqref{eq:sir-ode}) where
+To show that ternary plots have to potential to aid in the comparison of
+different epidemic's \(R_0\)s -in certain circumstances- we first need
+to introduce the classic, deterministic Kermack and McKendrick
+\cite{CITETHIS?} SIR model where the initial number in each state
+\((S(0), I(0), R(0))\) are known and \(N\) the population size is
+constant. The movement of individuals among the states is given by the
+following differential equations (Eq. \eqref{eq:sir-ode}) where
 \(\beta\) is the average infection rate and \(\gamma\) is the average
@@
   \end{align}
+
+\textcolor{violet}{SUGGESTION: define $R_0$ mathematically here (though referred to in figure 2 - which still needs to be defined there).}
 
@@
 The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
-result from \cite{} and is shown in detail in Appendix \ref{app:proof}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR
-models that have the same initial percent of individuals in each state
-and \(R_0\) then for every point on the epidemic path of the first SIR
-model is also a point on the epidemic path of the second SIR model.
+result from \cite{Harko2014} and is shown in detail in Appendix
+\ref{app:proof}. The consequence of Theorem \ref{thm:sir-scale} is that
+for two SIR models that have the same initial percent of individuals in
+each state and \(R_0\) then for every point on the epidemic path of the
+first SIR model is also a point on the epidemic path of the second SIR
+model. Taking the sample simulations from Fig.
+\ref{fig:different-scales-standard}, Fig.
+\ref{fig:different-scales-tern} presents these models in a ternary plot.
+
+\begin{figure}[!ht]
+%% Code to reproduce is in inst/paper_figs/different-time-scales.R
+    \centering
+    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
+    \caption{Ternary view of \# in each state.
+Both this Fig.
+and Fig.
+\ref{fig:different-scales-standard} display the same two sets of data.
+Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
+\textcolor{violet}{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view.
+\textcolor{violet}{I might suggest using opacity to show time scale here?} \textcolor{violet}{Make this reproducible too?}}
+    \label{fig:different-scales-tern}
+\end{figure}
+
+In \pkg{EpiCompare}, we also provide visualization tools to aid in
+comparing models in the ternary / time-invariant space, which is
+presented in more details in Section
+\ref{sec:time-invariant-in-simplex}.
+
+\subsection[Simplexes beyond ternary plots]{Simplexes beyond ternary
+plots and time-invariant comparison
+tools}\label{sec:time-invariant-in-simplex}
+
+Beyond ternary plots, and the associated SIR models, we can extend the
+time-invariant approach described in Section \ref{sec:ternary} to higher
+dimensions (i.e.~models with more states). The constraints in 3d that
+are met with the SIR model (that is
+\(\sum_{i=1}^3 (\text{number in state}(i)) = N(t)\)) actually represents
+a spaces of 3d simplexes, and the ternary plot specifically represents
+these after scaling to examine such values as proportions (ternary plots
+are known as a 3d unit simplex due to it's scaling). This same scaling
+for larger models (i.e.~with more states) can be done onto different
+simplexes. In this package we present tools to help compare models
+(mostly through similations), and this tool compares these objects after
+projecting them into a one-dimension less space through the simplexical
+structure of the data.
+
+In \pkg{EpiCompare} we allow for the compare epidemics in these higher
+dimensional spaces by first projecting onto these simplexes. Even though
+higher dimensional models may not be able to visualized, we provide
+multiple tools to aid in the comparison of models and epidemics. The
+first of which uses multiple simulations under specific model parameters
+to assess the bairabilty of the model fit. In \pkg{EpiCompare} we
+provide ways to create prediction regions for a true epidemic under the
+model assumptions using these simulations. These regions require
+representing multi-dimensional structures for functions to completely
+contain epidemics, and treat these simulations and epidmeics as
+filamental objects. We extend off of papers like \citet{Dalmasso2019a}
+to create these bands. These high dimensional bands allow the user to
+assess if the true epidemic is within the band (there-by assessing the
+model's representation of the epidemic), and also compare who different
+models are from each other through distances that compare sets. We
+recommend using the Hausdorff distance to compare such sets as it
+captures how much bigger the sets would have to expand to cover each
+other, and is defined mathematically as \[
+d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
+\]
 

### 2020-12-15 R/distance-psuedo-density.R

@@ -111,8 +111,8 @@ distance_psuedo_density_function.tidy_dist_mat <- function(x, x_new = NULL,
   if (is.null(x_new)){
     rnames <- rownames(x) # data.frame
     
-    kernel_dist_mat <- stats::dnorm(x/sigma)
-    psuedo_density <- apply(unclass(kernel_dist_mat), MARGIN = 1, mean)
+    kernel_dist_mat <- stats::dnorm(as.matrix(x)/sigma)
+    psuedo_density <- apply(kernel_dist_mat, MARGIN = 1, mean)
     
     
     if (df_out) {
@@ -132,8 +132,8 @@ distance_psuedo_density_function.tidy_dist_mat <- function(x, x_new = NULL,
   } else {
     rnames <- rownames(x_new) # data.frame
     
-    kernel_dist_mat <- stats::dnorm(x_new/sigma)
-    psuedo_density <- apply(unclass(kernel_dist_mat), MARGIN = 1, mean)
+    kernel_dist_mat <- stats::dnorm(as.matrix(x_new)/sigma)
+    psuedo_density <- apply(kernel_dist_mat, MARGIN = 1, mean)
     
     
     if (df_out) {

### 2020-12-15 jss/EpiCompare-paper/EpiCompare-paper.Rmd
update to paper (includes comparison of a true observation to simulationed observations). I've seen a different seed give .03 instead of 0 proportion above.

@@
 library(ggplot2)
+set.seed(2)
 
@@
 
-HERE BEN IS GOING TO LOOK AT THE EPIMODEL OUTPUT DIFFERENT PREDICTION BAND THINGS.
+```{r echo = F}
+simple_sir <- both_models %>% filter(Type == "Simple SIR") %>%
+  rename(S = "X0", I = "X1", R = "X2") %>%
+  select(Type, sim, t, S, I, R)
 
-\begin{itemize}
-\item thing
-\end{itemize}
+hagelloch_sir2 <- hagelloch_sir %>%
+  rename(t = "time") %>%
+  mutate(Type = "true observation",
+         sim = 0) %>%
+  select(Type, sim, t, S, I, R)
+```
+```{r}
+# after cleaning up and combining 
+all_together_df <- rbind(simple_sir,
+                         hagelloch_sir2)
+all_together_df[c(1:2, nrow(all_together_df) - c(1:0)),]
+
+compression_df <- all_together_df %>% group_by(Type, sim) %>% 
+  filament_compression(data_columns = c("S","I","R"), 
+                       number_points = 20)
+```
+
+```{r echo = F}
+# # -OR- for 2d onto the simplex (done under the hood in geom_prediction_band:
+# compression_df <- all_together_df %>% 
+#   as.data.frame() %>% # just to be sure...
+#   get_xy_coord(xyz_col = c("S", "I", "R")) %>% # to 2d simplex space
+#   group_by(Type, sim) %>%
+#   filament_compression(data_columns = c("x","y"), 
+#                        number_points = 20)
+
+```
+
+```{r}
+dmat <- compression_df %>% 
+  dist_matrix_innersq_direction(
+    position = c(1:length(compression_df))[
+      names(compression_df) %in% c("S","I", "R")],
+    id_as_columns = T)
+
+tdmat <- tidy_dist_mat(as.matrix(dmat[,-c(1:2)]),
+                       rownames_df = ungroup(dmat[,1:2]),
+                       colnames_df = ungroup(dmat[,1:2]))
+
+simple_sir_true_obs_info <- tdmat %>% 
+  compare_new_to_rest_via_distance(
+    new_name_id = data.frame(Type = "true observation", sim = 0),
+    distance_func = distance_psuedo_density_function, 
+    sigma = "20%")
+
+```
+
+```{r echo = F}
+# EpiModel simulations:
+epimodel_sir <- both_models %>% filter(Type == "EpiModel SIR") %>%
+  rename(S = "X0", I = "X1", R = "X2") %>%
+  select(Type, sim, t, S, I, R)
+
+hagelloch_sir2 <- hagelloch_sir %>%
+  rename(t = "time") %>%
+  mutate(Type = "true observation",
+         sim = 0) %>%
+  select(Type, sim, t, S, I, R)
+
+all_together_df <- rbind(epimodel_sir,
+                         hagelloch_sir2)
+
+compression_df <- all_together_df %>% group_by(Type, sim) %>% 
+  filament_compression(data_columns = c("S","I","R"), 
+                       number_points = 20)
+
+dmat <- compression_df %>% 
+  dist_matrix_innersq_direction(
+    position = c(1:length(compression_df))[
+      names(compression_df) %in% c("S","I", "R")],
+    id_as_columns = T)
+
+tdmat <- tidy_dist_mat(as.matrix(dmat[,-c(1:2)]),
+                       rownames_df = ungroup(dmat[,1:2]),
+                       colnames_df = ungroup(dmat[,1:2]))
+
+epimodel_sir_true_obs_info <- tdmat %>% 
+  compare_new_to_rest_via_distance(
+    new_name_id = data.frame(Type = "true observation", sim = 0),
+    distance_func = distance_psuedo_density_function, 
+    sigma = "20%") 
+```
+
+```{r hagelloch-extremeness, echo = FALSE}
+simple_sir_info <- simple_sir_true_obs_info %>%
+  select(-sim) %>%
+  mutate(Type = "Simple SIR")
+eimodel_sir_info <- epimodel_sir_true_obs_info %>%
+  select(-sim) %>%
+  mutate(Type = "EpiModel SIR")
+
+rbind(simple_sir_info, eimodel_sir_info)  %>%
+  kable(format = "latex", booktabs = TRUE, 
+        col.names = 
+          linebreak(c("Type",
+                      "simulations-based estimated psuedo-density",
+              "proportion of simulations with lower estimated psuedo-density"),
+              align = c("l")),
+        caption = paste("The extremeness of the true simulations based on",
+                        "comparing psuedo-density estimates between true",
+                        "vs simulated curves"), label = "hags-extreme") %>%
+  kable_styling(position = "center", latex_options = "hold_position") %>%
+  column_spec(2:3, width = "6cm")
+```

### 2020-12-15 jss/EpiCompare-paper/EpiCompare-paper.tex
update to paper (includes comparison of a true observation to simulationed observations). I've seen a different seed give .03 instead of 0 proportion above.

@@
 
-HERE BEN IS GOING TO LOOK AT THE EPIMODEL OUTPUT DIFFERENT PREDICTION
-BAND THINGS.
+\begin{CodeChunk}
+\begin{CodeInput}
+R> # after cleaning up and combining 
+R> all_together_df <- rbind(simple_sir,
++                          hagelloch_sir2)
+R> all_together_df[c(1:2, nrow(all_together_df) - c(1:0)),]
+\end{CodeInput}
+\begin{CodeOutput}
+# A tibble: 4 x 6
+# Groups:   sim [2]
+  Type               sim     t     S     I     R
+  <chr>            <dbl> <dbl> <dbl> <dbl> <dbl>
+1 Simple SIR           1     0   188     0     0
+2 Simple SIR           1     1   187     1     0
+3 true observation     0    54     1     0   187
+4 true observation     0    55     1     0   187
+\end{CodeOutput}
+\begin{CodeInput}
+R> compression_df <- all_together_df %>% group_by(Type, sim) %>% 
++   filament_compression(data_columns = c("S","I","R"), 
++                        number_points = 20)
+\end{CodeInput}
+\end{CodeChunk}
 
-\begin{itemize}
-\item thing
-\end{itemize}
+\begin{CodeChunk}
+\begin{CodeInput}
+R> dmat <- compression_df %>% 
++   dist_matrix_innersq_direction(
++     position = c(1:length(compression_df))[
++       names(compression_df) %in% c("S","I", "R")],
++     id_as_columns = T)
+R> 
+R> tdmat <- tidy_dist_mat(as.matrix(dmat[,-c(1:2)]),
++                        rownames_df = ungroup(dmat[,1:2]),
++                        colnames_df = ungroup(dmat[,1:2])) 
+R> 
+R> 
+R> simple_sir_true_obs_info <- tdmat %>% 
++   compare_new_to_rest_via_distance(
++     new_name_id = data.frame(Type = "true observation", sim = 0),
++     distance_func = distance_psuedo_density_function, 
++     sigma = "20%") 
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}[!h]
+
+\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing psuedo-density estimates between true vs simulated curves}
+\centering
+\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
+\toprule
+Type & simulations-based estimated psuedo-density & proportion of simulations with lower estimated psuedo-density\\
+\midrule
+Simple SIR & 0.0036733 & 0\\
+EpiModel SIR & 0.0028813 & 0\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
 

### 2020-12-15 R/distance-depth.R

@@ -1,3 +1,66 @@
+
+#' Title
+#'
+#' @param tdmat must be a \code{tidy_dist_mat} object
+#' @param new_name_id data.frame containing which rownames of the \code{tdmat}
+#' are associated with the new observations. Note that this data.frame must
+#' be of the same format as \code{rownames(tdmat)}.
+#' @param distance_func distance function to calculates some type of ordering
+#' between observations with the distance matrix. This package currently
+#' provides 3 functions for this:
+#' \itemize{
+#'   \item{\code{distance_psudeo_density_function}}
+#'   \item{\code{distance_depth_function}}
+#'   \item{\code{local_distance_depth_function}}
+#' }
+#' If you are trying to make your own, please see examples of these methods for
+#' the \code{tidy_dist_mat} class. It's minimum required parameters are \code{x}
+#' \code{x_new=NULL}, \code{df_out}. And get's passed \code{...}
+#' @param ... additional parameters to be passed to the dist_func
+#' @param df_out indicates if one should return a data.frame or a list,
+#' by default returns data.frame.
+#' 
+#' @return returns either a data.frame with information about quantile score 
+#' value for new observations, the proportion of the simulations that have more 
+#' extreme values than it and the observations names (if df_out = T), else it 
+#' returns a list of quatile scores and proportion of simulations that had more
+#' extreme values than it.
+#' @export
+#'
+compare_new_to_rest_via_distance <- function(tdmat, 
+                                             new_name_id = data.frame(id = "true observation"), 
+                                             distance_func = distance_psuedo_density_function, 
+                                             ..., df_out = T){
+  
+  tdmat_small <- tdmat[not(new_name_id)]
+  
+  ranking_original <- distance_func(tdmat_small, ..., df_out = F)
+  
+  
+  tdmat_selected <- tdmat[new_name_id, not(new_name_id)]
+  new_ranking <- distance_func(x = tdmat_small, x_new = tdmat_selected, ..., 
+                               df_out = df_out)
+  
+  
+  if (df_out){
+    output_column_name <- names(new_ranking)[!(names(new_ranking) %in% names(rownames(tdmat_selected)))]
+    new_ranking$proportion_above <- sapply(new_ranking[output_column_name], 
+                                           function(x_val) {
+                                             mean(x_val > ranking_original)})
+    return(new_ranking)
+  } else {
+    proportion_above <- sapply(new_ranking, function(x_val) {
[…]

### 2020-12-16 jss/EpiCompare-paper/EpiCompare-paper.Rmd
small update to final part of paper

@@
 library(ggplot2)
-set.seed(2)
+set.seed(1)
 
@@
 ```
+
 However, both models are not a good fit to the filamental path as opposed to the individual points in $(S, I, R)$-space.
-What we mean by that is that both models predict epidemic paths that generally have a single defined peak of infection whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
+This can be captures with the set of simulations both models predict, which all generally have a single defined peak of infection whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
+This observation is backed up by the below analysis that demonstrates that the estimated psuedo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme}.
 In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the EpiModel network model is a better fit than the Simple SIR model, and 2) the fit is only good at the individual point level as opposed to the epidemic path level.
@@
 ```{r}
-# after cleaning up and combining 
+#-- after cleaning up and combining --
 all_together_df <- rbind(simple_sir,
                          hagelloch_sir2)
-all_together_df[c(1:2, nrow(all_together_df) - c(1:0)),]
+```
 
+```{r echo = F}
+all_together_df[c(1:2, nrow(all_together_df) - c(1:0)),] %>% 
+  kable(booktabs = TRUE,
+        caption = paste("Top and bottom 2 rows of \\tt{all\\_together\\_df}\\text{,",
+                        "combining both simulated epidemics and the true",
+                        "epidemic.}"), label = "cif-all-together-df") %>%
+  kable_styling(position = "center", 
+                latex_options = "hold_position"
+                      )
+```
+
+```{r}
 compression_df <- all_together_df %>% group_by(Type, sim) %>% 

### 2020-12-16 jss/EpiCompare-paper/EpiCompare-paper.tex
small update to final part of paper

@@
 However, both models are not a good fit to the filamental path as
-opposed to the individual points in \((S, I, R)\)-space. What we mean by
-that is that both models predict epidemic paths that generally have a
-single defined peak of infection whereas the data certainly looks like
-it has two distinct peaks, likely caused by our assumed super-spreader
-event. In conclusion, \pkg{EpiCompare} makes it clear that, at a glance,
-1) the EpiModel network model is a better fit than the Simple SIR model,
-and 2) the fit is only good at the individual point level as opposed to
-the epidemic path level.
+opposed to the individual points in \((S, I, R)\)-space. This can be
+captures with the set of simulations both models predict, which all
+generally have a single defined peak of infection whereas the data
+certainly looks like it has two distinct peaks, likely caused by our
+assumed super-spreader event. This observation is backed up by the below
+analysis that demonstrates that the estimated psuedo-density of the
+observed epidemic (relative to the simulations from either model) is
+much less likely then \textbf{any} of the simulations (reported in Table
+\ref{tab:hags-extreme}. In conclusion, \pkg{EpiCompare} makes it clear
+that, at a glance, 1) the EpiModel network model is a better fit than
+the Simple SIR model, and 2) the fit is only good at the individual
+point level as opposed to the epidemic path level.
 
@@
 \begin{CodeInput}
-R> # after cleaning up and combining 
+R> #-- after cleaning up and combining --
 R> all_together_df <- rbind(simple_sir,
 +                          hagelloch_sir2)
-R> all_together_df[c(1:2, nrow(all_together_df) - c(1:0)),]
 \end{CodeInput}
-\begin{CodeOutput}
-# A tibble: 4 x 6
-# Groups:   sim [2]
-  Type               sim     t     S     I     R
-  <chr>            <dbl> <dbl> <dbl> <dbl> <dbl>
-1 Simple SIR           1     0   188     0     0
-2 Simple SIR           1     1   187     1     0
-3 true observation     0    54     1     0   187
-4 true observation     0    55     1     0   187
-\end{CodeOutput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}[!h]
+
+\caption{\label{tab:cif-all-together-df}Top and bottom 2 rows of \tt{all\_together\_df}\text{, combining both simulated epidemics and the true epidemic.}}
+\centering
+\begin{tabular}[t]{lrrrrr}
+\toprule
+Type & sim & t & S & I & R\\
+\midrule
+Simple SIR & 1 & 0 & 188 & 0 & 0\\
+Simple SIR & 1 & 1 & 187 & 1 & 0\\
+true observation & 0 & 54 & 1 & 0 & 187\\
+true observation & 0 & 55 & 1 & 0 & 187\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
+\begin{CodeChunk}
 \begin{CodeInput}

### 2020-12-20 jss/EpiCompare-paper/EpiCompare-paper.Rmd
distance functions now can return a tidy_dist_mat object

@@
 ```{r}
-dmat <- compression_df %>% 
+tdmat <- compression_df %>% 
   dist_matrix_innersq_direction(
@@
       names(compression_df) %in% c("S","I", "R")],
-    id_as_columns = T)
-
-tdmat <- tidy_dist_mat(as.matrix(dmat[,-c(1:2)]),
-                       rownames_df = ungroup(dmat[,1:2]),
-                       colnames_df = ungroup(dmat[,1:2]))
+    tdm_out = T)
 
@@
 
-dmat <- compression_df %>% 
+tdmat <- compression_df %>% 
   dist_matrix_innersq_direction(

### 2020-12-20 jss/EpiCompare-paper/EpiCompare-paper.tex
distance functions now can return a tidy_dist_mat object

@@
 \begin{CodeInput}
-R> dmat <- compression_df %>% 
+R> tdmat <- compression_df %>% 
 +   dist_matrix_innersq_direction(
@@
 +       names(compression_df) %in% c("S","I", "R")],
-+     id_as_columns = T)
-R> 
-R> tdmat <- tidy_dist_mat(as.matrix(dmat[,-c(1:2)]),
-+                        rownames_df = ungroup(dmat[,1:2]),
-+                        colnames_df = ungroup(dmat[,1:2])) 
-R> 
++     tdm_out = T)
 R> 

### 2020-12-20 tests/testthat/test-distance-between.R

@@ -334,7 +334,7 @@ test_that("test dist_matrix_innersq_2d basic checks", {
   testthat::expect_true(all(my_dist >= 0))
 })
 
-test_that("test dist_matrix_innersq_direction basic checks (2d)", {
+test_that("test dist_matrix_innersq_direction.list, basic checks (2d)", {
   function_list <- list(function(x) x^2, function(x) sqrt(x), function(x) x)
 
   my_list <- lapply(function_list, function(f) {
@@ -347,6 +347,159 @@ test_that("test dist_matrix_innersq_direction basic checks (2d)", {
   testthat::expect_true(all(diag(my_dist) == 0))
   testthat::expect_true(all(my_dist >= 0))
 })
+
+test_that("test dist_matrix_innersq_direction.grouped_df, basic checks (2d)", {
+  function_list <- list(function(x) x^2, function(x) sqrt(x), function(x) x)
+  
+  my_list <- lapply(function_list, function(f) {
+    data.frame(x = 1:20,
+               y = f(1:20))
+  })
+  
+  for (idx in 1:3){
+    my_list[[idx]] <- my_list[[idx]] %>% dplyr::mutate(id = idx)
+  }
+  my_grouped_df <- my_list %>% 
+    dplyr::bind_rows() %>%
+    group_by(.data$id)
+  
+  
+  my_dist <- dist_matrix_innersq_direction(my_grouped_df, position = 1:2)
+  
+  testthat::expect_equivalent(t(my_dist), my_dist)
+  testthat::expect_true(all(diag(my_dist) == 0))
+  testthat::expect_true(all(my_dist >= 0))
+  
+  my_tdm <- dist_matrix_innersq_direction(my_grouped_df, position = 1:2,
+                                          tdm_out = T)
+  
+  
+  testthat::expect_equivalent(as.matrix(my_tdm), my_dist)
+  testthat::expect_equivalent(rownames(my_tdm), data.frame(id = 1:3))
+})
+
+
+test_that("test dist_matrix_innersq_direction.grouped_df, basic checks, tdm (2d)", {
+  function_list <- list(function(x) x^2, function(x) sqrt(x), function(x) x)
+  
+  my_list <- lapply(function_list, function(f) {
+    data.frame(x = 1:20,
+               y = f(1:20))
+  })
+  
+  for (idx in 1:3){
+    my_list[[idx]] <- my_list[[idx]] %>% dplyr::mutate(id = idx)
[…]

### 2020-12-20 README.Rmd
attempt 1 for github action (dangerous) - check, covr and pkgdown all in 1

@@
 [![codecov](https://codecov.io/gh/skgallagher/EpiCompare/branch/master/graph/badge.svg)](https://codecov.io/gh/skgallagher/EpiCompare)
+[![R-CMD-check](https://github.com/skgallagher/EpiCompare/workflows/R-CMD-check/badge.svg)](https://github.com/skgallagher/EpiCompare/actions)
 <!-- badges: end -->

### 2020-12-21 README.Rmd
correct version of github actions

--- 2020-12-14 README.Rmd
@@
 <!-- badges: start -->
-[![Travis build status](https://travis-ci.org/skgallagher/EpiCompare.svg?branch=master)](https://travis-ci.org/skgallagher/EpiCompare)
+[![R-CMD-check, code coverage, pkgdown site creation](https://github.com/skgallagher/EpiCompare/workflows/R-CMD-check,%20code%20coverage,%20pkgdown%20site%20creation/badge.svg)](https://github.com/skgallagher/EpiCompare/actions)
+<!-- [![Travis build status](https://travis-ci.org/skgallagher/EpiCompare.svg?branch=master)](https://travis-ci.org/skgallagher/EpiCompare) -->
 [![codecov](https://codecov.io/gh/skgallagher/EpiCompare/branch/master/graph/badge.svg)](https://codecov.io/gh/skgallagher/EpiCompare)
+
 <!-- badges: end -->
@@
 ## Package Creation Notes:
+
+**We're transferring to ~github actions~ and away from Travis CI.
+Thanks Travis CI for the long run (During undergrad - probably around 2015, I got introduced to Travis CI and it has been a really great tool and CIs in general are great tools).
+Sadly, open source packages (like ours) no longer gets infinite free resources on Travis.
+[Dean Attali](https://deanattali.com/blog/migrating-travis-to-github/) and [ROpenSci](https://ropensci.org/technotes/2020/11/19/moving-away-travis/) have blog posts on the situation.
+As such, if you're looking to learn from our mistakes from Travis, then the comments below stop making sense after December 15th, 2020.
+
 1.
@@
 [stack overflow](https://stackoverflow.com/questions/11285496/r-cmd-check-warning-functions-methods-with-usage-in-documentation-object-bu) post on how to pass `check` for `rownames<-.tidy_dist_mat`.
+8.
+Transferring from Travis CI to github actions.
+We only use a single workflow file (although we use code ideas found in: [check-standard/`usethis::use_github_action_check_standard()`](https://github.com/r-lib/actions/blob/master/examples/check-standard.yaml)), [pkgdown/`usethis::use_github_action("pkgdown")`](https://github.com/r-lib/actions/blob/master/examples/pkgdown.yaml) and [test-coverage/`usethis::use_github_action("test-coverage")`](https://github.com/r-lib/actions/blob/master/examples/test-coverage.yaml).
+In our `.github/workflows/R-CMD-check-coverage-pkgdown.yaml` you'll find **a** (potentially not optional) approach to preform our complex checking approach (which tries to copy the ideas in our [old travis file](https://github.com/skgallagher/EpiCompare/blob/e1298f799d9248bb330885ba5e7b0fa17ea2c83e/.travis.yml)).
+The github action easily creates a *larger* matrix that allows us to run on all standard OS, R versions and our "tidyr-current" vs "tidyr-old" split.
+Our approach with 1 github action `yaml` file makes us only look at 8 matrix options - it's possible that, in the future, we'll go back to 3 files.
 

### 2020-12-22 README.Rmd
attempts to make sure tidyr 0.8.3 really is used in tests and coverage attempt 17

@@
 Transferring from Travis CI to github actions.
-We only use a single workflow file (although we use code ideas found in: [check-standard/`usethis::use_github_action_check_standard()`](https://github.com/r-lib/actions/blob/master/examples/check-standard.yaml)), [pkgdown/`usethis::use_github_action("pkgdown")`](https://github.com/r-lib/actions/blob/master/examples/pkgdown.yaml) and [test-coverage/`usethis::use_github_action("test-coverage")`](https://github.com/r-lib/actions/blob/master/examples/test-coverage.yaml).
-In our `.github/workflows/R-CMD-check-coverage-pkgdown.yaml` you'll find **a** (potentially not optional) approach to preform our complex checking approach (which tries to copy the ideas in our [old travis file](https://github.com/skgallagher/EpiCompare/blob/e1298f799d9248bb330885ba5e7b0fa17ea2c83e/.travis.yml)).
+We only use a single workflow file (although we use code ideas found in: [check-standard/`usethis::use_github_action_check_standard()`](https://github.com/r-lib/actions/blob/master/examples/check-standard.yaml)), [test-coverage/`usethis::use_github_action("test-coverage")`](https://github.com/r-lib/actions/blob/master/examples/test-coverage.yaml).
+In our `.github/workflows/R-CMD-check-coverage.yaml` you'll find **a** (potentially not optional) approach to preform our complex checking approach (which tries to copy the ideas in our [old travis file](https://github.com/skgallagher/EpiCompare/blob/e1298f799d9248bb330885ba5e7b0fa17ea2c83e/.travis.yml)).
 The github action easily creates a *larger* matrix that allows us to run on all standard OS, R versions and our "tidyr-current" vs "tidyr-old" split.
 Our approach with 1 github action `yaml` file makes us only look at 8 matrix options - it's possible that, in the future, we'll go back to 3 files.
+We currently compile our `pkgdown` site on our own computers and then push - and will continue to do so due to complications with `reticulate` and `python` packages on the virtual machines in github actions.
 

### 2020-12-22 README.Rmd
added an updated section about r0 and time invariant analysis

--- 2020-12-21 README.Rmd
@@
 The goal of `EpiCompare` is to provide the epidemiology community with easy-to-use tools to encourage comparing and assessing epidemics and epidemiology models in a time-free manner.
-All tools attempt to adhere to `tidyverse`/`ggplot2` style to enhance easy of use.
+All tools attempt to adhere to `tidyverse`/`ggplot2` style to enhance ease of use.
 
-Time free analysis allows for stronger comparison of epidemics and model based simulations avoiding different scaling and shifts of time that mess up time-based comparisons.
+Additionally, `EpiCompare` provides tools for time invariant analysis.
+This allows for 'fairer' comparison of epidemics and model based simulations by avoiding different scaling and shifting of time that can confound time-based comparisons.
 
-To achieve this goal, the package contains:
+To achieve these goals, the package contains:
 
@@
     - Convert and then compare models from *standard epidemic packages* like `EpiModels`, `pomp`, as well as internal agent-based models, and epidemics in a common framework. 
-    - Convert *agent-based information into aggregate* to compare in the aggregate framework described above.
+    - Convert *agent-based* information into *aggregate* to compare in the aggregate framework described above.
 
@@
 We followed ideas found in [tidyr: in-packages](https://tidyr.tidyverse.org/articles/in-packages.html), for the code, and - when working with Travis CI (using a matrix for multiple builds) - we leverage ideas in [tidyverse travis on github](https://github.com/tidyverse/design/issues/95) and [tidyverse principles](https://principles.tidyverse.org/changes-multivers.html).
-2. For writing your own `geom`s and `stat`s that works with `ggtern` (which are generally restricted), the following 2 stack-exchange articles helped use do so with ease:
+2. For writing your own `geom`s and `stat`s that works with `ggtern` (which are generally restricted), the following 2 stack-exchange articles helped us do so with ease:
     a. [stack-exchange: personal code idea](https://stackoverflow.com/questions/57175114/ternary-plot-scaling-opacity-across-groups)
     b.
-[stack-exchange: being able to access ggtern's element write away](https://stackoverflow.com/questions/31589479/how-to-fix-no-item-called-packagepkg-on-the-search-list-without-using-libra#comment51172891_31589479)
+[stack-exchange: being able to access ggtern's element right away](https://stackoverflow.com/questions/31589479/how-to-fix-no-item-called-packagepkg-on-the-search-list-without-using-libra#comment51172891_31589479)
 
@@
 `geom_prediction_band` required not just `compute_group` but `compute_layer` - there is very little documentation on how to approach this correctly.
-Basically - there are problems when the `compute_group` wants to make multiple `pieces`/`groups` - and it is similar to the problem that if you do something like `aes(color = var1, group = var2)` you may actually want to do `aes(color = var1, group = paste(var1, var2))`, if there are the same `var2` values across different `var1` values but they don't mean they should be grouped together.
+Basically - there are problems when the `compute_group` wants to make multiple `pieces`/`groups` - and it is similar to the problem that if you do something like `aes(color = var1, group = var2)` you may actually want to do `aes(color = var1, group = paste(var1, var2))`, if there are the same `var2` values across different `var1` value, but should not necessarily be grouped together.
 4.
-Now that `R` have come out with version >= 4.0.0, we now need to call `.S3method("method", "class")` to define the connection for `S3` methods (e.g.
+Now that `R` has come out with version >= 4.0.0, we now need to call `.S3method("method", "class")` to define the connection for `S3` methods (e.g.
 `method.class` function), which we have for the `contained` function.
 5.
-**Does 2 wrongs make a right?** As of 9/23 `ggtern` had an [issue](https://bitbucket.org/nicholasehamilton/ggtern/issues/13/ggtern-breaks-ggplots-plot) that it messed with `ggplot2`'s legends when loaded (it over-wrote the `print.ggplot` and other functions).
-To "correct" this, we've over-wrote `ggtern`'s `print.ggplot` to correct this problem when not producing ternary plots (code in `aaa.R`).
+**Do two wrongs make a right?** As of 9/23 `ggtern` had an [issue](https://bitbucket.org/nicholasehamilton/ggtern/issues/13/ggtern-breaks-ggplots-plot) that it messed with `ggplot2`'s legends when loaded (it over-wrote the `print.ggplot` and other functions).
+We've over-written `ggtern`'s `print.ggplot` to correct this problem when not producing ternary plots (code in `aaa.R`).
 6.

### 2020-12-22 jss/EpiCompare-paper/EpiCompare-paper.Rmd
added an updated section about r0 and time invariant analysis

@@
                       echo=TRUE, warning=FALSE, message=FALSE,
-                      fig.pos = "H")
+                      fig.pos = "H",
+                      cache = TRUE)
 
@@
 The current epidemic of COVID-19 also highlights that infectious disease  models are only one piece of the overall analysis pipeline.
-University based resources like John Hopkin's and government numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and visualization can be a important first step in the process (multiple \cite{}?).} \shannon{add comment about NYT or something} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individual / agent level, network structure or just aggregate flows (review paper \cite{}?).
+University based resources like John Hopkins' and government numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and visualization can be a important first step in the process (multiple \cite{}?).} \shannon{add comment about NYT or something} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individual / agent level, network structure or just aggregate flows (review paper \cite{}?).
 All placing individuals / proportions of the populations into different states (e.g.
@@
 
-## Motivation through $R_0$ {short-title="r0" #r0}
+## What $R_0$ has to do with time-invariant analysis {short-title="r0" #r0}
 
-$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[][]{Gallagher2020}.
+$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[][]{heesterbeek2003}.
+By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population.
 To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}.
-$R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, and is commonly associated with the Susceptible-Infectious-Recovered (SIR) data/models.
-The SIR framework, composed of data and statistical models where individuals pass from Susceptible to Infectious to Recovered states is a common, if not the most common, modeling framework in infectious disease epidemiology (seen in examples of recent published works in \cite{FILLIN}).
+$R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.
 
-In regards to traditional $X$ vs.
-time plots, $R_0$ is difficult to visualize.
-For example, in a SIR simulation with two models have the same value of $R_0$, one can see in Fig.
-\ref{fig:different-scales-standard} that there is no way of knowing that from looking at the graph (this figure will be commented on more below).
-
-\begin{figure}[!ht]
-%% Code to reproduce is in inst/paper_figs/different-time-scales.R
-    \centering
-    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
-    \caption{Bivariate view of \# in each state vs.
-time.
-Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\ben{We probably need to define $\beta/\gamma$.}
-\ben{NEW: recommendation - make this plot actually have 3 examples (the current one), one where we have an affine transformation of the time scale, and one with the halleolof data.} \ben{Also these figures should be reproducible.}}
-    \label{fig:different-scales-standard}
-\end{figure}
-
-## Ternary and time-invariant visualizations for $R_0$ {short-title="ternary visualizations" #sec:ternary}
-
-It is possible to identify, at a glance, which SIR epidemic paths, like those in Fig.
-\ref{fig:different-scales-tern}  have the same value of $R_0$ via ``time-invariant'' visualizations, under some circumstances.
-Specifically, for SIR models we propose using ternary to examine an epidemic's trajectory in a more time-invariant manner.
-
-Ternary plots are sometimes used in chemistry \citep[][]{Gillespie1976} but only rarely seen in the field of infectious disease epidemiology (to our knowledge, only seen in \cite{}).
-Ternary plots illustrate the relationship of constrained 3D data, namely of points $(a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]$ with $a_i+ b_i + c_i =1$ and as such, are situational.
-The SIR framework happens to be of the situational format required for ternary plots where $(S(t), I(t), R(t))$ are the number of susceptible, infectious, and recovered individuals in each state at time $t$ and $S(t) + I(t) + R(t) = N(t)$ where $N(t)$ is the population size at time $t$.
-
-Ternary plots are time-invariant in the sense that the temporal scale does not explicitly apear in the visualization and as a consequence, allow for direct comparisons of outbreaks on different time scales (e.g.
-days vs.
-years) or of SIR data of the same disease in different areas (our first case study in Sec.
-\ref{sec:ex1} focuses on this).
-
-To show that ternary plots have to potential to aid in the comparison of different epidemic's $R_0$s -in certain circumstances- we first need to introduce the classic, deterministic Kermack and McKendrick \cite{CITETHIS?} SIR model where the initial number in each state $(S(0), I(0), R(0))$ are known and $N$ the population size is constant.
-The movement of individuals among the states is given by the following differential equations (Eq.
-\eqref{eq:sir-ode}) where $\beta$ is the average infection rate and $\gamma$ is the average recovery rate,
+Estimator(s) for $R_0$ are dependent on the epidemic modeling framework, which consists of which states an individual can occupy (e.g.
+susceptible, infectious, recovered) and a description of how individuals move from one state to the next over time.
+A common epidemic modeling framework is the SIR model, originally introduced by @kermack1927.
+Transitions from one state to the next are defined by a series of ordinary differential equations, where $N$ is the (fixed) total number individuals, $\beta$ is the rate of infection, and $\gamma$ is rate of recovery,
 \begin{align}\label{eq:sir-ode}
@@
   \end{align}
+From this, $\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}$, the ratio of the estimated infection rate compared to the estimated recovery rate.
 
-\ben{SUGGESTION: define $R_0$ mathematically here (though referred to in figure 2 - which still needs to be defined there).}
+In regards to traditional $state$ vs.
+time plots, $R_0$ is difficult to visualize, especially with respect from one epidemic to another.
+For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = .3$ and $\gamma_1 = .15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = .24$ and $\gamma_1 = .12$ over 40 days.
+Both epidemics have the same value of $R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2$.
+The epidemic trajectories are shown in the $state$ vs.
+time plots in Figure \ref{fig:sir-ex}.
+At a glance, we may assume that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
+On the other hand, we may think Model 2 has a larger $R_0$ because it has a slightly larger peak of infection than Model 1.
 
-If we have two SIR models that follow the Kermack and McKendrick SIR model, have the same percentage of individuals in the initial states, and the same value of $R_0$ then the two models will create exactly overlapping paths in a ternary plot.
+```{r echo = FALSE}
+devtools::load_all()
+library(tidyr)
+library(dplyr)
+library(knitr)
+library(kableExtra)
+library(ggplot2)
+set.seed(1)
+
+theme_set(theme_bw(base_size = 24))
+```
+
+```{r echo = FALSE, fig.cap = "\\label{fig:sir-ex}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
+set.seed(1225)
+
+beta1 <- .8
+gamma1 <- .4
+scale <- .8
+sir1 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
+                            beta = beta1, gamma = gamma1,
+                            init_SIR = c(990, 10, 0)) %>%
+  group_by(sim) %>%
+  agents_to_aggregate(states = c(tI, tR)) %>%
+  group_by(t) %>%
+  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
+            .groups = "drop") %>%
+  mutate(type = "Model 1")
+
+sir2 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
+                            beta = beta1 * scale, gamma = gamma1 * scale,
+                            init_SIR = c(990, 10, 0)) %>%
+  group_by(sim) %>%
+    agents_to_aggregate(states = c(tI, tR)) %>%
+  group_by(t) %>%
+  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
+            .groups = "drop") %>%
+  mutate(type = "Model 2")
+
+df <- bind_rows(sir1, sir2)
+
+df_long <- df %>%
+         tidyr::pivot_longer(cols = c(S, I, R))  %>%
+  mutate(name = factor(name, levels = c("S", "I", "R")))
+
+ggplot(data = df_long,
+       aes(x = t, y = value, col = type)) + 
+  facet_wrap(~name, ncol = 1, scales = "free") +
+  geom_point() +
+  scale_color_manual(values = c("red", "blue"), name = "Epidemic") +
+  labs(x = "Time",
+       y = "# in state",
+       title = latex2exp::TeX("Comparing two epidemics over 50 days with same $R_0$"))
+```
+However, when we plot the trajectories as a single curve using the ternary plot in the time-invariant view, we immediately see a different story.
+In this time-invariant view in Fig.
+\ref{fig:ternary-ex}, the points seem to overlap and form the same trajectory.
+In other words, we can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
 
@@
 
-\begin{figure}[!ht]
-%% Code to reproduce is in inst/paper_figs/different-time-scales.R
-    \centering
-    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
-    \caption{Ternary view of \# in each state.
-Both this Fig.
-and Fig.
-\ref{fig:different-scales-standard} display the same two sets of data.
-Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
-\ben{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view.
-\ben{I might suggest using opacity to show time scale here?} \ben{Make this reproducible too?}}
-    \label{fig:different-scales-tern}
-\end{figure}
+```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:ternary-ex}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
+These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered."}
+ggplot(data = df) +
+  geom_point(aes(x = S, y = I, z = R,
+                     col = type)) +
+    coord_tern() + 
+  theme_sir() +
+  scale_color_manual(values = c("red", "blue"), name = "Epidemic") 
+```
 
-In \pkg{EpiCompare}, we also provide visualization tools to aid in comparing models in the ternary / time-invariant space, which is presented in more details in Section \ref{sec:time-invariant-in-simplex}.
-
-## Simplexes beyond ternary plots and time-invariant comparison tools {short-title="Simplexes beyond ternary plots" #sec:time-invariant-in-simplex}
+## Simplices beyond ternary plots and time-invariant comparison tools {short-title="Simplexes beyond ternary plots" #sec:time-invariant-in-simplex}
 
@@
 Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.
-
-```{r echo = FALSE}
-devtools::load_all()
-library(tidyr)
-library(dplyr)
-library(knitr)
-library(kableExtra)
-library(ggplot2)
-set.seed(1)
-
-theme_set(theme_bw(base_size = 24))
-```
 

### 2020-12-22 jss/EpiCompare-paper/EpiCompare-paper.tex
added an updated section about r0 and time invariant analysis

@@
 models are only one piece of the overall analysis pipeline. University
-based resources like John Hopkin's and government numerical dashboards
+based resources like John Hopkins' and government numerical dashboards
 (across all levels of government) during the COVID-19 epidemic remind us
@@
 
-\subsection[r0]{Motivation through \(R_0\)}\label{r0}
+\subsection[r0]{What \(R_0\) has to do with time-invariant
+analysis}\label{r0}
 
 \(R_0\), the initial reproduction number, has been called the ``most
-important quantity in epdiemiology'' \citep[][]{Gallagher2020}. To see
+important quantity in epdiemiology'\,' \citep[][]{heesterbeek2003}. By
+definition, \(R_0\) is the number of expected secondary infections when
+a primary infection is introduced to a susceptible population. To see
 the importance of \(R_0\), one need only read the newspaper
@@
 maybe, the most famous \textit{time-invariant} numerical summary of an
-epidemic, and is commonly associated with the
-Susceptible-Infectious-Recovered (SIR) data/models. The SIR framework,
-composed of data and statistical models where individuals pass from
-Susceptible to Infectious to Recovered states is a common, if not the
-most common, modeling framework in infectious disease epidemiology (seen
-in examples of recent published works in \cite{FILLIN}).
+epidemic, which allows epidemics to be compared to one another in both
+different time and geographic scales.
 
-In regards to traditional \(X\) vs.~time plots, \(R_0\) is difficult to
-visualize. For example, in a SIR simulation with two models have the
-same value of \(R_0\), one can see in Fig.
-\ref{fig:different-scales-standard} that there is no way of knowing that
-from looking at the graph (this figure will be commented on more below).
-
-\begin{figure}[!ht]
-%% Code to reproduce is in inst/paper_figs/different-time-scales.R
-    \centering
-    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
-    \caption{Bivariate view of \# in each state vs.
-time.
-Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\textcolor{violet}{We probably need to define $\beta/\gamma$.}
-\textcolor{violet}{NEW: recommendation - make this plot actually have 3 examples (the current one), one where we have an affine transformation of the time scale, and one with the halleolof data.} \textcolor{violet}{Also these figures should be reproducible.}}
-    \label{fig:different-scales-standard}
-\end{figure}
-
-\subsection[ternary visualizations]{Ternary and time-invariant
-visualizations for \(R_0\)}\label{sec:ternary}
-
-It is possible to identify, at a glance, which SIR epidemic paths, like
-those in Fig. \ref{fig:different-scales-tern} have the same value of
-\(R_0\) via ``time-invariant'' visualizations, under some circumstances.
-Specifically, for SIR models we propose using ternary to examine an
-epidemic's trajectory in a more time-invariant manner.
-
-Ternary plots are sometimes used in chemistry \citep[][]{Gillespie1976}
-but only rarely seen in the field of infectious disease epidemiology (to
-our knowledge, only seen in \cite{}). Ternary plots illustrate the
-relationship of constrained 3D data, namely of points
-\((a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]\) with
-\(a_i+ b_i + c_i =1\) and as such, are situational. The SIR framework
-happens to be of the situational format required for ternary plots where
-\((S(t), I(t), R(t))\) are the number of susceptible, infectious, and
-recovered individuals in each state at time \(t\) and
-\(S(t) + I(t) + R(t) = N(t)\) where \(N(t)\) is the population size at
-time \(t\).
-
-Ternary plots are time-invariant in the sense that the temporal scale
-does not explicitly apear in the visualization and as a consequence,
-allow for direct comparisons of outbreaks on different time scales
-(e.g.~days vs.~years) or of SIR data of the same disease in different
-areas (our first case study in Sec. \ref{sec:ex1} focuses on this).
-
-To show that ternary plots have to potential to aid in the comparison of
-different epidemic's \(R_0\)s -in certain circumstances- we first need
-to introduce the classic, deterministic Kermack and McKendrick
-\cite{CITETHIS?} SIR model where the initial number in each state
-\((S(0), I(0), R(0))\) are known and \(N\) the population size is
-constant. The movement of individuals among the states is given by the
-following differential equations (Eq. \eqref{eq:sir-ode}) where
-\(\beta\) is the average infection rate and \(\gamma\) is the average
-recovery rate, \begin{align}\label{eq:sir-ode}
+Estimator(s) for \(R_0\) are dependent on the epidemic modeling
+framework, which consists of which states an individual can occupy
+(e.g.~susceptible, infectious, recovered) and a description of how
+individuals move from one state to the next over time. A common epidemic
+modeling framework is the SIR model, originally introduced by
+\citet{kermack1927}. Transitions from one state to the next are defined
+by a series of ordinary differential equations, where \(N\) is the
+(fixed) total number individuals, \(\beta\) is the rate of infection,
+and \(\gamma\) is rate of recovery, \begin{align}\label{eq:sir-ode}
       S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
@@
       R^\prime(t) &= \gamma I(t) \nonumber.
-  \end{align}
+  \end{align} From this,
+\(\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}\), the ratio of the
+estimated infection rate compared to the estimated recovery rate.
 
-\textcolor{violet}{SUGGESTION: define $R_0$ mathematically here (though referred to in figure 2 - which still needs to be defined there).}
+In regards to traditional \(state\) vs.~time plots, \(R_0\) is difficult
+to visualize, especially with respect from one epidemic to another. For
+example, consider the scenarios where the first epidemic is generated
+from a SIR model with \((S(0) = 990, I(0) = 10)\), \(\beta_1 = .3\) and
+\(\gamma_1 = .15\), and the second epidemic is generated from a SIR
+model with \((S(0) = 990, I(0) = 10)\), \(\beta_2 = .24\) and
+\(\gamma_1 = .12\) over 40 days. Both epidemics have the same value of
+\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
+trajectories are shown in the \(state\) vs.~time plots in Figure
+\ref{fig:sir-ex}. At a glance, we may assume that Model 1 has a larger
+\(R_0\) than Model 2 because the peak of infection occurs more quickly
+than in Model 2. On the other hand, we may think Model 2 has a larger
+\(R_0\) because it has a slightly larger peak of infection than Model 1.
 
-If we have two SIR models that follow the Kermack and McKendrick SIR
-model, have the same percentage of individuals in the initial states,
-and the same value of \(R_0\) then the two models will create exactly
-overlapping paths in a ternary plot.
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-2-1}
+
+}
+
+\caption{\label{fig:sir-ex}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
+\end{figure}
+\end{CodeChunk}
+
+However, when we plot the trajectories as a single curve using the
+ternary plot in the time-invariant view, we immediately see a different
+story. In this time-invariant view in Fig. \ref{fig:ternary-ex}, the
+points seem to overlap and form the same trajectory. In other words, we
+can see there is something fundamentally linking these two different
+epidemics, and this fundamental link turns out to be \(R_0\).
 
@@
 
-\begin{figure}[!ht]
-%% Code to reproduce is in inst/paper_figs/different-time-scales.R
-    \centering
-    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
-    \caption{Ternary view of \# in each state.
-Both this Fig.
-and Fig.
-\ref{fig:different-scales-standard} display the same two sets of data.
-Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.
-\textcolor{violet}{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view.
-\textcolor{violet}{I might suggest using opacity to show time scale here?} \textcolor{violet}{Make this reproducible too?}}
-    \label{fig:different-scales-tern}
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-3-1}
+
+}
+
+\caption{\label{fig:ternary-ex}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
+These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
 \end{figure}
+\end{CodeChunk}
 
-In \pkg{EpiCompare}, we also provide visualization tools to aid in
-comparing models in the ternary / time-invariant space, which is
-presented in more details in Section
-\ref{sec:time-invariant-in-simplex}.
-
-\subsection[Simplexes beyond ternary plots]{Simplexes beyond ternary
+\subsection[Simplexes beyond ternary plots]{Simplices beyond ternary
 plots and time-invariant comparison
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-6-1}
+{\centering \includegraphics{Figs/unnamed-chunk-8-1}
 
@@
 \caption{\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.
-We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-6}
+We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-8}
 \end{figure}
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-8-1}
+{\centering \includegraphics{Figs/unnamed-chunk-10-1}
 
@@
 \caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.
-We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-8}
+We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-10}
 \end{figure}
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-9-1}
+{\centering \includegraphics{Figs/unnamed-chunk-11-1}
 
@@
 \caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.
-The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-9}
+The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-11}
 \end{figure}
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-16-1}
+{\centering \includegraphics{Figs/unnamed-chunk-18-1}
 
@@
 
-\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-16}
+\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-18}
 \end{figure}
@@
 \midrule
-Simple SIR & 0.0036733 & 0\\
-EpiModel SIR & 0.0028813 & 0\\
+Simple SIR & 0.0036733 & 0.00\\
+EpiModel SIR & 0.0149686 & 0.02\\
 \bottomrule

### 2020-12-22 README.Rmd
github action attempt plus *removal of tidyr 0.8.3 coverage*

--- 2020-12-22 README.Rmd
@@
 <!-- badges: start -->
-[![R-CMD-check, code coverage, pkgdown site creation](https://github.com/skgallagher/EpiCompare/workflows/R-CMD-check,%20code%20coverage,%20pkgdown%20site%20creation/badge.svg)](https://github.com/skgallagher/EpiCompare/actions)
+[![R-CMD-check, code coverage](https://github.com/skgallagher/EpiCompare/workflows/R-CMD-check,%20code%20coverage/badge.svg)](https://github.com/skgallagher/EpiCompare/actions)
 <!-- [![Travis build status](https://travis-ci.org/skgallagher/EpiCompare.svg?branch=master)](https://travis-ci.org/skgallagher/EpiCompare) -->
@@
 We followed ideas found in [tidyr: in-packages](https://tidyr.tidyverse.org/articles/in-packages.html), for the code, and - when working with Travis CI (using a matrix for multiple builds) - we leverage ideas in [tidyverse travis on github](https://github.com/tidyverse/design/issues/95) and [tidyverse principles](https://principles.tidyverse.org/changes-multivers.html).
+**This is no longer done (was removed 22 December 2020), as the rest of `tidyrverse` has moved on and now requires `tidyr >1.0.0`.
 2. For writing your own `geom`s and `stat`s that works with `ggtern` (which are generally restricted), the following 2 stack-exchange articles helped use do so with ease:

### 2020-12-22 R/bands.R

@@ -571,19 +571,13 @@ delta_ball_compute_group_paths_to_points <- function(data, scales, params,
                                     xrange = xrange,
                                     yrange = yrange,
                                     gridbreaks = NULL)
-  
-  if (tidyr_new_interface()){
-    update_gridpoints_mat <- tidyr::pivot_wider(updated_gridpoints,
-                                                names_from = "y",
-                                                values_from = "z") %>%
-      dplyr::select(-.data$x) %>% as.matrix
-  } else {
-    update_gridpoints_mat <- tidyr::spread(updated_gridpoints,
-                                           key = "y",
-                                           value = "z") %>%
-      dplyr::select(-.data$x) %>% as.matrix
-  }
-  
+
+  update_gridpoints_mat <- tidyr::pivot_wider(updated_gridpoints,
+                                              names_from = "y",
+                                              values_from = "z") %>%
+    dplyr::select(-.data$x) %>% as.matrix
+
+
   
   
   cl <- grDevices::contourLines(x = xrange, y = yrange,
@@ -779,18 +773,12 @@ StatPredBandSpherical <- ggplot2::ggproto("StatPredBandSpherical",
                                                             xrange = xrange,
                                                             yrange = yrange)
 
-        if (tidyr_new_interface()){
-          update_gridpoints_mat <- tidyr::pivot_wider(updated_gridpoints,
-                                                      names_from = "y",
-                                                      values_from = "included"
-          ) %>%
-            dplyr::select(-.data$x) %>% as.matrix
-        } else {
-          update_gridpoints_mat <- tidyr::spread(updated_gridpoints,
-                                                 key = "y",
-                                                 value = "included") %>%
-            dplyr::select(-.data$x) %>% as.matrix
-        }
+        update_gridpoints_mat <- tidyr::pivot_wider(updated_gridpoints,
+                                                    names_from = "y",
+                                                    values_from = "included"
+        ) %>%
+          dplyr::select(-.data$x) %>% as.matrix
+        
 
         cl <- grDevices::contourLines(x = xrange, y = yrange,
                                       z = update_gridpoints_mat,levels = c(1))

### 2020-12-22 R/containment-and-band-creation.R

@@ -43,37 +43,13 @@ filament_compression <- function(grouped_df, data_columns = NULL,
     data_columns <- data_columns[!(data_columns %in% group_columns)]
   }
 
-  if (tidyr_new_interface()){
-    
-    compression_df <- grouped_df %>%
-      tidyr::nest() %>%
-      dplyr::mutate(data =
-                      purrr::map(.data$data,
-                                 function(df) {equa_dist_points_direction(
-                                   df[, data_columns],num_splits = number_points)})) %>%
-      tidyr::unnest(cols = c(.data$data))
-    
-  } else {
-
-    compression_df <- grouped_df %>%
-      tidyr::nest() %>%
-      dplyr::mutate(data =
-                      purrr::map(.data$data,
-                                 function(df) {equa_dist_points_direction(
-                                   df[, data_columns], num_splits = number_points)})) %>%
-      tidyr::unnest()
-
-    if (length(group_columns) == 1) {
-      group_sym <- dplyr::sym(group_columns)
-      compression_df <- compression_df %>% dplyr::group_by(!!group_sym)
-    }
-    if (length(group_columns) > 1) {
-      group_sym <- dplyr::syms(group_columns)
-      compression_df <- compression_df %>% dplyr::group_by(!!!group_sym)
-    }
-
-  }
-
+  compression_df <- grouped_df %>%
+    tidyr::nest() %>%
+    dplyr::mutate(data =
+                    purrr::map(.data$data,
+                               function(df) {equa_dist_points_direction(
+                                 df[, data_columns],num_splits = number_points)})) %>%
+    tidyr::unnest(cols = c(.data$data))
 
   return(compression_df)
 }
@@ -197,41 +173,25 @@ grab_top_depth_filaments <- function(grouped_df, data_columns = NULL,
     stop("selected 'conf_level' returns 0 filaments")
   }
 
-  if (tidyr_new_interface()){
-    updated_df <- grouped_df %>% tidyr::nest() %>%
-      dplyr::ungroup() %>%
-      dplyr::mutate(depth = depth_vector) %>%
-      dplyr::filter(.data$depth >
-                      stats::quantile(depth_vector, probs = 1 - conf_level))
-    # ^conf_level = 1 - alpha
+
+  updated_df <- grouped_df %>% tidyr::nest() %>%
+    dplyr::ungroup() %>%
[…]

### 2020-12-22 R/distance-between.R

@@ -432,6 +432,9 @@ dist_matrix_innersq_direction.list <- function(x,  position = NULL,
   
   # what type of object do we return?
   if (tdm_out){
+    if (is.null(rnames)){
+      rnames <- 1:length(x)
+    }
     rownames_df <- data.frame(id = rnames)
     output_tdm <- tidy_dist_mat(output_mat,rownames_df,rownames_df)
     return(output_tdm)

### 2020-12-22 R/sir-array-to-df.R

@@ -31,18 +31,11 @@ sim_arr_to_df <- function(arr){
                                 agent_id = 1:n_agents)
     df <- as.data.frame.table(arr)
 
-
-    if (tidyr_new_interface()){
-        df_spread <- df %>% tidyr::pivot_wider(names_from = .data$agents_stat,
-                                               values_from = .data$Freq) %>%
-            dplyr::select(dplyr::one_of("init_state", "tI",
-                                        "tR", "sim", "agent_id"))
-    } else {
-        df_spread <- df %>% tidyr::spread(key = .data$agents_stat,
-                                          value = .data$Freq) %>%
-            dplyr::select(dplyr::one_of("init_state", "tI",
-                                        "tR", "sim", "agent_id"))
-    }
+    df_spread <- df %>% tidyr::pivot_wider(names_from = .data$agents_stat,
+                                           values_from = .data$Freq) %>%
+        dplyr::select(dplyr::one_of("init_state", "tI",
+                                    "tR", "sim", "agent_id"))
+    
     df <- df_spread %>% dplyr::mutate(tI = ifelse(.data$init_state != 0,
                                                   0, .data$tI),
                                       tR = ifelse(.data$init_state == 2,

### 2020-12-23 README.Rmd

--- 2020-12-22 README.Rmd
@@
 We followed ideas found in [tidyr: in-packages](https://tidyr.tidyverse.org/articles/in-packages.html), for the code, and - when working with Travis CI (using a matrix for multiple builds) - we leverage ideas in [tidyverse travis on github](https://github.com/tidyverse/design/issues/95) and [tidyverse principles](https://principles.tidyverse.org/changes-multivers.html).
-2. For writing your own `geom`s and `stat`s that works with `ggtern` (which are generally restricted), the following 2 stack-exchange articles helped us do so with ease:
+**This is no longer done (was removed 22 December 2020), as the rest of `tidyrverse` has moved on and now requires `tidyr >1.0.0`.
+2. For writing your own `geom`s and `stat`s that works with `ggtern` (which are generally restricted), the following 2 stack-exchange articles helped use do so with ease:
     a. [stack-exchange: personal code idea](https://stackoverflow.com/questions/57175114/ternary-plot-scaling-opacity-across-groups)

### 2020-12-23 README.Rmd
minor update to github action workflow name and badge

@@
 <!-- badges: start -->
-[![R-CMD-check, code coverage](https://github.com/skgallagher/EpiCompare/workflows/R-CMD-check,%20code%20coverage/badge.svg)](https://github.com/skgallagher/EpiCompare/actions)
+[![R-CMD-check, code coverage](https://github.com/skgallagher/EpiCompare/workflows/R-CMD-check%20and%20code%20coverage/badge.svg)](https://github.com/skgallagher/EpiCompare/actions)
 <!-- [![Travis build status](https://travis-ci.org/skgallagher/EpiCompare.svg?branch=master)](https://travis-ci.org/skgallagher/EpiCompare) -->

### 2020-12-24 README.Rmd
more detailed description of geom_prediction_band inputs values

@@
 
-```{r eval = F, echo = F}
-   Currently there are 13 data sets in which to explore the different `EpiCompare` functions.
-These correspond to different sources of raw data along, agent vs.
-aggregate format, and emphases of different processes such as vital dynamics (e.g.
-birth and).
-
-   Details of each can be found with `?{data}`.
-
-  - `agent_sims` this is 50 x 3 x 188 array where entry (i,j,k) looks at the ith simulation, the jth statistic, and the kth agent.  The statistics are (inititial state (0/1/2), SMax, IMax).
-
-  - `agent_sims_tidy` ("Simulations of Measles outbreaks for Hagelloch, Germany, 1861 (Tidy agent format)"): tidy version of `agent_sims` (9400 x 5), each row corresponds to an individual for a single simulation, and contains additional information on the individual's initial state, and SMax, IMax.
-
-  - `EpiModel_agg_bd` Example output from the `EpiModel package` for an ICM with birth and death rates.  The population $N$ is non-constant.
-
-  - `EpiModel_det` output from the `EpiModel` package for a deterministic model (aggregate)
-
-  - `EpiModel_icm` Output from the `EpiModel` package for a stochastic ICM (aggregate)
-
-  - `hagelloch_agents` -- One row is a "sufficient" statistic for each agent's infection.
-Each agent's infection is uniquely identified by an initial state, max time before infection (or T), and max time before recovery (or T).
-For the states, 0 = S, 1 = I, 2 = R.
-
-  - `hagelloch_aug_births` Like Hagelloch raw but augmented with births (it already had deaths).  Five fake people have been added who join the population over time
-
-  - `hagelloch_raw` -- One row is an agent.
-This is imported from the `surveillance` R package and the variable descriptions are found [here](https://rdrr.io/rforge/surveillance/man/hagelloch.html) where it is originally labeled `hagelloch.df`.
-We have renamed it here to help distinguish it from the other data sets we derive from it.
-
-  - `hagelloch_raw2` Like Hagelloch raw but with additional, fake people with NA values for `tI` and `tR`
-
-  - `hagelloch_sir`  -- One row is a state of $(t, s_t, i_t, r_t)$ where $s_t + i_t + r_t = N$ for $t = 0, \dots, T=94$
-
-  - `pomp_arr` Example SIR simulation output of class `array` from the `pomp` packagee
-
-  - `pomp_df` Example SIR simulation of class `data.frame` from the `pomp` package
-
-  - `pomp_pomp` Example SIR simulation output of class `pomp` from the `pomp` package
-
-  - `pomp_sir` Example output from the `pomp` package.
-```
 ## Example

### 2020-12-24 R/bands.R

@@ -909,19 +909,19 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #' @param over_delta defines small extension of box around actual points to define
 #' contour.
 #' @param dist_params list of parameters for distance based approaches (convex 
-#' hull and delta ball). Named parameters are \code{dist_approach} with options
-#' \code{c("auto", "temporal", "equa_dist")}, and \code{num_steps} as either
-#' an integer or "auto". 
+#' hull and delta ball). See details for more information
 #'
 #' @details
-#'
+#' 
+#' \strong{Prediction Bands:}
+#' 
 #' This stat/geom can create 1 of 4 prediction band structures. These approaches
 #' can be broken into 2 subgroups, "pointwise" and "uniform" prediction bands.
 #' The rational for these splits relate to containment properties and the
 #' 'original' ideas are discussed more here:
 #' \href{https://arxiv.org/abs/1906.08832}{Arvix: 1906.08832}
 #'
-#' \strong{Pointwise}:
+#' \strong{Prediction Bands - Pointwise:}
 #'
 #' \itemize{
 #' \item \code{spherical_ball}: This prediction band is defined
@@ -934,7 +934,7 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #' for \code{conf_level} relative to all points.
 #' }
 #'
-#' \strong{Uniform}:
+#' \strong{Prediction Bands - Uniform:}
 #'
 #' These approaches focus on containing the paths/curves/filaments in
 #' uniformity. This approach uses depth (specifically a distance-based depth
@@ -954,6 +954,44 @@ stat_prediction_band <- function(mapping = NULL, data = NULL, geom = "polygon",
 #' \code{conf_level} curves we just create a convex hull and define our
 #' prediction band as such.
 #' }
+#'
+#' \strong{Distance attributes}
+#' 
+#' The \code{dist_params} list parameter informs us about what type of 
+#' distance comparisons we do between simulations. The values in the list 
+#' include:
+#' \itemize{
+#' \item{\code{"dist_approach"}: }{Options are \code{c("auto", "equa_dist",
+#' "temporal")}. Naturally we recommend \code{"auto"} - which is just 
+#' \code{"equa_dist"}. This parameter defines we calculate the distance between 
+#' 2 curves using \eqn{l_2} distance. If \code{"equa_dist"} we re-define the 
[…]

### 2020-12-30 jss/EpiCompare-paper/EpiCompare-paper.Rmd
introduction update (message improvement)

@@
 
+devtools::load_all("../../")
+
 ```
@@
 
+The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology.
+Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.
+Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g.
+flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g.
+historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.
+...][]{}^[\@Shannon: CITE].
+At the same time, descriptive statistics and visualizations (from universities like John Hopkins \citep{}^[\@Shannon: CITE], many branches and levels of government, and new organizations \citep{}^[\@Shannon: CITE]) are an important first step of the process.
+
+With the many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center comparison guides \citep{}^[\@Shannon: CITE], this field has a lot of devices to aid an individual practitioner decide the correct approach.
+At the same time, analysis packages often only address a portion of the analysis pipeline.
+Modeling tools often don't provide easy ways to compare and assess their models on new data.
+Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways.
+To fill these gaps, we present our \proglang{R} package \pkg{EpiCompare}.
+Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models.
+In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next.
+\pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.
+
+<!-- # old draft
 The recent (and currently on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology.
@@
 \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.
+-->
 
@@
 
-This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
+This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} prxesents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
 
@@
 
-$R_0$, the initial reproduction number, has been called the ``most important quantity in epdiemiology'' \citep[][]{heesterbeek2003}.
+$R_0$, the initial reproduction number, has been called the "most important quantity in epdiemiology" \citep[][]{heesterbeek2003}.
 By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population.
@@
 
-Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:ternary} to higher dimensions (i.e.
+Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:r0} to higher dimensions (i.e.
 models with more states).

### 2020-12-30 jss/EpiCompare-paper/EpiCompare-paper.tex
introduction update (message improvement)

@@
 
-The recent (and currently on-going) COVID-19 global pandemic has
-galvanized public interest in understanding more about infectious
-disease modeling and has highlighted the usefulness of research in the
-area of infectious disease epidemiology. Infectious disease models
-typically attempt to do one or more of the following: 1) predict the
-spread of current and future epidemics
+The recent (and on-going) COVID-19 global pandemic has galvanized public
+interest in understanding more about infectious disease modeling and has
+highlighted the usefulness of research in the area of infectious disease
+epidemiology. Infectious diseases inflict enormous burdens on the world:
+millions of lives lost and trillions of dollars spent yearly. Infectious
+disease models typically attempt to do one or more of the following: 1)
+predict the spread of current and future epidemics
 \citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
@@
 project epidemic scenarios under pre-specified parameters
-\citep[e.g. ...][]{}. The COVID-19 pandemic highlights how all three
-goals are important both separately and taken as a whole. Infectious
-diseases inflict enormous burdens on the world: millions of lives lost
-and trillions of dollars spent yearly. Correctly analyzing and
-addressing these issues aids in prevention and mitigation of future
-outbreaks. \textcolor{orange}{really like this paragraph}
+\citep[e.g. ...][]{}\footnote{@Shannon: CITE}. At the same time,
+descriptive statistics and visualizations (from universities like John
+Hopkins \citep{}\footnote{@Shannon: CITE}, many branches and levels of
+government, and new organizations \citep{}\footnote{@Shannon: CITE}) are
+an important first step of the process.
 
-The current epidemic of COVID-19 also highlights that infectious disease
-models are only one piece of the overall analysis pipeline. University
-based resources like John Hopkins' and government numerical dashboards
-(across all levels of government) during the COVID-19 epidemic remind us
-that descriptive statistics and visualization can be a important first
-step in the process (multiple \cite{}?).\}
-\textcolor{orange}{add comment about NYT or something} Still, rightly
-so, a large amount of theoretical work goes into modeling epidemics,
-with different models focusing at the individual / agent level, network
-structure or just aggregate flows (review paper \cite{}?). All placing
-individuals / proportions of the populations into different states
-(e.g.~suspectible, exposed, infected, recovered, etc.). With all these
-models, review and comparison papers in the literature and through MIDAS
-(Models of Infectious Disease Agent Study) Control Center helps the
-individual practitioner decide the correct approach.
-\textcolor{orange}{along with expertise from healthcare professionals...}
-
-At the same time, analysis packages often only address a portion of the
-analysis pipeline. Modeling tools often don't provide easy ways to
-compare and assess their models on new data. Moreover, exploring and
-modeling epidemics require transforming and \textit{tidying} data in
-different ways. To fill these gaps, we present our our \proglang{R}
-package \pkg{EpiCompare}. Our package's primary focus is to aid and
-advance research in the area of comparison and assessment of epidemic \&
-epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the
-data analysis pipeline of infectious diseases as 1) data pre-processing,
-2) exploratory data analysis (EDA), 3) modeling and simulating, 4)
-post-processing, and 5) comparison and assessment; where each previous
-part of the pipeline influences the next. \pkg{EpiCompare} provides
-tools to aids practitioners in all areas of this pipeline.
+With the many visualization and exploratory tools, models and modeling
+paradigms, and reviews and comparisons in the literature and through
+MIDAS (Models of Infectious Disease Agent Study) Control Center
+comparison guides \citep{}\footnote{@Shannon: CITE}, this field has a
+lot of devices to aid an individual practitioner decide the correct
+approach. At the same time, analysis packages often only address a
+portion of the analysis pipeline. Modeling tools often don't provide
+easy ways to compare and assess their models on new data. Moreover,
+exploring and modeling epidemics require transforming and
+\textit{tidying} data in different ways. To fill these gaps, we present
+our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
+is to aid and advance research in the area of comparison and assessment
+of epidemic \& epidemiological models. In Figure \ref{fig:pipeline}, we
+illustrate the data analysis pipeline of infectious diseases as 1) data
+pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
+simulating, 4) post-processing, and 5) comparison and assessment; where
+each previous part of the pipeline influences the next. \pkg{EpiCompare}
+provides tools to aids practitioners in all areas of this pipeline.
 
@@
 \ref{sec:time-invariant} motivates and showcases tools of time-invariant
-analysis, section \ref{sec:overview} presents an outline of how
+analysis, section \ref{sec:overview} prxesents an outline of how
 \pkg{EpiCompare} aids a practitioner in every step of the pipeline and
@@
 \(R_0\), the initial reproduction number, has been called the ``most
-important quantity in epdiemiology'\,' \citep[][]{heesterbeek2003}. By
+important quantity in epdiemiology'' \citep[][]{heesterbeek2003}. By
 definition, \(R_0\) is the number of expected secondary infections when
@@
 Beyond ternary plots, and the associated SIR models, we can extend the
-time-invariant approach described in Section \ref{sec:ternary} to higher
+time-invariant approach described in Section \ref{sec:r0} to higher
 dimensions (i.e.~models with more states). The constraints in 3d that
@@
 \midrule
-Simple SIR & 0.0036733 & 0.00\\
-EpiModel SIR & 0.0149686 & 0.02\\
+Simple SIR & 0.0036733 & 0\\
+EpiModel SIR & 0.0028813 & 0\\
 \bottomrule

### 2021-01-06 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Added a couple paragraphs to section 2 and refined the simulations some.  also moved the section of the paper specific rmd commands to the end

@@
 Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g.
-flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g.
+flu prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g.
 historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.
@@
 Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play.
-Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
+Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of susceptible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
 Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches and in this section we present benefits of the time-invariant analysis, with 1) motivation for time-invariant analysis through $R_0$ (the initial reproduction number), 2) time-invariant visualization tools for 3-state models (e.g.
@@
 
-$R_0$, the initial reproduction number, has been called the "most important quantity in epdiemiology" \citep[][]{heesterbeek2003}.
  By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population.
-To see the importance of $R_0$, one need only read the newspaper \citep{Fisher2020} or look at the length of the table of estimated $R_0$ quantities for COVID-19 \citep{Aronson2020}.
 $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.
+For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks between 5-6.
 
@@
 
-In regards to traditional $state$ vs.
-time plots, $R_0$ is difficult to visualize, especially with respect from one epidemic to another.
-For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = .3$ and $\gamma_1 = .15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = .24$ and $\gamma_1 = .12$ over 40 days.
+With regards to traditional epidemic $state$ vs.
+$time$ plots, $R_0$ is difficult to visualize, especially with respect from one epidemic to another.
+For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = 0.3$ and $\gamma_1 = 0.15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = 0.24$ and $\gamma_1 = 0.12$ over 40 days.
 Both epidemics have the same value of $R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2$.
@@
 
-```{r echo = FALSE, fig.cap = "\\label{fig:sir-ex}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+```{r echo = FALSE, fig.cap = "\\label{fig:different-scales-standard}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
 Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
 set.seed(1225)
+theme_set(theme_bw(base_size = 24))
 
@@
        y = "# in state",
-       title = latex2exp::TeX("Comparing two epidemics over 50 days with same $R_0$"))
+       title = latex2exp::TeX("Comparing two epidemics over first 15 days with same $R_0$")) +
+  coord_cartesian(xlim = c(0, 15))
 ```
@@
 \ref{fig:different-scales-standard}, Fig.
-\ref{fig:different-scales-tern} presents these models in a ternary plot.
+\ref{fig:different-scales-tern} presents these two models in a ternary plot.
+This means, that at a glance, we can tell if two epidemics have different values of $R_0$.
 
-```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:ternary-ex}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
 Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
 These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered."}
-ggplot(data = df) +
+ggplot(data = df %>%
+         filter(t <= 15)) +
   geom_point(aes(x = S, y = I, z = R,
@@
 
-## Simplices beyond ternary plots and time-invariant comparison tools {short-title="Simplexes beyond ternary plots" #sec:time-invariant-in-simplex}
+## Beyond the Kermack and McKendrick SIR models
 
-Beyond ternary plots, and the associated SIR models, we can extend the time-invariant approach described in Section \ref{sec:r0} to higher dimensions (i.e.
-models with more states).
-The constraints in 3d that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a spaces of 3d simplexes, and the ternary plot specifically represents these after scaling to examine such values as proportions (ternary plots are known as a 3d unit simplex due to it's scaling).
+Although the result of Theorem \ref{thm:sir-scale} allows for easy visual comparison of $R_0$ in Kermack SIR models, it does require stringent assumptions of homogeneity of behavior in populations.
+The use of visualizing epidemics in a time-invariant lens via ternary plots extends beyond those of models that follow the ODEs in the Kermack-McKendrick equations.
+Any model with S, I, and R states can be visualized with ternary plots, regardless of birth and death dynamics and regardless of homogeneity of individual behavior.
+We can use ternary plots to compare the spread of a disease for groups within a population without time as a confounding factor.
+We show an example of this in a later section.
+
+But time-invariant analysis is also useful for epidemic models with more than three states.
+The constraints in three dimensions that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a space of 3d simplices, and the ternary plot specifically represents these simplices, after scaling to examine such values as proportions between 0 and 1 (ternary plots are known as a 3d unit simplex due to it's scaling).
 This same scaling for larger models (i.e.
 with more states) can be done onto different simplexes.
-In this package we present tools to help compare models (mostly through similations), and this tool compares these objects after projecting them into a one-dimension less space through the simplexical structure of the data.
+In this package we present tools to help compare models (mostly through simulations).
+s tool compares these objects after projecting them into a one-dimension-fewer space through the simplexical structure of the data.
 
-In \pkg{EpiCompare} we allow for the compare epidemics in these higher dimensional spaces by first projecting onto these simplexes.
+In \pkg{EpiCompare} we allow for the comparison epidemics in these higher dimensional spaces by first projecting onto these simplexes.
 Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics.
@@
 \]
-
-# Package tool overview {short-title="Package overview" #sec:overview}
-
-## RMD Code formatting info
-
-This is the Section \ref{sec:intro}. This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.
-
-In general, don't use Markdown, but use the more precise LaTeX commands instead:
-
-* \proglang{Java}
-* \pkg{plyr}
-
-One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).
-
-If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.
-
-## RMD \proglang{R} code {short-title="R code" #r-code}
-
-Can be inserted in regular R markdown blocks.
-
-hags hags hags \cite{Neal2004}
-
-```{r}
-x <- 1:10
-x
-```
 
@@
 ```{r echo = FALSE, results = 'hide'}
+data(hagelloch_raw)
 with(hagelloch_raw, sum(PRO > ERU, na.rm = TRUE))
@@
 ```
+
+# Package tool overview {short-title="Package overview" #sec:overview}
+
+## RMD Code formatting info
+
+This is the Section \ref{sec:intro}. This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.
+
+In general, don't use Markdown, but use the more precise LaTeX commands instead:
+
+* \proglang{Java}
+* \pkg{plyr}
+
+One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).
+
+If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.
+
+## RMD \proglang{R} code {short-title="R code" #r-code}
+
+Can be inserted in regular R markdown blocks.
+
+hags hags hags \cite{Neal2004}
+
+```{r}
+x <- 1:10
+x
+```

### 2021-01-06 jss/EpiCompare-paper/EpiCompare-paper.tex
Added a couple paragraphs to section 2 and refined the simulations some.  also moved the section of the paper specific rmd commands to the end

@@
 predict the spread of current and future epidemics
-\citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
+\citep[e.g. flu prediction][]{Biggerstaff2016}, 2) analyze past and
 current epidemics to increase scientific knowledge
@@
 many tools to examine the quantity of the population in each state along
-the infection process (for example: quantity of suspectible vs infected
+the infection process (for example: quantity of susceptible vs infected
 vs recovered individuals) don't always as intelligently capture the
@@
 
-\(R_0\), the initial reproduction number, has been called the ``most
-important quantity in epdiemiology'' \citep[][]{heesterbeek2003}. By
-definition, \(R_0\) is the number of expected secondary infections when
-a primary infection is introduced to a susceptible population. To see
-the importance of \(R_0\), one need only read the newspaper
-\citep{Fisher2020} or look at the length of the table of estimated
-\(R_0\) quantities for COVID-19 \citep{Aronson2020}. \(R_0\) is also,
-maybe, the most famous \textit{time-invariant} numerical summary of an
-epidemic, which allows epidemics to be compared to one another in both
-different time and geographic scales.
+By definition, \(R_0\) is the number of expected secondary infections
+when a primary infection is introduced to a susceptible population.
+\(R_0\) is also, maybe, the most famous \textit{time-invariant}
+numerical summary of an epidemic, which allows epidemics to be compared
+to one another in both different time and geographic scales. For
+example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
+influenza between 1.2-2, and modern measles outbreaks between 5-6.
 
@@
 
-In regards to traditional \(state\) vs.~time plots, \(R_0\) is difficult
-to visualize, especially with respect from one epidemic to another. For
-example, consider the scenarios where the first epidemic is generated
-from a SIR model with \((S(0) = 990, I(0) = 10)\), \(\beta_1 = .3\) and
-\(\gamma_1 = .15\), and the second epidemic is generated from a SIR
-model with \((S(0) = 990, I(0) = 10)\), \(\beta_2 = .24\) and
-\(\gamma_1 = .12\) over 40 days. Both epidemics have the same value of
+With regards to traditional epidemic \(state\) vs.~\(time\) plots,
+\(R_0\) is difficult to visualize, especially with respect from one
+epidemic to another. For example, consider the scenarios where the first
+epidemic is generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
+\(\beta_1 = 0.3\) and \(\gamma_1 = 0.15\), and the second epidemic is
+generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
+\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over 40 days. Both epidemics
+have the same value of
 \(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
@@
 
-\caption{\label{fig:sir-ex}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
 Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
@@
 \ref{fig:different-scales-standard}, Fig.
-\ref{fig:different-scales-tern} presents these models in a ternary plot.
+\ref{fig:different-scales-tern} presents these two models in a ternary
+plot. This means, that at a glance, we can tell if two epidemics have
+different values of \(R_0\).
 
@@
 
-\caption{\label{fig:ternary-ex}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
 Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
@@
 
-\subsection[Simplexes beyond ternary plots]{Simplices beyond ternary
-plots and time-invariant comparison
-tools}\label{sec:time-invariant-in-simplex}
+\hypertarget{beyond-the-kermack-and-mckendrick-sir-models}{%
+\subsection{Beyond the Kermack and McKendrick SIR
+models}\label{beyond-the-kermack-and-mckendrick-sir-models}}
 
-Beyond ternary plots, and the associated SIR models, we can extend the
-time-invariant approach described in Section \ref{sec:r0} to higher
-dimensions (i.e.~models with more states). The constraints in 3d that
-are met with the SIR model (that is
+Although the result of Theorem \ref{thm:sir-scale} allows for easy
+visual comparison of \(R_0\) in Kermack SIR models, it does require
+stringent assumptions of homogeneity of behavior in populations. The use
+of visualizing epidemics in a time-invariant lens via ternary plots
+extends beyond those of models that follow the ODEs in the
+Kermack-McKendrick equations. Any model with S, I, and R states can be
+visualized with ternary plots, regardless of birth and death dynamics
+and regardless of homogeneity of individual behavior. We can use ternary
+plots to compare the spread of a disease for groups within a population
+without time as a confounding factor. We show an example of this in a
+later section.
+
+But time-invariant analysis is also useful for epidemic models with more
+than three states. The constraints in three dimensions that are met with
+the SIR model (that is
 \(\sum_{i=1}^3 (\text{number in state}(i)) = N(t)\)) actually represents
-a spaces of 3d simplexes, and the ternary plot specifically represents
-these after scaling to examine such values as proportions (ternary plots
-are known as a 3d unit simplex due to it's scaling). This same scaling
-for larger models (i.e.~with more states) can be done onto different
-simplexes. In this package we present tools to help compare models
-(mostly through similations), and this tool compares these objects after
-projecting them into a one-dimension less space through the simplexical
-structure of the data.
+a space of 3d simplices, and the ternary plot specifically represents
+these simplices, after scaling to examine such values as proportions
+between 0 and 1 (ternary plots are known as a 3d unit simplex due to
+it's scaling). This same scaling for larger models (i.e.~with more
+states) can be done onto different simplexes. In this package we present
+tools to help compare models (mostly through simulations). s tool
+compares these objects after projecting them into a one-dimension-fewer
+space through the simplexical structure of the data.
 
-In \pkg{EpiCompare} we allow for the compare epidemics in these higher
-dimensional spaces by first projecting onto these simplexes. Even though
-higher dimensional models may not be able to visualized, we provide
-multiple tools to aid in the comparison of models and epidemics. The
-first of which uses multiple simulations under specific model parameters
-to assess the bairabilty of the model fit. In \pkg{EpiCompare} we
-provide ways to create prediction regions for a true epidemic under the
-model assumptions using these simulations. These regions require
-representing multi-dimensional structures for functions to completely
-contain epidemics, and treat these simulations and epidmeics as
-filamental objects. We extend off of papers like \citet{Dalmasso2019a}
-to create these bands. These high dimensional bands allow the user to
-assess if the true epidemic is within the band (there-by assessing the
-model's representation of the epidemic), and also compare who different
-models are from each other through distances that compare sets. We
-recommend using the Hausdorff distance to compare such sets as it
-captures how much bigger the sets would have to expand to cover each
-other, and is defined mathematically as \[
+In \pkg{EpiCompare} we allow for the comparison epidemics in these
+higher dimensional spaces by first projecting onto these simplexes. Even
+though higher dimensional models may not be able to visualized, we
+provide multiple tools to aid in the comparison of models and epidemics.
+The first of which uses multiple simulations under specific model
+parameters to assess the bairabilty of the model fit. In
+\pkg{EpiCompare} we provide ways to create prediction regions for a true
+epidemic under the model assumptions using these simulations. These
+regions require representing multi-dimensional structures for functions
+to completely contain epidemics, and treat these simulations and
+epidmeics as filamental objects. We extend off of papers like
+\citet{Dalmasso2019a} to create these bands. These high dimensional
+bands allow the user to assess if the true epidemic is within the band
+(there-by assessing the model's representation of the epidemic), and
+also compare who different models are from each other through distances
+that compare sets. We recommend using the Hausdorff distance to compare
+such sets as it captures how much bigger the sets would have to expand
+to cover each other, and is defined mathematically as \[
 d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
 \]
-
-\section[Package overview]{Package tool overview}\label{sec:overview}
-
-\hypertarget{rmd-code-formatting-info}{%
-\subsection{RMD Code formatting info}\label{rmd-code-formatting-info}}
-
-This is the Section \ref{sec:intro}. This template demonstrates some of
-the basic LaTeX that you need to know to create a JSS article.
-
-In general, don't use Markdown, but use the more precise LaTeX commands
-instead:
-
-\begin{itemize}
-\item
-  \proglang{Java}
-\item
-  \pkg{plyr}
-\end{itemize}
-
-One exception is inline code, which can be written inside a pair of
-backticks (i.e., using the Markdown syntax).
-
-If you want to use LaTeX commands in headers, you need to provide a
-\texttt{short-title} attribute. You can also provide a custom identifier
-if necessary. See the header of Section \ref{r-code} for example.
-
-\subsection[R code]{RMD \proglang{R} code}\label{r-code}
-
-Can be inserted in regular R markdown blocks.
-
-hags hags hags \cite{Neal2004}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> x <- 1:10
-R> x
-\end{CodeInput}
-\begin{CodeOutput}
- [1]  1  2  3  4  5  6  7  8  9 10
-\end{CodeOutput}
-\end{CodeChunk}
 
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-8-1}
+{\centering \includegraphics{Figs/unnamed-chunk-7-1}
 
@@
 \caption{\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.
-We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-8}
+We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-7}
 \end{figure}
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-10-1}
+{\centering \includegraphics{Figs/unnamed-chunk-9-1}
 
@@
 \caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.
-We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-10}
+We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-9}
 \end{figure}
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-11-1}
+{\centering \includegraphics{Figs/unnamed-chunk-10-1}
 
@@
 \caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.
-The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-11}
+The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-10}
 \end{figure}
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-18-1}
+{\centering \includegraphics{Figs/unnamed-chunk-17-1}
 
@@
 
-\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-18}
+\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-17}
 \end{figure}
@@
 
+\section[Package overview]{Package tool overview}\label{sec:overview}
+
+\hypertarget{rmd-code-formatting-info}{%
+\subsection{RMD Code formatting info}\label{rmd-code-formatting-info}}
+
+This is the Section \ref{sec:intro}. This template demonstrates some of
+the basic LaTeX that you need to know to create a JSS article.
+
+In general, don't use Markdown, but use the more precise LaTeX commands
+instead:
+
+\begin{itemize}
+\item
+  \proglang{Java}
+\item
+  \pkg{plyr}
+\end{itemize}
+
+One exception is inline code, which can be written inside a pair of
+backticks (i.e., using the Markdown syntax).
+
+If you want to use LaTeX commands in headers, you need to provide a
+\texttt{short-title} attribute. You can also provide a custom identifier
+if necessary. See the header of Section \ref{r-code} for example.
+
+\subsection[R code]{RMD \proglang{R} code}\label{r-code}
+
+Can be inserted in regular R markdown blocks.
+
+hags hags hags \cite{Neal2004}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> x <- 1:10
+R> x
+\end{CodeInput}
+\begin{CodeOutput}
+ [1]  1  2  3  4  5  6  7  8  9 10
+\end{CodeOutput}
+\end{CodeChunk}
+
 \bibliography{EpiCompare.bib}

### 2021-01-06 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Now with a changed intro paragraph.

@@
 
-This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} prxesents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
+This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
 
@@
 
-Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play.
-Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of susceptible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
-Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches and in this section we present benefits of the time-invariant analysis, with 1) motivation for time-invariant analysis through $R_0$ (the initial reproduction number), 2) time-invariant visualization tools for 3-state models (e.g.
-SIR models), and 3) potential for similar analysis for models with more states (e.g.
-SEIR models).
-\ben{Will need to rewrite this section "introduction".}
+Epidemics can be difficult to compare to one another due to differences in diseases, locations, or population behaviors, or times.
+We focus on comparisons between epidemics adjusting for time, both the relative unit of rate of infection (e.g.
+days, months, years) and the relative beginning and end of an epidemic.
+By adjusting for the unit of infection rate, we can focus on the ``lifetime'' of an epidemic, a view that is concerned more with the total number of lives affected than due to a time constraint.
+By adjusting for the relative beginning/end of an infection we can compare, for example the first part of infection from one city to another where this is some time gap in spread.
+
+One way to adjust for both facets of time may seem obvious -- scale the time between zero and one, based on the initial time of infection $t_0$ and the final time $t_F$.
+Upon inspection, this method of adjustment requires clear definition of what a beginning and end of an epidemic constitute which requires making one or more choices.
+We propose the use of time-invariant analysis as it appears in \pkg{EpiCompare}, which avoids the need of making such a choice.
 

### 2021-01-06 jss/EpiCompare-paper/EpiCompare-paper.tex
Now with a changed intro paragraph.

@@
 \ref{sec:time-invariant} motivates and showcases tools of time-invariant
-analysis, section \ref{sec:overview} prxesents an outline of how
+analysis, section \ref{sec:overview} presents an outline of how
 \pkg{EpiCompare} aids a practitioner in every step of the pipeline and
@@
 
-Although epidemics are by definition a process that evolves over time,
-epidemics, including the recent COVID-19, often need to be compared in a
-time-invariant way to understand the processes at play. Additionally,
-many tools to examine the quantity of the population in each state along
-the infection process (for example: quantity of susceptible vs infected
-vs recovered individuals) don't always as intelligently capture the
-natural connections between the proportion of individuals in these
-states. Tools in \pkg{EpiCompare} attempt to give the user the ability
-to extend their toolkit to evaluate epidemics to also include
-time-invariant approaches and in this section we present benefits of the
-time-invariant analysis, with 1) motivation for time-invariant analysis
-through \(R_0\) (the initial reproduction number), 2) time-invariant
-visualization tools for 3-state models (e.g.~SIR models), and 3)
-potential for similar analysis for models with more states (e.g.~SEIR
-models).
-\textcolor{violet}{Will need to rewrite this section "introduction".}
+Epidemics can be difficult to compare to one another due to differences
+in diseases, locations, or population behaviors, or times. We focus on
+comparisons between epidemics adjusting for time, both the relative unit
+of rate of infection (e.g.~days, months, years) and the relative
+beginning and end of an epidemic. By adjusting for the unit of infection
+rate, we can focus on the ``lifetime'\,' of an epidemic, a view that is
+concerned more with the total number of lives affected than due to a
+time constraint. By adjusting for the relative beginning/end of an
+infection we can compare, for example the first part of infection from
+one city to another where this is some time gap in spread.
+
+One way to adjust for both facets of time may seem obvious -- scale the
+time between zero and one, based on the initial time of infection
+\(t_0\) and the final time \(t_F\). Upon inspection, this method of
+adjustment requires clear definition of what a beginning and end of an
+epidemic constitute which requires making one or more choices. We
+propose the use of time-invariant analysis as it appears in
+\pkg{EpiCompare}, which avoids the need of making such a choice.
 

### 2021-01-12 jss/EpiCompare-paper/EpiCompare-paper.Rmd
minor update to paper with section 3 image

@@
 
+# Package tool overview {short-title="Package overview" #sec:overview}
+
+\ben{**The goals of this section is to present parts of the data science pipeline, introduce how we help, why they are useful and the ideas behind it.**}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width = 1\textwidth]{images/pipeline2.png}
+    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
+    \label{fig:pipeline2}
+\end{figure}
+
 # A tour of \pkg{EpiCompare} {short-title="Tour" #sec:tour}
 
-In this section, we highlight a number of the functionalities available in \proglang{EpiCompare}.
-These functionalities include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline REF\ref{}.
+In this section, we highlight a number of the functionalities available in \pkg{EpiCompare}.
+These functionalities include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline \ref{fig:pipeline}.
 We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner.
@@
 
-# Package tool overview {short-title="Package overview" #sec:overview}
+# Helpful Rmd tricks
 

### 2021-01-12 jss/EpiCompare-paper/EpiCompare-paper.tex
minor update to paper with section 3 image

@@
 beginning and end of an epidemic. By adjusting for the unit of infection
-rate, we can focus on the ``lifetime'\,' of an epidemic, a view that is
+rate, we can focus on the ``lifetime'' of an epidemic, a view that is
 concerned more with the total number of lives affected than due to a
@@
 
+\section[Package overview]{Package tool overview}\label{sec:overview}
+
+\textcolor{violet}{**The goals of this section is to present parts of the data science pipeline, introduce how we help, why they are useful and the ideas behind it.**}
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width = 1\textwidth]{images/pipeline2.png}
+    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
+    \label{fig:pipeline2}
+\end{figure}
+
 \section[Tour]{A tour of \pkg{EpiCompare}}\label{sec:tour}
@@
 In this section, we highlight a number of the functionalities available
-in \proglang{EpiCompare}. These functionalities include data cleaning,
+in \pkg{EpiCompare}. These functionalities include data cleaning,
 visualization, simulation, and comparison, in accordance with the data
-analysis pipeline REF\ref{}. We show a full data analysis from beginning
-to end that can be accomplished in a streamlined and standardized
-manner.
+analysis pipeline \ref{fig:pipeline}. We show a full data analysis from
+beginning to end that can be accomplished in a streamlined and
+standardized manner.
 
@@
 
-\section[Package overview]{Package tool overview}\label{sec:overview}
+\hypertarget{helpful-rmd-tricks}{%
+\section{Helpful Rmd tricks}\label{helpful-rmd-tricks}}
 

### 2021-01-13 jss/EpiCompare-paper/EpiCompare-paper.Rmd
update of section 3 and section 3 pipeline figure

@@
 
+In this section we will present the tools in this package and how they aid in the  data analysis pipeline.
+We present a modification of Figure \ref{fig:pipeline} from the Section \ref{sec:intro} with Figure \ref{fig:pipeline2} with uses of our package.
+We recommend the package user keep this image around as a reference.
+All user functions are aimed to be as friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
+There are 2 different ways \pkg{EpiCompare} can be incorporated in the data analysis pipeline for epidemics, either at the very beginning when pre-processing data and visualizing raw data, or after modeling has been done and a generative model is fit.
+Figure \ref{fig:pipeline2} captures these different paths, and we will highlight both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
+
 \begin{figure}[!ht]
@@
 \end{figure}
+
+## Data Preprocessing
+
+The first step of most data analysis is cleaning up the data to be explored.
+There are multiple ways to collect the epidemiological data.
+Sometimes individual records with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and-sub population information will be collected, whereas other data collections will focus population/sub-population counts of individuals in each epidemic state\footnote{\@Shannon include citations}.
+We focus on understanding the overall impact on the population/sub-populations we provide a function to transform information about each agent’s start time of each state (e.g.
+start of infection, etc).
+This transformation between from agent information to aggregate information is very useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
+Our tools aim to allow the user to easily group agents and define new subpopulations to explore.
+This is important as many case studies have highlighted the usefulness to identify differing subpopulations\footnote{\@Shannon cite} and many state based models provide for subpopulation based states in their analysis\footnote{\@Shannon cite}.
+
+We provide a "tidyverse"-styled function, `agents_to_aggregate` to transform agent information into aggregate state information,  As a "tidy" function, our function `agents_to_aggregate` allows the user to identify any subpopulations by first doing `group_by` from `dplyr`, and also allows for infinite epidemic states - which can be for models with assumptions of sub population based states (e.g.
+youth infections), as well as indicators for death and birth dates.
+Our function, `agents_to_aggregate` allows for states to be skipped - captured with `NA` values and multiple starts of a state - (e.g.
+multiple infections) as long as the original data is stored in columns that are constrained to ordering ideas\footnote{make this more clear?}.
+Currently this function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can make the time steps more (or less) .
+`agents_to_aggregate` then returns a standardized count for each state for each time point.
+We see `agents_to_aggregate` as a really powerful tool, but also very useful to quickly transform agents information to be used in aggregate state-based models and compare state-based models to observed data.
+
+## EDA
+
+With raw data, getting to know our data now-a-days very frequently means figuring out good combinations of visualizations, numerical summaries and subsetting.
+An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but we’ve also developed tools to rapidly explore your data if your have a 3 unique states models (like the SIR model).
+Our `geom_aggregate` provides a rapid way to explore different subpopulations experience of the epidemic, and combines the ideas behind agents_to_aggregate for the SIR case, with `geom_path` and `coord_term` to visualize any number of groups epidemic trajectory in 3d simplex space using `ggplot2` and `ggtern` \citep{Wickham2016, Hamilton2018}.
+Visualization tools for SIR models were developed because (1) SIR models are the most common and basic state-based models\footnote{\@Shannon: cite} and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+
+## Model Fitting and Simulations
+
+Although this package does not focus on estimating a model for the data, we do provide some power functions for simulation of basic state models with Bernoulli approximation of ODE state models encoded in simulate_agents and simulate_SIR_agents (which can naturally be combined with agents_to_aggregate ).
+**Description of these models, reference to Shannon R0 paper, etc...**
+
+## Post-processing
+
+...
+
+## Comparions and Assessment
+
+...
 

### 2021-01-13 jss/EpiCompare-paper/EpiCompare-paper.tex
update of section 3 and section 3 pipeline figure

@@
 
+In this section we will present the tools in this package and how they
+aid in the data analysis pipeline. We present a modification of Figure
+\ref{fig:pipeline} from the Section \ref{sec:intro} with Figure
+\ref{fig:pipeline2} with uses of our package. We recommend the package
+user keep this image around as a reference. All user functions are aimed
+to be as friendly as possible, and we focus on providing the user
+``tidyverse'' style functions, that encourage piping and also follow
+clear verb naming schemes \citep{Wickham2019}. There are 2 different
+ways \pkg{EpiCompare} can be incorporated in the data analysis pipeline
+for epidemics, either at the very beginning when pre-processing data and
+visualizing raw data, or after modeling has been done and a generative
+model is fit. Figure \ref{fig:pipeline2} captures these different paths,
+and we will highlight both approaches and how to leverage
+\pkg{EpiCompare} in the subsections below.
+
 \begin{figure}[!ht]
@@
 \end{figure}
+
+\hypertarget{data-preprocessing}{%
+\subsection{Data Preprocessing}\label{data-preprocessing}}
+
+The first step of most data analysis is cleaning up the data to be
+explored. There are multiple ways to collect the epidemiological data.
+Sometimes individual records with times of different states of the
+epidemic (infection, recovery, etc.) as well as individual information
+like network structure, location, and-sub population information will be
+collected, whereas other data collections will focus
+population/sub-population counts of individuals in each epidemic
+state\footnote{\@Shannon include citations}. We focus on understanding
+the overall impact on the population/sub-populations we provide a
+function to transform information about each agent's start time of each
+state (e.g.~start of infection, etc). This transformation between from
+agent information to aggregate information is very useful for seeing the
+overall trend of the epidemic (and how it impacts different
+subpopulations). Our tools aim to allow the user to easily group agents
+and define new subpopulations to explore. This is important as many case
+studies have highlighted the usefulness to identify differing
+subpopulations\footnote{\@Shannon cite} and many state based models
+provide for subpopulation based states in their
+analysis\footnote{\@Shannon cite}.
+
+We provide a ``tidyverse''-styled function,
+\texttt{agents\_to\_aggregate} to transform agent information into
+aggregate state information, As a ``tidy'' function, our function
+\texttt{agents\_to\_aggregate} allows the user to identify any
+subpopulations by first doing \texttt{group\_by} from \texttt{dplyr},
+and also allows for infinite epidemic states - which can be for models
+with assumptions of sub population based states (e.g.~youth infections),
+as well as indicators for death and birth dates. Our function,
+\texttt{agents\_to\_aggregate} allows for states to be skipped -
+captured with \texttt{NA} values and multiple starts of a state -
+(e.g.~multiple infections) as long as the original data is stored in
+columns that are constrained to ordering
+ideas\footnote{make this more clear?}. Currently this function is
+constrained to integer time steps (for example days), but
+transformations (linear or otherwise) of the time columns can make the
+time steps more (or less) . \texttt{agents\_to\_aggregate} then returns
+a standardized count for each state for each time point. We see
+\texttt{agents\_to\_aggregate} as a really powerful tool, but also very
+useful to quickly transform agents information to be used in aggregate
+state-based models and compare state-based models to observed data.
+
+\hypertarget{eda}{%
+\subsection{EDA}\label{eda}}
+
+With raw data, getting to know our data now-a-days very frequently means
+figuring out good combinations of visualizations, numerical summaries
+and subsetting. An expert coder can start from
+\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
+we've also developed tools to rapidly explore your data if your have a 3
+unique states models (like the SIR model). Our \texttt{geom\_aggregate}
+provides a rapid way to explore different subpopulations experience of
+the epidemic, and combines the ideas behind agents\_to\_aggregate for
+the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
+visualize any number of groups epidemic trajectory in 3d simplex space
+using \texttt{ggplot2} and \texttt{ggtern}
+\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
+were developed because (1) SIR models are the most common and basic
+state-based models\footnote{\@Shannon: cite} and (2) our simplex
+representation of these epidemics emphasizes a ``time-invarance''
+representation of the data (for a refresher see Section
+\ref{sec:time-invariant}).
+
+\hypertarget{model-fitting-and-simulations}{%
+\subsection{Model Fitting and
+Simulations}\label{model-fitting-and-simulations}}
+
+Although this package does not focus on estimating a model for the data,
+we do provide some power functions for simulation of basic state models
+with Bernoulli approximation of ODE state models encoded in
+simulate\_agents and simulate\_SIR\_agents (which can naturally be
+combined with agents\_to\_aggregate ). \textbf{Description of these
+models, reference to Shannon R0 paper, etc\ldots{}}
+
+\hypertarget{post-processing}{%
+\subsection{Post-processing}\label{post-processing}}
+
+\ldots{}
+
+\hypertarget{comparions-and-assessment}{%
+\subsection{Comparions and Assessment}\label{comparions-and-assessment}}
+
+\ldots{}
 

### 2021-01-14 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Section 1 and 2 changes and references

--- 2021-01-06 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
 flu prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g.
-historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.
-...][]{}^[\@Shannon: CITE].
-At the same time, descriptive statistics and visualizations (from universities like John Hopkins \citep{}^[\@Shannon: CITE], many branches and levels of government, and new organizations \citep{}^[\@Shannon: CITE]) are an important first step of the process.
+historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}.
+At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021} .
 
-With the many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center comparison guides \citep{}^[\@Shannon: CITE], this field has a lot of devices to aid an individual practitioner decide the correct approach.
-At the same time, analysis packages often only address a portion of the analysis pipeline.
-Modeling tools often don't provide easy ways to compare and assess their models on new data.
+With the many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021}, this field has a lot of devices to aid an individual practitioner decide the correct approach.
+For example,\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
+
+At the same time, analysis packages often only address a specific portion of the analysis pipeline, for instance focusing on certain types of models.
+Modeling tools, which usually require learning package-specific syntax, often don't provide easy ways to compare and assess their models on new data.
 Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways.
 To fill these gaps, we present our \proglang{R} package \pkg{EpiCompare}.
-Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models.
+Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic and epidemiological models.
 In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next.
@@
 
-This package also emphasizes the value of approach epidemics in a \textit{time-invariant} way, not constrained to things like a specific time scale, or initial date of the first infection.
-Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play.
-Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
-Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches.
+\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a \textit{time-invariant} way.
+Epidemics, despite by definition being a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales to understand the processes at play.
+Additionally, many tools designed to examine the quantity of the population in each epidemic state (for example: quantity of Susceptible vs Infectious vs Recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
+\shannon{I don't quite understand this.
+Is this because of underreporting?
+If so I'm not sure how EpiCompare helps us here.} Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens.
 The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
-\shannon{flesh out time invariance introduction to better explain to non-experts what we mean.
-perhaps mention R0}
 
 This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
-
-SHANNON: I THINK JUST INCLUDING LOTS OF LITERATURE IN INTRO + ETC. INSTEAD OF HAVING A LITERATURE REVIEW IS BETTER.
 
@@
 Epidemics can be difficult to compare to one another due to differences in diseases, locations, or population behaviors, or times.
-We focus on comparisons between epidemics adjusting for time, both the relative unit of rate of infection (e.g.
-days, months, years) and the relative beginning and end of an epidemic.
-By adjusting for the unit of infection rate, we can focus on the ``lifetime'' of an epidemic, a view that is concerned more with the total number of lives affected than due to a time constraint.
-By adjusting for the relative beginning/end of an infection we can compare, for example the first part of infection from one city to another where this is some time gap in spread.
+In \pkg{EpiCompare}, we emphasize comparisons between epidemics adjusting for the last component, time.
+Time-invariant analysis is beneficial because by adjusting for the unit of infection rate, we can focus on the "lifetime" of an epidemic, a view that is concerned more with the number of lives affected than as opposed to any specific time constraint.
+Time-invariant analysis is also beneficial when there are gaps in time between occurrences of outbreaks of a similar nature in different geographic regions.
+Finally, time-invariant analysis is beneficial because many studies talk about a period of ``exponential growth'' of the number of outbreaks \citep{chowell2007,wallinga2007generation,forsberg2008}.
 
-One way to adjust for both facets of time may seem obvious -- scale the time between zero and one, based on the initial time of infection $t_0$ and the final time $t_F$.
-Upon inspection, this method of adjustment requires clear definition of what a beginning and end of an epidemic constitute which requires making one or more choices.
-We propose the use of time-invariant analysis as it appears in \pkg{EpiCompare}, which avoids the need of making such a choice.
+Time-invariant analysis, as it appears in \pkg{EpiCompare} solves the above issues by observing functionals which attempt to capture the general shape of the epidemic with respect to the proportion of the population in each epidemic state.
+This allows us to compare scenarios as different as, for instance as a decades-long outbreak HIV in the US compared to a 10 day outbreak of norovirus on a cruise ship.
+Moreover, this tool avoids the need for choosing a 'beginning' $t_0$ or 'end' $t_F$ of an epidemic, choices \cite{gallagher2020} show that can heavily influence, for example,  estimates of peak infection height or the reproduction number $R_0$.
 
-## What $R_0$ has to do with time-invariant analysis {short-title="r0" #r0}
+## $R_0$  and time-invariant analysis {short-title="r0" #r0}
 
@@
 $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.
-For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks between 5-6.
+For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks between 5-6 \citep{midas2020,biggerstaff2014,namee2018}.
 
 Estimator(s) for $R_0$ are dependent on the epidemic modeling framework, which consists of which states an individual can occupy (e.g.
-susceptible, infectious, recovered) and a description of how individuals move from one state to the next over time.
+susceptible, infectious, recovered) and a description of how individuals move from one state to the next over time (see \cite{hethcote1994}).
 A common epidemic modeling framework is the SIR model, originally introduced by @kermack1927.
@@
 \ref{fig:different-scales-tern} presents these two models in a ternary plot.
-This means, that at a glance, we can tell if two epidemics have different values of $R_0$.
+This means with our time-invariant ternary plot, that at a glance, we can tell if two epidemics have different values of $R_0$.
 
@@
 
-In this section, we highlight a number of the functionalities available in \proglang{EpiCompare}.
+In this section, we highlight a number of the functionalities available in \kg{EpiCompare}.
 These functionalities include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline REF\ref{}.

### 2021-01-14 jss/EpiCompare-paper/EpiCompare-paper.tex
Section 1 and 2 changes and references

--- 2021-01-06 jss/EpiCompare-paper/EpiCompare-paper.tex
@@
 project epidemic scenarios under pre-specified parameters
-\citep[e.g. ...][]{}\footnote{@Shannon: CITE}. At the same time,
-descriptive statistics and visualizations (from universities like John
-Hopkins \citep{}\footnote{@Shannon: CITE}, many branches and levels of
-government, and new organizations \citep{}\footnote{@Shannon: CITE}) are
-an important first step of the process.
+\citep[e.g.][]{ferguson2020}. At the same time, descriptive statistics
+and visualizations from universities, many branches and levels of
+government, and news organizations are an important first step of the
+process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021} .
 
 With the many visualization and exploratory tools, models and modeling
-paradigms, and reviews and comparisons in the literature and through
-MIDAS (Models of Infectious Disease Agent Study) Control Center
-comparison guides \citep{}\footnote{@Shannon: CITE}, this field has a
-lot of devices to aid an individual practitioner decide the correct
-approach. At the same time, analysis packages often only address a
-portion of the analysis pipeline. Modeling tools often don't provide
-easy ways to compare and assess their models on new data. Moreover,
-exploring and modeling epidemics require transforming and
-\textit{tidying} data in different ways. To fill these gaps, we present
-our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
-is to aid and advance research in the area of comparison and assessment
-of epidemic \& epidemiological models. In Figure \ref{fig:pipeline}, we
-illustrate the data analysis pipeline of infectious diseases as 1) data
-pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
-simulating, 4) post-processing, and 5) comparison and assessment; where
-each previous part of the pipeline influences the next. \pkg{EpiCompare}
-provides tools to aids practitioners in all areas of this pipeline.
+paradigms, and reviews and comparisons in the literature and through the
+MIDAS (Models of Infectious Disease Agent Study) network
+\citep{midasnetwork2021}, this field has a lot of devices to aid an
+individual practitioner decide the correct approach. For
+example,\proglang{R} packages such as \pkg{surveillance},
+\pkg{EpiModel}, and \pkg{pomp} have all made significant steps in
+standardizing the flow of the data analysis pipeline for epidemic
+modeling through digitizing data sets, making accessible statistical
+models, and providing a plethora of educational material for both coding
+novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
+
+At the same time, analysis packages often only address a specific
+portion of the analysis pipeline, for instance focusing on certain types
+of models. Modeling tools, which usually require learning
+package-specific syntax, often don't provide easy ways to compare and
+assess their models on new data. Moreover, exploring and modeling
+epidemics require transforming and \textit{tidying} data in different
+ways. To fill these gaps, we present our \proglang{R} package
+\pkg{EpiCompare}. Our package's primary focus is to aid and advance
+research in the area of comparison and assessment of epidemic and
+epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the
+data analysis pipeline of infectious diseases as 1) data pre-processing,
+2) exploratory data analysis (EDA), 3) modeling and simulating, 4)
+post-processing, and 5) comparison and assessment; where each previous
+part of the pipeline influences the next. \pkg{EpiCompare} provides
+tools to aids practitioners in all areas of this pipeline.
 
@@
 
-This package also emphasizes the value of approach epidemics in a
-\textit{time-invariant} way, not constrained to things like a specific
-time scale, or initial date of the first infection. Although epidemics
-are by definition a process that evolves over time, epidemics, including
-the recent COVID-19, often need to be compared in a time-invariant way
-to understand the processes at play. Additionally, many tools to examine
-the quantity of the population in each state along the infection process
-(for example: quantity of suspectible vs infected vs recovered
-individuals) don't always as intelligently capture the natural
-connections between the proportion of individuals in these states. Tools
-in \pkg{EpiCompare} attempt to give the user the ability to extend their
-toolkit to evaluate epidemics to also include time-invariant approaches.
-The goal of \pkg{EpiCompare} is not to supplant existing infectious
-disease modeling tools and software but, rather, is a concerted effort
-to create standard and fair comparisons among models developed for
-disease outbreaks and outbreak data.
-\textcolor{orange}{flesh out time invariance introduction to better explain to non-experts what we mean.  perhaps mention R0}
+\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
+\textit{time-invariant} way. Epidemics, despite by definition being a
+process that evolves over time, often need to be compared in a way not
+constrained to initial times or time scales to understand the processes
+at play. Additionally, many tools designed to examine the quantity of
+the population in each epidemic state (for example: quantity of
+Susceptible vs Infectious vs Recovered individuals) don't always as
+intelligently capture the natural connections between the proportion of
+individuals in these states.
+\textcolor{orange}{I don't quite understand this.  Is this because of underreporting?  If so I'm not sure how EpiCompare helps us here.}
+Tools in \pkg{EpiCompare} give the user the ability to extend their
+toolkit to evaluate epidemics within a time-invariant lens. The goal of
+\pkg{EpiCompare} is not to supplant existing infectious disease modeling
+tools and software but, rather, is a concerted effort to create standard
+and fair comparisons among models developed for disease outbreaks and
+outbreak data.
 
@@
 
-SHANNON: I THINK JUST INCLUDING LOTS OF LITERATURE IN INTRO + ETC.
-INSTEAD OF HAVING A LITERATURE REVIEW IS BETTER.
-
 \section[Time-invariant]{Motivation and tools for time-invariant
@@
 Epidemics can be difficult to compare to one another due to differences
-in diseases, locations, or population behaviors, or times. We focus on
-comparisons between epidemics adjusting for time, both the relative unit
-of rate of infection (e.g.~days, months, years) and the relative
-beginning and end of an epidemic. By adjusting for the unit of infection
-rate, we can focus on the ``lifetime'\,' of an epidemic, a view that is
-concerned more with the total number of lives affected than due to a
-time constraint. By adjusting for the relative beginning/end of an
-infection we can compare, for example the first part of infection from
-one city to another where this is some time gap in spread.
+in diseases, locations, or population behaviors, or times. In
+\pkg{EpiCompare}, we emphasize comparisons between epidemics adjusting
+for the last component, time. Time-invariant analysis is beneficial
+because by adjusting for the unit of infection rate, we can focus on the
+``lifetime'' of an epidemic, a view that is concerned more with the
+number of lives affected than as opposed to any specific time
+constraint. Time-invariant analysis is also beneficial when there are
+gaps in time between occurrences of outbreaks of a similar nature in
+different geographic regions. Finally, time-invariant analysis is
+beneficial because many studies talk about a period of ``exponential
+growth'\,' of the number of outbreaks
+\citep{chowell2007,wallinga2007generation,forsberg2008}.
 
-One way to adjust for both facets of time may seem obvious -- scale the
-time between zero and one, based on the initial time of infection
-\(t_0\) and the final time \(t_F\). Upon inspection, this method of
-adjustment requires clear definition of what a beginning and end of an
-epidemic constitute which requires making one or more choices. We
-propose the use of time-invariant analysis as it appears in
-\pkg{EpiCompare}, which avoids the need of making such a choice.
+Time-invariant analysis, as it appears in \pkg{EpiCompare} solves the
+above issues by observing functionals which attempt to capture the
+general shape of the epidemic with respect to the proportion of the
+population in each epidemic state. This allows us to compare scenarios
+as different as, for instance as a decades-long outbreak HIV in the US
+compared to a 10 day outbreak of norovirus on a cruise ship. Moreover,
+this tool avoids the need for choosing a `beginning' \(t_0\) or `end'
+\(t_F\) of an epidemic, choices \cite{gallagher2020} show that can
+heavily influence, for example, estimates of peak infection height or
+the reproduction number \(R_0\).
 
-\subsection[r0]{What \(R_0\) has to do with time-invariant
-analysis}\label{r0}
+\subsection[r0]{\(R_0\) and time-invariant analysis}\label{r0}
 
@@
 example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
-influenza between 1.2-2, and modern measles outbreaks between 5-6.
+influenza between 1.2-2, and modern measles outbreaks between 5-6
+\citep{midas2020,biggerstaff2014,namee2018}.
 
@@
 (e.g.~susceptible, infectious, recovered) and a description of how
-individuals move from one state to the next over time. A common epidemic
-modeling framework is the SIR model, originally introduced by
-\citet{kermack1927}. Transitions from one state to the next are defined
-by a series of ordinary differential equations, where \(N\) is the
-(fixed) total number individuals, \(\beta\) is the rate of infection,
-and \(\gamma\) is rate of recovery, \begin{align}\label{eq:sir-ode}
+individuals move from one state to the next over time (see
+\cite{hethcote1994}). A common epidemic modeling framework is the SIR
+model, originally introduced by \citet{kermack1927}. Transitions from
+one state to the next are defined by a series of ordinary differential
+equations, where \(N\) is the (fixed) total number individuals,
+\(\beta\) is the rate of infection, and \(\gamma\) is rate of recovery,
+\begin{align}\label{eq:sir-ode}
       S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
@@
 \ref{fig:different-scales-tern} presents these two models in a ternary
-plot. This means, that at a glance, we can tell if two epidemics have
-different values of \(R_0\).
+plot. This means with our time-invariant ternary plot, that at a glance,
+we can tell if two epidemics have different values of \(R_0\).
 
@@
 In this section, we highlight a number of the functionalities available
-in \proglang{EpiCompare}. These functionalities include data cleaning,
+in \kg{EpiCompare}. These functionalities include data cleaning,
 visualization, simulation, and comparison, in accordance with the data

### 2021-01-14 jss/EpiCompare-paper/EpiCompare-paper.Rmd

@@
 
+# Package tool overview {short-title="Package overview" #sec:overview}
+
+\ben{**The goals of this section is to present parts of the data science pipeline, introduce how we help, why they are useful and the ideas behind it.**}
+
+In this section we will present the tools in this package and how they aid in the  data analysis pipeline.
+We present a modification of Figure \ref{fig:pipeline} from the Section \ref{sec:intro} with Figure \ref{fig:pipeline2} with uses of our package.
+We recommend the package user keep this image around as a reference.
+All user functions are aimed to be as friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
+There are 2 different ways \pkg{EpiCompare} can be incorporated in the data analysis pipeline for epidemics, either at the very beginning when pre-processing data and visualizing raw data, or after modeling has been done and a generative model is fit.
+Figure \ref{fig:pipeline2} captures these different paths, and we will highlight both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width = 1\textwidth]{images/pipeline2.png}
+    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
+    \label{fig:pipeline2}
+\end{figure}
+
+## Data Preprocessing
+
+The first step of most data analysis is cleaning up the data to be explored.
+There are multiple ways to collect the epidemiological data.
+Sometimes individual records with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and-sub population information will be collected, whereas other data collections will focus population/sub-population counts of individuals in each epidemic state\footnote{\@Shannon include citations}.
+We focus on understanding the overall impact on the population/sub-populations we provide a function to transform information about each agent’s start time of each state (e.g.
+start of infection, etc).
+This transformation between from agent information to aggregate information is very useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
+Our tools aim to allow the user to easily group agents and define new subpopulations to explore.
+This is important as many case studies have highlighted the usefulness to identify differing subpopulations\footnote{\@Shannon cite} and many state based models provide for subpopulation based states in their analysis\footnote{\@Shannon cite}.
+
+We provide a "tidyverse"-styled function, `agents_to_aggregate` to transform agent information into aggregate state information,  As a "tidy" function, our function `agents_to_aggregate` allows the user to identify any subpopulations by first doing `group_by` from `dplyr`, and also allows for infinite epidemic states - which can be for models with assumptions of sub population based states (e.g.
+youth infections), as well as indicators for death and birth dates.
+Our function, `agents_to_aggregate` allows for states to be skipped - captured with `NA` values and multiple starts of a state - (e.g.
+multiple infections) as long as the original data is stored in columns that are constrained to ordering ideas\footnote{make this more clear?}.
+Currently this function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can make the time steps more (or less) .
+`agents_to_aggregate` then returns a standardized count for each state for each time point.
+We see `agents_to_aggregate` as a really powerful tool, but also very useful to quickly transform agents information to be used in aggregate state-based models and compare state-based models to observed data.
+
+## EDA
+
+With raw data, getting to know our data now-a-days very frequently means figuring out good combinations of visualizations, numerical summaries and subsetting.
+An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but we’ve also developed tools to rapidly explore your data if your have a 3 unique states models (like the SIR model).
+Our `geom_aggregate` provides a rapid way to explore different subpopulations experience of the epidemic, and combines the ideas behind agents_to_aggregate for the SIR case, with `geom_path` and `coord_term` to visualize any number of groups epidemic trajectory in 3d simplex space using `ggplot2` and `ggtern` \citep{Wickham2016, Hamilton2018}.
+Visualization tools for SIR models were developed because (1) SIR models are the most common and basic state-based models\footnote{\@Shannon: cite} and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+
+## Model Fitting and Simulations
+
+Although this package does not focus on estimating a model for the data, we do provide some power functions for simulation of basic state models with Bernoulli approximation of ODE state models encoded in simulate_agents and simulate_SIR_agents (which can naturally be combined with agents_to_aggregate ).
+**Description of these models, reference to Shannon R0 paper, etc...**
+
+## Post-processing
+
+...
+
+## Comparions and Assessment
+
+...
+
 # A tour of \pkg{EpiCompare} {short-title="Tour" #sec:tour}
 
-In this section, we highlight a number of the functionalities available in \kg{EpiCompare}.
-These functionalities include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline REF\ref{}.
+In this section, we highlight a number of the functionalities available in \pkg{EpiCompare}.
+These functionalities include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline \ref{fig:pipeline}.
 We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner.

### 2021-01-14 jss/EpiCompare-paper/EpiCompare-paper.tex

@@
 
+\section[Package overview]{Package tool overview}\label{sec:overview}
+
+\textcolor{violet}{**The goals of this section is to present parts of the data science pipeline, introduce how we help, why they are useful and the ideas behind it.**}
+
+In this section we will present the tools in this package and how they
+aid in the data analysis pipeline. We present a modification of Figure
+\ref{fig:pipeline} from the Section \ref{sec:intro} with Figure
+\ref{fig:pipeline2} with uses of our package. We recommend the package
+user keep this image around as a reference. All user functions are aimed
+to be as friendly as possible, and we focus on providing the user
+``tidyverse'' style functions, that encourage piping and also follow
+clear verb naming schemes \citep{Wickham2019}. There are 2 different
+ways \pkg{EpiCompare} can be incorporated in the data analysis pipeline
+for epidemics, either at the very beginning when pre-processing data and
+visualizing raw data, or after modeling has been done and a generative
+model is fit. Figure \ref{fig:pipeline2} captures these different paths,
+and we will highlight both approaches and how to leverage
+\pkg{EpiCompare} in the subsections below.
+
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width = 1\textwidth]{images/pipeline2.png}
+    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
+    \label{fig:pipeline2}
+\end{figure}
+
+\hypertarget{data-preprocessing}{%
+\subsection{Data Preprocessing}\label{data-preprocessing}}
+
+The first step of most data analysis is cleaning up the data to be
+explored. There are multiple ways to collect the epidemiological data.
+Sometimes individual records with times of different states of the
+epidemic (infection, recovery, etc.) as well as individual information
+like network structure, location, and-sub population information will be
+collected, whereas other data collections will focus
+population/sub-population counts of individuals in each epidemic
+state\footnote{\@Shannon include citations}. We focus on understanding
+the overall impact on the population/sub-populations we provide a
+function to transform information about each agent's start time of each
+state (e.g.~start of infection, etc). This transformation between from
+agent information to aggregate information is very useful for seeing the
+overall trend of the epidemic (and how it impacts different
+subpopulations). Our tools aim to allow the user to easily group agents
+and define new subpopulations to explore. This is important as many case
+studies have highlighted the usefulness to identify differing
+subpopulations\footnote{\@Shannon cite} and many state based models
+provide for subpopulation based states in their
+analysis\footnote{\@Shannon cite}.
+
+We provide a ``tidyverse''-styled function,
+\texttt{agents\_to\_aggregate} to transform agent information into
+aggregate state information, As a ``tidy'' function, our function
+\texttt{agents\_to\_aggregate} allows the user to identify any
+subpopulations by first doing \texttt{group\_by} from \texttt{dplyr},
+and also allows for infinite epidemic states - which can be for models
+with assumptions of sub population based states (e.g.~youth infections),
+as well as indicators for death and birth dates. Our function,
+\texttt{agents\_to\_aggregate} allows for states to be skipped -
+captured with \texttt{NA} values and multiple starts of a state -
+(e.g.~multiple infections) as long as the original data is stored in
+columns that are constrained to ordering
+ideas\footnote{make this more clear?}. Currently this function is
+constrained to integer time steps (for example days), but
+transformations (linear or otherwise) of the time columns can make the
+time steps more (or less) . \texttt{agents\_to\_aggregate} then returns
+a standardized count for each state for each time point. We see
+\texttt{agents\_to\_aggregate} as a really powerful tool, but also very
+useful to quickly transform agents information to be used in aggregate
+state-based models and compare state-based models to observed data.
+
+\hypertarget{eda}{%
+\subsection{EDA}\label{eda}}
+
+With raw data, getting to know our data now-a-days very frequently means
+figuring out good combinations of visualizations, numerical summaries
+and subsetting. An expert coder can start from
+\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
+we've also developed tools to rapidly explore your data if your have a 3
+unique states models (like the SIR model). Our \texttt{geom\_aggregate}
+provides a rapid way to explore different subpopulations experience of
+the epidemic, and combines the ideas behind agents\_to\_aggregate for
+the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
+visualize any number of groups epidemic trajectory in 3d simplex space
+using \texttt{ggplot2} and \texttt{ggtern}
+\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
+were developed because (1) SIR models are the most common and basic
+state-based models\footnote{\@Shannon: cite} and (2) our simplex
+representation of these epidemics emphasizes a ``time-invarance''
+representation of the data (for a refresher see Section
+\ref{sec:time-invariant}).
+
+\hypertarget{model-fitting-and-simulations}{%
+\subsection{Model Fitting and
+Simulations}\label{model-fitting-and-simulations}}
+
+Although this package does not focus on estimating a model for the data,
+we do provide some power functions for simulation of basic state models
+with Bernoulli approximation of ODE state models encoded in
+simulate\_agents and simulate\_SIR\_agents (which can naturally be
+combined with agents\_to\_aggregate ). \textbf{Description of these
+models, reference to Shannon R0 paper, etc\ldots{}}
+
+\hypertarget{post-processing}{%
+\subsection{Post-processing}\label{post-processing}}
+
+\ldots{}
+
+\hypertarget{comparions-and-assessment}{%
+\subsection{Comparions and Assessment}\label{comparions-and-assessment}}
+
+\ldots{}
+
 \section[Tour]{A tour of \pkg{EpiCompare}}\label{sec:tour}
@@
 visualization, simulation, and comparison, in accordance with the data
-analysis pipeline REF\ref{}. We show a full data analysis from beginning
-to end that can be accomplished in a streamlined and standardized
-manner.
+analysis pipeline \ref{fig:pipeline}. We show a full data analysis from
+beginning to end that can be accomplished in a streamlined and
+standardized manner.
 

### 2021-01-15 jss/EpiCompare-paper/EpiCompare-paper.Rmd
mainly citations, cleaning up references, and added proof.

@@
   \usepackage{amsmath}
+  \usepackage{amsthm}
 output: 
@@
 $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.
-For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks between 5-6 \citep{midas2020,biggerstaff2014,namee2018}.
+For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks as big as 12 \citep{midas2020,biggerstaff2014,namee2018}.
 
@@
 The epidemic trajectories are shown in the $state$ vs.
-time plots in Figure \ref{fig:sir-ex}.
+time plots in Figure \ref{fig:different-scales-standard}.
 At a glance, we may assume that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
-On the other hand, we may think Model 2 has a larger $R_0$ because it has a slightly larger peak of infection than Model 1.
+On the other hand, we may think Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15.
 
@@
 In this time-invariant view in Fig.
-\ref{fig:ternary-ex}, the points seem to overlap and form the same trajectory.
+\ref{fig:different-scales-tern}, the points seem to overlap and form the same trajectory.
 In other words, we can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
 
-More formally, let two Kermack and McKendrick (see Eq.
-\eqref{eq:sir-ode}) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
+More formally, let two Kermack and McKendrick (see Eq.\eqref{eq:sir-ode} ) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
 Assume both models have initial values $(S(0), I(0), R(0))$.
@@
 
-The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Appendix \ref{app:proof}.
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
 The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.
@@
 But time-invariant analysis is also useful for epidemic models with more than three states.
-The constraints in three dimensions that are met with the SIR model (that is $\sum_{i=1}^3 (\text{number in state}(i)) = N(t)$) actually represents a space of 3d simplices, and the ternary plot specifically represents these simplices, after scaling to examine such values as proportions between 0 and 1 (ternary plots are known as a 3d unit simplex due to it's scaling).
+The constraints in three dimensions that are met with the SIR model (there are three epidemic states and for each time step, the sum of total number of individuals in each state is the population size at that time$) actually represents a space of 3d simplices, and the ternary plot specifically represents these simplices, after scaling to examine such values as proportions between 0 and 1 (ternary plots are known as a 3d unit simplex due to it's scaling).
 This same scaling for larger models (i.e.
@@
 There are multiple ways to collect the epidemiological data.
-Sometimes individual records with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and-sub population information will be collected, whereas other data collections will focus population/sub-population counts of individuals in each epidemic state\footnote{\@Shannon include citations}.
-We focus on understanding the overall impact on the population/sub-populations we provide a function to transform information about each agent’s start time of each state (e.g.
+Sometimes individual records with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information will be collected, whereas other data collections will focus aggregate counts of individuals in each epidemic state.
+In fact, usually only the number of new infections at each time step is observed and the compartment totals are then imputed from that along with other information about the disease and the population of interest.
+We focus on understanding the overall impact of an outbreak at the aggregate/population level.
+We provide a function to transform information about each individual/agent’s start time of each state (e.g.
 start of infection, etc).
-This transformation between from agent information to aggregate information is very useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
+This transformation between agent information to aggregate information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
 Our tools aim to allow the user to easily group agents and define new subpopulations to explore.
-This is important as many case studies have highlighted the usefulness to identify differing subpopulations\footnote{\@Shannon cite} and many state based models provide for subpopulation based states in their analysis\footnote{\@Shannon cite}.
+This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g.
+age, sex) and many state based models provide for subpopulation-based states in their analysis \citep{rvachev1985,anderson1992,worby2015}.
 
@@
 Our `geom_aggregate` provides a rapid way to explore different subpopulations experience of the epidemic, and combines the ideas behind agents_to_aggregate for the SIR case, with `geom_path` and `coord_term` to visualize any number of groups epidemic trajectory in 3d simplex space using `ggplot2` and `ggtern` \citep{Wickham2016, Hamilton2018}.
-Visualization tools for SIR models were developed because (1) SIR models are the most common and basic state-based models\footnote{\@Shannon: cite} and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+Visualization tools for SIR models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-We first try to model the Hagelloch data with an SIR model (see Eq.
-\ref{eq:sir}).
-In our vignette, we show how to fit a stochastic model via maximum likelihood and simulate from the model with those best fit parameters.
-Our function \code{simulate_agents()} generates individual level data according to discrete multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.
-For example, the below code generates 100 simulations of an outbreak of a diseease with one initial infector in a population of $n= 188$ individuals.
+We first try to model the Hagelloch data with a stochastic SIR model, which we refer to as the 'simple SIR.'  In our vignette, we show how to fit this simple SIR model via maximum likelihood and simulate from the model with those best fit parameters.
+Our function \code{simulate_agents()} generates individual level data according to discrete time multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.
+For example, the below code generates 100 simulations of an outbreak of a disease with one initial infector in a population of $n= 188$ individuals.
 
@@
 ```
-The result of our simulation is the object \code{agents} which is a `r nrow(agents)` $\times$ `r ncol(agents)` which details the time of entry into the $S$, $I$, and $R$ states for a given simulation.
+The result of our simulation is the object \code{agents} which is a `r nrow(agents)` $\times$ `r ncol(agents)` data frame, which details the time of entry into the $S$, $I$, and $R$ states for a given simulation.
 Before we examine the results of this simple SIR model, we will also examine another, more sophisticated SIR model, this time from the package \pkg{EpiModel}.
@@
 
-The output of this model is \code{epimodel_sir}, an object of class `r class(epimodel_sir)`, which contains a plethora of modeling information.
+The output of this model is \code{epimodel_sir}, an object of class \code{`r class(epimodel_sir)`}, which contains a plethora of modeling information.
 We provide the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output and transform it into a tidy-style data frame.
@@
 
-# Helpful Rmd tricks
+# A. Appendix {-}
 
-## RMD Code formatting info
+## A.1 Proof of Theorem \ref{thm:sir-scale} {-}
 
-This is the Section \ref{sec:intro}. This template demonstrates some of the basic LaTeX that you need to know to create a JSS article.
-
-In general, don't use Markdown, but use the more precise LaTeX commands instead:
-
-* \proglang{Java}
-* \pkg{plyr}
-
-One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax).
-
-If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example.
-
-## RMD \proglang{R} code {short-title="R code" #r-code}
-
-Can be inserted in regular R markdown blocks.
-
-hags hags hags \cite{Neal2004}
-
-```{r}
-x <- 1:10
-x
-```
+\begin{proof}\label{proof:thm}
+\cite{Harko2014} provide an analytical solution for the Kermack and McKendrick equations (Eq.
+\eqref{eq:sir-ode}) by reparameterizing the ODEs so that $\mathcal{S}(u) = S(t)$, $\mathcal{I}(u) = S(t)$, and $\mathcal{R}(u) = R(t)$ for $0< u_T < 1$ with
+\begin{align}\label{eq:harko-odes}
+\mathcal{S}(u) &= S(0)u\\
+\mathcal{I}(u) &= N - R(0) + NR_0^{-1}\log u - S(0)u \nonumber\\
+\mathcal{R}(u) &= R(0) - NR_0^{-1} \log u, \nonumber
+\end{align}
+and $u$ and t are related by the following integral,
+\begin{align*}
+    t &= \int_{u}^1 \frac{N}{\beta \tau (N - R(0) + R_{0}^{-1} \log \tau - S(0)\tau)}d\tau \\
+    &= \int_{u}^1 \frac{1}{\beta f(S(0), R(0), N, R_0, \tau)} d \tau\\
+    &= \int_{u}^1 \frac{1}{\beta f(\tau)} d\tau,
+\end{align*}
+where we have made the denominator of the integral a function of $N$, the initial values, $R_0$, and $\tau$, which we further condense to $f(\tau)$ for brevity.
+Then for a given $t$ we want to find $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
+Or equivalently, for a fixed $u$ want to find $v$ such that  $\mathcal{S}_1(u) = \mathcal{S}_2(v)$ and then the corresponding $t$ and $s$ are given by
+\begin{align*}
+    t & = \int_{u}^1 \frac{1}{\beta_1 f(\tau)} d\tau \\
+    s & = \int_{v}^1 \frac{1}{\beta_2 f(\tau)} d\tau.
+\end{align*}
+Note that since the equations in Eq. \eqref{eq:harko-odes} are functions of the initial values and $R_0$, then $u = v$. We then can find a relation for $s$,
+    \begin{align*}
+    s & = \int_{u}^1 \frac{1}{\beta_2 f(\tau)} d\tau  \\
+    & = \int_{u}^1 \frac{1}{a\beta_1 f(\tau)} d\tau \\ 
+    &= \frac{1}{a}\int_{u}^1 \frac{1}{\beta_1 f(\tau)} d\tau \\
+    &= \frac{1}{a}t.
+\end{align*}
+\end{proof}

### 2021-01-15 jss/EpiCompare-paper/EpiCompare-paper.tex
mainly citations, cleaning up references, and added proof.

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath} \usepackage{amsthm}
 
@@
 example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
-influenza between 1.2-2, and modern measles outbreaks between 5-6
+influenza between 1.2-2, and modern measles outbreaks as big as 12
 \citep{midas2020,biggerstaff2014,namee2018}.
@@
 trajectories are shown in the \(state\) vs.~time plots in Figure
-\ref{fig:sir-ex}. At a glance, we may assume that Model 1 has a larger
-\(R_0\) than Model 2 because the peak of infection occurs more quickly
-than in Model 2. On the other hand, we may think Model 2 has a larger
-\(R_0\) because it has a slightly larger peak of infection than Model 1.
+\ref{fig:different-scales-standard}. At a glance, we may assume that
+Model 1 has a larger \(R_0\) than Model 2 because the peak of infection
+occurs more quickly than in Model 2. On the other hand, we may think
+Model 2 has a larger \(R_0\) because we may think the number of
+infections in that model has not yet peaked at time 15.
 
@@
 ternary plot in the time-invariant view, we immediately see a different
-story. In this time-invariant view in Fig. \ref{fig:ternary-ex}, the
-points seem to overlap and form the same trajectory. In other words, we
-can see there is something fundamentally linking these two different
-epidemics, and this fundamental link turns out to be \(R_0\).
+story. In this time-invariant view in Fig.
+\ref{fig:different-scales-tern}, the points seem to overlap and form the
+same trajectory. In other words, we can see there is something
+fundamentally linking these two different epidemics, and this
+fundamental link turns out to be \(R_0\).
 
-More formally, let two Kermack and McKendrick (see Eq.
-\eqref{eq:sir-ode}) SIR models be denoted \((S_1(t), I_1(t), R_1(t))\)
-and \((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume
-both models have initial values \((S(0), I(0), R(0))\). Let
+More formally, let two Kermack and McKendrick (see Eq.\eqref{eq:sir-ode}
+) SIR models be denoted \((S_1(t), I_1(t), R_1(t))\) and
+\((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume both
+models have initial values \((S(0), I(0), R(0))\). Let
 \(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
@@
 The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
-result from \cite{Harko2014} and is shown in detail in Appendix
-\ref{app:proof}. The consequence of Theorem \ref{thm:sir-scale} is that
+result from \cite{Harko2014} and is shown in detail in Proof
+\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
 for two SIR models that have the same initial percent of individuals in
@@
 than three states. The constraints in three dimensions that are met with
-the SIR model (that is
-\(\sum_{i=1}^3 (\text{number in state}(i)) = N(t)\)) actually represents
-a space of 3d simplices, and the ternary plot specifically represents
-these simplices, after scaling to examine such values as proportions
-between 0 and 1 (ternary plots are known as a 3d unit simplex due to
-it's scaling). This same scaling for larger models (i.e.~with more
-states) can be done onto different simplexes. In this package we present
-tools to help compare models (mostly through simulations). s tool
-compares these objects after projecting them into a one-dimension-fewer
-space through the simplexical structure of the data.
+the SIR model (there are three epidemic states and for each time step,
+the sum of total number of individuals in each state is the population
+size at that time\$) actually represents a space of 3d simplices, and
+the ternary plot specifically represents these simplices, after scaling
+to examine such values as proportions between 0 and 1 (ternary plots are
+known as a 3d unit simplex due to it's scaling). This same scaling for
+larger models (i.e.~with more states) can be done onto different
+simplexes. In this package we present tools to help compare models
+(mostly through simulations). s tool compares these objects after
+projecting them into a one-dimension-fewer space through the simplexical
+structure of the data.
 
@@
 epidemic (infection, recovery, etc.) as well as individual information
-like network structure, location, and-sub population information will be
-collected, whereas other data collections will focus
-population/sub-population counts of individuals in each epidemic
-state\footnote{\@Shannon include citations}. We focus on understanding
-the overall impact on the population/sub-populations we provide a
-function to transform information about each agent's start time of each
-state (e.g.~start of infection, etc). This transformation between from
-agent information to aggregate information is very useful for seeing the
-overall trend of the epidemic (and how it impacts different
-subpopulations). Our tools aim to allow the user to easily group agents
-and define new subpopulations to explore. This is important as many case
-studies have highlighted the usefulness to identify differing
-subpopulations\footnote{\@Shannon cite} and many state based models
-provide for subpopulation based states in their
-analysis\footnote{\@Shannon cite}.
+like network structure, location, and sub-population information will be
+collected, whereas other data collections will focus aggregate counts of
+individuals in each epidemic state. In fact, usually only the number of
+new infections at each time step is observed and the compartment totals
+are then imputed from that along with other information about the
+disease and the population of interest. We focus on understanding the
+overall impact of an outbreak at the aggregate/population level. We
+provide a function to transform information about each
+individual/agent's start time of each state (e.g.~start of infection,
+etc). This transformation between agent information to aggregate
+information is useful for seeing the overall trend of the epidemic (and
+how it impacts different subpopulations). Our tools aim to allow the
+user to easily group agents and define new subpopulations to explore.
+This is important as case studies highlight the usefulness of
+identifying differing subpopulations (e.g.~age, sex) and many state
+based models provide for subpopulation-based states in their analysis
+\citep{rvachev1985,anderson1992,worby2015}.
 
@@
 \citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
-were developed because (1) SIR models are the most common and basic
-state-based models\footnote{\@Shannon: cite} and (2) our simplex
-representation of these epidemics emphasizes a ``time-invarance''
-representation of the data (for a refresher see Section
-\ref{sec:time-invariant}).
+were developed because (1) SIR models are some of the most common and
+basic epidemic state-based models and (2) our simplex representation of
+these epidemics emphasizes a ``time-invarance'' representation of the
+data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 In this section, we highlight a number of the functionalities available
-in \kg{EpiCompare}. These functionalities include data cleaning,
+in \pkg{EpiCompare}. These functionalities include data cleaning,
 visualization, simulation, and comparison, in accordance with the data
@@
 
-We first try to model the Hagelloch data with an SIR model (see Eq.
-\ref{eq:sir}). In our vignette, we show how to fit a stochastic model
-via maximum likelihood and simulate from the model with those best fit
-parameters. Our function \code{simulate_agents()} generates individual
-level data according to discrete multinomial draws, which depend on the
-number of individuals in each state at the previous time step and a
-matrix of transition probabilities. For example, the below code
-generates 100 simulations of an outbreak of a diseease with one initial
-infector in a population of \(n= 188\) individuals.
+We first try to model the Hagelloch data with a stochastic SIR model,
+which we refer to as the `simple SIR.' In our vignette, we show how to
+fit this simple SIR model via maximum likelihood and simulate from the
+model with those best fit parameters. Our function
+\code{simulate_agents()} generates individual level data according to
+discrete time multinomial draws, which depend on the number of
+individuals in each state at the previous time step and a matrix of
+transition probabilities. For example, the below code generates 100
+simulations of an outbreak of a disease with one initial infector in a
+population of \(n= 188\) individuals.
 
@@
 The result of our simulation is the object \code{agents} which is a
-18800 \(\times\) 5 which details the time of entry into the \(S\),
-\(I\), and \(R\) states for a given simulation. Before we examine the
-results of this simple SIR model, we will also examine another, more
+18800 \(\times\) 5 data frame, which details the time of entry into the
+\(S\), \(I\), and \(R\) states for a given simulation. Before we examine
+the results of this simple SIR model, we will also examine another, more
 sophisticated SIR model, this time from the package \pkg{EpiModel}.
@@
 The output of this model is \code{epimodel_sir}, an object of class
-netsim, which contains a plethora of modeling information. We provide
-the function \code{fortify_aggregate()}, which can take objects from
-specialized classes of modeling output and transform it into a
+\code{netsim}, which contains a plethora of modeling information. We
+provide the function \code{fortify_aggregate()}, which can take objects
+from specialized classes of modeling output and transform it into a
 tidy-style data frame.
@@
 
-\hypertarget{helpful-rmd-tricks}{%
-\section{Helpful Rmd tricks}\label{helpful-rmd-tricks}}
+\hypertarget{a.-appendix}{%
+\section*{A. Appendix}\label{a.-appendix}}
+\addcontentsline{toc}{section}{A. Appendix}
 
-\hypertarget{rmd-code-formatting-info}{%
-\subsection{RMD Code formatting info}\label{rmd-code-formatting-info}}
+\hypertarget{a.1-proof-of-theorem}{%
+\subsection*{\texorpdfstring{A.1 Proof of Theorem
+\ref{thm:sir-scale}}{A.1 Proof of Theorem }}\label{a.1-proof-of-theorem}}
+\addcontentsline{toc}{subsection}{A.1 Proof of Theorem
+\ref{thm:sir-scale}}
 
-This is the Section \ref{sec:intro}. This template demonstrates some of
-the basic LaTeX that you need to know to create a JSS article.
-
-In general, don't use Markdown, but use the more precise LaTeX commands
-instead:
-
-\begin{itemize}
-\item
-  \proglang{Java}
-\item
-  \pkg{plyr}
-\end{itemize}
-
-One exception is inline code, which can be written inside a pair of
-backticks (i.e., using the Markdown syntax).
-
-If you want to use LaTeX commands in headers, you need to provide a
-\texttt{short-title} attribute. You can also provide a custom identifier
-if necessary. See the header of Section \ref{r-code} for example.
-
-\subsection[R code]{RMD \proglang{R} code}\label{r-code}
-
-Can be inserted in regular R markdown blocks.
-
-hags hags hags \cite{Neal2004}
-
-\begin{CodeChunk}
-\begin{CodeInput}
-R> x <- 1:10
-R> x
-\end{CodeInput}
-\begin{CodeOutput}
- [1]  1  2  3  4  5  6  7  8  9 10
-\end{CodeOutput}
-\end{CodeChunk}
+\begin{proof}\label{proof:thm}
+\cite{Harko2014} provide an analytical solution for the Kermack and McKendrick equations (Eq.
+\eqref{eq:sir-ode}) by reparameterizing the ODEs so that $\mathcal{S}(u) = S(t)$, $\mathcal{I}(u) = S(t)$, and $\mathcal{R}(u) = R(t)$ for $0< u_T < 1$ with
+\begin{align}\label{eq:harko-odes}
+\mathcal{S}(u) &= S(0)u\\
+\mathcal{I}(u) &= N - R(0) + NR_0^{-1}\log u - S(0)u \nonumber\\
+\mathcal{R}(u) &= R(0) - NR_0^{-1} \log u, \nonumber
+\end{align}
+and $u$ and t are related by the following integral,
+\begin{align*}
+    t &= \int_{u}^1 \frac{N}{\beta \tau (N - R(0) + R_{0}^{-1} \log \tau - S(0)\tau)}d\tau \\
+    &= \int_{u}^1 \frac{1}{\beta f(S(0), R(0), N, R_0, \tau)} d \tau\\
+    &= \int_{u}^1 \frac{1}{\beta f(\tau)} d\tau,
+\end{align*}
+where we have made the denominator of the integral a function of $N$, the initial values, $R_0$, and $\tau$, which we further condense to $f(\tau)$ for brevity.
+Then for a given $t$ we want to find $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
+Or equivalently, for a fixed $u$ want to find $v$ such that  $\mathcal{S}_1(u) = \mathcal{S}_2(v)$ and then the corresponding $t$ and $s$ are given by
+\begin{align*}
+    t & = \int_{u}^1 \frac{1}{\beta_1 f(\tau)} d\tau \\
+    s & = \int_{v}^1 \frac{1}{\beta_2 f(\tau)} d\tau.
+\end{align*}
+Note that since the equations in Eq. \eqref{eq:harko-odes} are functions of the initial values and $R_0$, then $u = v$. We then can find a relation for $s$,
+    \begin{align*}
+    s & = \int_{u}^1 \frac{1}{\beta_2 f(\tau)} d\tau  \\
+    & = \int_{u}^1 \frac{1}{a\beta_1 f(\tau)} d\tau \\ 
+    &= \frac{1}{a}\int_{u}^1 \frac{1}{\beta_1 f(\tau)} d\tau \\
+    &= \frac{1}{a}t.
+\end{align*}
+\end{proof}
 

### 2021-01-19 jss/EpiCompare-paper/EpiCompare-paper.tex
edits and updates to 2.2

@@
 example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
-influenza between 1.2-2, and modern measles outbreaks as big as 12
+influenza between 1.2-2, and modern measles outbreaks as large as 12
 \citep{midas2020,biggerstaff2014,namee2018}.
 
-Estimator(s) for \(R_0\) are dependent on the epidemic modeling
-framework, which consists of which states an individual can occupy
+Estimators for \(R_0\) are dependent on the epidemic modeling framework,
+which consists of which states an individual can occupy
 (e.g.~susceptible, infectious, recovered) and a description of how
@@
 generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
-\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over 40 days. Both epidemics
-have the same value of
+\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over the first 15 days. Both
+epidemics have the same value of
 \(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
@@
 \ref{fig:different-scales-tern}, the points seem to overlap and form the
-same trajectory. In other words, we can see there is something
-fundamentally linking these two different epidemics, and this
-fundamental link turns out to be \(R_0\).
+same trajectory. Now it seems to be that Model 2 is following the same
+trajectory as Model 1 but is not as far along in the infection process.
+We can see there is something fundamentally linking these two different
+epidemics, and this fundamental link turns out to be \(R_0\).
 
-More formally, let two Kermack and McKendrick (see Eq.\eqref{eq:sir-ode}
-) SIR models be denoted \((S_1(t), I_1(t), R_1(t))\) and
+More formally, let two Kermack and McKendrick SIR models(see
+Eq.\eqref{eq:sir-ode} ) be denoted \((S_1(t), I_1(t), R_1(t))\) and
 \((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume both
@@
 \begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
 \end{theorem}
@@
 
-But time-invariant analysis is also useful for epidemic models with more
-than three states. The constraints in three dimensions that are met with
-the SIR model (there are three epidemic states and for each time step,
-the sum of total number of individuals in each state is the population
-size at that time\$) actually represents a space of 3d simplices, and
-the ternary plot specifically represents these simplices, after scaling
-to examine such values as proportions between 0 and 1 (ternary plots are
-known as a 3d unit simplex due to it's scaling). This same scaling for
-larger models (i.e.~with more states) can be done onto different
-simplexes. In this package we present tools to help compare models
-(mostly through simulations). s tool compares these objects after
-projecting them into a one-dimension-fewer space through the simplexical
-structure of the data.
+Moreover, time-invariant analysis is also useful for epidemic models
+with more than three epidemic states. The constraints in three
+dimensions that are met with the SIR model (there are three epidemic
+states and for each time step, the sum of total number of individuals in
+each state is the population size at that time) represents a space of 3d
+simplices, and the ternary plot specifically represents these simplices,
+after scaling the values in each state as proportions of the total
+population at that time (ternary plots are known as a 3d unit simplex
+due to it's scaling). This same scaling for larger models (i.e.~with
+more states) can be done onto different simplexes. In this package we
+present tools to help compare models (mostly through simulations).
+\pkg{EpiCompare}'s tools permits comparisons of \(d\)-dimensional
+objects by projecting them into (\(d-1\))-dimensional space.
 
-In \pkg{EpiCompare} we allow for the comparison epidemics in these
-higher dimensional spaces by first projecting onto these simplexes. Even
-though higher dimensional models may not be able to visualized, we
-provide multiple tools to aid in the comparison of models and epidemics.
-The first of which uses multiple simulations under specific model
-parameters to assess the bairabilty of the model fit. In
-\pkg{EpiCompare} we provide ways to create prediction regions for a true
-epidemic under the model assumptions using these simulations. These
-regions require representing multi-dimensional structures for functions
-to completely contain epidemics, and treat these simulations and
-epidmeics as filamental objects. We extend off of papers like
-\citet{Dalmasso2019a} to create these bands. These high dimensional
-bands allow the user to assess if the true epidemic is within the band
-(there-by assessing the model's representation of the epidemic), and
-also compare who different models are from each other through distances
-that compare sets. We recommend using the Hausdorff distance to compare
-such sets as it captures how much bigger the sets would have to expand
-to cover each other, and is defined mathematically as \[
-d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
+Even though higher dimensional models may not be able to visualized
+simply, we provide a number of tools to aid in the comparison of models
+and epidemics. The first of which uses multiple simulations under
+specific model parameters to assess the \textcolor{orange}{variability?}
+bairabilty of the model fit. In \pkg{EpiCompare}, we provide ways to
+create prediction regions for a `true' epidemic under a fully specified
+model. These regions require representing multi-dimensional structures
+for functions to completely contain epidemics. \pkg{EpiCompare} treats
+these simulations and epidemics as \textit{filamental} objects --
+\textcolor{orange}{need a description for filamental, Ben feel free to change}
+where not only the outbreak trajectory points are important but also
+their form which can be considered a function.
+\textcolor{orange}{could be useful to have a picture/example here?
+like  three curves and 2 metrics one where closer under a pointwise distance and the other closer in filamental distance?
+I'll sketch up an idea unless you have something in your proposal already}
+
+We extend off of papers like \citet{Dalmasso2019a} to create these
+bands. These high dimensional bands allow the user to assess if the true
+epidemic is within the band (thereby assessing the model's
+representation of the epidemic). With these bands, we can compare
+\textit{models} to \textit{models} by assessing the distance between the
+two bands of model representation. We recommend using the Hausdorff
+distance to compare two bands to one another, as this distance captures
+how much bigger the sets would have to expand to cover each other. The
+Hausdorff distance is defined mathematically as \[
+d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y),\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
 \]

### 2021-01-19 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Update to section 3.

@@
   - \usepackage{xcolor}
+  - \usepackage{rotating}
 ---
@@
 $R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.
-For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks as big as 12 \citep{midas2020,biggerstaff2014,namee2018}.
+For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks as large as 12 \citep{midas2020,biggerstaff2014,namee2018}.
 
-Estimator(s) for $R_0$ are dependent on the epidemic modeling framework, which consists of which states an individual can occupy (e.g.
+Estimators for $R_0$ are dependent on the epidemic modeling framework, which consists of which states an individual can occupy (e.g.
 susceptible, infectious, recovered) and a description of how individuals move from one state to the next over time (see \cite{hethcote1994}).
@@
 $time$ plots, $R_0$ is difficult to visualize, especially with respect from one epidemic to another.
-For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = 0.3$ and $\gamma_1 = 0.15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = 0.24$ and $\gamma_1 = 0.12$ over 40 days.
+For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = 0.3$ and $\gamma_1 = 0.15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = 0.24$ and $\gamma_1 = 0.12$ over the first 15 days.
 Both epidemics have the same value of $R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2$.
@@
 \ref{fig:different-scales-tern}, the points seem to overlap and form the same trajectory.
-In other words, we can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
+Now it seems to be that Model 2 is following the same trajectory as Model 1 but is not as far along in the infection process.
+We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
 
-More formally, let two Kermack and McKendrick (see Eq.\eqref{eq:sir-ode} ) SIR models be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
+More formally, let two Kermack and McKendrick  SIR models(see Eq.\eqref{eq:sir-ode} )  be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
 Assume both models have initial values $(S(0), I(0), R(0))$.
@@
 
-But time-invariant analysis is also useful for epidemic models with more than three states.
-The constraints in three dimensions that are met with the SIR model (there are three epidemic states and for each time step, the sum of total number of individuals in each state is the population size at that time$) actually represents a space of 3d simplices, and the ternary plot specifically represents these simplices, after scaling to examine such values as proportions between 0 and 1 (ternary plots are known as a 3d unit simplex due to it's scaling).
+Moreover, time-invariant analysis is also useful for epidemic models with more than three epidemic states.
+The constraints in three dimensions that are met with the SIR model (there are three epidemic states and for each time step, the sum of total number of individuals in each state is the population size at that time) represents a space of 3d simplices, and the ternary plot specifically represents these simplices, after scaling the values in each state as proportions of the total population at that time (ternary plots are known as a 3d unit simplex due to it's scaling).
 This same scaling for larger models (i.e.
@@
 In this package we present tools to help compare models (mostly through simulations).
-s tool compares these objects after projecting them into a one-dimension-fewer space through the simplexical structure of the data.
+\pkg{EpiCompare}'s tools permits comparisons of $d$-dimensional objects by projecting them into ($d-1$)-dimensional space.
 
-In \pkg{EpiCompare} we allow for the comparison epidemics in these higher dimensional spaces by first projecting onto these simplexes.
-Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics.
-The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit.
-In \pkg{EpiCompare} we provide ways to create prediction regions for a true epidemic under the model assumptions using these simulations.
-These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects.
+ Even though higher dimensional models may not be able to visualized simply, we provide a number of tools to aid in the comparison of models and epidemics.
+The first of which uses multiple simulations under specific model parameters to assess the \shannon{variability?} bairabilty of the model fit.
+In \pkg{EpiCompare}, we provide ways to create prediction regions for a `true' epidemic under a fully specified model.
+These regions require representing multi-dimensional structures for functions to completely contain epidemics.
+\pkg{EpiCompare} treats these simulations and epidemics as \textit{filamental} objects -- \shannon{need a description for filamental, Ben feel free to change} where not only the outbreak trajectory points are important but also their form which can be considered a function.
+\shannon{could be useful to have a picture/example here?
+like  three curves and 2 metrics one where closer under a pointwise distance and the other closer in filamental distance?
+I'll sketch up an idea unless you have something in your proposal already}
+
  We extend off of papers like \citet{Dalmasso2019a} to create these bands.
-These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets.
-We recommend using the Hausdorff distance to compare such sets as it captures how much bigger the sets would have to expand to cover each other, and is defined mathematically as
+These high dimensional bands allow the user to assess if the true epidemic is within the band (thereby assessing the model's representation of the epidemic).
+With these bands, we can compare \textit{models} to \textit{models} by assessing the distance between the two bands of model representation.
+We recommend using the Hausdorff distance to compare two bands to one another, as this distance captures how much bigger the sets would have to expand to cover each other.
+The Hausdorff distance is defined mathematically as
 \[
-d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
+d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y),\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
 \]
 
-# Package tool overview {short-title="Package overview" #sec:overview}
+# Overview of \pkg{EpiCompare} {short-title="Package overview" #sec:overview}
 
@@
 
-In this section we will present the tools in this package and how they aid in the  data analysis pipeline.
-We present a modification of Figure \ref{fig:pipeline} from the Section \ref{sec:intro} with Figure \ref{fig:pipeline2} with uses of our package.
-We recommend the package user keep this image around as a reference.
-All user functions are aimed to be as friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
-There are 2 different ways \pkg{EpiCompare} can be incorporated in the data analysis pipeline for epidemics, either at the very beginning when pre-processing data and visualizing raw data, or after modeling has been done and a generative model is fit.
+In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.
+In Fig.
+\ref{fig:pipeline2}, we illustrate how our package fits into the data analysis pipeline introduced in Fig.
+\ref{fig:pipeline}.
+All front-facing functions are aimed to be as user-friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
+Although users can typically incorporate \pkg{EpiCompare} into any step in the data analysis process, there are two primary points of entry.
+The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation.
 Figure \ref{fig:pipeline2} captures these different paths, and we will highlight both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
 
-\begin{figure}[!ht]
+\begin{sidewaysfigure}[!ht]
     \centering
-    \includegraphics[width = 1\textwidth]{images/pipeline2.png}
+    \includegraphics[width = 1\textwidth]{images/pipeline2_v2.pdf}
     \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
     \label{fig:pipeline2}
-\end{figure}
+\end{sidewaysfigure}
 
-## Data Preprocessing
+**Data Pre-processing**
 
@@
 There are multiple ways to collect the epidemiological data.
-Sometimes individual records with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information will be collected, whereas other data collections will focus aggregate counts of individuals in each epidemic state.
+Sometimes individual records with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information will be collected, whereas other data collections will focus on aggregate counts of individuals in each epidemic state.
 In fact, usually only the number of new infections at each time step is observed and the compartment totals are then imputed from that along with other information about the disease and the population of interest.
 We focus on understanding the overall impact of an outbreak at the aggregate/population level.
-We provide a function to transform information about each individual/agent’s start time of each state (e.g.
-start of infection, etc).
-This transformation between agent information to aggregate information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
-Our tools aim to allow the user to easily group agents and define new subpopulations to explore.
+In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g.
+start of infection, start of recovery, etc).
+This transformation between \textit{agent} information to \textit{aggregate} information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
+Our tools allow the user to easily group agents and define new subpopulations to explore.
 This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g.
@@
 
-We provide a "tidyverse"-styled function, `agents_to_aggregate` to transform agent information into aggregate state information,  As a "tidy" function, our function `agents_to_aggregate` allows the user to identify any subpopulations by first doing `group_by` from `dplyr`, and also allows for infinite epidemic states - which can be for models with assumptions of sub population based states (e.g.
-youth infections), as well as indicators for death and birth dates.
-Our function, `agents_to_aggregate` allows for states to be skipped - captured with `NA` values and multiple starts of a state - (e.g.
-multiple infections) as long as the original data is stored in columns that are constrained to ordering ideas\footnote{make this more clear?}.
-Currently this function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can make the time steps more (or less) .
-`agents_to_aggregate` then returns a standardized count for each state for each time point.
-We see `agents_to_aggregate` as a really powerful tool, but also very useful to quickly transform agents information to be used in aggregate state-based models and compare state-based models to observed data.
+We provide a "tidyverse"-styled function, `agents_to_aggregate()` to transform agent information into aggregate state information.
+As a "tidy" function, our function `agents_to_aggregate()` allows the user to identify and analyze subpopulations by first executing a `group_by()` from \pkg{dplyr}.
+The function `agents_to_aggregate` is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates.
+Our function, `agents_to_aggregate()` does not require agents to pass through every state listed and agents are allowed to start in any state.
+This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g.
+hours or minutes).
+The function `agents_to_aggregate()` returns the total number of individuals in each state at each time point.
+We view `agents_to_aggregate()` as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.
 
-## EDA
+**EDA**
 
-With raw data, getting to know our data now-a-days very frequently means figuring out good combinations of visualizations, numerical summaries and subsetting.
-An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but we’ve also developed tools to rapidly explore your data if your have a 3 unique states models (like the SIR model).
-Our `geom_aggregate` provides a rapid way to explore different subpopulations experience of the epidemic, and combines the ideas behind agents_to_aggregate for the SIR case, with `geom_path` and `coord_term` to visualize any number of groups epidemic trajectory in 3d simplex space using `ggplot2` and `ggtern` \citep{Wickham2016, Hamilton2018}.
-Visualization tools for SIR models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+With raw data,  exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings.
+An expert coder can start from `agents_to_aggregate()` to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).
+Our `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens.
+It combines the ideas behind `agents_to_aggregate()` with \pkg{ggplot2}'s `geom_path()` and \pkg{ggtern}'s `coord_tern()` to visualize any number of groups epidemic trajectory in 3d simplex \citep{Wickham2016, Hamilton2018}.
+We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).
 
-## Model Fitting and Simulations
+**Model Fitting and Simulations**
 
-Although this package does not focus on estimating a model for the data, we do provide some power functions for simulation of basic state models with Bernoulli approximation of ODE state models encoded in simulate_agents and simulate_SIR_agents (which can naturally be combined with agents_to_aggregate ).
-**Description of these models, reference to Shannon R0 paper, etc...**
+Although this package does not focus on estimating a model for the data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models with Bernoulli/Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next.
+These functions can be naturally combined with `agents_to_aggregate()` to proceed in the pipeline.
+The function `simulate_SIR_agents()` simulates an SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
+This function allows for easy access to SIR model analysis and comparison.
+Beyond SIR models, the function `simulate_agents()` takes as input a user-specified transition matrix and other epidemic parameters to allow the user to create simulations of an outbreak for \textit{any} number of states and any number of transitions among them.
+This allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
+For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.
 
-## Post-processing
+**Post-processing**
 
-...
+Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.
+For example, comma separated value files (`.csv`) are a  standard way to share information within tables.
+However, model output is often far more complicated than what a traditional `.csv` would allow.
+As a result, a number of epidemic modeling packages return a special class, specific to their modeling.
+The special classes often contain a plethora of information from residuals, model diagnostics, input parameters, and more.
+While incredibly useful, these special classes can be difficult for novice coders to navigate.
 
-## Comparions and Assessment
+To this end, we have adapted a series of `fortify`-style functions, called `fortify_aggregate()` which transform output from packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
+These fortify functions have output that is consistent with that of `agents_to_aggregate()`.
 
-...
+\shannon{Here Ben talks about filament compression and tidy_dist mats, etc}
+
+**Comparisons and Assessment**
+
+Comparison and assessment of model fit or comparisons of one model to another model can be performed in a variety of ways including mean square error, AIC, plots, and more.
+Perhaps the most useful tool \pkg{EpiCompare} has to offer to the expert, for comparison and assessment of models, is in its post-processing tools which create a standard output.
+It is then a matter of writing a script or function made for that standard output to assess the results from multiple models in the way the user desires.
+
+However, for those who like more concrete tools, \pkg{EpiCompare} offers functions to compare prediction regions to one another including `geom_prediction_band()` (which plots the region), and `create_{convex_hull,delta_ball}_structure()`(which returns the `R` output for the given structure), and `contained()` (which allows the user to determine if one set of points is contained in a prediction band).
+Additionally, we offer ways to determine if model outputs are compatible with one another, that is how extreme one output is to another.
+\shannon{Ben says something about distance}
 

### 2021-01-19 jss/EpiCompare-paper/EpiCompare-paper.tex
Update to section 3.

@@
 \usepackage{xcolor}
+\usepackage{rotating}
 
@@
 
-\section[Package overview]{Package tool overview}\label{sec:overview}
+\section[Package overview]{Overview of
+\pkg{EpiCompare}}\label{sec:overview}
 
@@
 
-In this section we will present the tools in this package and how they
-aid in the data analysis pipeline. We present a modification of Figure
-\ref{fig:pipeline} from the Section \ref{sec:intro} with Figure
-\ref{fig:pipeline2} with uses of our package. We recommend the package
-user keep this image around as a reference. All user functions are aimed
-to be as friendly as possible, and we focus on providing the user
-``tidyverse'' style functions, that encourage piping and also follow
-clear verb naming schemes \citep{Wickham2019}. There are 2 different
-ways \pkg{EpiCompare} can be incorporated in the data analysis pipeline
-for epidemics, either at the very beginning when pre-processing data and
-visualizing raw data, or after modeling has been done and a generative
-model is fit. Figure \ref{fig:pipeline2} captures these different paths,
-and we will highlight both approaches and how to leverage
-\pkg{EpiCompare} in the subsections below.
+In this section, we present the tools implemented in \pkg{EpiCompare}
+and explain how they aid in the data analysis pipeline. In Fig.
+\ref{fig:pipeline2}, we illustrate how our package fits into the data
+analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
+front-facing functions are aimed to be as user-friendly as possible, and
+we focus on providing the user ``tidyverse'' style functions, that
+encourage piping and also follow clear verb naming schemes
+\citep{Wickham2019}. Although users can typically incorporate
+\pkg{EpiCompare} into any step in the data analysis process, there are
+two primary points of entry. The first point of entry is the very
+beginning with pre-processing and visualizing raw data, and the second
+point of entry is after modeling and simulation. Figure
+\ref{fig:pipeline2} captures these different paths, and we will
+highlight both approaches and how to leverage \pkg{EpiCompare} in the
+subsections below.
 
-\begin{figure}[!ht]
+\begin{sidewaysfigure}[!ht]
     \centering
-    \includegraphics[width = 1\textwidth]{images/pipeline2.png}
+    \includegraphics[width = 1\textwidth]{images/pipeline2_v2.pdf}
     \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
     \label{fig:pipeline2}
-\end{figure}
+\end{sidewaysfigure}
 
-\hypertarget{data-preprocessing}{%
-\subsection{Data Preprocessing}\label{data-preprocessing}}
+\textbf{Data Pre-processing}
 
@@
 like network structure, location, and sub-population information will be
-collected, whereas other data collections will focus aggregate counts of
-individuals in each epidemic state. In fact, usually only the number of
-new infections at each time step is observed and the compartment totals
-are then imputed from that along with other information about the
+collected, whereas other data collections will focus on aggregate counts
+of individuals in each epidemic state. In fact, usually only the number
+of new infections at each time step is observed and the compartment
+totals are then imputed from that along with other information about the
 disease and the population of interest. We focus on understanding the
-overall impact of an outbreak at the aggregate/population level. We
-provide a function to transform information about each
-individual/agent's start time of each state (e.g.~start of infection,
-etc). This transformation between agent information to aggregate
-information is useful for seeing the overall trend of the epidemic (and
-how it impacts different subpopulations). Our tools aim to allow the
-user to easily group agents and define new subpopulations to explore.
-This is important as case studies highlight the usefulness of
-identifying differing subpopulations (e.g.~age, sex) and many state
-based models provide for subpopulation-based states in their analysis
+overall impact of an outbreak at the aggregate/population level. In
+\pkg{EpiCompare}, we provide a function to transform information about
+each individual/agent's initial time of entry into each state
+(e.g.~start of infection, start of recovery, etc). This transformation
+between \textit{agent} information to \textit{aggregate} information is
+useful for seeing the overall trend of the epidemic (and how it impacts
+different subpopulations). Our tools allow the user to easily group
+agents and define new subpopulations to explore. This is important as
+case studies highlight the usefulness of identifying differing
+subpopulations (e.g.~age, sex) and many state based models provide for
+subpopulation-based states in their analysis
 \citep{rvachev1985,anderson1992,worby2015}.
@@
 We provide a ``tidyverse''-styled function,
-\texttt{agents\_to\_aggregate} to transform agent information into
-aggregate state information, As a ``tidy'' function, our function
-\texttt{agents\_to\_aggregate} allows the user to identify any
-subpopulations by first doing \texttt{group\_by} from \texttt{dplyr},
-and also allows for infinite epidemic states - which can be for models
-with assumptions of sub population based states (e.g.~youth infections),
-as well as indicators for death and birth dates. Our function,
-\texttt{agents\_to\_aggregate} allows for states to be skipped -
-captured with \texttt{NA} values and multiple starts of a state -
-(e.g.~multiple infections) as long as the original data is stored in
-columns that are constrained to ordering
-ideas\footnote{make this more clear?}. Currently this function is
-constrained to integer time steps (for example days), but
-transformations (linear or otherwise) of the time columns can make the
-time steps more (or less) . \texttt{agents\_to\_aggregate} then returns
-a standardized count for each state for each time point. We see
-\texttt{agents\_to\_aggregate} as a really powerful tool, but also very
-useful to quickly transform agents information to be used in aggregate
-state-based models and compare state-based models to observed data.
+\texttt{agents\_to\_aggregate()} to transform agent information into
+aggregate state information. As a ``tidy'' function, our function
+\texttt{agents\_to\_aggregate()} allows the user to identify and analyze
+subpopulations by first executing a \texttt{group\_by()} from
+\pkg{dplyr}. The function \texttt{agents\_to\_aggregate} is flexible in
+that there is no limit on the amount of permitted epidemiological states
+so users can incorporate information from compartments such as
+``Exposed'', ``iMmune'', or ``Hospitalized'' groups, for instance, and
+also permits indicators for death/exit and birth/entry dates. Our
+function, \texttt{agents\_to\_aggregate()} does not require agents to
+pass through every state listed and agents are allowed to start in any
+state. This function is constrained to integer time steps (for example
+days), but transformations (linear or otherwise) of the time columns can
+account for finer grained time steps (e.g.~hours or minutes). The
+function \texttt{agents\_to\_aggregate()} returns the total number of
+individuals in each state at each time point. We view
+\texttt{agents\_to\_aggregate()} as powerful tool in the step of
+pre-processing data, one that quickly and simply transforms individual
+level information into an aggregate view that can be processed and
+analyzed in a simpler manner.
 
-\hypertarget{eda}{%
-\subsection{EDA}\label{eda}}
+\textbf{EDA}
 
-With raw data, getting to know our data now-a-days very frequently means
-figuring out good combinations of visualizations, numerical summaries
-and subsetting. An expert coder can start from
-\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
-we've also developed tools to rapidly explore your data if your have a 3
-unique states models (like the SIR model). Our \texttt{geom\_aggregate}
-provides a rapid way to explore different subpopulations experience of
-the epidemic, and combines the ideas behind agents\_to\_aggregate for
-the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
-visualize any number of groups epidemic trajectory in 3d simplex space
-using \texttt{ggplot2} and \texttt{ggtern}
-\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
-were developed because (1) SIR models are some of the most common and
-basic epidemic state-based models and (2) our simplex representation of
-these epidemics emphasizes a ``time-invarance'' representation of the
-data (for a refresher see Section \ref{sec:time-invariant}).
+With raw data, exploratory analysis frequently means figuring out good
+combinations of visualizations, numerical summaries, and groupings. An
+expert coder can start from \texttt{agents\_to\_aggregate()} to
+successfully accomplish EDA in many ways, but we have also developed
+tools to allow a novice coder to rapidly explore data, as long as there
+three unique epidemiological states (like the SIR model). Our
+\texttt{geom\_aggregate()} provides a rapid way to explore different
+subpopulations' experiences of the epidemic in the ternary lens. It
+combines the ideas behind \texttt{agents\_to\_aggregate()} with
+\pkg{ggplot2}'s \texttt{geom\_path()} and \pkg{ggtern}'s
+\texttt{coord\_tern()} to visualize any number of groups epidemic
+trajectory in 3d simplex \citep{Wickham2016, Hamilton2018}. We developed
+these out-of-the-box visualization tools for SIR models because (1) SIR
+models are some of the most common and basic epidemic state-based models
+and (2) our simplex representation of these epidemics emphasizes a
+``time-invarance'' representation of the data (see Section
+\ref{sec:time-invariant}).
 
-\hypertarget{model-fitting-and-simulations}{%
-\subsection{Model Fitting and
-Simulations}\label{model-fitting-and-simulations}}
+\textbf{Model Fitting and Simulations}
 
 Although this package does not focus on estimating a model for the data,
-we do provide some power functions for simulation of basic state models
-with Bernoulli approximation of ODE state models encoded in
-simulate\_agents and simulate\_SIR\_agents (which can naturally be
-combined with agents\_to\_aggregate ). \textbf{Description of these
-models, reference to Shannon R0 paper, etc\ldots{}}
+we do provide some flexible functions for simulation of basic
+discrete-time epidemic-state models with Bernoulli/Multinomial
+transitions which can be used to (noisily) approximate ODEs which
+describe transitions from one state to the next. These functions can be
+naturally combined with \texttt{agents\_to\_aggregate()} to proceed in
+the pipeline. The function \texttt{simulate\_SIR\_agents()} simulates an
+SIR epidemic with user inputs for the number of simulations, the initial
+number in each state, the infection and recovery parameters
+\((\beta, \gamma\)), and the total number of discrete time steps. This
+function allows for easy access to SIR model analysis and comparison.
+Beyond SIR models, the function \texttt{simulate\_agents()} takes as
+input a user-specified transition matrix and other epidemic parameters
+to allow the user to create simulations of an outbreak for \textit{any}
+number of states and any number of transitions among them. This allows
+for users to explore the space of models in an intuitive way without
+getting bogged down by too much mathematical detail. For consistency, we
+have made output from \texttt{simulate\_agents()} and
+\texttt{simulate\_SIR\_agents()} compatible with
+\texttt{agents\_to\_aggregate()} so aggregate information may easily be
+accessed.
 
-\hypertarget{post-processing}{%
-\subsection{Post-processing}\label{post-processing}}
+\textbf{Post-processing}
 
-\ldots{}
+Post-processing of modeling and simulation consists of making summary
+statistics, plots, tables, and other ways to disseminate information to
+the public. For example, comma separated value files (\texttt{.csv}) are
+a standard way to share information within tables. However, model output
+is often far more complicated than what a traditional \texttt{.csv}
+would allow. As a result, a number of epidemic modeling packages return
+a special class, specific to their modeling. The special classes often
+contain a plethora of information from residuals, model diagnostics,
+input parameters, and more. While incredibly useful, these special
+classes can be difficult for novice coders to navigate.
 
-\hypertarget{comparions-and-assessment}{%
-\subsection{Comparions and Assessment}\label{comparions-and-assessment}}
+To this end, we have adapted a series of \texttt{fortify}-style
+functions, called \texttt{fortify\_aggregate()} which transform output
+from packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data
+frames which contain information about the total number of individuals
+in each state at a given time, for a given simulation. These fortify
+functions have output that is consistent with that of
+\texttt{agents\_to\_aggregate()}.
 
-\ldots{}
+\textbf{Comparisons and Assessment}
+
+Comparison and assessment of model fit or comparisons of one model to
+another model can be performed in a variety of ways including mean
+square error, AIC, plots, and more. Perhaps the most useful tool
+\pkg{EpiCompare} has to offer to the expert, for comparison and
+assessment of models, is in its post-processing tools which create a
+standard output. It is then a matter of writing a script or function
+made for that standard output to assess the results from multiple models
+in the way the user desires.
+
+However, for those who like more concrete tools, \pkg{EpiCompare} offers
+functions to compare prediction regions to one another including
+\texttt{geom\_prediction\_band()} (which plots the region), and
+\texttt{create\_\{convex\_hull,delta\_ball\}\_structure()}(which returns
+the \texttt{R} output for the given structure), and \texttt{contained()}
+(which allows the user to determine if one set of points is contained in
+a prediction band). Additionally, we offer ways to determine if model
+outputs are compatible with one another, that is how extreme one output
+is to another. \textcolor{orange}{Ben says something about distance}
 

### 2021-01-20 jss/EpiCompare-paper/EpiCompare-paper.Rmd

--- 2021-01-15 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
 In this package we present tools to help compare models (mostly through simulations).
-s tool compares these objects after projecting them into a one-dimension-fewer space through the simplexical structure of the data.
+\pkg{EpiCompare}'s tools permits comparisons of $d$-dimensional objects by projecting them into ($d-1$)-dimensional space.
 
-In \pkg{EpiCompare} we allow for the comparison epidemics in these higher dimensional spaces by first projecting onto these simplexes.
-Even though higher dimensional models may not be able to visualized, we provide multiple tools to aid in the comparison of models and epidemics.
-The first of which uses multiple simulations under specific model parameters to assess the bairabilty of the model fit.
-In \pkg{EpiCompare} we provide ways to create prediction regions for a true epidemic under the model assumptions using these simulations.
-These regions require representing multi-dimensional structures for functions to completely contain epidemics, and treat these simulations and epidmeics as filamental objects.
+ Even though higher dimensional models may not be able to visualized simply, we provide a number of tools to aid in the comparison of models and epidemics.
+The first of which uses multiple simulations under specific model parameters to assess the \shannon{variability?} bairabilty of the model fit.
+In \pkg{EpiCompare}, we provide ways to create prediction regions for a `true' epidemic under a fully specified model.
+These regions require representing multi-dimensional structures for functions to completely contain epidemics.
+\pkg{EpiCompare} treats these simulations and epidemics as \textit{filamental} objects -- \shannon{need a description for filamental, Ben feel free to change} where not only the outbreak trajectory points are important but also their form which can be considered a function.
+\shannon{could be useful to have a picture/example here?
+like  three curves and 2 metrics one where closer under a pointwise distance and the other closer in filamental distance?
+I'll sketch up an idea unless you have something in your proposal already}
+
  We extend off of papers like \citet{Dalmasso2019a} to create these bands.
-These high dimensional bands allow the user to assess if the true epidemic is within the band (there-by assessing the model's representation of the epidemic), and also compare who different models are from each other through distances that compare sets.
-We recommend using the Hausdorff distance to compare such sets as it captures how much bigger the sets would have to expand to cover each other, and is defined mathematically as
+These high dimensional bands allow the user to assess if the true epidemic is within the band (thereby assessing the model's representation of the epidemic).
+With these bands, we can compare \textit{models} to \textit{models} by assessing the distance between the two bands of model representation.
+We recommend using the Hausdorff distance to compare two bands to one another, as this distance captures how much bigger the sets would have to expand to cover each other.
+The Hausdorff distance is defined mathematically as
 \[
-d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y) \;,\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
+d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y),\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
 \]

### 2021-01-20 README.Rmd
rcpp speed up, clean up of arrangeGrob + grid.arrange, and additional comment in readme

@@
 We currently compile our `pkgdown` site on our own computers and then push - and will continue to do so due to complications with `reticulate` and `python` packages on the virtual machines in github actions.
+9.
+We've speed up some of the interval functions with `Rcpp`.
+The setup to use this functions was slightly complex, we drafted code using `cppFunction('...')` and then needed to pipe them over to the package setting.
+Approaches required (1) `usethis::use_rcpp()`, then copying C++ code into `src/code.cpp` (needed to make sure we had [`// [[Rcpp::export]]`](http://adv-r.had.co.nz/Rcpp.html#rcpp-sugar) above each function.
+For make sure the functions would be compiled and available we needed to update the package documentation with [2 roxygen tags](https://r-pkgs.org/src.html#cpp):
+
+```
+#' @useDynLib your-package-name
+#' @importFrom Rcpp sourceCpp
+```
 

### 2021-01-20 R/aaa.R

@@ -94,7 +94,7 @@ print.ggplot <- function(x, newpage = is.null(vp), vp = NULL, ...){
 arrangeGrob = function (..., grobs = list(...), layout_matrix, vp = NULL, name = "arrange", 
                         as.table = TRUE, respect = FALSE, clip = "off", nrow = NULL, 
                         ncol = NULL, widths = NULL, heights = NULL, top = NULL, bottom = NULL, 
-                        left = NULL, right = NULL, padding = unit(0.5, "line")) 
+                        left = NULL, right = NULL, padding = grid::unit(0.5, "line")) ## BL (grid::)
 {
   n <- length(grobs)
   if (!is.null(ncol) && !is.null(widths)) {
@@ -139,7 +139,8 @@ arrangeGrob = function (..., grobs = list(...), layout_matrix, vp = NULL, name =
   if (any(inherit.trellis)) {
     stopifnot(requireNamespace("lattice", quietly = TRUE))
     toconv <- which(inherit.trellis)
-    grobs[toconv] <- lapply(grobs[toconv], ggint$latticeGrob)
+    latticeGrob <- imports_hidden_from("gridExtra", "latticeGrob")
+    grobs[toconv] <- lapply(grobs[toconv], latticeGrob)
   }
   if (missing(layout_matrix)) {
     positions <- expand.grid(t = seq_len(nrow), l = seq_len(ncol))
@@ -162,53 +163,53 @@ arrangeGrob = function (..., grobs = list(...), layout_matrix, vp = NULL, name =
     nrow <- max(positions$b)
   }
   if (is.null(widths)) 
-    widths <- unit(rep(1, ncol), "null")
+    widths <- grid::unit(rep(1, ncol), "null") ## BL (grid::)
   if (is.null(heights)) 
-    heights <- unit(rep(1, nrow), "null")
+    heights <- grid::unit(rep(1, nrow), "null") ## BL (grid::)
   if (!grid::is.unit(widths)) ## BL (grid::)
-    widths <- unit(widths, "null")
+    widths <- grid::unit(widths, "null") ## BL (grid::)
   if (!grid::is.unit(heights)) ## BL (grid::)
-    heights <- unit(heights, "null")
+    heights <- grid::unit(heights, "null") ## BL (grid::)
   gt <- gtable::gtable(name = name, respect = respect, heights = heights, ##NH
[…]

### 2021-01-20 src/code.cpp

@@ -1,0 +1,51 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+// [[Rcpp::export]]
+double l2filamentdist(NumericMatrix m1, NumericMatrix m2) {
+  int nrow1 = m1.nrow();
+  int ncol = m1.ncol();
+  
+  if (ncol != m2.ncol()) {
+    throw std::runtime_error("Incompatible number of dimensions");
+  }
+  
+  double out = 0;
+  
+  for (int r1 = 0; r1 < nrow1; r1++) {
+    double total = 0;
+    for (int c12 = 0; c12 < ncol; c12++) {
+      total += pow(m1(r1, c12) - m2(r1, c12), 2);
+    }
+    out += total;
+  }
+  
+  return out;
+}
+
+
+// [[Rcpp::export]]
+List distalongpath(NumericMatrix m1){
+  int nrow = m1.nrow();
+  int ncol = m1.ncol();
+  
+  List out(2);
+  NumericVector out1(nrow-1);
+  NumericMatrix out2(nrow-1,ncol);
+  
+  double innerdistpart;
+  
+  for(int r1 = 0; r1 < nrow - 1; r1++) {
+    double innerdist = 0;
+    for (int c1 = 0; c1 < ncol; c1++) {
+      innerdistpart = m1(r1+1, c1) - m1(r1,c1);
+      out2(r1,c1) = innerdistpart;
+      innerdist += pow(innerdistpart, 2);
+    }
+    out1(r1) = sqrt(innerdist);
+  }
+  out(0) = out1;
+  out(1) = out2;
+  
+  return out;
+}

### 2021-01-26 jss/EpiCompare-paper/EpiCompare-paper.Rmd
rcpp speed up, clean up of arrangeGrob + grid.arrange, and additional comment in readme
section 3
epicompare section 2 rewrite + update of pipeline2 image

--- 2021-01-19 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
 abstract: >
-  We present \pkg{EpiCompare}, an \proglang{R} package that suppliments and enhance current infectious disease modeling analysis pipelines as well as to encourage comparisons across these pipelines.
+  We present \pkg{EpiCompare}, an \proglang{R} package that suppliments and enhances current infectious disease analysis pipelines and encourages comparisons across models and epidemics.
 A major contribution of this work is the set of novel \textit{time-invariate} tools for model and epidemic comparisons - including time-invariate prediction bands.
-\pkg{EpiCompare} encorporates \proglang{R}'s \textit{tidy} coding style to aid it rapid and easy use.
+\pkg{EpiCompare} embraces \proglang{R}'s \textit{tidy} coding style to make adoption of the package easier and analysis faster.
 This paper provides an overview of both the tools in and intuition behind \pkg{EpiCompare} and a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
@@
   \usepackage{amsthm}
+  \usepackage{afterpage} 
 output: 
@@
 # Introduction {short-title="Intro" #sec:intro}
+
+\ben{I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful.
+Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.}
 
@@
 Epidemics, despite by definition being a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales to understand the processes at play.
-Additionally, many tools designed to examine the quantity of the population in each epidemic state (for example: quantity of Susceptible vs Infectious vs Recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
+<!-- Additionally, many tools designed to examine the quantity of the population in each epidemic state (for example: quantity of Susceptible vs Infectious vs Recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
 \shannon{I don't quite understand this.
 Is this because of underreporting?
-If so I'm not sure how EpiCompare helps us here.} Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens.
+If so I'm not sure how EpiCompare helps us here.} -->
+\ben{Moreover, many current comparison tools for state-space models (e.g.
+SIR models) highlight the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
+This approach may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.} Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens.
 The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
 
-This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
+This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
 
@@
 
-Epidemics can be difficult to compare to one another due to differences in diseases, locations, or population behaviors, or times.
+\ben{the vast majority of this section has been changed / needs more updates.}
+
+<!--Epidemics can be difficult to compare to one another due to differences in diseases, locations, or population behaviors, or times.
 In \pkg{EpiCompare}, we emphasize comparisons between epidemics adjusting for the last component, time.
@@
 Moreover, this tool avoids the need for choosing a 'beginning' $t_0$ or 'end' $t_F$ of an epidemic, choices \cite{gallagher2020} show that can heavily influence, for example,  estimates of peak infection height or the reproduction number $R_0$.
+-->
 
-## $R_0$  and time-invariant analysis {short-title="r0" #r0}
+\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by (1) taking a global, not marginal view of how epidemics move through time and (2) by treating full epidemics as filaments, not functions (breaking away from a time-centric view).
+The following section aims to highlight the strengths of \textit{time-invariance} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
+
+Epidemics are complex objects, and hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the effected population reacts, and the time aspects (including start of the epidemic, speed of infection and more).
+Time-invariant analysis attempts to make different epidemics easier to compare by removing many time dependent aspects of an epidemic and instead focus on the overall "life" of an epidemic, emphasizing the number of lives affected.
+
+## Motivating time-invariance through $R_0$ {short-title="motivating through R0" #r0}
+
+Time-invariance analysis, as it appears in \pkg{EpiCompare}, can be seen as an attempt to escape  many difficulties comparing different epidemics.
+With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is still possible.
+Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproductive number $R_0$, which \citet{Gallagher2020} showed can be heavily impacted by choices of beginning and ending time points of an epidemic.
+
+\ben{
+Story: 
+(1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic.
+Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020).
+}
+
+$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is associated with the Susceptible-Infectious-Recovered (SIR) model.
+For a demonstration of $R_0$ getting lost in a time-dependent analysis we introduce the \citet{Kermack1927}’s common SIR model.
+This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
+
+\begin{align}\label{eq:sir-ode}
+    S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
+    I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
+    R^\prime(t) &= \gamma I(t) \nonumber.
+\end{align}
+
+From this model, $R_0 = \beta/\gamma$, aka the ratio of the estimated infection rate compared to the estimated recovery rate.
+
+Interestingly, with regards to traditional epidemic $state$ vs.
+$time$ plots, when comparing different epidemics their $R_0$s are often impossible to compare.
+This can be seen is a basic example, where we have 2 different epidemics that start in a population of 1000 with 10 individuals infected, but with different parameters related to equation \ref{eq:ODE}.
+Specifically suppose that $\beta_1, \gamma_1 = (a,b)$ and $\beta_2, \gamma_2 = (c,d)$ and that we observe both of these epidemics for 15 days.
+The epidemic trajectories are shown in the $state$ vs.
+time plots in Figure \ref{fig:different-scales-standard}.
+At a glance, we may assume that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
+On the other hand, we may think Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15.
+Yet both epidemics have the same $R_0$ of 2.
+
+However, when we treat both epidemics a functional data through a space that tells us the proportion of the population each each class, as is seen in figure 3 we see a different story.
+The points seem to overlap and form the same trajectory.
+Now it seems to be that Model 2 is following the same trajectory as Model 1 but is not as far along in the infection process.
+We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
+
+We can see this mathematically if we let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.
+As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
+And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.
+
+\begin{theorem}\label{thm:sir-scale}
+Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+\end{theorem}
+
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.
+
+<!-- ## $R_0$  and time-invariant analysis {short-title="r0" #r0}
 
@@
 On the other hand, we may think Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15.
+-->
 
@@
 ```
+
+<!--
 However, when we plot the trajectories as a single curve using the ternary plot in the time-invariant view, we immediately see a different story.
@@
 ```
+-->
 
+##  Beyond $R_0$ and Kerman’s and McKendrick SIR Models {short-title="Beyond R0 and SIR" #beyond-r0-sir}
+
+Through $R_0$ we saw that treating epidemics more like a function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models), then we could better compare the overall structure of the epidemic and how it impacted the population.
+Without knowing the generative structure underlying these two epidemics, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
+
+\ben{
+to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset.
+Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.
+}
+
+<!--
 ## Beyond the Kermack and McKendrick SIR models
@@
 
+-->
+
 # Overview of \pkg{EpiCompare} {short-title="Package overview" #sec:overview}
@@
 
+\afterpage{\clearpage}
 \begin{sidewaysfigure}[!ht]
     \centering
-    \includegraphics[width = 1\textwidth]{images/pipeline2_v2.pdf}
+    \includegraphics[width = 1\textwidth]{images/pipeline2.pdf}
     \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}

### 2021-01-26 jss/EpiCompare-paper/EpiCompare-paper.tex
rcpp speed up, clean up of arrangeGrob + grid.arrange, and additional comment in readme
section 3
epicompare section 2 rewrite + update of pipeline2 image

@@
 We present \pkg{EpiCompare}, an \proglang{R} package that suppliments
-and enhance current infectious disease modeling analysis pipelines as
-well as to encourage comparisons across these pipelines. A major
-contribution of this work is the set of novel \textit{time-invariate}
-tools for model and epidemic comparisons - including time-invariate
-prediction bands. \pkg{EpiCompare} encorporates \proglang{R}'s
-\textit{tidy} coding style to aid it rapid and easy use. This paper
+and enhances current infectious disease analysis pipelines and
+encourages comparisons across models and epidemics. A major contribution
+of this work is the set of novel \textit{time-invariate} tools for model
+and epidemic comparisons - including time-invariate prediction bands.
+\pkg{EpiCompare} embraces \proglang{R}'s \textit{tidy} coding style to
+make adoption of the package easier and analysis faster. This paper
 provides an overview of both the tools in and intuition behind
@@
 
-\usepackage{amsmath} \usepackage{amsthm}
+\usepackage{amsmath} \usepackage{amsthm} \usepackage{afterpage}
 
@@
 \section[Intro]{Introduction}\label{sec:intro}
+
+\textcolor{violet}{I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful.
+Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.}
 
@@
 constrained to initial times or time scales to understand the processes
-at play. Additionally, many tools designed to examine the quantity of
-the population in each epidemic state (for example: quantity of
-Susceptible vs Infectious vs Recovered individuals) don't always as
-intelligently capture the natural connections between the proportion of
-individuals in these states.
-\textcolor{orange}{I don't quite understand this.  Is this because of underreporting?  If so I'm not sure how EpiCompare helps us here.}
+at play.
+\textcolor{violet}{Moreover, many current comparison tools for state-space models (e.g.
+SIR models) highlight the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
+This approach may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.}
 Tools in \pkg{EpiCompare} give the user the ability to extend their
@@
 
-This paper is broken up into the following sections, section
+This paper is broken up into the following sections; section
 \ref{sec:time-invariant} motivates and showcases tools of time-invariant
@@
 
-Epidemics can be difficult to compare to one another due to differences
-in diseases, locations, or population behaviors, or times. In
-\pkg{EpiCompare}, we emphasize comparisons between epidemics adjusting
-for the last component, time. Time-invariant analysis is beneficial
-because by adjusting for the unit of infection rate, we can focus on the
-``lifetime'' of an epidemic, a view that is concerned more with the
-number of lives affected than as opposed to any specific time
-constraint. Time-invariant analysis is also beneficial when there are
-gaps in time between occurrences of outbreaks of a similar nature in
-different geographic regions. Finally, time-invariant analysis is
-beneficial because many studies talk about a period of ``exponential
-growth'\,' of the number of outbreaks
-\citep{chowell2007,wallinga2007generation,forsberg2008}.
+\textcolor{violet}{the vast majority of this section has been changed / needs more updates.}
 
-Time-invariant analysis, as it appears in \pkg{EpiCompare} solves the
-above issues by observing functionals which attempt to capture the
-general shape of the epidemic with respect to the proportion of the
-population in each epidemic state. This allows us to compare scenarios
-as different as, for instance as a decades-long outbreak HIV in the US
-compared to a 10 day outbreak of norovirus on a cruise ship. Moreover,
-this tool avoids the need for choosing a `beginning' \(t_0\) or `end'
-\(t_F\) of an epidemic, choices \cite{gallagher2020} show that can
-heavily influence, for example, estimates of peak infection height or
-the reproduction number \(R_0\).
+\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by
+(1) taking a global, not marginal view of how epidemics move through
+time and (2) by treating full epidemics as filaments, not functions
+(breaking away from a time-centric view). The following section aims to
+highlight the strengths of \textit{time-invariance} and define the
+mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
 
-\subsection[r0]{\(R_0\) and time-invariant analysis}\label{r0}
+Epidemics are complex objects, and hard to assess and compare to one
+another due to the differences in the diseases, the location where the
+outbreak occurs, how the effected population reacts, and the time
+aspects (including start of the epidemic, speed of infection and more).
+Time-invariant analysis attempts to make different epidemics easier to
+compare by removing many time dependent aspects of an epidemic and
+instead focus on the overall ``life'' of an epidemic, emphasizing the
+number of lives affected.
 
-By definition, \(R_0\) is the number of expected secondary infections
-when a primary infection is introduced to a susceptible population.
-\(R_0\) is also, maybe, the most famous \textit{time-invariant}
-numerical summary of an epidemic, which allows epidemics to be compared
-to one another in both different time and geographic scales. For
-example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
-influenza between 1.2-2, and modern measles outbreaks as large as 12
-\citep{midas2020,biggerstaff2014,namee2018}.
+\subsection[motivating through R0]{Motivating time-invariance through
+\(R_0\)}\label{r0}
 
-Estimators for \(R_0\) are dependent on the epidemic modeling framework,
-which consists of which states an individual can occupy
-(e.g.~susceptible, infectious, recovered) and a description of how
-individuals move from one state to the next over time (see
-\cite{hethcote1994}). A common epidemic modeling framework is the SIR
-model, originally introduced by \citet{kermack1927}. Transitions from
-one state to the next are defined by a series of ordinary differential
-equations, where \(N\) is the (fixed) total number individuals,
-\(\beta\) is the rate of infection, and \(\gamma\) is rate of recovery,
+Time-invariance analysis, as it appears in \pkg{EpiCompare}, can be seen
+as an attempt to escape many difficulties comparing different epidemics.
+With time-invariant analysis, comparing the decades-long outbreak of HIV
+in the US to a 10 day outbreak of norovirus on a cruise ship is still
+possible. Time-dependent problems can arise when estimation of certain
+epidemiological parameters, including the reproductive number \(R_0\),
+which \citet{Gallagher2020} showed can be heavily impacted by choices of
+beginning and ending time points of an epidemic.
+
+\textcolor{violet}{
+Story: 
+(1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic.
+Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020).
+}
+
+\(R_0\) is probably the most famous time-invariant numerical summary of
+an epidemic and is associated with the Susceptible-Infectious-Recovered
+(SIR) model. For a demonstration of \(R_0\) getting lost in a
+time-dependent analysis we introduce the \citet{Kermack1927}'s common
+SIR model. This model captures the transitions from one state to the
+next as a system of ordinary differential equations, where \(N\) is the
+total number of individuals, \(\beta\) is the rate of infection, and
+\(\gamma\) is the rate of recovery,
+
 \begin{align}\label{eq:sir-ode}
@@
     R^\prime(t) &= \gamma I(t) \nonumber.
-  \end{align} From this,
-\(\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}\), the ratio of the
-estimated infection rate compared to the estimated recovery rate.
+\end{align}
 
-With regards to traditional epidemic \(state\) vs.~\(time\) plots,
-\(R_0\) is difficult to visualize, especially with respect from one
-epidemic to another. For example, consider the scenarios where the first
-epidemic is generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
-\(\beta_1 = 0.3\) and \(\gamma_1 = 0.15\), and the second epidemic is
-generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
-\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over the first 15 days. Both
-epidemics have the same value of
-\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
+From this model, \(R_0 = \beta/\gamma\), aka the ratio of the estimated
+infection rate compared to the estimated recovery rate.
+
+Interestingly, with regards to traditional epidemic \(state\)
+vs.~\(time\) plots, when comparing different epidemics their \(R_0\)s
+are often impossible to compare. This can be seen is a basic example,
+where we have 2 different epidemics that start in a population of 1000
+with 10 individuals infected, but with different parameters related to
+equation \ref{eq:ODE}. Specifically suppose that
+\(\beta_1, \gamma_1 = (a,b)\) and \(\beta_2, \gamma_2 = (c,d)\) and that
+we observe both of these epidemics for 15 days. The epidemic
 trajectories are shown in the \(state\) vs.~time plots in Figure
@@
 Model 2 has a larger \(R_0\) because we may think the number of
-infections in that model has not yet peaked at time 15.
+infections in that model has not yet peaked at time 15. Yet both
+epidemics have the same \(R_0\) of 2.
+
+However, when we treat both epidemics a functional data through a space
+that tells us the proportion of the population each each class, as is
+seen in figure 3 we see a different story. The points seem to overlap
+and form the same trajectory. Now it seems to be that Model 2 is
+following the same trajectory as Model 1 but is not as far along in the
+infection process. We can see there is something fundamentally linking
+these two different epidemics, and this fundamental link turns out to be
+\(R_0\).
+
+We can see this mathematically if we let our two epidemics be presented
+as \(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
+\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
+example, assume both models have the same initial values
+\((S(0), I(0), R(0))\), and let
+\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
+\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
+rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
+the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
+\(\gamma_2 = a \gamma_1\).
+
+\begin{theorem}\label{thm:sir-scale}
+Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+\end{theorem}
+
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
+result from \cite{Harko2014} and is shown in detail in Proof
+\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
+for two SIR models that have the same initial percent of individuals in
+each state and \(R_0\) then for every point on the epidemic path of the
+first SIR model is also a point on the epidemic path of the second SIR
+model.
 
@@
 
-However, when we plot the trajectories as a single curve using the
-ternary plot in the time-invariant view, we immediately see a different
-story. In this time-invariant view in Fig.
-\ref{fig:different-scales-tern}, the points seem to overlap and form the
-same trajectory. Now it seems to be that Model 2 is following the same
-trajectory as Model 1 but is not as far along in the infection process.
-We can see there is something fundamentally linking these two different
-epidemics, and this fundamental link turns out to be \(R_0\).
+\subsection[Beyond R0 and SIR]{Beyond \(R_0\) and Kerman's and
+McKendrick SIR Models}\label{beyond-r0-sir}
 
-More formally, let two Kermack and McKendrick SIR models(see
-Eq.\eqref{eq:sir-ode} ) be denoted \((S_1(t), I_1(t), R_1(t))\) and
-\((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume both
-models have initial values \((S(0), I(0), R(0))\). Let
-\(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
-\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
-rate, respectively, for SIR model \(i=1, 2\). Equivalently,
-\(\beta_2 = a \beta_1\) if and only if \(\gamma_2 = a \gamma_1\) for
-some \(a > 0\).
+Through \(R_0\) we saw that treating epidemics more like a function of
+the time points in some higher dimensional space (nicely visualizable
+with a ternary plot for the SIR models), then we could better compare
+the overall structure of the epidemic and how it impacted the
+population. Without knowing the generative structure underlying these
+two epidemics, we propose that time-invariant tools can help compare
+them in smart ways, and that these ideas can extend past the CITE
+kermack1927 model and to epidemics with more states.
 
-\begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
-\end{theorem}
-
-The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
-result from \cite{Harko2014} and is shown in detail in Proof
-\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
-for two SIR models that have the same initial percent of individuals in
-each state and \(R_0\) then for every point on the epidemic path of the
-first SIR model is also a point on the epidemic path of the second SIR
-model. Taking the sample simulations from Fig.
-\ref{fig:different-scales-standard}, Fig.
-\ref{fig:different-scales-tern} presents these two models in a ternary
-plot. This means with our time-invariant ternary plot, that at a glance,
-we can tell if two epidemics have different values of \(R_0\).
-
-\begin{CodeChunk}
-\begin{figure}[H]
-
-{\centering \includegraphics{Figs/unnamed-chunk-3-1}
-
+\textcolor{violet}{
+to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset.
+Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.
 }
-
-\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
-Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
-These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
-\end{figure}
-\end{CodeChunk}
-
-\hypertarget{beyond-the-kermack-and-mckendrick-sir-models}{%
-\subsection{Beyond the Kermack and McKendrick SIR
-models}\label{beyond-the-kermack-and-mckendrick-sir-models}}
-
-Although the result of Theorem \ref{thm:sir-scale} allows for easy
-visual comparison of \(R_0\) in Kermack SIR models, it does require
-stringent assumptions of homogeneity of behavior in populations. The use
-of visualizing epidemics in a time-invariant lens via ternary plots
-extends beyond those of models that follow the ODEs in the
-Kermack-McKendrick equations. Any model with S, I, and R states can be
-visualized with ternary plots, regardless of birth and death dynamics
-and regardless of homogeneity of individual behavior. We can use ternary
-plots to compare the spread of a disease for groups within a population
-without time as a confounding factor. We show an example of this in a
-later section.
-
-Moreover, time-invariant analysis is also useful for epidemic models
-with more than three epidemic states. The constraints in three
-dimensions that are met with the SIR model (there are three epidemic
-states and for each time step, the sum of total number of individuals in
-each state is the population size at that time) represents a space of 3d
-simplices, and the ternary plot specifically represents these simplices,
-after scaling the values in each state as proportions of the total
-population at that time (ternary plots are known as a 3d unit simplex
-due to it's scaling). This same scaling for larger models (i.e.~with
-more states) can be done onto different simplexes. In this package we
-present tools to help compare models (mostly through simulations).
-\pkg{EpiCompare}'s tools permits comparisons of \(d\)-dimensional
-objects by projecting them into (\(d-1\))-dimensional space.
-
-Even though higher dimensional models may not be able to visualized
-simply, we provide a number of tools to aid in the comparison of models
-and epidemics. The first of which uses multiple simulations under
-specific model parameters to assess the \textcolor{orange}{variability?}
-bairabilty of the model fit. In \pkg{EpiCompare}, we provide ways to
-create prediction regions for a `true' epidemic under a fully specified
-model. These regions require representing multi-dimensional structures
-for functions to completely contain epidemics. \pkg{EpiCompare} treats
-these simulations and epidemics as \textit{filamental} objects --
-\textcolor{orange}{need a description for filamental, Ben feel free to change}
-where not only the outbreak trajectory points are important but also
-their form which can be considered a function.
-\textcolor{orange}{could be useful to have a picture/example here?
-like  three curves and 2 metrics one where closer under a pointwise distance and the other closer in filamental distance?
-I'll sketch up an idea unless you have something in your proposal already}
-
-We extend off of papers like \citet{Dalmasso2019a} to create these
-bands. These high dimensional bands allow the user to assess if the true
-epidemic is within the band (thereby assessing the model's
-representation of the epidemic). With these bands, we can compare
-\textit{models} to \textit{models} by assessing the distance between the
-two bands of model representation. We recommend using the Hausdorff
-distance to compare two bands to one another, as this distance captures
-how much bigger the sets would have to expand to cover each other. The
-Hausdorff distance is defined mathematically as \[
-d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y),\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
-\]
 

### 2021-03-12 jss/EpiCompare-paper/EpiCompare-paper.Rmd
updates to section 3 (2/3rds of it anyway).

@@
   \usepackage{afterpage}
+  \usepackage[normalem]{ulem} 
 output: 
@@
 
-\ben{**The goals of this section is to present parts of the data science pipeline, introduce how we help, why they are useful and the ideas behind it.**}
-
-In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.
-In Fig.
-\ref{fig:pipeline2}, we illustrate how our package fits into the data analysis pipeline introduced in Fig.
-\ref{fig:pipeline}.
-All front-facing functions are aimed to be as user-friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
-Although users can typically incorporate \pkg{EpiCompare} into any step in the data analysis process, there are two primary points of entry.
-The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation.
-Figure \ref{fig:pipeline2} captures these different paths, and we will highlight both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
-
 \afterpage{\clearpage}
@@
 
+In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.
+In Fig.
+\ref{fig:pipeline2}, we illustrate how our \ben{\sout{package fits} package's functions fit} into the data analysis pipeline introduced in Fig.
+\ref{fig:pipeline}.
+All front-facing functions are aimed to be as user-friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
+Although users can typically incorporate \pkg{EpiCompare} into any step in the data analysis process, there are two primary points of entry.
+The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation.
+Figure \ref{fig:pipeline2} captures these different paths, and we will highlight both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
+
 **Data Pre-processing**
 
-The first step of most data analysis is cleaning up the data to be explored.
+The first step of most data analysis is cleaning up the data \ben{so it can \sout{to}} be explored.
 There are multiple ways to collect the epidemiological data.
-Sometimes individual records with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information will be collected, whereas other data collections will focus on aggregate counts of individuals in each epidemic state.
-In fact, usually only the number of new infections at each time step is observed and the compartment totals are then imputed from that along with other information about the disease and the population of interest.
-We focus on understanding the overall impact of an outbreak at the aggregate/population level.
-In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g.
+Sometimes individual records \ben{are collected}, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information\ben{.\sout{will be collected, whereas}} \ben{O}ther data collections will focus on aggregate counts of individuals in each epidemic state.
+In fact, usually only the number of new infections at each time step is observed and the compartment totals \ben{(amounts of individuals in each state)} are then imputed from that along with other information about the disease and the population of interest.
+\ben{In \pkg{EpiCompare}, w}e focus on understanding the overall impact of an outbreak at the aggregate/population level, \ben{which allows examination of overall trends of an epidemic more easily}.
+
+\ben{[Ben says: the following rewrite was done because the original text seemed to repeat the same thing 2x, once in text and the second time with the function }`agents_to_aggregate`\ben{.
+The following two paragraphs try to merge the introduction of the function and it's capabilities into one story.]}
+
+\ben{[Ben says: this whole paragraph is new]} In order to help the practitioner to a point in which they can examine their epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`.
+This function transforms data that contains information about individual/agents' initial entry into each state (e.g.
+start of infection, start of recovery, etc) to an aggregate view of how many individuals where in a state at a given time.
+There are often situations where grouping agents into subpopulations (e.g.
+subpopulations defined by age or sex) can highlight different aggregate level trends.
+Papers like \citet{rvachev1985,anderson1992,worby2015} also develop state-based models that account for subpopulations in their models\footnote{\ben{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
+This function allows the user to easily group agents into new subpopulations by using the \pkg{tidyverse} paradigm to first `group_by()`^[The `group_by` function is from the \pkg{dplyr} package.] the subpopulation defining features and then aggregate (using `agents_to_aggregate`) conditional on these subpopulations.
+
+\ben{[Ben says: this whole paragraph is new]} The `agents_to_aggregate` function is also flexible to deal with a wide range of information about each individual.
+It can account for infinitely many states, allowing the practioner to aggregate information relative to the standard "Suspectible", "Infectious", and "Recovered" states (CITE) or also add states like "Exposed", "iMmune", "Hospitalized" and more.
+Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates (that change the overall number of individuals in the population).
+This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data.
+For these reasons, we view `agents_to_aggregate()` as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.
+
+\ben{\sout{In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g.
 start of infection, start of recovery, etc).
@@
 This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g.
-age, sex) and many state based models provide for subpopulation-based states in their analysis \citep{rvachev1985,anderson1992,worby2015}.
+age, sex) and many state based models provide for subpopulation-based states in their analysis}} \citep{rvachev1985,anderson1992,worby2015}.
 
-We provide a "tidyverse"-styled function, `agents_to_aggregate()` to transform agent information into aggregate state information.
-As a "tidy" function, our function `agents_to_aggregate()` allows the user to identify and analyze subpopulations by first executing a `group_by()` from \pkg{dplyr}.
-The function `agents_to_aggregate` is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates.
-Our function, `agents_to_aggregate()` does not require agents to pass through every state listed and agents are allowed to start in any state.
+\ben{\sout{We provide a "tidyverse"-styled function, }`agents_to_aggregate()`\ben{\sout{ to transform agent information into aggregate state information.
+As a "tidy" function, our function }}`agents_to_aggregate()` \ben{\sout{allows the user to identify and analyze subpopulations by first executing a }}`group_by()` \ben{\sout{{ from \pkg{dplyr}.
+The function }}`agents_to_aggregate`\ben{\sout{is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates.
+Our function,}} `agents_to_aggregate()`\ben{\sout{ does not require agents to pass through every state listed and agents are allowed to start in any state.
 This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g.
 hours or minutes).
-The function `agents_to_aggregate()` returns the total number of individuals in each state at each time point.
-We view `agents_to_aggregate()` as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.
+The function }}`agents_to_aggregate()`\ben{\sout{ returns the total number of individuals in each state at each time point.
+We view }}`agents_to_aggregate()`\ben{\sout{ as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.}}
 
@@
 
-With raw data,  exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings.
-An expert coder can start from `agents_to_aggregate()` to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).
-Our `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens.
-It combines the ideas behind `agents_to_aggregate()` with \pkg{ggplot2}'s `geom_path()` and \pkg{ggtern}'s `coord_tern()` to visualize any number of groups epidemic trajectory in 3d simplex \citep{Wickham2016, Hamilton2018}.
-We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).
+\ben{[Ben says: I'm went back and thought that the version before this version was clearer.
+Wanna take a look and discuss?
+Side by side: [github](https://github.com/skgallagher/EpiCompare/commit/7dbb4996d1d3169d3e3a426bc9820c417ea87beb)]}
+
+\ben{[Ben says: this is actually an older version:]} With raw data, getting to know our data now-a-days very frequently means figuring out good combinations of visualizations, numerical summaries and subsetting.
+An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but we’ve also developed tools to rapidly explore your data if your have a 3 unique states models (like the SIR model).
+Our `geom_aggregate` provides a rapid way to explore different subpopulations experience of the epidemic, and combines the ideas behind agents_to_aggregate for the SIR case, with `geom_path` and `coord_term` to visualize any number of groups epidemic trajectory in 3d simplex space using `ggplot2` and `ggtern` \citep{Wickham2016, Hamilton2018}.
+Visualization tools for SIR models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+
+\ben{[Ben says: The 2 things I really like about the earlier (now current) version is (1) the introduction seems to make more sense in getting at what EDA really is, (2) the way it disucsses SIR models only 1 potential model instead of putting it as *the* model.]}
+
+\ben{\sout{With raw data, exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings.
+An expert coder can start from}} `agents_to_aggregate()` \ben{\sout{to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).
+Our}} `geom_aggregate()` \ben{\sout{provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens.
+It combines the ideas behind}} `agents_to_aggregate()`  \ben{\sout{with \pkg{ggplot2}'s} `geom_path()`  \ben{\sout{and \pkg{ggtern}'s}} `coord_tern()`  \ben{\sout{to visualize any number of groups epidemic trajectory in 3d simplex}} \citep{Wickham2016, Hamilton2018}.\ben{\sout{We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).}}
 
 **Model Fitting and Simulations**
+\ben{[Ben says: see footnote 3 about the introduction to this paragraph.]}
 
-Although this package does not focus on estimating a model for the data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models with Bernoulli/Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next.
+Although this package does not focus on estimating a model for the data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models with Bernoulli/Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next^[The original sentence seemed clearer to me (it may even be a run-on sentence now).
+I'm wondering if you didn't like the original sentence relative to the "truth-ful-ness" of the statement?
+Here's the original sentence slightly edited: "Although this package does not focus on estimating a model for the data, we do provide some power functions for simulation of basic state models with Bernoulli approximation of ODE state models encoded in simulate_agents and simulate_SIR_agents".].
 These functions can be naturally combined with `agents_to_aggregate()` to proceed in the pipeline.

### 2021-03-12 jss/EpiCompare-paper/EpiCompare-paper.tex
updates to section 3 (2/3rds of it anyway).

@@
 
-\usepackage{amsmath} \usepackage{amsthm} \usepackage{afterpage}
+\usepackage{amsmath} \usepackage{amsthm} \usepackage{afterpage} \usepackage[normalem]{ulem}
 
@@
 
-\textcolor{violet}{**The goals of this section is to present parts of the data science pipeline, introduce how we help, why they are useful and the ideas behind it.**}
+\afterpage{\clearpage}
+\begin{sidewaysfigure}[!ht]
+    \centering
+    \includegraphics[width = 1\textwidth]{images/pipeline2.pdf}
+    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
+    \label{fig:pipeline2}
+\end{sidewaysfigure}
 
@@
 and explain how they aid in the data analysis pipeline. In Fig.
-\ref{fig:pipeline2}, we illustrate how our package fits into the data
-analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
+\ref{fig:pipeline2}, we illustrate how our
+\textcolor{violet}{\sout{package fits} package's functions fit} into the
+data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
 front-facing functions are aimed to be as user-friendly as possible, and
@@
 
-\afterpage{\clearpage}
-\begin{sidewaysfigure}[!ht]
-    \centering
-    \includegraphics[width = 1\textwidth]{images/pipeline2.pdf}
-    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
-    \label{fig:pipeline2}
-\end{sidewaysfigure}
-
 \textbf{Data Pre-processing}
 
-The first step of most data analysis is cleaning up the data to be
-explored. There are multiple ways to collect the epidemiological data.
-Sometimes individual records with times of different states of the
+The first step of most data analysis is cleaning up the data
+\textcolor{violet}{so it can \sout{to}} be explored. There are multiple
+ways to collect the epidemiological data. Sometimes individual records
+\textcolor{violet}{are collected}, with times of different states of the
 epidemic (infection, recovery, etc.) as well as individual information
-like network structure, location, and sub-population information will be
-collected, whereas other data collections will focus on aggregate counts
-of individuals in each epidemic state. In fact, usually only the number
-of new infections at each time step is observed and the compartment
-totals are then imputed from that along with other information about the
-disease and the population of interest. We focus on understanding the
-overall impact of an outbreak at the aggregate/population level. In
-\pkg{EpiCompare}, we provide a function to transform information about
-each individual/agent's initial time of entry into each state
-(e.g.~start of infection, start of recovery, etc). This transformation
-between \textit{agent} information to \textit{aggregate} information is
-useful for seeing the overall trend of the epidemic (and how it impacts
-different subpopulations). Our tools allow the user to easily group
-agents and define new subpopulations to explore. This is important as
-case studies highlight the usefulness of identifying differing
-subpopulations (e.g.~age, sex) and many state based models provide for
-subpopulation-based states in their analysis
-\citep{rvachev1985,anderson1992,worby2015}.
+like network structure, location, and sub-population
+information\textcolor{violet}{.\sout{will be collected, whereas}}
+\textcolor{violet}{O}ther data collections will focus on aggregate
+counts of individuals in each epidemic state. In fact, usually only the
+number of new infections at each time step is observed and the
+compartment totals
+\textcolor{violet}{(amounts of individuals in each state)} are then
+imputed from that along with other information about the disease and the
+population of interest. \textcolor{violet}{In \pkg{EpiCompare}, w}e
+focus on understanding the overall impact of an outbreak at the
+aggregate/population level,
+\textcolor{violet}{which allows examination of overall trends of an epidemic more easily}.
 
-We provide a ``tidyverse''-styled function,
-\texttt{agents\_to\_aggregate()} to transform agent information into
-aggregate state information. As a ``tidy'' function, our function
-\texttt{agents\_to\_aggregate()} allows the user to identify and analyze
-subpopulations by first executing a \texttt{group\_by()} from
-\pkg{dplyr}. The function \texttt{agents\_to\_aggregate} is flexible in
-that there is no limit on the amount of permitted epidemiological states
-so users can incorporate information from compartments such as
-``Exposed'', ``iMmune'', or ``Hospitalized'' groups, for instance, and
-also permits indicators for death/exit and birth/entry dates. Our
-function, \texttt{agents\_to\_aggregate()} does not require agents to
-pass through every state listed and agents are allowed to start in any
-state. This function is constrained to integer time steps (for example
-days), but transformations (linear or otherwise) of the time columns can
-account for finer grained time steps (e.g.~hours or minutes). The
-function \texttt{agents\_to\_aggregate()} returns the total number of
-individuals in each state at each time point. We view
+\textcolor{violet}{[Ben says: the following rewrite was done because the original text seemed to repeat the same thing 2x, once in text and the second time with the function }\texttt{agents\_to\_aggregate}\textcolor{violet}{.
+The following two paragraphs try to merge the introduction of the function and it's capabilities into one story.]}
+
+\textcolor{violet}{[Ben says: this whole paragraph is new]} In order to
+help the practitioner to a point in which they can examine their
+epidemics from an aggregate/ population lens, we provide a function
+called \texttt{agents\_to\_aggregate}. This function transforms data
+that contains information about individual/agents' initial entry into
+each state (e.g.~start of infection, start of recovery, etc) to an
+aggregate view of how many individuals where in a state at a given time.
+There are often situations where grouping agents into subpopulations
+(e.g.~subpopulations defined by age or sex) can highlight different
+aggregate level trends. Papers like
+\citet{rvachev1985,anderson1992,worby2015} also develop state-based
+models that account for subpopulations in their
+models\footnote{\textcolor{violet}{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
+This function allows the user to easily group agents into new
+subpopulations by using the \pkg{tidyverse} paradigm to first
+\texttt{group\_by()}\footnote{The \texttt{group\_by} function is from
+  the \pkg{dplyr} package.} the subpopulation defining features and then
+aggregate (using \texttt{agents\_to\_aggregate}) conditional on these
+subpopulations.
+
+\textcolor{violet}{[Ben says: this whole paragraph is new]} The
+\texttt{agents\_to\_aggregate} function is also flexible to deal with a
+wide range of information about each individual. It can account for
+infinitely many states, allowing the practioner to aggregate information
+relative to the standard ``Suspectible'', ``Infectious'', and
+``Recovered'' states (CITE) or also add states like ``Exposed'',
+``iMmune'', ``Hospitalized'' and more. Additionally,
+\texttt{agents\_to\_aggregate} also permits indicators for death/exit
+and birth/entry dates (that change the overall number of individuals in
+the population). This function is the highest tested function in this
+package, and can deal with many nuances that might come-about in the
+individual data. For these reasons, we view
 \texttt{agents\_to\_aggregate()} as powerful tool in the step of
@@
 
+\textcolor{violet}{\sout{In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g.
+start of infection, start of recovery, etc).
+This transformation between \textit{agent} information to \textit{aggregate} information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
+Our tools allow the user to easily group agents and define new subpopulations to explore.
+This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g.
+age, sex) and many state based models provide for subpopulation-based states in their analysis}}
+\citep{rvachev1985,anderson1992,worby2015}.
+
+\textbackslash ben\{\sout{We provide a "tidyverse"-styled function, }\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ to transform agent information into aggregate state information.
+As a "tidy" function, our function }}\texttt{agents\_to\_aggregate()}
+\textcolor{violet}{\sout{allows the user to identify and analyze subpopulations by first executing a }}\texttt{group\_by()}
+\textbackslash ben\{\sout{{ from \pkg{dplyr}.
+The function }}\texttt{agents\_to\_aggregate}\textcolor{violet}{\sout{is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates.
+Our function,}}
+\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ does not require agents to pass through every state listed and agents are allowed to start in any state.
+This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g.
+hours or minutes).
+The function }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ returns the total number of individuals in each state at each time point.
+We view }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.}}
+
 \textbf{EDA}
 
-With raw data, exploratory analysis frequently means figuring out good
-combinations of visualizations, numerical summaries, and groupings. An
-expert coder can start from \texttt{agents\_to\_aggregate()} to
-successfully accomplish EDA in many ways, but we have also developed
-tools to allow a novice coder to rapidly explore data, as long as there
-three unique epidemiological states (like the SIR model). Our
-\texttt{geom\_aggregate()} provides a rapid way to explore different
-subpopulations' experiences of the epidemic in the ternary lens. It
-combines the ideas behind \texttt{agents\_to\_aggregate()} with
-\pkg{ggplot2}'s \texttt{geom\_path()} and \pkg{ggtern}'s
-\texttt{coord\_tern()} to visualize any number of groups epidemic
-trajectory in 3d simplex \citep{Wickham2016, Hamilton2018}. We developed
-these out-of-the-box visualization tools for SIR models because (1) SIR
-models are some of the most common and basic epidemic state-based models
-and (2) our simplex representation of these epidemics emphasizes a
-``time-invarance'' representation of the data (see Section
-\ref{sec:time-invariant}).
+\textcolor{violet}{[Ben says: I'm went back and thought that the version before this version was clearer.
+Wanna take a look and discuss?
+Side by side: [github](https://github.com/skgallagher/EpiCompare/commit/7dbb4996d1d3169d3e3a426bc9820c417ea87beb)]}
+
+\textcolor{violet}{[Ben says: this is actually an older version:]} With
+raw data, getting to know our data now-a-days very frequently means
+figuring out good combinations of visualizations, numerical summaries
+and subsetting. An expert coder can start from
+\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
+we've also developed tools to rapidly explore your data if your have a 3
+unique states models (like the SIR model). Our \texttt{geom\_aggregate}
+provides a rapid way to explore different subpopulations experience of
+the epidemic, and combines the ideas behind agents\_to\_aggregate for
+the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
+visualize any number of groups epidemic trajectory in 3d simplex space
+using \texttt{ggplot2} and \texttt{ggtern}
+\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
+were developed because (1) SIR models are some of the most common and
+basic epidemic state-based models and (2) our simplex representation of
+these epidemics emphasizes a ``time-invarance'' representation of the
+data (for a refresher see Section \ref{sec:time-invariant}).
+
+\textcolor{violet}{[Ben says: The 2 things I really like about the earlier (now current) version is (1) the introduction seems to make more sense in getting at what EDA really is, (2) the way it disucsses SIR models only 1 potential model instead of putting it as *the* model.]}
+
+\textcolor{violet}{\sout{With raw data, exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings. An expert coder can start from}}
+\texttt{agents\_to\_aggregate()}
+\textcolor{violet}{\sout{to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).
+Our}}
+\texttt{geom\_aggregate()}
+\textcolor{violet}{\sout{provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens. It combines the ideas behind}}
+\texttt{agents\_to\_aggregate()}
+\textbackslash ben\{\sout{with \pkg{ggplot2}'s} \texttt{geom\_path()}
+\textcolor{violet}{\sout{and \pkg{ggtern}'s}} \texttt{coord\_tern()}
+\textcolor{violet}{\sout{to visualize any number of groups epidemic trajectory in 3d simplex}}
+\citep{Wickham2016, Hamilton2018}.\textcolor{violet}{\sout{We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).}}
 
 \textbf{Model Fitting and Simulations}
+\textcolor{violet}{[Ben says: see footnote 3 about the introduction to this paragraph.]}
 
@@
 transitions which can be used to (noisily) approximate ODEs which
-describe transitions from one state to the next. These functions can be
-naturally combined with \texttt{agents\_to\_aggregate()} to proceed in
-the pipeline. The function \texttt{simulate\_SIR\_agents()} simulates an
-SIR epidemic with user inputs for the number of simulations, the initial
-number in each state, the infection and recovery parameters
-\((\beta, \gamma\)), and the total number of discrete time steps. This
-function allows for easy access to SIR model analysis and comparison.
-Beyond SIR models, the function \texttt{simulate\_agents()} takes as
-input a user-specified transition matrix and other epidemic parameters
-to allow the user to create simulations of an outbreak for \textit{any}
-number of states and any number of transitions among them. This allows
-for users to explore the space of models in an intuitive way without
-getting bogged down by too much mathematical detail. For consistency, we
-have made output from \texttt{simulate\_agents()} and
-\texttt{simulate\_SIR\_agents()} compatible with
-\texttt{agents\_to\_aggregate()} so aggregate information may easily be
-accessed.
+describe transitions from one state to the next\footnote{The original
+  sentence seemed clearer to me (it may even be a run-on sentence now).
+  I'm wondering if you didn't like the original sentence relative to the
+  ``truth-ful-ness'' of the statement? Here's the original sentence
+  slightly edited: ``Although this package does not focus on estimating
+  a model for the data, we do provide some power functions for
+  simulation of basic state models with Bernoulli approximation of ODE
+  state models encoded in simulate\_agents and simulate\_SIR\_agents''.}.
+These functions can be naturally combined with
+\texttt{agents\_to\_aggregate()} to proceed in the pipeline. The
+function \texttt{simulate\_SIR\_agents()} simulates an SIR epidemic with
+user inputs for the number of simulations, the initial number in each
+state, the infection and recovery parameters \((\beta, \gamma\)), and
+the total number of discrete time steps. This function allows for easy
+access to SIR model analysis and comparison. Beyond SIR models, the
+function \texttt{simulate\_agents()} takes as input a user-specified
+transition matrix and other epidemic parameters to allow the user to
+create simulations of an outbreak for \textit{any} number of states and
+any number of transitions among them. This allows for users to explore
+the space of models in an intuitive way without getting bogged down by
+too much mathematical detail. For consistency, we have made output from
+\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
+compatible with \texttt{agents\_to\_aggregate()} so aggregate
+information may easily be accessed.
 

### 2021-03-18 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Small edits in paper writing for Ben

@@
 \ref{fig:pipeline}.
-All front-facing functions are aimed to be as user-friendly as possible, and we focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
+All front-facing functions are aimed to be as user-friendly as possible.
+We also focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
 Although users can typically incorporate \pkg{EpiCompare} into any step in the data analysis process, there are two primary points of entry.
@@
 
-The first step of most data analysis is cleaning up the data \ben{so it can \sout{to}} be explored.
-There are multiple ways to collect the epidemiological data.
-Sometimes individual records \ben{are collected}, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information\ben{.\sout{will be collected, whereas}} \ben{O}ther data collections will focus on aggregate counts of individuals in each epidemic state.
-In fact, usually only the number of new infections at each time step is observed and the compartment totals \ben{(amounts of individuals in each state)} are then imputed from that along with other information about the disease and the population of interest.
+The first step of most data analysis is cleaning the data \ben{so it can \sout{to}} be explored.
+There are multiple ways to collect epidemiological data.
+Sometimes individual records \ben{are collected}, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information\ben{.\sout{will be collected, whereas}} \ben{O}ther data collections focus on aggregate counts of individuals in each epidemic state.
+In fact, usually only the number of new infections at each time step (e.g.
+weekly case counts) is observed.
+Compartment totals \ben{(amounts of individuals in each state)} are then imputed from those case counts along with other information about the disease and the population of interest.
 \ben{In \pkg{EpiCompare}, w}e focus on understanding the overall impact of an outbreak at the aggregate/population level, \ben{which allows examination of overall trends of an epidemic more easily}.
@@
 
-\ben{[Ben says: this whole paragraph is new]} In order to help the practitioner to a point in which they can examine their epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`.
-This function transforms data that contains information about individual/agents' initial entry into each state (e.g.
+\ben{[Ben says: this whole paragraph is new]} In order to help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`.
+This function transforms data about individual/agents' initial entry into each state (e.g.
 start of infection, start of recovery, etc) to an aggregate view of how many individuals where in a state at a given time.
@@
 subpopulations defined by age or sex) can highlight different aggregate level trends.
-Papers like \citet{rvachev1985,anderson1992,worby2015} also develop state-based models that account for subpopulations in their models\footnote{\ben{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
+For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for subpopulations in their models\footnote{\ben{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
 This function allows the user to easily group agents into new subpopulations by using the \pkg{tidyverse} paradigm to first `group_by()`^[The `group_by` function is from the \pkg{dplyr} package.] the subpopulation defining features and then aggregate (using `agents_to_aggregate`) conditional on these subpopulations.
+\shannon{I think these last two sentences are repetitive}
 
 \ben{[Ben says: this whole paragraph is new]} The `agents_to_aggregate` function is also flexible to deal with a wide range of information about each individual.
-It can account for infinitely many states, allowing the practioner to aggregate information relative to the standard "Suspectible", "Infectious", and "Recovered" states (CITE) or also add states like "Exposed", "iMmune", "Hospitalized" and more.
-Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates (that change the overall number of individuals in the population).
-This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data.
-For these reasons, we view `agents_to_aggregate()` as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.
+It can, theoretically, account for infinitely many states, allowing the practitioner to aggregate information relative to the standard "Suspectible", "Infectious", and "Recovered" states (CITE) or also add states like "Exposed", "iMmune", "Hospitalized" and more.
+Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates (that change the overall population size).
+\shannon{Don't think this is necessary \sout{This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data.}} The function `agents_to_aggregate()` is a powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.
 
@@
 
-\ben{[Ben says: this is actually an older version:]} With raw data, getting to know our data now-a-days very frequently means figuring out good combinations of visualizations, numerical summaries and subsetting.
-An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but we’ve also developed tools to rapidly explore your data if your have a 3 unique states models (like the SIR model).
-Our `geom_aggregate` provides a rapid way to explore different subpopulations experience of the epidemic, and combines the ideas behind agents_to_aggregate for the SIR case, with `geom_path` and `coord_term` to visualize any number of groups epidemic trajectory in 3d simplex space using `ggplot2` and `ggtern` \citep{Wickham2016, Hamilton2018}.
+\ben{[Ben says: this is actually an older version:]} With raw data, "getting to know" our data currently means figuring out good combinations of visualizations, numerical summaries and subsets.
+An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but \pkg{EpiCompare} also includes tools to rapidly explore your data if your has at least three epidemic states (although these tools work best with exactly three epidemic states).
+Our `geom_aggregate` provides a rapid way to explore different subpopulations' experience of the epidemic, and combines the ideas behind `agents_to_aggregate` for the SIR case, with `geom_path` and `coord_tern` to visualize any number of groups epidemic trajectory in 3d simplex space using \pkg{ggplot2} and \pkg{ggtern} \citep{Wickham2016, Hamilton2018}.
 Visualization tools for SIR models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
@@
 
-Although this package does not focus on estimating a model for the data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models with Bernoulli/Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next^[The original sentence seemed clearer to me (it may even be a run-on sentence now).
+Although this package does not focus on fitting a model to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.\sout{ with Bernoulli or Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next}^[The original sentence seemed clearer to me (it may even be a run-on sentence now).
 I'm wondering if you didn't like the original sentence relative to the "truth-ful-ness" of the statement?
@@
 However, model output is often far more complicated than what a traditional `.csv` would allow.
-As a result, a number of epidemic modeling packages return a special class, specific to their modeling.
+\shannon{Do we need three sentences about csvs?} As a result, a number of epidemic modeling packages return a special class, specific to their modeling.
 The special classes often contain a plethora of information from residuals, model diagnostics, input parameters, and more.

### 2021-03-18 jss/EpiCompare-paper/EpiCompare-paper.tex
Small edits in paper writing for Ben

@@
   }
+
+% Pandoc citation processing
 
@@
 data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
-front-facing functions are aimed to be as user-friendly as possible, and
-we focus on providing the user ``tidyverse'' style functions, that
+front-facing functions are aimed to be as user-friendly as possible. We
+also focus on providing the user ``tidyverse'' style functions, that
 encourage piping and also follow clear verb naming schemes
@@
 
-The first step of most data analysis is cleaning up the data
+The first step of most data analysis is cleaning the data
 \textcolor{violet}{so it can \sout{to}} be explored. There are multiple
-ways to collect the epidemiological data. Sometimes individual records
+ways to collect epidemiological data. Sometimes individual records
 \textcolor{violet}{are collected}, with times of different states of the
@@
 information\textcolor{violet}{.\sout{will be collected, whereas}}
-\textcolor{violet}{O}ther data collections will focus on aggregate
-counts of individuals in each epidemic state. In fact, usually only the
-number of new infections at each time step is observed and the
-compartment totals
+\textcolor{violet}{O}ther data collections focus on aggregate counts of
+individuals in each epidemic state. In fact, usually only the number of
+new infections at each time step (e.g.~weekly case counts) is observed.
+Compartment totals
 \textcolor{violet}{(amounts of individuals in each state)} are then
-imputed from that along with other information about the disease and the
-population of interest. \textcolor{violet}{In \pkg{EpiCompare}, w}e
-focus on understanding the overall impact of an outbreak at the
-aggregate/population level,
+imputed from those case counts along with other information about the
+disease and the population of interest.
+\textcolor{violet}{In \pkg{EpiCompare}, w}e focus on understanding the
+overall impact of an outbreak at the aggregate/population level,
 \textcolor{violet}{which allows examination of overall trends of an epidemic more easily}.
@@
 \textcolor{violet}{[Ben says: this whole paragraph is new]} In order to
-help the practitioner to a point in which they can examine their
-epidemics from an aggregate/ population lens, we provide a function
-called \texttt{agents\_to\_aggregate}. This function transforms data
-that contains information about individual/agents' initial entry into
+help the practitioner examine epidemics from an aggregate/ population
+lens, we provide a function called \texttt{agents\_to\_aggregate}. This
+function transforms data about individual/agents' initial entry into
 each state (e.g.~start of infection, start of recovery, etc) to an
@@
 (e.g.~subpopulations defined by age or sex) can highlight different
-aggregate level trends. Papers like
-\citet{rvachev1985,anderson1992,worby2015} also develop state-based
-models that account for subpopulations in their
+aggregate level trends. For example, research by
+\citet{rvachev1985,anderson1992,worby2015} develop state-based models
+that account for subpopulations in their
 models\footnote{\textcolor{violet}{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
@@
 subpopulations.
+\textcolor{orange}{I think these last two sentences are repetitive}
 
@@
 \texttt{agents\_to\_aggregate} function is also flexible to deal with a
-wide range of information about each individual. It can account for
-infinitely many states, allowing the practioner to aggregate information
-relative to the standard ``Suspectible'', ``Infectious'', and
-``Recovered'' states (CITE) or also add states like ``Exposed'',
-``iMmune'', ``Hospitalized'' and more. Additionally,
+wide range of information about each individual. It can, theoretically,
+account for infinitely many states, allowing the practitioner to
+aggregate information relative to the standard ``Suspectible'',
+``Infectious'', and ``Recovered'' states (CITE) or also add states like
+``Exposed'', ``iMmune'', ``Hospitalized'' and more. Additionally,
 \texttt{agents\_to\_aggregate} also permits indicators for death/exit
-and birth/entry dates (that change the overall number of individuals in
-the population). This function is the highest tested function in this
-package, and can deal with many nuances that might come-about in the
-individual data. For these reasons, we view
-\texttt{agents\_to\_aggregate()} as powerful tool in the step of
-pre-processing data, one that quickly and simply transforms individual
-level information into an aggregate view that can be processed and
-analyzed in a simpler manner.
+and birth/entry dates (that change the overall population size).
+\textcolor{orange}{Don't think this is necessary \sout{This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data.}}
+The function \texttt{agents\_to\_aggregate()} is a powerful tool in the
+step of pre-processing data, one that quickly and simply transforms
+individual level information into an aggregate view that can be
+processed and analyzed in a simpler manner.
 
@@
 \textcolor{violet}{[Ben says: this is actually an older version:]} With
-raw data, getting to know our data now-a-days very frequently means
-figuring out good combinations of visualizations, numerical summaries
-and subsetting. An expert coder can start from
-\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
-we've also developed tools to rapidly explore your data if your have a 3
-unique states models (like the SIR model). Our \texttt{geom\_aggregate}
-provides a rapid way to explore different subpopulations experience of
-the epidemic, and combines the ideas behind agents\_to\_aggregate for
-the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
-visualize any number of groups epidemic trajectory in 3d simplex space
-using \texttt{ggplot2} and \texttt{ggtern}
-\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
-were developed because (1) SIR models are some of the most common and
-basic epidemic state-based models and (2) our simplex representation of
-these epidemics emphasizes a ``time-invarance'' representation of the
-data (for a refresher see Section \ref{sec:time-invariant}).
+raw data, ``getting to know'' our data currently means figuring out good
+combinations of visualizations, numerical summaries and subsets. An
+expert coder can start from \texttt{agents\_to\_aggregate} to
+successfully do this in many ways, but \pkg{EpiCompare} also includes
+tools to rapidly explore your data if your has at least three epidemic
+states (although these tools work best with exactly three epidemic
+states). Our \texttt{geom\_aggregate} provides a rapid way to explore
+different subpopulations' experience of the epidemic, and combines the
+ideas behind \texttt{agents\_to\_aggregate} for the SIR case, with
+\texttt{geom\_path} and \texttt{coord\_tern} to visualize any number of
+groups epidemic trajectory in 3d simplex space using \pkg{ggplot2} and
+\pkg{ggtern} \citep{Wickham2016, Hamilton2018}. Visualization tools for
+SIR models were developed because (1) SIR models are some of the most
+common and basic epidemic state-based models and (2) our simplex
+representation of these epidemics emphasizes a ``time-invarance''
+representation of the data (for a refresher see Section
+\ref{sec:time-invariant}).
 
@@
 
-Although this package does not focus on estimating a model for the data,
-we do provide some flexible functions for simulation of basic
-discrete-time epidemic-state models with Bernoulli/Multinomial
-transitions which can be used to (noisily) approximate ODEs which
-describe transitions from one state to the next\footnote{The original
-  sentence seemed clearer to me (it may even be a run-on sentence now).
-  I'm wondering if you didn't like the original sentence relative to the
-  ``truth-ful-ness'' of the statement? Here's the original sentence
-  slightly edited: ``Although this package does not focus on estimating
-  a model for the data, we do provide some power functions for
-  simulation of basic state models with Bernoulli approximation of ODE
-  state models encoded in simulate\_agents and simulate\_SIR\_agents''.}.
-These functions can be naturally combined with
-\texttt{agents\_to\_aggregate()} to proceed in the pipeline. The
+Although this package does not focus on fitting a model to data, we do
+provide some flexible functions for simulation of basic discrete-time
+epidemic-state
+models.\sout{ with Bernoulli or Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next}\footnote{The
+  original sentence seemed clearer to me (it may even be a run-on
+  sentence now). I'm wondering if you didn't like the original sentence
+  relative to the ``truth-ful-ness'' of the statement? Here's the
+  original sentence slightly edited: ``Although this package does not
+  focus on estimating a model for the data, we do provide some power
+  functions for simulation of basic state models with Bernoulli
+  approximation of ODE state models encoded in simulate\_agents and
+  simulate\_SIR\_agents''.}. These functions can be naturally combined
+with \texttt{agents\_to\_aggregate()} to proceed in the pipeline. The
 function \texttt{simulate\_SIR\_agents()} simulates an SIR epidemic with
@@
 is often far more complicated than what a traditional \texttt{.csv}
-would allow. As a result, a number of epidemic modeling packages return
-a special class, specific to their modeling. The special classes often
-contain a plethora of information from residuals, model diagnostics,
-input parameters, and more. While incredibly useful, these special
-classes can be difficult for novice coders to navigate.
+would allow. \textcolor{orange}{Do we need three sentences about csvs?}
+As a result, a number of epidemic modeling packages return a special
+class, specific to their modeling. The special classes often contain a
+plethora of information from residuals, model diagnostics, input
+parameters, and more. While incredibly useful, these special classes can
+be difficult for novice coders to navigate.
 

### 2021-03-18 jss/EpiCompare-paper/EpiCompare-paper.Rmd
thursday chat edits

@@
 Compartment totals \ben{(amounts of individuals in each state)} are then imputed from those case counts along with other information about the disease and the population of interest.
-\ben{In \pkg{EpiCompare}, w}e focus on understanding the overall impact of an outbreak at the aggregate/population level, \ben{which allows examination of overall trends of an epidemic more easily}.
+\ben{In \pkg{EpiCompare}, w}e focus on understanding the overall impact of an outbreak at the aggregate/population level, \ben{which allows for examination of overall trends of an epidemic more easily}.
 
@@
 This function transforms data about individual/agents' initial entry into each state (e.g.
-start of infection, start of recovery, etc) to an aggregate view of how many individuals where in a state at a given time.
+start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
 There are often situations where grouping agents into subpopulations (e.g.
 subpopulations defined by age or sex) can highlight different aggregate level trends.
-For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for subpopulations in their models\footnote{\ben{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
-This function allows the user to easily group agents into new subpopulations by using the \pkg{tidyverse} paradigm to first `group_by()`^[The `group_by` function is from the \pkg{dplyr} package.] the subpopulation defining features and then aggregate (using `agents_to_aggregate`) conditional on these subpopulations.
-\shannon{I think these last two sentences are repetitive}
+For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for subpopulations.
+In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining \pkg{dplyr}::`group_by` and `agent_to_aggregate` to provide aggregation at a group level.
 
-\ben{[Ben says: this whole paragraph is new]} The `agents_to_aggregate` function is also flexible to deal with a wide range of information about each individual.
-It can, theoretically, account for infinitely many states, allowing the practitioner to aggregate information relative to the standard "Suspectible", "Infectious", and "Recovered" states (CITE) or also add states like "Exposed", "iMmune", "Hospitalized" and more.
-Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates (that change the overall population size).
-\shannon{Don't think this is necessary \sout{This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data.}} The function `agents_to_aggregate()` is a powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.
+\ben{[Ben says: this whole paragraph is new]} The `agents_to_aggregate` function is flexible and can deal with a wide range of information about each individual.
+It can, theoretically, account for infinitely many states.
+This functionality allows the practitioner to aggregate information relative to the standard states (e.g.
+"Susceptible", "Infectious", and "Recovered") as well as add states (e.g.
+"Exposed", "iMmune", "Hospitalized") (CITE).
+Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates.
+Overall, the function `agents_to_aggregate()` is a powerful tool for pre-processing data.
 
@@
 \ben{[Ben says: this is actually an older version:]} With raw data, "getting to know" our data currently means figuring out good combinations of visualizations, numerical summaries and subsets.
-An expert coder can start from `agents_to_aggregate` to successfully do this in many ways, but \pkg{EpiCompare} also includes tools to rapidly explore your data if your has at least three epidemic states (although these tools work best with exactly three epidemic states).
-Our `geom_aggregate` provides a rapid way to explore different subpopulations' experience of the epidemic, and combines the ideas behind `agents_to_aggregate` for the SIR case, with `geom_path` and `coord_tern` to visualize any number of groups epidemic trajectory in 3d simplex space using \pkg{ggplot2} and \pkg{ggtern} \citep{Wickham2016, Hamilton2018}.
-Visualization tools for SIR models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+An expert coder has many ways to successfully explore the data in an aggregate lens using `agents_to_aggregate`.
+\pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.
+Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our `geom_aggregate` provides a rapid way to explore different subpopulations' experience of an epidemic by combining the ideas behind `agents_to_aggregate` for three-state models to examine subpopulation trajectories in a 3d simplex space \citep{Wickham2016, Hamilton2018}.
+\shannon{come back?} Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+\shannon{make sure SIR is defined before}
 
@@
 **Model Fitting and Simulations**
-\ben{[Ben says: see footnote 3 about the introduction to this paragraph.]}
 
-Although this package does not focus on fitting a model to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.\sout{ with Bernoulli or Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next}^[The original sentence seemed clearer to me (it may even be a run-on sentence now).
-I'm wondering if you didn't like the original sentence relative to the "truth-ful-ness" of the statement?
-Here's the original sentence slightly edited: "Although this package does not focus on estimating a model for the data, we do provide some power functions for simulation of basic state models with Bernoulli approximation of ODE state models encoded in simulate_agents and simulate_SIR_agents".].
-These functions can be naturally combined with `agents_to_aggregate()` to proceed in the pipeline.
+Although this package does not focus on fitting a model to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
+These functions produce individual level information and can be naturally combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
 The function `simulate_SIR_agents()` simulates an SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
@@
 Beyond SIR models, the function `simulate_agents()` takes as input a user-specified transition matrix and other epidemic parameters to allow the user to create simulations of an outbreak for \textit{any} number of states and any number of transitions among them.
+This flexibility in states can be used to also reflect group-based dynamics.
 This allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.

### 2021-03-23 jss/EpiCompare-paper/EpiCompare-paper.Rmd
A new take on Section 2

@@
 
+\shannon{time invariance vs. time invariant analysis??}
+
+\shannon{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
+The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot; and 4) talk about models beyond SIR.}
+
+\shannon{I noticed the ternary plot kinda fell out of this section and it's not really detailed anywhere else in the paper.
+So I put it back in.
+It's a non-standard plot and a feature of EpiCompare so I think it's worth talking about, and I think it's appropriate to introduce in this section.}
+
 <!--Epidemics can be difficult to compare to one another due to differences in diseases, locations, or population behaviors, or times.
@@
 
-\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by (1) taking a global, not marginal view of how epidemics move through time and (2) by treating full epidemics as filaments, not functions (breaking away from a time-centric view).
+\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by (1) taking a global, not marginal view of how epidemics move through \sout{time} \shannon{populations} and (2) by treating full epidemics as filaments\sout{ not functions (breaking away from a time-centric view)} and not points produced by functions of time.
 The following section aims to highlight the strengths of \textit{time-invariance} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
 
-Epidemics are complex objects, and hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the effected population reacts, and the time aspects (including start of the epidemic, speed of infection and more).
-Time-invariant analysis attempts to make different epidemics easier to compare by removing many time dependent aspects of an epidemic and instead focus on the overall "life" of an epidemic, emphasizing the number of lives affected.
+\shannon{Mathematically}, epidemics are complex objects.
+They are hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the \shannon{a?}ffected population reacts, and the timefeatures (including start of the epidemic, speed of infection and more).
+Time-invariant analysis \sout{attempts to make} \shannon{makes} different epidemics easier to compare by removing many time dependent aspects of an epidemic.
+Instead \shannon{it} focuses on the overall \sout{"life"} \shannon{trajectory} of an epidemic, emphasizing the number of lives affected.
+\shannon{really like the lives affected part; I think that's a powerful statement}
 
@@
 
-Time-invariance analysis, as it appears in \pkg{EpiCompare}, can be seen as an attempt to escape  many difficulties comparing different epidemics.
-With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is still possible.
-Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproductive number $R_0$, which \citet{Gallagher2020} showed can be heavily impacted by choices of beginning and ending time points of an epidemic.
+Time-invariance analysis, as it appears in \pkg{EpiCompare}, \sout{can be seen as an attempt to escape} \shannon{bypasses}  many difficulties \shannon{in} comparing different epidemics.
+With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible.
+Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproductive number $R_0$, which \citet{Gallagher2020} show can be \sout{heavily impacted} \shannon{strongly effected} by choices of beginning and ending time points of an epidemic.
 
@@
 
-$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is associated with the Susceptible-Infectious-Recovered (SIR) model.
+$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is \shannon{often} associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.
 For a demonstration of $R_0$ getting lost in a time-dependent analysis we introduce the \citet{Kermack1927}’s common SIR model.
@@
 From this model, $R_0 = \beta/\gamma$, aka the ratio of the estimated infection rate compared to the estimated recovery rate.
+Since $\beta$ and $\gamma$ are both rates (of infection and recovery, respectively), the ratio of the two, $R_0$, is a time-invariant quantity.
+\shannon{Once $R_0$ is estimated, then important epidemic quantities can be inferred such as a the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic and the total number of infections, the final size.
+Moreover, $R_0$ allows us to compare different disease outbreaks and different outbreaks on the same scale.
+}
 
-Interestingly, with regards to traditional epidemic $state$ vs.
-$time$ plots, when comparing different epidemics their $R_0$s are often impossible to compare.
-This can be seen is a basic example, where we have 2 different epidemics that start in a population of 1000 with 10 individuals infected, but with different parameters related to equation \ref{eq:ODE}.
-Specifically suppose that $\beta_1, \gamma_1 = (a,b)$ and $\beta_2, \gamma_2 = (c,d)$ and that we observe both of these epidemics for 15 days.
+## Visualizing $R_0$ {short-title="visualizing r0" #vis-r0}
+
+With regards to traditional epidemic $state$ vs.
+$time$ plots, when \sout{comparing}\shannon{examining} different epidemics their $R_0$s are often impossible to compare.
+This can be seen in a basic example, where we have two different epidemics that start in two populations of 1000 people with 10 individuals \shannon{initially} infected, but with \sout{different parameters} \shannon{different infection and recovery rates (see Eq.
+\ref{eq:sir-ODE}}).
+Specifically suppose that $\beta_1, \gamma_1 = (a,b)$ and $\beta_2, \gamma_2 = (ca,cb)$ where $a, b, c > 0$.
+Further suppose we observe both of these epidemics for 15 days.
 The epidemic trajectories are shown in the $state$ vs.
 time plots in Figure \ref{fig:different-scales-standard}.
-At a glance, we may assume that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
-On the other hand, we may think Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15.
-Yet both epidemics have the same $R_0$ of 2.
+At a glance, we may \sout{assume} \shannon{believe} that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
+On the other hand, we may \sout{think} \shannon{believe} Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15.
+Yet, both epidemics have the same $R_0$ of 2.
 
-However, when we treat both epidemics a functional data through a space that tells us the proportion of the population each each class, as is seen in figure 3 we see a different story.
-The points seem to overlap and form the same trajectory.
-Now it seems to be that Model 2 is following the same trajectory as Model 1 but is not as far along in the infection process.
+\shannon{Introduce ternary plots here.  The below paragraphs are mostly new}
+
+However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.
+We call the epidemic trajectory of number of individuals in each state a \textit{filament}.
+Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,
+$$
+\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
+$$
+
+Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then $\psi$ will lay in a 2d triangular plane, i.e.
+a ternary plot.
+As a result, we can visualize the global trajectory of an epidemic in a single 2d plot and ultimately $R_0$.
+\shannon{Show pic here?}
+
+We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
+Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.
+Now, it is much clearer that Model 2 is following the same trajectory as Model 1 but is not as far along in the infection process.
 We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
 
-We can see this mathematically if we let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.
+We can show this mathematically if we let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.
 As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
@@
          tidyr::pivot_longer(cols = c(S, I, R))  %>%
-  mutate(name = factor(name, levels = c("S", "I", "R")))
+  mutate(name = factor(name, levels = c("S", "I", "R"))) %>%
+  filter(t <= 15)
 
@@
   facet_wrap(~name, ncol = 1, scales = "free") +
+  geom_line() +
   geom_point() +
@@
 This means with our time-invariant ternary plot, that at a glance, we can tell if two epidemics have different values of $R_0$.
+-->
 
@@
          filter(t <= 15)) +
+  geom_path(aes(x = S, y = I, z = R,
+                     col = type)) +
   geom_point(aes(x = S, y = I, z = R,
@@
 ```
--->
 
-##  Beyond $R_0$ and Kerman’s and McKendrick SIR Models {short-title="Beyond R0 and SIR" #beyond-r0-sir}
+##  Beyond $R_0$ and Kermack’s and McKendrick SIR Models {short-title="Beyond R0 and SIR" #beyond-r0-sir}
 
-Through $R_0$ we saw that treating epidemics more like a function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models), then we could better compare the overall structure of the epidemic and how it impacted the population.
-Without knowing the generative structure underlying these two epidemics, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
+Through $R_0$ \shannon{example}, we see that treating epidemics more like a \sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)} \shannon{filaments embedded in a lower dimensional space}, then we can better compare the overall structure of the epidemic and \sout{how it impacted the population} \shannon{see how the population was directly impacted}.
+\sout{Without knowing the generative structure underlying these two epidemics}, \shannon{When the underlying generative model is unknown for epidemic trajectories}, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
 
@@
 
-In this section, we highlight a number of the functionalities available in \pkg{EpiCompare}.
-These functionalities include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline \ref{fig:pipeline}.
+In this section, we highlight a number of the functionality available in \pkg{EpiCompare}.
+These functionality include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline \ref{fig:pipeline}.
 We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner.

### 2021-03-23 jss/EpiCompare-paper/EpiCompare-paper.tex
A new take on Section 2

@@
 
+\textcolor{orange}{time invariance vs. time invariant analysis??}
+
+\textcolor{orange}{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
+The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot; and 4) talk about models beyond SIR.}
+
+\textcolor{orange}{I noticed the ternary plot kinda fell out of this section and it's not really detailed anywhere else in the paper.
+So I put it back in.
+It's a non-standard plot and a feature of EpiCompare so I think it's worth talking about, and I think it's appropriate to introduce in this section.}
+
 \pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by
 (1) taking a global, not marginal view of how epidemics move through
-time and (2) by treating full epidemics as filaments, not functions
-(breaking away from a time-centric view). The following section aims to
-highlight the strengths of \textit{time-invariance} and define the
+\sout{time} \textcolor{orange}{populations} and (2) by treating full
+epidemics as
+filaments\sout{ not functions (breaking away from a time-centric view)}
+and not points produced by functions of time. The following section aims
+to highlight the strengths of \textit{time-invariance} and define the
 mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
 
-Epidemics are complex objects, and hard to assess and compare to one
-another due to the differences in the diseases, the location where the
-outbreak occurs, how the effected population reacts, and the time
-aspects (including start of the epidemic, speed of infection and more).
-Time-invariant analysis attempts to make different epidemics easier to
-compare by removing many time dependent aspects of an epidemic and
-instead focus on the overall ``life'' of an epidemic, emphasizing the
-number of lives affected.
+\textcolor{orange}{Mathematically}, epidemics are complex objects. They
+are hard to assess and compare to one another due to the differences in
+the diseases, the location where the outbreak occurs, how the
+\textcolor{orange}{a?}ffected population reacts, and the timefeatures
+(including start of the epidemic, speed of infection and more).
+Time-invariant analysis \sout{attempts to make}
+\textcolor{orange}{makes} different epidemics easier to compare by
+removing many time dependent aspects of an epidemic. Instead
+\textcolor{orange}{it} focuses on the overall \sout{"life"}
+\textcolor{orange}{trajectory} of an epidemic, emphasizing the number of
+lives affected.
+\textcolor{orange}{really like the lives affected part; I think that's a powerful statement}
 
@@
 
-Time-invariance analysis, as it appears in \pkg{EpiCompare}, can be seen
-as an attempt to escape many difficulties comparing different epidemics.
+Time-invariance analysis, as it appears in \pkg{EpiCompare},
+\sout{can be seen as an attempt to escape} \textcolor{orange}{bypasses}
+many difficulties \textcolor{orange}{in} comparing different epidemics.
 With time-invariant analysis, comparing the decades-long outbreak of HIV
-in the US to a 10 day outbreak of norovirus on a cruise ship is still
-possible. Time-dependent problems can arise when estimation of certain
-epidemiological parameters, including the reproductive number \(R_0\),
-which \citet{Gallagher2020} showed can be heavily impacted by choices of
-beginning and ending time points of an epidemic.
+in the US to a 10 day outbreak of norovirus on a cruise ship is
+\sout{still} possible. Time-dependent problems can arise when estimation
+of certain epidemiological parameters, including the reproductive number
+\(R_0\), which \citet{Gallagher2020} show can be \sout{heavily impacted}
+\textcolor{orange}{strongly effected} by choices of beginning and ending
+time points of an epidemic.
 
@@
 \(R_0\) is probably the most famous time-invariant numerical summary of
-an epidemic and is associated with the Susceptible-Infectious-Recovered
-(SIR) model. For a demonstration of \(R_0\) getting lost in a
-time-dependent analysis we introduce the \citet{Kermack1927}'s common
-SIR model. This model captures the transitions from one state to the
-next as a system of ordinary differential equations, where \(N\) is the
-total number of individuals, \(\beta\) is the rate of infection, and
-\(\gamma\) is the rate of recovery,
+an epidemic and is \textcolor{orange}{often} associated with the
+Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}. For a
+demonstration of \(R_0\) getting lost in a time-dependent analysis we
+introduce the \citet{Kermack1927}'s common SIR model. This model
+captures the transitions from one state to the next as a system of
+ordinary differential equations, where \(N\) is the total number of
+individuals, \(\beta\) is the rate of infection, and \(\gamma\) is the
+rate of recovery,
 
@@
 From this model, \(R_0 = \beta/\gamma\), aka the ratio of the estimated
-infection rate compared to the estimated recovery rate.
-
-Interestingly, with regards to traditional epidemic \(state\)
-vs.~\(time\) plots, when comparing different epidemics their \(R_0\)s
-are often impossible to compare. This can be seen is a basic example,
-where we have 2 different epidemics that start in a population of 1000
-with 10 individuals infected, but with different parameters related to
-equation \ref{eq:ODE}. Specifically suppose that
-\(\beta_1, \gamma_1 = (a,b)\) and \(\beta_2, \gamma_2 = (c,d)\) and that
+infection rate compared to the estimated recovery rate. Since \(\beta\)
+and \(\gamma\) are both rates (of infection and recovery, respectively),
+the ratio of the two, \(R_0\), is a time-invariant quantity.
+\textcolor{orange}{Once $R_0$ is estimated, then important epidemic quantities can be inferred such as a the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic and the total number of infections, the final size.
+Moreover, $R_0$ allows us to compare different disease outbreaks and different outbreaks on the same scale.
+}
+
+\subsection[visualizing r0]{Visualizing \(R_0\)}\label{vis-r0}
+
+With regards to traditional epidemic \(state\) vs.~\(time\) plots, when
+\sout{comparing}\textcolor{orange}{examining} different epidemics their
+\(R_0\)s are often impossible to compare. This can be seen in a basic
+example, where we have two different epidemics that start in two
+populations of 1000 people with 10 individuals
+\textcolor{orange}{initially} infected, but with
+\sout{different parameters}
+\textcolor{orange}{different infection and recovery rates (see Eq. \ref{eq:sir-ODE}}).
+Specifically suppose that \(\beta_1, \gamma_1 = (a,b)\) and
+\(\beta_2, \gamma_2 = (ca,cb)\) where \(a, b, c > 0\). Further suppose
 we observe both of these epidemics for 15 days. The epidemic
 trajectories are shown in the \(state\) vs.~time plots in Figure
-\ref{fig:different-scales-standard}. At a glance, we may assume that
-Model 1 has a larger \(R_0\) than Model 2 because the peak of infection
-occurs more quickly than in Model 2. On the other hand, we may think
-Model 2 has a larger \(R_0\) because we may think the number of
-infections in that model has not yet peaked at time 15. Yet both
-epidemics have the same \(R_0\) of 2.
-
-However, when we treat both epidemics a functional data through a space
-that tells us the proportion of the population each each class, as is
-seen in figure 3 we see a different story. The points seem to overlap
-and form the same trajectory. Now it seems to be that Model 2 is
-following the same trajectory as Model 1 but is not as far along in the
-infection process. We can see there is something fundamentally linking
-these two different epidemics, and this fundamental link turns out to be
-\(R_0\).
-
-We can see this mathematically if we let our two epidemics be presented
+\ref{fig:different-scales-standard}. At a glance, we may \sout{assume}
+\textcolor{orange}{believe} that Model 1 has a larger \(R_0\) than Model
+2 because the peak of infection occurs more quickly than in Model 2. On
+the other hand, we may \sout{think} \textcolor{orange}{believe} Model 2
+has a larger \(R_0\) because we may think the number of infections in
+that model has not yet peaked at time 15. Yet, both epidemics have the
+same \(R_0\) of 2.
+
+\textcolor{orange}{Introduce ternary plots here.  The below paragraphs are mostly new}
+
+However, time-invariant centric analysis allows us to visualize \(R_0\)
+for the Kermack and McKendrick SIR model. We call the epidemic
+trajectory of number of individuals in each state a \textit{filament}.
+Mathematically, a filament \(\psi\) for the SIR model \(\psi\) this is
+the set of points of the number in each state for each time point, \[
+\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
+\]
+
+Since the number in each state is non-negative and the sum over the
+three states for a given time point sums to \(N\), then \(\psi\) will
+lay in a 2d triangular plane, i.e.~a ternary plot. As a result, we can
+visualize the global trajectory of an epidemic in a single 2d plot and
+ultimately \(R_0\). \textcolor{orange}{Show pic here?}
+
+We visualize the two epidemics in a global, ternary view in Figure
+\ref{fig:different-scales-tern}. Without getting into too much detail of
+the intricacies of this plot, we immediately see the points of the two
+filaments \(\psi\) seem to form the same trajectory. Now, it is much
+clearer that Model 2 is following the same trajectory as Model 1 but is
+not as far along in the infection process. We can see there is something
+fundamentally linking these two different epidemics, and this
+fundamental link turns out to be \(R_0\).
+
+We can show this mathematically if we let our two epidemics be presented
 as \(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
@@
 
-\subsection[Beyond R0 and SIR]{Beyond \(R_0\) and Kerman's and
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-3-1}
+
+}
+
+\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
+These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
+\end{figure}
+\end{CodeChunk}
+
+\subsection[Beyond R0 and SIR]{Beyond \(R_0\) and Kermack's and
 McKendrick SIR Models}\label{beyond-r0-sir}
 
-Through \(R_0\) we saw that treating epidemics more like a function of
-the time points in some higher dimensional space (nicely visualizable
-with a ternary plot for the SIR models), then we could better compare
-the overall structure of the epidemic and how it impacted the
-population. Without knowing the generative structure underlying these
-two epidemics, we propose that time-invariant tools can help compare
-them in smart ways, and that these ideas can extend past the CITE
-kermack1927 model and to epidemics with more states.
+Through \(R_0\) \textcolor{orange}{example}, we see that treating
+epidemics more like a
+\sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)}
+\textcolor{orange}{filaments embedded in a lower dimensional space},
+then we can better compare the overall structure of the epidemic and
+\sout{how it impacted the population}
+\textcolor{orange}{see how the population was directly impacted}.
+\sout{Without knowing the generative structure underlying these two epidemics},
+\textcolor{orange}{When the underlying generative model is unknown for epidemic trajectories},
+we propose that time-invariant tools can help compare them in smart
+ways, and that these ideas can extend past the CITE kermack1927 model
+and to epidemics with more states.
 
@@
 overall impact of an outbreak at the aggregate/population level,
-\textcolor{violet}{which allows examination of overall trends of an epidemic more easily}.
+\textcolor{violet}{which allows for examination of overall trends of an epidemic more easily}.
 
@@
 function transforms data about individual/agents' initial entry into
-each state (e.g.~start of infection, start of recovery, etc) to an
-aggregate view of how many individuals where in a state at a given time.
+each state (e.g.~start of infection, start of recovery, etc.) to an
+aggregate view of how many individuals were in a state at a given time.
 There are often situations where grouping agents into subpopulations
@@
 \citet{rvachev1985,anderson1992,worby2015} develop state-based models
-that account for subpopulations in their
-models\footnote{\textcolor{violet}{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
-This function allows the user to easily group agents into new
-subpopulations by using the \pkg{tidyverse} paradigm to first
-\texttt{group\_by()}\footnote{The \texttt{group\_by} function is from
-  the \pkg{dplyr} package.} the subpopulation defining features and then
-aggregate (using \texttt{agents\_to\_aggregate}) conditional on these
-subpopulations.
-\textcolor{orange}{I think these last two sentences are repetitive}
+that account for subpopulations. In \pkg{EpiCompare}, we facilitate
+subpopulation analysis by combining \pkg{dplyr}::\texttt{group\_by} and
+\texttt{agent\_to\_aggregate} to provide aggregation at a group level.
 
 \textcolor{violet}{[Ben says: this whole paragraph is new]} The
-\texttt{agents\_to\_aggregate} function is also flexible to deal with a
+\texttt{agents\_to\_aggregate} function is flexible and can deal with a
 wide range of information about each individual. It can, theoretically,
-account for infinitely many states, allowing the practitioner to
-aggregate information relative to the standard ``Suspectible'',
-``Infectious'', and ``Recovered'' states (CITE) or also add states like
-``Exposed'', ``iMmune'', ``Hospitalized'' and more. Additionally,
-\texttt{agents\_to\_aggregate} also permits indicators for death/exit
-and birth/entry dates (that change the overall population size).
-\textcolor{orange}{Don't think this is necessary \sout{This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data.}}
-The function \texttt{agents\_to\_aggregate()} is a powerful tool in the
-step of pre-processing data, one that quickly and simply transforms
-individual level information into an aggregate view that can be
-processed and analyzed in a simpler manner.
+account for infinitely many states. This functionality allows the
+practitioner to aggregate information relative to the standard states
+(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as add
+states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'') (CITE).
+Additionally, \texttt{agents\_to\_aggregate} also permits indicators for
+death/exit and birth/entry dates. Overall, the function
+\texttt{agents\_to\_aggregate()} is a powerful tool for pre-processing
+data.
 
@@
 combinations of visualizations, numerical summaries and subsets. An
-expert coder can start from \texttt{agents\_to\_aggregate} to
-successfully do this in many ways, but \pkg{EpiCompare} also includes
-tools to rapidly explore your data if your has at least three epidemic
-states (although these tools work best with exactly three epidemic
-states). Our \texttt{geom\_aggregate} provides a rapid way to explore
-different subpopulations' experience of the epidemic, and combines the
-ideas behind \texttt{agents\_to\_aggregate} for the SIR case, with
-\texttt{geom\_path} and \texttt{coord\_tern} to visualize any number of
-groups epidemic trajectory in 3d simplex space using \pkg{ggplot2} and
-\pkg{ggtern} \citep{Wickham2016, Hamilton2018}. Visualization tools for
-SIR models were developed because (1) SIR models are some of the most
-common and basic epidemic state-based models and (2) our simplex
-representation of these epidemics emphasizes a ``time-invarance''
-representation of the data (for a refresher see Section
-\ref{sec:time-invariant}).
+expert coder has many ways to successfully explore the data in an
+aggregate lens using \texttt{agents\_to\_aggregate}. \pkg{EpiCompare}
+also includes tools to rapidly explore data that has three epidemic
+states. Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our
+\texttt{geom\_aggregate} provides a rapid way to explore different
+subpopulations' experience of an epidemic by combining the ideas behind
+\texttt{agents\_to\_aggregate} for three-state models to examine
+subpopulation trajectories in a 3d simplex space
+\citep{Wickham2016, Hamilton2018}. \textcolor{orange}{come back?}
+Visualization tools for three-state models were developed because (1)
+SIR models are some of the most common and basic epidemic state-based
+models and (2) our simplex representation of these epidemics emphasizes
+a ``time-invarance'' representation of the data (for a refresher see
+Section \ref{sec:time-invariant}).
+\textcolor{orange}{make sure SIR is defined before}
 
@@
 \textbf{Model Fitting and Simulations}
-\textcolor{violet}{[Ben says: see footnote 3 about the introduction to this paragraph.]}
 
@@
 provide some flexible functions for simulation of basic discrete-time
-epidemic-state
-models.\sout{ with Bernoulli or Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next}\footnote{The
-  original sentence seemed clearer to me (it may even be a run-on
-  sentence now). I'm wondering if you didn't like the original sentence
-  relative to the ``truth-ful-ness'' of the statement? Here's the
-  original sentence slightly edited: ``Although this package does not
-  focus on estimating a model for the data, we do provide some power
-  functions for simulation of basic state models with Bernoulli
-  approximation of ODE state models encoded in simulate\_agents and
-  simulate\_SIR\_agents''.}. These functions can be naturally combined
-with \texttt{agents\_to\_aggregate()} to proceed in the pipeline. The
-function \texttt{simulate\_SIR\_agents()} simulates an SIR epidemic with
-user inputs for the number of simulations, the initial number in each
-state, the infection and recovery parameters \((\beta, \gamma\)), and
-the total number of discrete time steps. This function allows for easy
-access to SIR model analysis and comparison. Beyond SIR models, the
-function \texttt{simulate\_agents()} takes as input a user-specified
-transition matrix and other epidemic parameters to allow the user to
-create simulations of an outbreak for \textit{any} number of states and
-any number of transitions among them. This allows for users to explore
-the space of models in an intuitive way without getting bogged down by
-too much mathematical detail. For consistency, we have made output from
-\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
-compatible with \texttt{agents\_to\_aggregate()} so aggregate
-information may easily be accessed.
+epidemic-state models. These functions produce individual level
+information and can be naturally combined with
+\texttt{agents\_to\_aggregate()} to view these simulations through an
+aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
+an SIR epidemic with user inputs for the number of simulations, the
+initial number in each state, the infection and recovery parameters
+\((\beta, \gamma\)), and the total number of discrete time steps. This
+function allows for easy access to SIR model analysis and comparison.
+Beyond SIR models, the function \texttt{simulate\_agents()} takes as
+input a user-specified transition matrix and other epidemic parameters
+to allow the user to create simulations of an outbreak for \textit{any}
+number of states and any number of transitions among them. This
+flexibility in states can be used to also reflect group-based dynamics.
+This allows for users to explore the space of models in an intuitive way
+without getting bogged down by too much mathematical detail. For
+consistency, we have made output from \texttt{simulate\_agents()} and
+\texttt{simulate\_SIR\_agents()} compatible with
+\texttt{agents\_to\_aggregate()} so aggregate information may easily be
+accessed.
 
@@
 
-In this section, we highlight a number of the functionalities available
-in \pkg{EpiCompare}. These functionalities include data cleaning,
+In this section, we highlight a number of the functionality available in
+\pkg{EpiCompare}. These functionality include data cleaning,
 visualization, simulation, and comparison, in accordance with the data

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.Rmd
partial comments on section 2 rewrite and other clean-up

@@
 
-\ben{I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful.
-Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.}
+\ben{[Ben says: I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful.
+Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.]}
 
@@
 historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}.
-At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021} .
+At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.]}
 
 With the many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021}, this field has a lot of devices to aid an individual practitioner decide the correct approach.
-For example,\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
+For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
 
-At the same time, analysis packages often only address a specific portion of the analysis pipeline, for instance focusing on certain types of models.
-Modeling tools, which usually require learning package-specific syntax, often don't provide easy ways to compare and assess their models on new data.
-Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways.
+At the same time, analysis packages often only address a specific portion of the analysis pipeline\ben{\sout{, for instance focusing on certain types of models.}} \ben{These m}odeling tools\ben{\sout{, which}} usually require learning package-specific syntax\ben{\sout{,}} and often don't provide easy ways to compare and assess their models on new data.
+Moreover, exploring\ben{, \sout{and}} modeling \ben{and comparing} epidemics require transforming and \textit{tidying} data in different ways.
 To fill these gaps, we present our \proglang{R} package \pkg{EpiCompare}.
@@
 If so I'm not sure how EpiCompare helps us here.} -->
-\ben{Moreover, many current comparison tools for state-space models (e.g.
-SIR models) highlight the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
-This approach may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.} Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens.
+\ben{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion.
+\sout{Moreover, m} M}any current \ben{time-dependent} comparison tools for state-space models (e.g.
+SIR models) \ben{\sout{highlight} examine} the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
+\ben{These \sout{This}} approach\ben{es} may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.
+Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens.
 The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
 
-This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
+This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a \ben{\sout{thorough}} demonstrating of the tools through a detailed example of a full data analysis pipeline.
 
 # Motivation and tools for time-invariant analysis {short-title="Time-invariant" #sec:time-invariant}
-
-\ben{the vast majority of this section has been changed / needs more updates.}
 
@@
 \shannon{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
-The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot; and 4) talk about models beyond SIR.}
+The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot\footnote{\ben{[Ben says: defend how this fits into the narrative?]}} and 4) talk }\ben{Time invariance in} \sout{about}\shannon{ models beyond SIR.}
 
@@
 
-\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by (1) taking a global, not marginal view of how epidemics move through \sout{time} \shannon{populations} and (2) by treating full epidemics as filaments\sout{ not functions (breaking away from a time-centric view)} and not points produced by functions of time.
+\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filaments^[\ben{[Ben says: Do we need a footnote explaining what this is more clearly?]}] and not points produced by functions of time.
 The following section aims to highlight the strengths of \textit{time-invariance} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
 
-\shannon{Mathematically}, epidemics are complex objects.
-They are hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the \shannon{a?}ffected population reacts, and the timefeatures (including start of the epidemic, speed of infection and more).
-Time-invariant analysis \sout{attempts to make} \shannon{makes} different epidemics easier to compare by removing many time dependent aspects of an epidemic.
-Instead \shannon{it} focuses on the overall \sout{"life"} \shannon{trajectory} of an epidemic, emphasizing the number of lives affected.
-\shannon{really like the lives affected part; I think that's a powerful statement}
+Mathematically, epidemics are complex objects.
+They are hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time \ben{related} features (including start of the epidemic, speed of infection and more).
+Time-invariant analysis make different epidemics easier to compare by removing many time dependent aspects of an epidemic.
+Instead it focuses on the overall trajectory of an epidemic, emphasizing the number of lives affected.
+\shannon{[Shannon says: really like the lives affected part; I think that's a powerful statement.]}
 
@@
 In Fig.
-\ref{fig:pipeline2}, we illustrate how our \ben{\sout{package fits} package's functions fit} into the data analysis pipeline introduced in Fig.
+\ref{fig:pipeline2}, we illustrate how our package's functions fit into the data analysis pipeline introduced in Fig.
 \ref{fig:pipeline}.
@@
 The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation.
-Figure \ref{fig:pipeline2} captures these different paths, and we will highlight both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
+Figure \ref{fig:pipeline2} captures these different paths, and we will highlight\footnote{\ben{[Ben says: we need to make sure we actually do highlight ]}} both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
 
@@
 
-The first step of most data analysis is cleaning the data \ben{so it can \sout{to}} be explored.
+The first step of most data analysis is cleaning the data so it can be explored.
 There are multiple ways to collect epidemiological data.
-Sometimes individual records \ben{are collected}, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information\ben{.\sout{will be collected, whereas}} \ben{O}ther data collections focus on aggregate counts of individuals in each epidemic state.
+Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+Other data collections focus on aggregate counts of individuals in each epidemic state.
 In fact, usually only the number of new infections at each time step (e.g.
 weekly case counts) is observed.
-Compartment totals \ben{(amounts of individuals in each state)} are then imputed from those case counts along with other information about the disease and the population of interest.
-\ben{In \pkg{EpiCompare}, w}e focus on understanding the overall impact of an outbreak at the aggregate/population level, \ben{which allows for examination of overall trends of an epidemic more easily}.
+Compartment totals (amounts of individuals in each state) are then imputed from those case counts along with other information about the disease and the population of interest.
+In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for examination of overall trends of an epidemic more easily.
 
-\ben{[Ben says: the following rewrite was done because the original text seemed to repeat the same thing 2x, once in text and the second time with the function }`agents_to_aggregate`\ben{.
-The following two paragraphs try to merge the introduction of the function and it's capabilities into one story.]}
-
-\ben{[Ben says: this whole paragraph is new]} In order to help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`.
+In order to help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`.
 This function transforms data about individual/agents' initial entry into each state (e.g.
@@
 
-\ben{[Ben says: this whole paragraph is new]} The `agents_to_aggregate` function is flexible and can deal with a wide range of information about each individual.
+The `agents_to_aggregate` function is flexible and can deal with a wide range of information about each individual.
 It can, theoretically, account for infinitely many states.
@@
 "Susceptible", "Infectious", and "Recovered") as well as add states (e.g.
-"Exposed", "iMmune", "Hospitalized") (CITE).
+"Exposed", "iMmune", "Hospitalized") (CITE\footnote{\ben{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}}).
 Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates.
@@
 
-\ben{\sout{In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g.
-start of infection, start of recovery, etc).
-This transformation between \textit{agent} information to \textit{aggregate} information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
-Our tools allow the user to easily group agents and define new subpopulations to explore.
-This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g.
-age, sex) and many state based models provide for subpopulation-based states in their analysis}} \citep{rvachev1985,anderson1992,worby2015}.
-
-\ben{\sout{We provide a "tidyverse"-styled function, }`agents_to_aggregate()`\ben{\sout{ to transform agent information into aggregate state information.
-As a "tidy" function, our function }}`agents_to_aggregate()` \ben{\sout{allows the user to identify and analyze subpopulations by first executing a }}`group_by()` \ben{\sout{{ from \pkg{dplyr}.
-The function }}`agents_to_aggregate`\ben{\sout{is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates.
-Our function,}} `agents_to_aggregate()`\ben{\sout{ does not require agents to pass through every state listed and agents are allowed to start in any state.
-This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g.
-hours or minutes).
-The function }}`agents_to_aggregate()`\ben{\sout{ returns the total number of individuals in each state at each time point.
-We view }}`agents_to_aggregate()`\ben{\sout{ as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.}}
-
 **EDA**
 
-\ben{[Ben says: I'm went back and thought that the version before this version was clearer.
-Wanna take a look and discuss?
-Side by side: [github](https://github.com/skgallagher/EpiCompare/commit/7dbb4996d1d3169d3e3a426bc9820c417ea87beb)]}
-
-\ben{[Ben says: this is actually an older version:]} With raw data, "getting to know" our data currently means figuring out good combinations of visualizations, numerical summaries and subsets.
+With raw data, "getting to know" our data currently means figuring out good combinations of visualizations, numerical summaries and subsets.
 An expert coder has many ways to successfully explore the data in an aggregate lens using `agents_to_aggregate`.
@@
 Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our `geom_aggregate` provides a rapid way to explore different subpopulations' experience of an epidemic by combining the ideas behind `agents_to_aggregate` for three-state models to examine subpopulation trajectories in a 3d simplex space \citep{Wickham2016, Hamilton2018}.
-\shannon{come back?} Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
-\shannon{make sure SIR is defined before}
-
-\ben{[Ben says: The 2 things I really like about the earlier (now current) version is (1) the introduction seems to make more sense in getting at what EDA really is, (2) the way it disucsses SIR models only 1 potential model instead of putting it as *the* model.]}
-
-\ben{\sout{With raw data, exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings.
-An expert coder can start from}} `agents_to_aggregate()` \ben{\sout{to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).
-Our}} `geom_aggregate()` \ben{\sout{provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens.
-It combines the ideas behind}} `agents_to_aggregate()`  \ben{\sout{with \pkg{ggplot2}'s} `geom_path()`  \ben{\sout{and \pkg{ggtern}'s}} `coord_tern()`  \ben{\sout{to visualize any number of groups epidemic trajectory in 3d simplex}} \citep{Wickham2016, Hamilton2018}.\ben{\sout{We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).}}
+\shannon{[Shannon says: come back?]}\footnote{\ben{[Ben says: what does this note mean?]}} Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+\shannon{[Shannon says: make sure SIR is defined before.]}
 

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.tex
partial comments on section 2 rewrite and other clean-up

@@
 
-\textcolor{violet}{I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful.
-Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.}
+\textcolor{violet}{[Ben says: I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful.
+Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.]}
 
@@
 government, and news organizations are an important first step of the
-process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021} .
+process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.]}
 
@@
 \citep{midasnetwork2021}, this field has a lot of devices to aid an
-individual practitioner decide the correct approach. For
-example,\proglang{R} packages such as \pkg{surveillance},
-\pkg{EpiModel}, and \pkg{pomp} have all made significant steps in
-standardizing the flow of the data analysis pipeline for epidemic
-modeling through digitizing data sets, making accessible statistical
-models, and providing a plethora of educational material for both coding
-novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
+individual practitioner decide the correct approach. For example,
+\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
+\pkg{pomp} have all made significant steps in standardizing the flow of
+the data analysis pipeline for epidemic modeling through digitizing data
+sets, making accessible statistical models, and providing a plethora of
+educational material for both coding novices and experts alike
+\citep{surveillance2017,Jenness2018,King2016}.
 
 At the same time, analysis packages often only address a specific
-portion of the analysis pipeline, for instance focusing on certain types
-of models. Modeling tools, which usually require learning
-package-specific syntax, often don't provide easy ways to compare and
-assess their models on new data. Moreover, exploring and modeling
-epidemics require transforming and \textit{tidying} data in different
-ways. To fill these gaps, we present our \proglang{R} package
-\pkg{EpiCompare}. Our package's primary focus is to aid and advance
-research in the area of comparison and assessment of epidemic and
-epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the
-data analysis pipeline of infectious diseases as 1) data pre-processing,
-2) exploratory data analysis (EDA), 3) modeling and simulating, 4)
-post-processing, and 5) comparison and assessment; where each previous
-part of the pipeline influences the next. \pkg{EpiCompare} provides
-tools to aids practitioners in all areas of this pipeline.
+portion of the analysis
+pipeline\textcolor{violet}{\sout{, for instance focusing on certain types of models.}}
+\textcolor{violet}{These m}odeling
+tools\textcolor{violet}{\sout{, which}} usually require learning
+package-specific syntax\textcolor{violet}{\sout{,}} and often don't
+provide easy ways to compare and assess their models on new data.
+Moreover, exploring\textcolor{violet}{, \sout{and}} modeling
+\textcolor{violet}{and comparing} epidemics require transforming and
+\textit{tidying} data in different ways. To fill these gaps, we present
+our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus
+is to aid and advance research in the area of comparison and assessment
+of epidemic and epidemiological models. In Figure \ref{fig:pipeline}, we
+illustrate the data analysis pipeline of infectious diseases as 1) data
+pre-processing, 2) exploratory data analysis (EDA), 3) modeling and
+simulating, 4) post-processing, and 5) comparison and assessment; where
+each previous part of the pipeline influences the next. \pkg{EpiCompare}
+provides tools to aids practitioners in all areas of this pipeline.
 
@@
 at play.
-\textcolor{violet}{Moreover, many current comparison tools for state-space models (e.g.
-SIR models) highlight the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
-This approach may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.}
-Tools in \pkg{EpiCompare} give the user the ability to extend their
-toolkit to evaluate epidemics within a time-invariant lens. The goal of
-\pkg{EpiCompare} is not to supplant existing infectious disease modeling
-tools and software but, rather, is a concerted effort to create standard
-and fair comparisons among models developed for disease outbreaks and
-outbreak data.
+\textcolor{violet}{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion. \sout{Moreover, m} M}any
+current \textcolor{violet}{time-dependent} comparison tools for
+state-space models (e.g.~SIR models)
+\textcolor{violet}{\sout{highlight} examine} the proportion of
+individuals in each state (at a given time) in a piece-wise / marginal
+fashion. \textcolor{violet}{These \sout{This}}
+approach\textcolor{violet}{es} may reduce the amount of connections that
+can be seen, similar to projections of a multidimensional distribution
+onto a single axis at a time. Tools in \pkg{EpiCompare} give the user
+the ability to extend their toolkit to evaluate epidemics within a
+time-invariant lens. The goal of \pkg{EpiCompare} is not to supplant
+existing infectious disease modeling tools and software but, rather, is
+a concerted effort to create standard and fair comparisons among models
+developed for disease outbreaks and outbreak data.
 
@@
 \pkg{EpiCompare} aids a practitioner in every step of the pipeline and
-section \ref{sec:tour} provides a thorough demonstrating of the tools
-through a detailed example of a full data analysis pipeline.
+section \ref{sec:tour} provides a \textcolor{violet}{\sout{thorough}}
+demonstrating of the tools through a detailed example of a full data
+analysis pipeline.
 
@@
 
-\textcolor{violet}{the vast majority of this section has been changed / needs more updates.}
-
 \textcolor{orange}{time invariance vs. time invariant analysis??}
@@
 \textcolor{orange}{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
-The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot; and 4) talk about models beyond SIR.}
+The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot\footnote{\textcolor{violet}{[Ben says: defend how this fits into the narrative?]}} and 4) talk }\textcolor{violet}{Time invariance in}
+\sout{about}\textcolor{orange}{ models beyond SIR.}
 
@@
 (1) taking a global, not marginal view of how epidemics move through
-\sout{time} \textcolor{orange}{populations} and (2) by treating full
-epidemics as
-filaments\sout{ not functions (breaking away from a time-centric view)}
+populations and (2) by treating full epidemics as filaments\footnote{\textcolor{violet}{[Ben says: Do we need a footnote explaining what this is more clearly?]}}
 and not points produced by functions of time. The following section aims
@@
 
-\textcolor{orange}{Mathematically}, epidemics are complex objects. They
-are hard to assess and compare to one another due to the differences in
-the diseases, the location where the outbreak occurs, how the
-\textcolor{orange}{a?}ffected population reacts, and the timefeatures
-(including start of the epidemic, speed of infection and more).
-Time-invariant analysis \sout{attempts to make}
-\textcolor{orange}{makes} different epidemics easier to compare by
-removing many time dependent aspects of an epidemic. Instead
-\textcolor{orange}{it} focuses on the overall \sout{"life"}
-\textcolor{orange}{trajectory} of an epidemic, emphasizing the number of
-lives affected.
-\textcolor{orange}{really like the lives affected part; I think that's a powerful statement}
+Mathematically, epidemics are complex objects. They are hard to assess
+and compare to one another due to the differences in the diseases, the
+location where the outbreak occurs, how the affected population reacts,
+and the time \textcolor{violet}{related} features (including start of
+the epidemic, speed of infection and more). Time-invariant analysis make
+different epidemics easier to compare by removing many time dependent
+aspects of an epidemic. Instead it focuses on the overall trajectory of
+an epidemic, emphasizing the number of lives affected.
+\textcolor{orange}{[Shannon says: really like the lives affected part; I think that's a powerful statement.]}
 
@@
 and explain how they aid in the data analysis pipeline. In Fig.
-\ref{fig:pipeline2}, we illustrate how our
-\textcolor{violet}{\sout{package fits} package's functions fit} into the
-data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
+\ref{fig:pipeline2}, we illustrate how our package's functions fit into
+the data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
 front-facing functions are aimed to be as user-friendly as possible. We
@@
 \ref{fig:pipeline2} captures these different paths, and we will
-highlight both approaches and how to leverage \pkg{EpiCompare} in the
-subsections below.
+highlight\footnote{\textcolor{violet}{[Ben says: we need to make sure we actually do highlight ]}}
+both approaches and how to leverage \pkg{EpiCompare} in the subsections
+below.
 
@@
 
-The first step of most data analysis is cleaning the data
-\textcolor{violet}{so it can \sout{to}} be explored. There are multiple
-ways to collect epidemiological data. Sometimes individual records
-\textcolor{violet}{are collected}, with times of different states of the
-epidemic (infection, recovery, etc.) as well as individual information
-like network structure, location, and sub-population
-information\textcolor{violet}{.\sout{will be collected, whereas}}
-\textcolor{violet}{O}ther data collections focus on aggregate counts of
+The first step of most data analysis is cleaning the data so it can be
+explored. There are multiple ways to collect epidemiological data.
+Sometimes individual records are collected, with times of different
+states of the epidemic (infection, recovery, etc.) as well as individual
+information like network structure, location, and sub-population
+information. Other data collections focus on aggregate counts of
 individuals in each epidemic state. In fact, usually only the number of
 new infections at each time step (e.g.~weekly case counts) is observed.
-Compartment totals
-\textcolor{violet}{(amounts of individuals in each state)} are then
+Compartment totals (amounts of individuals in each state) are then
 imputed from those case counts along with other information about the
-disease and the population of interest.
-\textcolor{violet}{In \pkg{EpiCompare}, w}e focus on understanding the
-overall impact of an outbreak at the aggregate/population level,
-\textcolor{violet}{which allows for examination of overall trends of an epidemic more easily}.
-
-\textcolor{violet}{[Ben says: the following rewrite was done because the original text seemed to repeat the same thing 2x, once in text and the second time with the function }\texttt{agents\_to\_aggregate}\textcolor{violet}{.
-The following two paragraphs try to merge the introduction of the function and it's capabilities into one story.]}
-
-\textcolor{violet}{[Ben says: this whole paragraph is new]} In order to
-help the practitioner examine epidemics from an aggregate/ population
-lens, we provide a function called \texttt{agents\_to\_aggregate}. This
-function transforms data about individual/agents' initial entry into
-each state (e.g.~start of infection, start of recovery, etc.) to an
-aggregate view of how many individuals were in a state at a given time.
-There are often situations where grouping agents into subpopulations
-(e.g.~subpopulations defined by age or sex) can highlight different
-aggregate level trends. For example, research by
-\citet{rvachev1985,anderson1992,worby2015} develop state-based models
-that account for subpopulations. In \pkg{EpiCompare}, we facilitate
-subpopulation analysis by combining \pkg{dplyr}::\texttt{group\_by} and
-\texttt{agent\_to\_aggregate} to provide aggregation at a group level.
-
-\textcolor{violet}{[Ben says: this whole paragraph is new]} The
-\texttt{agents\_to\_aggregate} function is flexible and can deal with a
-wide range of information about each individual. It can, theoretically,
-account for infinitely many states. This functionality allows the
-practitioner to aggregate information relative to the standard states
-(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as add
-states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'') (CITE).
+disease and the population of interest. In \pkg{EpiCompare}, we focus on
+understanding the overall impact of an outbreak at the
+aggregate/population level, which allows for examination of overall
+trends of an epidemic more easily.
+
+In order to help the practitioner examine epidemics from an aggregate/
+population lens, we provide a function called
+\texttt{agents\_to\_aggregate}. This function transforms data about
+individual/agents' initial entry into each state (e.g.~start of
+infection, start of recovery, etc.) to an aggregate view of how many
+individuals were in a state at a given time. There are often situations
+where grouping agents into subpopulations (e.g.~subpopulations defined
+by age or sex) can highlight different aggregate level trends. For
+example, research by \citet{rvachev1985,anderson1992,worby2015} develop
+state-based models that account for subpopulations. In \pkg{EpiCompare},
+we facilitate subpopulation analysis by combining
+\pkg{dplyr}::\texttt{group\_by} and \texttt{agent\_to\_aggregate} to
+provide aggregation at a group level.
+
+The \texttt{agents\_to\_aggregate} function is flexible and can deal
+with a wide range of information about each individual. It can,
+theoretically, account for infinitely many states. This functionality
+allows the practitioner to aggregate information relative to the
+standard states (e.g.~``Susceptible'', ``Infectious'', and
+``Recovered'') as well as add states (e.g.~``Exposed'', ``iMmune'',
+``Hospitalized'')
+(CITE\footnote{\textcolor{violet}{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}}).
 Additionally, \texttt{agents\_to\_aggregate} also permits indicators for
@@
 
-\textcolor{violet}{\sout{In \pkg{EpiCompare}, we provide a function to transform information about each individual/agent’s initial time of entry into each state (e.g.
-start of infection, start of recovery, etc).
-This transformation between \textit{agent} information to \textit{aggregate} information is useful for seeing the overall trend of the epidemic (and how it impacts different subpopulations).
-Our tools allow the user to easily group agents and define new subpopulations to explore.
-This is important as case studies highlight the usefulness of identifying differing subpopulations (e.g.
-age, sex) and many state based models provide for subpopulation-based states in their analysis}}
-\citep{rvachev1985,anderson1992,worby2015}.
-
-\textbackslash ben\{\sout{We provide a "tidyverse"-styled function, }\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ to transform agent information into aggregate state information.
-As a "tidy" function, our function }}\texttt{agents\_to\_aggregate()}
-\textcolor{violet}{\sout{allows the user to identify and analyze subpopulations by first executing a }}\texttt{group\_by()}
-\textbackslash ben\{\sout{{ from \pkg{dplyr}.
-The function }}\texttt{agents\_to\_aggregate}\textcolor{violet}{\sout{is flexible in that there is no limit on the amount of permitted epidemiological states so users can incorporate information from compartments such as "Exposed", "iMmune", or "Hospitalized" groups, for instance, and also permits indicators for death/exit and birth/entry dates.
-Our function,}}
-\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ does not require agents to pass through every state listed and agents are allowed to start in any state.
-This function is constrained to integer time steps (for example days), but transformations (linear or otherwise) of the time columns can account for finer grained time steps (e.g.
-hours or minutes).
-The function }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ returns the total number of individuals in each state at each time point.
-We view }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{ as powerful tool in the step of pre-processing data, one that quickly and simply transforms individual level information into an aggregate view that can be processed and analyzed in a simpler manner.}}
-
 \textbf{EDA}
 
-\textcolor{violet}{[Ben says: I'm went back and thought that the version before this version was clearer.
-Wanna take a look and discuss?
-Side by side: [github](https://github.com/skgallagher/EpiCompare/commit/7dbb4996d1d3169d3e3a426bc9820c417ea87beb)]}
-
-\textcolor{violet}{[Ben says: this is actually an older version:]} With
-raw data, ``getting to know'' our data currently means figuring out good
-combinations of visualizations, numerical summaries and subsets. An
+With raw data, ``getting to know'' our data currently means figuring out
+good combinations of visualizations, numerical summaries and subsets. An
 expert coder has many ways to successfully explore the data in an
@@
 subpopulation trajectories in a 3d simplex space
-\citep{Wickham2016, Hamilton2018}. \textcolor{orange}{come back?}
+\citep{Wickham2016, Hamilton2018}.
+\textcolor{orange}{[Shannon says: come back?]}\footnote{\textcolor{violet}{[Ben says: what does this note mean?]}}
 Visualization tools for three-state models were developed because (1)
@@
 Section \ref{sec:time-invariant}).
-\textcolor{orange}{make sure SIR is defined before}
-
-\textcolor{violet}{[Ben says: The 2 things I really like about the earlier (now current) version is (1) the introduction seems to make more sense in getting at what EDA really is, (2) the way it disucsses SIR models only 1 potential model instead of putting it as *the* model.]}
-
-\textcolor{violet}{\sout{With raw data, exploratory analysis frequently means figuring out good combinations of visualizations, numerical summaries, and groupings. An expert coder can start from}}
-\texttt{agents\_to\_aggregate()}
-\textcolor{violet}{\sout{to successfully accomplish EDA in many ways, but we have also developed tools to allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).
-Our}}
-\texttt{geom\_aggregate()}
-\textcolor{violet}{\sout{provides a rapid way to explore different subpopulations' experiences of the epidemic in the ternary lens. It combines the ideas behind}}
-\texttt{agents\_to\_aggregate()}
-\textbackslash ben\{\sout{with \pkg{ggplot2}'s} \texttt{geom\_path()}
-\textcolor{violet}{\sout{and \pkg{ggtern}'s}} \texttt{coord\_tern()}
-\textcolor{violet}{\sout{to visualize any number of groups epidemic trajectory in 3d simplex}}
-\citep{Wickham2016, Hamilton2018}.\textcolor{violet}{\sout{We developed these out-of-the-box visualization tools for SIR models because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (see Section \ref{sec:time-invariant}).}}
+\textcolor{orange}{[Shannon says: make sure SIR is defined before.]}
 

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Added a few sentences to 2.1 to address some of the purple text.

@@
 With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible.
-Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproductive number $R_0$, which \citet{Gallagher2020} show can be \sout{heavily impacted} \shannon{strongly effected} by choices of beginning and ending time points of an epidemic.
+Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproduction number $R_0$.
 
@@
 $R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is \shannon{often} associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.
-For a demonstration of $R_0$ getting lost in a time-dependent analysis we introduce the \citet{Kermack1927}’s common SIR model.
+\shannon{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a naive population \citep{anderson1992}.
+$R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}.
+} To demonstrate the difficulty of discerning $R_0$in a time-dependent analysis, we \shannon{first} introduce the \citet{Kermack1927}’s common SIR model.
 This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.tex
Added a few sentences to 2.1 to address some of the purple text.

@@
 \sout{still} possible. Time-dependent problems can arise when estimation
-of certain epidemiological parameters, including the reproductive number
-\(R_0\), which \citet{Gallagher2020} show can be \sout{heavily impacted}
-\textcolor{orange}{strongly effected} by choices of beginning and ending
-time points of an epidemic.
+of certain epidemiological parameters, including the reproduction number
+\(R_0\).
 
@@
 an epidemic and is \textcolor{orange}{often} associated with the
-Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}. For a
-demonstration of \(R_0\) getting lost in a time-dependent analysis we
-introduce the \citet{Kermack1927}'s common SIR model. This model
-captures the transitions from one state to the next as a system of
-ordinary differential equations, where \(N\) is the total number of
-individuals, \(\beta\) is the rate of infection, and \(\gamma\) is the
-rate of recovery,
+Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.
+\textcolor{orange}{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a naive population \citep{anderson1992}.
+$R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}.
+}
+To demonstrate the difficulty of discerning \(R_0\)in a time-dependent
+analysis, we \textcolor{orange}{first} introduce the
+\citet{Kermack1927}'s common SIR model. This model captures the
+transitions from one state to the next as a system of ordinary
+differential equations, where \(N\) is the total number of individuals,
+\(\beta\) is the rate of infection, and \(\gamma\) is the rate of
+recovery,
 

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.Rmd
updates to section 2 by Ben (after shannon's updates)

@@
   \usepackage{amsmath}
+  \usepackage{amssymb}
   \usepackage{amsthm}
@@
 
-## Motivating time-invariance through $R_0$ {short-title="motivating through R0" #r0}
+## Motivating time-invariance through $R_0$^[\ben{Is this still a relevant title for this section?}] {short-title="motivating through R0" #r0}
 
@@
 Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproduction number $R_0$.
+\ben{We will use $R_0$ to motivate the usefulness of time-invariaten analysis in this section.}
 
-\ben{
-Story: 
-(1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic.
+<!-- \ben{ -->
+<!-- Story:  -->
+<!-- (1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic.
 Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020).
-}
+-->
+<!-- } -->
 
 $R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is \shannon{often} associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.
-\shannon{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a naive population \citep{anderson1992}.
+\shannon{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a \ben{\sout{naive} completely susceptible}\footnote{\ben{[Ben says: I feel like naive is unclear - even if that's the technical term.]}} population \citep{anderson1992}.
 $R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}.
-} To demonstrate the difficulty of discerning $R_0$in a time-dependent analysis, we \shannon{first} introduce the \citet{Kermack1927}’s common SIR model.
+}\ben{[Ben says: could we expand on this difficulty / show it really is difficult with 1 more sentence?]} To demonstrate the difficulty of discerning $R_0$in a time-dependent analysis, we \shannon{first} introduce the \citet{Kermack1927}’s \ben{\sout{common}}\footnote{\ben{[Ben says: is it common?
+maybe just remove it?]}} SIR model.
 This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
@@
 
-## Visualizing $R_0$ {short-title="visualizing r0" #vis-r0}
+## \ben{Time invariant visualization of differences in $R_0}\sout{ Visualizing $R_0$} {short-title="visualizing r0" #vis-r0}
 
-With regards to traditional epidemic $state$ vs.
+\ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
+Also, I feel like the story is weak here.
+The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$.
+In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
+
+\ben{[Ben says: replacement of following paragraph: presentation rearranged to emphasis the figure first (when presenting the problem), and then discussing it.]} Traditional $state$ vs.
+$time$ plots, like seen in Fig.
+\ref{fig:different-scales-standard}, present information about epidemics in a time-dependent and piece-wise fashion, and make comparing different epidemic $R_0$s impossible.
+In Fig.
+\ref{fig:different-scales-standard} we compare [two different epidemics that start in two populations of 1000 people with 10 individuals initially infected, but with different infection and recovery rates (see Eq.
+\ref{eq:sir-ode})].
+More specifically, we've observed both epidemics for 15 days, with the first epidemic\footnote{In the plot it says model - probably want to say epidemic?} (in red) having parameters $\beta_1, \gamma_1 = (1,2)$ and the second epidemic (in blue) having  parameters $\beta_2, \gamma_2 = (2,4)$.
+[At a glance, we may believe that \sout{Model}\ben{epidemic} 1 has a larger $R_0$ than \sout{Model}\ben{epidemic} 2 because the peak of infection occurs more quickly than in Model 2 \ben{(remember $R_0$ relates to the rate of infection)}.
+On the other hand, we may believe \sout{Model}\ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15.
+Yet, both epidemics have the same $R_0$ of 2].
+
+\ben{[Ben says: bring Fig. \ref{fig:different-scales-standard} here.]}
+
+\ben{[Ben says: this paragraph is replaced with the one above]}With regards to traditional epidemic $state$ vs.
 $time$ plots, when \sout{comparing}\shannon{examining} different epidemics their $R_0$s are often impossible to compare.
 This can be seen in a basic example, where we have two different epidemics that start in two populations of 1000 people with 10 individuals \shannon{initially} infected, but with \sout{different parameters} \shannon{different infection and recovery rates (see Eq.
-\ref{eq:sir-ODE}}).
+\ref{eq:sir-ode}}).
 Specifically suppose that $\beta_1, \gamma_1 = (a,b)$ and $\beta_2, \gamma_2 = (ca,cb)$ where $a, b, c > 0$.
@@
 The epidemic trajectories are shown in the $state$ vs.
-time plots in Figure \ref{fig:different-scales-standard}.
+$time$ plots in Figure \ref{fig:different-scales-standard}.
 At a glance, we may \sout{assume} \shannon{believe} that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
@@
 
-However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.
+\ben{[Ben says: another rewrite of the paragraph below.]} A time-invariant approach to visualizing epidemics, which will be described more below, provides a simple way to compare $R_0$ values of epidemics that are generated from the Kermack and McKendrick SIR model.
+In these visuals, we treat epidemics as \textit{trajectories}.
+A trajectory can be mathematically viewed as a set of points is space that have an ordering, and that all points on the line between these order points are also contained in the geometric object.
+For a SIR epidemic, we can represent the associated filament $\psi$ as
+
+$$
+\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
+$$
+where the an mapping $\xi : t \to \mathbb{R}$ that is monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi = \psi$ where :
+
+$$
+\psi_\xi = \left \{(S(\xi(t)), I(\xi(t)), R(\xi(t))): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]}.
+$$
+[Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a \ben{\sout{2d triangular plane, i.e.
+a ternary plot}} \ben{two-dimensional triangular plane in the three-dimensional space, which can be visualized in a two-dimensional ternary plot}.
+As a result, we can visualize the global trajectory of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.
+\shannon{[Shannon says: Show pic here?]}]
+
+\ben{[Ben says: this paragraph was replaced with the paragraph above]} However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.
 We call the epidemic trajectory of number of individuals in each state a \textit{filament}.
@@
 
-We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
+\ben{[This section could be a bit less wordy.
+But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
 Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.
-Now, it is much clearer that Model 2 is following the same trajectory as Model 1 but is not as far along in the infection process.
+Now, it is much clearer that \ben{\sout{Model} epidemic} 2 is following the same trajectory as \ben{\sout{Model} epidemic} 1 but is not as far along in the infection process.
 We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
 
-We can show this mathematically if we let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.
-As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
+We can show this mathematically\ben{.\sout{ if we} L}et our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.\footnote{\ben{[Ben says: Originally this was a sentence fragment.]}}  As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
 And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.
@@
 The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model.\ben{[We've changed how we describe these two objects when they are related to the theorem - is that desirable?]}
 
@@
 
-##  Beyond $R_0$ and Kermack’s and McKendrick SIR Models {short-title="Beyond R0 and SIR" #beyond-r0-sir}
+##  \ben{Time-invariant analysis} Beyond $R_0$ and Kermack’s and McKendrick SIR Models {short-title="Beyond R0 and SIR" #beyond-r0-sir}
 
-Through $R_0$ \shannon{example}, we see that treating epidemics more like a \sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)} \shannon{filaments embedded in a lower dimensional space}, then we can better compare the overall structure of the epidemic and \sout{how it impacted the population} \shannon{see how the population was directly impacted}.
+Through \ben{the} $R_0$ \shannon{example}, we see that treating epidemics \ben{\sout{more}} like a \sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)} \shannon{filaments embedded in a lower dimensional space}\ben{\sout{, then we can} allows us to} better compare the overall structure of the epidemic and \sout{how it impacted the population} \shannon{see how the population was directly impacted}.
 \sout{Without knowing the generative structure underlying these two epidemics}, \shannon{When the underlying generative model is unknown for epidemic trajectories}, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
@@
     \centering
-    \includegraphics[width = 1\textwidth]{images/pipeline2.pdf}
+    \includegraphics[width = 1\textwidth]{images/pipeline2_1.pdf}
     \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.tex
updates to section 2 by Ben (after shannon's updates)

@@
 
-\usepackage{amsmath} \usepackage{amsthm} \usepackage{afterpage} \usepackage[normalem]{ulem}
+\usepackage{amsmath} \usepackage{amssymb} \usepackage{amsthm} \usepackage{afterpage} \usepackage[normalem]{ulem}
 
@@
 \subsection[motivating through R0]{Motivating time-invariance through
-\(R_0\)}\label{r0}
+\(R_0\)\footnote{\textcolor{violet}{Is this still a relevant title for this section?}}}\label{r0}
 
@@
 \(R_0\).
-
-\textcolor{violet}{
-Story: 
-(1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic.
-Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020).
-}
+\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariaten analysis in this section.}
 
@@
 Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.
-\textcolor{orange}{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a naive population \citep{anderson1992}.
+\textcolor{orange}{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a \textcolor{violet}{\sout{naive} completely susceptible}\footnote{\textcolor{violet}{[Ben says: I feel like naive is unclear - even if that's the technical term.]}} population \citep{anderson1992}.
 $R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}.
-}
+}\textcolor{violet}{[Ben says: could we expand on this difficulty / show it really is difficult with 1 more sentence?]}
 To demonstrate the difficulty of discerning \(R_0\)in a time-dependent
 analysis, we \textcolor{orange}{first} introduce the
-\citet{Kermack1927}'s common SIR model. This model captures the
-transitions from one state to the next as a system of ordinary
-differential equations, where \(N\) is the total number of individuals,
-\(\beta\) is the rate of infection, and \(\gamma\) is the rate of
-recovery,
+\citet{Kermack1927}'s
+\textcolor{violet}{\sout{common}}\footnote{\textcolor{violet}{[Ben says: is it common? maybe just remove it?]}}
+SIR model. This model captures the transitions from one state to the
+next as a system of ordinary differential equations, where \(N\) is the
+total number of individuals, \(\beta\) is the rate of infection, and
+\(\gamma\) is the rate of recovery,
 
@@
 
-\subsection[visualizing r0]{Visualizing \(R_0\)}\label{vis-r0}
-
-With regards to traditional epidemic \(state\) vs.~\(time\) plots, when
+\subsection[visualizing r0]{\sout{ Visualizing $R_0$}}\label{vis-r0}
+
+\textcolor{violet}{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
+Also, I feel like the story is weak here.
+The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$.
+In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
+
+\textcolor{violet}{[Ben says: replacement of following paragraph: presentation rearranged to emphasis the figure first (when presenting the problem), and then discussing it.]}
+Traditional \(state\) vs.~\(time\) plots, like seen in Fig.
+\ref{fig:different-scales-standard}, present information about epidemics
+in a time-dependent and piece-wise fashion, and make comparing different
+epidemic \(R_0\)s impossible. In Fig.
+\ref{fig:different-scales-standard} we compare {[}two different
+epidemics that start in two populations of 1000 people with 10
+individuals initially infected, but with different infection and
+recovery rates (see Eq. \ref{eq:sir-ode}){]}. More specifically, we've
+observed both epidemics for 15 days, with the first
+epidemic\footnote{In the plot it says model - probably want to say epidemic?}
+(in red) having parameters \(\beta_1, \gamma_1 = (1,2)\) and the second
+epidemic (in blue) having parameters \(\beta_2, \gamma_2 = (2,4)\).
+{[}At a glance, we may believe that
+\sout{Model}\textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
+\sout{Model}\textcolor{violet}{epidemic} 2 because the peak of infection
+occurs more quickly than in Model 2
+\textcolor{violet}{(remember $R_0$ relates to the rate of infection)}.
+On the other hand, we may believe
+\sout{Model}\textcolor{violet}{epidemic} 2 has a larger \(R_0\) because
+its unclear if the number of infections in that
+\sout{model}\textcolor{violet}{epidemic} has not yet peaked at time 15.
+Yet, both epidemics have the same \(R_0\) of 2{]}.
+
+\textcolor{violet}{[Ben says: bring Fig. \ref{fig:different-scales-standard} here.]}
+
+\textcolor{violet}{[Ben says: this paragraph is replaced with the one above]}With
+regards to traditional epidemic \(state\) vs.~\(time\) plots, when
 \sout{comparing}\textcolor{orange}{examining} different epidemics their
@@
 \sout{different parameters}
-\textcolor{orange}{different infection and recovery rates (see Eq. \ref{eq:sir-ODE}}).
+\textcolor{orange}{different infection and recovery rates (see Eq. \ref{eq:sir-ode}}).
 Specifically suppose that \(\beta_1, \gamma_1 = (a,b)\) and
@@
 we observe both of these epidemics for 15 days. The epidemic
-trajectories are shown in the \(state\) vs.~time plots in Figure
+trajectories are shown in the \(state\) vs.~\(time\) plots in Figure
 \ref{fig:different-scales-standard}. At a glance, we may \sout{assume}
@@
 
+\textcolor{violet}{[Ben says: another rewrite of the paragraph below.]}
+A time-invariant approach to visualizing epidemics, which will be
+described more below, provides a simple way to compare \(R_0\) values of
+epidemics that are generated from the Kermack and McKendrick SIR model.
+In these visuals, we treat epidemics as \textit{trajectories}. A
+trajectory can be mathematically viewed as a set of points is space that
+have an ordering, and that all points on the line between these order
+points are also contained in the geometric object. For a SIR epidemic,
+we can represent the associated filament \(\psi\) as
+
+\[
+\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
+\] where the an mapping \(\xi : t \to \mathbb{R}\) that is monotonically
+increasing would not change the definition of \(\psi\),
+i.e.~\(\psi_\xi = \psi\) where :
+
+\[
+\psi_\xi = \left \{(S(\xi(t)), I(\xi(t)), R(\xi(t))): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]}.
+\] {[}Since the number in each state is non-negative and the sum over
+the three states for a given time point sums to \(N\), then
+\textcolor{violet}{all points in} \(\psi\) will lay in a
+\textcolor{violet}{\sout{2d triangular plane, i.e. a ternary plot}}
+\textcolor{violet}{two-dimensional triangular plane in the three-dimensional space, which can be visualized in a two-dimensional ternary plot}.
+As a result, we can visualize the global trajectory of an epidemic in a
+single
+\textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
+\textcolor{orange}{[Shannon says: Show pic here?]}{]}
+
+\textcolor{violet}{[Ben says: this paragraph was replaced with the paragraph above]}
 However, time-invariant centric analysis allows us to visualize \(R_0\)
@@
 
+\textcolor{violet}{[This section could be a bit less wordy. But generally good.]}
 We visualize the two epidemics in a global, ternary view in Figure
@@
 filaments \(\psi\) seem to form the same trajectory. Now, it is much
-clearer that Model 2 is following the same trajectory as Model 1 but is
-not as far along in the infection process. We can see there is something
-fundamentally linking these two different epidemics, and this
+clearer that \textcolor{violet}{\sout{Model} epidemic} 2 is following
+the same trajectory as \textcolor{violet}{\sout{Model} epidemic} 1 but
+is not as far along in the infection process. We can see there is
+something fundamentally linking these two different epidemics, and this
 fundamental link turns out to be \(R_0\).
 
-We can show this mathematically if we let our two epidemics be presented
-as \(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
-\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
-example, assume both models have the same initial values
+We can show this mathematically\textcolor{violet}{.\sout{ if we} L}et
+our two epidemics be presented as
+\(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
+\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\)
+respectively.\footnote{\textcolor{violet}{[Ben says: Originally this was a sentence fragment.]}}
+As with the example, assume both models have the same initial values
 \((S(0), I(0), R(0))\), and let
@@
 each state and \(R_0\) then for every point on the epidemic path of the
-first SIR model is also a point on the epidemic path of the second SIR
-model.
+first SIR model \textcolor{violet}{\sout{is also} can be mapped to} a
+point on the epidemic path of the second SIR
+model.\textcolor{violet}{[We've changed how we describe these two objects when they are related to the theorem - is that desirable?]}
 
@@
 
-\subsection[Beyond R0 and SIR]{Beyond \(R_0\) and Kermack's and
-McKendrick SIR Models}\label{beyond-r0-sir}
-
-Through \(R_0\) \textcolor{orange}{example}, we see that treating
-epidemics more like a
+\subsection[Beyond R0 and SIR]{\textcolor{violet}{Time-invariant analysis}
+Beyond \(R_0\) and Kermack's and McKendrick SIR
+Models}\label{beyond-r0-sir}
+
+Through \textcolor{violet}{the} \(R_0\) \textcolor{orange}{example}, we
+see that treating epidemics \textcolor{violet}{\sout{more}} like a
 \sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)}
-\textcolor{orange}{filaments embedded in a lower dimensional space},
-then we can better compare the overall structure of the epidemic and
+\textcolor{orange}{filaments embedded in a lower dimensional space}\textcolor{violet}{\sout{, then we can} allows us to}
+better compare the overall structure of the epidemic and
 \sout{how it impacted the population}

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.Rmd
some updates to section 3's final parts.

@@
 
-Although this package does not focus on fitting a model to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
-These functions produce individual level information and can be naturally combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
+\ben{[Ben says: think about this section and if it highlights that we can bring in outside models...]}
+
+\ben{After getting a good sense of what a past or current epidemice looks like through EDA, the next step is often model fitting and/or simulations.
+In this step, and the next step (post-processing), we discuss how to include practioners' use of tools to create models and simulations that exist outside of \pkg{EpiCompare}.} This package does not focus on fitting a model to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
+\ben{After estimating or predicting transition rates between states,} these functions produce individual level \ben{simulations\sout{information}} and can be \ben{\sout{naturally}} combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
 The function `simulate_SIR_agents()` simulates an SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
@@
 
-Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.
+\ben{[Ben says: I think we should remind the reader that we care more about simulations, in order to compare fitted models between themselves and the true epidemics.]}
+
+\ben{[Ben says: this replaces the first paragraph below]}
+\ben{If the practitioner wishes to compare between models and with their true epidemics, one needs to process their models and simulations.} \ben{In general, [p}ost-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.]  \ben{The summaries can be very complex, and a}[s a result, a number of epidemic modeling \ben{\proglang{R}} packages return a special class, specific to their modeling.
+The special classes often contain a plethora of information from residuals, model diagnostics, input parameters, and more.
+While incredibly useful, these special classes can be difficult for novice coders to navigate.]
+
+\ben{[Ben says: replaced with above paragraph]} Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.
 For example, comma separated value files (`.csv`) are a  standard way to share information within tables.
@@
 
-To this end, we have adapted a series of `fortify`-style functions, called `fortify_aggregate()` which transform output from packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
+To this end, we have \ben{provide\sout{adapted}} a series of `fortify`-style \ben{methods\sout{functions}}\footnote{\ben{[Ben says: Not tied to this change.]}}, called `fortify_aggregate()` which transform output from packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
 These fortify functions have output that is consistent with that of `agents_to_aggregate()`.
 
-\shannon{Here Ben talks about filament compression and tidy_dist mats, etc}
+\shannon{[Shannon says: Here Ben talks about filament compression and tidy_dist mats, etc.]} \ben{[Ben says: not currently thinking about including tidy_dist stuff - will need to think about this as it is inlcudding in the pipeline 2 image...]}
+
+\ben{[Ben says: new paragraph]} To utilize simulations epidemics in later time-invariant analysis we also provide a function to convert temporally defined epidemics to filamental representations.
+Specifically, we provide the function `filament_compression` to convert simulation(s) to filaments as expressed by presenting the epidemic as a ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space\footnote{\ben{[Ben: this should be cleaned up.]}}.
 
 **Comparisons and Assessment**
+
+\ben{[This still needs to be worked on - just a reminder.]}
 

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.tex
some updates to section 3's final parts.

@@
 
-Although this package does not focus on fitting a model to data, we do
-provide some flexible functions for simulation of basic discrete-time
-epidemic-state models. These functions produce individual level
-information and can be naturally combined with
+\textcolor{violet}{[Ben says: think about this section and if it highlights that we can bring in outside models...]}
+
+\textcolor{violet}{After getting a good sense of what a past or current epidemice looks like through EDA, the next step is often model fitting and/or simulations.
+In this step, and the next step (post-processing), we discuss how to include practioners' use of tools to create models and simulations that exist outside of \pkg{EpiCompare}.}
+This package does not focus on fitting a model to data, we do provide
+some flexible functions for simulation of basic discrete-time
+epidemic-state models.
+\textcolor{violet}{After estimating or predicting transition rates between states,}
+these functions produce individual level
+\textcolor{violet}{simulations\sout{information}} and can be
+\textcolor{violet}{\sout{naturally}} combined with
 \texttt{agents\_to\_aggregate()} to view these simulations through an
@@
 
+\textcolor{violet}{[Ben says: I think we should remind the reader that we care more about simulations, in order to compare fitted models between themselves and the true epidemics.]}
+
+\textcolor{violet}{[Ben says: this replaces the first paragraph below]}
+
+\textcolor{violet}{If the practitioner wishes to compare between models and with their true epidemics, one needs to process their models and simulations.}
+\textcolor{violet}{In general, [p}ost-processing of modeling and
+simulation consists of making summary statistics, plots, tables, and
+other ways to disseminate information to the public.{]}
+\textcolor{violet}{The summaries can be very complex, and a}{[}s a
+result, a number of epidemic modeling \textcolor{violet}{\proglang{R}}
+packages return a special class, specific to their modeling. The special
+classes often contain a plethora of information from residuals, model
+diagnostics, input parameters, and more. While incredibly useful, these
+special classes can be difficult for novice coders to navigate.{]}
+
+\textcolor{violet}{[Ben says: replaced with above paragraph]}
 Post-processing of modeling and simulation consists of making summary
@@
 
-To this end, we have adapted a series of \texttt{fortify}-style
-functions, called \texttt{fortify\_aggregate()} which transform output
-from packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data
-frames which contain information about the total number of individuals
-in each state at a given time, for a given simulation. These fortify
-functions have output that is consistent with that of
+To this end, we have \textcolor{violet}{provide\sout{adapted}} a series
+of \texttt{fortify}-style
+\textcolor{violet}{methods\sout{functions}}\footnote{\textcolor{violet}{[Ben says: Not tied to this change.]}},
+called \texttt{fortify\_aggregate()} which transform output from
+packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
+which contain information about the total number of individuals in each
+state at a given time, for a given simulation. These fortify functions
+have output that is consistent with that of
 \texttt{agents\_to\_aggregate()}.
 
+\textcolor{violet}{[Ben says: new paragraph]} To utilize simulations
+epidemics in later time-invariant analysis we also provide a function to
+convert temporally defined epidemics to filamental representations.
+Specifically, we provide the function \texttt{filament\_compression} to
+convert simulation(s) to filaments as expressed by presenting the
+epidemic as a ordering of some common fixed number of points so that
+they are equally spaced along the original path in the proportional
+state
+space\footnote{\textcolor{violet}{[Ben: this should be cleaned up.]}}.
+
 \textbf{Comparisons and Assessment}
+
+\textcolor{violet}{[This still needs to be worked on - just a reminder.]}
 

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.Rmd
minor last comments on section 3

@@
 
+\ben{[Brainstorming: (intro) discuss model vs model comparison tools, (a) highlight that standard comparisons can be done if models share metrics or parameters (and similarly some comparisons can be done to the real epidemics).
+(b) Present time-invariance tools that allow us to compare more types of models (and can also give new tools to assess models beyond the standard ones).
+(1) introduce prediction regions (a) geom_prediction_bands,  (2) discuss different extremeness measures.]}
+
 Comparison and assessment of model fit or comparisons of one model to another model can be performed in a variety of ways including mean square error, AIC, plots, and more.
 Perhaps the most useful tool \pkg{EpiCompare} has to offer to the expert, for comparison and assessment of models, is in its post-processing tools which create a standard output.
-It is then a matter of writing a script or function made for that standard output to assess the results from multiple models in the way the user desires.
+It is then a matter of writing a script or function made for that standard output to assess the results from multiple models in the way the user desires\footnote{\ben{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}}.
 

### 2021-03-25 jss/EpiCompare-paper/EpiCompare-paper.tex
Auto stash before merge of "master" and "origin/master"
minor last comments on section 3

@@
 made for that standard output to assess the results from multiple models
-in the way the user desires.
+in the way the user
+desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}}.
 

### 2021-03-29 jss/EpiCompare-paper/EpiCompare-paper.tex
first draft of final section of part 2.

@@
 
-\textcolor{violet}{[Ben says: I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful.
-Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.]}
-
 The recent (and on-going) COVID-19 global pandemic has galvanized public
@@
 government, and news organizations are an important first step of the
-process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
-\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.]}
+process
+\textcolor{violet}{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}}
 
@@
 
-\textcolor{orange}{time invariance vs. time invariant analysis??}
-
-\textcolor{orange}{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
-The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot\footnote{\textcolor{violet}{[Ben says: defend how this fits into the narrative?]}} and 4) talk }\textcolor{violet}{Time invariance in}
-\sout{about}\textcolor{orange}{ models beyond SIR.}
-
-\textcolor{orange}{I noticed the ternary plot kinda fell out of this section and it's not really detailed anywhere else in the paper.
-So I put it back in.
-It's a non-standard plot and a feature of EpiCompare so I think it's worth talking about, and I think it's appropriate to introduce in this section.}
-
-\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by
-(1) taking a global, not marginal view of how epidemics move through
-populations and (2) by treating full epidemics as filaments\footnote{\textcolor{violet}{[Ben says: Do we need a footnote explaining what this is more clearly?]}}
-and not points produced by functions of time. The following section aims
-to highlight the strengths of \textit{time-invariance} and define the
-mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
-
-Mathematically, epidemics are complex objects. They are hard to assess
-and compare to one another due to the differences in the diseases, the
-location where the outbreak occurs, how the affected population reacts,
-and the time \textcolor{violet}{related} features (including start of
-the epidemic, speed of infection and more). Time-invariant analysis make
+\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking
+a global, not marginal view of how epidemics move through populations
+and (2) by treating full epidemics as filamental trajectories\footnote{\textcolor{violet}{[Ben says: Do we need a footnote explaining what this is more clearly?]}}
+and not points produced by functions of
+time\footnote{\textcolor{orange}{[Shannon says: this notation needs to be cleaned up.]}}
+(this will be explained in Section \ref{sec:r0}). The following section
+aims to highlight the strengths of \textit{time-invariance} and define
+the mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
+
+Mathematically, epidemics are complex objects. They can be hard to
+assess and compare to one another due to the differences in the
+diseases, the location where the outbreak occurs, how the affected
+population reacts, and the time related features (including start of the
+epidemic, speed of infection and more). Time-invariant analysis make
 different epidemics easier to compare by removing many time dependent
@@
 an epidemic, emphasizing the number of lives affected.
-\textcolor{orange}{[Shannon says: really like the lives affected part; I think that's a powerful statement.]}
 
 \subsection[motivating through R0]{Motivating time-invariance through
-\(R_0\)\footnote{\textcolor{violet}{Is this still a relevant title for this section?}}}\label{r0}
+\(R_0\)\footnote{\textcolor{violet}{Is this still a relevant title for this section? Think about this header after the rewrite.}}}\label{r0}
 
@@
 in the US to a 10 day outbreak of norovirus on a cruise ship is
-\sout{still} possible. Time-dependent problems can arise when estimation
-of certain epidemiological parameters, including the reproduction number
-\(R_0\).
-\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariaten analysis in this section.}
+\sout{still} possible. Time-dependent problems can arise when estimating
+epidemiological parameters, including the reproduction number \(R_0\).
+\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}
 
 \(R_0\) is probably the most famous time-invariant numerical summary of
-an epidemic and is \textcolor{orange}{often} associated with the
-Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.
-\textcolor{orange}{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a \textcolor{violet}{\sout{naive} completely susceptible}\footnote{\textcolor{violet}{[Ben says: I feel like naive is unclear - even if that's the technical term.]}} population \citep{anderson1992}.
-$R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}.
-}\textcolor{violet}{[Ben says: could we expand on this difficulty / show it really is difficult with 1 more sentence?]}
-To demonstrate the difficulty of discerning \(R_0\)in a time-dependent
-analysis, we \textcolor{orange}{first} introduce the
-\citet{Kermack1927}'s
-\textcolor{violet}{\sout{common}}\footnote{\textcolor{violet}{[Ben says: is it common? maybe just remove it?]}}
-SIR model. This model captures the transitions from one state to the
-next as a system of ordinary differential equations, where \(N\) is the
-total number of individuals, \(\beta\) is the rate of infection, and
-\(\gamma\) is the rate of recovery,
+an epidemic and is often associated with the
+Susceptible-Infectious-Recovered (SIR) model
+\citep{hethcote2001}\footnote{[Shannon: correct CITation]}. \(R_0\) is a
+one-number summary of a disease and is defined as the expected number of
+infections caused by a single infector who is added to a completely
+susceptible population \citep{anderson1992}. \(R_0\) can be a difficult
+quantity to estimate and is typically sensitive to time-based parameters
+such as the beginning and end of an epidemic \citep{Gallagher2020}.
+\textcolor{violet}{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence?
+It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}
+To demonstrate the difficulty of discerning \(R_0\) in a time-dependent
+analysis, we first introduce the \citet{Kermack1927}'s SIR model. This
+model captures the transitions from one state to the next as a system of
+ordinary differential equations, where \(N\) is the total number of
+individuals, \(\beta\) is the rate of infection, and \(\gamma\) is the
+rate of recovery,
 
@@
 
-From this model, \(R_0 = \beta/\gamma\), aka the ratio of the estimated
-infection rate compared to the estimated recovery rate. Since \(\beta\)
-and \(\gamma\) are both rates (of infection and recovery, respectively),
-the ratio of the two, \(R_0\), is a time-invariant quantity.
-\textcolor{orange}{Once $R_0$ is estimated, then important epidemic quantities can be inferred such as a the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic and the total number of infections, the final size.
-Moreover, $R_0$ allows us to compare different disease outbreaks and different outbreaks on the same scale.
-}
+From this model, \(R_0 = \beta/\gamma\), aka the ratio of the infection
+rate compared to the recovery rate. Since \(\beta\) and \(\gamma\) are
+both rates, the ratio of the two, \(R_0\), is a time-invariant quantity.
+Once \(R_0\) is estimated, practioners can infer important epidemic
+quantities such as the total number of infections or the percent of a
+population needed to be vacccinated to stop the sustained spread of an
+epidemic. Moreover, \(R_0\) allows us to compare different diseases and
+different instances of outbreaks on the same scale.
 
@@
 epidemics that are generated from the Kermack and McKendrick SIR model.
-In these visuals, we treat epidemics as \textit{trajectories}. A
-trajectory can be mathematically viewed as a set of points is space that
-have an ordering, and that all points on the line between these order
-points are also contained in the geometric object. For a SIR epidemic,
-we can represent the associated filament \(\psi\) as
+In these visuals, we treat epidemics as
+\textit{filamental trajectories}. A filamental trajectory can be
+mathematically viewed as a set of points is space that have an ordering,
+and that all points on the line between these order points are also
+contained in the geometric object. For a SIR epidemic, we can represent
+the associated filamental trajectory \(\psi\) as
 
@@
 \textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
-\textcolor{orange}{[Shannon says: Show pic here?]}{]}
+\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?}
 
 \textcolor{violet}{[Ben says: this paragraph was replaced with the paragraph above]}
-However, time-invariant centric analysis allows us to visualize \(R_0\)
-for the Kermack and McKendrick SIR model. We call the epidemic
-trajectory of number of individuals in each state a \textit{filament}.
-Mathematically, a filament \(\psi\) for the SIR model \(\psi\) this is
-the set of points of the number in each state for each time point, \[
+\sout{However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.
+We call the epidemic trajectory of number of individuals in each state a \textit{filament}.
+Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,}
+\[
 \psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
@@
 
-Since the number in each state is non-negative and the sum over the
-three states for a given time point sums to \(N\), then \(\psi\) will
-lay in a 2d triangular plane, i.e.~a ternary plot. As a result, we can
-visualize the global trajectory of an epidemic in a single 2d plot and
-ultimately \(R_0\). \textcolor{orange}{Show pic here?}
+\sout{Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then $\psi$ will lay in a 2d triangular plane, i.e.
+a ternary plot.
+As a result, we can visualize the global trajectory of an epidemic in a single 2d plot and ultimately $R_0$.
+\textcolor{orange}{Show pic here?}}
 
@@
 Beyond \(R_0\) and Kermack's and McKendrick SIR
-Models}\label{beyond-r0-sir}
-
-Through \textcolor{violet}{the} \(R_0\) \textcolor{orange}{example}, we
-see that treating epidemics \textcolor{violet}{\sout{more}} like a
-\sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)}
-\textcolor{orange}{filaments embedded in a lower dimensional space}\textcolor{violet}{\sout{, then we can} allows us to}
-better compare the overall structure of the epidemic and
-\sout{how it impacted the population}
-\textcolor{orange}{see how the population was directly impacted}.
-\sout{Without knowing the generative structure underlying these two epidemics},
-\textcolor{orange}{When the underlying generative model is unknown for epidemic trajectories},
-we propose that time-invariant tools can help compare them in smart
-ways, and that these ideas can extend past the CITE kermack1927 model
-and to epidemics with more states.
-
-\textcolor{violet}{
+Models\footnote{Probably will need to change this title...}}\label{beyond-r0-sir}
+
+Through the \(R_0\) example, we see that treating epidemics like a
+filamental trajectories embedded in a lower dimensional space allows us
+to better compare the overall structure of the epidemic and see how the
+population was directly impacted.
+\textcolor{violet}{In this section we will present time-invariant tools that can be useful even when the underlying generative model for the epidemic is unknown.
+We will also present how this tools can be used when the epidemics have more than 3 states.}
+\sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states. }
+
+\textcolor{violet}{New paragraph} Viewing epidemics as filamental
+trajectories provide a lot new ways to compare and examine epidemics in
+a time-invariant manner. For epidemics that have ``played'' themselves
+out, one way to represent their filamental trajectories to avoid impacts
+of temporal structure is to define them as a sequence of points their
+trajectory with equi-distance between each point. This also creates a
+natural distance between this type of representation between epidemics,
+specifically: \[
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
+\] where \(\psi_i'(s)\) the point along \(\psi_1\) that is \(s\)
+fraction of \(|\psi_1|\) distance away from the start of \(\psi_i\).
+This distance is naturally time-invariant, and can be plugged into
+multiple distance-based assessment tools to examine the overall
+extremeness of points, including psuedo-density estimators and
+depth/local depth functions
+\citep[for examples see][]{Ciollaro2016, Geenens2017}. This can be very
+useful when comparing between a set of simulation epidemics and the true
+epidemic, and does not constrain the number of states of the model,
+though we recommend projecting the points into the unit
+simplex\footnote{\textcolor{violet}{I'm not sure we've talked about this before...}}
+(by making all values the proportion of the population in the given
+state).
+
+\textcolor{violet}{New paragraph:} If the set of epidemics that one is
+examining have only gone through a single cycle of the outbreak (before
+the population of individuals have become susceptible again), then
+additional time-invariant tools, including prediction regions can we
+leveraged. In these setting we go a step farther and treat epidemics
+more like geometric filaments than filamental trajectories. In
+\pkg{EpiCompare} we create prediction regions that contain a the top
+(\(1-\alpha\)) proportion of simulated curves by defining geometric
+regions defined by the union of small filaments around the subset of
+simulations (subset by measures like the above psuedo-density estimates
+or depth estimates). These regions look at where in the state-space we
+expect the epidemic to traverse, and we can compare prediction regions
+defined by different models using the Haussdorf distance as well as
+examining how well the truth epidemic matches the simulations by
+examining if the epidemic's trajectory lies within the prediction
+region. All these geometrics structures and distances notations apply to
+epidemics with any number of states, and at the end of Section
+\ref{sec:overview} we also highlight how these prediction regions can
+aid in visual comparisons for epidemics with 3 states (like the SIR
+models).
+
+\textcolor{violet}{[Ben says: Old recommendations:
 to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset.
-Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.
-}
+Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.]}
 
@@
 \citep{Wickham2016, Hamilton2018}.
-\textcolor{orange}{[Shannon says: come back?]}\footnote{\textcolor{violet}{[Ben says: what does this note mean?]}}
+\textcolor{orange}{[Shannon says: come back - think about spaces and trajectories?]}\footnote{\textcolor{violet}{[Ben says: what does this note mean?]}}
 Visualization tools for three-state models were developed because (1)

### 2021-03-29 jss/EpiCompare-paper/EpiCompare-paper.Rmd

@@
 
-\ben{[Ben says: I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful.
-Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.]}
-
 The recent (and on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology.
@@
 historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}.
-At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
-\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.]}
+At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process \ben{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+\footnote{\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.
+*This is less so the case now.]}}
 
@@
 
-\shannon{time invariance vs. time invariant analysis??}
+<!-- \shannon{time invariance vs. time invariant analysis??} -->
 
-\shannon{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
-The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot\footnote{\ben{[Ben says: defend how this fits into the narrative?]}} and 4) talk }\ben{Time invariance in} \sout{about}\shannon{ models beyond SIR.}
+<!-- \shannon{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
+The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot\footnote{\ben{[Ben says: defend how this fits into the narrative?]}} and 4) talk }\ben{Time invariance in} \sout{about}\shannon{ models beyond SIR.} -->
 
-\shannon{I noticed the ternary plot kinda fell out of this section and it's not really detailed anywhere else in the paper.
+<!-- \shannon{I noticed the ternary plot kinda fell out of this section and it's not really detailed anywhere else in the paper.
 So I put it back in.
-It's a non-standard plot and a feature of EpiCompare so I think it's worth talking about, and I think it's appropriate to introduce in this section.}
+It's a non-standard plot and a feature of EpiCompare so I think it's worth talking about, and I think it's appropriate to introduce in this section.} -->
 
@@
 
-\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filaments^[\ben{[Ben says: Do we need a footnote explaining what this is more clearly?]}] and not points produced by functions of time.
+\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories^[\ben{[Ben says: Do we need a footnote explaining what this is more clearly?]}] and not points produced by functions of time\footnote{\shannon{[Shannon says: this notation needs to be cleaned up.]}} (this will be explained in Section \ref{sec:r0}).
 The following section aims to highlight the strengths of \textit{time-invariance} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
@@
 Mathematically, epidemics are complex objects.
-They are hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time \ben{related} features (including start of the epidemic, speed of infection and more).
+They can be hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time related features (including start of the epidemic, speed of infection and more).
 Time-invariant analysis make different epidemics easier to compare by removing many time dependent aspects of an epidemic.
 Instead it focuses on the overall trajectory of an epidemic, emphasizing the number of lives affected.
-\shannon{[Shannon says: really like the lives affected part; I think that's a powerful statement.]}
 
-## Motivating time-invariance through $R_0$^[\ben{Is this still a relevant title for this section?}] {short-title="motivating through R0" #r0}
+## Motivating time-invariance through $R_0$^[\ben{Is this still a relevant title for this section? Think about this header after the rewrite.}] {short-title="motivating through R0" #r0}
 
@@
 With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible.
-Time-dependent problems can arise when estimation of certain epidemiological parameters, including the reproduction number $R_0$.
-\ben{We will use $R_0$ to motivate the usefulness of time-invariaten analysis in this section.}
+Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$.
+\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}
 
@@
 
-$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is \shannon{often} associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}.
-\shannon{$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a \ben{\sout{naive} completely susceptible}\footnote{\ben{[Ben says: I feel like naive is unclear - even if that's the technical term.]}} population \citep{anderson1992}.
+$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}\footnote{[Shannon: correct CITation]}.
+$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.
 $R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}.
-}\ben{[Ben says: could we expand on this difficulty / show it really is difficult with 1 more sentence?]} To demonstrate the difficulty of discerning $R_0$in a time-dependent analysis, we \shannon{first} introduce the \citet{Kermack1927}’s \ben{\sout{common}}\footnote{\ben{[Ben says: is it common?
-maybe just remove it?]}} SIR model.
+\ben{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence?
+It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]} To demonstrate the difficulty of discerning $R_0$ in a time-dependent analysis, we first introduce the \citet{Kermack1927}’s SIR model.
 This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
@@
 
-From this model, $R_0 = \beta/\gamma$, aka the ratio of the estimated infection rate compared to the estimated recovery rate.
-Since $\beta$ and $\gamma$ are both rates (of infection and recovery, respectively), the ratio of the two, $R_0$, is a time-invariant quantity.
-\shannon{Once $R_0$ is estimated, then important epidemic quantities can be inferred such as a the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic and the total number of infections, the final size.
-Moreover, $R_0$ allows us to compare different disease outbreaks and different outbreaks on the same scale.
-}
+From this model, $R_0 = \beta/\gamma$, aka the ratio of the infection rate compared to the recovery rate.
+Since $\beta$ and $\gamma$ are both rates, the ratio of the two, $R_0$, is a time-invariant quantity.
+Once $R_0$ is estimated, practioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
+Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
 
@@
 \ben{[Ben says: another rewrite of the paragraph below.]} A time-invariant approach to visualizing epidemics, which will be described more below, provides a simple way to compare $R_0$ values of epidemics that are generated from the Kermack and McKendrick SIR model.
-In these visuals, we treat epidemics as \textit{trajectories}.
-A trajectory can be mathematically viewed as a set of points is space that have an ordering, and that all points on the line between these order points are also contained in the geometric object.
-For a SIR epidemic, we can represent the associated filament $\psi$ as
+In these visuals, we treat epidemics as \textit{filamental trajectories}.
+A filamental trajectory can be mathematically viewed as a set of points is space that have an ordering, and that all points on the line between these order points are also contained in the geometric object.
+For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as
 
@@
 As a result, we can visualize the global trajectory of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.
-\shannon{[Shannon says: Show pic here?]}]
+\shannon{[Shannon says: Show pic here?]}]\footnote{Which?}
 
-\ben{[Ben says: this paragraph was replaced with the paragraph above]} However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.
+\ben{[Ben says: this paragraph was replaced with the paragraph above]} \sout{However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.
 We call the epidemic trajectory of number of individuals in each state a \textit{filament}.
-Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,
+Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,}
 $$
@@
 
-Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then $\psi$ will lay in a 2d triangular plane, i.e.
+\sout{Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then $\psi$ will lay in a 2d triangular plane, i.e.
 a ternary plot.
 As a result, we can visualize the global trajectory of an epidemic in a single 2d plot and ultimately $R_0$.
-\shannon{Show pic here?}
+\shannon{Show pic here?}}
 
@@
 
-##  \ben{Time-invariant analysis} Beyond $R_0$ and Kermack’s and McKendrick SIR Models {short-title="Beyond R0 and SIR" #beyond-r0-sir}
+##  \ben{Time-invariant analysis} Beyond $R_0$ and Kermack’s and McKendrick SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #beyond-r0-sir}
 
-Through \ben{the} $R_0$ \shannon{example}, we see that treating epidemics \ben{\sout{more}} like a \sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)} \shannon{filaments embedded in a lower dimensional space}\ben{\sout{, then we can} allows us to} better compare the overall structure of the epidemic and \sout{how it impacted the population} \shannon{see how the population was directly impacted}.
-\sout{Without knowing the generative structure underlying these two epidemics}, \shannon{When the underlying generative model is unknown for epidemic trajectories}, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
+Through the $R_0$ example, we see that treating epidemics like a filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted.
+\ben{In this section we will present time-invariant tools that can be useful even when the underlying generative model for the epidemic is unknown.
+We will also present how this tools can be used when the epidemics have more than 3 states.} \sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
+}
 
-\ben{
+\ben{New paragraph} Viewing epidemics as filamental trajectories provide a lot new ways to compare and examine epidemics in a time-invariant manner.
+For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid impacts of temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point.
+This also creates a natural distance between this type of representation between epidemics, specifically:
+\[
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
+\]
+where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$.
+This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall extremeness of points, including psuedo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
+This can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model, though we recommend projecting the points into the unit simplex\footnote{\ben{I'm not sure we've talked about this before...}} (by making all values the proportion of the population in the given state).
+
+\ben{New paragraph:} If the set of epidemics that one is examining have only gone through a single cycle of the outbreak (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged.
+In these setting we go a step farther and treat epidemics more like geometric filaments than filamental trajectories.
+In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above psuedo-density estimates or depth estimates).
+These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using the Haussdorf distance as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
+All these geometrics structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we also highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).
+
+\ben{[Ben says: Old recommendations:
 to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset.
-Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.
-}
+Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.]}
 
@@
 Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our `geom_aggregate` provides a rapid way to explore different subpopulations' experience of an epidemic by combining the ideas behind `agents_to_aggregate` for three-state models to examine subpopulation trajectories in a 3d simplex space \citep{Wickham2016, Hamilton2018}.
-\shannon{[Shannon says: come back?]}\footnote{\ben{[Ben says: what does this note mean?]}} Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+\shannon{[Shannon says: come back - think about spaces and trajectories?]}\footnote{\ben{[Ben says: what does this note mean?]}} Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 \shannon{[Shannon says: make sure SIR is defined before.]}

### 2021-03-30 jss/EpiCompare-paper/EpiCompare-paper.Rmd
We are moving right along.  Continued editing Section 2.

@@
 
-\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories^[\ben{[Ben says: Do we need a footnote explaining what this is more clearly?]}] and not points produced by functions of time\footnote{\shannon{[Shannon says: this notation needs to be cleaned up.]}} (this will be explained in Section \ref{sec:r0}).
-The following section aims to highlight the strengths of \textit{time-invariance} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
+\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories and not points produced by functions of time \shannon{(these concepts will be explained in  explained in Section \ref{sec:r0})}.
+The following section aims to highlight the strengths of \textit{time-invariant} \shannon{analysis} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
 
@@
 They can be hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time related features (including start of the epidemic, speed of infection and more).
-Time-invariant analysis make different epidemics easier to compare by removing many time dependent aspects of an epidemic.
-Instead it focuses on the overall trajectory of an epidemic, emphasizing the number of lives affected.
+Time-invariant analysis makes different epidemics easier to compare by removing many time dependent aspects of an epidemic.
+Instead, \shannon{time-invariant analysis} focuses on the overall \shannon{shape and direction, a filamental trajctory,} \sout{trajectory of }an epidemic.
+\shannon{These filamental trajectories,  in turn,} emphasize the number of lives affected.
 
-## Motivating time-invariance through $R_0$^[\ben{Is this still a relevant title for this section? Think about this header after the rewrite.}] {short-title="motivating through R0" #r0}
+## Motivating time-invariant analysis through \shannon{the reproduction number} $R_0$^[\ben{Is this still a relevant title for this section?
+Think about this header after the rewrite.}] {short-title="motivating through R0" #sec:r0}
 
-Time-invariance analysis, as it appears in \pkg{EpiCompare}, \sout{can be seen as an attempt to escape} \shannon{bypasses} many difficulties \shannon{in} comparing different epidemics.
+Time-invariant analysis, as it appears in \pkg{EpiCompare}, \shannon{bypasses} many difficulties \shannon{in} comparing different epidemics.
 With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible.
 Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$.
-\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}
+\sout{\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}^[\shannon{I don't think this is a necessary sentence}]
 
@@
 
-$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}\footnote{[Shannon: correct CITation]}.
+$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
 $R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.
-$R_0$ can be a difficult quantity to estimate and is typically sensitive to time-based parameters such as the beginning and end of an epidemic \citep{Gallagher2020}.
-\ben{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence?
-It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]} To demonstrate the difficulty of discerning $R_0$ in a time-dependent analysis, we first introduce the \citet{Kermack1927}’s SIR model.
+\shannon{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
+However, data informing models are often temporally anchored (e.g.
+weekly case counts).
+As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.
+For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally do not have precise definitions.} \ben{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence?
+It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}\shannon{[Shannon says I added the few sentences to more smoothly introduce r0]} To demonstrate the difficulty of discerning $R_0$ in a time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model.
 This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
@@
 
-## \ben{Time invariant visualization of differences in $R_0}\sout{ Visualizing $R_0$} {short-title="visualizing r0" #vis-r0}
-
 \ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
@@
 In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
+\shannon{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
+also highlighted tie to TIA and r0.
+I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}
 
-\ben{[Ben says: replacement of following paragraph: presentation rearranged to emphasis the figure first (when presenting the problem), and then discussing it.]} Traditional $state$ vs.
-$time$ plots, like seen in Fig.
-\ref{fig:different-scales-standard}, present information about epidemics in a time-dependent and piece-wise fashion, and make comparing different epidemic $R_0$s impossible.
+\shannon{Time-invariant analysis helps practicioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
+Tne first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
+Both epidemics have populations of 1000 people with 10 individuals initially infected.
+Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.
+We plot the epidemics with traditional $state$ vs.
+$time$ plots.
 In Fig.
-\ref{fig:different-scales-standard} we compare [two different epidemics that start in two populations of 1000 people with 10 individuals initially infected, but with different infection and recovery rates (see Eq.
-\ref{eq:sir-ode})].
-More specifically, we've observed both epidemics for 15 days, with the first epidemic\footnote{In the plot it says model - probably want to say epidemic?} (in red) having parameters $\beta_1, \gamma_1 = (1,2)$ and the second epidemic (in blue) having  parameters $\beta_2, \gamma_2 = (2,4)$.
-[At a glance, we may believe that \sout{Model}\ben{epidemic} 1 has a larger $R_0$ than \sout{Model}\ben{epidemic} 2 because the peak of infection occurs more quickly than in Model 2 \ben{(remember $R_0$ relates to the rate of infection)}.
-On the other hand, we may believe \sout{Model}\ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15.
-Yet, both epidemics have the same $R_0$ of 2].
+\ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.
+In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.
+On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15.
+In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$.
 
-\ben{[Ben says: bring Fig. \ref{fig:different-scales-standard} here.]}
+```{r echo = FALSE}
+devtools::load_all()
+library(tidyr)
+library(dplyr)
+library(knitr)
+library(kableExtra)
+library(ggplot2)
+set.seed(1)
 
-\ben{[Ben says: this paragraph is replaced with the one above]}With regards to traditional epidemic $state$ vs.
-$time$ plots, when \sout{comparing}\shannon{examining} different epidemics their $R_0$s are often impossible to compare.
-This can be seen in a basic example, where we have two different epidemics that start in two populations of 1000 people with 10 individuals \shannon{initially} infected, but with \sout{different parameters} \shannon{different infection and recovery rates (see Eq.
-\ref{eq:sir-ode}}).
-Specifically suppose that $\beta_1, \gamma_1 = (a,b)$ and $\beta_2, \gamma_2 = (ca,cb)$ where $a, b, c > 0$.
-Further suppose we observe both of these epidemics for 15 days.
-The epidemic trajectories are shown in the $state$ vs.
-$time$ plots in Figure \ref{fig:different-scales-standard}.
-At a glance, we may \sout{assume} \shannon{believe} that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
-On the other hand, we may \sout{think} \shannon{believe} Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15.
-Yet, both epidemics have the same $R_0$ of 2.
+theme_set(theme_bw(base_size = 24))
+```
 
-\shannon{Introduce ternary plots here.  The below paragraphs are mostly new}
+```{r echo = FALSE, fig.cap = "\\label{fig:different-scales-standard}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both epidemics are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
+set.seed(1225)
+theme_set(theme_bw(base_size = 24))
 
-\ben{[Ben says: another rewrite of the paragraph below.]} A time-invariant approach to visualizing epidemics, which will be described more below, provides a simple way to compare $R_0$ values of epidemics that are generated from the Kermack and McKendrick SIR model.
-In these visuals, we treat epidemics as \textit{filamental trajectories}.
-A filamental trajectory can be mathematically viewed as a set of points is space that have an ordering, and that all points on the line between these order points are also contained in the geometric object.
+beta1 <- .8
+gamma1 <- .4
+scale <- .8
+sir1 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
+                            beta = beta1, gamma = gamma1,
+                            init_SIR = c(990, 10, 0)) %>%
+  group_by(sim) %>%
+  agents_to_aggregate(states = c(tI, tR)) %>%
+  group_by(t) %>%
+  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
+            .groups = "drop") %>%
+  mutate(type = "Epidemic 1")
+
+sir2 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
+                            beta = beta1 * scale, gamma = gamma1 * scale,
+                            init_SIR = c(990, 10, 0)) %>%
+  group_by(sim) %>%
+    agents_to_aggregate(states = c(tI, tR)) %>%
+  group_by(t) %>%
+  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
+            .groups = "drop") %>%
+  mutate(type = "Epidemic 2")
+
+df <- bind_rows(sir1, sir2)
+
+df_long <- df %>%
+         tidyr::pivot_longer(cols = c(S, I, R))  %>%
+  mutate(name = factor(name, levels = c("S", "I", "R"))) %>%
+  filter(t <= 15)
+
+ggplot(data = df_long,
+       aes(x = t, y = value, col = type)) + 
+  facet_wrap(~name, ncol = 1, scales = "free") +
+  geom_line() +
+  geom_point() +
+  scale_color_manual(values = c("red", "blue"), name = "Epidemic") +
+  labs(x = "Time",
+       y = "# in state",
+       title = latex2exp::TeX("Comparing two epidemics over first 15 days with same $R_0$")) +
+  coord_cartesian(xlim = c(0, 15))
+```
+
+\shannon{NEW} A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.  In Fig. \ref{fig:different-scales-tern}
+we plot the filamental trajectories of the two epidemics in a time-invariant view.
+We will explain how and why this works shortly.
+The important takeaway is that in this time-invariant view, it is apparent that these epidemics are on "the same path." In this case, this indicates that two epidemics have the same value of $R_0$.
+
+```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
+These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered."}
+ggplot(data = df %>%
+         filter(t <= 15)) +
+  geom_path(aes(x = S, y = I, z = R,
+                     col = type)) +
+  geom_point(aes(x = S, y = I, z = R,
+                     col = type)) +
+    coord_tern() + 
+  theme_sir() +
+  scale_color_manual(values = c("red", "blue"), name = "Epidemic") 
+```
+
+\shannon{The reason why we can visually compare $R_0$ in Fig.
+\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.} A filamental trajectory can be mathematically viewed as a set of points in space that have an ordering, and that all points on the line between these ordered points are also contained in the geometric object.
 For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as
@@
 $$
-where the an mapping $\xi : t \to \mathbb{R}$ that is monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi = \psi$ where :
+where a mapping $\xi : s \to \mathbb{R}$ that is monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi = \psi$ where :
 
 $$
-\psi_\xi = \left \{(S(\xi(t)), I(\xi(t)), R(\xi(t))): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]}.
+\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
 $$
-[Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a \ben{\sout{2d triangular plane, i.e.
-a ternary plot}} \ben{two-dimensional triangular plane in the three-dimensional space, which can be visualized in a two-dimensional ternary plot}.
-As a result, we can visualize the global trajectory of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.
-\shannon{[Shannon says: Show pic here?]}]\footnote{Which?}
+Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.
+\sout{We can then which can be visualize the \shannon{full filamental trajectory} in a two-dimensional ternary plot}  As a result, we can visualize \shannon{the full filamental trajectory} of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.
+\shannon{[Shannon says: Show pic here?]}]\footnote{Which?
+\shannon{3d space one?
+But I'm leaning against it now.
+3d never looks good in a paper.}}
 
-\ben{[Ben says: this paragraph was replaced with the paragraph above]} \sout{However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.
-We call the epidemic trajectory of number of individuals in each state a \textit{filament}.
-Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,}
-$$
-\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
-$$
-
-\sout{Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then $\psi$ will lay in a 2d triangular plane, i.e.
-a ternary plot.
-As a result, we can visualize the global trajectory of an epidemic in a single 2d plot and ultimately $R_0$.
-\shannon{Show pic here?}}
-
-\ben{[This section could be a bit less wordy.
+\sout{\ben{[This section could be a bit less wordy.
 But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
@@
 Now, it is much clearer that \ben{\sout{Model} epidemic} 2 is following the same trajectory as \ben{\sout{Model} epidemic} 1 but is not as far along in the infection process.
-We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
+}
 
-We can show this mathematically\ben{.\sout{ if we} L}et our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.\footnote{\ben{[Ben says: Originally this was a sentence fragment.]}}  As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
+The filamental trajectories in Fig.
+\ref{fig:different-scales-tern} seem to overlap, and we may suspect that something is fundamentally linking these two different epidemics together.
+Mathematically, we can show this fundamental link turns is $R_0$.
+Let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.
+As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
 And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.
@@
 The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model.\ben{[We've changed how we describe these two objects when they are related to the theorem - is that desirable?]}
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model.
+\shannon{In other words, the two epidemics form the same filamental trajectory.
+For SIR models with similar initial state percentages, time-invariant analysis, time-invariant analysis allows practicioners to compare values of $R_0$ at a glance.}  \ben{[We've changed how we describe these two objects when they are related to the theorem - is that desirable?
+]}\shannon{added that sentence to for that}
 
@@
 
-```{r echo = FALSE}
-devtools::load_all()
-library(tidyr)
-library(dplyr)
-library(knitr)
-library(kableExtra)
-library(ggplot2)
-set.seed(1)
-
-theme_set(theme_bw(base_size = 24))
-```
-
-```{r echo = FALSE, fig.cap = "\\label{fig:different-scales-standard}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
-Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
-set.seed(1225)
-theme_set(theme_bw(base_size = 24))
-
-beta1 <- .8
-gamma1 <- .4
-scale <- .8
-sir1 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
-                            beta = beta1, gamma = gamma1,
-                            init_SIR = c(990, 10, 0)) %>%
-  group_by(sim) %>%
-  agents_to_aggregate(states = c(tI, tR)) %>%
-  group_by(t) %>%
-  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
-            .groups = "drop") %>%
-  mutate(type = "Model 1")
-
-sir2 <- simulate_SIR_agents(n_sims = 100, n_time_steps = 40,
-                            beta = beta1 * scale, gamma = gamma1 * scale,
-                            init_SIR = c(990, 10, 0)) %>%
-  group_by(sim) %>%
-    agents_to_aggregate(states = c(tI, tR)) %>%
-  group_by(t) %>%
-  summarize(S = mean(X0), I = mean(X1), R = mean(X2),
-            .groups = "drop") %>%
-  mutate(type = "Model 2")
-
-df <- bind_rows(sir1, sir2)
-
-df_long <- df %>%
-         tidyr::pivot_longer(cols = c(S, I, R))  %>%
-  mutate(name = factor(name, levels = c("S", "I", "R"))) %>%
-  filter(t <= 15)
-
-ggplot(data = df_long,
-       aes(x = t, y = value, col = type)) + 
-  facet_wrap(~name, ncol = 1, scales = "free") +
-  geom_line() +
-  geom_point() +
-  scale_color_manual(values = c("red", "blue"), name = "Epidemic") +
-  labs(x = "Time",
-       y = "# in state",
-       title = latex2exp::TeX("Comparing two epidemics over first 15 days with same $R_0$")) +
-  coord_cartesian(xlim = c(0, 15))
-```
-
 <!--
@@
 
-```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
-Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
-These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered."}
-ggplot(data = df %>%
-         filter(t <= 15)) +
-  geom_path(aes(x = S, y = I, z = R,
-                     col = type)) +
-  geom_point(aes(x = S, y = I, z = R,
-                     col = type)) +
-    coord_tern() + 
-  theme_sir() +
-  scale_color_manual(values = c("red", "blue"), name = "Epidemic") 
-```
-
 ##  \ben{Time-invariant analysis} Beyond $R_0$ and Kermack’s and McKendrick SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #beyond-r0-sir}
 
-Through the $R_0$ example, we see that treating epidemics like a filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted.
-\ben{In this section we will present time-invariant tools that can be useful even when the underlying generative model for the epidemic is unknown.
-We will also present how this tools can be used when the epidemics have more than 3 states.} \sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
+Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted.
+\ben{\sout{In this section we will present} time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown \shannon{or have more than three epidemic states}.
+\sout{We will also present how this tools can be used when the epidemics have more than 3 states.} \sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
 }
 
-\ben{New paragraph} Viewing epidemics as filamental trajectories provide a lot new ways to compare and examine epidemics in a time-invariant manner.
-For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid impacts of temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point.
-This also creates a natural distance between this type of representation between epidemics, specifically:
+\ben{New paragraph} Viewing epidemics as filamental trajectories provide \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner.
+For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{are we missing a few words in this definition?}.
+This \shannon{representation induces}\sout{ also creates} a natural distance between this type of representation between epidemics, specifically:
 \[
@@
 This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall extremeness of points, including psuedo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
-This can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model, though we recommend projecting the points into the unit simplex\footnote{\ben{I'm not sure we've talked about this before...}} (by making all values the proportion of the population in the given state).
+This \shannon{what is this?} can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model, though we recommend projecting the points into the unit simplex\footnote{\ben{I'm not sure we've talked about this before...
+\shannon{I don't think we have but am wondering if we're getting in the weeds}}} (by making all values the proportion of the population in the given state).
 
-\ben{New paragraph:} If the set of epidemics that one is examining have only gone through a single cycle of the outbreak (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged.
-In these setting we go a step farther and treat epidemics more like geometric filaments than filamental trajectories.
-In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above psuedo-density estimates or depth estimates).
-These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using the Haussdorf distance as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
-All these geometrics structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we also highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).
+\ben{New paragraph:} If the set of epidemics that one is examining have only gone through a single cycle of the outbreak (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?}}].
+In these settings we go a step farther and treat epidemics more like geometric filaments \shannon{(i.e.
+filamental trajectories without an ordering of points)} than filamental trajectories.
+In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above pseudo-density estimates or depth estimates).
+These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using the Haussdorf \shannon{why Haussdorf specifically?} distance as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
+All these geometric structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).
 
@@
 Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.]}
+
+\shannon{I like your completion.  I think we can make it flow a bit more smoothly but the content we want to talk about is all there.}
 

### 2021-03-30 jss/EpiCompare-paper/EpiCompare-paper.tex
We are moving right along.  Continued editing Section 2.

@@
 a global, not marginal view of how epidemics move through populations
-and (2) by treating full epidemics as filamental trajectories\footnote{\textcolor{violet}{[Ben says: Do we need a footnote explaining what this is more clearly?]}}
-and not points produced by functions of
-time\footnote{\textcolor{orange}{[Shannon says: this notation needs to be cleaned up.]}}
-(this will be explained in Section \ref{sec:r0}). The following section
-aims to highlight the strengths of \textit{time-invariance} and define
-the mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
+and (2) by treating full epidemics as filamental trajectories and not
+points produced by functions of time
+\textcolor{orange}{(these concepts will be explained in  explained in Section \ref{sec:r0})}.
+The following section aims to highlight the strengths of
+\textit{time-invariant} \textcolor{orange}{analysis} and define the
+mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
 
@@
 population reacts, and the time related features (including start of the
-epidemic, speed of infection and more). Time-invariant analysis make
+epidemic, speed of infection and more). Time-invariant analysis makes
 different epidemics easier to compare by removing many time dependent
-aspects of an epidemic. Instead it focuses on the overall trajectory of
-an epidemic, emphasizing the number of lives affected.
-
-\subsection[motivating through R0]{Motivating time-invariance through
-\(R_0\)\footnote{\textcolor{violet}{Is this still a relevant title for this section? Think about this header after the rewrite.}}}\label{r0}
-
-Time-invariance analysis, as it appears in \pkg{EpiCompare},
-\sout{can be seen as an attempt to escape} \textcolor{orange}{bypasses}
-many difficulties \textcolor{orange}{in} comparing different epidemics.
-With time-invariant analysis, comparing the decades-long outbreak of HIV
-in the US to a 10 day outbreak of norovirus on a cruise ship is
-\sout{still} possible. Time-dependent problems can arise when estimating
-epidemiological parameters, including the reproduction number \(R_0\).
-\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}
+aspects of an epidemic. Instead,
+\textcolor{orange}{time-invariant analysis} focuses on the overall
+\textcolor{orange}{shape and direction, a filamental trajctory,}
+\sout{trajectory of }an epidemic.
+\textcolor{orange}{These filamental trajectories,  in turn,} emphasize
+the number of lives affected.
+
+\subsection[motivating through R0]{Motivating time-invariant analysis
+through \textcolor{orange}{the reproduction number}
+\(R_0\)\footnote{\textcolor{violet}{Is this still a relevant title for this section? Think about this header after the rewrite.}}}\label{sec:r0}
+
+Time-invariant analysis, as it appears in \pkg{EpiCompare},
+\textcolor{orange}{bypasses} many difficulties \textcolor{orange}{in}
+comparing different epidemics. With time-invariant analysis, comparing
+the decades-long outbreak of HIV in the US to a 10 day outbreak of
+norovirus on a cruise ship is \sout{still} possible. Time-dependent
+problems can arise when estimating epidemiological parameters, including
+the reproduction number \(R_0\).
+\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence}}
 
@@
 an epidemic and is often associated with the
-Susceptible-Infectious-Recovered (SIR) model
-\citep{hethcote2001}\footnote{[Shannon: correct CITation]}. \(R_0\) is a
-one-number summary of a disease and is defined as the expected number of
-infections caused by a single infector who is added to a completely
-susceptible population \citep{anderson1992}. \(R_0\) can be a difficult
-quantity to estimate and is typically sensitive to time-based parameters
-such as the beginning and end of an epidemic \citep{Gallagher2020}.
+Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
+\(R_0\) is a one-number summary of a disease and is defined as the
+expected number of infections caused by a single infector who is added
+to a completely susceptible population \citep{anderson1992}.
+\textcolor{orange}{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
+However, data informing models are often temporally anchored (e.g.
+weekly case counts).
+As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.
+For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally do not have precise definitions.}
 \textcolor{violet}{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence?
-It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}
+It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}\textcolor{orange}{[Shannon says I added the few sentences to more smoothly introduce r0]}
 To demonstrate the difficulty of discerning \(R_0\) in a time-dependent
-analysis, we first introduce the \citet{Kermack1927}'s SIR model. This
-model captures the transitions from one state to the next as a system of
+analysis, we first introduce \citet{Kermack1927}'s SIR model. This model
+captures the transitions from one state to the next as a system of
 ordinary differential equations, where \(N\) is the total number of
@@
 
-\subsection[visualizing r0]{\sout{ Visualizing $R_0$}}\label{vis-r0}
-
 \textcolor{violet}{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
@@
 In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
-
-\textcolor{violet}{[Ben says: replacement of following paragraph: presentation rearranged to emphasis the figure first (when presenting the problem), and then discussing it.]}
-Traditional \(state\) vs.~\(time\) plots, like seen in Fig.
-\ref{fig:different-scales-standard}, present information about epidemics
-in a time-dependent and piece-wise fashion, and make comparing different
-epidemic \(R_0\)s impossible. In Fig.
-\ref{fig:different-scales-standard} we compare {[}two different
-epidemics that start in two populations of 1000 people with 10
-individuals initially infected, but with different infection and
-recovery rates (see Eq. \ref{eq:sir-ode}){]}. More specifically, we've
-observed both epidemics for 15 days, with the first
-epidemic\footnote{In the plot it says model - probably want to say epidemic?}
-(in red) having parameters \(\beta_1, \gamma_1 = (1,2)\) and the second
-epidemic (in blue) having parameters \(\beta_2, \gamma_2 = (2,4)\).
-{[}At a glance, we may believe that
-\sout{Model}\textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
-\sout{Model}\textcolor{violet}{epidemic} 2 because the peak of infection
-occurs more quickly than in Model 2
-\textcolor{violet}{(remember $R_0$ relates to the rate of infection)}.
-On the other hand, we may believe
-\sout{Model}\textcolor{violet}{epidemic} 2 has a larger \(R_0\) because
-its unclear if the number of infections in that
+\textcolor{orange}{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
+also highlighted tie to TIA and r0.
+I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}
+
+\textcolor{orange}{Time-invariant analysis helps practicioners to more easily compare $R_0$ from different outbreaks.}
+For example, consider two epidemics generated from the Kermack and
+McKendrick SIR equations. Tne first epidemic has parameters
+\(\beta_1, \gamma_1 = (0.8,0.4)\) and the second has
+\(\beta_2, \gamma_2 = (0.64,0.32)\). Both epidemics have populations of
+1000 people with 10 individuals initially infected. Additionally note
+that the two reproduction numbers are the same for each epidemic,
+\(R_0 = 2=0.8/0.4 = 0.64/0.32\). We plot the epidemics with traditional
+\(state\) vs.~\(time\) plots. In Fig.
+\ref{fig:different-scales-standard} we show the time-based paths for the
+\(S\), \(I\), and \(R\) states for the first 15 days of observed data.
+In this time-variant view, we may believe that
+\textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
+\textcolor{violet}{epidemic} 2 because the peak of infection occurs more
+quickly than in Epidemic 2. On the other hand, we may believe
+\textcolor{violet}{epidemic} 2 has a larger \(R_0\) because its unclear
+if the number of infections in that
 \sout{model}\textcolor{violet}{epidemic} has not yet peaked at time 15.
-Yet, both epidemics have the same \(R_0\) of 2{]}.
-
-\textcolor{violet}{[Ben says: bring Fig. \ref{fig:different-scales-standard} here.]}
-
-\textcolor{violet}{[Ben says: this paragraph is replaced with the one above]}With
-regards to traditional epidemic \(state\) vs.~\(time\) plots, when
-\sout{comparing}\textcolor{orange}{examining} different epidemics their
-\(R_0\)s are often impossible to compare. This can be seen in a basic
-example, where we have two different epidemics that start in two
-populations of 1000 people with 10 individuals
-\textcolor{orange}{initially} infected, but with
-\sout{different parameters}
-\textcolor{orange}{different infection and recovery rates (see Eq. \ref{eq:sir-ode}}).
-Specifically suppose that \(\beta_1, \gamma_1 = (a,b)\) and
-\(\beta_2, \gamma_2 = (ca,cb)\) where \(a, b, c > 0\). Further suppose
-we observe both of these epidemics for 15 days. The epidemic
-trajectories are shown in the \(state\) vs.~\(time\) plots in Figure
-\ref{fig:different-scales-standard}. At a glance, we may \sout{assume}
-\textcolor{orange}{believe} that Model 1 has a larger \(R_0\) than Model
-2 because the peak of infection occurs more quickly than in Model 2. On
-the other hand, we may \sout{think} \textcolor{orange}{believe} Model 2
-has a larger \(R_0\) because we may think the number of infections in
-that model has not yet peaked at time 15. Yet, both epidemics have the
-same \(R_0\) of 2.
-
-\textcolor{orange}{Introduce ternary plots here.  The below paragraphs are mostly new}
-
-\textcolor{violet}{[Ben says: another rewrite of the paragraph below.]}
-A time-invariant approach to visualizing epidemics, which will be
-described more below, provides a simple way to compare \(R_0\) values of
-epidemics that are generated from the Kermack and McKendrick SIR model.
-In these visuals, we treat epidemics as
-\textit{filamental trajectories}. A filamental trajectory can be
-mathematically viewed as a set of points is space that have an ordering,
-and that all points on the line between these order points are also
-contained in the geometric object. For a SIR epidemic, we can represent
-the associated filamental trajectory \(\psi\) as
+In this time-variant view, we cannot determine if one epidemic has
+larger value of \(R_0\).
+
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-2-1}
+
+}
+
+\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both epidemics are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
+\end{figure}
+\end{CodeChunk}
+
+\textcolor{orange}{NEW} A time-invariant approach to visualizing
+epidemics, in comparison, allows us to directly compare \(R_0\) from a
+single plot. In Fig. \ref{fig:different-scales-tern} we plot the
+filamental trajectories of the two epidemics in a time-invariant view.
+We will explain how and why this works shortly. The important takeaway
+is that in this time-invariant view, it is apparent that these epidemics
+are on ``the same path.'' In this case, this indicates that two
+epidemics have the same value of \(R_0\).
+
+\begin{CodeChunk}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-3-1}
+
+}
+
+\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
+These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
+\end{figure}
+\end{CodeChunk}
+
+\textcolor{orange}{The reason why we can visually compare $R_0$ in Fig.
+\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}
+A filamental trajectory can be mathematically viewed as a set of points
+in space that have an ordering, and that all points on the line between
+these ordered points are also contained in the geometric object. For a
+SIR epidemic, we can represent the associated filamental trajectory
+\(\psi\) as
 
@@
 \psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
-\] where the an mapping \(\xi : t \to \mathbb{R}\) that is monotonically
+\] where a mapping \(\xi : s \to \mathbb{R}\) that is monotonically
 increasing would not change the definition of \(\psi\),
@@
 \[
-\psi_\xi = \left \{(S(\xi(t)), I(\xi(t)), R(\xi(t))): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]}.
-\] {[}Since the number in each state is non-negative and the sum over
-the three states for a given time point sums to \(N\), then
+\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
+\] Since the number in each state is non-negative and the sum over the
+three states for a given time point sums to \(N\), then
 \textcolor{violet}{all points in} \(\psi\) will lay in a
-\textcolor{violet}{\sout{2d triangular plane, i.e. a ternary plot}}
-\textcolor{violet}{two-dimensional triangular plane in the three-dimensional space, which can be visualized in a two-dimensional ternary plot}.
-As a result, we can visualize the global trajectory of an epidemic in a
+\textbackslash ben\{two-dimensional triangular plane in
+three-dimensional space.
+\sout{We can then which can be visualize the \textcolor{orange}{full filamental trajectory} in a two-dimensional ternary plot}
+As a result, we can visualize
+\textcolor{orange}{the full filamental trajectory} of an epidemic in a
 single
 \textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
-\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?}
-
-\textcolor{violet}{[Ben says: this paragraph was replaced with the paragraph above]}
-\sout{However, time-invariant centric analysis allows us to visualize $R_0$ for the Kermack and McKendrick SIR model.
-We call the epidemic trajectory of number of individuals in each state a \textit{filament}.
-Mathematically, a filament $\psi$ for the SIR model  $\psi$ this is the set of points of the number in each state for each time point,}
-\[
-\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
-\]
-
-\sout{Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then $\psi$ will lay in a 2d triangular plane, i.e.
-a ternary plot.
-As a result, we can visualize the global trajectory of an epidemic in a single 2d plot and ultimately $R_0$.
-\textcolor{orange}{Show pic here?}}
-
-\textcolor{violet}{[This section could be a bit less wordy. But generally good.]}
-We visualize the two epidemics in a global, ternary view in Figure
-\ref{fig:different-scales-tern}. Without getting into too much detail of
-the intricacies of this plot, we immediately see the points of the two
-filaments \(\psi\) seem to form the same trajectory. Now, it is much
-clearer that \textcolor{violet}{\sout{Model} epidemic} 2 is following
-the same trajectory as \textcolor{violet}{\sout{Model} epidemic} 1 but
-is not as far along in the infection process. We can see there is
-something fundamentally linking these two different epidemics, and this
-fundamental link turns out to be \(R_0\).
-
-We can show this mathematically\textcolor{violet}{.\sout{ if we} L}et
-our two epidemics be presented as
+\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?  \textcolor{orange}{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}
+
+\sout{\textcolor{violet}{[This section could be a bit less wordy.
+But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
+Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.
+Now, it is much clearer that \textcolor{violet}{\sout{Model} epidemic} 2 is following the same trajectory as \textcolor{violet}{\sout{Model} epidemic} 1 but is not as far along in the infection process.
+}
+
+The filamental trajectories in Fig. \ref{fig:different-scales-tern} seem
+to overlap, and we may suspect that something is fundamentally linking
+these two different epidemics together. Mathematically, we can show this
+fundamental link turns is \(R_0\). Let our two epidemics be presented as
 \(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
-\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\)
-respectively.\footnote{\textcolor{violet}{[Ben says: Originally this was a sentence fragment.]}}
-As with the example, assume both models have the same initial values
+\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
+example, assume both models have the same initial values
 \((S(0), I(0), R(0))\), and let
@@
 first SIR model \textcolor{violet}{\sout{is also} can be mapped to} a
-point on the epidemic path of the second SIR
-model.\textcolor{violet}{[We've changed how we describe these two objects when they are related to the theorem - is that desirable?]}
-
-\begin{CodeChunk}
-\begin{figure}[H]
-
-{\centering \includegraphics{Figs/unnamed-chunk-2-1}
-
-}
-
-\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
-Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}[H]
-
-{\centering \includegraphics{Figs/unnamed-chunk-3-1}
-
-}
-
-\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
-Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
-These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
-\end{figure}
-\end{CodeChunk}
+point on the epidemic path of the second SIR model.
+\textcolor{orange}{In other words, the two epidemics form the same filamental trajectory.
+For SIR models with similar initial state percentages, time-invariant analysis, time-invariant analysis allows practicioners to compare values of $R_0$ at a glance.}
+\textcolor{violet}{[We've changed how we describe these two objects when they are related to the theorem - is that desirable? ]}\textcolor{orange}{added that sentence to for that}
 
@@
 
-Through the \(R_0\) example, we see that treating epidemics like a
+Through the \(R_0\) example, we see that treating epidemics like
 filamental trajectories embedded in a lower dimensional space allows us
@@
 population was directly impacted.
-\textcolor{violet}{In this section we will present time-invariant tools that can be useful even when the underlying generative model for the epidemic is unknown.
-We will also present how this tools can be used when the epidemics have more than 3 states.}
+\textbackslash ben\{\sout{In this section we will present}
+time-invariant tools \sout{that} can be useful even when the underlying
+generative model for the epidemic is unknown
+\textcolor{orange}{or have more than three epidemic states}.
+\sout{We will also present how this tools can be used when the epidemics have more than 3 states.}
 \sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states. }
@@
 \textcolor{violet}{New paragraph} Viewing epidemics as filamental
-trajectories provide a lot new ways to compare and examine epidemics in
-a time-invariant manner. For epidemics that have ``played'' themselves
-out, one way to represent their filamental trajectories to avoid impacts
-of temporal structure is to define them as a sequence of points their
-trajectory with equi-distance between each point. This also creates a
-natural distance between this type of representation between epidemics,
+trajectories provide \sout{a lot }new ways to compare and examine
+epidemics in a time-invariant manner. For epidemics that have ``played''
+themselves out, one way to represent their filamental trajectories to
+avoid \textcolor{orange}{confusion stemming from}\sout{impacts of}
+temporal structure is to define them as a sequence of points their
+trajectory with equi-distance between each point
+\textcolor{orange}{are we missing a few words in this definition?}. This
+\textcolor{orange}{representation induces}\sout{ also creates} a natural
+distance between this type of representation between epidemics,
 specifically: \[
@@
 depth/local depth functions
-\citep[for examples see][]{Ciollaro2016, Geenens2017}. This can be very
-useful when comparing between a set of simulation epidemics and the true
-epidemic, and does not constrain the number of states of the model,
-though we recommend projecting the points into the unit
-simplex\footnote{\textcolor{violet}{I'm not sure we've talked about this before...}}
+\citep[for examples see][]{Ciollaro2016, Geenens2017}. This
+\textcolor{orange}{what is this?} can be very useful when comparing
+between a set of simulation epidemics and the true epidemic, and does
+not constrain the number of states of the model, though we recommend
+projecting the points into the unit
+simplex\footnote{\textcolor{violet}{I'm not sure we've talked about this before... \textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}
 (by making all values the proportion of the population in the given
@@
 additional time-invariant tools, including prediction regions can we
-leveraged. In these setting we go a step farther and treat epidemics
-more like geometric filaments than filamental trajectories. In
-\pkg{EpiCompare} we create prediction regions that contain a the top
-(\(1-\alpha\)) proportion of simulated curves by defining geometric
-regions defined by the union of small filaments around the subset of
-simulations (subset by measures like the above psuedo-density estimates
-or depth estimates). These regions look at where in the state-space we
-expect the epidemic to traverse, and we can compare prediction regions
-defined by different models using the Haussdorf distance as well as
-examining how well the truth epidemic matches the simulations by
+leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done?}\}}.
+In these settings we go a step farther and treat epidemics more like
+geometric filaments
+\textcolor{orange}{(i.e. filamental trajectories without an ordering of points)}
+than filamental trajectories. In \pkg{EpiCompare} we create prediction
+regions that contain a the top (\(1-\alpha\)) proportion of simulated
+curves by defining geometric regions defined by the union of small
+filaments around the subset of simulations (subset by measures like the
+above pseudo-density estimates or depth estimates). These regions look
+at where in the state-space we expect the epidemic to traverse, and we
+can compare prediction regions defined by different models using the
+Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance as
+well as examining how well the truth epidemic matches the simulations by
 examining if the epidemic's trajectory lies within the prediction
-region. All these geometrics structures and distances notations apply to
+region. All these geometric structures and distances notations apply to
 epidemics with any number of states, and at the end of Section
-\ref{sec:overview} we also highlight how these prediction regions can
-aid in visual comparisons for epidemics with 3 states (like the SIR
+\ref{sec:overview} we \sout{also} highlight how these prediction regions
+can aid in visual comparisons for epidemics with 3 states (like the SIR
 models).
@@
 Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.]}
+
+\textcolor{orange}{I like your completion.  I think we can make it flow a bit more smoothly but the content we want to talk about is all there.}
 

### 2021-03-30 jss/EpiCompare-paper/EpiCompare-paper.Rmd
attempts at section 3

--- 2021-03-29 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
 
-\ben{[Brainstorming: (intro) discuss model vs model comparison tools, (a) highlight that standard comparisons can be done if models share metrics or parameters (and similarly some comparisons can be done to the real epidemics).
+<!-- \ben{[Brainstorming: (intro) discuss model vs model comparison tools, (a) highlight that standard comparisons can be done if models share metrics or parameters (and similarly some comparisons can be done to the real epidemics).
 (b) Present time-invariance tools that allow us to compare more types of models (and can also give new tools to assess models beyond the standard ones).
-(1) introduce prediction regions (a) geom_prediction_bands,  (2) discuss different extremeness measures.]}
+(1) introduce prediction regions (a) geom_prediction_bands,  (2) discuss different extremeness measures.]} -->
 
-Comparison and assessment of model fit or comparisons of one model to another model can be performed in a variety of ways including mean square error, AIC, plots, and more.
+As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of potential for time-invariant tools to help compare and assess epidemics and models/simulations.
+In \pkg{EpiCompare} we provide a set of comparison and assessment tools for models that extend beyond the standard performance metrics (e.g.
+means squared error, AIC) and focus on assessing the structural information the models capture.
+This approaches work well on models where online one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again).
+Epidemics are complex objects, and we provide tools to create prediction regions out of simulated epidemics.
+For 3-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations (where the simulations treated as filaments).
+In \pkg{Epicompare} we also provide these prediction regions for epidemic models with with more than 3 states.
+With `create_convex_hull_structure` and ``create_delta_ball_structure` that will create different geometric representations of prediction regions and an assoicated `contained` method that assesses if a new epidemic (seen as a filament) is contained in the prediction region.
+
+We also provide functions to calculate the extremeness of a true epidemic compared to simulated epidemics through the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
+Specifically, functions like  `distance_psuedo_density_function` can calculate a psuedo-density estimate of the true epidemic relative to simulated ones.
+Functions `distance_depth_function` and `local_distance_depth_function` provide depth scores that suggest how geometrically central an epidemic is to simulations.
+
+\ben{[Ben says: Replaced with paragraphs above.]} Comparison and assessment of model fit or comparisons of one model to another model can be performed in a variety of ways including mean square error, AIC, plots, and more.
 Perhaps the most useful tool \pkg{EpiCompare} has to offer to the expert, for comparison and assessment of models, is in its post-processing tools which create a standard output.
@@
 
-However, for those who like more concrete tools, \pkg{EpiCompare} offers functions to compare prediction regions to one another including `geom_prediction_band()` (which plots the region), and `create_{convex_hull,delta_ball}_structure()`(which returns the `R` output for the given structure), and `contained()` (which allows the user to determine if one set of points is contained in a prediction band).
+\ben{[Ben says: Replaced with paragraphs above.]} However, for those who like more concrete tools, \pkg{EpiCompare} offers functions to compare prediction regions to one another including `geom_prediction_band()` (which plots the region), and `create_{convex_hull,delta_ball}_structure()`(which returns the `R` output for the given structure), and `contained()` (which allows the user to determine if one set of points is contained in a prediction band).
 Additionally, we offer ways to determine if model outputs are compatible with one another, that is how extreme one output is to another.

### 2021-03-31 jss/EpiCompare-paper/EpiCompare-paper.Rmd
some section 2 edits plus add of plotly vis

@@
 
-## Motivating time-invariant analysis through \shannon{the reproduction number} $R_0$^[\ben{Is this still a relevant title for this section?
-Think about this header after the rewrite.}] {short-title="motivating through R0" #sec:r0}
+## Motivating time-invariant analysis through the reproduction number $R_0$ {short-title="motivating through R0" #sec:r0}
 
@@
 Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$.
-\sout{\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}^[\shannon{I don't think this is a necessary sentence}]
+\sout{\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}^[\shannon{I don't think this is a necessary sentence.} \ben{I still think it adds value to the story and I'm not sure people really read section titles that are long.}]
 
@@
 $R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.
-\shannon{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
+\ben{This means that $R_0$ is a time-invariant parameter yet is estimated with time-based data, which can make it a [difficult quantity to estimate.]}  \ben{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
 However, data informing models are often temporally anchored (e.g.
-weekly case counts).
+weekly case counts).}}\footnote{\ben{This was a bit unclear including defining a new term.
 As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.
-For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally do not have precise definitions.} \ben{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence?
-It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}\shannon{[Shannon says I added the few sentences to more smoothly introduce r0]} To demonstrate the difficulty of discerning $R_0$ in a time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model.
+For example,}}\shannon{ \cite{Gallagher2020} demonstrate\ben{s} how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally}\ben{hard to define precisely.\sout{do not have precise definitions}}.
+To demonstrate the difficulty of discerning $R_0$ in \ben{\sout{a}other}\footnote{\ben{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}} time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model.
 This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
@@
 
-From this model, $R_0 = \beta/\gamma$, aka the ratio of the infection rate compared to the recovery rate.
-Since $\beta$ and $\gamma$ are both rates, the ratio of the two, $R_0$, is a time-invariant quantity.
-Once $R_0$ is estimated, practioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
+From this model, $R_0 = \beta/\gamma$, aka the ratio of the infection rate compared to the recovery rate. \ben{From this definition, given 
+\sout{Since}} $\beta$ and $\gamma$ are both rates, \ben{it should be clear that} the ratio of the two, $R_0$, is a time-invariant quantity.\footnote{\ben{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}  Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
 Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
@@
 \shannon{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
-also highlighted tie to TIA and r0.
+also highlighted tie to time-invariant analysis and $R_0$.
 I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}
 
-\shannon{Time-invariant analysis helps practicioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
-Tne first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
+\shannon{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
+The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
 Both epidemics have populations of 1000 people with 10 individuals initially infected.
@@
 Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
-These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered."}
-ggplot(data = df %>%
+These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.", fig.show="hold", out.width="49%"}
+tern_vis <- ggplot(data = df %>%
          filter(t <= 15)) +
@@
   scale_color_manual(values = c("red", "blue"), name = "Epidemic")
+
+df_tri <- data.frame(t = 0:3,
+                     S = c(1,0,0,1),
+                     I = c(0,1,0,0),
+                     R = c(0,0,1,0),
+                     name = c("S", "I", "R", "S"),
+                     type = "border") %>% 
+  mutate(type = as.character(type))
+df2 <- df %>% filter(t <= 15) %>%
+  mutate(N = S+I+R,
+        S = S/N,
+        I = I/N,
+        R = R/N) %>%
+  select(-N)
+
+if (F){
+  plotly_vis <- df2 %>% 
+    plotly::plot_ly(x = ~S, y = ~R, z = ~I, color = ~type) %>%
+    plotly::add_trace(type = 'scatter3d', mode = "lines+markers",
+                      marker = list(size = 5),
+                      line = list(width = 4)) %>%
+    plotly::add_trace(data = df_tri, 
+                      x = ~S, 
+                      y = ~R,
+                      z = ~I,
+                      type = "mesh3d", color = I("blue"), opacity = .1) %>%
+    plotly::add_text(data = df_tri[1:3,], 
+                      x = ~S, 
+                      y = ~R,
+                      z = ~I, 
+                     text= ~name, color = I("black")) %>%
+    plotly::layout(scene = list(
+        camera = list(eye = list(x = 1.5, y = 1.5, z = 1.2)),
+        aspectratio = list(x = 1, y = 1, z = 1),
+        xaxis = list(title = ""),
+        yaxis = list(title = ""),
+        zaxis = list(title = "")),
+        showlegend = FALSE)
+    #plotly::add_surface(data = df_tri, x = ~S, y = ~I, z = ~R, color = ~type)
+
+  Sys.setenv("PATH" = paste(Sys.getenv("PATH"), 
+                            "/Users/benjaminleroy/miniconda3/bin", 
+                            sep = .Platform$path.sep))
+  plotly::orca(plotly_vis, file = "images/vis3d.png",width = 8 * 120, height = 8 * 120)
+}
+
+knitr::include_graphics("images/vis3d.png")
+plot(tern_vis)
 ```
@@
 $$
-where a mapping $\xi : s \to \mathbb{R}$ that is monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi = \psi$ where :
+where a mapping $\xi : s \to \mathbb{R}$ that is strictly monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi \equiv \psi$ where :
 
@@
 \shannon{In other words, the two epidemics form the same filamental trajectory.
-For SIR models with similar initial state percentages, time-invariant analysis, time-invariant analysis allows practicioners to compare values of $R_0$ at a glance.}  \ben{[We've changed how we describe these two objects when they are related to the theorem - is that desirable?
-]}\shannon{added that sentence to for that}
+For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}
 

### 2021-03-31 jss/EpiCompare-paper/EpiCompare-paper.tex
some section 2 edits plus add of plotly vis

@@
 \subsection[motivating through R0]{Motivating time-invariant analysis
-through \textcolor{orange}{the reproduction number}
-\(R_0\)\footnote{\textcolor{violet}{Is this still a relevant title for this section? Think about this header after the rewrite.}}}\label{sec:r0}
+through the reproduction number \(R_0\)}\label{sec:r0}
 
@@
 the reproduction number \(R_0\).
-\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence}}
+\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence.}
+  \textcolor{violet}{I still think it adds value to the story and I'm not sure people really read section titles that are long.}}
 
@@
 to a completely susceptible population \citep{anderson1992}.
-\textcolor{orange}{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
+\textcolor{violet}{This means that $R_0$ is a time-invariant parameter yet is estimated with time-based data, which can make it a [difficult quantity to estimate.]}
+\textcolor{violet}{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
 However, data informing models are often temporally anchored (e.g.
-weekly case counts).
+weekly case counts).}}\footnote{\textcolor{violet}{This was a bit unclear including defining a new term.
 As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.
-For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally do not have precise definitions.}
-\textcolor{violet}{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence?
-It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}\textcolor{orange}{[Shannon says I added the few sentences to more smoothly introduce r0]}
-To demonstrate the difficulty of discerning \(R_0\) in a time-dependent
-analysis, we first introduce \citet{Kermack1927}'s SIR model. This model
-captures the transitions from one state to the next as a system of
-ordinary differential equations, where \(N\) is the total number of
-individuals, \(\beta\) is the rate of infection, and \(\gamma\) is the
-rate of recovery,
+For example,}}\textcolor{orange}{ \cite{Gallagher2020} demonstrate\textcolor{violet}{s} how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
+To demonstrate the difficulty of discerning \(R_0\) in
+\textcolor{violet}{\sout{a}other}\footnote{\textcolor{violet}{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}}
+time-dependent analysis, we first introduce \citet{Kermack1927}'s SIR
+model. This model captures the transitions from one state to the next as
+a system of ordinary differential equations, where \(N\) is the total
+number of individuals, \(\beta\) is the rate of infection, and
+\(\gamma\) is the rate of recovery,
 
@@
 From this model, \(R_0 = \beta/\gamma\), aka the ratio of the infection
-rate compared to the recovery rate. Since \(\beta\) and \(\gamma\) are
-both rates, the ratio of the two, \(R_0\), is a time-invariant quantity.
-Once \(R_0\) is estimated, practioners can infer important epidemic
+rate compared to the recovery rate.
+\textcolor{violet}{From this definition, given 
+\sout{Since}} \(\beta\) and \(\gamma\) are both rates,
+\textcolor{violet}{it should be clear that} the ratio of the two,
+\(R_0\), is a time-invariant
+quantity.\footnote{\textcolor{violet}{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}
+Once \(R_0\) is estimated, practitioners can infer important epidemic
 quantities such as the total number of infections or the percent of a
@@
 \textcolor{orange}{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
-also highlighted tie to TIA and r0.
+also highlighted tie to time-invariant analysis and $R_0$.
 I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}
 
-\textcolor{orange}{Time-invariant analysis helps practicioners to more easily compare $R_0$ from different outbreaks.}
+\textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.}
 For example, consider two epidemics generated from the Kermack and
-McKendrick SIR equations. Tne first epidemic has parameters
+McKendrick SIR equations. The first epidemic has parameters
 \(\beta_1, \gamma_1 = (0.8,0.4)\) and the second has
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-3-1}
+{\centering \includegraphics[width=0.49\linewidth]{images/vis3d} \includegraphics[width=0.49\linewidth]{Figs/unnamed-chunk-3-2}
 
@@
 \psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
-\] where a mapping \(\xi : s \to \mathbb{R}\) that is monotonically
-increasing would not change the definition of \(\psi\),
-i.e.~\(\psi_\xi = \psi\) where :
+\] where a mapping \(\xi : s \to \mathbb{R}\) that is strictly
+monotonically increasing would not change the definition of \(\psi\),
+i.e.~\(\psi_\xi \equiv \psi\) where :
 
@@
 \textcolor{orange}{In other words, the two epidemics form the same filamental trajectory.
-For SIR models with similar initial state percentages, time-invariant analysis, time-invariant analysis allows practicioners to compare values of $R_0$ at a glance.}
-\textcolor{violet}{[We've changed how we describe these two objects when they are related to the theorem - is that desirable? ]}\textcolor{orange}{added that sentence to for that}
+For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}
 
@@
 
+As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of
+potential for time-invariant tools to help compare and assess epidemics
+and models/simulations. In \pkg{EpiCompare} we provide a set of
+comparison and assessment tools for models that extend beyond the
+standard performance metrics (e.g.~means squared error, AIC) and focus
+on assessing the structural information the models capture. This
+approaches work well on models where online one ``cycle'' of the
+epidemic has occurred (no recovered individuals have been susceptible
+again). Epidemics are complex objects, and we provide tools to create
+prediction regions out of simulated epidemics. For 3-state epidemic
+models, we provide the \texttt{ggplot}/\texttt{ggtern} extension
+\texttt{geom\_prediction\_band()} which creates a prediction region
+around the top \(1-\alpha\) proportion of the simulations (where the
+simulations treated as filaments). In \pkg{Epicompare} we also provide
+these prediction regions for epidemic models with with more than 3
+states. With \texttt{create\_convex\_hull\_structure} and
+`\texttt{create\_delta\_ball\_structure} that will create different
+geometric representations of prediction regions and an assoicated
+\texttt{contained} method that assesses if a new epidemic (seen as a
+filament) is contained in the prediction region.
+
+We also provide functions to calculate the extremeness of a true
+epidemic compared to simulated epidemics through the equi-distance
+filamental trajectory representation as mentioned in Section
+\ref{sec:beyond-r0-sir}. Specifically, functions like
+\texttt{distance\_psuedo\_density\_function} can calculate a
+psuedo-density estimate of the true epidemic relative to simulated ones.
+Functions \texttt{distance\_depth\_function} and
+\texttt{local\_distance\_depth\_function} provide depth scores that
+suggest how geometrically central an epidemic is to simulations.
+
+\textcolor{violet}{[Ben says: Replaced with paragraphs above.]}
 Comparison and assessment of model fit or comparisons of one model to
@@
 
-However, for those who like more concrete tools, \pkg{EpiCompare} offers
+\textcolor{violet}{[Ben says: Replaced with paragraphs above.]} However,
+for those who like more concrete tools, \pkg{EpiCompare} offers
 functions to compare prediction regions to one another including

### 2021-04-01 jss/EpiCompare-paper/EpiCompare-paper.Rmd
correction of new order of ternary plot in section 2.

@@
 
-\shannon{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
+\ben{[Ben says: this paragraph needs to still be looked at.
+Also I'm not sure why this particular paragraph was changed so much at all.
+Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \shannon{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
 The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
@@
 We plot the epidemics with traditional $state$ vs.
-$time$ plots.
+$time$ plots\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
 In Fig.
@@
 In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.
-On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15.
-In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$.
+On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \ben{epidemic} has not yet peaked at time 15.
+In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$\footnote{\ben{This sentnce doesn't connect wit previous examples.}}.
 
@@
 
-\shannon{NEW} A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.  In Fig. \ref{fig:different-scales-tern}
+\shannon{NEW} A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.
+\ben{For every time point $t$ we have $S(t),I(t), R(t)$, so we can treat epdiemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.
+For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point ina a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}'s right subplot.
+\sout{In Fig.
+\ref{fig:different-scales-tern}
 we plot the filamental trajectories of the two epidemics in a time-invariant view.
 We will explain how and why this works shortly.
-The important takeaway is that in this time-invariant view, it is apparent that these epidemics are on "the same path." In this case, this indicates that two epidemics have the same value of $R_0$.
+The important takeaway is that in this time-invariant view, i}I}t is apparent that these epidemics are on "the same path." In this case, this indicates that two epidemics have the same value of $R_0$.
 
@@
 
-\shannon{The reason why we can visually compare $R_0$ in Fig.
-\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.} A filamental trajectory can be mathematically viewed as a set of points in space that have an ordering, and that all points on the line between these ordered points are also contained in the geometric object.
+\ben{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig.
+\ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space.
+\sout{The reason why we can visually compare $R_0$ in Fig.
+\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}} A filamental trajectory can be mathematically viewed as a set of points in space that have an ordering, and that all points on the line between these ordered points are also contained in the geometric object.
 For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as
@@
 $$
-Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.
+\ben{[Ben says: removed this paragraph now]} Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.
 \sout{We can then which can be visualize the \shannon{full filamental trajectory} in a two-dimensional ternary plot}  As a result, we can visualize \shannon{the full filamental trajectory} of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.
@@
 
-The filamental trajectories in Fig.
+\ben{As suggested a few paragraphs back, t}he filamental trajectories in Fig.
 \ref{fig:different-scales-tern} seem to overlap, and we may suspect that something is fundamentally linking these two different epidemics together.

### 2021-04-01 jss/EpiCompare-paper/EpiCompare-paper.tex
correction of new order of ternary plot in section 2.

@@
 
+\textcolor{violet}{[Ben says: this paragraph needs to still be looked at.
+Also I'm not sure why this particular paragraph was changed so much at all.
+Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]}
 \textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.}
@@
 \(R_0 = 2=0.8/0.4 = 0.64/0.32\). We plot the epidemics with traditional
-\(state\) vs.~\(time\) plots. In Fig.
-\ref{fig:different-scales-standard} we show the time-based paths for the
-\(S\), \(I\), and \(R\) states for the first 15 days of observed data.
-In this time-variant view, we may believe that
+\(state\) vs.~\(time\)
+plots\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
+In Fig. \ref{fig:different-scales-standard} we show the time-based paths
+for the \(S\), \(I\), and \(R\) states for the first 15 days of observed
+data. In this time-variant view, we may believe that
 \textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
@@
 \textcolor{violet}{epidemic} 2 has a larger \(R_0\) because its unclear
-if the number of infections in that
-\sout{model}\textcolor{violet}{epidemic} has not yet peaked at time 15.
-In this time-variant view, we cannot determine if one epidemic has
-larger value of \(R_0\).
+if the number of infections in that \textcolor{violet}{epidemic} has not
+yet peaked at time 15. In this time-variant view, we cannot determine if
+one epidemic has larger value of
+\(R_0\)\footnote{\textcolor{violet}{This sentnce doesn't connect wit previous examples.}}.
 
@@
 epidemics, in comparison, allows us to directly compare \(R_0\) from a
-single plot. In Fig. \ref{fig:different-scales-tern} we plot the
-filamental trajectories of the two epidemics in a time-invariant view.
-We will explain how and why this works shortly. The important takeaway
-is that in this time-invariant view, it is apparent that these epidemics
-are on ``the same path.'' In this case, this indicates that two
-epidemics have the same value of \(R_0\).
+single plot.
+\textcolor{violet}{For every time point $t$ we have $S(t),I(t), R(t)$, so we can treat epdiemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.
+For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point ina a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}'s right subplot.
+\sout{In Fig.
+\ref{fig:different-scales-tern}
+we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, i}I}t
+is apparent that these epidemics are on ``the same path.'' In this case,
+this indicates that two epidemics have the same value of \(R_0\).
 
@@
 
-\textcolor{orange}{The reason why we can visually compare $R_0$ in Fig.
-\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}
+\textcolor{violet}{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig.
+\ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space.
+\sout{The reason why we can visually compare $R_0$ in Fig.
+\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}}
 A filamental trajectory can be mathematically viewed as a set of points
@@
 \psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
-\] Since the number in each state is non-negative and the sum over the
-three states for a given time point sums to \(N\), then
+\] \textcolor{violet}{[Ben says: removed this paragraph now]} Since the
+number in each state is non-negative and the sum over the three states
+for a given time point sums to \(N\), then
 \textcolor{violet}{all points in} \(\psi\) will lay in a
@@
 
-The filamental trajectories in Fig. \ref{fig:different-scales-tern} seem
-to overlap, and we may suspect that something is fundamentally linking
-these two different epidemics together. Mathematically, we can show this
+\textcolor{violet}{As suggested a few paragraphs back, t}he filamental
+trajectories in Fig. \ref{fig:different-scales-tern} seem to overlap,
+and we may suspect that something is fundamentally linking these two
+different epidemics together. Mathematically, we can show this
 fundamental link turns is \(R_0\). Let our two epidemics be presented as

### 2021-04-01 jss/EpiCompare-paper/EpiCompare-paper.Rmd
ternary plots 3d update.

@@
 df_tri <- data.frame(t = 0:3,
-                     S = c(1,0,0,1),
-                     I = c(0,1,0,0),
-                     R = c(0,0,1,0),
+                     S = c(1,0,0,1)*1000,
+                     I = c(0,1,0,0)*1000,
+                     R = c(0,0,1,0)*1000,
                      name = c("S", "I", "R", "S"),
@@
   mutate(type = as.character(type))
-df2 <- df %>% filter(t <= 15) %>%
-  mutate(N = S+I+R,
-        S = S/N,
-        I = I/N,
-        R = R/N) %>%
-  select(-N)
+df2 <- df %>% filter(t <= 15) #%>%
+  #mutate(N = S+I+R,
+  #      S = S/N,
+  #      I = I/N,
+  #      R = R/N) %>%
+  #select(-N)
 
@@
                       z = ~I, 
-                     text= ~name, color = I("black")) %>%
+                     text= ~name, color = I("black"), textfont = list(size = 20)) %>%
     plotly::layout(scene = list(

### 2021-04-02 jss/EpiCompare-paper/EpiCompare-paper.Rmd
update to 2.2 and comparisons and assessment in section 3.

@@
 
-##  \ben{Time-invariant analysis} Beyond $R_0$ and Kermack’s and McKendrick SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #beyond-r0-sir}
+##  Time-invariant analysis beyond $R_0$ and Kermack’s and McKendrick SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #sec:beyond-r0-sir}
 
 Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted.
-\ben{\sout{In this section we will present} time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown \shannon{or have more than three epidemic states}.
-\sout{We will also present how this tools can be used when the epidemics have more than 3 states.} \sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states.
-}
+Time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown or have more than three epidemic states.
 
 \ben{New paragraph} Viewing epidemics as filamental trajectories provide \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner.
-For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{are we missing a few words in this definition?}.
-This \shannon{representation induces}\sout{ also creates} a natural distance between this type of representation between epidemics, specifically:
+\ben{For epidemics that have ended, one way to examine their filamental trajectories is to redefine the trajectory as a finite sequence of points on the filamental trajectory that are equally spaced (i.e.
+equa-distance between pairs of ordered points).} \sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{[Shannon says: are we missing a few words in this definition?]}}.
+This representation induces a natural distance between this type of representation between epidemics, specifically:
 \[
@@
 This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall extremeness of points, including psuedo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
-This \shannon{what is this?} can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model, though we recommend projecting the points into the unit simplex\footnote{\ben{I'm not sure we've talked about this before...
-\shannon{I don't think we have but am wondering if we're getting in the weeds}}} (by making all values the proportion of the population in the given state).
-
-\ben{New paragraph:} If the set of epidemics that one is examining have only gone through a single cycle of the outbreak (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?}}].
+These extremeness estiamtors can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\ben{I'm not sure we've talked about this before...
+\shannon{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
+
+\ben{New paragraph:} \ben{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If one is interested in understanding epeidmeics through a single cycle of their outbreak} (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?}}].
 In these settings we go a step farther and treat epidemics more like geometric filaments \shannon{(i.e.
@@
 In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above pseudo-density estimates or depth estimates).
-These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using the Haussdorf \shannon{why Haussdorf specifically?} distance as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
+These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using \ben{many set difference distances \sout{the Haussdorf \shannon{why Haussdorf specifically?} distance}} as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
 All these geometric structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).
-
-\ben{[Ben says: Old recommendations:
-to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset.
-Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.]}
-
-\shannon{I like your completion.  I think we can make it flow a bit more smoothly but the content we want to talk about is all there.}
 
@@
 
-\ben{[This still needs to be worked on - just a reminder.]}
-
 <!-- \ben{[Brainstorming: (intro) discuss model vs model comparison tools, (a) highlight that standard comparisons can be done if models share metrics or parameters (and similarly some comparisons can be done to the real epidemics).
@@
 means squared error, AIC) and focus on assessing the structural information the models capture.
-This approaches work well on models where online one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again).
+This approaches work well on models where online one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again)\footnote{\ben{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}}.
 Epidemics are complex objects, and we provide tools to create prediction regions out of simulated epidemics.
-For 3-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations (where the simulations treated as filaments).
-In \pkg{Epicompare} we also provide these prediction regions for epidemic models with with more than 3 states.
-With `create_convex_hull_structure` and ``create_delta_ball_structure` that will create different geometric representations of prediction regions and an assoicated `contained` method that assesses if a new epidemic (seen as a filament) is contained in the prediction region.
+For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations (where the simulations treated as filaments).
+In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
+The functions `create_convex_hull_structure` and `create_delta_ball_structure` create different geometric representations of prediction regions for any state-based model.
+For both of these geometric objects, we provide functions to check if a path is contained (`contained`) and the ability to assess the Hassdorf distance between prediction regions based on simulations from different model (`hassdorf_dist`).
+\ben{[Ben says: should we discuss the details of `geom_prediction_bands` more?]}
 
@@
 Functions `distance_depth_function` and `local_distance_depth_function` provide depth scores that suggest how geometrically central an epidemic is to simulations.
+\ben{[Ben says: should we define these more in this paragraph?
+Introducted in 2.2.
+as well and not defined.]}
 

### 2021-04-02 jss/EpiCompare-paper/EpiCompare-paper.tex
Auto stash before merge of "master" and "origin/master"
update to 2.2 and comparisons and assessment in section 3.

@@
 
-\subsection[Beyond R0 and SIR]{\textcolor{violet}{Time-invariant analysis}
-Beyond \(R_0\) and Kermack's and McKendrick SIR
-Models\footnote{Probably will need to change this title...}}\label{beyond-r0-sir}
+\subsection[Beyond R0 and SIR]{Time-invariant analysis beyond \(R_0\)
+and Kermack's and McKendrick SIR
+Models\footnote{Probably will need to change this title...}}\label{sec:beyond-r0-sir}
 
@@
 to better compare the overall structure of the epidemic and see how the
-population was directly impacted.
-\textbackslash ben\{\sout{In this section we will present}
-time-invariant tools \sout{that} can be useful even when the underlying
-generative model for the epidemic is unknown
-\textcolor{orange}{or have more than three epidemic states}.
-\sout{We will also present how this tools can be used when the epidemics have more than 3 states.}
-\sout{trajectories, we propose that time-invariant tools can help compare them in smart ways, and that these ideas can extend past the CITE kermack1927 model and to epidemics with more states. }
+population was directly impacted. Time-invariant tools \sout{that} can
+be useful even when the underlying generative model for the epidemic is
+unknown or have more than three epidemic states.
 
@@
 trajectories provide \sout{a lot }new ways to compare and examine
-epidemics in a time-invariant manner. For epidemics that have ``played''
-themselves out, one way to represent their filamental trajectories to
-avoid \textcolor{orange}{confusion stemming from}\sout{impacts of}
-temporal structure is to define them as a sequence of points their
-trajectory with equi-distance between each point
-\textcolor{orange}{are we missing a few words in this definition?}. This
-\textcolor{orange}{representation induces}\sout{ also creates} a natural
-distance between this type of representation between epidemics,
-specifically: \[
+epidemics in a time-invariant manner.
+\textcolor{violet}{For epidemics that have ended, one way to examine their filamental trajectories is to redefine the trajectory as a finite sequence of points on the filamental trajectory that are equally spaced (i.e.
+equa-distance between pairs of ordered points).}
+\sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \textcolor{orange}{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \textcolor{orange}{[Shannon says: are we missing a few words in this definition?]}}.
+This representation induces a natural distance between this type of
+representation between epidemics, specifically: \[
 d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
@@
 depth/local depth functions
-\citep[for examples see][]{Ciollaro2016, Geenens2017}. This
-\textcolor{orange}{what is this?} can be very useful when comparing
-between a set of simulation epidemics and the true epidemic, and does
-not constrain the number of states of the model, though we recommend
-projecting the points into the unit
-simplex\footnote{\textcolor{violet}{I'm not sure we've talked about this before... \textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}
-(by making all values the proportion of the population in the given
-state).
-
-\textcolor{violet}{New paragraph:} If the set of epidemics that one is
-examining have only gone through a single cycle of the outbreak (before
-the population of individuals have become susceptible again), then
-additional time-invariant tools, including prediction regions can we
-leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done?}\}}.
+\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
+estiamtors can be very useful when comparing between a set of simulation
+epidemics and the true epidemic, and does not constrain the number of
+states of the
+model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\textcolor{violet}{I'm not sure we've talked about this before...
+\textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
+
+\textcolor{violet}{New paragraph:}
+\textcolor{violet}{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If one is interested in understanding epeidmeics through a single cycle of their outbreak}
+(before the population of individuals have become susceptible again),
+then additional time-invariant tools, including prediction regions can
+we leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done?}\}}.
 In these settings we go a step farther and treat epidemics more like
@@
 at where in the state-space we expect the epidemic to traverse, and we
-can compare prediction regions defined by different models using the
-Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance as
-well as examining how well the truth epidemic matches the simulations by
-examining if the epidemic's trajectory lies within the prediction
+can compare prediction regions defined by different models using
+\textcolor{violet}{many set difference distances \sout{the Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance}}
+as well as examining how well the truth epidemic matches the simulations
+by examining if the epidemic's trajectory lies within the prediction
 region. All these geometric structures and distances notations apply to
@@
 models).
-
-\textcolor{violet}{[Ben says: Old recommendations:
-to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset.
-Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.]}
-
-\textcolor{orange}{I like your completion.  I think we can make it flow a bit more smoothly but the content we want to talk about is all there.}
 
@@
 
-\textcolor{violet}{[This still needs to be worked on - just a reminder.]}
-
 As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of
@@
 epidemic has occurred (no recovered individuals have been susceptible
-again). Epidemics are complex objects, and we provide tools to create
-prediction regions out of simulated epidemics. For 3-state epidemic
-models, we provide the \texttt{ggplot}/\texttt{ggtern} extension
+again)\footnote{\textcolor{violet}{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}}.
+Epidemics are complex objects, and we provide tools to create prediction
+regions out of simulated epidemics. For three-state epidemic models, we
+provide the \texttt{ggplot}/\texttt{ggtern} extension
 \texttt{geom\_prediction\_band()} which creates a prediction region
 around the top \(1-\alpha\) proportion of the simulations (where the
-simulations treated as filaments). In \pkg{Epicompare} we also provide
-these prediction regions for epidemic models with with more than 3
-states. With \texttt{create\_convex\_hull\_structure} and
-`\texttt{create\_delta\_ball\_structure} that will create different
-geometric representations of prediction regions and an assoicated
-\texttt{contained} method that assesses if a new epidemic (seen as a
-filament) is contained in the prediction region.
+simulations treated as filaments). In \pkg{EpiCompare} we also provide
+these prediction regions for epidemic models with with more than three
+states. The functions \texttt{create\_convex\_hull\_structure} and
+\texttt{create\_delta\_ball\_structure} create different geometric
+representations of prediction regions for any state-based model. For
+both of these geometric objects, we provide functions to check if a path
+is contained (\texttt{contained}) and the ability to assess the Hassdorf
+distance between prediction regions based on simulations from different
+model (\texttt{hassdorf\_dist}).
 
@@
 suggest how geometrically central an epidemic is to simulations.
+\textcolor{violet}{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}
 

### 2021-04-06 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Edits to section 2.

@@
 
-\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories and not points produced by functions of time \shannon{(these concepts will be explained in  explained in Section \ref{sec:r0})}.
+\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories\sout{ and not points produced by functions of time} \shannon{(this concept will be explained in  explained in Section \ref{sec:r0})}.
 The following section aims to highlight the strengths of \textit{time-invariant} \shannon{analysis} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
@@
 Mathematically, epidemics are complex objects.
-They can be hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time related features (including start of the epidemic, speed of infection and more).
+They can be hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time \sout{related}\shannon{related} features (including start of the epidemic, speed of infection and more).
 Time-invariant analysis makes different epidemics easier to compare by removing many time dependent aspects of an epidemic.
-Instead, \shannon{time-invariant analysis} focuses on the overall \shannon{shape and direction, a filamental trajctory,} \sout{trajectory of }an epidemic.
-\shannon{These filamental trajectories,  in turn,} emphasize the number of lives affected.
+Instead, \shannon{time-invariant analysis} focuses on \sout{the overall \shannon{shape and direction, a filamental trajctory,} \sout{trajectory of }an epidemic.
+\shannon{These filamental trajectories,  in turn,} } \shannon{filamental trajectories, which emphasize the number of lives affected.}
 
@@
 
-$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
-$R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.
-\ben{This means that $R_0$ is a time-invariant parameter yet is estimated with time-based data, which can make it a [difficult quantity to estimate.]}  \ben{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
+$R_0$ is probably the most famous \sout{time-invariant} numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
+$R_0$ is \sout{a one-number summary of a disease and is }defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.
+\ben{This \shannon{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\shannon{.
+Yet $R_0$ is} estimated with time-\sout{based}\shannon{dependent} data, which can make it a [difficult quantity to estimate.]} \ben{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
 However, data informing models are often temporally anchored (e.g.
@@
 As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.
-For example,}}\shannon{ \cite{Gallagher2020} demonstrate\ben{s} how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally}\ben{hard to define precisely.\sout{do not have precise definitions}}.
+For example,}}\shannon{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\shannon{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally}\ben{hard to define precisely.\sout{do not have precise definitions}}.
 To demonstrate the difficulty of discerning $R_0$ in \ben{\sout{a}other}\footnote{\ben{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}} time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model.
@@
 
-From this model, $R_0 = \beta/\gamma$, aka the ratio of the infection rate compared to the recovery rate. \ben{From this definition, given 
-\sout{Since}} $\beta$ and $\gamma$ are both rates, \ben{it should be clear that} the ratio of the two, $R_0$, is a time-invariant quantity.\footnote{\ben{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}  Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
+From this model,\shannon{the reproduction number is the ratio of the infection rate to the recovery rate,} $R_0 = \beta/\gamma$\sout{, aka the ratio of the infection rate compared to the recovery rate.} \ben{From this definition, given
+\sout{Since}} $\beta$ and $\gamma$ are both rates, \ben{it should be clear that} the ratio of the two, $R_0$, is a time-invariant quantity.\footnote{\ben{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}} \sout{ Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
 Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
+}^[cool facts about r0, but not the central point]
 
@@
 also highlighted tie to time-invariant analysis and $R_0$.
-I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}
-
-\ben{[Ben says: this paragraph needs to still be looked at.
+I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}\shannon{ Shannon tries again in \textcolor{blue}{blue}}
+
+\sout{\ben{[Ben says: this paragraph needs to still be looked at.
 Also I'm not sure why this particular paragraph was changed so much at all.
@@
 We plot the epidemics with traditional $state$ vs.
-$time$ plots\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
-In Fig.
+$time$ plots}\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
+\sout{In Fig.
 \ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.
@@
 On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \ben{epidemic} has not yet peaked at time 15.
-In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$\footnote{\ben{This sentnce doesn't connect wit previous examples.}}.
+In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$}\footnote{\ben{This sentnce doesn't connect wit previous examples.}}.
+
+\textcolor{blue}{Since $R_0$ is an important value, it would be helpful to have more intuitive ways of comparing one $R_0$ to another.
+Usually numerical summaries of $R_0$ are presented, which while overall very helpful, may be confusing when presented along side epidemic data that are visualized in a traditional, time-dependent manner.}
+
+\textcolor{blue}{For example, consider two epidemics generated from the Kermack and McKendrick SIR equations where both have the same value of $R_0$.
+The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
+Both epidemics have populations of 1000 people with 10 individuals initially infected.
+An analysis may present an estimate of $\hat{R}_0 = 2$ alongside state vs.
+time plots like those shown in Figure \ref{fig:different-scales-standard}.
+The paths of the epidemics in the state vs.
+time view seem to differ from one another including having different infection peaks.
+From these traditional time-based plots, there is no intuitive way to conclude that these two epidemics have the same value of $R_0$.}
 
@@
 
-\shannon{NEW} A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.
-\ben{For every time point $t$ we have $S(t),I(t), R(t)$, so we can treat epdiemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.
-For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point ina a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}'s right subplot.
+\textcolor{blue}{\pkg{EpiCompare} provides a time-invariant tool to visualize these epidemics in a more intuitive manner, at least in regards to comparing values of $R_0$.} \sout{A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.}  \ben{For every time \sout{point} $t$ we have a \shannon{point} $(S(t),I(t), R(t))$\sout{, so we can treat epidemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.} \shannon{so we can visualize the trajectory of the epidemic in three-dimensional space (see Fig.
+\ref{fig:different-scales-tern} (left)).} For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern} (right).
 \sout{In Fig.
@@
 
-```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
-Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
-These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.", fig.show="hold", out.width="49%"}
+```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Left: trajectory of epidemic in three-dimensional space, plotting $(S(t), I(t), R(t))$.
+Right: the gray-shaded region and epidemic trajectory shown from (left) now shown in two-dimensional space.
+This is more commonly known as a ternary plot.", fig.show="hold", out.width="49%"}
 tern_vis <- ggplot(data = df %>%
@@
 $$
-\ben{[Ben says: removed this paragraph now]} Since the number in each state is non-negative and the sum over the three states for a given time point sums to $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.
+\ben{[Ben says: removed this paragraph now]} Since the number \shannon{of individuals} in each state is non-negative and the sum over the three states for a given time point \sout{sums to}\shannon{is} $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.
 \sout{We can then which can be visualize the \shannon{full filamental trajectory} in a two-dimensional ternary plot}  As a result, we can visualize \shannon{the full filamental trajectory} of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.
@@
 
-\ben{As suggested a few paragraphs back, t}he filamental trajectories in Fig.
-\ref{fig:different-scales-tern} seem to overlap, and we may suspect that something is fundamentally linking these two different epidemics together.
-Mathematically, we can show this fundamental link turns is $R_0$.
+\sout{\ben{As suggested a few paragraphs back, t}}The filamental trajectories in Fig.
+\ref{fig:different-scales-tern} \sout{seem to overlap}, and we may suspect that something is fundamentally linking these two different epidemics together.
+Mathematically, we can show this fundamental link \sout{turns} is $R_0$.
 Let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.
@@
 
-##  Time-invariant analysis beyond $R_0$ and Kermack’s and McKendrick SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #sec:beyond-r0-sir}
-
-Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to better compare the overall structure of the epidemic and see how the population was directly impacted.
+##  Time-invariant analysis beyond $R_0$ and \sout{Kermack’s and McKendrick} SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #sec:beyond-r0-sir}
+
+Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to \sout{better}\shannon{more fully} compare the overall structure of the epidemic and see how the population is directly impacted.
 Time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown or have more than three epidemic states.
 
-\ben{New paragraph} Viewing epidemics as filamental trajectories provide \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner.
-\ben{For epidemics that have ended, one way to examine their filamental trajectories is to redefine the trajectory as a finite sequence of points on the filamental trajectory that are equally spaced (i.e.
+\ben{New paragraph} Viewing epidemics as filamental trajectories provides \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner.
+\ben{For \shannon{completed?}epidemics that have ended, one way to examine their filamental trajectories is to \sout{redefine}\shannon{represent} the \shannon{filamental} trajectory as a \shannon{finite sequence of equally spaced points.} \sout{finite sequence of points on the filamental trajectory that are equally spaced (i.e.
 equa-distance between pairs of ordered points).} \sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{[Shannon says: are we missing a few words in this definition?]}}.
@@
 \]
-where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$.
-This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall extremeness of points, including psuedo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
-These extremeness estiamtors can be very useful when comparing between a set of simulation epidemics and the true epidemic, and does not constrain the number of states of the model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\ben{I'm not sure we've talked about this before...
+where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$.\footnote{\shannon{I think you're trying to say something about a distance based on the equally space points.
+Some clarifying questions:  is $\psi^\prime$ the derivative?
+Does proportion make more sense than fraction?
+or simply $\frac{|\psi|}{s}$?
+It's only naturally time-invariant if we have a well defined ending point, right?}} This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
+These extremeness estimators can be \sout{very} useful when comparing \sout{between a setof simulation} \shannon{a set of simulated} epidemics and the true epidemic\shannon{.
+Moreover these extremness estimators,} \sout{and does} \shannon{} not constrain the number of states of the model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\ben{I'm not sure we've talked about this before...
 \shannon{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
 
-\ben{New paragraph:} \ben{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If one is interested in understanding epeidmeics through a single cycle of their outbreak} (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can we leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?}}].
-In these settings we go a step farther and treat epidemics more like geometric filaments \shannon{(i.e.
+\ben{New paragraph:} \ben{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If \sout{one} \shannon{a practicioner} is interested in understanding \shannon{an} epidemic\sout{s} through a single \sout{cycle}\shannon{realization} of \sout{their}\shannon{its} outbreak} (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can be leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?
+Update 4/6: I'm satisfied.}}].
+In these settings, \shannon{\pkg{EpiCompare} goes} \sout{we go }a step further and treats epidemics more like geometric filaments \shannon{(i.e.
 filamental trajectories without an ordering of points)} than filamental trajectories.
-In \pkg{EpiCompare} we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small filaments around the subset of simulations (subset by measures like the above pseudo-density estimates or depth estimates).
-These regions look at where in the state-space we expect the epidemic to traverse, and we can compare prediction regions defined by different models using \ben{many set difference distances \sout{the Haussdorf \shannon{why Haussdorf specifically?} distance}} as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
-All these geometric structures and distances notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).
+In \pkg{EpiCompare}, we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small \shannon{geometric?} filaments around the subset of simulations (\sout{subset} \shannon{grouped} by measures like the above pseudo-density estimates or depth estimates).
+These regions \sout{look at}\shannon{show} where in the state-space we expect the epidemic to traverse\sout{, and }.
+\shannon{Additionally, }we can compare prediction regions defined by different models using \ben{many set difference distances \sout{the Haussdorf \shannon{why Haussdorf specifically?} distance}} as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
+All these \shannon{mentioned?} geometric structures and distance notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).
 
@@
 
+\shannon{READ}
 **Post-processing**

### 2021-04-06 jss/EpiCompare-paper/EpiCompare-paper.tex
Edits to section 2.

@@
 a global, not marginal view of how epidemics move through populations
-and (2) by treating full epidemics as filamental trajectories and not
-points produced by functions of time
-\textcolor{orange}{(these concepts will be explained in  explained in Section \ref{sec:r0})}.
+and (2) by treating full epidemics as filamental
+trajectories\sout{ and not points produced by functions of time}
+\textcolor{orange}{(this concept will be explained in  explained in Section \ref{sec:r0})}.
 The following section aims to highlight the strengths of
@@
 diseases, the location where the outbreak occurs, how the affected
-population reacts, and the time related features (including start of the
-epidemic, speed of infection and more). Time-invariant analysis makes
-different epidemics easier to compare by removing many time dependent
-aspects of an epidemic. Instead,
-\textcolor{orange}{time-invariant analysis} focuses on the overall
-\textcolor{orange}{shape and direction, a filamental trajctory,}
-\sout{trajectory of }an epidemic.
-\textcolor{orange}{These filamental trajectories,  in turn,} emphasize
-the number of lives affected.
+population reacts, and the time
+\sout{related}\textcolor{orange}{related} features (including start of
+the epidemic, speed of infection and more). Time-invariant analysis
+makes different epidemics easier to compare by removing many time
+dependent aspects of an epidemic. Instead,
+\textcolor{orange}{time-invariant analysis} focuses on
+\sout{the overall \textcolor{orange}{shape and direction, a filamental trajctory,} \sout{trajectory of }an epidemic.  \textcolor{orange}{These filamental trajectories,  in turn,} }
+\textcolor{orange}{filamental trajectories, which emphasize the number of lives affected.}
 
@@
 
-\(R_0\) is probably the most famous time-invariant numerical summary of
-an epidemic and is often associated with the
+\(R_0\) is probably the most famous \sout{time-invariant} numerical
+summary of an epidemic and is often associated with the
 Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
-\(R_0\) is a one-number summary of a disease and is defined as the
-expected number of infections caused by a single infector who is added
-to a completely susceptible population \citep{anderson1992}.
-\textcolor{violet}{This means that $R_0$ is a time-invariant parameter yet is estimated with time-based data, which can make it a [difficult quantity to estimate.]}
+\(R_0\) is \sout{a one-number summary of a disease and is }defined as
+the expected number of infections caused by a single infector who is
+added to a completely susceptible population \citep{anderson1992}.
+\textcolor{violet}{This \textcolor{orange}{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\textcolor{orange}{.
+Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a [difficult quantity to estimate.]}
 \textcolor{violet}{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
@@
 As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.
-For example,}}\textcolor{orange}{ \cite{Gallagher2020} demonstrate\textcolor{violet}{s} how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
+For example,}}\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
 To demonstrate the difficulty of discerning \(R_0\) in
@@
 
-From this model, \(R_0 = \beta/\gamma\), aka the ratio of the infection
-rate compared to the recovery rate.
+From this
+model,\textcolor{orange}{the reproduction number is the ratio of the infection rate to the recovery rate,}
+\(R_0 = \beta/\gamma\)\sout{, aka the ratio of the infection rate compared to the recovery rate.}
 \textcolor{violet}{From this definition, given 
@@
 quantity.\footnote{\textcolor{violet}{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}
-Once \(R_0\) is estimated, practitioners can infer important epidemic
-quantities such as the total number of infections or the percent of a
-population needed to be vacccinated to stop the sustained spread of an
-epidemic. Moreover, \(R_0\) allows us to compare different diseases and
-different instances of outbreaks on the same scale.
+\sout{ Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
+Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
+}\footnote{cool
+  facts about r0, but not the central point}
 
@@
 also highlighted tie to time-invariant analysis and $R_0$.
-I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}
-
-\textcolor{violet}{[Ben says: this paragraph needs to still be looked at.
+I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}\textcolor{orange}{ Shannon tries again in \textcolor{blue}{blue}}
+
+\sout{\textcolor{violet}{[Ben says: this paragraph needs to still be looked at.
 Also I'm not sure why this particular paragraph was changed so much at all.
-Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]}
-\textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.}
-For example, consider two epidemics generated from the Kermack and
-McKendrick SIR equations. The first epidemic has parameters
-\(\beta_1, \gamma_1 = (0.8,0.4)\) and the second has
-\(\beta_2, \gamma_2 = (0.64,0.32)\). Both epidemics have populations of
-1000 people with 10 individuals initially infected. Additionally note
-that the two reproduction numbers are the same for each epidemic,
-\(R_0 = 2=0.8/0.4 = 0.64/0.32\). We plot the epidemics with traditional
-\(state\) vs.~\(time\)
-plots\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
-In Fig. \ref{fig:different-scales-standard} we show the time-based paths
-for the \(S\), \(I\), and \(R\) states for the first 15 days of observed
-data. In this time-variant view, we may believe that
-\textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
-\textcolor{violet}{epidemic} 2 because the peak of infection occurs more
-quickly than in Epidemic 2. On the other hand, we may believe
-\textcolor{violet}{epidemic} 2 has a larger \(R_0\) because its unclear
-if the number of infections in that \textcolor{violet}{epidemic} has not
-yet peaked at time 15. In this time-variant view, we cannot determine if
-one epidemic has larger value of
-\(R_0\)\footnote{\textcolor{violet}{This sentnce doesn't connect wit previous examples.}}.
+Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
+The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
+Both epidemics have populations of 1000 people with 10 individuals initially infected.
+Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.
+We plot the epidemics with traditional $state$ vs.
+$time$ plots}\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
+\sout{In Fig.
+\ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.
+In this time-variant view, we may believe that \textcolor{violet}{epidemic} 1 has a larger $R_0$ than \textcolor{violet}{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.
+On the other hand, we may believe \textcolor{violet}{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \textcolor{violet}{epidemic} has not yet peaked at time 15.
+In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$}\footnote{\textcolor{violet}{This sentnce doesn't connect wit previous examples.}}.
+
+\textcolor{blue}{Since $R_0$ is an important value, it would be helpful to have more intuitive ways of comparing one $R_0$ to another.
+Usually numerical summaries of $R_0$ are presented, which while overall very helpful, may be confusing when presented along side epidemic data that are visualized in a traditional, time-dependent manner.}
+
+\textcolor{blue}{For example, consider two epidemics generated from the Kermack and McKendrick SIR equations where both have the same value of $R_0$.
+The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
+Both epidemics have populations of 1000 people with 10 individuals initially infected.
+An analysis may present an estimate of $\hat{R}_0 = 2$ alongside state vs.
+time plots like those shown in Figure \ref{fig:different-scales-standard}.
+The paths of the epidemics in the state vs.
+time view seem to differ from one another including having different infection peaks.
+From these traditional time-based plots, there is no intuitive way to conclude that these two epidemics have the same value of $R_0$.}
 
@@
 
-\textcolor{orange}{NEW} A time-invariant approach to visualizing
-epidemics, in comparison, allows us to directly compare \(R_0\) from a
-single plot.
-\textcolor{violet}{For every time point $t$ we have $S(t),I(t), R(t)$, so we can treat epdiemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.
-For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point ina a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}'s right subplot.
+\textcolor{blue}{\pkg{EpiCompare} provides a time-invariant tool to visualize these epidemics in a more intuitive manner, at least in regards to comparing values of $R_0$.}
+\sout{A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.}
+\textcolor{violet}{For every time \sout{point} $t$ we have a \textcolor{orange}{point} $(S(t),I(t), R(t))$\sout{, so we can treat epidemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.} \textcolor{orange}{so we can visualize the trajectory of the epidemic in three-dimensional space (see Fig.
+\ref{fig:different-scales-tern} (left)).} For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern} (right).
 \sout{In Fig.
@@
 
-\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
-Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.
-These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
+\caption{\label{fig:different-scales-tern}Left: trajectory of epidemic in three-dimensional space, plotting $(S(t), I(t), R(t))$.
+Right: the gray-shaded region and epidemic trajectory shown from (left) now shown in two-dimensional space.
+This is more commonly known as a ternary plot.}\label{fig:unnamed-chunk-3}
 \end{figure}
@@
 \] \textcolor{violet}{[Ben says: removed this paragraph now]} Since the
-number in each state is non-negative and the sum over the three states
-for a given time point sums to \(N\), then
+number \textcolor{orange}{of individuals} in each state is non-negative
+and the sum over the three states for a given time point
+\sout{sums to}\textcolor{orange}{is} \(N\), then
 \textcolor{violet}{all points in} \(\psi\) will lay in a
@@
 
-\textcolor{violet}{As suggested a few paragraphs back, t}he filamental
-trajectories in Fig. \ref{fig:different-scales-tern} seem to overlap,
-and we may suspect that something is fundamentally linking these two
-different epidemics together. Mathematically, we can show this
-fundamental link turns is \(R_0\). Let our two epidemics be presented as
+\sout{\textcolor{violet}{As suggested a few paragraphs back, t}}The
+filamental trajectories in Fig. \ref{fig:different-scales-tern}
+\sout{seem to overlap}, and we may suspect that something is
+fundamentally linking these two different epidemics together.
+Mathematically, we can show this fundamental link \sout{turns} is
+\(R_0\). Let our two epidemics be presented as
 \(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
@@
 \subsection[Beyond R0 and SIR]{Time-invariant analysis beyond \(R_0\)
-and Kermack's and McKendrick SIR
+and \sout{Kermack’s and McKendrick} SIR
 Models\footnote{Probably will need to change this title...}}\label{sec:beyond-r0-sir}
@@
 filamental trajectories embedded in a lower dimensional space allows us
-to better compare the overall structure of the epidemic and see how the
-population was directly impacted. Time-invariant tools \sout{that} can
-be useful even when the underlying generative model for the epidemic is
-unknown or have more than three epidemic states.
+to \sout{better}\textcolor{orange}{more fully} compare the overall
+structure of the epidemic and see how the population is directly
+impacted. Time-invariant tools \sout{that} can be useful even when the
+underlying generative model for the epidemic is unknown or have more
+than three epidemic states.
 
 \textcolor{violet}{New paragraph} Viewing epidemics as filamental
-trajectories provide \sout{a lot }new ways to compare and examine
-epidemics in a time-invariant manner.
-\textcolor{violet}{For epidemics that have ended, one way to examine their filamental trajectories is to redefine the trajectory as a finite sequence of points on the filamental trajectory that are equally spaced (i.e.
-equa-distance between pairs of ordered points).}
+trajectories provides \sout{a lot }new ways to compare and examine
+epidemics in a time-invariant manner. \textbackslash ben\{For
+\textcolor{orange}{completed?}epidemics that have ended, one way to
+examine their filamental trajectories is to
+\sout{redefine}\textcolor{orange}{represent} the
+\textcolor{orange}{filamental} trajectory as a
+\textcolor{orange}{finite sequence of equally spaced points.}
+\sout{finite sequence of points on the filamental trajectory that are equally spaced (i.e. equa-distance between pairs of ordered points).}
 \sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \textcolor{orange}{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \textcolor{orange}{[Shannon says: are we missing a few words in this definition?]}}.
@@
 \] where \(\psi_i'(s)\) the point along \(\psi_1\) that is \(s\)
-fraction of \(|\psi_1|\) distance away from the start of \(\psi_i\).
+fraction of \(|\psi_1|\) distance away from the start of
+\(\psi_i\).\footnote{\textcolor{orange}{I think you're trying to say something about a distance based on the equally space points.
+Some clarifying questions:  is $\psi^\prime$ the derivative?
+Does proportion make more sense than fraction?
+or simply $\frac{|\psi|}{s}$?
+It's only naturally time-invariant if we have a well defined ending point, right?}}
 This distance is naturally time-invariant, and can be plugged into
 multiple distance-based assessment tools to examine the overall
-extremeness of points, including psuedo-density estimators and
+``extremeness'\,' of points, including pseudo-density estimators and
 depth/local depth functions
 \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
-estiamtors can be very useful when comparing between a set of simulation
-epidemics and the true epidemic, and does not constrain the number of
-states of the
+estimators can be \sout{very} useful when comparing
+\sout{between a setof simulation} \textcolor{orange}{a set of simulated}
+epidemics and the true
+epidemic\textcolor{orange}{. Moreover these extremness estimators,}
+\sout{and does} \textcolor{orange}{} not constrain the number of states
+of the
 model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\textcolor{violet}{I'm not sure we've talked about this before...
@@
 \textcolor{violet}{New paragraph:}
-\textcolor{violet}{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If one is interested in understanding epeidmeics through a single cycle of their outbreak}
+\textcolor{violet}{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If \sout{one} \textcolor{orange}{a practicioner} is interested in understanding \textcolor{orange}{an} epidemic\sout{s} through a single \sout{cycle}\textcolor{orange}{realization} of \sout{their}\textcolor{orange}{its} outbreak}
 (before the population of individuals have become susceptible again),
 then additional time-invariant tools, including prediction regions can
-we leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done?}\}}.
-In these settings we go a step farther and treat epidemics more like
-geometric filaments
+be leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done? Update 4/6: I'm satisfied.}\}}.
+In these settings, \textcolor{orange}{\pkg{EpiCompare} goes}
+\sout{we go }a step further and treats epidemics more like geometric
+filaments
 \textcolor{orange}{(i.e. filamental trajectories without an ordering of points)}
-than filamental trajectories. In \pkg{EpiCompare} we create prediction
+than filamental trajectories. In \pkg{EpiCompare}, we create prediction
 regions that contain a the top (\(1-\alpha\)) proportion of simulated
 curves by defining geometric regions defined by the union of small
-filaments around the subset of simulations (subset by measures like the
-above pseudo-density estimates or depth estimates). These regions look
-at where in the state-space we expect the epidemic to traverse, and we
-can compare prediction regions defined by different models using
+\textcolor{orange}{geometric?} filaments around the subset of
+simulations (\sout{subset} \textcolor{orange}{grouped} by measures like
+the above pseudo-density estimates or depth estimates). These regions
+\sout{look at}\textcolor{orange}{show} where in the state-space we
+expect the epidemic to traverse\sout{, and }.
+\textcolor{orange}{Additionally, }we can compare prediction regions
+defined by different models using
 \textcolor{violet}{many set difference distances \sout{the Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance}}
@@
 by examining if the epidemic's trajectory lies within the prediction
-region. All these geometric structures and distances notations apply to
-epidemics with any number of states, and at the end of Section
-\ref{sec:overview} we \sout{also} highlight how these prediction regions
-can aid in visual comparisons for epidemics with 3 states (like the SIR
-models).
+region. All these \textcolor{orange}{mentioned?} geometric structures
+and distance notations apply to epidemics with any number of states, and
+at the end of Section \ref{sec:overview} we \sout{also} highlight how
+these prediction regions can aid in visual comparisons for epidemics
+with 3 states (like the SIR models).
 
@@
 
+\textcolor{orange}{READ}
+
 \textbf{Post-processing}

### 2021-04-14 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Edits to section 3.  I think it now lines up really well with the figure and gives a good picture of what EpiCompare does!

@@
 
-\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories\sout{ and not points produced by functions of time} \shannon{(this concept will be explained in  explained in Section \ref{sec:r0})}.
+\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories.
 The following section aims to highlight the strengths of \textit{time-invariant} \shannon{analysis} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
@@
 Time-invariant analysis makes different epidemics easier to compare by removing many time dependent aspects of an epidemic.
-Instead, \shannon{time-invariant analysis} focuses on \sout{the overall \shannon{shape and direction, a filamental trajctory,} \sout{trajectory of }an epidemic.
-\shannon{These filamental trajectories,  in turn,} } \shannon{filamental trajectories, which emphasize the number of lives affected.}
+Instead, \shannon{time-invariant analysis} focuses \shannon{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.} \ben{[Ben wants to try this sentence again.]}
 
@@
 \ben{This \shannon{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\shannon{.
-Yet $R_0$ is} estimated with time-\sout{based}\shannon{dependent} data, which can make it a [difficult quantity to estimate.]} \ben{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
-However, data informing models are often temporally anchored (e.g.
-weekly case counts).}}\footnote{\ben{This was a bit unclear including defining a new term.
-As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.
-For example,}}\shannon{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\shannon{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally}\ben{hard to define precisely.\sout{do not have precise definitions}}.
+Yet $R_0$ is} estimated with time-\sout{based}\shannon{dependent} data, which can make it a difficult quantity to estimate.} \shannon{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\shannon{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\ben{hard to define precisely.\sout{do not have precise definitions}}.
 To demonstrate the difficulty of discerning $R_0$ in \ben{\sout{a}other}\footnote{\ben{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}} time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model.
@@
 All front-facing functions are aimed to be as user-friendly as possible.
-We also focus on providing the user "tidyverse" style functions, that encourage piping and also follow clear verb naming schemes \citep{Wickham2019}.
-Although users can typically incorporate \pkg{EpiCompare} into any step in the data analysis process, there are two primary points of entry.
+We also focus on providing the user "tidyverse" style functions, that encourage piping objects from one function to the next and follow clear "verb" naming schemes \citep{Wickham2019}.
+Although users can \sout{typically} incorporate \pkg{EpiCompare} into any step in the data analysis \sout{process}\shannon{pipeline}, there are two primary points of entry.
 The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation.
-Figure \ref{fig:pipeline2} captures these different paths, and we will highlight\footnote{\ben{[Ben says: we need to make sure we actually do highlight ]}} both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
+Figure \ref{fig:pipeline2} captures these different paths, and we highlight\footnote{\ben{[Ben says: we need to make sure we actually do highlight ]}} both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
 
@@
 
-The first step of most data analysis is cleaning the data so it can be explored.
-There are multiple ways to collect epidemiological data.
-Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+The first step of most data analysis is "cleaning" the data \shannon{to a format that is friendly for both computers and programmers} so it can be explored.
+\sout{There are multiple ways to collect epidemiological data.} \shannon{Epidemiological data are collected in many different formats.} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
 Other data collections focus on aggregate counts of individuals in each epidemic state.
-In fact, usually only the number of new infections at each time step (e.g.
+In fact, \sout{usually}\shannon{many times} only the number of new infections at each time step (e.g.
 weekly case counts) is observed.
-Compartment totals (amounts of individuals in each state) are then imputed from those case counts along with other information about the disease and the population of interest.
-In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for examination of overall trends of an epidemic more easily.
-
-In order to help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate`.
+Compartment totals (amounts of individuals in each state) are then imputed from those case counts\sout{ along with}\shannon{and} other information about the disease and the population of interest.
+In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for \shannon{streamlined} examination of overall trends of an epidemic.\sout{ more easily.}
+
+\sout{In order t},To help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate()`.
 This function transforms data about individual/agents' initial entry into each state (e.g.
@@
 There are often situations where grouping agents into subpopulations (e.g.
-subpopulations defined by age or sex) can highlight different aggregate level trends.
-For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining \pkg{dplyr}::`group_by` and `agent_to_aggregate` to provide aggregation at a group level.
-
-The `agents_to_aggregate` function is flexible and can deal with a wide range of information about each individual.
-It can, theoretically, account for infinitely many states.
-This functionality allows the practitioner to aggregate information relative to the standard states (e.g.
-"Susceptible", "Infectious", and "Recovered") as well as add states (e.g.
-"Exposed", "iMmune", "Hospitalized") (CITE\footnote{\ben{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}}).
-Additionally, `agents_to_aggregate` also permits indicators for death/exit and birth/entry dates.
-Overall, the function `agents_to_aggregate()` is a powerful tool for pre-processing data.
+sub\sout{populations}\shannon{ groups} defined by age or sex) can highlight different aggregate level trends.
+For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for \shannon{differing disease dynamics in different} subpopulations.
+In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining \shannon{the function} \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation \sout{at a}\shannon{by} group level.
+
+The `agents_to_aggregate()` function is flexible and can deal with a wide range of information about each individual.
+\sout{It can,} Theoretically, \shannon{this `agents_to_aggregate()` can} account for infinitely many states.
+This functionality allows the practitioner to aggregate information relative to \sout{the standard}\shannon{common} states (e.g.
+"Susceptible", "Infectious", and "Recovered") as well as \sout{add}\shannon{more complex} states (e.g.
+"Exposed", "iMmune", "Hospitalized") (CITE\footnote{\ben{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}\shannon{Shannon says: I changed 'standard' to 'common' so maybe that's a less loaded word.
+Otherwise I think I'd just cite anderson and may 1992}}).
+Additionally, `agents_to_aggregate()` \sout{also} permits indicators for death/exit and birth/entry dates.
+Overall, this function is a powerful tool for pre-processing data, \shannon{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\shannon{Shannon says: I added this in theme with 'highlighting a point of entry'}}
 
@@
 
-With raw data, "getting to know" our data currently means figuring out good combinations of visualizations, numerical summaries and subsets.
-An expert coder has many ways to successfully explore the data in an aggregate lens using `agents_to_aggregate`.
-\pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.
-Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our `geom_aggregate` provides a rapid way to explore different subpopulations' experience of an epidemic by combining the ideas behind `agents_to_aggregate` for three-state models to examine subpopulation trajectories in a 3d simplex space \citep{Wickham2016, Hamilton2018}.
-\shannon{[Shannon says: come back - think about spaces and trajectories?]}\footnote{\ben{[Ben says: what does this note mean?]}} Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
-\shannon{[Shannon says: make sure SIR is defined before.]}
+\sout{With raw data, }"Getting to know" \sout{our} data currently means figuring out \sout{good}\shannon{useful} combinations of visualizations, numerical summaries and \sout{subsets}\shannon{groupings}.
+An expert coder has many ways to successfully explore the data in an aggregate lens using `agents_to_aggregate()`.
+\shannon{For less experienced coders,} \pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.
+Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, \sout{our}\pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of an epidemic \citep{Wickham2016, Hamilton2018}.
+\sout{By combining the ideas behind agents_to_aggregate for three-state models to examine subpopulation trajectories in a 3d simplex space} \citep{Wickham2016, Hamilton2018}\footnote{\shannon{I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.}}.
+Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our \shannon{3d} simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-\ben{After getting a good sense of what a past or current epidemice looks like through EDA, the next step is often model fitting and/or simulations.
-In this step, and the next step (post-processing), we discuss how to include practioners' use of tools to create models and simulations that exist outside of \pkg{EpiCompare}.} This package does not focus on fitting a model to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
-\ben{After estimating or predicting transition rates between states,} these functions produce individual level \ben{simulations\sout{information}} and can be \ben{\sout{naturally}} combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
-The function `simulate_SIR_agents()` simulates an SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
+\ben{After getting a good sense of what a past or current epidemic looks like with EDA, the next step is often model fitting and/or simulations.
+In this step and the next step (post-processing), we discuss how to \shannon{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.} \sout{include practioners' use of tools to create models and simulations that exist outside of \pkg{EpiCompare}.} \shannon{While \pkg{EpiCompare}}\sout{This package }does not focus on fitting model(s) to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
+\ben{After estimating or predicting transition rates between states,} these functions produce individual-level \ben{simulations\sout{information}} and can be \ben{\sout{naturally}} combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
+The function `simulate_SIR_agents()` simulates a \shannon{basic} SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
 This function allows for easy access to SIR model analysis and comparison.
-Beyond SIR models, the function `simulate_agents()` takes as input a user-specified transition matrix and other epidemic parameters to allow the user to create simulations of an outbreak for \textit{any} number of states and any number of transitions among them.
+Beyond SIR models, the function `simulate_agents()` takes as input a user-specified \shannon{state-}transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.
 This flexibility in states can be used to also reflect group-based dynamics.
-This allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
+\shannon{In turn, t}his allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
 For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.
 
-\shannon{READ}
 **Post-processing**
@@
 \ben{[Ben says: this replaces the first paragraph below]}
-\ben{If the practitioner wishes to compare between models and with their true epidemics, one needs to process their models and simulations.} \ben{In general, [p}ost-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.]  \ben{The summaries can be very complex, and a}[s a result, a number of epidemic modeling \ben{\proglang{R}} packages return a special class, specific to their modeling.
+\ben{If the practitioner wishes to compare \shannon{models-to-observations or even models-to-models }\sout{between models and with their true epidemics,} she needs to \shannon{post}-process her models and simulations}\shannon{to disseminate the results in an easily digestable format.} \ben{In general, [p}ost-processing of modeling and simulation consists of making summary statistics, plots, tables\sout{, and other ways to disseminate information to the public.]}  \ben{The summaries can be very complex, and a}[s a result, a number of epidemic modeling \ben{\proglang{R}} packages return a special class\sout{}, specific to their modeling}.
 The special classes often contain a plethora of information from residuals, model diagnostics, input parameters, and more.
@@
 
-\ben{[Ben says: replaced with above paragraph]} Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.
+[\ben{[Ben says: replaced with above paragraph]} Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.
 For example, comma separated value files (`.csv`) are a  standard way to share information within tables.
@@
 While incredibly useful, these special classes can be difficult for novice coders to navigate.
-
-To this end, we have \ben{provide\sout{adapted}} a series of `fortify`-style \ben{methods\sout{functions}}\footnote{\ben{[Ben says: Not tied to this change.]}}, called `fortify_aggregate()` which transform output from packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
+]
+
+To this end, \pkg{EpiCompare} \ben{provides\sout{adapted}} a series of `fortify`-style \ben{methods\sout{functions}}\footnote{\ben{[Ben says: Not tied to this change.]}}, called `fortify_aggregate()` which transform output from \shannon{infectious disease modeling and simulation} packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
 These fortify functions have output that is consistent with that of `agents_to_aggregate()`.
@@
 
-\ben{[Ben says: new paragraph]} To utilize simulations epidemics in later time-invariant analysis we also provide a function to convert temporally defined epidemics to filamental representations.
-Specifically, we provide the function `filament_compression` to convert simulation(s) to filaments as expressed by presenting the epidemic as a ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space\footnote{\ben{[Ben: this should be cleaned up.]}}.
+\ben{[Ben says: new paragraph]} To utilize simulations of epidemics in later time-invariant analysis we also provide a function to convert temporally defined epidemics to filamental representations.
+Specifically, we provide the function `filament_compression()` to convert simulation(s) to filaments as expressed by presenting the epidemic as a ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space\footnote{\ben{[Ben: this should be cleaned up.
+I also 'highlighted' why this makes epicompare a good entry point]}}
+
+\shannon{Shannon's attempt at above paragraph is below..
+I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.}
+[\pkg{EpiCompare} also provides a tool to convert time-dependent epidemic simulations into their time-invariant filamental representations.
+Specifically, `filament_compression()` converts simulations to filaments (see Section \ref{sec:time-invariant})] so practitioners can view model simulations and results through a time-invariant lens.
+These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis is already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.]
 
@@
 
-As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of potential for time-invariant tools to help compare and assess epidemics and models/simulations.
+\shannon{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}  As introduced in Section \ref{sec:beyond-r0-sir} there's a \sout{lot of}\shannon{much} potential for time-invariant tools to help compare and assess epidemics and models/simulations.
 In \pkg{EpiCompare} we provide a set of comparison and assessment tools for models that extend beyond the standard performance metrics (e.g.
-means squared error, AIC) and focus on assessing the structural information the models capture.
-This approaches work well on models where online one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again)\footnote{\ben{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}}.
-Epidemics are complex objects, and we provide tools to create prediction regions out of simulated epidemics.
-For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations (where the simulations treated as filaments).
+mean squared error or AIC) and focus on assessing the structural information the models capture.
+This approaches work well on models where \sout{online} one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again)\footnote{\ben{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}\shannon{Shannon says I don't think I understand what you are trying to say there.
+We should chat about it.}}.
+Epidemics are complex objects, and we provide tools to create prediction regions \shannon{with differing desired characteristics}\footnote{\shannon{Shannon says: I want to connect epidemics being complex to the fact that prediction is hard and not all prediction regions tell you the same thing}} from simulated epidemics.
+For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion [\shannon{good place for Ben paper cite?}] of the simulations (where the simulations treated as filaments).
 In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
-The functions `create_convex_hull_structure` and `create_delta_ball_structure` create different geometric representations of prediction regions for any state-based model.
-For both of these geometric objects, we provide functions to check if a path is contained (`contained`) and the ability to assess the Hassdorf distance between prediction regions based on simulations from different model (`hassdorf_dist`).
-\ben{[Ben says: should we discuss the details of `geom_prediction_bands` more?]}
-
-We also provide functions to calculate the extremeness of a true epidemic compared to simulated epidemics through the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
-Specifically, functions like  `distance_psuedo_density_function` can calculate a psuedo-density estimate of the true epidemic relative to simulated ones.
-Functions `distance_depth_function` and `local_distance_depth_function` provide depth scores that suggest how geometrically central an epidemic is to simulations.
+The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any state-based model.
+For both of these geometric objects, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hassdorf_dist()`).
+\ben{[Ben says: should we discuss the details of `geom_prediction_bands` more?
+]}\shannon{I don't think so, best left to the vignette and data example}
+
+We also provide functions to calculate the "extremeness" of a true epidemic [\shannon{trajectory?}] compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
+Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
+Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
 \ben{[Ben says: should we define these more in this paragraph?
 Introducted in 2.2.
-as well and not defined.]}
+as well and not defined.]}\shannon{[I'd throw a citation or two but again limit the detail.]}
 
@@
 Perhaps the most useful tool \pkg{EpiCompare} has to offer to the expert, for comparison and assessment of models, is in its post-processing tools which create a standard output.
-It is then a matter of writing a script or function made for that standard output to assess the results from multiple models in the way the user desires\footnote{\ben{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}}.
+It is then a matter of writing a script or function made for that standard output to assess the results from multiple models in the way the user desires\footnote{\ben{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}\shannon{I agree with you now that this section is nice and filled in.
+Let's drop this paragraph.}}.
 
@@
 
-In this section, we highlight a number of the functionality available in \pkg{EpiCompare}.
-These functionality include data cleaning, visualization, simulation, and comparison, in accordance with the data analysis pipeline \ref{fig:pipeline}.
-We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner.
+In this section, we highlight many of the tools available in \pkg{EpiCompare}.
+As previously discussed, these tools include data cleaning; visualization; modeling and simulation; post-processing; and comparison and model assessment, in accordance with the data analysis pipeline (Fig.
+\ref{fig:pipeline}).
+We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner via \pkg{EpiCompare}.
 
@@
 
-With \pkg{EpiCompare}, we can easily obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate}.
-The function \code{agents_to_aggregate} is a key component of \pkg{EpiCompare}, allowing the user to easily switch from an individual-level (i.e.
+With \pkg{EpiCompare}, we can easily obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate()}.
+The function \code{agents_to_aggregate()} is a key component of \pkg{EpiCompare}, allowing the user to easily switch from an individual-level (i.e.
 an agent) view of a disease to an aggregate level.
@@
 the initial symptoms.
-We do this with the below code, and a part of the cumulative incidence data output are shown in Table \ref{tab:cif-rash}.
+We do this with the below code, and a part of the cumulative incidence data output is shown in Table \ref{tab:cif-rash}.
 The argument \code{integer_time_expansion} indicates whether we should include all time points in the recorded range of the data or only when there is a change in the incidence.
@@
 
-One question of interest is the duration between initial onset of prodromes or symptoms and the appearance of the measles rash.
-Since \code{agent_to_aggregate} outputs a tidy-style data frame, it is a simple task to plot the two sets of incidence curves on the same graph (Fig.
+One possible question of interest is the duration between initial onset of prodromes  and the appearance of the measles rash.
+Since \code{agent_to_aggregate()} outputs a tidy-style data frame, it is a simple task to plot the two sets of incidence curves on the same graph (Fig.
 \ref{fig:cifs}).
@@
 Immediately, we see that time invariant infection curve is different for the pre-school class compared to the 1st class.
-In the 1st class, we see about 95\% of the class become infected and less than 10\% of them having recovered, which is indicative of a super-spreading event.
+In the 1st class, we see about 95\% of the class become infected and less than 10\% of them having recovered, which may be indicative of a super-spreading event.
 This suspicion is further confirmed in that `r max(table(hagelloch_raw$IFTO[hagelloch_raw$CL == "1st class"]))` of the `r nrow(hagelloch_raw[hagelloch_raw$CL == "1st class",])` 1st class students have been reportedly infected by the same individual.
@@
 Up to this point, we have used \pkg{EpiCompare} in the context of observed data.
-We also want to compare statistical models, and \pkg{EpiCompare} aids in that process via a simple but dynamic individual-level data generator, conversion tools for popular epidemic model packages, and model assessments.
+We also want to compare statistical models, and \pkg{EpiCompare} aids in that process via a simple yet flexible individual-level simulator, conversion tools for popular epidemic model packages, and model assessments.
 We demonstrate an example here.
@@
 However, both models are not a good fit to the filamental path as opposed to the individual points in $(S, I, R)$-space.
-This can be captures with the set of simulations both models predict, which all generally have a single defined peak of infection whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
-This observation is backed up by the below analysis that demonstrates that the estimated psuedo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme}.
-In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the EpiModel network model is a better fit than the Simple SIR model, and 2) the fit is only good at the individual point level as opposed to the epidemic path level.
+This can be captured with the set of simulations both models predict (gray lines), which all generally have a single defined peak of infection whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
+This observation is backed up by the below analysis that demonstrates that the estimated pseudo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme}).
+In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the \pkg{EpiModel} network model is a better fit than the Simple SIR model, and 2) the fit is only good at the geometric filamental level as opposed to the epidemic trajectory filamental level.
 
@@
 ```
+
+Overall, \pkg{EpiCompare} aids in the data analysis pipeline for both novice and expert practitioners and coders alike.
+These tools encourage model and simulation exploration of many of the existing and well-supported packages that already exist, and side-by-side comparison thereof.
+Finally, we hope that practicioners will consider using time-invariant analysis when trying to assess and compare epidemics and epidemic models.
 

### 2021-04-14 jss/EpiCompare-paper/EpiCompare-paper.tex
Edits to section 3.  I think it now lines up really well with the figure and gives a good picture of what EpiCompare does!

@@
 a global, not marginal view of how epidemics move through populations
-and (2) by treating full epidemics as filamental
-trajectories\sout{ and not points produced by functions of time}
-\textcolor{orange}{(this concept will be explained in  explained in Section \ref{sec:r0})}.
-The following section aims to highlight the strengths of
+and (2) by treating full epidemics as filamental trajectories. The
+following section aims to highlight the strengths of
 \textit{time-invariant} \textcolor{orange}{analysis} and define the
@@
 dependent aspects of an epidemic. Instead,
-\textcolor{orange}{time-invariant analysis} focuses on
-\sout{the overall \textcolor{orange}{shape and direction, a filamental trajctory,} \sout{trajectory of }an epidemic.  \textcolor{orange}{These filamental trajectories,  in turn,} }
-\textcolor{orange}{filamental trajectories, which emphasize the number of lives affected.}
+\textcolor{orange}{time-invariant analysis} focuses
+\textcolor{orange}{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.}
+\textcolor{violet}{[Ben wants to try this sentence again.]}
 
@@
 \textcolor{violet}{This \textcolor{orange}{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\textcolor{orange}{.
-Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a [difficult quantity to estimate.]}
-\textcolor{violet}{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.
-However, data informing models are often temporally anchored (e.g.
-weekly case counts).}}\footnote{\textcolor{violet}{This was a bit unclear including defining a new term.
-As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.
-For example,}}\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
+Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a difficult quantity to estimate.}
+\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
 To demonstrate the difficulty of discerning \(R_0\) in
@@
 also focus on providing the user ``tidyverse'' style functions, that
-encourage piping and also follow clear verb naming schemes
-\citep{Wickham2019}. Although users can typically incorporate
-\pkg{EpiCompare} into any step in the data analysis process, there are
-two primary points of entry. The first point of entry is the very
-beginning with pre-processing and visualizing raw data, and the second
-point of entry is after modeling and simulation. Figure
-\ref{fig:pipeline2} captures these different paths, and we will
+encourage piping objects from one function to the next and follow clear
+``verb'' naming schemes \citep{Wickham2019}. Although users can
+\sout{typically} incorporate \pkg{EpiCompare} into any step in the data
+analysis \sout{process}\textcolor{orange}{pipeline}, there are two
+primary points of entry. The first point of entry is the very beginning
+with pre-processing and visualizing raw data, and the second point of
+entry is after modeling and simulation. Figure \ref{fig:pipeline2}
+captures these different paths, and we
 highlight\footnote{\textcolor{violet}{[Ben says: we need to make sure we actually do highlight ]}}
@@
 
-The first step of most data analysis is cleaning the data so it can be
-explored. There are multiple ways to collect epidemiological data.
+The first step of most data analysis is ``cleaning'' the data
+\textcolor{orange}{to a format that is friendly for both computers and programmers}
+so it can be explored.
+\sout{There are multiple ways to collect epidemiological data.}
+\textcolor{orange}{Epidemiological data are collected in many different formats.}
 Sometimes individual records are collected, with times of different
@@
 information. Other data collections focus on aggregate counts of
-individuals in each epidemic state. In fact, usually only the number of
-new infections at each time step (e.g.~weekly case counts) is observed.
+individuals in each epidemic state. In fact,
+\sout{usually}\textcolor{orange}{many times} only the number of new
+infections at each time step (e.g.~weekly case counts) is observed.
 Compartment totals (amounts of individuals in each state) are then
-imputed from those case counts along with other information about the
-disease and the population of interest. In \pkg{EpiCompare}, we focus on
-understanding the overall impact of an outbreak at the
-aggregate/population level, which allows for examination of overall
-trends of an epidemic more easily.
-
-In order to help the practitioner examine epidemics from an aggregate/
-population lens, we provide a function called
-\texttt{agents\_to\_aggregate}. This function transforms data about
+imputed from those case counts\sout{ along with}\textcolor{orange}{and}
+other information about the disease and the population of interest. In
+\pkg{EpiCompare}, we focus on understanding the overall impact of an
+outbreak at the aggregate/population level, which allows for
+\textcolor{orange}{streamlined} examination of overall trends of an
+epidemic.\sout{ more easily.}
+
+\sout{In order t},To help the practitioner examine epidemics from an
+aggregate/ population lens, we provide a function called
+\texttt{agents\_to\_aggregate()}. This function transforms data about
 individual/agents' initial entry into each state (e.g.~start of
@@
 individuals were in a state at a given time. There are often situations
-where grouping agents into subpopulations (e.g.~subpopulations defined
-by age or sex) can highlight different aggregate level trends. For
-example, research by \citet{rvachev1985,anderson1992,worby2015} develop
-state-based models that account for subpopulations. In \pkg{EpiCompare},
-we facilitate subpopulation analysis by combining
-\pkg{dplyr}::\texttt{group\_by} and \texttt{agent\_to\_aggregate} to
-provide aggregation at a group level.
-
-The \texttt{agents\_to\_aggregate} function is flexible and can deal
-with a wide range of information about each individual. It can,
-theoretically, account for infinitely many states. This functionality
-allows the practitioner to aggregate information relative to the
-standard states (e.g.~``Susceptible'', ``Infectious'', and
-``Recovered'') as well as add states (e.g.~``Exposed'', ``iMmune'',
-``Hospitalized'')
-(CITE\footnote{\textcolor{violet}{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}}).
-Additionally, \texttt{agents\_to\_aggregate} also permits indicators for
-death/exit and birth/entry dates. Overall, the function
-\texttt{agents\_to\_aggregate()} is a powerful tool for pre-processing
-data.
+where grouping agents into subpopulations
+(e.g.~sub\sout{populations}\textcolor{orange}{ groups} defined by age or
+sex) can highlight different aggregate level trends. For example,
+research by \citet{rvachev1985,anderson1992,worby2015} develop
+state-based models that account for
+\textcolor{orange}{differing disease dynamics in different}
+subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation
+analysis by combining \textcolor{orange}{the function}
+\pkg{dplyr}::\texttt{group\_by()} and \texttt{agent\_to\_aggregate()} to
+provide aggregation \sout{at a}\textcolor{orange}{by} group level.
+
+The \texttt{agents\_to\_aggregate()} function is flexible and can deal
+with a wide range of information about each individual. \sout{It can,}
+Theoretically, account for infinitely many states. This functionality
+allows the practitioner to aggregate information relative to
+\sout{the standard}\textcolor{orange}{common} states
+(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as
+\sout{add}\textcolor{orange}{more complex} states (e.g.~``Exposed'',
+``iMmune'', ``Hospitalized'')
+(CITE\footnote{\textcolor{violet}{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}\textcolor{orange}{Shannon says: I changed 'standard' to 'common' so maybe that's a less loaded word.
+Otherwise I think I'd just cite anderson and may 1992}}).
+Additionally, \texttt{agents\_to\_aggregate()} \sout{also} permits
+indicators for death/exit and birth/entry dates. Overall, this function
+is a powerful tool for pre-processing data,
+\textcolor{orange}{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\textcolor{orange}{Shannon says: I added this in theme with 'highlighting a point of entry'}}
 
@@
 
-With raw data, ``getting to know'' our data currently means figuring out
-good combinations of visualizations, numerical summaries and subsets. An
-expert coder has many ways to successfully explore the data in an
-aggregate lens using \texttt{agents\_to\_aggregate}. \pkg{EpiCompare}
-also includes tools to rapidly explore data that has three epidemic
-states. Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our
-\texttt{geom\_aggregate} provides a rapid way to explore different
-subpopulations' experience of an epidemic by combining the ideas behind
-\texttt{agents\_to\_aggregate} for three-state models to examine
-subpopulation trajectories in a 3d simplex space
-\citep{Wickham2016, Hamilton2018}.
-\textcolor{orange}{[Shannon says: come back - think about spaces and trajectories?]}\footnote{\textcolor{violet}{[Ben says: what does this note mean?]}}
+\sout{With raw data, }``Getting to know'' \sout{our} data currently
+means figuring out \sout{good}\textcolor{orange}{useful} combinations of
+visualizations, numerical summaries and
+\sout{subsets}\textcolor{orange}{groupings}. An expert coder has many
+ways to successfully explore the data in an aggregate lens using
+\texttt{agents\_to\_aggregate()}.
+\textcolor{orange}{For less experienced coders,} \pkg{EpiCompare} also
+includes tools to rapidly explore data that has three epidemic states.
+Building on the tools in \pkg{ggplot2} and \pkg{ggtern},
+\sout{our}\pkg{EpiCompare}`s \texttt{geom\_aggregate()} provides a rapid
+way to explore different subpopulations' experiences of an epidemic
+\citep{Wickham2016, Hamilton2018}. \textbackslash sout\{By combining the
+ideas behind agents\_to\_aggregate for three-state models to examine
+subpopulation trajectories in a 3d simplex space\}
+\citep{Wickham2016, Hamilton2018}\footnote{\textcolor{orange}{I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.}}.
 Visualization tools for three-state models were developed because (1)
 SIR models are some of the most common and basic epidemic state-based
-models and (2) our simplex representation of these epidemics emphasizes
-a ``time-invarance'' representation of the data (for a refresher see
-Section \ref{sec:time-invariant}).
-\textcolor{orange}{[Shannon says: make sure SIR is defined before.]}
+models and (2) our \textcolor{orange}{3d} simplex representation of
+these epidemics emphasizes a ``time-invarance'' representation of the
+data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-\textcolor{violet}{After getting a good sense of what a past or current epidemice looks like through EDA, the next step is often model fitting and/or simulations.
-In this step, and the next step (post-processing), we discuss how to include practioners' use of tools to create models and simulations that exist outside of \pkg{EpiCompare}.}
-This package does not focus on fitting a model to data, we do provide
-some flexible functions for simulation of basic discrete-time
-epidemic-state models.
-\textcolor{violet}{After estimating or predicting transition rates between states,}
-these functions produce individual level
-\textcolor{violet}{simulations\sout{information}} and can be
-\textcolor{violet}{\sout{naturally}} combined with
-\texttt{agents\_to\_aggregate()} to view these simulations through an
-aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
-an SIR epidemic with user inputs for the number of simulations, the
-initial number in each state, the infection and recovery parameters
-\((\beta, \gamma\)), and the total number of discrete time steps. This
-function allows for easy access to SIR model analysis and comparison.
-Beyond SIR models, the function \texttt{simulate\_agents()} takes as
-input a user-specified transition matrix and other epidemic parameters
-to allow the user to create simulations of an outbreak for \textit{any}
-number of states and any number of transitions among them. This
-flexibility in states can be used to also reflect group-based dynamics.
-This allows for users to explore the space of models in an intuitive way
-without getting bogged down by too much mathematical detail. For
-consistency, we have made output from \texttt{simulate\_agents()} and
-\texttt{simulate\_SIR\_agents()} compatible with
-\texttt{agents\_to\_aggregate()} so aggregate information may easily be
-accessed.
-
-\textcolor{orange}{READ}
-
-\textbf{Post-processing}
-
-\textcolor{violet}{[Ben says: I think we should remind the reader that we care more about simulations, in order to compare fitted models between themselves and the true epidemics.]}
-
-\textcolor{violet}{[Ben says: this replaces the first paragraph below]}
-
-\textcolor{violet}{If the practitioner wishes to compare between models and with their true epidemics, one needs to process their models and simulations.}
-\textcolor{violet}{In general, [p}ost-processing of modeling and
-simulation consists of making summary statistics, plots, tables, and
-other ways to disseminate information to the public.{]}
-\textcolor{violet}{The summaries can be very complex, and a}{[}s a
-result, a number of epidemic modeling \textcolor{violet}{\proglang{R}}
-packages return a special class, specific to their modeling. The special
-classes often contain a plethora of information from residuals, model
-diagnostics, input parameters, and more. While incredibly useful, these
-special classes can be difficult for novice coders to navigate.{]}
-
-\textcolor{violet}{[Ben says: replaced with above paragraph]}
+. The special classes often contain a plethora of information from
+residuals, model diagnostics, input parameters, and more. While
+incredibly useful, these special classes can be difficult for novice
+coders to navigate.{]}
+
+{[}\textcolor{violet}{[Ben says: replaced with above paragraph]}
 Post-processing of modeling and simulation consists of making summary
@@
 parameters, and more. While incredibly useful, these special classes can
-be difficult for novice coders to navigate.
-
-To this end, we have \textcolor{violet}{provide\sout{adapted}} a series
-of \texttt{fortify}-style
+be difficult for novice coders to navigate. {]}
+
+To this end, \pkg{EpiCompare} \textcolor{violet}{provides\sout{adapted}}
+a series of \texttt{fortify}-style
 \textcolor{violet}{methods\sout{functions}}\footnote{\textcolor{violet}{[Ben says: Not tied to this change.]}},
 called \texttt{fortify\_aggregate()} which transform output from
-packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
-which contain information about the total number of individuals in each
-state at a given time, for a given simulation. These fortify functions
-have output that is consistent with that of
-\texttt{agents\_to\_aggregate()}.
-
-\textcolor{violet}{[Ben says: new paragraph]} To utilize simulations
+\textcolor{orange}{infectious disease modeling and simulation} packages
+like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which
+contain information about the total number of individuals in each state
+at a given time, for a given simulation. These fortify functions have
+output that is consistent with that of \texttt{agents\_to\_aggregate()}.
+
+\textcolor{violet}{[Ben says: new paragraph]} To utilize simulations of
 epidemics in later time-invariant analysis we also provide a function to
 convert temporally defined epidemics to filamental representations.
-Specifically, we provide the function \texttt{filament\_compression} to
-convert simulation(s) to filaments as expressed by presenting the
+Specifically, we provide the function \texttt{filament\_compression()}
+to convert simulation(s) to filaments as expressed by presenting the
 epidemic as a ordering of some common fixed number of points so that
@@
 state
-space\footnote{\textcolor{violet}{[Ben: this should be cleaned up.]}}.
+space\footnote{\textcolor{violet}{[Ben: this should be cleaned up.  I also 'highlighted' why this makes epicompare a good entry point]}}
+
+\textcolor{orange}{Shannon's attempt at above paragraph is below..
+I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.}
+
+{[}\pkg{EpiCompare} also provides a tool to convert time-dependent
+epidemic simulations into their time-invariant filamental
+representations. Specifically, \texttt{filament\_compression()} converts
+simulations to filaments (see Section \ref{sec:time-invariant}){]} so
+practitioners can view model simulations and results through a
+time-invariant lens. These tools were developed to provide another
+natural entry point into the \pkg{EpiCompare} data analysis pipeline for
+situations where modeling and analysis is already completed and
+practitioners are looking for simple and transparent tools to help
+understand and disseminate results.{]}
 
@@
 
-As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of
-potential for time-invariant tools to help compare and assess epidemics
-and models/simulations. In \pkg{EpiCompare} we provide a set of
-comparison and assessment tools for models that extend beyond the
-standard performance metrics (e.g.~means squared error, AIC) and focus
-on assessing the structural information the models capture. This
-approaches work well on models where online one ``cycle'' of the
-epidemic has occurred (no recovered individuals have been susceptible
-again)\footnote{\textcolor{violet}{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}}.
+\textcolor{orange}{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}
+As introduced in Section \ref{sec:beyond-r0-sir} there's a
+\sout{lot of}\textcolor{orange}{much} potential for time-invariant tools
+to help compare and assess epidemics and models/simulations. In
+\pkg{EpiCompare} we provide a set of comparison and assessment tools for
+models that extend beyond the standard performance metrics (e.g.~mean
+squared error or AIC) and focus on assessing the structural information
+the models capture. This approaches work well on models where
+\sout{online} one ``cycle'' of the epidemic has occurred (no recovered
+individuals have been susceptible
+again)\footnote{\textcolor{violet}{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}\textcolor{orange}{Shannon says I don't think I understand what you are trying to say there.
+We should chat about it.}}.
 Epidemics are complex objects, and we provide tools to create prediction
-regions out of simulated epidemics. For three-state epidemic models, we
-provide the \texttt{ggplot}/\texttt{ggtern} extension
+regions
+\textcolor{orange}{with differing desired characteristics}\footnote{\textcolor{orange}{Shannon says: I want to connect epidemics being complex to the fact that prediction is hard and not all prediction regions tell you the same thing}}
+from simulated epidemics. For three-state epidemic models, we provide
+the \texttt{ggplot}/\texttt{ggtern} extension
 \texttt{geom\_prediction\_band()} which creates a prediction region
-around the top \(1-\alpha\) proportion of the simulations (where the
-simulations treated as filaments). In \pkg{EpiCompare} we also provide
-these prediction regions for epidemic models with with more than three
-states. The functions \texttt{create\_convex\_hull\_structure} and
-\texttt{create\_delta\_ball\_structure} create different geometric
+around the top \(1-\alpha\) proportion
+{[}\textcolor{orange}{good place for Ben paper cite?}{]} of the
+simulations (where the simulations treated as filaments). In
+\pkg{EpiCompare} we also provide these prediction regions for epidemic
+models with with more than three states. The functions
+\texttt{create\_convex\_hull\_structure()} and
+\texttt{create\_delta\_ball\_structure()} create different geometric
 representations of prediction regions for any state-based model. For
 both of these geometric objects, we provide functions to check if a path
-is contained (\texttt{contained}) and the ability to assess the Hassdorf
-distance between prediction regions based on simulations from different
-model (\texttt{hassdorf\_dist}).
-
-We also provide functions to calculate the extremeness of a true
-epidemic compared to simulated epidemics through the equi-distance
-filamental trajectory representation as mentioned in Section
-\ref{sec:beyond-r0-sir}. Specifically, functions like
-\texttt{distance\_psuedo\_density\_function} can calculate a
-psuedo-density estimate of the true epidemic relative to simulated ones.
-Functions \texttt{distance\_depth\_function} and
-\texttt{local\_distance\_depth\_function} provide depth scores that
+is contained (\texttt{contained()}) and the ability to assess the
+Hausdorff distance between prediction regions based on simulations from
+different model (\texttt{hassdorf\_dist()}).
+\textcolor{orange}{I don't think so, best left to the vignette and data example}
+
+We also provide functions to calculate the ``extremeness'' of a true
+epidemic {[}\textcolor{orange}{trajectory?}{]} compared to simulated
+epidemics via the equi-distance filamental trajectory representation as
+mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions
+like \texttt{distance\_pseudo\_density\_function()} can calculate a
+pseudo-density estimate of the true epidemic relative to simulated ones.
+Functions \texttt{distance\_depth\_function()} and
+\texttt{local\_distance\_depth\_function()} provide depth scores that
 suggest how geometrically central an epidemic is to simulations.
-\textcolor{violet}{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}
+\textcolor{violet}{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}\textcolor{orange}{[I'd throw a citation or two but again limit the detail.]}
 
@@
 in the way the user
-desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}}.
+desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}\textcolor{orange}{I agree with you now that this section is nice and filled in.
+Let's drop this paragraph.}}.
 
@@
 
-In this section, we highlight a number of the functionality available in
-\pkg{EpiCompare}. These functionality include data cleaning,
-visualization, simulation, and comparison, in accordance with the data
-analysis pipeline \ref{fig:pipeline}. We show a full data analysis from
+In this section, we highlight many of the tools available in
+\pkg{EpiCompare}. As previously discussed, these tools include data
+cleaning; visualization; modeling and simulation; post-processing; and
+comparison and model assessment, in accordance with the data analysis
+pipeline (Fig. \ref{fig:pipeline}). We show a full data analysis from
 beginning to end that can be accomplished in a streamlined and
-standardized manner.
+standardized manner via \pkg{EpiCompare}.
 
@@
 \code{ERU}) with the following tidy-style function,
-\code{agents_to_aggregate}. The function \code{agents_to_aggregate} is a
-key component of \pkg{EpiCompare}, allowing the user to easily switch
-from an individual-level (i.e.~an agent) view of a disease to an
+\code{agents_to_aggregate()}. The function \code{agents_to_aggregate()}
+is a key component of \pkg{EpiCompare}, allowing the user to easily
+switch from an individual-level (i.e.~an agent) view of a disease to an
 aggregate level. For example, the below code shows how we can convert
@@
 incidence of the prodromes, i.e.~the initial symptoms. We do this with
-the below code, and a part of the cumulative incidence data output are
+the below code, and a part of the cumulative incidence data output is
 shown in Table \ref{tab:cif-rash}. The argument
@@
 
-One question of interest is the duration between initial onset of
-prodromes or symptoms and the appearance of the measles rash. Since
-\code{agent_to_aggregate} outputs a tidy-style data frame, it is a
+One possible question of interest is the duration between initial onset
+of prodromes and the appearance of the measles rash. Since
+\code{agent_to_aggregate()} outputs a tidy-style data frame, it is a
 simple task to plot the two sets of incidence curves on the same graph
@@
 1st class. In the 1st class, we see about 95\% of the class become
-infected and less than 10\% of them having recovered, which is
+infected and less than 10\% of them having recovered, which may be
 indicative of a super-spreading event. This suspicion is further
@@
 observed data. We also want to compare statistical models, and
-\pkg{EpiCompare} aids in that process via a simple but dynamic
-individual-level data generator, conversion tools for popular epidemic
-model packages, and model assessments. We demonstrate an example here.
+\pkg{EpiCompare} aids in that process via a simple yet flexible
+individual-level simulator, conversion tools for popular epidemic model
+packages, and model assessments. We demonstrate an example here.
 
@@
 opposed to the individual points in \((S, I, R)\)-space. This can be
-captures with the set of simulations both models predict, which all
-generally have a single defined peak of infection whereas the data
-certainly looks like it has two distinct peaks, likely caused by our
-assumed super-spreader event. This observation is backed up by the below
-analysis that demonstrates that the estimated psuedo-density of the
-observed epidemic (relative to the simulations from either model) is
+captured with the set of simulations both models predict (gray lines),
+which all generally have a single defined peak of infection whereas the
+data certainly looks like it has two distinct peaks, likely caused by
+our assumed super-spreader event. This observation is backed up by the
+below analysis that demonstrates that the estimated pseudo-density of
+the observed epidemic (relative to the simulations from either model) is
 much less likely then \textbf{any} of the simulations (reported in Table
-\ref{tab:hags-extreme}. In conclusion, \pkg{EpiCompare} makes it clear
-that, at a glance, 1) the EpiModel network model is a better fit than
-the Simple SIR model, and 2) the fit is only good at the individual
-point level as opposed to the epidemic path level.
+\ref{tab:hags-extreme}). In conclusion, \pkg{EpiCompare} makes it clear
+that, at a glance, 1) the \pkg{EpiModel} network model is a better fit
+than the Simple SIR model, and 2) the fit is only good at the geometric
+filamental level as opposed to the epidemic trajectory filamental level.
 
@@
 \end{CodeChunk}
+
+Overall, \pkg{EpiCompare} aids in the data analysis pipeline for both
+novice and expert practitioners and coders alike. These tools encourage
+model and simulation exploration of many of the existing and
+well-supported packages that already exist, and side-by-side comparison
+thereof. Finally, we hope that practicioners will consider using
+time-invariant analysis when trying to assess and compare epidemics and
+epidemic models.
 

### 2021-04-15 jss/EpiCompare-paper/EpiCompare-paper.Rmd
ben responses to comments on section 3 - minus last section which we will discuss in person (first round of edits after completition of section 3)

@@
 We also focus on providing the user "tidyverse" style functions, that encourage piping objects from one function to the next and follow clear "verb" naming schemes \citep{Wickham2019}.
-Although users can \sout{typically} incorporate \pkg{EpiCompare} into any step in the data analysis \sout{process}\shannon{pipeline}, there are two primary points of entry.
+Although users can incorporate \pkg{EpiCompare} into any step in the data analysis pipeline, there are two primary points of entry.
 The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation.
@@
 
-The first step of most data analysis is "cleaning" the data \shannon{to a format that is friendly for both computers and programmers} so it can be explored.
-\sout{There are multiple ways to collect epidemiological data.} \shannon{Epidemiological data are collected in many different formats.} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+The first step of most data analysis is "cleaning" the data \shannon{to a format that is friendly for both computers and programmers}\footnote{\ben{[Ben says: I'm unsure this is needed and naturally adds more text.
+Willing to accept...]}} so it can be explored.
+\sout{There are multiple ways to collect epidemiological data.} \ben{In epidemology, there are multiple different formats the data can arrive in.} \footnote{\shannon{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \ben{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
 Other data collections focus on aggregate counts of individuals in each epidemic state.
-In fact, \sout{usually}\shannon{many times} only the number of new infections at each time step (e.g.
+In fact, many times only the number of new infections at each time step (e.g.
 weekly case counts) is observed.
-Compartment totals (amounts of individuals in each state) are then imputed from those case counts\sout{ along with}\shannon{and} other information about the disease and the population of interest.
-In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for \shannon{streamlined} examination of overall trends of an epidemic.\sout{ more easily.}
-
-\sout{In order t},To help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate()`.
+Compartment totals (amounts of individuals in each state) are then imputed from those case counts\sout{ along with}\shannon{and}\footnote{\ben{[Ben says: Declined.
+I'm unclear why we should.
+Is the sentence unclear?
+Maybe it's the way it's a compound sentence?
+- If so a rewrite would be better?]}} other information about the disease and the population of interest.
+In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.
+
+To help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate()`.
 This function transforms data about individual/agents' initial entry into each state (e.g.
 start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
-There are often situations where grouping agents into subpopulations (e.g.
-sub\sout{populations}\shannon{ groups} defined by age or sex) can highlight different aggregate level trends.
-For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for \shannon{differing disease dynamics in different} subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining \shannon{the function} \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation \sout{at a}\shannon{by} group level.
+There are often situations where \ben{\sout{grouping} aggregating}\footnote{\ben{[Ben says: reason - connect to languare from sentence before.]}} agents into subpopulations (e.g.
+sub\sout{populations}\shannon{ groups}\footnote{\ben{[Ben says: declined.
+I'm unclear why we'd do this.]}} defined by age or sex) can highlight different aggregate level trends.
+For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations.
+In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation \sout{at a}\shannon{by}\footnote{\ben{[Ben says: declined.
+unclear about this change.
+Also would it not be "by groups"?
+Orignial seems to flow more.]}} group level.
 
 The `agents_to_aggregate()` function is flexible and can deal with a wide range of information about each individual. 
-\sout{It can,} Theoretically, \shannon{this `agents_to_aggregate()` can} account for infinitely many states.
-This functionality allows the practitioner to aggregate information relative to \sout{the standard}\shannon{common} states (e.g.
-"Susceptible", "Infectious", and "Recovered") as well as \sout{add}\shannon{more complex} states (e.g.
-"Exposed", "iMmune", "Hospitalized") (CITE\footnote{\ben{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}\shannon{Shannon says: I changed 'standard' to 'common' so maybe that's a less loaded word.
-Otherwise I think I'd just cite anderson and may 1992}}).
-Additionally, `agents_to_aggregate()` \sout{also} permits indicators for death/exit and birth/entry dates.
-Overall, this function is a powerful tool for pre-processing data, \shannon{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\shannon{Shannon says: I added this in theme with 'highlighting a point of entry'}}
+\ben{\sout{It can,} In fact, this function can}\footnote{\shannon{[Shannon suggests: "Theoretically, this agents_to_aggregate() can"]} \ben{[Ben says: I just removed the "theoretically" since it seems to suggest more than we want.
+Also I didn't see your original edit.
+both edits connect to the previous sentence better.]}} account for infinitely many states.
+This functionality allows the practitioner to aggregate information relative to common states (e.g.
+"Susceptible", "Infectious", and "Recovered") as well as more complex states (e.g.
+"Exposed", "iMmune", "Hospitalized").
+Additionally, `agents_to_aggregate()` permits indicators for death/exit and birth/entry dates.
+Overall, this function is a powerful tool for pre-processing data, \shannon{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\shannon{[Shannon says: I added this in theme with 'highlighting a point of entry']} \ben{[Ben says: I'm unclear about this comment but I'm find with the addition...]}}
 
@@
 
-\sout{With raw data, }"Getting to know" \sout{our} data currently means figuring out \sout{good}\shannon{useful} combinations of visualizations, numerical summaries and \sout{subsets}\shannon{groupings}.
-An expert coder has many ways to successfully explore the data in an aggregate lens using `agents_to_aggregate()`.
-\shannon{For less experienced coders,} \pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.
-Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, \sout{our}\pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of an epidemic \citep{Wickham2016, Hamilton2018}.
-\sout{By combining the ideas behind agents_to_aggregate for three-state models to examine subpopulation trajectories in a 3d simplex space} \citep{Wickham2016, Hamilton2018}\footnote{\shannon{I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.}}.
-Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our \shannon{3d} simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+\ben{\sout{With raw data, "Getting to know" our} In the early stages of a project, getting to know the}\footnote{\shannon{[Shannon suggested: "Getting to know" our]} \ben{[Ben says: even though I was the original one with the quotes - one should always stray away from it.
+I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}} data currently means figuring out useful combinations of visualizations \ben{\sout{,}and} numerical summaries and \sout{\sout{subsets}\shannon{groupings}}\ben{exploring different groupings of the data}\footnote{\ben{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}}.
+\ben{\sout{An expert coder has many ways to successfully explore the data in an aggregate lens using }}`agents_to_aggregate()`\ben{\sout{.
+For less experienced coders, \pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.} An expert coder can start from } `agents_to_aggregate()` \ben{to successfully accomplish EDA in many ways, but \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).}\footnote{\ben{[Ben says: I'm reverting to the old text - it's much clearer.
+- happy to have a a discussion on it.
+The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}} Building on \ben{\sout{the tools in}} \pkg{ggplot2} and \pkg{ggtern} \ben{packages}, \pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of an epidemic \citep{Wickham2016, Hamilton2018}.
+\ben{The function }`geom_aggregate()` \ben{provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.}  \footnote{\ben{[Original text: "By combiing the ideas behind agent2aggregate for three-state models to examine subpopulation trajcetories in 2d simplex space."]} \shannon{[Shannons says: I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.]}\ben{[Ben says: I'm unclear why this warrants just deleting it.
+I've provided a newly written sentence...]}}  Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-\ben{After getting a good sense of what a past or current epidemic looks like with EDA, the next step is often model fitting and/or simulations.
-In this step and the next step (post-processing), we discuss how to \shannon{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.} \sout{include practioners' use of tools to create models and simulations that exist outside of \pkg{EpiCompare}.} \shannon{While \pkg{EpiCompare}}\sout{This package }does not focus on fitting model(s) to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
-\ben{After estimating or predicting transition rates between states,} these functions produce individual-level \ben{simulations\sout{information}} and can be \ben{\sout{naturally}} combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
-The function `simulate_SIR_agents()` simulates a \shannon{basic} SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
+After getting a good sense of what a past or current epidemic looks like with EDA, the next step is often model fitting and/or simulations.
+\sout{In this step and the next step (post-processing), we discuss how to \shannon{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\ben{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\ben{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}} While \pkg{EpiCompare} does not focus on fitting model(s) to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
+\ben{These functions simulation individual-level information based on practitioner estimated transition rates between states and} can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
+The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
 This function allows for easy access to SIR model analysis and comparison.
-Beyond SIR models, the function `simulate_agents()` takes as input a user-specified \shannon{state-}transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.
+Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.
 This flexibility in states can be used to also reflect group-based dynamics.
-\shannon{In turn, t}his allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
+In turn, this allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
 For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.
@@
 \ben{[Ben says: this replaces the first paragraph below]}
-\ben{If the practitioner wishes to compare \shannon{models-to-observations or even models-to-models }\sout{between models and with their true epidemics,} she needs to \shannon{post}-process her models and simulations}\shannon{to disseminate the results in an easily digestable format.} \ben{In general, [p}ost-processing of modeling and simulation consists of making summary statistics, plots, tables\sout{, and other ways to disseminate information to the public.]}  \ben{The summaries can be very complex, and a}[s a result, a number of epidemic modeling \ben{\proglang{R}} packages return a special class\sout{}, specific to their modeling}.
-The special classes often contain a plethora of information from residuals, model diagnostics, input parameters, and more.
-While incredibly useful, these special classes can be difficult for novice coders to navigate.]
-
-[\ben{[Ben says: replaced with above paragraph]} Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.
+If the practitioner wishes to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestable format.
+In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.
+\ben{Model \sout{The}} summaries can be very complex, and as a result, a number of epidemic modeling \\proglang{R} packages return \ben{\sout{a}} special class \ben{objects}.
+The special class\ben{\sout{es} objects} often contain a plethora of information from residuals, model diagnostics, input parameters, and more.
+While incredibly useful, these special classes can be difficult for novice coders to \ben{\sout{navigate} handle}.
+
+<!-- [\ben{[Ben says: replaced with above paragraph]} Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.
 For example, comma separated value files (`.csv`) are a  standard way to share information within tables.
@@
 While incredibly useful, these special classes can be difficult for novice coders to navigate.
-]
-
-To this end, \pkg{EpiCompare} \ben{provides\sout{adapted}} a series of `fortify`-style \ben{methods\sout{functions}}\footnote{\ben{[Ben says: Not tied to this change.]}}, called `fortify_aggregate()` which transform output from \shannon{infectious disease modeling and simulation} packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
+] -->
+
+To this end, \pkg{EpiCompare} provides a series of `fortify`-style methods, called `fortify_aggregate()` which transform output from infectious disease modeling and simulation packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
 These fortify functions have output that is consistent with that of `agents_to_aggregate()`.
 
-\shannon{[Shannon says: Here Ben talks about filament compression and tidy_dist mats, etc.]} \ben{[Ben says: not currently thinking about including tidy_dist stuff - will need to think about this as it is inlcudding in the pipeline 2 image...]}
-
-\ben{[Ben says: new paragraph]} To utilize simulations of epidemics in later time-invariant analysis we also provide a function to convert temporally defined epidemics to filamental representations.
-Specifically, we provide the function `filament_compression()` to convert simulation(s) to filaments as expressed by presenting the epidemic as a ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space\footnote{\ben{[Ben: this should be cleaned up.
-I also 'highlighted' why this makes epicompare a good entry point]}}
-
-\shannon{Shannon's attempt at above paragraph is below..
-I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.}
-[\pkg{EpiCompare} also provides a tool to convert time-dependent epidemic simulations into their time-invariant filamental representations.
-Specifically, `filament_compression()` converts simulations to filaments (see Section \ref{sec:time-invariant})] so practitioners can view model simulations and results through a time-invariant lens.
-These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis is already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.]
+\shannon{[Shannon says: Here Ben talks about filament compression and tidy\_dist mats, etc.]} \ben{[Ben says: not currently thinking about including tidy\_dist stuff - will need to think about this as it is including in the pipeline 2 image...]} \footnote{\ben{[Ben says: are these comments still relevant?]}}
+
+To utilize simulations of epidemics in later time-invariant analysis we also provide a function to convert temporally defined epidemics to filamental representations.
+Specifically, we provide the function `filament_compression()` to convert simulation(s) to filaments as expressed by presenting the epidemic as a ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space.
+
+\shannon{[Shannon says: Shannon's attempt at above paragraph is below..
+I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.]}
+
+\pkg{EpiCompare} also provides a tool to convert time-dependent epidemic simulations into their time-invariant filamental representations.
+Specifically, `filament_compression()` converts simulations to filaments (see Section \ref{sec:time-invariant}) \ben{\sout{so practitioners can view model simulations and results through a time-invariant lens.}which allows practitioner to later apply time-invariant tools to these compressed epidemiological objects.}\footnote{\ben{[Ben says: I worry about this statement - it is very broad / seems to claim a lot but not very clearly...]}}  These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis is already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.
 
 **Comparisons and Assessment**
+
+\ben{[Ben says: I didn't look at this section - it looks like your comments are more "let's" talk about this and not suggestions / feedback.]}
 
@@
 For both of these geometric objects, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hassdorf_dist()`).
-\ben{[Ben says: should we discuss the details of `geom_prediction_bands` more?
-]}\shannon{I don't think so, best left to the vignette and data example}
+\footnote{\ben{[Ben says: should we discuss the details of geom\_prediction\_bands more?
+]}\shannon{[Shannon says: I don't think so, best left to the vignette and data example]}}
 
@@
 Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
-\ben{[Ben says: should we define these more in this paragraph?
+\footnote{\ben{[Ben says: should we define these more in this paragraph?
 Introducted in 2.2.
-as well and not defined.]}\shannon{[I'd throw a citation or two but again limit the detail.]}
+as well and not defined.]}\shannon{[Shannon says: I'd throw a citation or two but again limit the detail.]}}
 

### 2021-04-15 jss/EpiCompare-paper/EpiCompare-paper.tex
ben responses to comments on section 3 - minus last section which we will discuss in person (first round of edits after completition of section 3)

@@
 multiple distance-based assessment tools to examine the overall
-``extremeness'\,' of points, including pseudo-density estimators and
+``extremeness'' of points, including pseudo-density estimators and
 depth/local depth functions
@@
 ``verb'' naming schemes \citep{Wickham2019}. Although users can
-\sout{typically} incorporate \pkg{EpiCompare} into any step in the data
-analysis \sout{process}\textcolor{orange}{pipeline}, there are two
-primary points of entry. The first point of entry is the very beginning
-with pre-processing and visualizing raw data, and the second point of
-entry is after modeling and simulation. Figure \ref{fig:pipeline2}
-captures these different paths, and we
+incorporate \pkg{EpiCompare} into any step in the data analysis
+pipeline, there are two primary points of entry. The first point of
+entry is the very beginning with pre-processing and visualizing raw
+data, and the second point of entry is after modeling and simulation.
+Figure \ref{fig:pipeline2} captures these different paths, and we
 highlight\footnote{\textcolor{violet}{[Ben says: we need to make sure we actually do highlight ]}}
@@
 The first step of most data analysis is ``cleaning'' the data
-\textcolor{orange}{to a format that is friendly for both computers and programmers}
+\textcolor{orange}{to a format that is friendly for both computers and programmers}\footnote{\textcolor{violet}{[Ben says: I'm unsure this is needed and naturally adds more text.
+Willing to accept...]}}
 so it can be explored.
 \sout{There are multiple ways to collect epidemiological data.}
-\textcolor{orange}{Epidemiological data are collected in many different formats.}
+\textcolor{violet}{In epidemology, there are multiple different formats the data can arrive in.}
+\footnote{\textcolor{orange}{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \textcolor{violet}{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}}
 Sometimes individual records are collected, with times of different
@@
 information. Other data collections focus on aggregate counts of
-individuals in each epidemic state. In fact,
-\sout{usually}\textcolor{orange}{many times} only the number of new
-infections at each time step (e.g.~weekly case counts) is observed.
-Compartment totals (amounts of individuals in each state) are then
-imputed from those case counts\sout{ along with}\textcolor{orange}{and}
+individuals in each epidemic state. In fact, many times only the number
+of new infections at each time step (e.g.~weekly case counts) is
+observed. Compartment totals (amounts of individuals in each state) are
+then imputed from those case
+counts\sout{ along with}\textcolor{orange}{and}\footnote{\textcolor{violet}{[Ben says: Declined.
+I'm unclear why we should.
+Is the sentence unclear?
+Maybe it's the way it's a compound sentence?
+- If so a rewrite would be better?]}}
 other information about the disease and the population of interest. In
 \pkg{EpiCompare}, we focus on understanding the overall impact of an
-outbreak at the aggregate/population level, which allows for
-\textcolor{orange}{streamlined} examination of overall trends of an
-epidemic.\sout{ more easily.}
-
-\sout{In order t},To help the practitioner examine epidemics from an
-aggregate/ population lens, we provide a function called
-\texttt{agents\_to\_aggregate()}. This function transforms data about
-individual/agents' initial entry into each state (e.g.~start of
-infection, start of recovery, etc.) to an aggregate view of how many
-individuals were in a state at a given time. There are often situations
-where grouping agents into subpopulations
-(e.g.~sub\sout{populations}\textcolor{orange}{ groups} defined by age or
-sex) can highlight different aggregate level trends. For example,
-research by \citet{rvachev1985,anderson1992,worby2015} develop
-state-based models that account for
-\textcolor{orange}{differing disease dynamics in different}
-subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation
-analysis by combining \textcolor{orange}{the function}
+outbreak at the aggregate/population level, which allows for streamlined
+examination of overall trends of an epidemic.
+
+To help the practitioner examine epidemics from an aggregate/ population
+lens, we provide a function called \texttt{agents\_to\_aggregate()}.
+This function transforms data about individual/agents' initial entry
+into each state (e.g.~start of infection, start of recovery, etc.) to an
+aggregate view of how many individuals were in a state at a given time.
+There are often situations where
+\textcolor{violet}{\sout{grouping} aggregating}\footnote{\textcolor{violet}{[Ben says: reason - connect to languare from sentence before.]}}
+agents into subpopulations
+(e.g.~sub\sout{populations}\textcolor{orange}{ groups}\footnote{\textcolor{violet}{[Ben says: declined. I'm unclear why we'd do this.]}}
+defined by age or sex) can highlight different aggregate level trends.
+For example, research by \citet{rvachev1985,anderson1992,worby2015}
+develop state-based models that account for differing disease dynamics
+in different subpopulations. In \pkg{EpiCompare}, we facilitate
+subpopulation analysis by combining the function
 \pkg{dplyr}::\texttt{group\_by()} and \texttt{agent\_to\_aggregate()} to
-provide aggregation \sout{at a}\textcolor{orange}{by} group level.
+provide aggregation
+\sout{at a}\textcolor{orange}{by}\footnote{\textcolor{violet}{[Ben says: declined. unclear about this change. Also would it not be "by groups"? Orignial seems to flow more.]}}
+group level.
 
 The \texttt{agents\_to\_aggregate()} function is flexible and can deal
-with a wide range of information about each individual. \sout{It can,}
-Theoretically, account for infinitely many states. This functionality
-allows the practitioner to aggregate information relative to
-\sout{the standard}\textcolor{orange}{common} states
+with a wide range of information about each individual.
+\textcolor{violet}{\sout{It can,} In fact, this function can} account
+for infinitely many states. This functionality allows the practitioner
+to aggregate information relative to common states
 (e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as
-\sout{add}\textcolor{orange}{more complex} states (e.g.~``Exposed'',
-``iMmune'', ``Hospitalized'')
-(CITE\footnote{\textcolor{violet}{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}\textcolor{orange}{Shannon says: I changed 'standard' to 'common' so maybe that's a less loaded word.
-Otherwise I think I'd just cite anderson and may 1992}}).
-Additionally, \texttt{agents\_to\_aggregate()} \sout{also} permits
-indicators for death/exit and birth/entry dates. Overall, this function
-is a powerful tool for pre-processing data,
-\textcolor{orange}{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\textcolor{orange}{Shannon says: I added this in theme with 'highlighting a point of entry'}}
+more complex states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'').
+Additionally, \texttt{agents\_to\_aggregate()} permits indicators for
+death/exit and birth/entry dates. Overall, this function is a powerful
+tool for pre-processing data,
+\textcolor{orange}{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\textcolor{orange}{[Shannon says: I added this in theme with 'highlighting a point of entry']} \textcolor{violet}{[Ben says: I'm unclear about this comment but I'm find with the addition...]}}
 
@@
 
-\sout{With raw data, }``Getting to know'' \sout{our} data currently
-means figuring out \sout{good}\textcolor{orange}{useful} combinations of
-visualizations, numerical summaries and
-\sout{subsets}\textcolor{orange}{groupings}. An expert coder has many
-ways to successfully explore the data in an aggregate lens using
-\texttt{agents\_to\_aggregate()}.
-\textcolor{orange}{For less experienced coders,} \pkg{EpiCompare} also
-includes tools to rapidly explore data that has three epidemic states.
-Building on the tools in \pkg{ggplot2} and \pkg{ggtern},
-\sout{our}\pkg{EpiCompare}`s \texttt{geom\_aggregate()} provides a rapid
-way to explore different subpopulations' experiences of an epidemic
-\citep{Wickham2016, Hamilton2018}. \textbackslash sout\{By combining the
-ideas behind agents\_to\_aggregate for three-state models to examine
-subpopulation trajectories in a 3d simplex space\}
-\citep{Wickham2016, Hamilton2018}\footnote{\textcolor{orange}{I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.}}.
+\textcolor{violet}{\sout{With raw data, "Getting to know" our} In the early stages of a project, getting to know the}\footnote{\textcolor{orange}{[Shannon suggested: "Getting to know" our]} \textcolor{violet}{[Ben says: even though I was the original one with the quotes - one should always stray away from it.
+I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}}
+data currently means figuring out useful combinations of visualizations
+\textcolor{violet}{\sout{,}and} numerical summaries and
+\sout{\sout{subsets}\textcolor{orange}{groupings}}\textcolor{violet}{exploring different groupings of the data}\footnote{\textcolor{violet}{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}}.
+\textcolor{violet}{\sout{An expert coder has many ways to successfully explore the data in an aggregate lens using }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{.
+For less experienced coders, \pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.} An expert coder can start from }
+\texttt{agents\_to\_aggregate()}
+\textcolor{violet}{to successfully accomplish EDA in many ways, but \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).}\footnote{\textcolor{violet}{[Ben says: I'm reverting to the old text - it's much clearer.
+- happy to have a a discussion on it.
+The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}}
+Building on \textcolor{violet}{\sout{the tools in}} \pkg{ggplot2} and
+\pkg{ggtern} \textcolor{violet}{packages}, \pkg{EpiCompare}`s
+\texttt{geom\_aggregate()} provides a rapid way to explore different
+subpopulations' experiences of an epidemic
+\citep{Wickham2016, Hamilton2018}.
+\textcolor{violet}{The function }\texttt{geom\_aggregate()}
+\textcolor{violet}{provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.}
+\footnote{\textcolor{violet}{[Original text: "By combiing the ideas behind agent2aggregate for three-state models to examine subpopulation trajcetories in 2d simplex space."]} \textcolor{orange}{[Shannons says: I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.]}\textcolor{violet}{[Ben says: I'm unclear why this warrants just deleting it.
+I've provided a newly written sentence...]}}
 Visualization tools for three-state models were developed because (1)
 SIR models are some of the most common and basic epidemic state-based
-models and (2) our \textcolor{orange}{3d} simplex representation of
-these epidemics emphasizes a ``time-invarance'' representation of the
-data (for a refresher see Section \ref{sec:time-invariant}).
+models and (2) our three-dimensional simplex representation of these
+epidemics emphasizes a ``time-invarance'' representation of the data
+(for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-. The special classes often contain a plethora of information from
-residuals, model diagnostics, input parameters, and more. While
-incredibly useful, these special classes can be difficult for novice
-coders to navigate.{]}
-
-{[}\textcolor{violet}{[Ben says: replaced with above paragraph]}
-Post-processing of modeling and simulation consists of making summary
-statistics, plots, tables, and other ways to disseminate information to
-the public. For example, comma separated value files (\texttt{.csv}) are
-a standard way to share information within tables. However, model output
-is often far more complicated than what a traditional \texttt{.csv}
-would allow. \textcolor{orange}{Do we need three sentences about csvs?}
-As a result, a number of epidemic modeling packages return a special
-class, specific to their modeling. The special classes often contain a
-plethora of information from residuals, model diagnostics, input
-parameters, and more. While incredibly useful, these special classes can
-be difficult for novice coders to navigate. {]}
-
-To this end, \pkg{EpiCompare} \textcolor{violet}{provides\sout{adapted}}
-a series of \texttt{fortify}-style
-\textcolor{violet}{methods\sout{functions}}\footnote{\textcolor{violet}{[Ben says: Not tied to this change.]}},
-called \texttt{fortify\_aggregate()} which transform output from
-\textcolor{orange}{infectious disease modeling and simulation} packages
-like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which
-contain information about the total number of individuals in each state
-at a given time, for a given simulation. These fortify functions have
-output that is consistent with that of \texttt{agents\_to\_aggregate()}.
-
-\textcolor{violet}{[Ben says: new paragraph]} To utilize simulations of
-epidemics in later time-invariant analysis we also provide a function to
-convert temporally defined epidemics to filamental representations.
-Specifically, we provide the function \texttt{filament\_compression()}
-to convert simulation(s) to filaments as expressed by presenting the
-epidemic as a ordering of some common fixed number of points so that
-they are equally spaced along the original path in the proportional
-state
-space\footnote{\textcolor{violet}{[Ben: this should be cleaned up.  I also 'highlighted' why this makes epicompare a good entry point]}}
-
-\textcolor{orange}{Shannon's attempt at above paragraph is below..
-I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.}
-
-{[}\pkg{EpiCompare} also provides a tool to convert time-dependent
-epidemic simulations into their time-invariant filamental
-representations. Specifically, \texttt{filament\_compression()} converts
-simulations to filaments (see Section \ref{sec:time-invariant}){]} so
-practitioners can view model simulations and results through a
-time-invariant lens. These tools were developed to provide another
-natural entry point into the \pkg{EpiCompare} data analysis pipeline for
-situations where modeling and analysis is already completed and
-practitioners are looking for simple and transparent tools to help
-understand and disseminate results.{]}
+After getting a good sense of what a past or current epidemic looks like
+with EDA, the next step is often model fitting and/or simulations.
+\sout{In this step and the next step (post-processing), we discuss how to \textcolor{orange}{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\textcolor{violet}{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\textcolor{violet}{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}}
+While \pkg{EpiCompare} does not focus on fitting model(s) to data, we do
+provide some flexible functions for simulation of basic discrete-time
+epidemic-state models.
+\textcolor{violet}{These functions simulation individual-level information based on practitioner estimated transition rates between states and}
+can be combined with \texttt{agents\_to\_aggregate()} to view these
+simulations through an aggregate lens. The function
+\texttt{simulate\_SIR\_agents()} simulates a basic SIR epidemic with
+user inputs for the number of simulations, the initial number in each
+state, the infection and recovery parameters \((\beta, \gamma\)), and
+the total number of discrete time steps. This function allows for easy
+access to SIR model analysis and comparison. Beyond SIR models, the
+function \texttt{simulate\_agents()} takes as input a user-specified
+state-transition matrix and other epidemic parameters to allow the user
+to create simulations for an outbreak with \textit{any} number of states
+and any number of transitions among them. This flexibility in states can
+be used to also reflect group-based dynamics. In turn, this allows for
+users to explore the space of models in an intuitive way without getting
+bogged down by too much mathematical detail. For consistency, we have
+made output from \texttt{simulate\_agents()} and
+\texttt{simulate\_SIR\_agents()} compatible with
+\texttt{agents\_to\_aggregate()} so aggregate information may easily be
+accessed.
+
+\textbf{Post-processing}
+
+\textcolor{violet}{[Ben says: I think we should remind the reader that we care more about simulations, in order to compare fitted models between themselves and the true epidemics.]}
+
+\textcolor{violet}{[Ben says: this replaces the first paragraph below]}
+
+If the practitioner wishes to compare models-to-observations or even
+models-to-models, they needs to post-process their models and
+simulations to disseminate the results in an easily digestable format.
+In general, post-processing of modeling and simulation consists of
+making summary statistics, plots, and tables.
+\textcolor{violet}{Model \sout{The}} summaries can be very complex, and
+as a result, a number of epidemic modeling \textbackslash proglang\{R\}
+packages return \textcolor{violet}{\sout{a}} special class
+\textcolor{violet}{objects}. The special
+class\textcolor{violet}{\sout{es} objects} often contain a plethora of
+information from residuals, model diagnostics, input parameters, and
+more. While incredibly useful, these special classes can be difficult
+for novice coders to \textcolor{violet}{\sout{navigate} handle}.
+
+To this end, \pkg{EpiCompare} provides a series of
+\texttt{fortify}-style methods, called \texttt{fortify\_aggregate()}
+which transform output from infectious disease modeling and simulation
+packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
+which contain information about the total number of individuals in each
+state at a given time, for a given simulation. These fortify functions
+have output that is consistent with that of
+\texttt{agents\_to\_aggregate()}.
+
+\textcolor{orange}{[Shannon says: Here Ben talks about filament compression and tidy\_dist mats, etc.]}
+\textcolor{violet}{[Ben says: not currently thinking about including tidy\_dist stuff - will need to think about this as it is including in the pipeline 2 image...]}
+\footnote{\textcolor{violet}{[Ben says: are these comments still relevant?]}}
+
+To utilize simulations of epidemics in later time-invariant analysis we
+also provide a function to convert temporally defined epidemics to
+filamental representations. Specifically, we provide the function
+\texttt{filament\_compression()} to convert simulation(s) to filaments
+as expressed by presenting the epidemic as a ordering of some common
+fixed number of points so that they are equally spaced along the
+original path in the proportional state space.
+
+\textcolor{orange}{[Shannon says: Shannon's attempt at above paragraph is below..
+I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.]}
+
+\pkg{EpiCompare} also provides a tool to convert time-dependent epidemic
+simulations into their time-invariant filamental representations.
+Specifically, \texttt{filament\_compression()} converts simulations to
+filaments (see Section \ref{sec:time-invariant})
+\textcolor{violet}{\sout{so practitioners can view model simulations and results through a time-invariant lens.}which allows practitioner to later apply time-invariant tools to these compressed epidemiological objects.}\footnote{\textcolor{violet}{[Ben says: I worry about this statement - it is very broad / seems to claim a lot but not very clearly...]}}
+These tools were developed to provide another natural entry point into
+the \pkg{EpiCompare} data analysis pipeline for situations where
+modeling and analysis is already completed and practitioners are looking
+for simple and transparent tools to help understand and disseminate
+results.
 
 \textbf{Comparisons and Assessment}
+
+\textcolor{violet}{[Ben says: I didn't look at this section - it looks like your comments are more "let's" talk about this and not suggestions / feedback.]}
 
@@
 different model (\texttt{hassdorf\_dist()}).
-\textcolor{orange}{I don't think so, best left to the vignette and data example}
+\footnote{\textcolor{violet}{[Ben says: should we discuss the details of geom\_prediction\_bands more?
+]}\textcolor{orange}{[Shannon says: I don't think so, best left to the vignette and data example]}}
 
@@
 suggest how geometrically central an epidemic is to simulations.
-\textcolor{violet}{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}\textcolor{orange}{[I'd throw a citation or two but again limit the detail.]}
+\footnote{\textcolor{violet}{[Ben says: should we define these more in this paragraph?
+Introducted in 2.2.
+as well and not defined.]}\textcolor{orange}{[Shannon says: I'd throw a citation or two but again limit the detail.]}}
 

### 2021-04-27 jss/EpiCompare-paper/EpiCompare-paper.tex
updated

@@
 multiple distance-based assessment tools to examine the overall
-``extremeness'' of points, including pseudo-density estimators and
+``extremeness'\,' of points, including pseudo-density estimators and
 depth/local depth functions

### 2021-04-27 jss/EpiCompare-paper/EpiCompare-paper.Rmd
updated again

@@
 \newcommand{\shannon}[1]{\textcolor{orange}{#1}}
+\newcommand{\shan}[1]{\textcolor{brown}{#1}}
 \newcommand{\ben}[1]{\textcolor{violet}{#1}}
@@
 Time-invariant analysis makes different epidemics easier to compare by removing many time dependent aspects of an epidemic.
-Instead, \shannon{time-invariant analysis} focuses \shannon{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.} \ben{[Ben wants to try this sentence again.]}
+\sout{Instead,} \shannon{Time-invariant analysis} focuses \shannon{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.} \ben{[Ben wants to try this sentence again.]}
 
@@
 The first step of most data analysis is "cleaning" the data \shannon{to a format that is friendly for both computers and programmers}\footnote{\ben{[Ben says: I'm unsure this is needed and naturally adds more text.
-Willing to accept...]}} so it can be explored.
-\sout{There are multiple ways to collect epidemiological data.} \ben{In epidemology, there are multiple different formats the data can arrive in.} \footnote{\shannon{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \ben{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+Willing to accept...]}\shan{When in doubt, remove.}} so it can be explored.
+\sout{There are multiple ways to collect epidemiological data.} \ben{In epidemology, there are multiple different formats the data can arrive in.} \footnote{\shannon{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \ben{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}\shan{What about adding a linking sentence "Before data can be explored, it must be colleected.
+Epidemiological data are collected in many different formats."}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
 Other data collections focus on aggregate counts of individuals in each epidemic state.
@@
 Maybe it's the way it's a compound sentence?
-- If so a rewrite would be better?]}} other information about the disease and the population of interest.
+- If so a rewrite would be better?]}\shan{Upon reflection, I think it's fine.}} other information about the disease and the population of interest.
 In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.
@@
 start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
-There are often situations where \ben{\sout{grouping} aggregating}\footnote{\ben{[Ben says: reason - connect to languare from sentence before.]}} agents into subpopulations (e.g.
+There are often situations where \ben{\sout{grouping} aggregating}\footnote{\ben{[Ben says: reason - connect to languare from sentence before.]}\shan{fine}} agents into subpopulations (e.g.
 sub\sout{populations}\shannon{ groups}\footnote{\ben{[Ben says: declined.
-I'm unclear why we'd do this.]}} defined by age or sex) can highlight different aggregate level trends.
+I'm unclear why we'd do this.]}\shan{fine}} defined by age or sex) can highlight different aggregate level trends.
 For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations.
@@
 Also would it not be "by groups"?
-Orignial seems to flow more.]}} group level.
+Orignial seems to flow more.]}\shan{fine}} group level.
 
@@
 Additionally, `agents_to_aggregate()` permits indicators for death/exit and birth/entry dates.
-Overall, this function is a powerful tool for pre-processing data, \shannon{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\shannon{[Shannon says: I added this in theme with 'highlighting a point of entry']} \ben{[Ben says: I'm unclear about this comment but I'm find with the addition...]}}
+Overall, this function is a powerful tool for pre-processing data, \shannon{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\shannon{[Shannon says: I added this in theme with 'highlighting a point of entry']} \ben{[Ben says: I'm unclear about this comment but I'm find with the addition...]}\shan{We say in the intro about highlighting two points of entry into EpiCompare.}}
 
@@
 \ben{\sout{With raw data, "Getting to know" our} In the early stages of a project, getting to know the}\footnote{\shannon{[Shannon suggested: "Getting to know" our]} \ben{[Ben says: even though I was the original one with the quotes - one should always stray away from it.
-I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}} data currently means figuring out useful combinations of visualizations \ben{\sout{,}and} numerical summaries and \sout{\sout{subsets}\shannon{groupings}}\ben{exploring different groupings of the data}\footnote{\ben{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}}.
+I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}\shan{I don't like getting to know without quotes -- it's too colloquial.
+What about 'The practicioner famliarizes herself with data through...' }} data currently means figuring out useful combinations of visualizations \ben{\sout{,}and} numerical summaries and \sout{\sout{subsets}\shannon{groupings}}\ben{exploring different groupings of the data}\footnote{\ben{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}\shan{fine}}.
 \ben{\sout{An expert coder has many ways to successfully explore the data in an aggregate lens using }}`agents_to_aggregate()`\ben{\sout{.
@@
 - happy to have a a discussion on it.
-The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}} Building on \ben{\sout{the tools in}} \pkg{ggplot2} and \pkg{ggtern} \ben{packages}, \pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of an epidemic \citep{Wickham2016, Hamilton2018}.
+The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}\shan{fine}} Building on \ben{\sout{the tools in}} \pkg{ggplot2} and \pkg{ggtern} \ben{packages}, \pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of an epidemic \citep{Wickham2016, Hamilton2018}.
 \ben{The function }`geom_aggregate()` \ben{provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.}  \footnote{\ben{[Original text: "By combiing the ideas behind agent2aggregate for three-state models to examine subpopulation trajcetories in 2d simplex space."]} \shannon{[Shannons says: I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.]}\ben{[Ben says: I'm unclear why this warrants just deleting it.
@@
 After getting a good sense of what a past or current epidemic looks like with EDA, the next step is often model fitting and/or simulations.
-\sout{In this step and the next step (post-processing), we discuss how to \shannon{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\ben{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\ben{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}} While \pkg{EpiCompare} does not focus on fitting model(s) to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
+\sout{In this step and the next step (post-processing), we discuss how to \shannon{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\ben{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\ben{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}\shan{fine}} While \pkg{EpiCompare} does not focus on fitting model(s) to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
 \ben{These functions simulation individual-level information based on practitioner estimated transition rates between states and} can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
@@
 
-\shannon{[Shannon says: Here Ben talks about filament compression and tidy\_dist mats, etc.]} \ben{[Ben says: not currently thinking about including tidy\_dist stuff - will need to think about this as it is including in the pipeline 2 image...]} \footnote{\ben{[Ben says: are these comments still relevant?]}}
+\shannon{[Shannon says: Here Ben talks about filament compression and tidy\_dist mats, etc.]} \ben{[Ben says: not currently thinking about including tidy\_dist stuff - will need to think about this as it is including in the pipeline 2 image...]} \footnote{\ben{[Ben says: are these comments still relevant?]}\shan{These comments are resolved}}
 
@@
 Additionally, we offer ways to determine if model outputs are compatible with one another, that is how extreme one output is to another.
-\shannon{Ben says something about distance}
 

### 2021-04-27 jss/EpiCompare-paper/EpiCompare-paper.tex
updated again

@@
 makes different epidemics easier to compare by removing many time
-dependent aspects of an epidemic. Instead,
-\textcolor{orange}{time-invariant analysis} focuses
+dependent aspects of an epidemic. \sout{Instead,}
+\textcolor{orange}{Time-invariant analysis} focuses
 \textcolor{orange}{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.}
@@
 \textcolor{orange}{to a format that is friendly for both computers and programmers}\footnote{\textcolor{violet}{[Ben says: I'm unsure this is needed and naturally adds more text.
-Willing to accept...]}}
+Willing to accept...]}\textcolor{brown}{When in doubt, remove.}}
 so it can be explored.
@@
 \textcolor{violet}{In epidemology, there are multiple different formats the data can arrive in.}
-\footnote{\textcolor{orange}{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \textcolor{violet}{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}}
+\footnote{\textcolor{orange}{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \textcolor{violet}{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}\textcolor{brown}{What about adding a linking sentence "Before data can be explored, it must be colleected.
+Epidemiological data are collected in many different formats."}}
 Sometimes individual records are collected, with times of different
@@
 Maybe it's the way it's a compound sentence?
-- If so a rewrite would be better?]}}
+- If so a rewrite would be better?]}\textcolor{brown}{Upon reflection, I think it's fine.}}
 other information about the disease and the population of interest. In
@@
 There are often situations where
-\textcolor{violet}{\sout{grouping} aggregating}\footnote{\textcolor{violet}{[Ben says: reason - connect to languare from sentence before.]}}
+\textcolor{violet}{\sout{grouping} aggregating}\footnote{\textcolor{violet}{[Ben says: reason - connect to languare from sentence before.]}\textcolor{brown}{fine}}
 agents into subpopulations
-(e.g.~sub\sout{populations}\textcolor{orange}{ groups}\footnote{\textcolor{violet}{[Ben says: declined. I'm unclear why we'd do this.]}}
+(e.g.~sub\sout{populations}\textcolor{orange}{ groups}\footnote{\textcolor{violet}{[Ben says: declined. I'm unclear why we'd do this.]}\textcolor{brown}{fine}}
 defined by age or sex) can highlight different aggregate level trends.
@@
 provide aggregation
-\sout{at a}\textcolor{orange}{by}\footnote{\textcolor{violet}{[Ben says: declined. unclear about this change. Also would it not be "by groups"? Orignial seems to flow more.]}}
+\sout{at a}\textcolor{orange}{by}\footnote{\textcolor{violet}{[Ben says: declined. unclear about this change. Also would it not be "by groups"? Orignial seems to flow more.]}\textcolor{brown}{fine}}
 group level.
@@
 tool for pre-processing data,
-\textcolor{orange}{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\textcolor{orange}{[Shannon says: I added this in theme with 'highlighting a point of entry']} \textcolor{violet}{[Ben says: I'm unclear about this comment but I'm find with the addition...]}}
+\textcolor{orange}{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\textcolor{orange}{[Shannon says: I added this in theme with 'highlighting a point of entry']} \textcolor{violet}{[Ben says: I'm unclear about this comment but I'm find with the addition...]}\textcolor{brown}{We say in the intro about highlighting two points of entry into EpiCompare.}}
 
@@
 \textcolor{violet}{\sout{With raw data, "Getting to know" our} In the early stages of a project, getting to know the}\footnote{\textcolor{orange}{[Shannon suggested: "Getting to know" our]} \textcolor{violet}{[Ben says: even though I was the original one with the quotes - one should always stray away from it.
-I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}}
+I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}\textcolor{brown}{I don't like getting to know without quotes -- it's too colloquial.
+What about 'The practicioner famliarizes herself with data through...' }}
 data currently means figuring out useful combinations of visualizations
 \textcolor{violet}{\sout{,}and} numerical summaries and
-\sout{\sout{subsets}\textcolor{orange}{groupings}}\textcolor{violet}{exploring different groupings of the data}\footnote{\textcolor{violet}{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}}.
+\sout{\sout{subsets}\textcolor{orange}{groupings}}\textcolor{violet}{exploring different groupings of the data}\footnote{\textcolor{violet}{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}\textcolor{brown}{fine}}.
 \textcolor{violet}{\sout{An expert coder has many ways to successfully explore the data in an aggregate lens using }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{.
@@
 - happy to have a a discussion on it.
-The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}}
+The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}\textcolor{brown}{fine}}
 Building on \textcolor{violet}{\sout{the tools in}} \pkg{ggplot2} and
@@
 with EDA, the next step is often model fitting and/or simulations.
-\sout{In this step and the next step (post-processing), we discuss how to \textcolor{orange}{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\textcolor{violet}{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\textcolor{violet}{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}}
+\sout{In this step and the next step (post-processing), we discuss how to \textcolor{orange}{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\textcolor{violet}{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\textcolor{violet}{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}\textcolor{brown}{fine}}
 While \pkg{EpiCompare} does not focus on fitting model(s) to data, we do
@@
 \textcolor{violet}{[Ben says: not currently thinking about including tidy\_dist stuff - will need to think about this as it is including in the pipeline 2 image...]}
-\footnote{\textcolor{violet}{[Ben says: are these comments still relevant?]}}
+\footnote{\textcolor{violet}{[Ben says: are these comments still relevant?]}\textcolor{brown}{These comments are resolved}}
 
@@
 outputs are compatible with one another, that is how extreme one output
-is to another. \textcolor{orange}{Ben says something about distance}
+is to another.
 

### 2021-04-27 jss/EpiCompare-paper/EpiCompare-paper.Rmd
edit to Section 3's final part - clarifications for first paragraph

@@
 
-\ben{[Ben says: I didn't look at this section - it looks like your comments are more "let's" talk about this and not suggestions / feedback.]}
-
 <!-- \ben{[Brainstorming: (intro) discuss model vs model comparison tools, (a) highlight that standard comparisons can be done if models share metrics or parameters (and similarly some comparisons can be done to the real epidemics).
@@
 
-\shannon{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}  As introduced in Section \ref{sec:beyond-r0-sir} there's a \sout{lot of}\shannon{much} potential for time-invariant tools to help compare and assess epidemics and models/simulations.
+\ben{Finally,} [In \pkg{EpiCompare} we provide a set of comparison and assessment tools for models \ben{(and model's simulations)} that extend beyond the standard performance metrics (e.g.
+mean squared error or AIC).] \ben{Aligned with the discussion in Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
+We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of the epidemic has occurred (e.g.
+one flue season).}\footnote{\ben{[Ben says: Shannon - is this a clear enough use of season / cycle?]}}
+
+\ben{One tool we provide to assess models is through the creation of geometric prediction regions, useful when we treat epidemics like filaments.
+If we have a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epdiemice in the state space.} [For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.] \ben{In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done visually.} [In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
+The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any \ben{dimensional} state-based model.
+For both of these geometric objects, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hassdorf_dist()`).]
+
+\ben{[Ben says: this paragraph is replaced by the two above, see "[" segments - they refer to items below.]}\footnote{\ben{Ben says: Here are some thoughts for the replacment work: (1) The introduction is cleaned up the confusion around "cycle" - hopefully?
+(2) Shannon's proposed intro wasn't used due to not wanting to claim that this step is always the "last step" of the pipeline.
+(3) more focus was placed on why we are providing this tools.}} \shannon{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}  As introduced in Section \ref{sec:beyond-r0-sir} there's a \sout{lot of}\shannon{much} potential for time-invariant tools to help compare and assess epidemics and models/simulations.
 In \pkg{EpiCompare} we provide a set of comparison and assessment tools for models that extend beyond the standard performance metrics (e.g.
@@
 For both of these geometric objects, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hassdorf_dist()`).
-\footnote{\ben{[Ben says: should we discuss the details of geom\_prediction\_bands more?
-]}\shannon{[Shannon says: I don't think so, best left to the vignette and data example]}}
-
-We also provide functions to calculate the "extremeness" of a true epidemic [\shannon{trajectory?}] compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
+
+\ben{[Ben says: this paragraph is still kept.]} We also provide functions to calculate the "extremeness" of a true epidemic trajectory\footnote{\ben{Ben says: accepted.}} compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
 Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
 Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
-\footnote{\ben{[Ben says: should we define these more in this paragraph?
-Introducted in 2.2.
-as well and not defined.]}\shannon{[Shannon says: I'd throw a citation or two but again limit the detail.]}}
-
-\ben{[Ben says: Replaced with paragraphs above.]} Comparison and assessment of model fit or comparisons of one model to another model can be performed in a variety of ways including mean square error, AIC, plots, and more.
+
+<!--\ben{[Ben says: Replaced with paragraphs above.]} Comparison and assessment of model fit or comparisons of one model to another model can be performed in a variety of ways including mean square error, AIC, plots, and more.
 Perhaps the most useful tool \pkg{EpiCompare} has to offer to the expert, for comparison and assessment of models, is in its post-processing tools which create a standard output.
@@
 Additionally, we offer ways to determine if model outputs are compatible with one another, that is how extreme one output is to another.
+-->
 

### 2021-04-27 jss/EpiCompare-paper/EpiCompare-paper.tex
edit to Section 3's final part - clarifications for first paragraph

@@
 
-\textcolor{violet}{[Ben says: I didn't look at this section - it looks like your comments are more "let's" talk about this and not suggestions / feedback.]}
-
+\textcolor{violet}{Finally,} {[}In \pkg{EpiCompare} we provide a set of
+comparison and assessment tools for models
+\textcolor{violet}{(and model's simulations)} that extend beyond the
+standard performance metrics (e.g.~mean squared error or AIC).{]}
+\textcolor{violet}{Aligned with the discussion in Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
+We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of the epidemic has occurred (e.g.
+one flue season).}\footnote{\textcolor{violet}{[Ben says: Shannon - is this a clear enough use of season / cycle?]}}
+
+\textcolor{violet}{One tool we provide to assess models is through the creation of geometric prediction regions, useful when we treat epidemics like filaments.
+If we have a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epdiemice in the state space.}
+{[}For three-state epidemic models, we provide the
+\texttt{ggplot}/\texttt{ggtern} extension
+\texttt{geom\_prediction\_band()} which creates a prediction region
+around the top \(1-\alpha\) proportion of the simulations.{]}
+\textcolor{violet}{In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done visually.}
+{[}In \pkg{EpiCompare} we also provide these prediction regions for
+epidemic models with with more than three states. The functions
+\texttt{create\_convex\_hull\_structure()} and
+\texttt{create\_delta\_ball\_structure()} create different geometric
+representations of prediction regions for any
+\textcolor{violet}{dimensional} state-based model. For both of these
+geometric objects, we provide functions to check if a path is contained
+(\texttt{contained()}) and the ability to assess the Hausdorff distance
+between prediction regions based on simulations from different model
+(\texttt{hassdorf\_dist()}).{]}
+
+\textcolor{violet}{[Ben says: this paragraph is replaced by the two above, see "[" segments - they refer to items below.]}\footnote{\textcolor{violet}{Ben says: Here are some thoughts for the replacment work: (1) The introduction is cleaned up the confusion around "cycle" - hopefully?
+(2) Shannon's proposed intro wasn't used due to not wanting to claim that this step is always the "last step" of the pipeline.
+(3) more focus was placed on why we are providing this tools.}}
 \textcolor{orange}{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}
@@
 different model (\texttt{hassdorf\_dist()}).
-\footnote{\textcolor{violet}{[Ben says: should we discuss the details of geom\_prediction\_bands more?
-]}\textcolor{orange}{[Shannon says: I don't think so, best left to the vignette and data example]}}
-
-We also provide functions to calculate the ``extremeness'' of a true
-epidemic {[}\textcolor{orange}{trajectory?}{]} compared to simulated
-epidemics via the equi-distance filamental trajectory representation as
-mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions
-like \texttt{distance\_pseudo\_density\_function()} can calculate a
+
+\textcolor{violet}{[Ben says: this paragraph is still kept.]} We also
+provide functions to calculate the ``extremeness'' of a true epidemic
+trajectory\footnote{\textcolor{violet}{Ben says: accepted.}} compared to
+simulated epidemics via the equi-distance filamental trajectory
+representation as mentioned in Section \ref{sec:beyond-r0-sir}.
+Specifically, functions like
+\texttt{distance\_pseudo\_density\_function()} can calculate a
 pseudo-density estimate of the true epidemic relative to simulated ones.
@@
 suggest how geometrically central an epidemic is to simulations.
-\footnote{\textcolor{violet}{[Ben says: should we define these more in this paragraph?
-Introducted in 2.2.
-as well and not defined.]}\textcolor{orange}{[Shannon says: I'd throw a citation or two but again limit the detail.]}}
-
-\textcolor{violet}{[Ben says: Replaced with paragraphs above.]}
-Comparison and assessment of model fit or comparisons of one model to
-another model can be performed in a variety of ways including mean
-square error, AIC, plots, and more. Perhaps the most useful tool
-\pkg{EpiCompare} has to offer to the expert, for comparison and
-assessment of models, is in its post-processing tools which create a
-standard output. It is then a matter of writing a script or function
-made for that standard output to assess the results from multiple models
-in the way the user
-desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}\textcolor{orange}{I agree with you now that this section is nice and filled in.
-Let's drop this paragraph.}}.
-
-\textcolor{violet}{[Ben says: Replaced with paragraphs above.]} However,
-for those who like more concrete tools, \pkg{EpiCompare} offers
-functions to compare prediction regions to one another including
-\texttt{geom\_prediction\_band()} (which plots the region), and
-\texttt{create\_\{convex\_hull,delta\_ball\}\_structure()}(which returns
-the \texttt{R} output for the given structure), and \texttt{contained()}
-(which allows the user to determine if one set of points is contained in
-a prediction band). Additionally, we offer ways to determine if model
-outputs are compatible with one another, that is how extreme one output
-is to another.
 

### 2021-05-10 jss/EpiCompare-paper/EpiCompare-paper.Rmd
pass 1 at section 3 edits

--- 2021-04-15 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
 \ref{fig:pipeline}.
-All front-facing functions are aimed to be as user-friendly as possible.
+All front-facing functions in \pkg{EpiCompare} are aimed to be as user-friendly as possible.
 We also focus on providing the user "tidyverse" style functions, that encourage piping objects from one function to the next and follow clear "verb" naming schemes \citep{Wickham2019}.
@@
 The first point of entry is the very beginning with pre-processing and visualizing raw data, and the second point of entry is after modeling and simulation.
-Figure \ref{fig:pipeline2} captures these different paths, and we highlight\footnote{\ben{[Ben says: we need to make sure we actually do highlight ]}} both approaches and how to leverage \pkg{EpiCompare} in the subsections below.
+Figure \ref{fig:pipeline2} captures these different paths, and we highlight how to leverage \pkg{EpiCompare} functionalities in the subsections below.
 
@@
 
-The first step of most data analysis is "cleaning" the data \shannon{to a format that is friendly for both computers and programmers}\footnote{\ben{[Ben says: I'm unsure this is needed and naturally adds more text.
-Willing to accept...]}} so it can be explored.
-\sout{There are multiple ways to collect epidemiological data.} \ben{In epidemology, there are multiple different formats the data can arrive in.} \footnote{\shannon{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \ben{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+The first step of most data analysis is "cleaning" the raw data so it can be explored.
+Before data can be explored, they must be collected.
+Raw data can appear in many different formats.
+Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
 Other data collections focus on aggregate counts of individuals in each epidemic state.
@@
 weekly case counts) is observed.
-Compartment totals (amounts of individuals in each state) are then imputed from those case counts\sout{ along with}\shannon{and}\footnote{\ben{[Ben says: Declined.
-I'm unclear why we should.
-Is the sentence unclear?
-Maybe it's the way it's a compound sentence?
-- If so a rewrite would be better?]}} other information about the disease and the population of interest.
+Compartment totals (amounts of individuals in each state) are then imputed from those case counts other information about the disease and the population of interest.
 In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.
 
-To help the practitioner examine epidemics from an aggregate/ population lens, we provide a function called `agents_to_aggregate()`.
+To help the practitioner examine epidemics from an aggregate/population lens, we provide a function called `agents_to_aggregate()`.
 This function transforms data about individual/agents' initial entry into each state (e.g.
 start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
-There are often situations where \ben{\sout{grouping} aggregating}\footnote{\ben{[Ben says: reason - connect to languare from sentence before.]}} agents into subpopulations (e.g.
-sub\sout{populations}\shannon{ groups}\footnote{\ben{[Ben says: declined.
-I'm unclear why we'd do this.]}} defined by age or sex) can highlight different aggregate level trends.
+There are often situations where aggregating agents into subpopulations (e.g.
+subpopulations defined by age or sex) can highlight different aggregate level trends.
 For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation \sout{at a}\shannon{by}\footnote{\ben{[Ben says: declined.
-unclear about this change.
-Also would it not be "by groups"?
-Orignial seems to flow more.]}} group level.
-
-The `agents_to_aggregate()` function is flexible and can deal with a wide range of information about each individual. 
-\ben{\sout{It can,} In fact, this function can}\footnote{\shannon{[Shannon suggests: "Theoretically, this agents_to_aggregate() can"]} \ben{[Ben says: I just removed the "theoretically" since it seems to suggest more than we want.
-Also I didn't see your original edit.
-both edits connect to the previous sentence better.]}} account for infinitely many states.
+In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation at a subpopulation level.
+
+\shannon{Besides aiding subpopulation analysis,} `agents_to_aggregate()` can deal with a wide range of information about each individual. 
+In fact, this function account for infinitely many states.
 This functionality allows the practitioner to aggregate information relative to common states (e.g.
@@
 Additionally, `agents_to_aggregate()` permits indicators for death/exit and birth/entry dates.
-Overall, this function is a powerful tool for pre-processing data, \shannon{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\shannon{[Shannon says: I added this in theme with 'highlighting a point of entry']} \ben{[Ben says: I'm unclear about this comment but I'm find with the addition...]}}
+Overall, this function is a powerful tool for pre-processing data.
 
@@
 
-\ben{\sout{With raw data, "Getting to know" our} In the early stages of a project, getting to know the}\footnote{\shannon{[Shannon suggested: "Getting to know" our]} \ben{[Ben says: even though I was the original one with the quotes - one should always stray away from it.
-I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}} data currently means figuring out useful combinations of visualizations \ben{\sout{,}and} numerical summaries and \sout{\sout{subsets}\shannon{groupings}}\ben{exploring different groupings of the data}\footnote{\ben{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}}.
-\ben{\sout{An expert coder has many ways to successfully explore the data in an aggregate lens using }}`agents_to_aggregate()`\ben{\sout{.
-For less experienced coders, \pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.} An expert coder can start from } `agents_to_aggregate()` \ben{to successfully accomplish EDA in many ways, but \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).}\footnote{\ben{[Ben says: I'm reverting to the old text - it's much clearer.
-- happy to have a a discussion on it.
-The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}} Building on \ben{\sout{the tools in}} \pkg{ggplot2} and \pkg{ggtern} \ben{packages}, \pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore different subpopulations' experiences of an epidemic \citep{Wickham2016, Hamilton2018}.
-\ben{The function }`geom_aggregate()` \ben{provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.}  \footnote{\ben{[Original text: "By combiing the ideas behind agent2aggregate for three-state models to examine subpopulation trajcetories in 2d simplex space."]} \shannon{[Shannons says: I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.]}\ben{[Ben says: I'm unclear why this warrants just deleting it.
-I've provided a newly written sentence...]}}  Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+ In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data \shannon{both at population and subpopulation level.}  An expert coder can start with `agents_to_aggregate()` to successfully accomplish EDA in many ways.
+\pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there three unique epidemiological states (like in the SIR model).
+Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
+The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.
+Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-\ben{[Ben says: think about this section and if it highlights that we can bring in outside models...]}
-
-After getting a good sense of what a past or current epidemic looks like with EDA, the next step is often model fitting and/or simulations.
-\sout{In this step and the next step (post-processing), we discuss how to \shannon{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\ben{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\ben{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}} While \pkg{EpiCompare} does not focus on fitting model(s) to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
-\ben{These functions simulation individual-level information based on practitioner estimated transition rates between states and} can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
+After getting a sense of what a past or current epidemic looks like with EDA, often the next step in the data analysis pipeline is model fitting and/or simulation.
+In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.
+While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
+These functions simulate individual-level information based on practitioner estimated transition rates between states and can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
 The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
@@
 
-\ben{[Ben says: I think we should remind the reader that we care more about simulations, in order to compare fitted models between themselves and the true epidemics.]}
-
-\ben{[Ben says: this replaces the first paragraph below]}
-If the practitioner wishes to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestable format.
+If practitioners wish to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestible format.
 In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.
-\ben{Model \sout{The}} summaries can be very complex, and as a result, a number of epidemic modeling \\proglang{R} packages return \ben{\sout{a}} special class \ben{objects}.
-The special class\ben{\sout{es} objects} often contain a plethora of information from residuals, model diagnostics, input parameters, and more.
-While incredibly useful, these special classes can be difficult for novice coders to \ben{\sout{navigate} handle}.
-
-<!-- [\ben{[Ben says: replaced with above paragraph]} Post-processing of modeling and simulation consists of making summary statistics, plots, tables, and other ways to disseminate information to the public.
-For example, comma separated value files (`.csv`) are a  standard way to share information within tables.
-However, model output is often far more complicated than what a traditional `.csv` would allow.
-\shannon{Do we need three sentences about csvs?} As a result, a number of epidemic modeling packages return a special class, specific to their modeling.
-The special classes often contain a plethora of information from residuals, model diagnostics, input parameters, and more.
-While incredibly useful, these special classes can be difficult for novice coders to navigate.
-] -->
+Model output summaries can be very complex, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.
+The special classes often contain a plethora of information about residuals, model diagnostics, input parameters, and more.
+While incredibly useful, these special classes can be difficult for novice coders to handle.
 
@@
 
-\shannon{[Shannon says: Here Ben talks about filament compression and tidy\_dist mats, etc.]} \ben{[Ben says: not currently thinking about including tidy\_dist stuff - will need to think about this as it is including in the pipeline 2 image...]} \footnote{\ben{[Ben says: are these comments still relevant?]}}
-
-To utilize simulations of epidemics in later time-invariant analysis we also provide a function to convert temporally defined epidemics to filamental representations.
-Specifically, we provide the function `filament_compression()` to convert simulation(s) to filaments as expressed by presenting the epidemic as a ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space.
-
-\shannon{[Shannon says: Shannon's attempt at above paragraph is below..
-I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.]}
-
-\pkg{EpiCompare} also provides a tool to convert time-dependent epidemic simulations into their time-invariant filamental representations.
-Specifically, `filament_compression()` converts simulations to filaments (see Section \ref{sec:time-invariant}) \ben{\sout{so practitioners can view model simulations and results through a time-invariant lens.}which allows practitioner to later apply time-invariant tools to these compressed epidemiological objects.}\footnote{\ben{[Ben says: I worry about this statement - it is very broad / seems to claim a lot but not very clearly...]}}  These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis is already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.
+To utilize simulations of epidemics in  time-invariant analysis, we also provide a function to convert temporally defined epidemics to filamental representations.
+Specifically, we provide the function `filament_compression()` to convert simulation(s) to filaments as expressed by presenting the epidemic as an ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space.
+These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis are already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.
 
@@
 
-\ben{[Ben says: I didn't look at this section - it looks like your comments are more "let's" talk about this and not suggestions / feedback.]}
-
-<!-- \ben{[Brainstorming: (intro) discuss model vs model comparison tools, (a) highlight that standard comparisons can be done if models share metrics or parameters (and similarly some comparisons can be done to the real epidemics).
-(b) Present time-invariance tools that allow us to compare more types of models (and can also give new tools to assess models beyond the standard ones).
-(1) introduce prediction regions (a) geom_prediction_bands,  (2) discuss different extremeness measures.]} -->
-
-\shannon{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}  As introduced in Section \ref{sec:beyond-r0-sir} there's a \sout{lot of}\shannon{much} potential for time-invariant tools to help compare and assess epidemics and models/simulations.
-In \pkg{EpiCompare} we provide a set of comparison and assessment tools for models that extend beyond the standard performance metrics (e.g.
-mean squared error or AIC) and focus on assessing the structural information the models capture.
-This approaches work well on models where \sout{online} one "cycle" of the epidemic has occurred (no recovered individuals have been susceptible again)\footnote{\ben{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}\shannon{Shannon says I don't think I understand what you are trying to say there.
-We should chat about it.}}.
-Epidemics are complex objects, and we provide tools to create prediction regions \shannon{with differing desired characteristics}\footnote{\shannon{Shannon says: I want to connect epidemics being complex to the fact that prediction is hard and not all prediction regions tell you the same thing}} from simulated epidemics.
-For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion [\shannon{good place for Ben paper cite?}] of the simulations (where the simulations treated as filaments).
-In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
-The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any state-based model.
-For both of these geometric objects, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hassdorf_dist()`).
-\footnote{\ben{[Ben says: should we discuss the details of geom\_prediction\_bands more?
-]}\shannon{[Shannon says: I don't think so, best left to the vignette and data example]}}
-
-We also provide functions to calculate the "extremeness" of a true epidemic [\shannon{trajectory?}] compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
+Finally, in \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
+mean squared error or AIC).
+Corresponding to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
+We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of the epidemic has occurred.
+
+One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
+If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space.
+For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
+In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done with the 'eye test.' In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
+The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model.
+For both of these geometric structures, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hausdorff_dist()`).
+
+We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
 Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
 Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
-\footnote{\ben{[Ben says: should we define these more in this paragraph?
-Introducted in 2.2.
-as well and not defined.]}\shannon{[Shannon says: I'd throw a citation or two but again limit the detail.]}}
-
-\ben{[Ben says: Replaced with paragraphs above.]} Comparison and assessment of model fit or comparisons of one model to another model can be performed in a variety of ways including mean square error, AIC, plots, and more.
-Perhaps the most useful tool \pkg{EpiCompare} has to offer to the expert, for comparison and assessment of models, is in its post-processing tools which create a standard output.
-It is then a matter of writing a script or function made for that standard output to assess the results from multiple models in the way the user desires\footnote{\ben{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}\shannon{I agree with you now that this section is nice and filled in.
-Let's drop this paragraph.}}.
-
-\ben{[Ben says: Replaced with paragraphs above.]} However, for those who like more concrete tools, \pkg{EpiCompare} offers functions to compare prediction regions to one another including `geom_prediction_band()` (which plots the region), and `create_{convex_hull,delta_ball}_structure()`(which returns the `R` output for the given structure), and `contained()` (which allows the user to determine if one set of points is contained in a prediction band).
-Additionally, we offer ways to determine if model outputs are compatible with one another, that is how extreme one output is to another.
-\shannon{Ben says something about distance}
 

### 2021-05-10 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
new version of section 2

--- 2021-04-27 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
 
-\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a \textit{time-invariant} way.
-Epidemics, despite by definition being a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales to understand the processes at play.
-<!-- Additionally, many tools designed to examine the quantity of the population in each epidemic state (for example: quantity of Susceptible vs Infectious vs Recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
-\shannon{I don't quite understand this.
-Is this because of underreporting?
-If so I'm not sure how EpiCompare helps us here.} -->
-\ben{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion.
-\sout{Moreover, m} M}any current \ben{time-dependent} comparison tools for state-space models (e.g.
-SIR models) \ben{\sout{highlight} examine} the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
-\ben{These \sout{This}} approach\ben{es} may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.
+[One of \pkg{EpiCompare}’s main contribution to] comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments.
+Epidemics, despite being defined a a process that evolves overt time, often needs to be compared in a way not constrained to initial times or time scales to understand the processes at play.] [With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible.]\footnote{\ben{[Ben says: brought this sentence from a later paragraph as it’s much more introductory.]}} Compared to time-dependent comparison tools for state-space modeling,  [time-invariant analysis can make it easier to compare state-space \ben{\sout{models} epidemic representations} in a more global, holistic fashion.] Many time-dependent comparison tools only [examine the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.] [These \ben{time-dependent approaches can} reduce the amount of \ben{\sout{connections that can be seen} insights that can be drawn}, similar to \ben{examining} projections of a multidimensional distribution onto a single axis \ben{one} at a time.
 Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens.
-The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
+\ben{At the same time, t}he goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.]
 
@@
 
-# Motivation and tools for time-invariant analysis {short-title="Time-invariant" #sec:time-invariant}
+# Time-invariant analysis
 
-<!-- \shannon{time invariance vs. time invariant analysis??} -->
+[EpiCompare emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis.
+In this section we will highlight weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take.
+To accomplish this goal we will first demonstrate the inability of time-dependent tools to well quantify a classic epidemic parameter, the reproduction number $R_0$. 
+Then we will motivate new time-invariant approaches for more complex situations where $R_0$ doesn’t capture the complexities of outbreaks.
+This will lead us to our final subsection that defines ways to view epidemics in a time-invariant lense and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.
 
-<!-- \shannon{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
-The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot\footnote{\ben{[Ben says: defend how this fits into the narrative?]}} and 4) talk }\ben{Time invariance in} \sout{about}\shannon{ models beyond SIR.} -->
+## Motivation of time-invariant analysis through the reproduction number $R_0$ {short-title="Motivation through the reproduction number $R_0$" #sec:r0_subsection}
 
-<!-- \shannon{I noticed the ternary plot kinda fell out of this section and it's not really detailed anywhere else in the paper.
-So I put it back in.
-It's a non-standard plot and a feature of EpiCompare so I think it's worth talking about, and I think it's appropriate to introduce in this section.} -->
+A lot of epidemiological research is interested in understanding intrinsic properties of the epidemic, not depending upon exactly when the epidemic occurred or how frequently data was collected.
+One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$, a time-invariant value that is [defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population.] Interestingly, \citet{Gallagher2020} has shown that the estimation of $R_0$  [can be sensitive to time-dependent parameters] like estimation of [the beginning and end of an epidemic].
+This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable properties of an epidemic.
 
-<!--Epidemics can be difficult to compare to one another due to differences in diseases, locations, or population behaviors, or times.
-In \pkg{EpiCompare}, we emphasize comparisons between epidemics adjusting for the last component, time.
-Time-invariant analysis is beneficial because by adjusting for the unit of infection rate, we can focus on the "lifetime" of an epidemic, a view that is concerned more with the number of lives affected than as opposed to any specific time constraint.
-Time-invariant analysis is also beneficial when there are gaps in time between occurrences of outbreaks of a similar nature in different geographic regions.
-Finally, time-invariant analysis is beneficial because many studies talk about a period of ``exponential growth'' of the number of outbreaks \citep{chowell2007,wallinga2007generation,forsberg2008}.
-
-Time-invariant analysis, as it appears in \pkg{EpiCompare} solves the above issues by observing functionals which attempt to capture the general shape of the epidemic with respect to the proportion of the population in each epidemic state.
-This allows us to compare scenarios as different as, for instance as a decades-long outbreak HIV in the US compared to a 10 day outbreak of norovirus on a cruise ship.
-Moreover, this tool avoids the need for choosing a 'beginning' $t_0$ or 'end' $t_F$ of an epidemic, choices \cite{gallagher2020} show that can heavily influence, for example,  estimates of peak infection height or the reproduction number $R_0$.
--->
-
-\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories.
-The following section aims to highlight the strengths of \textit{time-invariant} \shannon{analysis} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
-
-Mathematically, epidemics are complex objects.
-They can be hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time \sout{related}\shannon{related} features (including start of the epidemic, speed of infection and more).
-Time-invariant analysis makes different epidemics easier to compare by removing many time dependent aspects of an epidemic.
-\sout{Instead,} \shannon{Time-invariant analysis} focuses \shannon{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.} \ben{[Ben wants to try this sentence again.]}
-
-## Motivating time-invariant analysis through the reproduction number $R_0$ {short-title="motivating through R0" #sec:r0}
-
-Time-invariant analysis, as it appears in \pkg{EpiCompare}, \shannon{bypasses} many difficulties \shannon{in} comparing different epidemics.
-With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible.
-Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$.
-\sout{\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}^[\shannon{I don't think this is a necessary sentence.} \ben{I still think it adds value to the story and I'm not sure people really read section titles that are long.}]
-
-<!-- \ben{ -->
-<!-- Story:  -->
-<!-- (1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic.
-Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020).
--->
-<!-- } -->
-
-$R_0$ is probably the most famous \sout{time-invariant} numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
-$R_0$ is \sout{a one-number summary of a disease and is }defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.
-\ben{This \shannon{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\shannon{.
-Yet $R_0$ is} estimated with time-\sout{based}\shannon{dependent} data, which can make it a difficult quantity to estimate.} \shannon{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\shannon{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\ben{hard to define precisely.\sout{do not have precise definitions}}.
-To demonstrate the difficulty of discerning $R_0$ in \ben{\sout{a}other}\footnote{\ben{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}} time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model.
-This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
+In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproducible number.
+One common time-dependent tool is a series of time series line plots that track the proportion of the population in a given state (e.g.
+infected) at a given time.
+Figure \ref{fig:different-scales-standard} visualizes two different simulated epidemics with population states (s)usceptible, (i)nfectious and (r)ecovered.
+These simulations are actually generated under a discrete approximation of  \citet{Kermack1927}’s SIR model.
+[This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,]
 
@@
 
-From this model,\shannon{the reproduction number is the ratio of the infection rate to the recovery rate,} $R_0 = \beta/\gamma$\sout{, aka the ratio of the infection rate compared to the recovery rate.} \ben{From this definition, given
-\sout{Since}} $\beta$ and $\gamma$ are both rates, \ben{it should be clear that} the ratio of the two, $R_0$, is a time-invariant quantity.\footnote{\ben{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}} \sout{ Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
-Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
-}^[cool facts about r0, but not the central point]
-
-\ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
-Also, I feel like the story is weak here.
-The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$.
-In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
-\shannon{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
-also highlighted tie to time-invariant analysis and $R_0$.
-I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}\shannon{ Shannon tries again in \textcolor{blue}{blue}}
-
-\sout{\ben{[Ben says: this paragraph needs to still be looked at.
-Also I'm not sure why this particular paragraph was changed so much at all.
-Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \shannon{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
-The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
-Both epidemics have populations of 1000 people with 10 individuals initially infected.
-Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.
-We plot the epidemics with traditional $state$ vs.
-$time$ plots}\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
-\sout{In Fig.
-\ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.
-In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.
-On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \ben{epidemic} has not yet peaked at time 15.
-In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$}\footnote{\ben{This sentnce doesn't connect wit previous examples.}}.
-
-\textcolor{blue}{Since $R_0$ is an important value, it would be helpful to have more intuitive ways of comparing one $R_0$ to another.
-Usually numerical summaries of $R_0$ are presented, which while overall very helpful, may be confusing when presented along side epidemic data that are visualized in a traditional, time-dependent manner.}
-
-\textcolor{blue}{For example, consider two epidemics generated from the Kermack and McKendrick SIR equations where both have the same value of $R_0$.
-The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
-Both epidemics have populations of 1000 people with 10 individuals initially infected.
-An analysis may present an estimate of $\hat{R}_0 = 2$ alongside state vs.
-time plots like those shown in Figure \ref{fig:different-scales-standard}.
-The paths of the epidemics in the state vs.
-time view seem to differ from one another including having different infection peaks.
-From these traditional time-based plots, there is no intuitive way to conclude that these two epidemics have the same value of $R_0$.}
+[From this model, the reproduction number is the ratio of the infection rate to the recovery rate, $R_0 = \beta/\gamma$.]
 
@@
 
-```{r echo = FALSE, fig.cap = "\\label{fig:different-scales-standard}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+```{r echo = FALSE, fig.cap = "\\label{fig:different-scales-standard}Example of two epidemics with different $\\beta$ and $\\gamma$ parameters but the same initial reproduction number $R_0$ = 2.
 Both epidemics are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$."  }
@@
 
-\textcolor{blue}{\pkg{EpiCompare} provides a time-invariant tool to visualize these epidemics in a more intuitive manner, at least in regards to comparing values of $R_0$.} \sout{A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.}  \ben{For every time \sout{point} $t$ we have a \shannon{point} $(S(t),I(t), R(t))$\sout{, so we can treat epidemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.} \shannon{so we can visualize the trajectory of the epidemic in three-dimensional space (see Fig.
-\ref{fig:different-scales-tern} (left)).} For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern} (right).
-\sout{In Fig.
-\ref{fig:different-scales-tern}
-we plot the filamental trajectories of the two epidemics in a time-invariant view.
-We will explain how and why this works shortly.
-The important takeaway is that in this time-invariant view, i}I}t is apparent that these epidemics are on "the same path." In this case, this indicates that two epidemics have the same value of $R_0$.
+Even though Figure \ref{fig:different-scales-standard}’s visual analysis might be able to present more complex properties of the epidemic, one might wish to understand how these two simulated epidemics’ $R_0$ values compare.
+Even though these two simulated epidemics appear different in these figures ([including having different infection peaks]), there is no real intuitive way to compare the epidemic’s $R_0$.
+In fact, both of these simulated epidemics started out the same population (1000 people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$).
+Even with simple generative models, this time-dependent visualization tools is not able to be used to compare a very simple but powerful numerical summary.
+
+## Ternary plots: a time-invariant visualization tool
+
+The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state could be examined at a given time.
+A more holistic approach to visualizing the overall path of our simulated epidemics would be to examine each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space.
+Figure 3’s left plot does just that, presenting the trajectories of the simulated epidemics in this three-dimensional space.
+[For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot.] In this both of Figure  \ref{fig:different-scales-tern}’s plots we an observe that the epidemics are on the same trajectory.
+In this simple generative example setting, [this indicates that \ben{the} two epidemics have the same $R_0$ value].
+This can be proven mathematically.
 
 ```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Left: trajectory of epidemic in three-dimensional space, plotting $(S(t), I(t), R(t))$.
-Right: the gray-shaded region and epidemic trajectory shown from (left) now shown in two-dimensional space.
+Right: the gray-shaded region and epidemic trajectories shown from (left) now shown in two-dimensional space.
 This is more commonly known as a ternary plot.", fig.show="hold", out.width="49%"}
@@
 
-\ben{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig.
-\ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space.
-\sout{The reason why we can visually compare $R_0$ in Fig.
-\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}} A filamental trajectory can be mathematically viewed as a set of points in space that have an ordering, and that all points on the line between these ordered points are also contained in the geometric object.
-For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as
-
-$$
-\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
-$$
-where a mapping $\xi : s \to \mathbb{R}$ that is strictly monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi \equiv \psi$ where :
-
-$$
-\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
-$$
-\ben{[Ben says: removed this paragraph now]} Since the number \shannon{of individuals} in each state is non-negative and the sum over the three states for a given time point \sout{sums to}\shannon{is} $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.
-\sout{We can then which can be visualize the \shannon{full filamental trajectory} in a two-dimensional ternary plot}  As a result, we can visualize \shannon{the full filamental trajectory} of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.
-\shannon{[Shannon says: Show pic here?]}]\footnote{Which?
-\shannon{3d space one?
-But I'm leaning against it now.
-3d never looks good in a paper.}}
-
-\sout{\ben{[This section could be a bit less wordy.
-But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
-Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.
-Now, it is much clearer that \ben{\sout{Model} epidemic} 2 is following the same trajectory as \ben{\sout{Model} epidemic} 1 but is not as far along in the infection process.
-}
-
-\sout{\ben{As suggested a few paragraphs back, t}}The filamental trajectories in Fig.
-\ref{fig:different-scales-tern} \sout{seem to overlap}, and we may suspect that something is fundamentally linking these two different epidemics together.
-Mathematically, we can show this fundamental link \sout{turns} is $R_0$.
-Let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.
-As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
-And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.
-
 \begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+ Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (aka $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ and define $a > 0$ such that $\beta_2 = a\beta_1$.
+[Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
+Moreover, $s = \frac{1}{a}t$.]
 \end{theorem}
 
-The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model.
-\shannon{In other words, the two epidemics form the same filamental trajectory.
-For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}
+[The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model can be mapped to a point on the epidemic path of the second SIR model.
+In other words, the two epidemics form the same filamental trajectory.]
 
-<!-- ## $R_0$  and time-invariant analysis {short-title="r0" #r0}
+## Time-invariance beyond SIR models: Trajectories and Filaments
 
- By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population.
-$R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.
-For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks as large as 12 \citep{midas2020,biggerstaff2014,namee2018}.
+[Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us  better compare the overall structure of the epidemic and see how the population is directly impacted.] In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states.
+These tools leverage the idea that an epidemic can be viewed as a trajectory and that many properties of the epidemic are well captured when we do so.
+This approach is useful when the epidemic of interest has only gone [through a single realization of its outbreak (before the population of individuals become susceptible again)].
 
-Estimators for $R_0$ are dependent on the epidemic modeling framework, which consists of which states an individual can occupy (e.g.
-susceptible, infectious, recovered) and a description of how individuals move from one state to the next over time (see \cite{hethcote1994}).
-A common epidemic modeling framework is the SIR model, originally introduced by @kermack1927.
-Transitions from one state to the next are defined by a series of ordinary differential equations, where $N$ is the (fixed) total number individuals, $\beta$ is the rate of infection, and $\gamma$ is rate of recovery,
-\begin{align}\label{eq:sir-ode}
-      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
-      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
-      R^\prime(t) &= \gamma I(t) \nonumber.
-  \end{align}
-From this, $\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}$, the ratio of the estimated infection rate compared to the estimated recovery rate.
+The first set of tools allows a practioner to define distances between epidemics whose time features don’t well align.
+For completed epidemics, one way to better examine their properties is to represent their filamental trajectories as a finite sequence of equally spaced points.
+[This representation induces a natural distance \ben{\sout{between this type of representation}} between epidemics, specifically:]
 
-With regards to traditional epidemic $state$ vs.
-$time$ plots, $R_0$ is difficult to visualize, especially with respect from one epidemic to another.
-For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = 0.3$ and $\gamma_1 = 0.15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = 0.24$ and $\gamma_1 = 0.12$ over the first 15 days.
-Both epidemics have the same value of $R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2$.
-The epidemic trajectories are shown in the $state$ vs.
-time plots in Figure \ref{fig:different-scales-standard}.
-At a glance, we may assume that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
-On the other hand, we may think Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15.
--->
+\[
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
+\]
+where $\tilde{\psi}_i(s)$ the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$.
+This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
+These extremeness estimators can be useful when comparing the true epidemic to a a set of simulated epidemics,] and practitioners can interpret the univariate epidemic's extremeness score relative to the extremeness scores of the simulations very easily.
 
-<!--
-However, when we plot the trajectories as a single curve using the ternary plot in the time-invariant view, we immediately see a different story.
-In this time-invariant view in Fig.
-\ref{fig:different-scales-tern}, the points seem to overlap and form the same trajectory.
-Now it seems to be that Model 2 is following the same trajectory as Model 1 but is not as far along in the infection process.
-We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
+When the goal is to predict the future impact / trajectory of a epidemic, time-invariant tools can help describe state-space regions in which we expect the true epidemic to traverse.
+In settings where the epidemic only goes a single outbreak, these regions can be very telling if simulation models well capture the epidemic’s structure.
+In \pkg{EpiCompare} we create geometric prediction regions around all but the $\alpha$ proportion of most extreme simulated epidemics.
+These geometric regions can also be used to compared simulation models that have different time scales, parameters and even different statistical philosophies, through set different distances.
+Although visualization is the easiest when the epidemic has three states, prediction regions can be useful to assess and compare simulation models where the epidemics have multiple states.
 
-More formally, let two Kermack and McKendrick  SIR models(see Eq.\eqref{eq:sir-ode} )  be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
-Assume both models have initial values $(S(0), I(0), R(0))$.
-Let $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
-Equivalently, $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$ for some $a > 0$.
-
-\begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
-\end{theorem}
-
-The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.
-Taking the sample simulations from Fig.
-\ref{fig:different-scales-standard}, Fig.
-\ref{fig:different-scales-tern} presents these two models in a ternary plot.
-This means with our time-invariant ternary plot, that at a glance, we can tell if two epidemics have different values of $R_0$.
--->
-
-##  Time-invariant analysis beyond $R_0$ and \sout{Kermack’s and McKendrick} SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #sec:beyond-r0-sir}
-
-Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to \sout{better}\shannon{more fully} compare the overall structure of the epidemic and see how the population is directly impacted.
-Time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown or have more than three epidemic states.
-
-\ben{New paragraph} Viewing epidemics as filamental trajectories provides \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner.
-\ben{For \shannon{completed?}epidemics that have ended, one way to examine their filamental trajectories is to \sout{redefine}\shannon{represent} the \shannon{filamental} trajectory as a \shannon{finite sequence of equally spaced points.} \sout{finite sequence of points on the filamental trajectory that are equally spaced (i.e.
-equa-distance between pairs of ordered points).} \sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{[Shannon says: are we missing a few words in this definition?]}}.
-This representation induces a natural distance between this type of representation between epidemics, specifically:
-\[
-d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
-\]
-where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$.\footnote{\shannon{I think you're trying to say something about a distance based on the equally space points.
-Some clarifying questions:  is $\psi^\prime$ the derivative?
-Does proportion make more sense than fraction?
-or simply $\frac{|\psi|}{s}$?
-It's only naturally time-invariant if we have a well defined ending point, right?}} This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
-These extremeness estimators can be \sout{very} useful when comparing \sout{between a setof simulation} \shannon{a set of simulated} epidemics and the true epidemic\shannon{.
-Moreover these extremness estimators,} \sout{and does} \shannon{} not constrain the number of states of the model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\ben{I'm not sure we've talked about this before...
-\shannon{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
-
-\ben{New paragraph:} \ben{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If \sout{one} \shannon{a practicioner} is interested in understanding \shannon{an} epidemic\sout{s} through a single \sout{cycle}\shannon{realization} of \sout{their}\shannon{its} outbreak} (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can be leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?
-Update 4/6: I'm satisfied.}}].
-In these settings, \shannon{\pkg{EpiCompare} goes} \sout{we go }a step further and treats epidemics more like geometric filaments \shannon{(i.e.
-filamental trajectories without an ordering of points)} than filamental trajectories.
-In \pkg{EpiCompare}, we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small \shannon{geometric?} filaments around the subset of simulations (\sout{subset} \shannon{grouped} by measures like the above pseudo-density estimates or depth estimates).
-These regions \sout{look at}\shannon{show} where in the state-space we expect the epidemic to traverse\sout{, and }.
-\shannon{Additionally, }we can compare prediction regions defined by different models using \ben{many set difference distances \sout{the Haussdorf \shannon{why Haussdorf specifically?} distance}} as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
-All these \shannon{mentioned?} geometric structures and distance notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).
-
-<!--
-## Beyond the Kermack and McKendrick SIR models
-
-Although the result of Theorem \ref{thm:sir-scale} allows for easy visual comparison of $R_0$ in Kermack SIR models, it does require stringent assumptions of homogeneity of behavior in populations.
-The use of visualizing epidemics in a time-invariant lens via ternary plots extends beyond those of models that follow the ODEs in the Kermack-McKendrick equations.
-Any model with S, I, and R states can be visualized with ternary plots, regardless of birth and death dynamics and regardless of homogeneity of individual behavior.
-We can use ternary plots to compare the spread of a disease for groups within a population without time as a confounding factor.
-We show an example of this in a later section.
-
-Moreover, time-invariant analysis is also useful for epidemic models with more than three epidemic states.
-The constraints in three dimensions that are met with the SIR model (there are three epidemic states and for each time step, the sum of total number of individuals in each state is the population size at that time) represents a space of 3d simplices, and the ternary plot specifically represents these simplices, after scaling the values in each state as proportions of the total population at that time (ternary plots are known as a 3d unit simplex due to it's scaling).
-This same scaling for larger models (i.e.
-with more states) can be done onto different simplexes.
-In this package we present tools to help compare models (mostly through simulations).
-\pkg{EpiCompare}'s tools permits comparisons of $d$-dimensional objects by projecting them into ($d-1$)-dimensional space.
-
- Even though higher dimensional models may not be able to visualized simply, we provide a number of tools to aid in the comparison of models and epidemics.
-The first of which uses multiple simulations under specific model parameters to assess the \shannon{variability?} bairabilty of the model fit.
-In \pkg{EpiCompare}, we provide ways to create prediction regions for a `true' epidemic under a fully specified model.
-These regions require representing multi-dimensional structures for functions to completely contain epidemics.
-\pkg{EpiCompare} treats these simulations and epidemics as \textit{filamental} objects -- \shannon{need a description for filamental, Ben feel free to change} where not only the outbreak trajectory points are important but also their form which can be considered a function.
-\shannon{could be useful to have a picture/example here?
-like  three curves and 2 metrics one where closer under a pointwise distance and the other closer in filamental distance?
-I'll sketch up an idea unless you have something in your proposal already}
-
- We extend off of papers like \citet{Dalmasso2019a} to create these bands.
-These high dimensional bands allow the user to assess if the true epidemic is within the band (thereby assessing the model's representation of the epidemic).
-With these bands, we can compare \textit{models} to \textit{models} by assessing the distance between the two bands of model representation.
-We recommend using the Hausdorff distance to compare two bands to one another, as this distance captures how much bigger the sets would have to expand to cover each other.
-The Hausdorff distance is defined mathematically as
-\[
-d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y),\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
-\]
-
--->
+Overall, there are many tools to aid in the assessment and comparison of epidemics and models that avoid being affected by time-based parameters.
+We believe the time-invariant analysis provides many insights and should be in the toolkit of many epidemiologists.
+\pkg{EpiCompare} provides a strong starting point to do just that.
 

### 2021-05-10 jss/EpiCompare-paper/EpiCompare-paper2.tex
new version of section 2

--- 2021-04-27 jss/EpiCompare-paper/EpiCompare-paper.tex
@@
 
-\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
-\textit{time-invariant} way. Epidemics, despite by definition being a
-process that evolves over time, often need to be compared in a way not
-constrained to initial times or time scales to understand the processes
-at play.
-\textcolor{violet}{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion. \sout{Moreover, m} M}any
-current \textcolor{violet}{time-dependent} comparison tools for
-state-space models (e.g.~SIR models)
-\textcolor{violet}{\sout{highlight} examine} the proportion of
-individuals in each state (at a given time) in a piece-wise / marginal
-fashion. \textcolor{violet}{These \sout{This}}
-approach\textcolor{violet}{es} may reduce the amount of connections that
-can be seen, similar to projections of a multidimensional distribution
-onto a single axis at a time. Tools in \pkg{EpiCompare} give the user
-the ability to extend their toolkit to evaluate epidemics within a
-time-invariant lens. The goal of \pkg{EpiCompare} is not to supplant
-existing infectious disease modeling tools and software but, rather, is
-a concerted effort to create standard and fair comparisons among models
-developed for disease outbreaks and outbreak data.
+{[}One of \pkg{EpiCompare}'s main contribution to{]} comparison and
+assessment of epidemics is through tools that provide
+\textit{time-invariant} assessments. Epidemics, despite being defined a
+a process that evolves overt time, often needs to be compared in a way
+not constrained to initial times or time scales to understand the
+processes at play.{]} {[}With time-invariant analysis, comparing
+decades-long outbreaks of HIV in the US to a 10 day outbreak of
+norovirus on a cruise ship is
+possible.{]}\footnote{\textcolor{violet}{[Ben says: brought this sentence from a later paragraph as it’s much more introductory.]}}
+Compared to time-dependent comparison tools for state-space modeling,
+{[}time-invariant analysis can make it easier to compare state-space
+\textcolor{violet}{\sout{models} epidemic representations} in a more
+global, holistic fashion.{]} Many time-dependent comparison tools only
+{[}examine the proportion of individuals in each state (at a given time)
+in a piece-wise / marginal fashion.{]} {[}These
+\textcolor{violet}{time-dependent approaches can} reduce the amount of
+\textcolor{violet}{\sout{connections that can be seen} insights that can be drawn},
+similar to \textcolor{violet}{examining} projections of a
+multidimensional distribution onto a single axis \textcolor{violet}{one}
+at a time. Tools in \pkg{EpiCompare} give the user the ability to extend
+their toolkit to evaluate epidemics within a time-invariant lens.
+\textcolor{violet}{At the same time, t}he goal of \pkg{EpiCompare} is
+not to supplant existing infectious disease modeling tools and software
+but, rather, is a concerted effort to create standard and fair
+comparisons among models developed for disease outbreaks and outbreak
+data.{]}
 
@@
 
-\section[Time-invariant]{Motivation and tools for time-invariant
-analysis}\label{sec:time-invariant}
-
-\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking
-a global, not marginal view of how epidemics move through populations
-and (2) by treating full epidemics as filamental trajectories. The
-following section aims to highlight the strengths of
-\textit{time-invariant} \textcolor{orange}{analysis} and define the
-mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
-
-Mathematically, epidemics are complex objects. They can be hard to
-assess and compare to one another due to the differences in the
-diseases, the location where the outbreak occurs, how the affected
-population reacts, and the time
-\sout{related}\textcolor{orange}{related} features (including start of
-the epidemic, speed of infection and more). Time-invariant analysis
-makes different epidemics easier to compare by removing many time
-dependent aspects of an epidemic. \sout{Instead,}
-\textcolor{orange}{Time-invariant analysis} focuses
-\textcolor{orange}{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.}
-\textcolor{violet}{[Ben wants to try this sentence again.]}
-
-\subsection[motivating through R0]{Motivating time-invariant analysis
-through the reproduction number \(R_0\)}\label{sec:r0}
-
-Time-invariant analysis, as it appears in \pkg{EpiCompare},
-\textcolor{orange}{bypasses} many difficulties \textcolor{orange}{in}
-comparing different epidemics. With time-invariant analysis, comparing
-the decades-long outbreak of HIV in the US to a 10 day outbreak of
-norovirus on a cruise ship is \sout{still} possible. Time-dependent
-problems can arise when estimating epidemiological parameters, including
-the reproduction number \(R_0\).
-\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence.}
-  \textcolor{violet}{I still think it adds value to the story and I'm not sure people really read section titles that are long.}}
-
-\(R_0\) is probably the most famous \sout{time-invariant} numerical
-summary of an epidemic and is often associated with the
-Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
-\(R_0\) is \sout{a one-number summary of a disease and is }defined as
-the expected number of infections caused by a single infector who is
-added to a completely susceptible population \citep{anderson1992}.
-\textcolor{violet}{This \textcolor{orange}{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\textcolor{orange}{.
-Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a difficult quantity to estimate.}
-\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
-To demonstrate the difficulty of discerning \(R_0\) in
-\textcolor{violet}{\sout{a}other}\footnote{\textcolor{violet}{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}}
-time-dependent analysis, we first introduce \citet{Kermack1927}'s SIR
-model. This model captures the transitions from one state to the next as
-a system of ordinary differential equations, where \(N\) is the total
-number of individuals, \(\beta\) is the rate of infection, and
-\(\gamma\) is the rate of recovery,
+\hypertarget{time-invariant-analysis}{%
+\section{Time-invariant analysis}\label{time-invariant-analysis}}
+
+{[}EpiCompare emphasizes the value of analyzing epidemics in a
+\textit{time-invariant} way - approaches that remove some or all of the
+impact of start/end times and recording time scales when preforming the
+analysis. In this section we will highlight weaknesses of
+time-\textit{de}pendent analysis and define the mathematical
+underpinning of the time-invariant approach we take. To accomplish this
+goal we will first demonstrate the inability of time-dependent tools to
+well quantify a classic epidemic parameter, the reproduction number
+\(R_0\). Then we will motivate new time-invariant approaches for more
+complex situations where \(R_0\) doesn't capture the complexities of
+outbreaks. This will lead us to our final subsection that defines ways
+to view epidemics in a time-invariant lense and discusses natural
+properties of these representations that allow for clearer ways to
+compare models and epidemics.
+
+\subsection[Motivation through the reproduction number $R_0$]{Motivation
+of time-invariant analysis through the reproduction number
+\(R_0\)}\label{sec:r0_subsection}
+
+A lot of epidemiological research is interested in understanding
+intrinsic properties of the epidemic, not depending upon exactly when
+the epidemic occurred or how frequently data was collected. One of the
+most famous numerical summaries of an epidemic is the reproduction
+number - \(R_0\), a time-invariant value that is {[}defined as the
+expected number of infections caused by a single infect or who is added
+to a completely susceptible population.{]} Interestingly,
+\citet{Gallagher2020} has shown that the estimation of \(R_0\) {[}can be
+sensitive to time-dependent parameters{]} like estimation of {[}the
+beginning and end of an epidemic{]}. This time-dependent sensitivity in
+estimating \(R_0\) reflects a general problem in epidemiology
+surrounding how to transform time-dependent information into insights
+into desirable properties of an epidemic.
+
+In many situations, epidemiologists want a much deeper understand of an
+epidemic than just the reproducible number. One common time-dependent
+tool is a series of time series line plots that track the proportion of
+the population in a given state (e.g.~infected) at a given time. Figure
+\ref{fig:different-scales-standard} visualizes two different simulated
+epidemics with population states (s)usceptible, (i)nfectious and
+(r)ecovered. These simulations are actually generated under a discrete
+approximation of \citet{Kermack1927}'s SIR model. {[}This model captures
+the transitions from one state to the next as a system of ordinary
+differential equations, where \(N\) is the total number of individuals,
+\(\beta\) is the rate of infection, and \(\gamma\) is the rate of
+recovery,{]}
 
@@
 
-From this
-model,\textcolor{orange}{the reproduction number is the ratio of the infection rate to the recovery rate,}
-\(R_0 = \beta/\gamma\)\sout{, aka the ratio of the infection rate compared to the recovery rate.}
-\textcolor{violet}{From this definition, given 
-\sout{Since}} \(\beta\) and \(\gamma\) are both rates,
-\textcolor{violet}{it should be clear that} the ratio of the two,
-\(R_0\), is a time-invariant
-quantity.\footnote{\textcolor{violet}{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}
-\sout{ Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
-Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
-}\footnote{cool
-  facts about r0, but not the central point}
-
-\textcolor{violet}{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
-Also, I feel like the story is weak here.
-The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$.
-In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
-\textcolor{orange}{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
-also highlighted tie to time-invariant analysis and $R_0$.
-I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}\textcolor{orange}{ Shannon tries again in \textcolor{blue}{blue}}
-
-\sout{\textcolor{violet}{[Ben says: this paragraph needs to still be looked at.
-Also I'm not sure why this particular paragraph was changed so much at all.
-Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
-The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
-Both epidemics have populations of 1000 people with 10 individuals initially infected.
-Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.
-We plot the epidemics with traditional $state$ vs.
-$time$ plots}\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
-\sout{In Fig.
-\ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.
-In this time-variant view, we may believe that \textcolor{violet}{epidemic} 1 has a larger $R_0$ than \textcolor{violet}{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.
-On the other hand, we may believe \textcolor{violet}{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \textcolor{violet}{epidemic} has not yet peaked at time 15.
-In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$}\footnote{\textcolor{violet}{This sentnce doesn't connect wit previous examples.}}.
-
-\textcolor{blue}{Since $R_0$ is an important value, it would be helpful to have more intuitive ways of comparing one $R_0$ to another.
-Usually numerical summaries of $R_0$ are presented, which while overall very helpful, may be confusing when presented along side epidemic data that are visualized in a traditional, time-dependent manner.}
-
-\textcolor{blue}{For example, consider two epidemics generated from the Kermack and McKendrick SIR equations where both have the same value of $R_0$.
-The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
-Both epidemics have populations of 1000 people with 10 individuals initially infected.
-An analysis may present an estimate of $\hat{R}_0 = 2$ alongside state vs.
-time plots like those shown in Figure \ref{fig:different-scales-standard}.
-The paths of the epidemics in the state vs.
-time view seem to differ from one another including having different infection peaks.
-From these traditional time-based plots, there is no intuitive way to conclude that these two epidemics have the same value of $R_0$.}
+{[}From this model, the reproduction number is the ratio of the
+infection rate to the recovery rate, \(R_0 = \beta/\gamma\).{]}
 
@@
 
-\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
+\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ parameters but the same initial reproduction number $R_0$ = 2.
 Both epidemics are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
@@
 
-\textcolor{blue}{\pkg{EpiCompare} provides a time-invariant tool to visualize these epidemics in a more intuitive manner, at least in regards to comparing values of $R_0$.}
-\sout{A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.}
-\textcolor{violet}{For every time \sout{point} $t$ we have a \textcolor{orange}{point} $(S(t),I(t), R(t))$\sout{, so we can treat epidemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.} \textcolor{orange}{so we can visualize the trajectory of the epidemic in three-dimensional space (see Fig.
-\ref{fig:different-scales-tern} (left)).} For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern} (right).
-\sout{In Fig.
-\ref{fig:different-scales-tern}
-we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, i}I}t
-is apparent that these epidemics are on ``the same path.'' In this case,
-this indicates that two epidemics have the same value of \(R_0\).
+Even though Figure \ref{fig:different-scales-standard}'s visual analysis
+might be able to present more complex properties of the epidemic, one
+might wish to understand how these two simulated epidemics' \(R_0\)
+values compare. Even though these two simulated epidemics appear
+different in these figures ({[}including having different infection
+peaks{]}), there is no real intuitive way to compare the epidemic's
+\(R_0\). In fact, both of these simulated epidemics started out the same
+population (1000 people with 10 infected) and have the same \(R_0\)
+value, just with slightly different infection and recovery rates
+(\(\beta_1, \gamma_1 = 0.8, 0.4\), and
+\(\beta_2, \gamma_2 = .64, .32\)). Even with simple generative models,
+this time-dependent visualization tools is not able to be used to
+compare a very simple but powerful numerical summary.
+
+\hypertarget{ternary-plots-a-time-invariant-visualization-tool}{%
+\subsection{Ternary plots: a time-invariant visualization
+tool}\label{ternary-plots-a-time-invariant-visualization-tool}}
+
+The faceted time series plot, like that seen in Figure
+\ref{fig:different-scales-standard}, not only fails to allow the
+practitioner to compare simulated epidemics' \(R_0\) values but also
+presented epidemics' trajectory data so that only the marginal
+information for each state could be examined at a given time. A more
+holistic approach to visualizing the overall path of our simulated
+epidemics would be to examine each epidemic traverses the
+three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
+just that, presenting the trajectories of the simulated epidemics in
+this three-dimensional space. {[}For state space models like in our
+example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
+to \(N\) (the total population size) we can visual these point in a a
+two-dimesional \textit{ternary} plot, as seen in Figure
+\ref{fig:different-scales-tern}'s right plot.{]} In this both of Figure
+\ref{fig:different-scales-tern}'s plots we an observe that the epidemics
+are on the same trajectory. In this simple generative example setting,
+{[}this indicates that \textcolor{violet}{the} two epidemics have the
+same \(R_0\) value{]}. This can be proven mathematically.
 
@@
 \caption{\label{fig:different-scales-tern}Left: trajectory of epidemic in three-dimensional space, plotting $(S(t), I(t), R(t))$.
-Right: the gray-shaded region and epidemic trajectory shown from (left) now shown in two-dimensional space.
+Right: the gray-shaded region and epidemic trajectories shown from (left) now shown in two-dimensional space.
 This is more commonly known as a ternary plot.}\label{fig:unnamed-chunk-3}
@@
 
-\textcolor{violet}{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig.
-\ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space.
-\sout{The reason why we can visually compare $R_0$ in Fig.
-\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}}
-A filamental trajectory can be mathematically viewed as a set of points
-in space that have an ordering, and that all points on the line between
-these ordered points are also contained in the geometric object. For a
-SIR epidemic, we can represent the associated filamental trajectory
-\(\psi\) as
-
-\[
-\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
-\] where a mapping \(\xi : s \to \mathbb{R}\) that is strictly
-monotonically increasing would not change the definition of \(\psi\),
-i.e.~\(\psi_\xi \equiv \psi\) where :
-
-\[
-\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
-\] \textcolor{violet}{[Ben says: removed this paragraph now]} Since the
-number \textcolor{orange}{of individuals} in each state is non-negative
-and the sum over the three states for a given time point
-\sout{sums to}\textcolor{orange}{is} \(N\), then
-\textcolor{violet}{all points in} \(\psi\) will lay in a
-\textbackslash ben\{two-dimensional triangular plane in
-three-dimensional space.
-\sout{We can then which can be visualize the \textcolor{orange}{full filamental trajectory} in a two-dimensional ternary plot}
-As a result, we can visualize
-\textcolor{orange}{the full filamental trajectory} of an epidemic in a
-single
-\textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
-\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?  \textcolor{orange}{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}
-
-\sout{\textcolor{violet}{[This section could be a bit less wordy.
-But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
-Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.
-Now, it is much clearer that \textcolor{violet}{\sout{Model} epidemic} 2 is following the same trajectory as \textcolor{violet}{\sout{Model} epidemic} 1 but is not as far along in the infection process.
-}
-
-\sout{\textcolor{violet}{As suggested a few paragraphs back, t}}The
-filamental trajectories in Fig. \ref{fig:different-scales-tern}
-\sout{seem to overlap}, and we may suspect that something is
-fundamentally linking these two different epidemics together.
-Mathematically, we can show this fundamental link \sout{turns} is
-\(R_0\). Let our two epidemics be presented as
-\(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
-\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
-example, assume both models have the same initial values
-\((S(0), I(0), R(0))\), and let
-\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
-\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
-rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
-the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
-\(\gamma_2 = a \gamma_1\).
-
 \begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+ Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (aka $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ and define $a > 0$ such that $\beta_2 = a\beta_1$.
+[Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
+Moreover, $s = \frac{1}{a}t$.]
 \end{theorem}
 
-The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
+{[}The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
 result from \cite{Harko2014} and is shown in detail in Proof
@@
 each state and \(R_0\) then for every point on the epidemic path of the
-first SIR model \textcolor{violet}{\sout{is also} can be mapped to} a
-point on the epidemic path of the second SIR model.
-\textcolor{orange}{In other words, the two epidemics form the same filamental trajectory.
-For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}
-
-\subsection[Beyond R0 and SIR]{Time-invariant analysis beyond \(R_0\)
-and \sout{Kermack’s and McKendrick} SIR
-Models\footnote{Probably will need to change this title...}}\label{sec:beyond-r0-sir}
-
-Through the \(R_0\) example, we see that treating epidemics like
+first SIR model can be mapped to a point on the epidemic path of the
+second SIR model. In other words, the two epidemics form the same
+filamental trajectory.{]}
+
+\hypertarget{time-invariance-beyond-sir-models-trajectories-and-filaments}{%
+\subsection{Time-invariance beyond SIR models: Trajectories and
+Filaments}\label{time-invariance-beyond-sir-models-trajectories-and-filaments}}
+
+{[}Through the \(R_0\) example, we see that treating epidemics like
 filamental trajectories embedded in a lower dimensional space allows us
-to \sout{better}\textcolor{orange}{more fully} compare the overall
-structure of the epidemic and see how the population is directly
-impacted. Time-invariant tools \sout{that} can be useful even when the
-underlying generative model for the epidemic is unknown or have more
-than three epidemic states.
-
-\textcolor{violet}{New paragraph} Viewing epidemics as filamental
-trajectories provides \sout{a lot }new ways to compare and examine
-epidemics in a time-invariant manner. \textbackslash ben\{For
-\textcolor{orange}{completed?}epidemics that have ended, one way to
-examine their filamental trajectories is to
-\sout{redefine}\textcolor{orange}{represent} the
-\textcolor{orange}{filamental} trajectory as a
-\textcolor{orange}{finite sequence of equally spaced points.}
-\sout{finite sequence of points on the filamental trajectory that are equally spaced (i.e. equa-distance between pairs of ordered points).}
-\sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \textcolor{orange}{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \textcolor{orange}{[Shannon says: are we missing a few words in this definition?]}}.
-This representation induces a natural distance between this type of
-representation between epidemics, specifically: \[
-d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
-\] where \(\psi_i'(s)\) the point along \(\psi_1\) that is \(s\)
-fraction of \(|\psi_1|\) distance away from the start of
-\(\psi_i\).\footnote{\textcolor{orange}{I think you're trying to say something about a distance based on the equally space points.
-Some clarifying questions:  is $\psi^\prime$ the derivative?
-Does proportion make more sense than fraction?
-or simply $\frac{|\psi|}{s}$?
-It's only naturally time-invariant if we have a well defined ending point, right?}}
-This distance is naturally time-invariant, and can be plugged into
-multiple distance-based assessment tools to examine the overall
-``extremeness'' of points, including pseudo-density estimators and
-depth/local depth functions
-\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
-estimators can be \sout{very} useful when comparing
-\sout{between a setof simulation} \textcolor{orange}{a set of simulated}
-epidemics and the true
-epidemic\textcolor{orange}{. Moreover these extremness estimators,}
-\sout{and does} \textcolor{orange}{} not constrain the number of states
-of the
-model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\textcolor{violet}{I'm not sure we've talked about this before...
-\textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
-
-\textcolor{violet}{New paragraph:}
-\textcolor{violet}{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If \sout{one} \textcolor{orange}{a practicioner} is interested in understanding \textcolor{orange}{an} epidemic\sout{s} through a single \sout{cycle}\textcolor{orange}{realization} of \sout{their}\textcolor{orange}{its} outbreak}
-(before the population of individuals have become susceptible again),
-then additional time-invariant tools, including prediction regions can
-be leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done? Update 4/6: I'm satisfied.}\}}.
-In these settings, \textcolor{orange}{\pkg{EpiCompare} goes}
-\sout{we go }a step further and treats epidemics more like geometric
-filaments
-\textcolor{orange}{(i.e. filamental trajectories without an ordering of points)}
-than filamental trajectories. In \pkg{EpiCompare}, we create prediction
-regions that contain a the top (\(1-\alpha\)) proportion of simulated
-curves by defining geometric regions defined by the union of small
-\textcolor{orange}{geometric?} filaments around the subset of
-simulations (\sout{subset} \textcolor{orange}{grouped} by measures like
-the above pseudo-density estimates or depth estimates). These regions
-\sout{look at}\textcolor{orange}{show} where in the state-space we
-expect the epidemic to traverse\sout{, and }.
-\textcolor{orange}{Additionally, }we can compare prediction regions
-defined by different models using
-\textcolor{violet}{many set difference distances \sout{the Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance}}
-as well as examining how well the truth epidemic matches the simulations
-by examining if the epidemic's trajectory lies within the prediction
-region. All these \textcolor{orange}{mentioned?} geometric structures
-and distance notations apply to epidemics with any number of states, and
-at the end of Section \ref{sec:overview} we \sout{also} highlight how
-these prediction regions can aid in visual comparisons for epidemics
-with 3 states (like the SIR models).
+better compare the overall structure of the epidemic and see how the
+population is directly impacted.{]} In this section we present
+time-invariant tools that can be applied to complex epidemics where the
+epidemic's generative process is unknown and can have more than three
+states. These tools leverage the idea that an epidemic can be viewed as
+a trajectory and that many properties of the epidemic are well captured
+when we do so. This approach is useful when the epidemic of interest has
+only gone {[}through a single realization of its outbreak (before the
+population of individuals become susceptible again){]}.
+
+The first set of tools allows a practioner to define distances between
+epidemics whose time features don't well align. For completed epidemics,
+one way to better examine their properties is to represent their
+filamental trajectories as a finite sequence of equally spaced points.
+{[}This representation induces a natural distance
+\textcolor{violet}{\sout{between this type of representation}} between
+epidemics, specifically:{]}
+
+\[
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
+\] where \(\tilde{\psi}_i(s)\) the point along \(\psi_i\) that is
+\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\). This
+distance is naturally time-invariant, and can be plugged into multiple
+distance-based assessment tools to examine the overall ``extremeness''
+of points, including pseudo-density estimators and depth/local depth
+functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. These
+extremeness estimators can be useful when comparing the true epidemic to
+a a set of simulated epidemics,{]} and practitioners can interpret the
+univariate epidemic's extremeness score relative to the extremeness
+scores of the simulations very easily.
+
+When the goal is to predict the future impact / trajectory of a
+epidemic, time-invariant tools can help describe state-space regions in
+which we expect the true epidemic to traverse. In settings where the
+epidemic only goes a single outbreak, these regions can be very telling
+if simulation models well capture the epidemic's structure. In
+\pkg{EpiCompare} we create geometric prediction regions around all but
+the \(\alpha\) proportion of most extreme simulated epidemics. These
+geometric regions can also be used to compared simulation models that
+have different time scales, parameters and even different statistical
+philosophies, through set different distances. Although visualization is
+the easiest when the epidemic has three states, prediction regions can
+be useful to assess and compare simulation models where the epidemics
+have multiple states.
+
+Overall, there are many tools to aid in the assessment and comparison of
+epidemics and models that avoid being affected by time-based parameters.
+We believe the time-invariant analysis provides many insights and should
+be in the toolkit of many epidemiologists. \pkg{EpiCompare} provides a
+strong starting point to do just that.
 
@@
 \midrule
-Simple SIR & 0.0036733 & 0\\
-EpiModel SIR & 0.0028813 & 0\\
+Simple SIR & 0.0117621 & 0.01\\
+EpiModel SIR & 0.0356503 & 0.01\\
 \bottomrule

### 2021-05-10 jss/EpiCompare-paper/EpiCompare-paper.tex

--- 2021-04-27 jss/EpiCompare-paper/EpiCompare-paper.tex
@@
 the data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
-front-facing functions are aimed to be as user-friendly as possible. We
-also focus on providing the user ``tidyverse'' style functions, that
-encourage piping objects from one function to the next and follow clear
-``verb'' naming schemes \citep{Wickham2019}. Although users can
-incorporate \pkg{EpiCompare} into any step in the data analysis
-pipeline, there are two primary points of entry. The first point of
-entry is the very beginning with pre-processing and visualizing raw
-data, and the second point of entry is after modeling and simulation.
-Figure \ref{fig:pipeline2} captures these different paths, and we
-highlight\footnote{\textcolor{violet}{[Ben says: we need to make sure we actually do highlight ]}}
-both approaches and how to leverage \pkg{EpiCompare} in the subsections
-below.
+front-facing functions in \pkg{EpiCompare} are aimed to be as
+user-friendly as possible. We also focus on providing the user
+``tidyverse'' style functions, that encourage piping objects from one
+function to the next and follow clear ``verb'' naming schemes
+\citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare}
+into any step in the data analysis pipeline, there are two primary
+points of entry. The first point of entry is the very beginning with
+pre-processing and visualizing raw data, and the second point of entry
+is after modeling and simulation. Figure \ref{fig:pipeline2} captures
+these different paths, and we highlight how to leverage \pkg{EpiCompare}
+functionalities in the subsections below.
 
@@
 
-The first step of most data analysis is ``cleaning'' the data
-\textcolor{orange}{to a format that is friendly for both computers and programmers}\footnote{\textcolor{violet}{[Ben says: I'm unsure this is needed and naturally adds more text.
-Willing to accept...]}}
-so it can be explored.
-\sout{There are multiple ways to collect epidemiological data.}
-\textcolor{violet}{In epidemology, there are multiple different formats the data can arrive in.}
-\footnote{\textcolor{orange}{[Shannon suggests: "Epidemiological data are collected in many different formats."]} \textcolor{violet}{[Ben says: I see the point you're trying to make and made a new edit - but it's less connected to the previous sentence since "format" is the end.]}}
-Sometimes individual records are collected, with times of different
-states of the epidemic (infection, recovery, etc.) as well as individual
-information like network structure, location, and sub-population
-information. Other data collections focus on aggregate counts of
-individuals in each epidemic state. In fact, many times only the number
-of new infections at each time step (e.g.~weekly case counts) is
-observed. Compartment totals (amounts of individuals in each state) are
-then imputed from those case
-counts\sout{ along with}\textcolor{orange}{and}\footnote{\textcolor{violet}{[Ben says: Declined.
-I'm unclear why we should.
-Is the sentence unclear?
-Maybe it's the way it's a compound sentence?
-- If so a rewrite would be better?]}}
-other information about the disease and the population of interest. In
-\pkg{EpiCompare}, we focus on understanding the overall impact of an
-outbreak at the aggregate/population level, which allows for streamlined
-examination of overall trends of an epidemic.
-
-To help the practitioner examine epidemics from an aggregate/ population
+The first step of most data analysis is ``cleaning'' the raw data so it
+can be explored. Before data can be explored, they must be collected.
+Raw data can appear in many different formats. Sometimes individual
+records are collected, with times of different states of the epidemic
+(infection, recovery, etc.) as well as individual information like
+network structure, location, and sub-population information. Other data
+collections focus on aggregate counts of individuals in each epidemic
+state. In fact, many times only the number of new infections at each
+time step (e.g.~weekly case counts) is observed. Compartment totals
+(amounts of individuals in each state) are then imputed from those case
+counts other information about the disease and the population of
+interest. In \pkg{EpiCompare}, we focus on understanding the overall
+impact of an outbreak at the aggregate/population level, which allows
+for streamlined examination of overall trends of an epidemic.
+
+To help the practitioner examine epidemics from an aggregate/population
 lens, we provide a function called \texttt{agents\_to\_aggregate()}.
@@
 aggregate view of how many individuals were in a state at a given time.
-There are often situations where
-\textcolor{violet}{\sout{grouping} aggregating}\footnote{\textcolor{violet}{[Ben says: reason - connect to languare from sentence before.]}}
-agents into subpopulations
-(e.g.~sub\sout{populations}\textcolor{orange}{ groups}\footnote{\textcolor{violet}{[Ben says: declined. I'm unclear why we'd do this.]}}
-defined by age or sex) can highlight different aggregate level trends.
-For example, research by \citet{rvachev1985,anderson1992,worby2015}
-develop state-based models that account for differing disease dynamics
-in different subpopulations. In \pkg{EpiCompare}, we facilitate
-subpopulation analysis by combining the function
-\pkg{dplyr}::\texttt{group\_by()} and \texttt{agent\_to\_aggregate()} to
-provide aggregation
-\sout{at a}\textcolor{orange}{by}\footnote{\textcolor{violet}{[Ben says: declined. unclear about this change. Also would it not be "by groups"? Orignial seems to flow more.]}}
-group level.
-
-The \texttt{agents\_to\_aggregate()} function is flexible and can deal
-with a wide range of information about each individual.
-\textcolor{violet}{\sout{It can,} In fact, this function can} account
-for infinitely many states. This functionality allows the practitioner
-to aggregate information relative to common states
-(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as
-more complex states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'').
-Additionally, \texttt{agents\_to\_aggregate()} permits indicators for
-death/exit and birth/entry dates. Overall, this function is a powerful
-tool for pre-processing data,
-\textcolor{orange}{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\textcolor{orange}{[Shannon says: I added this in theme with 'highlighting a point of entry']} \textcolor{violet}{[Ben says: I'm unclear about this comment but I'm find with the addition...]}}
+There are often situations where aggregating agents into subpopulations
+(e.g.~subpopulations defined by age or sex) can highlight different
+aggregate level trends. For example, research by
+\citet{rvachev1985,anderson1992,worby2015} develop state-based models
+that account for differing disease dynamics in different subpopulations.
+In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining
+the function \pkg{dplyr}::\texttt{group\_by()} and
+\texttt{agent\_to\_aggregate()} to provide aggregation at a
+subpopulation level.
+
+\textcolor{orange}{Besides aiding subpopulation analysis,}
+\texttt{agents\_to\_aggregate()} can deal with a wide range of
+information about each individual. In fact, this function account for
+infinitely many states. This functionality allows the practitioner to
+aggregate information relative to common states (e.g.~``Susceptible'',
+``Infectious'', and ``Recovered'') as well as more complex states
+(e.g.~``Exposed'', ``iMmune'', ``Hospitalized''). Additionally,
+\texttt{agents\_to\_aggregate()} permits indicators for death/exit and
+birth/entry dates. Overall, this function is a powerful tool for
+pre-processing data.
 
@@
 
-\textcolor{violet}{\sout{With raw data, "Getting to know" our} In the early stages of a project, getting to know the}\footnote{\textcolor{orange}{[Shannon suggested: "Getting to know" our]} \textcolor{violet}{[Ben says: even though I was the original one with the quotes - one should always stray away from it.
-I kept the implicit connection to the section we're in given I'm not sure people are reading the titles super well.]}}
-data currently means figuring out useful combinations of visualizations
-\textcolor{violet}{\sout{,}and} numerical summaries and
-\sout{\sout{subsets}\textcolor{orange}{groupings}}\textcolor{violet}{exploring different groupings of the data}\footnote{\textcolor{violet}{[Ben says: the action "figuring out" and "groupings" was a bit unclear what actually was going to happen - so I changed it.]}}.
-\textcolor{violet}{\sout{An expert coder has many ways to successfully explore the data in an aggregate lens using }}\texttt{agents\_to\_aggregate()}\textcolor{violet}{\sout{.
-For less experienced coders, \pkg{EpiCompare} also includes tools to rapidly explore data that has three epidemic states.} An expert coder can start from }
-\texttt{agents\_to\_aggregate()}
-\textcolor{violet}{to successfully accomplish EDA in many ways, but \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, as long as there three unique epidemiological states (like the SIR model).}\footnote{\textcolor{violet}{[Ben says: I'm reverting to the old text - it's much clearer.
-- happy to have a a discussion on it.
-The crossed out text didn't capture what an expert coder was really going to do and how that differed.]}}
-Building on \textcolor{violet}{\sout{the tools in}} \pkg{ggplot2} and
-\pkg{ggtern} \textcolor{violet}{packages}, \pkg{EpiCompare}`s
-\texttt{geom\_aggregate()} provides a rapid way to explore different
-subpopulations' experiences of an epidemic
-\citep{Wickham2016, Hamilton2018}.
-\textcolor{violet}{The function }\texttt{geom\_aggregate()}
-\textcolor{violet}{provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.}
-\footnote{\textcolor{violet}{[Original text: "By combiing the ideas behind agent2aggregate for three-state models to examine subpopulation trajcetories in 2d simplex space."]} \textcolor{orange}{[Shannons says: I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.]}\textcolor{violet}{[Ben says: I'm unclear why this warrants just deleting it.
-I've provided a newly written sentence...]}}
-Visualization tools for three-state models were developed because (1)
-SIR models are some of the most common and basic epidemic state-based
-models and (2) our three-dimensional simplex representation of these
-epidemics emphasizes a ``time-invarance'' representation of the data
-(for a refresher see Section \ref{sec:time-invariant}).
+In the early stages of a project, familiarizing oneself with the data
+usually means figuring out useful combinations of visualizations and
+numerical summaries of the data
+\textcolor{orange}{both at population and subpopulation level.} An
+expert coder can start with \texttt{agents\_to\_aggregate()} to
+successfully accomplish EDA in many ways. \pkg{EpiCompare} also includes
+tools that allow a novice coder to rapidly explore data, provided there
+three unique epidemiological states (like in the SIR model). Building on
+\pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s
+\texttt{geom\_aggregate()} provides a rapid way to explore how different
+subpopulations experience of an epidemic
+\citep{Wickham2016, Hamilton2018}. The function
+\texttt{geom\_aggregate()} provides a visualization tool to holistically
+examine aggregate level information across different subpopulations by
+visualizing each subpopulation's epidemic trajectory in the
+three-dimensional state space. Visualization tools for three-state
+models were developed because (1) SIR models are some of the most common
+and basic epidemic state-based models and (2) our three-dimensional
+simplex representation of these epidemics emphasizes a
+``time-invarance'' representation of the data (for a refresher see
+Section \ref{sec:time-invariant}).
 
@@
 
-\textcolor{violet}{[Ben says: think about this section and if it highlights that we can bring in outside models...]}
-
-After getting a good sense of what a past or current epidemic looks like
-with EDA, the next step is often model fitting and/or simulations.
-\sout{In this step and the next step (post-processing), we discuss how to \textcolor{orange}{easily include external models and simulations originating outside into the \pkg{EpiCompare} data analysis pipeline.}}\textcolor{violet}{In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}\footnote{\textcolor{violet}{[Ben says: the original rewrite didn't seem to clear - specifically in how the practitioner would be using outside models in this step.]}}
-While \pkg{EpiCompare} does not focus on fitting model(s) to data, we do
-provide some flexible functions for simulation of basic discrete-time
-epidemic-state models.
-\textcolor{violet}{These functions simulation individual-level information based on practitioner estimated transition rates between states and}
-can be combined with \texttt{agents\_to\_aggregate()} to view these
-simulations through an aggregate lens. The function
-\texttt{simulate\_SIR\_agents()} simulates a basic SIR epidemic with
-user inputs for the number of simulations, the initial number in each
-state, the infection and recovery parameters \((\beta, \gamma\)), and
-the total number of discrete time steps. This function allows for easy
-access to SIR model analysis and comparison. Beyond SIR models, the
-function \texttt{simulate\_agents()} takes as input a user-specified
-state-transition matrix and other epidemic parameters to allow the user
-to create simulations for an outbreak with \textit{any} number of states
-and any number of transitions among them. This flexibility in states can
-be used to also reflect group-based dynamics. In turn, this allows for
-users to explore the space of models in an intuitive way without getting
-bogged down by too much mathematical detail. For consistency, we have
-made output from \texttt{simulate\_agents()} and
-\texttt{simulate\_SIR\_agents()} compatible with
-\texttt{agents\_to\_aggregate()} so aggregate information may easily be
-accessed.
+After getting a sense of what a past or current epidemic looks like with
+EDA, often the next step in the data analysis pipeline is model fitting
+and/or simulation. In this step and the next step (post-proessing), we
+highlight how practitioners can pair models and simulations of epidemics
+from outside of \pkg{EpiCompare} with analysis and simulation tools in
+\pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting
+models to data, we do provide some flexible functions for simulation of
+basic discrete-time epidemic-state models. These functions simulate
+individual-level information based on practitioner estimated transition
+rates between states and can be combined with
+\texttt{agents\_to\_aggregate()} to view these simulations through an
+aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
+a basic SIR epidemic with user inputs for the number of simulations, the
+initial number in each state, the infection and recovery parameters
+\((\beta, \gamma\)), and the total number of discrete time steps. This
+function allows for easy access to SIR model analysis and comparison.
+Beyond SIR models, the function \texttt{simulate\_agents()} takes as
+input a user-specified state-transition matrix and other epidemic
+parameters to allow the user to create simulations for an outbreak with
+\textit{any} number of states and any number of transitions among them.
+This flexibility in states can be used to also reflect group-based
+dynamics. In turn, this allows for users to explore the space of models
+in an intuitive way without getting bogged down by too much mathematical
+detail. For consistency, we have made output from
+\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
+compatible with \texttt{agents\_to\_aggregate()} so aggregate
+information may easily be accessed.
 
@@
 
-\textcolor{violet}{[Ben says: I think we should remind the reader that we care more about simulations, in order to compare fitted models between themselves and the true epidemics.]}
-
-\textcolor{violet}{[Ben says: this replaces the first paragraph below]}
-
-If the practitioner wishes to compare models-to-observations or even
+If practitioners wish to compare models-to-observations or even
 models-to-models, they needs to post-process their models and
-simulations to disseminate the results in an easily digestable format.
+simulations to disseminate the results in an easily digestible format.
 In general, post-processing of modeling and simulation consists of
-making summary statistics, plots, and tables.
-\textcolor{violet}{Model \sout{The}} summaries can be very complex, and
-as a result, a number of epidemic modeling \textbackslash proglang\{R\}
-packages return \textcolor{violet}{\sout{a}} special class
-\textcolor{violet}{objects}. The special
-class\textcolor{violet}{\sout{es} objects} often contain a plethora of
-information from residuals, model diagnostics, input parameters, and
-more. While incredibly useful, these special classes can be difficult
-for novice coders to \textcolor{violet}{\sout{navigate} handle}.
+making summary statistics, plots, and tables. Model output summaries can
+be very complex, and as a result, a number of epidemic modeling
+\proglang{R} packages return a special class. The special classes often
+contain a plethora of information about residuals, model diagnostics,
+input parameters, and more. While incredibly useful, these special
+classes can be difficult for novice coders to handle.
 
@@
 
-\textcolor{orange}{[Shannon says: Here Ben talks about filament compression and tidy\_dist mats, etc.]}
-\textcolor{violet}{[Ben says: not currently thinking about including tidy\_dist stuff - will need to think about this as it is including in the pipeline 2 image...]}
-\footnote{\textcolor{violet}{[Ben says: are these comments still relevant?]}}
-
-To utilize simulations of epidemics in later time-invariant analysis we
-also provide a function to convert temporally defined epidemics to
-filamental representations. Specifically, we provide the function
+To utilize simulations of epidemics in time-invariant analysis, we also
+provide a function to convert temporally defined epidemics to filamental
+representations. Specifically, we provide the function
 \texttt{filament\_compression()} to convert simulation(s) to filaments
-as expressed by presenting the epidemic as a ordering of some common
+as expressed by presenting the epidemic as an ordering of some common
 fixed number of points so that they are equally spaced along the
-original path in the proportional state space.
-
-\textcolor{orange}{[Shannon says: Shannon's attempt at above paragraph is below..
-I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.]}
-
-\pkg{EpiCompare} also provides a tool to convert time-dependent epidemic
-simulations into their time-invariant filamental representations.
-Specifically, \texttt{filament\_compression()} converts simulations to
-filaments (see Section \ref{sec:time-invariant})
-\textcolor{violet}{\sout{so practitioners can view model simulations and results through a time-invariant lens.}which allows practitioner to later apply time-invariant tools to these compressed epidemiological objects.}\footnote{\textcolor{violet}{[Ben says: I worry about this statement - it is very broad / seems to claim a lot but not very clearly...]}}
-These tools were developed to provide another natural entry point into
-the \pkg{EpiCompare} data analysis pipeline for situations where
-modeling and analysis is already completed and practitioners are looking
-for simple and transparent tools to help understand and disseminate
-results.
+original path in the proportional state space. These tools were
+developed to provide another natural entry point into the
+\pkg{EpiCompare} data analysis pipeline for situations where modeling
+and analysis are already completed and practitioners are looking for
+simple and transparent tools to help understand and disseminate results.
 
@@
 
-\textcolor{violet}{[Ben says: I didn't look at this section - it looks like your comments are more "let's" talk about this and not suggestions / feedback.]}
-
-\textcolor{orange}{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}
-As introduced in Section \ref{sec:beyond-r0-sir} there's a
-\sout{lot of}\textcolor{orange}{much} potential for time-invariant tools
-to help compare and assess epidemics and models/simulations. In
-\pkg{EpiCompare} we provide a set of comparison and assessment tools for
-models that extend beyond the standard performance metrics (e.g.~mean
-squared error or AIC) and focus on assessing the structural information
-the models capture. This approaches work well on models where
-\sout{online} one ``cycle'' of the epidemic has occurred (no recovered
-individuals have been susceptible
-again)\footnote{\textcolor{violet}{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}\textcolor{orange}{Shannon says I don't think I understand what you are trying to say there.
-We should chat about it.}}.
-Epidemics are complex objects, and we provide tools to create prediction
-regions
-\textcolor{orange}{with differing desired characteristics}\footnote{\textcolor{orange}{Shannon says: I want to connect epidemics being complex to the fact that prediction is hard and not all prediction regions tell you the same thing}}
-from simulated epidemics. For three-state epidemic models, we provide
-the \texttt{ggplot}/\texttt{ggtern} extension
+Finally, in \pkg{EpiCompare} we provide a set of comparison and
+assessment tools for model and simulation results that extend beyond the
+standard performance metrics (e.g.~mean squared error or AIC).
+Corresponding to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare}
+provides a set of time-invariant tools to compare and evaluate epidemic
+models and simulations. We have found these tools to be specifically
+applicable for situations where only one ``season'' or ``cycle'' of the
+epidemic has occurred.
+
+One tool we provide to assess models is through the creation of
+geometric prediction regions, which are useful when we treat epidemics
+like filaments. If there is a set of simulated epidemics from a model,
+we can create a geometric prediction region for the expected trajectory
+of the epidemic in the state space. For three-state epidemic models, we
+provide the \texttt{ggplot}/\texttt{ggtern} extension
 \texttt{geom\_prediction\_band()} which creates a prediction region
-around the top \(1-\alpha\) proportion
-{[}\textcolor{orange}{good place for Ben paper cite?}{]} of the
-simulations (where the simulations treated as filaments). In
-\pkg{EpiCompare} we also provide these prediction regions for epidemic
-models with with more than three states. The functions
+around the top \(1-\alpha\) proportion of the simulations. In this
+visual setting, comparing this prediction to the true epidemic
+trajectory or comparing the prediction regions defined by two different
+models' simulations can be done with the `eye test.' In \pkg{EpiCompare}
+we also provide these prediction regions for epidemic models with with
+more than three states. The functions
 \texttt{create\_convex\_hull\_structure()} and
 \texttt{create\_delta\_ball\_structure()} create different geometric
-representations of prediction regions for any state-based model. For
-both of these geometric objects, we provide functions to check if a path
-is contained (\texttt{contained()}) and the ability to assess the
-Hausdorff distance between prediction regions based on simulations from
-different model (\texttt{hassdorf\_dist()}).
-\footnote{\textcolor{violet}{[Ben says: should we discuss the details of geom\_prediction\_bands more?
-]}\textcolor{orange}{[Shannon says: I don't think so, best left to the vignette and data example]}}
+representations of prediction regions for any dimensional state-based
+model. For both of these geometric structures, we provide functions to
+check if a path is contained (\texttt{contained()}) and the ability to
+assess the Hausdorff distance between prediction regions based on
+simulations from different model (\texttt{hausdorff\_dist()}).
 
 We also provide functions to calculate the ``extremeness'' of a true
-epidemic {[}\textcolor{orange}{trajectory?}{]} compared to simulated
-epidemics via the equi-distance filamental trajectory representation as
-mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions
-like \texttt{distance\_pseudo\_density\_function()} can calculate a
+epidemic trajectory compared to simulated epidemics via the
+equi-distance filamental trajectory representation as mentioned in
+Section \ref{sec:beyond-r0-sir}. Specifically, functions like
+\texttt{distance\_pseudo\_density\_function()} can calculate a
 pseudo-density estimate of the true epidemic relative to simulated ones.
@@
 suggest how geometrically central an epidemic is to simulations.
-\footnote{\textcolor{violet}{[Ben says: should we define these more in this paragraph?
-Introducted in 2.2.
-as well and not defined.]}\textcolor{orange}{[Shannon says: I'd throw a citation or two but again limit the detail.]}}
-
-\textcolor{violet}{[Ben says: Replaced with paragraphs above.]}
-Comparison and assessment of model fit or comparisons of one model to
-another model can be performed in a variety of ways including mean
-square error, AIC, plots, and more. Perhaps the most useful tool
-\pkg{EpiCompare} has to offer to the expert, for comparison and
-assessment of models, is in its post-processing tools which create a
-standard output. It is then a matter of writing a script or function
-made for that standard output to assess the results from multiple models
-in the way the user
-desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}\textcolor{orange}{I agree with you now that this section is nice and filled in.
-Let's drop this paragraph.}}.
-
-\textcolor{violet}{[Ben says: Replaced with paragraphs above.]} However,
-for those who like more concrete tools, \pkg{EpiCompare} offers
-functions to compare prediction regions to one another including
-\texttt{geom\_prediction\_band()} (which plots the region), and
-\texttt{create\_\{convex\_hull,delta\_ball\}\_structure()}(which returns
-the \texttt{R} output for the given structure), and \texttt{contained()}
-(which allows the user to determine if one set of points is contained in
-a prediction band). Additionally, we offer ways to determine if model
-outputs are compatible with one another, that is how extreme one output
-is to another. \textcolor{orange}{Ben says something about distance}
 

### 2021-05-12 jss/EpiCompare-paper/EpiCompare-paper.Rmd
footnotes comments on section 3 at a more high level

@@
 In Fig.
-\ref{fig:pipeline2}, we illustrate how our package's functions fit into the data analysis pipeline introduced in Fig.
+\ref{fig:pipeline2}, we \ben{\sout{illustrate} show}\footnote{\ben{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
+I'm happy with other synonyms as well.}} how our package's functions fit into the data analysis pipeline introduced in Fig.
 \ref{fig:pipeline}.
@@
 Before data can be explored, they must be collected.
-Raw data can appear in many different formats.
-Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+Raw data can appear in many different formats.\footnote{\ben{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
 Other data collections focus on aggregate counts of individuals in each epidemic state.
@@
 start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
-There are often situations where aggregating agents into subpopulations (e.g.
+There are often situations where \ben{\sout{aggregating} grouping}\footnote{\ben{[We use aggregating two times in this sentence.
+I'm open to other synonyms as well, but it seems repetitive.]}} agents into subpopulations (e.g.
 subpopulations defined by age or sex) can highlight different aggregate level trends.
 For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation at a subpopulation level.
-
-\shannon{Besides aiding subpopulation analysis,} `agents_to_aggregate()` can deal with a wide range of information about each individual. 
+In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
+]}}.
+
+Besides aiding subpopulation analysis, `agents_to_aggregate()` can \ben{\sout{deal with} accommodate}\footnote{\ben{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
+Again, I'm open to other synonyms.]}} a wide range of information about each individual.
 In fact, this function account for infinitely many states.
@@
 
-**EDA**
-
- In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data \shannon{both at population and subpopulation level.}  An expert coder can start with `agents_to_aggregate()` to successfully accomplish EDA in many ways.
-\pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there three unique epidemiological states (like in the SIR model).
-Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a rapid way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
+**\ben{Exploratory Data Analysis (}EDA\ben{)}**
+
+ In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data both at population and subpopulation level.
+An expert coder can start with `agents_to_aggregate()` to successfully accomplish \ben{exploratory data analysis (}EDA\ben{)} in many ways.
+\pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there \ben{are} three unique epidemiological states (like in the SIR model).
+Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a \ben{\sout{rapid}}\footnote{\ben{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]}} way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
 The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.
-Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a "time-invarance" representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a \ben{\sout{"time-invarance"}time-invariant}\footnote{\ben{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]}} representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-After getting a sense of what a past or current epidemic looks like with EDA, often the next step in the data analysis pipeline is model fitting and/or simulation.
-In this step and the next step (post-proessing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.
+After getting a sense of what a past or current epidemic looks like with EDA, \ben{\sout{often}} the next step in the data analysis pipeline is \ben{often}\footnote{\ben{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence.
+Maybe it's just me.]}} model fitting and/or simulation.
+In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.
 While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
@@
 
+\footnote{\ben{[Ben says: I found the structure of this section slightly confusing.
+It appears that the 2 subsections are (1) model based transformation and (2) simulation based transformations.
+The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}}
 If practitioners wish to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestible format.
@@
 To utilize simulations of epidemics in  time-invariant analysis, we also provide a function to convert temporally defined epidemics to filamental representations.
-Specifically, we provide the function `filament_compression()` to convert simulation(s) to filaments as expressed by presenting the epidemic as an ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space.
-These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis are already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.
+Specifically, we provide the function `filament_compression()` to convert simulation(s) to filamental trajectories as expressed by presenting the epidemic as an ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space\footnote{\ben{[Ben says: there's a lot going on in this sentence, and the phrase "as expressed by presenting the epidemic as an ordering of some common fixed number of points" is between the subject and our verb "so that".]}}.
+These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis are already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.\footnote{\ben{[Ben says: This sentence still doesn't seem to fit into the paragraph or section too well.
+First, given it's in this paragraph: is it just the simulation based "tools"?
+I might suggest moving it ot the first introductory sentence - but probably needs some word-smithing as well - I feel like it might be overselling the package a bit.]}}
 
@@
 mean squared error or AIC).
-Corresponding to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
-We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of the epidemic has occurred.
-
-One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
+Corresponding\footnote{\ben{[Ben says: "corresponding to" doesn't seem to be the correct phrase here.
+If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect work choice.]}} to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
+We have found these tools to be specifically applicable for situations where only one "season" or "cycle"\footnote{\ben{[Ben says: I think we can remove the paratheses here.]}} of the epidemic has occurred.
+
+\footnote{\ben{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}}One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
 If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space.
 For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
-In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done with the 'eye test.' In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
+In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done \ben{by eye.\sout{with the 'eye test.'}} In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
 The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model.
@@
 
-We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
+\footnote{\ben{[Ben says: random thought (not needed) - maybe we do provide some equations for these measures and also maybe mathematically express how we'd examine extremeness?]}}We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
 Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
@@
 
-In this section, we highlight many of the tools available in \pkg{EpiCompare}.
+\footnote{\ben{[Ben says: Shannon, would you mind reading this whole section over again once we've finished edits for section 2 and 3?
+This initial paragraph seems to be stating section 3's story.]}}In this section, we highlight many of the tools available in \pkg{EpiCompare}.
 As previously discussed, these tools include data cleaning; visualization; modeling and simulation; post-processing; and comparison and model assessment, in accordance with the data analysis pipeline (Fig.

### 2021-05-12 jss/EpiCompare-paper/EpiCompare-paper.tex
footnotes comments on section 3 at a more high level

@@
 and explain how they aid in the data analysis pipeline. In Fig.
-\ref{fig:pipeline2}, we illustrate how our package's functions fit into
-the data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
-front-facing functions in \pkg{EpiCompare} are aimed to be as
-user-friendly as possible. We also focus on providing the user
-``tidyverse'' style functions, that encourage piping objects from one
-function to the next and follow clear ``verb'' naming schemes
-\citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare}
-into any step in the data analysis pipeline, there are two primary
-points of entry. The first point of entry is the very beginning with
-pre-processing and visualizing raw data, and the second point of entry
-is after modeling and simulation. Figure \ref{fig:pipeline2} captures
-these different paths, and we highlight how to leverage \pkg{EpiCompare}
-functionalities in the subsections below.
+\ref{fig:pipeline2}, we
+\textcolor{violet}{\sout{illustrate} show}\footnote{\textcolor{violet}{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
+I'm happy with other synonyms as well.}}
+how our package's functions fit into the data analysis pipeline
+introduced in Fig. \ref{fig:pipeline}. All front-facing functions in
+\pkg{EpiCompare} are aimed to be as user-friendly as possible. We also
+focus on providing the user ``tidyverse'' style functions, that
+encourage piping objects from one function to the next and follow clear
+``verb'' naming schemes \citep{Wickham2019}. Although users can
+incorporate \pkg{EpiCompare} into any step in the data analysis
+pipeline, there are two primary points of entry. The first point of
+entry is the very beginning with pre-processing and visualizing raw
+data, and the second point of entry is after modeling and simulation.
+Figure \ref{fig:pipeline2} captures these different paths, and we
+highlight how to leverage \pkg{EpiCompare} functionalities in the
+subsections below.
 
@@
 can be explored. Before data can be explored, they must be collected.
-Raw data can appear in many different formats. Sometimes individual
-records are collected, with times of different states of the epidemic
-(infection, recovery, etc.) as well as individual information like
-network structure, location, and sub-population information. Other data
-collections focus on aggregate counts of individuals in each epidemic
-state. In fact, many times only the number of new infections at each
-time step (e.g.~weekly case counts) is observed. Compartment totals
-(amounts of individuals in each state) are then imputed from those case
-counts other information about the disease and the population of
-interest. In \pkg{EpiCompare}, we focus on understanding the overall
-impact of an outbreak at the aggregate/population level, which allows
-for streamlined examination of overall trends of an epidemic.
+Raw data can appear in many different
+formats.\footnote{\textcolor{violet}{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}}
+Sometimes individual records are collected, with times of different
+states of the epidemic (infection, recovery, etc.) as well as individual
+information like network structure, location, and sub-population
+information. Other data collections focus on aggregate counts of
+individuals in each epidemic state. In fact, many times only the number
+of new infections at each time step (e.g.~weekly case counts) is
+observed. Compartment totals (amounts of individuals in each state) are
+then imputed from those case counts other information about the disease
+and the population of interest. In \pkg{EpiCompare}, we focus on
+understanding the overall impact of an outbreak at the
+aggregate/population level, which allows for streamlined examination of
+overall trends of an epidemic.
 
@@
 aggregate view of how many individuals were in a state at a given time.
-There are often situations where aggregating agents into subpopulations
-(e.g.~subpopulations defined by age or sex) can highlight different
-aggregate level trends. For example, research by
+There are often situations where
+\textcolor{violet}{\sout{aggregating} grouping}\footnote{\textcolor{violet}{[We use aggregating two times in this sentence. I'm open to other synonyms as well, but it seems repetitive.]}}
+agents into subpopulations (e.g.~subpopulations defined by age or sex)
+can highlight different aggregate level trends. For example, research by
 \citet{rvachev1985,anderson1992,worby2015} develop state-based models
@@
 \texttt{agent\_to\_aggregate()} to provide aggregation at a
-subpopulation level.
-
-\textcolor{orange}{Besides aiding subpopulation analysis,}
-\texttt{agents\_to\_aggregate()} can deal with a wide range of
-information about each individual. In fact, this function account for
-infinitely many states. This functionality allows the practitioner to
-aggregate information relative to common states (e.g.~``Susceptible'',
-``Infectious'', and ``Recovered'') as well as more complex states
-(e.g.~``Exposed'', ``iMmune'', ``Hospitalized''). Additionally,
-\texttt{agents\_to\_aggregate()} permits indicators for death/exit and
-birth/entry dates. Overall, this function is a powerful tool for
-pre-processing data.
-
-\textbf{EDA}
+subpopulation
+level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}}.
+
+Besides aiding subpopulation analysis, \texttt{agents\_to\_aggregate()}
+can
+\textcolor{violet}{\sout{deal with} accommodate}\footnote{\textcolor{violet}{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
+Again, I'm open to other synonyms.]}}
+a wide range of information about each individual. In fact, this
+function account for infinitely many states. This functionality allows
+the practitioner to aggregate information relative to common states
+(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as
+more complex states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'').
+Additionally, \texttt{agents\_to\_aggregate()} permits indicators for
+death/exit and birth/entry dates. Overall, this function is a powerful
+tool for pre-processing data.
+
+\textbf{\textcolor{violet}{Exploratory Data Analysis (}EDA\textcolor{violet}{)}}
 
@@
 usually means figuring out useful combinations of visualizations and
-numerical summaries of the data
-\textcolor{orange}{both at population and subpopulation level.} An
-expert coder can start with \texttt{agents\_to\_aggregate()} to
-successfully accomplish EDA in many ways. \pkg{EpiCompare} also includes
-tools that allow a novice coder to rapidly explore data, provided there
+numerical summaries of the data both at population and subpopulation
+level. An expert coder can start with \texttt{agents\_to\_aggregate()}
+to successfully accomplish
+\textcolor{violet}{exploratory data analysis (}EDA\textcolor{violet}{)}
+in many ways. \pkg{EpiCompare} also includes tools that allow a novice
+coder to rapidly explore data, provided there \textcolor{violet}{are}
 three unique epidemiological states (like in the SIR model). Building on
 \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s
-\texttt{geom\_aggregate()} provides a rapid way to explore how different
-subpopulations experience of an epidemic
+\texttt{geom\_aggregate()} provides a
+\textcolor{violet}{\sout{rapid}}\footnote{\textcolor{violet}{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]}}
+way to explore how different subpopulations experience of an epidemic
 \citep{Wickham2016, Hamilton2018}. The function
@@
 simplex representation of these epidemics emphasizes a
-``time-invarance'' representation of the data (for a refresher see
-Section \ref{sec:time-invariant}).
+\textcolor{violet}{\sout{"time-invarance"}time-invariant}\footnote{\textcolor{violet}{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]}}
+representation of the data (for a refresher see Section
+\ref{sec:time-invariant}).
 
@@
 After getting a sense of what a past or current epidemic looks like with
-EDA, often the next step in the data analysis pipeline is model fitting
-and/or simulation. In this step and the next step (post-proessing), we
-highlight how practitioners can pair models and simulations of epidemics
-from outside of \pkg{EpiCompare} with analysis and simulation tools in
-\pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting
-models to data, we do provide some flexible functions for simulation of
-basic discrete-time epidemic-state models. These functions simulate
-individual-level information based on practitioner estimated transition
-rates between states and can be combined with
-\texttt{agents\_to\_aggregate()} to view these simulations through an
-aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
-a basic SIR epidemic with user inputs for the number of simulations, the
-initial number in each state, the infection and recovery parameters
-\((\beta, \gamma\)), and the total number of discrete time steps. This
-function allows for easy access to SIR model analysis and comparison.
-Beyond SIR models, the function \texttt{simulate\_agents()} takes as
-input a user-specified state-transition matrix and other epidemic
-parameters to allow the user to create simulations for an outbreak with
-\textit{any} number of states and any number of transitions among them.
-This flexibility in states can be used to also reflect group-based
-dynamics. In turn, this allows for users to explore the space of models
-in an intuitive way without getting bogged down by too much mathematical
-detail. For consistency, we have made output from
-\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
-compatible with \texttt{agents\_to\_aggregate()} so aggregate
-information may easily be accessed.
+EDA, \textcolor{violet}{\sout{often}} the next step in the data analysis
+pipeline is
+\textcolor{violet}{often}\footnote{\textcolor{violet}{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence. Maybe it's just me.]}}
+model fitting and/or simulation. In this step and the next step
+(post-processing), we highlight how practitioners can pair models and
+simulations of epidemics from outside of \pkg{EpiCompare} with analysis
+and simulation tools in \pkg{EpiCompare}. While \pkg{EpiCompare} does
+not focus on fitting models to data, we do provide some flexible
+functions for simulation of basic discrete-time epidemic-state models.
+These functions simulate individual-level information based on
+practitioner estimated transition rates between states and can be
+combined with \texttt{agents\_to\_aggregate()} to view these simulations
+through an aggregate lens. The function \texttt{simulate\_SIR\_agents()}
+simulates a basic SIR epidemic with user inputs for the number of
+simulations, the initial number in each state, the infection and
+recovery parameters \((\beta, \gamma\)), and the total number of
+discrete time steps. This function allows for easy access to SIR model
+analysis and comparison. Beyond SIR models, the function
+\texttt{simulate\_agents()} takes as input a user-specified
+state-transition matrix and other epidemic parameters to allow the user
+to create simulations for an outbreak with \textit{any} number of states
+and any number of transitions among them. This flexibility in states can
+be used to also reflect group-based dynamics. In turn, this allows for
+users to explore the space of models in an intuitive way without getting
+bogged down by too much mathematical detail. For consistency, we have
+made output from \texttt{simulate\_agents()} and
+\texttt{simulate\_SIR\_agents()} compatible with
+\texttt{agents\_to\_aggregate()} so aggregate information may easily be
+accessed.
 
@@
 
+\footnote{\textcolor{violet}{[Ben says: I found the structure of this section slightly confusing.
+It appears that the 2 subsections are (1) model based transformation and (2) simulation based transformations.
+The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}}
 If practitioners wish to compare models-to-observations or even
@@
 representations. Specifically, we provide the function
-\texttt{filament\_compression()} to convert simulation(s) to filaments
-as expressed by presenting the epidemic as an ordering of some common
-fixed number of points so that they are equally spaced along the
-original path in the proportional state space. These tools were
-developed to provide another natural entry point into the
-\pkg{EpiCompare} data analysis pipeline for situations where modeling
-and analysis are already completed and practitioners are looking for
-simple and transparent tools to help understand and disseminate results.
+\texttt{filament\_compression()} to convert simulation(s) to filamental
+trajectories as expressed by presenting the epidemic as an ordering of
+some common fixed number of points so that they are equally spaced along
+the original path in the proportional state
+space\footnote{\textcolor{violet}{[Ben says: there's a lot going on in this sentence, and the phrase "as expressed by presenting the epidemic as an ordering of some common fixed number of points" is between the subject and our verb "so that".]}}.
+These tools were developed to provide another natural entry point into
+the \pkg{EpiCompare} data analysis pipeline for situations where
+modeling and analysis are already completed and practitioners are
+looking for simple and transparent tools to help understand and
+disseminate
+results.\footnote{\textcolor{violet}{[Ben says: This sentence still doesn't seem to fit into the paragraph or section too well.
+First, given it's in this paragraph: is it just the simulation based "tools"?
+I might suggest moving it ot the first introductory sentence - but probably needs some word-smithing as well - I feel like it might be overselling the package a bit.]}}
 
@@
 standard performance metrics (e.g.~mean squared error or AIC).
-Corresponding to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare}
-provides a set of time-invariant tools to compare and evaluate epidemic
-models and simulations. We have found these tools to be specifically
-applicable for situations where only one ``season'' or ``cycle'' of the
-epidemic has occurred.
-
-One tool we provide to assess models is through the creation of
-geometric prediction regions, which are useful when we treat epidemics
-like filaments. If there is a set of simulated epidemics from a model,
-we can create a geometric prediction region for the expected trajectory
-of the epidemic in the state space. For three-state epidemic models, we
-provide the \texttt{ggplot}/\texttt{ggtern} extension
+Corresponding\footnote{\textcolor{violet}{[Ben says: "corresponding to" doesn't seem to be the correct phrase here.
+If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect work choice.]}}
+to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of
+time-invariant tools to compare and evaluate epidemic models and
+simulations. We have found these tools to be specifically applicable for
+situations where only one ``season'' or
+``cycle''\footnote{\textcolor{violet}{[Ben says: I think we can remove the paratheses here.]}}
+of the epidemic has occurred.
+
+\footnote{\textcolor{violet}{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}}One
+tool we provide to assess models is through the creation of geometric
+prediction regions, which are useful when we treat epidemics like
+filaments. If there is a set of simulated epidemics from a model, we can
+create a geometric prediction region for the expected trajectory of the
+epidemic in the state space. For three-state epidemic models, we provide
+the \texttt{ggplot}/\texttt{ggtern} extension
 \texttt{geom\_prediction\_band()} which creates a prediction region
@@
 trajectory or comparing the prediction regions defined by two different
-models' simulations can be done with the `eye test.' In \pkg{EpiCompare}
-we also provide these prediction regions for epidemic models with with
-more than three states. The functions
+models' simulations can be done
+\textcolor{violet}{by eye.\sout{with the 'eye test.'}} In
+\pkg{EpiCompare} we also provide these prediction regions for epidemic
+models with with more than three states. The functions
 \texttt{create\_convex\_hull\_structure()} and
@@
 
-We also provide functions to calculate the ``extremeness'' of a true
+\footnote{\textcolor{violet}{[Ben says: random thought (not needed) - maybe we do provide some equations for these measures and also maybe mathematically express how we'd examine extremeness?]}}We
+also provide functions to calculate the ``extremeness'' of a true
 epidemic trajectory compared to simulated epidemics via the
@@
 
-In this section, we highlight many of the tools available in
+\footnote{\textcolor{violet}{[Ben says: Shannon, would you mind reading this whole section over again once we've finished edits for section 2 and 3?
+This initial paragraph seems to be stating section 3's story.]}}In
+this section, we highlight many of the tools available in
 \pkg{EpiCompare}. As previously discussed, these tools include data

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Added slight edits to section 2 (and fixed typos) and added comments for section 3.

@@
 Before data can be explored, they must be collected.
-Raw data can appear in many different formats.\footnote{\ben{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+Raw data can appear in many different formats.\footnote{\ben{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\shannon{I think we can get rid of the 'raw data can appear...' sentence.}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
 Other data collections focus on aggregate counts of individuals in each epidemic state.
@@
 There are often situations where \ben{\sout{aggregating} grouping}\footnote{\ben{[We use aggregating two times in this sentence.
-I'm open to other synonyms as well, but it seems repetitive.]}} agents into subpopulations (e.g.
+I'm open to other synonyms as well, but it seems repetitive.
+]} \shannon{It was grouping before ;).
+Fine with me.} agents into subpopulations (e.g.
 subpopulations defined by age or sex) can highlight different aggregate level trends.
@@
 In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
-]}}.
+]}\shannon{agreed.
+I like 'to provide summaries at a'}}.
 
 Besides aiding subpopulation analysis, `agents_to_aggregate()` can \ben{\sout{deal with} accommodate}\footnote{\ben{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
-Again, I'm open to other synonyms.]}} a wide range of information about each individual.
-In fact, this function account for infinitely many states.
+Again, I'm open to other synonyms.]}\shannon{I like accomodate}} a wide range of information about each individual.
+In fact, this function can account for infinitely many states.
 This functionality allows the practitioner to aggregate information relative to common states (e.g.
@@
 \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there \ben{are} three unique epidemiological states (like in the SIR model).
-Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a \ben{\sout{rapid}}\footnote{\ben{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]}} way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
+Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a \ben{\sout{rapid}}\footnote{\ben{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \shannon{accepted}} way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
 The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.
-Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a \ben{\sout{"time-invarance"}time-invariant}\footnote{\ben{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]}} representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a \ben{\sout{"time-invarance"}time-invariant}\footnote{\ben{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \shannon{agreed}} representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 It appears that the 2 subsections are (1) model based transformation and (2) simulation based transformations.
-The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}}
+The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}\shannon{agreed that this section needs some work.
+but to me those aren't the clear points, it seems like (1) put external model results in the same format and (2) transform model output to filaments.
+that said i don't see how we're over-selling the package.
+would 'transform and disseminate' be better?}}
 If practitioners wish to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestible format.
@@
 Corresponding\footnote{\ben{[Ben says: "corresponding to" doesn't seem to be the correct phrase here.
-If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect work choice.]}} to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
+If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect word choice.]}\shannon{"To support the ideas presented in Sec."?}} to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
 We have found these tools to be specifically applicable for situations where only one "season" or "cycle"\footnote{\ben{[Ben says: I think we can remove the paratheses here.]}} of the epidemic has occurred.
 
-\footnote{\ben{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}}One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
+\footnote{\ben{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\shannon{I'm not saying 'no' but I'd do it in an appendix and not the main text.
+I don't think equations belong in Section 3.}}One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
 If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space.

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper.tex
Added slight edits to section 2 (and fixed typos) and added comments for section 3.

@@
 Raw data can appear in many different
-formats.\footnote{\textcolor{violet}{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}}
+formats.\footnote{\textcolor{violet}{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\textcolor{orange}{I think we can get rid of the 'raw data can appear...' sentence.}}
 Sometimes individual records are collected, with times of different
@@
 There are often situations where
-\textcolor{violet}{\sout{aggregating} grouping}\footnote{\textcolor{violet}{[We use aggregating two times in this sentence. I'm open to other synonyms as well, but it seems repetitive.]}}
-agents into subpopulations (e.g.~subpopulations defined by age or sex)
-can highlight different aggregate level trends. For example, research by
+\textcolor{violet}{\sout{aggregating} grouping}\textbackslash footnote\{\textcolor{violet}{[We use aggregating two times in this sentence.
+I'm open to other synonyms as well, but it seems repetitive.
+]}
+\textcolor{orange}{It was grouping before ;).  Fine with me.} agents
+into subpopulations (e.g.~subpopulations defined by age or sex) can
+highlight different aggregate level trends. For example, research by
 \citet{rvachev1985,anderson1992,worby2015} develop state-based models
@@
 subpopulation
-level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}}.
+level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed.  I like 'to provide summaries at a'}}.
 
@@
 \textcolor{violet}{\sout{deal with} accommodate}\footnote{\textcolor{violet}{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
-Again, I'm open to other synonyms.]}}
+Again, I'm open to other synonyms.]}\textcolor{orange}{I like accomodate}}
 a wide range of information about each individual. In fact, this
-function account for infinitely many states. This functionality allows
-the practitioner to aggregate information relative to common states
-(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as
-more complex states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'').
+function can account for infinitely many states. This functionality
+allows the practitioner to aggregate information relative to common
+states (e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well
+as more complex states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'').
 Additionally, \texttt{agents\_to\_aggregate()} permits indicators for
@@
 \texttt{geom\_aggregate()} provides a
-\textcolor{violet}{\sout{rapid}}\footnote{\textcolor{violet}{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]}}
+\textcolor{violet}{\sout{rapid}}\footnote{\textcolor{violet}{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \textcolor{orange}{accepted}}
 way to explore how different subpopulations experience of an epidemic
@@
 simplex representation of these epidemics emphasizes a
-\textcolor{violet}{\sout{"time-invarance"}time-invariant}\footnote{\textcolor{violet}{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]}}
+\textcolor{violet}{\sout{"time-invarance"}time-invariant}\footnote{\textcolor{violet}{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \textcolor{orange}{agreed}}
 representation of the data (for a refresher see Section
@@
 It appears that the 2 subsections are (1) model based transformation and (2) simulation based transformations.
-The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}}
+The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}\textcolor{orange}{agreed that this section needs some work.
+but to me those aren't the clear points, it seems like (1) put external model results in the same format and (2) transform model output to filaments.
+that said i don't see how we're over-selling the package.
+would 'transform and disseminate' be better?}}
 If practitioners wish to compare models-to-observations or even
@@
 Corresponding\footnote{\textcolor{violet}{[Ben says: "corresponding to" doesn't seem to be the correct phrase here.
-If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect work choice.]}}
+If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect word choice.]}\textcolor{orange}{"To support the ideas presented in Sec."?}}
 to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of
@@
 
-\footnote{\textcolor{violet}{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}}One
+\footnote{\textcolor{violet}{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\textcolor{orange}{I'm not saying 'no' but I'd do it in an appendix and not the main text.
+I don't think equations belong in Section 3.}}One
 tool we provide to assess models is through the creation of geometric

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
Added slight edits to section 2 (and fixed typos) and added comments for section 3.

@@
 
-[EpiCompare emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis.
+EpiCompare emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis.
 In this section we will highlight weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take.
@@
 
-A lot of epidemiological research is interested in understanding intrinsic properties of the epidemic, not depending upon exactly when the epidemic occurred or how frequently data was collected.
-One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$, a time-invariant value that is [defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population.] Interestingly, \citet{Gallagher2020} has shown that the estimation of $R_0$  [can be sensitive to time-dependent parameters] like estimation of [the beginning and end of an epidemic].
-This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable properties of an epidemic.
+Epidemiological research is often interested in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected.
+One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.
+This intrinsic property\footnote{added to shorten this long sentence and connect that it is an intrinsic property} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}.
+\citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic.
+This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, intrinsic properties of an epidemic.
 
-In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproducible number.
-One common time-dependent tool is a series of time series line plots that track the proportion of the population in a given state (e.g.
+In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproduction number.
+One common time-dependent tool \shannon{used to better understand epidemics} is a series of time series line plots that track the proportion of the population in a given state (e.g.
 infected) at a given time.
-Figure \ref{fig:different-scales-standard} visualizes two different simulated epidemics with population states (s)usceptible, (i)nfectious and (r)ecovered.
-These simulations are actually generated under a discrete approximation of  \citet{Kermack1927}’s SIR model.
-[This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,]
+Figure \ref{fig:different-scales-standard} visualizes two different simulated epidemics with population states (S)usceptible, (I)nfectious and (R)ecovered.
+These simulations have been generated under a discrete approximation of  \citet{Kermack1927}’s SIR model.
+The Kermack and McKendrick model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
 
@@
 
-[From this model, the reproduction number is the ratio of the infection rate to the recovery rate, $R_0 = \beta/\gamma$.]
+From this model, the reproduction number is the ratio of the infection rate to the recovery rate, $R_0 = \beta/\gamma$.
 
@@
 Even though Figure \ref{fig:different-scales-standard}’s visual analysis might be able to present more complex properties of the epidemic, one might wish to understand how these two simulated epidemics’ $R_0$ values compare.
-Even though these two simulated epidemics appear different in these figures ([including having different infection peaks]), there is no real intuitive way to compare the epidemic’s $R_0$.
-In fact, both of these simulated epidemics started out the same population (1000 people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$).
-Even with simple generative models, this time-dependent visualization tools is not able to be used to compare a very simple but powerful numerical summary.
+Even though these two simulated epidemics appear different in these figures (including having different infection peaks), there is no real intuitive way to compare the epidemics' $R_0$ values.
+In fact, both of these simulated epidemics started with the same population (1000 total people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$).
+Even for simple generative models, this time-dependent visualization cannot be used to compare a very simple but powerful numerical summary, $R_0$.
 
@@
 
-The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state could be examined at a given time.
-A more holistic approach to visualizing the overall path of our simulated epidemics would be to examine each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space.
+The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state can be examined at a given time.
+A more holistic approach to visualizing the overall path of our simulated epidemics is to examine how each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space.
 Figure 3’s left plot does just that, presenting the trajectories of the simulated epidemics in this three-dimensional space.
-[For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot.] In this both of Figure  \ref{fig:different-scales-tern}’s plots we an observe that the epidemics are on the same trajectory.
-In this simple generative example setting, [this indicates that \ben{the} two epidemics have the same $R_0$ value].
+For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimensional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot.
+In this both of Figure  \ref{fig:different-scales-tern}’s plots we observe that the two epidemics are on the same trajectory in this three-dimensional space.
+In this simple generative example setting, being on the same trajectory indicates that \ben{the} two epidemics have the same $R_0$ value.
 This can be proven mathematically.
@@
 \begin{theorem}\label{thm:sir-scale}
- Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (aka $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ and define $a > 0$ such that $\beta_2 = a\beta_1$.
-[Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
-Moreover, $s = \frac{1}{a}t$.]
+ Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (i.e.
+$R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$) and define $a > 0$ such that $\beta_2 = a\beta_1$.
+Then for all $t > 0$ there exists $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
+Moreover, $s = \frac{1}{a}t$.
 \end{theorem}
 
-[The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model can be mapped to a point on the epidemic path of the second SIR model.
-In other words, the two epidemics form the same filamental trajectory.]
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$, then every point on the epidemic path of the first SIR model can be mapped to a point on the epidemic path of the second SIR model.
+In other words, the two epidemics form the same filamental trajectory.
 
@@
 
-[Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us  better compare the overall structure of the epidemic and see how the population is directly impacted.] In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states.
+Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space (i.e.
+through a time-invariant lens) allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted.
+In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states.
 These tools leverage the idea that an epidemic can be viewed as a trajectory and that many properties of the epidemic are well captured when we do so.
-This approach is useful when the epidemic of interest has only gone [through a single realization of its outbreak (before the population of individuals become susceptible again)].
+This approach is useful when the epidemic of interest has only gone through a single realization of its outbreak (before the population of individuals become susceptible again).
 
-The first set of tools allows a practioner to define distances between epidemics whose time features don’t well align.
+The first set of tools allows a practioner to define distances between epidemics whose time features do not align.
 For completed epidemics, one way to better examine their properties is to represent their filamental trajectories as a finite sequence of equally spaced points.
-[This representation induces a natural distance \ben{\sout{between this type of representation}} between epidemics, specifically:]
+This representation induces a natural distance between epidemics, specifically:
 
@@
 \]
-where $\tilde{\psi}_i(s)$ the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$.
+where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$, where $|\psi_i|$ is the length of $\psi_i$.
 This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
-These extremeness estimators can be useful when comparing the true epidemic to a a set of simulated epidemics,] and practitioners can interpret the univariate epidemic's extremeness score relative to the extremeness scores of the simulations very easily.
+These extremeness estimators can be useful when comparing the true epidemic to a a set of simulated epidemics, and practitioners can interpret the epidemic's extremeness score relative to the extremeness scores of the simulations very easily.
 
-When the goal is to predict the future impact / trajectory of a epidemic, time-invariant tools can help describe state-space regions in which we expect the true epidemic to traverse.
-In settings where the epidemic only goes a single outbreak, these regions can be very telling if simulation models well capture the epidemic’s structure.
-In \pkg{EpiCompare} we create geometric prediction regions around all but the $\alpha$ proportion of most extreme simulated epidemics.
+\sout{When the goal is to predict the future impact / trajectory of a epidemic,} In addition to being used to better understand completed epidemics, time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse.
+In settings where the epidemic only generates a single outbreak, these regions can be very telling if simulation models capture the epidemic’s structure.
+In \pkg{EpiCompare} we create geometric prediction regions around all but the $\alpha$ proportion of most extreme simulated trajectories.
 These geometric regions can also be used to compared simulation models that have different time scales, parameters and even different statistical philosophies, through set different distances.

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.tex
Added slight edits to section 2 (and fixed typos) and added comments for section 3.

@@
 
-{[}EpiCompare emphasizes the value of analyzing epidemics in a
+EpiCompare emphasizes the value of analyzing epidemics in a
 \textit{time-invariant} way - approaches that remove some or all of the
@@
 
-A lot of epidemiological research is interested in understanding
-intrinsic properties of the epidemic, not depending upon exactly when
-the epidemic occurred or how frequently data was collected. One of the
-most famous numerical summaries of an epidemic is the reproduction
-number - \(R_0\), a time-invariant value that is {[}defined as the
-expected number of infections caused by a single infect or who is added
-to a completely susceptible population.{]} Interestingly,
-\citet{Gallagher2020} has shown that the estimation of \(R_0\) {[}can be
-sensitive to time-dependent parameters{]} like estimation of {[}the
-beginning and end of an epidemic{]}. This time-dependent sensitivity in
-estimating \(R_0\) reflects a general problem in epidemiology
-surrounding how to transform time-dependent information into insights
-into desirable properties of an epidemic.
+Epidemiological research is often interested in understanding intrinsic
+properties of the epidemic, independent from when the epidemic occurred
+or how frequently data was collected. One of the most famous numerical
+summaries of an epidemic is the reproduction number - \(R_0\). This
+intrinsic
+property\footnote{added to shorten this long sentence and connect that it is an intrinsic property}
+is a time-invariant value that is defined as the expected number of
+infections caused by a single infect or who is added to a completely
+susceptible population \citep{anderson1992}. \citet{Gallagher2020}
+showed that the estimation of \(R_0\) can be sensitive to time-dependent
+parameters like estimation of the beginning and end of an epidemic. This
+time-dependent sensitivity in estimating \(R_0\) reflects a general
+problem in epidemiology surrounding how to transform time-dependent
+information into insights into desirable, intrinsic properties of an
+epidemic.
 
 In many situations, epidemiologists want a much deeper understand of an
-epidemic than just the reproducible number. One common time-dependent
-tool is a series of time series line plots that track the proportion of
-the population in a given state (e.g.~infected) at a given time. Figure
+epidemic than just the reproduction number. One common time-dependent
+tool \textcolor{orange}{used to better understand epidemics} is a series
+of time series line plots that track the proportion of the population in
+a given state (e.g.~infected) at a given time. Figure
 \ref{fig:different-scales-standard} visualizes two different simulated
-epidemics with population states (s)usceptible, (i)nfectious and
-(r)ecovered. These simulations are actually generated under a discrete
-approximation of \citet{Kermack1927}'s SIR model. {[}This model captures
-the transitions from one state to the next as a system of ordinary
-differential equations, where \(N\) is the total number of individuals,
-\(\beta\) is the rate of infection, and \(\gamma\) is the rate of
-recovery,{]}
+epidemics with population states (S)usceptible, (I)nfectious and
+(R)ecovered. These simulations have been generated under a discrete
+approximation of \citet{Kermack1927}'s SIR model. The Kermack and
+McKendrick model captures the transitions from one state to the next as
+a system of ordinary differential equations, where \(N\) is the total
+number of individuals, \(\beta\) is the rate of infection, and
+\(\gamma\) is the rate of recovery,
 
@@
 
-{[}From this model, the reproduction number is the ratio of the
-infection rate to the recovery rate, \(R_0 = \beta/\gamma\).{]}
+From this model, the reproduction number is the ratio of the infection
+rate to the recovery rate, \(R_0 = \beta/\gamma\).
 
@@
 values compare. Even though these two simulated epidemics appear
-different in these figures ({[}including having different infection
-peaks{]}), there is no real intuitive way to compare the epidemic's
-\(R_0\). In fact, both of these simulated epidemics started out the same
-population (1000 people with 10 infected) and have the same \(R_0\)
-value, just with slightly different infection and recovery rates
+different in these figures (including having different infection peaks),
+there is no real intuitive way to compare the epidemics' \(R_0\) values.
+In fact, both of these simulated epidemics started with the same
+population (1000 total people with 10 infected) and have the same
+\(R_0\) value, just with slightly different infection and recovery rates
 (\(\beta_1, \gamma_1 = 0.8, 0.4\), and
-\(\beta_2, \gamma_2 = .64, .32\)). Even with simple generative models,
-this time-dependent visualization tools is not able to be used to
-compare a very simple but powerful numerical summary.
+\(\beta_2, \gamma_2 = .64, .32\)). Even for simple generative models,
+this time-dependent visualization cannot be used to compare a very
+simple but powerful numerical summary, \(R_0\).
 
@@
 presented epidemics' trajectory data so that only the marginal
-information for each state could be examined at a given time. A more
+information for each state can be examined at a given time. A more
 holistic approach to visualizing the overall path of our simulated
-epidemics would be to examine each epidemic traverses the
+epidemics is to examine how each epidemic traverses the
 three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
 just that, presenting the trajectories of the simulated epidemics in
-this three-dimensional space. {[}For state space models like in our
+this three-dimensional space. For state space models like in our
 example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
 to \(N\) (the total population size) we can visual these point in a a
-two-dimesional \textit{ternary} plot, as seen in Figure
-\ref{fig:different-scales-tern}'s right plot.{]} In this both of Figure
-\ref{fig:different-scales-tern}'s plots we an observe that the epidemics
-are on the same trajectory. In this simple generative example setting,
-{[}this indicates that \textcolor{violet}{the} two epidemics have the
-same \(R_0\) value{]}. This can be proven mathematically.
+two-dimensional \textit{ternary} plot, as seen in Figure
+\ref{fig:different-scales-tern}'s right plot. In this both of Figure
+\ref{fig:different-scales-tern}'s plots we observe that the two
+epidemics are on the same trajectory in this three-dimensional space. In
+this simple generative example setting, being on the same trajectory
+indicates that \textcolor{violet}{the} two epidemics have the same
+\(R_0\) value. This can be proven mathematically.
 
@@
 \begin{theorem}\label{thm:sir-scale}
- Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (aka $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ and define $a > 0$ such that $\beta_2 = a\beta_1$.
-[Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
-Moreover, $s = \frac{1}{a}t$.]
+ Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (i.e.
+$R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$) and define $a > 0$ such that $\beta_2 = a\beta_1$.
+Then for all $t > 0$ there exists $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
+Moreover, $s = \frac{1}{a}t$.
 \end{theorem}
 
-{[}The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
+The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
 result from \cite{Harko2014} and is shown in detail in Proof
@@
 for two SIR models that have the same initial percent of individuals in
-each state and \(R_0\) then for every point on the epidemic path of the
+each state and \(R_0\), then every point on the epidemic path of the
 first SIR model can be mapped to a point on the epidemic path of the
 second SIR model. In other words, the two epidemics form the same
-filamental trajectory.{]}
+filamental trajectory.
 
@@
 
-{[}Through the \(R_0\) example, we see that treating epidemics like
-filamental trajectories embedded in a lower dimensional space allows us
-better compare the overall structure of the epidemic and see how the
-population is directly impacted.{]} In this section we present
-time-invariant tools that can be applied to complex epidemics where the
-epidemic's generative process is unknown and can have more than three
-states. These tools leverage the idea that an epidemic can be viewed as
-a trajectory and that many properties of the epidemic are well captured
-when we do so. This approach is useful when the epidemic of interest has
-only gone {[}through a single realization of its outbreak (before the
-population of individuals become susceptible again){]}.
+Through the \(R_0\) example, we see that treating epidemics like
+filamental trajectories embedded in a lower dimensional space
+(i.e.~through a time-invariant lens) allows us to better compare the
+overall structure of the epidemic and see how the population is directly
+impacted. In this section we present time-invariant tools that can be
+applied to complex epidemics where the epidemic's generative process is
+unknown and can have more than three states. These tools leverage the
+idea that an epidemic can be viewed as a trajectory and that many
+properties of the epidemic are well captured when we do so. This
+approach is useful when the epidemic of interest has only gone through a
+single realization of its outbreak (before the population of individuals
+become susceptible again).
 
 The first set of tools allows a practioner to define distances between
-epidemics whose time features don't well align. For completed epidemics,
-one way to better examine their properties is to represent their
-filamental trajectories as a finite sequence of equally spaced points.
-{[}This representation induces a natural distance
-\textcolor{violet}{\sout{between this type of representation}} between
-epidemics, specifically:{]}
+epidemics whose time features do not align. For completed epidemics, one
+way to better examine their properties is to represent their filamental
+trajectories as a finite sequence of equally spaced points. This
+representation induces a natural distance between epidemics,
+specifically:
 
@@
 d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
-\] where \(\tilde{\psi}_i(s)\) the point along \(\psi_i\) that is
-\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\). This
-distance is naturally time-invariant, and can be plugged into multiple
-distance-based assessment tools to examine the overall ``extremeness''
-of points, including pseudo-density estimators and depth/local depth
-functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. These
-extremeness estimators can be useful when comparing the true epidemic to
-a a set of simulated epidemics,{]} and practitioners can interpret the
-univariate epidemic's extremeness score relative to the extremeness
-scores of the simulations very easily.
-
-When the goal is to predict the future impact / trajectory of a
-epidemic, time-invariant tools can help describe state-space regions in
-which we expect the true epidemic to traverse. In settings where the
-epidemic only goes a single outbreak, these regions can be very telling
-if simulation models well capture the epidemic's structure. In
-\pkg{EpiCompare} we create geometric prediction regions around all but
-the \(\alpha\) proportion of most extreme simulated epidemics. These
-geometric regions can also be used to compared simulation models that
-have different time scales, parameters and even different statistical
-philosophies, through set different distances. Although visualization is
-the easiest when the epidemic has three states, prediction regions can
-be useful to assess and compare simulation models where the epidemics
-have multiple states.
+\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
+\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\), where
+\(|\psi_i|\) is the length of \(\psi_i\). This distance is naturally
+time-invariant, and can be plugged into multiple distance-based
+assessment tools to examine the overall ``extremeness'\,' of points,
+including pseudo-density estimators and depth/local depth functions
+\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
+estimators can be useful when comparing the true epidemic to a a set of
+simulated epidemics, and practitioners can interpret the epidemic's
+extremeness score relative to the extremeness scores of the simulations
+very easily.
+
+\sout{When the goal is to predict the future impact / trajectory of a epidemic,}
+In addition to being used to better understand completed epidemics,
+time-invariant tools can be used to aid prediction of future epidemics
+by providing a state-space regions in which we expect the true epidemic
+to traverse. In settings where the epidemic only generates a single
+outbreak, these regions can be very telling if simulation models capture
+the epidemic's structure. In \pkg{EpiCompare} we create geometric
+prediction regions around all but the \(\alpha\) proportion of most
+extreme simulated trajectories. These geometric regions can also be used
+to compared simulation models that have different time scales,
+parameters and even different statistical philosophies, through set
+different distances. Although visualization is the easiest when the
+epidemic has three states, prediction regions can be useful to assess
+and compare simulation models where the epidemics have multiple states.
 
@@
 \midrule
-Simple SIR & 0.0117621 & 0.01\\
-EpiModel SIR & 0.0356503 & 0.01\\
+Simple SIR & 0.0036733 & 0.00\\
+EpiModel SIR & 0.0149686 & 0.02\\
 \bottomrule

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
highlights

@@
 
-EpiCompare emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis.
-In this section we will highlight weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take.
-To accomplish this goal we will first demonstrate the inability of time-dependent tools to well quantify a classic epidemic parameter, the reproduction number $R_0$. 
-Then we will motivate new time-invariant approaches for more complex situations where $R_0$ doesn’t capture the complexities of outbreaks.
-This will lead us to our final subsection that defines ways to view epidemics in a time-invariant lense and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.
+\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis.
+In this section we highlight some weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take.
+To accomplish this goal we  first demonstrate the inability of time-dependent tools to \sout{well quantify}\shannon{simply capture} a classic epidemic parameter, the reproduction number $R_0$. 
+Then we motivate new time-invariant approaches for more complex situations where $R_0$ does not capture the complexities of outbreaks.
+This leads to the final subsection that defines ways to view epidemics in a time-invariant lens and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.
 
@@
 
-Epidemiological research is often interested in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected.
+Epidemiological research is \shannon{often interested}\footnote{\shannon{I used often intsead of "a lot of epidemiological research" which sounds too colloquial}} in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected.
 One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.
-This intrinsic property\footnote{added to shorten this long sentence and connect that it is an intrinsic property} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}.
+\shannon{} This intrinsic property}\footnote{\shannon{added to shorten this long sentence and connect that it is an intrinsic property}} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}.
 \citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic.
-This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, intrinsic properties of an epidemic.
+This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, \shannon{intrinsic} properties of an epidemic.
 
@@
 These simulations have been generated under a discrete approximation of  \citet{Kermack1927}’s SIR model.
-The Kermack and McKendrick model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
+j\shannon{The Kermack and McKendrick} model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
 
@@
 In fact, both of these simulated epidemics started with the same population (1000 total people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$).
-Even for simple generative models, this time-dependent visualization cannot be used to compare a very simple but powerful numerical summary, $R_0$.
+Even for simple generative models, this time-dependent visualization cannot be used to compare a very simple but powerful numerical summary, \shannon{$R_0$}.
 
@@
 The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state can be examined at a given time.
-A more holistic approach to visualizing the overall path of our simulated epidemics is to examine how each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space.
+A more holistic approach to visualizing the overall path of our simulated epidemics is to examine \shannon{how} each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space.
 Figure 3’s left plot does just that, presenting the trajectories of the simulated epidemics in this three-dimensional space.
@@
 In this both of Figure  \ref{fig:different-scales-tern}’s plots we observe that the two epidemics are on the same trajectory in this three-dimensional space.
-In this simple generative example setting, being on the same trajectory indicates that \ben{the} two epidemics have the same $R_0$ value.
+In this simple generative example setting, \shannon{being on the same trajectory} indicates that \ben{the} two epidemics have the same $R_0$ value.
 This can be proven mathematically.
@@
 
-\sout{When the goal is to predict the future impact / trajectory of a epidemic,} In addition to being used to better understand completed epidemics, time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse.
+\sout{When the goal is to predict the future impact / trajectory of a epidemic,} \shannon{In addition to being used to better understand completed epidemics,} time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse.
 In settings where the epidemic only generates a single outbreak, these regions can be very telling if simulation models capture the epidemic’s structure.

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.tex
highlights

@@
 
-EpiCompare emphasizes the value of analyzing epidemics in a
+\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a
 \textit{time-invariant} way - approaches that remove some or all of the
 impact of start/end times and recording time scales when preforming the
-analysis. In this section we will highlight weaknesses of
+analysis. In this section we highlight some weaknesses of
 time-\textit{de}pendent analysis and define the mathematical
 underpinning of the time-invariant approach we take. To accomplish this
-goal we will first demonstrate the inability of time-dependent tools to
-well quantify a classic epidemic parameter, the reproduction number
-\(R_0\). Then we will motivate new time-invariant approaches for more
-complex situations where \(R_0\) doesn't capture the complexities of
-outbreaks. This will lead us to our final subsection that defines ways
-to view epidemics in a time-invariant lense and discusses natural
-properties of these representations that allow for clearer ways to
-compare models and epidemics.
+goal we first demonstrate the inability of time-dependent tools to
+\sout{well quantify}\textcolor{orange}{simply capture} a classic
+epidemic parameter, the reproduction number \(R_0\). Then we motivate
+new time-invariant approaches for more complex situations where \(R_0\)
+does not capture the complexities of outbreaks. This leads to the final
+subsection that defines ways to view epidemics in a time-invariant lens
+and discusses natural properties of these representations that allow for
+clearer ways to compare models and epidemics.
 
@@
 
-Epidemiological research is often interested in understanding intrinsic
-properties of the epidemic, independent from when the epidemic occurred
-or how frequently data was collected. One of the most famous numerical
-summaries of an epidemic is the reproduction number - \(R_0\). This
-intrinsic
-property\footnote{added to shorten this long sentence and connect that it is an intrinsic property}
+Epidemiological research is
+\textcolor{orange}{often interested}\footnote{\textcolor{orange}{I used often intsead of "a lot of epidemiological research" which sounds too colloquial}}
+in understanding intrinsic properties of the epidemic, independent from
+when the epidemic occurred or how frequently data was collected. One of
+the most famous numerical summaries of an epidemic is the reproduction
+number - \(R_0\). \textcolor{orange}{} This intrinsic
+property\}\footnote{\textcolor{orange}{added to shorten this long sentence and connect that it is an intrinsic property}}
 is a time-invariant value that is defined as the expected number of
@@
 problem in epidemiology surrounding how to transform time-dependent
-information into insights into desirable, intrinsic properties of an
-epidemic.
+information into insights into desirable, \textcolor{orange}{intrinsic}
+properties of an epidemic.
 
@@
 (R)ecovered. These simulations have been generated under a discrete
-approximation of \citet{Kermack1927}'s SIR model. The Kermack and
-McKendrick model captures the transitions from one state to the next as
-a system of ordinary differential equations, where \(N\) is the total
-number of individuals, \(\beta\) is the rate of infection, and
-\(\gamma\) is the rate of recovery,
+approximation of \citet{Kermack1927}'s SIR model.
+j\textcolor{orange}{The Kermack and McKendrick} model captures the
+transitions from one state to the next as a system of ordinary
+differential equations, where \(N\) is the total number of individuals,
+\(\beta\) is the rate of infection, and \(\gamma\) is the rate of
+recovery,
 
@@
 this time-dependent visualization cannot be used to compare a very
-simple but powerful numerical summary, \(R_0\).
+simple but powerful numerical summary, \textcolor{orange}{$R_0$}.
 
@@
 holistic approach to visualizing the overall path of our simulated
-epidemics is to examine how each epidemic traverses the
-three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
-just that, presenting the trajectories of the simulated epidemics in
-this three-dimensional space. For state space models like in our
+epidemics is to examine \textcolor{orange}{how} each epidemic traverses
+the three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot
+does just that, presenting the trajectories of the simulated epidemics
+in this three-dimensional space. For state space models like in our
 example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
@@
 epidemics are on the same trajectory in this three-dimensional space. In
-this simple generative example setting, being on the same trajectory
-indicates that \textcolor{violet}{the} two epidemics have the same
-\(R_0\) value. This can be proven mathematically.
+this simple generative example setting,
+\textcolor{orange}{being on the same trajectory} indicates that
+\textcolor{violet}{the} two epidemics have the same \(R_0\) value. This
+can be proven mathematically.
 
@@
 \sout{When the goal is to predict the future impact / trajectory of a epidemic,}
-In addition to being used to better understand completed epidemics,
+\textcolor{orange}{In addition to being used to better understand completed epidemics,}
 time-invariant tools can be used to aid prediction of future epidemics

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
minor updates to section 3 based on discussion and shannon suggested edits

--- 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
@@
 
-EpiCompare emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis.
-In this section we will highlight weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take.
-To accomplish this goal we will first demonstrate the inability of time-dependent tools to well quantify a classic epidemic parameter, the reproduction number $R_0$. 
-Then we will motivate new time-invariant approaches for more complex situations where $R_0$ doesn’t capture the complexities of outbreaks.
-This will lead us to our final subsection that defines ways to view epidemics in a time-invariant lense and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.
+\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis.
+In this section we highlight some weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take.
+To accomplish this goal we  first demonstrate the inability of time-dependent tools to \ben{adequately} quantify\footnote{\ben{[Ben says: original suggestion was "simply capture", and original phrase was "well quantify".]}} a classic epidemic parameter, the reproduction number $R_0$.
+Then we motivate new time-invariant approaches for more complex situations where $R_0$ does not capture the complexities of outbreaks.
+This leads to the final subsection that defines ways to view epidemics in a time-invariant lens and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.
 
@@
 One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.
-This intrinsic property\footnote{added to shorten this long sentence and connect that it is an intrinsic property} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}.
+This \ben{numerical summary} \sout{intrinsic property} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}.
 \citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic.
@@
 In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproduction number.
-One common time-dependent tool \shannon{used to better understand epidemics} is a series of time series line plots that track the proportion of the population in a given state (e.g.
+One common time-dependent tool used to better understand epidemics is a series of time series line plots that track the proportion of the population in a given state (e.g.
 infected) at a given time.
@@
 These simulations have been generated under a discrete approximation of  \citet{Kermack1927}’s SIR model.
-The Kermack and McKendrick model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
+The Kermack and McKendrick model captures the transitions from one state to the next as a system of ordinary differential equations
 
@@
     I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
-    R^\prime(t) &= \gamma I(t) \nonumber.
+    R^\prime(t) &= \gamma I(t) \nonumber,
 \end{align}
+
+where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery.
 
@@
 For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimensional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot.
-In this both of Figure  \ref{fig:different-scales-tern}’s plots we observe that the two epidemics are on the same trajectory in this three-dimensional space.
-In this simple generative example setting, being on the same trajectory indicates that \ben{the} two epidemics have the same $R_0$ value.
+In both of Figure  \ref{fig:different-scales-tern}’s plots we observe that the two epidemics are on the same trajectory.
+In this simple generative example setting, being on the same trajectory indicates that the two epidemics have the same $R_0$ value.
 This can be proven mathematically.
@@
 
-Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space (i.e.
-through a time-invariant lens) allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted.
+Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space \sout{(i.e.
+through a time-invariant lens)}\footnote{\ben{[Ben says: the next sentence mentions time-invariance - where you thinking something more global?]}} allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted.
 In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states.
@@
 
-The first set of tools allows a practioner to define distances between epidemics whose time features do not align.
+The first set of tools allows a practitioner to define distances between epidemics whose time features do not align.
 For completed epidemics, one way to better examine their properties is to represent their filamental trajectories as a finite sequence of equally spaced points.
@@
 \[
-d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
 \]
-where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$, where $|\psi_i|$ is the length of $\psi_i$.
+where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$ (where $|\psi_i|$ is the length of $\psi_i$).
 This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
-These extremeness estimators can be useful when comparing the true epidemic to a a set of simulated epidemics, and practitioners can interpret the epidemic's extremeness score relative to the extremeness scores of the simulations very easily.
+These extremeness estimators can be useful when comparing the true epidemic to a set of simulated epidemics, and practitioners can interpret the epidemic's extremeness score relative to the extremeness scores of the simulations very easily.
 
-\sout{When the goal is to predict the future impact / trajectory of a epidemic,} In addition to being used to better understand completed epidemics, time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse.
+In addition to being used to better understand completed epidemics, time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse.
 In settings where the epidemic only generates a single outbreak, these regions can be very telling if simulation models capture the epidemic’s structure.
@@
     \centering
-    \includegraphics[width = 1\textwidth]{images/pipeline2_1.pdf}
+    \includegraphics[width = 1\textwidth]{images/pipeline2_2.pdf}
     \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.tex
minor updates to section 3 based on discussion and shannon suggested edits

--- 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.tex
@@
 
-EpiCompare emphasizes the value of analyzing epidemics in a
+\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a
 \textit{time-invariant} way - approaches that remove some or all of the
 impact of start/end times and recording time scales when preforming the
-analysis. In this section we will highlight weaknesses of
+analysis. In this section we highlight some weaknesses of
 time-\textit{de}pendent analysis and define the mathematical
 underpinning of the time-invariant approach we take. To accomplish this
-goal we will first demonstrate the inability of time-dependent tools to
-well quantify a classic epidemic parameter, the reproduction number
-\(R_0\). Then we will motivate new time-invariant approaches for more
-complex situations where \(R_0\) doesn't capture the complexities of
-outbreaks. This will lead us to our final subsection that defines ways
-to view epidemics in a time-invariant lense and discusses natural
-properties of these representations that allow for clearer ways to
-compare models and epidemics.
+goal we first demonstrate the inability of time-dependent tools to
+\textcolor{violet}{adequately}
+quantify\footnote{\textcolor{violet}{[Ben says: original suggestion was "simply capture", and original phrase was "well quantify".]}}
+a classic epidemic parameter, the reproduction number \(R_0\). Then we
+motivate new time-invariant approaches for more complex situations where
+\(R_0\) does not capture the complexities of outbreaks. This leads to
+the final subsection that defines ways to view epidemics in a
+time-invariant lens and discusses natural properties of these
+representations that allow for clearer ways to compare models and
+epidemics.
 
@@
 summaries of an epidemic is the reproduction number - \(R_0\). This
-intrinsic
-property\footnote{added to shorten this long sentence and connect that it is an intrinsic property}
-is a time-invariant value that is defined as the expected number of
+\textcolor{violet}{numerical summary} \sout{intrinsic property} is a
+time-invariant value that is defined as the expected number of
 infections caused by a single infect or who is added to a completely
@@
 epidemic than just the reproduction number. One common time-dependent
-tool \textcolor{orange}{used to better understand epidemics} is a series
-of time series line plots that track the proportion of the population in
-a given state (e.g.~infected) at a given time. Figure
+tool used to better understand epidemics is a series of time series line
+plots that track the proportion of the population in a given state
+(e.g.~infected) at a given time. Figure
 \ref{fig:different-scales-standard} visualizes two different simulated
@@
 McKendrick model captures the transitions from one state to the next as
-a system of ordinary differential equations, where \(N\) is the total
-number of individuals, \(\beta\) is the rate of infection, and
-\(\gamma\) is the rate of recovery,
+a system of ordinary differential equations
 
@@
     I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
-    R^\prime(t) &= \gamma I(t) \nonumber.
+    R^\prime(t) &= \gamma I(t) \nonumber,
 \end{align}
+
+where \(N\) is the total number of individuals, \(\beta\) is the rate of
+infection, and \(\gamma\) is the rate of recovery.
 
@@
 two-dimensional \textit{ternary} plot, as seen in Figure
-\ref{fig:different-scales-tern}'s right plot. In this both of Figure
+\ref{fig:different-scales-tern}'s right plot. In both of Figure
 \ref{fig:different-scales-tern}'s plots we observe that the two
-epidemics are on the same trajectory in this three-dimensional space. In
-this simple generative example setting, being on the same trajectory
-indicates that \textcolor{violet}{the} two epidemics have the same
-\(R_0\) value. This can be proven mathematically.
+epidemics are on the same trajectory. In this simple generative example
+setting, being on the same trajectory indicates that the two epidemics
+have the same \(R_0\) value. This can be proven mathematically.
 
@@
 filamental trajectories embedded in a lower dimensional space
-(i.e.~through a time-invariant lens) allows us to better compare the
-overall structure of the epidemic and see how the population is directly
-impacted. In this section we present time-invariant tools that can be
-applied to complex epidemics where the epidemic's generative process is
-unknown and can have more than three states. These tools leverage the
-idea that an epidemic can be viewed as a trajectory and that many
-properties of the epidemic are well captured when we do so. This
-approach is useful when the epidemic of interest has only gone through a
-single realization of its outbreak (before the population of individuals
-become susceptible again).
-
-The first set of tools allows a practioner to define distances between
+\sout{(i.e. through a time-invariant lens)}\footnote{\textcolor{violet}{[Ben says: the next sentence mentions time-invariance - where you thinking something more global?]}}
+allows us to better compare the overall structure of the epidemic and
+see how the population is directly impacted. In this section we present
+time-invariant tools that can be applied to complex epidemics where the
+epidemic's generative process is unknown and can have more than three
+states. These tools leverage the idea that an epidemic can be viewed as
+a trajectory and that many properties of the epidemic are well captured
+when we do so. This approach is useful when the epidemic of interest has
+only gone through a single realization of its outbreak (before the
+population of individuals become susceptible again).
+
+The first set of tools allows a practitioner to define distances between
 epidemics whose time features do not align. For completed epidemics, one
@@
 \[
-d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
 \] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
-\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\), where
-\(|\psi_i|\) is the length of \(\psi_i\). This distance is naturally
+\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (where
+\(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
 time-invariant, and can be plugged into multiple distance-based
-assessment tools to examine the overall ``extremeness'\,' of points,
+assessment tools to examine the overall ``extremeness'' of points,
 including pseudo-density estimators and depth/local depth functions
 \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
-estimators can be useful when comparing the true epidemic to a a set of
+estimators can be useful when comparing the true epidemic to a set of
 simulated epidemics, and practitioners can interpret the epidemic's
@@
 
-\sout{When the goal is to predict the future impact / trajectory of a epidemic,}
 In addition to being used to better understand completed epidemics,
@@
 \midrule
-Simple SIR & 0.0036733 & 0.00\\
-EpiModel SIR & 0.0149686 & 0.02\\
+Simple SIR & 0.0117621 & 0.01\\
+EpiModel SIR & 0.0356503 & 0.01\\
 \bottomrule

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Auto stash before merge of "master" and "origin/master"

@@
 \ref{fig:pipeline2}, we \ben{\sout{illustrate} show}\footnote{\ben{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
-I'm happy with other synonyms as well.}} how our package's functions fit into the data analysis pipeline introduced in Fig.
+I'm happy with other synonyms as well.} \shannon{accept}} how our package's functions fit into the data analysis pipeline introduced in Fig.
 \ref{fig:pipeline}.
@@
 ]} \shannon{It was grouping before ;).
-Fine with me.} agents into subpopulations (e.g.
+Fine with me.}} agents into subpopulations (e.g.
 subpopulations defined by age or sex) can highlight different aggregate level trends.
@@
 ]}\shannon{agreed.
-I like 'to provide summaries at a'}}.
+aggregation at a group level}}.
 
@@
 In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.
+
 Model output summaries can be very complex, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.
@@
 While incredibly useful, these special classes can be difficult for novice coders to handle.
+\shannon{motivates for fortify but not filamental compression}
 
@@
 If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect word choice.]}\shannon{"To support the ideas presented in Sec."?}} to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
-We have found these tools to be specifically applicable for situations where only one "season" or "cycle"\footnote{\ben{[Ben says: I think we can remove the paratheses here.]}} of the epidemic has occurred.
-
-\footnote{\ben{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\shannon{I'm not saying 'no' but I'd do it in an appendix and not the main text.
-I don't think equations belong in Section 3.}}One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
+We have found these tools to be specifically applicable for situations where only one "season" or "cycle"\footnote{\ben{[Ben says: I think we can remove the paratheses here.]} \shannon{accepted}} of the epidemic has occurred.
+
+\footnote{\ben{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\shannon{motivate better}}One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
 If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space.
 For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
-In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done \ben{by eye.\sout{with the 'eye test.'}} In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
+In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done \ben{by eye.\sout{with the 'eye test.'}\shannon{I'm not saying 'no' but I'd do it in an appendix and not the main text.
+I don't think equations belong in Section 3.}} In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
 The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model.

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.Rmd
section 3 edits - tried again at last two steps in pipeline.

@@
 In Fig.
-\ref{fig:pipeline2}, we \ben{\sout{illustrate} show}\footnote{\ben{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
+\ref{fig:pipeline2}, we show\footnote{\ben{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
 I'm happy with other synonyms as well.} \shannon{accept}} how our package's functions fit into the data analysis pipeline introduced in Fig.
@@
 Before data can be explored, they must be collected.
-Raw data can appear in many different formats.\footnote{\ben{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\shannon{I think we can get rid of the 'raw data can appear...' sentence.}} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+\sout{Raw data can appear in many different formats.}\footnote{\ben{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\shannon{I think we can get rid of the 'raw data can appear...' sentence.} RESOLVED 5/17} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
 Other data collections focus on aggregate counts of individuals in each epidemic state.
@@
 start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
-There are often situations where \ben{\sout{aggregating} grouping}\footnote{\ben{[We use aggregating two times in this sentence.
-I'm open to other synonyms as well, but it seems repetitive.
-]} \shannon{It was grouping before ;).
-Fine with me.}} agents into subpopulations (e.g.
+There are often situations where \ben{\sout{aggregating} grouping}\footnote{RESOLVED 5/17 (aggregating)} agents into subpopulations (e.g.
 subpopulations defined by age or sex) can highlight different aggregate level trends.
 For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining the function \pkg{dplyr}::`group_by()` and `agent_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
+In \pkg{EpiCompare}, we facilitate subpopulation analysis by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`, the tidyverse approach to the split-apply-combine paradigm for data analysis \citep{wickham2011}.\footnote{Changed to try to clarify how how agents_to_aggregate works with group_by and why (split-apply-combine)}
+
+\shannon{old text below}
+[by combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
 ]}\shannon{agreed.
-aggregation at a group level}}.
-
-Besides aiding subpopulation analysis, `agents_to_aggregate()` can \ben{\sout{deal with} accommodate}\footnote{\ben{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
-Again, I'm open to other synonyms.]}\shannon{I like accomodate}} a wide range of information about each individual.
+aggregation at a group level}}.]
+
+Besides aiding subpopulation analysis, `agents_to_aggregate()` can  accommodate\footnote{\ben{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
+Again, I'm open to other synonyms.]}\shannon{I like accomodate}  RESOLVED 5/17} a wide range of information about each individual.
 In fact, this function can account for infinitely many states.
@@
 
-**\ben{Exploratory Data Analysis (}EDA\ben{)}**
+**Exploratory Data Analysis (EDA)*
 
  In the early stages of a project, familiarizing oneself with the data usually means figuring out useful combinations of visualizations and numerical summaries of the data both at population and subpopulation level.
-An expert coder can start with `agents_to_aggregate()` to successfully accomplish \ben{exploratory data analysis (}EDA\ben{)} in many ways.
+An expert coder can start with `agents_to_aggregate()` to successfully accomplish exploratory data analysis (EDA) in many ways.
 \pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there \ben{are} three unique epidemiological states (like in the SIR model).
-Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a \ben{\sout{rapid}}\footnote{\ben{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \shannon{accepted}} way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
+Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a \footnote{\ben{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \shannon{accepted}.
+RESOLVED 5/17} way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
 The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.
-Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a \ben{\sout{"time-invarance"}time-invariant}\footnote{\ben{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \shannon{agreed}} representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a time-invariant\footnote{\ben{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \shannon{agreed} RESOLVED 5/17} representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-After getting a sense of what a past or current epidemic looks like with EDA, \ben{\sout{often}} the next step in the data analysis pipeline is \ben{often}\footnote{\ben{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence.
-Maybe it's just me.]}} model fitting and/or simulation.
+After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often\footnote{\ben{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence.
+Maybe it's just me.]}RESOLVED 5/17} model fitting and/or simulation.
 In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.
@@
 
+If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format.
+In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.
+In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{new signpost.
+There are two paragraphs below corresponding to the two points.
+The second paragraph could use another sentence maybe.
+I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}
+
+Modeling and simulation output can be very complex objects, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.
+The special classes often contain a plethora of information about residuals, model diagnostics, input parameters, and more.
+While incredibly useful, these special classes can be difficult for novice coders to handle.
+\shannon{motivates for fortify but not filamental compression}.
+To this end, \pkg{EpiCompare} provides a series of `fortify`-style methods, called `fortify_aggregate()` which transform output from infectious disease modeling and simulation packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
+These fortify functions have output that is consistent with that of `agents_to_aggregate()`.
+These standardized outputs can then be piped to summaries, tables, and plots.
+
+Because the fortified output contains temporal information and we provide the function, `filament_compression()`, to transform temporally defined epidemics to their filamental representations.
+These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.
+
+[[\shannon{OLD TEXT BELOW}
+
 \footnote{\ben{[Ben says: I found the structure of this section slightly confusing.
@@
 
+]]
+
 **Comparisons and Assessment**
+
+Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare different models, and simulations to one another.
+In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
+mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{a bit of an awk transition but I wanted to use past wording about TIA.}  We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of an epidemic has occurred.
+
+One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
+If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space.
+For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
+In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done by eye.
+In \pkg{EpiCompare}, we also provide these prediction regions for epidemic models with with more than three states.
+The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model.
+For both of these geometric structures, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hausdorff_dist()`).
+The vignettes for `EpiCompare` go into much more mathematical detail about how these functions can be used in conjunction with time-invariant analysis.
+
+[[\shannon{OLD TEXT}
 
@@
 Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
+]]
 
 # A tour of \pkg{EpiCompare} {short-title="Tour" #sec:tour}
+
+Finally, in this section we show how tools from \pkg{EpiCompare} can be used in each step of the data analysis pipeline shown in Fig.
+\ref{fig:pipeline}.
+We analyze an outbreak of measles in the town of Hagelloch, Germany from 1861-1862, a data set organized by \cite{pfeilsticker1863}.
+The data was later made visible by \cite{oesterle1992} and made available in an \proglang{R} by \cite{surveillance2017}.
+The Hagelloch data includes a rich set of features including household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.
+A subset of the data is shown in Table \ref{tab:hags-people}.
+Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.
+
+[[\shannon{OLD TEXT}
 
@@
 We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner via \pkg{EpiCompare}.
+]]
 
 \subsection{Data and exploratory analysis}
-
-We analyze an outbreak of measles in the town of Hagelloch, Germany from 1861-1862, a data set organized by \cite{pfeilsticker1863}.
-The data was later made visible by \cite{oesterle1992} and made available in an \proglang{R} by \cite{surveillance2017}.
-The Hagelloch data includes a rich set of features including household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.
-A subset of the data is shown in Table \ref{tab:hags-people}.
-Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.
 

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.tex
section 3 edits - tried again at last two steps in pipeline.

@@
 \ref{fig:pipeline2}, we
-\textcolor{violet}{\sout{illustrate} show}\footnote{\textcolor{violet}{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
-I'm happy with other synonyms as well.}}
+show\footnote{\textcolor{violet}{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
+I'm happy with other synonyms as well.} \textcolor{orange}{accept}}
 how our package's functions fit into the data analysis pipeline
@@
 can be explored. Before data can be explored, they must be collected.
-Raw data can appear in many different
-formats.\footnote{\textcolor{violet}{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\textcolor{orange}{I think we can get rid of the 'raw data can appear...' sentence.}}
+\sout{Raw data can appear in many different formats.}\footnote{\textcolor{violet}{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\textcolor{orange}{I think we can get rid of the 'raw data can appear...' sentence.} RESOLVED 5/17}
 Sometimes individual records are collected, with times of different
@@
 There are often situations where
-\textcolor{violet}{\sout{aggregating} grouping}\textbackslash footnote\{\textcolor{violet}{[We use aggregating two times in this sentence.
-I'm open to other synonyms as well, but it seems repetitive.
-]}
-\textcolor{orange}{It was grouping before ;).  Fine with me.} agents
-into subpopulations (e.g.~subpopulations defined by age or sex) can
-highlight different aggregate level trends. For example, research by
+\textcolor{violet}{\sout{aggregating} grouping}\footnote{RESOLVED 5/17 (aggregating)}
+agents into subpopulations (e.g.~subpopulations defined by age or sex)
+can highlight different aggregate level trends. For example, research by
 \citet{rvachev1985,anderson1992,worby2015} develop state-based models
 that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis by combining
-the function \pkg{dplyr}::\texttt{group\_by()} and
-\texttt{agent\_to\_aggregate()} to provide aggregation at a
+In \pkg{EpiCompare}, we facilitate subpopulation analysis by extending
+\texttt{agents\_to\_aggregate()} to work with
+\pkg{dplyr}::\texttt{group\_by()}, the tidyverse approach to the
+split-apply-combine paradigm for data analysis
+\citep{wickham2011}.\textbackslash footnote\{Changed to try to clarify
+how how agents\_to\_aggregate works with group\_by and why
+(split-apply-combine)\}
+
+\textcolor{orange}{old text below}
+
+{[}by combining the function \pkg{dplyr}::\texttt{group\_by()} and
+\texttt{agents\_to\_aggregate()} to provide aggregation at a
 subpopulation
-level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed.  I like 'to provide summaries at a'}}.
+level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed. aggregation at a group level}}.{]}
 
@@
 can
-\textcolor{violet}{\sout{deal with} accommodate}\footnote{\textcolor{violet}{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
-Again, I'm open to other synonyms.]}\textcolor{orange}{I like accomodate}}
+accommodate\footnote{\textcolor{violet}{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
+Again, I'm open to other synonyms.]}\textcolor{orange}{I like accomodate}  RESOLVED 5/17}
 a wide range of information about each individual. In fact, this
@@
 
-\textbf{\textcolor{violet}{Exploratory Data Analysis (}EDA\textcolor{violet}{)}}
+**Exploratory Data Analysis (EDA)*
 
@@
 level. An expert coder can start with \texttt{agents\_to\_aggregate()}
-to successfully accomplish
-\textcolor{violet}{exploratory data analysis (}EDA\textcolor{violet}{)}
-in many ways. \pkg{EpiCompare} also includes tools that allow a novice
-coder to rapidly explore data, provided there \textcolor{violet}{are}
-three unique epidemiological states (like in the SIR model). Building on
+to successfully accomplish exploratory data analysis (EDA) in many ways.
+\pkg{EpiCompare} also includes tools that allow a novice coder to
+rapidly explore data, provided there \textcolor{violet}{are} three
+unique epidemiological states (like in the SIR model). Building on
 \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s
 \texttt{geom\_aggregate()} provides a
-\textcolor{violet}{\sout{rapid}}\footnote{\textcolor{violet}{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \textcolor{orange}{accepted}}
+\footnote{\textcolor{violet}{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \textcolor{orange}{accepted}. RESOLVED 5/17}
 way to explore how different subpopulations experience of an epidemic
@@
 simplex representation of these epidemics emphasizes a
-\textcolor{violet}{\sout{"time-invarance"}time-invariant}\footnote{\textcolor{violet}{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \textcolor{orange}{agreed}}
+time-invariant\footnote{\textcolor{violet}{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \textcolor{orange}{agreed} RESOLVED 5/17}
 representation of the data (for a refresher see Section
@@
 After getting a sense of what a past or current epidemic looks like with
-EDA, \textcolor{violet}{\sout{often}} the next step in the data analysis
-pipeline is
-\textcolor{violet}{often}\footnote{\textcolor{violet}{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence. Maybe it's just me.]}}
+EDA, the next step in the data analysis pipeline is
+often\footnote{\textcolor{violet}{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence. Maybe it's just me.]}RESOLVED 5/17}
 model fitting and/or simulation. In this step and the next step
@@
 
+If practitioners wish to compare models-to-observations or even
+models-to-models, they need to post-process their models and simulations
+to disseminate the results in an easily digestible format. In general,
+post-processing of modeling and simulation consists of making summary
+statistics, plots, and tables. In \pkg{EpiCompare}, we provide (1)
+functions to standardize simulation and model output from external
+packages and (2) a function to transform standardized simulation and
+model output into a format amenable to time-invariant
+analysis.\footnote{new signpost.
+There are two paragraphs below corresponding to the two points.
+The second paragraph could use another sentence maybe.
+I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}
+
+Modeling and simulation output can be very complex objects, and as a
+result, a number of epidemic modeling \proglang{R} packages return a
+special class. The special classes often contain a plethora of
+information about residuals, model diagnostics, input parameters, and
+more. While incredibly useful, these special classes can be difficult
+for novice coders to handle.
+\textcolor{orange}{motivates for fortify but not filamental compression}.
+To this end, \pkg{EpiCompare} provides a series of
+\texttt{fortify}-style methods, called \texttt{fortify\_aggregate()}
+which transform output from infectious disease modeling and simulation
+packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
+which contain information about the total number of individuals in each
+state at a given time, for a given simulation. These fortify functions
+have output that is consistent with that of
+\texttt{agents\_to\_aggregate()}. These standardized outputs can then be
+piped to summaries, tables, and plots.
+
+Because the fortified output contains temporal information and we
+provide the function, \texttt{filament\_compression()}, to transform
+temporally defined epidemics to their filamental representations. These
+filaments can then be fairly compared to one another or passed to
+further time-invariant analysis tools described below.
+
+{[}{[}\textcolor{orange}{OLD TEXT BELOW}
+
 \footnote{\textcolor{violet}{[Ben says: I found the structure of this section slightly confusing.
@@
 In general, post-processing of modeling and simulation consists of
-making summary statistics, plots, and tables. Model output summaries can
-be very complex, and as a result, a number of epidemic modeling
-\proglang{R} packages return a special class. The special classes often
-contain a plethora of information about residuals, model diagnostics,
-input parameters, and more. While incredibly useful, these special
-classes can be difficult for novice coders to handle.
+making summary statistics, plots, and tables.
+
+Model output summaries can be very complex, and as a result, a number of
+epidemic modeling \proglang{R} packages return a special class. The
+special classes often contain a plethora of information about residuals,
+model diagnostics, input parameters, and more. While incredibly useful,
+these special classes can be difficult for novice coders to handle.
+\textcolor{orange}{motivates for fortify but not filamental compression}
 
@@
 
+{]}{]}
+
 \textbf{Comparisons and Assessment}
+
+Finally, the data analysis pipeline often ends with plots, tables, and
+summary statistics that are used to compare different models, and
+simulations to one another. In \pkg{EpiCompare} we provide a set of
+comparison and assessment tools for model and simulation results that
+extend beyond the standard performance metrics (e.g.~mean squared error
+or AIC) and into the lens of time-invariant
+analysis.\footnote{a bit of an awk transition but I wanted to use past wording about TIA.}
+We have found these tools to be specifically applicable for situations
+where only one ``season'' or ``cycle'' of an epidemic has occurred.
+
+One tool we provide to assess models is through the creation of
+geometric prediction regions, which are useful when we treat epidemics
+like filaments. If there is a set of simulated epidemics from a model,
+we can create a geometric prediction region for the expected trajectory
+of the epidemic in the state space. For three-state epidemic models, we
+provide the \texttt{ggplot}/\texttt{ggtern} extension
+\texttt{geom\_prediction\_band()} which creates a prediction region
+around the top \(1-\alpha\) proportion of the simulations. In this
+visual setting, comparing this prediction to the true epidemic
+trajectory or comparing the prediction regions defined by two different
+models' simulations can be done by eye. In \pkg{EpiCompare}, we also
+provide these prediction regions for epidemic models with with more than
+three states. The functions \texttt{create\_convex\_hull\_structure()}
+and \texttt{create\_delta\_ball\_structure()} create different geometric
+representations of prediction regions for any dimensional state-based
+model. For both of these geometric structures, we provide functions to
+check if a path is contained (\texttt{contained()}) and the ability to
+assess the Hausdorff distance between prediction regions based on
+simulations from different model (\texttt{hausdorff\_dist()}). The
+vignettes for \texttt{EpiCompare} go into much more mathematical detail
+about how these functions can be used in conjunction with time-invariant
+analysis.
+
+{[}{[}\textcolor{orange}{OLD TEXT}
 
@@
 situations where only one ``season'' or
-``cycle''\footnote{\textcolor{violet}{[Ben says: I think we can remove the paratheses here.]}}
+``cycle''\footnote{\textcolor{violet}{[Ben says: I think we can remove the paratheses here.]} \textcolor{orange}{accepted}}
 of the epidemic has occurred.
 
-\footnote{\textcolor{violet}{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\textcolor{orange}{I'm not saying 'no' but I'd do it in an appendix and not the main text.
-I don't think equations belong in Section 3.}}One
+\footnote{\textcolor{violet}{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\textcolor{orange}{motivate better}}One
 tool we provide to assess models is through the creation of geometric
@@
 models' simulations can be done
-\textcolor{violet}{by eye.\sout{with the 'eye test.'}} In
-\pkg{EpiCompare} we also provide these prediction regions for epidemic
-models with with more than three states. The functions
+\textcolor{violet}{by eye.\sout{with the 'eye test.'}\textcolor{orange}{I'm not saying 'no' but I'd do it in an appendix and not the main text.  I don't think equations belong in Section 3.}}
+In \pkg{EpiCompare} we also provide these prediction regions for
+epidemic models with with more than three states. The functions
 \texttt{create\_convex\_hull\_structure()} and
@@
 \texttt{local\_distance\_depth\_function()} provide depth scores that
-suggest how geometrically central an epidemic is to simulations.
+suggest how geometrically central an epidemic is to simulations. {]}{]}
 
 \section[Tour]{A tour of \pkg{EpiCompare}}\label{sec:tour}
+
+Finally, in this section we show how tools from \pkg{EpiCompare} can be
+used in each step of the data analysis pipeline shown in Fig.
+\ref{fig:pipeline}. We analyze an outbreak of measles in the town of
+Hagelloch, Germany from 1861-1862, a data set organized by
+\cite{pfeilsticker1863}. The data was later made visible by
+\cite{oesterle1992} and made available in an \proglang{R} by
+\cite{surveillance2017}. The Hagelloch data includes a rich set of
+features including household members, school level, household locations,
+date of first symptoms (prodromes), date of measles rash, and even the
+alleged infector. A subset of the data is shown in Table
+\ref{tab:hags-people}. Because of these rich features, this data set has
+been an ideal testing ground methodology in infectious disease
+epidemiology and is used in work by
+\cite{Neal2004,britton2011,groendyke2012,becker2016}.
+
+{[}{[}OLD TEXT
 
@@
 beginning to end that can be accomplished in a streamlined and
-standardized manner via \pkg{EpiCompare}.
+standardized manner via \pkg{EpiCompare}. {]}{]}
 
 \subsection{Data and exploratory analysis}
-
-We analyze an outbreak of measles in the town of Hagelloch, Germany from
-1861-1862, a data set organized by \cite{pfeilsticker1863}. The data was
-later made visible by \cite{oesterle1992} and made available in an
-\proglang{R} by \cite{surveillance2017}. The Hagelloch data includes a
-rich set of features including household members, school level,
-household locations, date of first symptoms (prodromes), date of measles
-rash, and even the alleged infector. A subset of the data is shown in
-Table \ref{tab:hags-people}. Because of these rich features, this data
-set has been an ideal testing ground methodology in infectious disease
-epidemiology and is used in work by
-\cite{Neal2004,britton2011,groendyke2012,becker2016}.
 

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Section 4 edits.

@@
 The data was later made visible by \cite{oesterle1992} and made available in an \proglang{R} by \cite{surveillance2017}.
-The Hagelloch data includes a rich set of features including household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.
-A subset of the data is shown in Table \ref{tab:hags-people}.
-Because of these rich features, this data set has been an ideal testing ground  methodology in infectious disease epidemiology and is used in work by \cite{Neal2004,britton2011,groendyke2012,becker2016}.
+This data set includes a rich collection of features and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\shannon{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's goingon.}}
 
@@
 
-\subsection{Data and exploratory analysis}
+\subsection{Pre-processing and EDA}
 
@@
 
-With \pkg{EpiCompare}, we can easily obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate()}.
+The Hagelloch data include a rich set of features at the individual level.
+Recorded features include household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.
+A subset of the data is shown in Table \ref{tab:hags-people}.
+With \pkg{EpiCompare}, we can easily \shannon{pre-process the data to} obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate()}.
 The function \code{agents_to_aggregate()} is a key component of \pkg{EpiCompare}, allowing the user to easily switch from an individual-level (i.e.
-an agent) view of a disease to an aggregate level.
-For example, the below code shows how we can convert the agent data to a cumulative incidence of the measles rash, in order to see how the disease spread through the population over time.
-We can then compare the cumulative incidence of the rash to the cumulative incidence of the prodromes, i.e.
+an agent) \sout{view} lens of a disease to an aggregate \sout{level} lens.
+For example, the below code shows how we can convert the agent data to a cumulative incidence \shannon{plot} of the measles rash\sout{, in order to see how the disease spread through the population over time.} We can then compare the cumulative incidence of the rash to the cumulative incidence of the prodromes, i.e.
 the initial symptoms.
@@
 One possible question of interest is the duration between initial onset of prodromes  and the appearance of the measles rash.
-Since \code{agent_to_aggregate()} outputs a tidy-style data frame, it is a simple task to plot the two sets of incidence curves on the same graph (Fig.
+Since \code{agents_to_aggregate()} outputs a tidy-style data frame, it is a simple task to plot the two sets of incidence curves on the same graph (Fig.
 \ref{fig:cifs}).
@@
 
-Along with multiple epidemic states, the function \code{agents_to_aggregate} can also be extended to populations with vital dynamics (e.g.
+Along with multiple epidemic states, the function \code{agents_to_aggregate()} can also be extended to populations with vital dynamics (e.g.
 birth and death) and examples of this are shown in the package vignette.
@@
 
+## Modeling and Simulation
+
+\footnote{\shannon{section headings to align with our pipeline}}
+
 Up to this point, we have used \pkg{EpiCompare} in the context of observed data.
-We also want to compare statistical models, and \pkg{EpiCompare} aids in that process via a simple yet flexible individual-level simulator, conversion tools for popular epidemic model packages, and model assessments.
-We demonstrate an example here.
-
-We first try to model the Hagelloch data with a stochastic SIR model, which we refer to as the 'simple SIR.'  In our vignette, we show how to fit this simple SIR model via maximum likelihood and simulate from the model with those best fit parameters.
+We also want to compare statistical models, and \pkg{EpiCompare} aids in that process via a simple yet flexible individual-level simulator\sout{, conversion tools for popular epidemic model packages, and model assessments.}  We demonstrate an example here.
+
+We first try to model the Hagelloch data with a stochastic SIR model, which we refer to as the 'simple SIR.'  In our full vignette \shannon{(available online),} we show how to fit this simple SIR model via maximum likelihood and simulate from the model with those best fit parameters.
 Our function \code{simulate_agents()} generates individual level data according to discrete time multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.
@@
 The result of our simulation is the object \code{agents} which is a `r nrow(agents)` $\times$ `r ncol(agents)` data frame, which details the time of entry into the $S$, $I$, and $R$ states for a given simulation.
-Before we examine the results of this simple SIR model, we will also examine another, more sophisticated SIR model, this time from the package \pkg{EpiModel}.
-Briefly, this model first fits a contact network to the set of indivduals, where the class of the student is a covariate.
+Before we examine the results of this simple SIR model, we will also examine another, more sophisticated SIR model, this time from the package \pkg{EpiModel} \citep{Jenness2018}.
+Briefly, this model first fits a contact network to the set of individuals, where the class of the child is a covariate.
 The model then simulates a SIR-epidemic on that network.
@@
 The output of this model is \code{epimodel_sir}, an object of class \code{`r class(epimodel_sir)`}, which contains a plethora of modeling information.
-We provide the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output and transform it into a tidy-style data frame.
+
+## Post-processing and comparison
+
+The next step is to compare the Simple SIR model to the EpiModel SIR model.
+We provide the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transform it into a tidy-style data frame.
 
@@
 We also see that the prediction band is very large, covering up a large area of the ternary plot.
-On the other hand, for the \pkg{EpiModel} model, we see that the prediction band covers the data quite well and takes up less area.
+On the other hand, for the EpiModel network model, we see that the prediction band covers the data quite well and takes up less area.
 
@@
 This can be captured with the set of simulations both models predict (gray lines), which all generally have a single defined peak of infection whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
-This observation is backed up by the below analysis that demonstrates that the estimated pseudo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme}).
-In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the \pkg{EpiModel} network model is a better fit than the Simple SIR model, and 2) the fit is only good at the geometric filamental level as opposed to the epidemic trajectory filamental level.
+This observation is backed up by the below analysis that demonstrates that the estimated pseudo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme})\footnote{\shannon{Ben, do we want to add another sentence or two explaining the two columns in the table?
+The second one I think makes sense to me but not the first.}} In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the EpiModel network model is a better fit than the Simple SIR model, and 2) the fit is only good at the \sout{geometric filamental level as opposed to the epidemic trajectory filamental level.} \shannon{individual point level as opposed to the geometric filamental level.}
 
@@
   kable(booktabs = TRUE,
-        caption = paste("Top and bottom 2 rows of \\tt{all\\_together\\_df}\\text{,",
+        caption = paste("Top and bottom 2 rows of \\tt{all\\_together\\_df}\\textnormal{,",
                         "combining both simulated epidemics and the true",
@@
           linebreak(c("Type",
-                      "simulations-based estimated psuedo-density",
-              "proportion of simulations with lower estimated psuedo-density"),
+                      "simulations-based estimated pseudo-density",
+              "proportion of simulations with lower estimated pseudo-density"),
               align = c("l")),
         caption = paste("The extremeness of the true simulations based on",
-                        "comparing psuedo-density estimates between true",
+                        "comparing pseudo-density estimates between true",
                         "vs simulated curves"), label = "hags-extreme") %>%

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.tex
Section 4 edits.

@@
 \cite{oesterle1992} and made available in an \proglang{R} by
-\cite{surveillance2017}. The Hagelloch data includes a rich set of
-features including household members, school level, household locations,
-date of first symptoms (prodromes), date of measles rash, and even the
-alleged infector. A subset of the data is shown in Table
-\ref{tab:hags-people}. Because of these rich features, this data set has
-been an ideal testing ground methodology in infectious disease
-epidemiology and is used in work by
-\cite{Neal2004,britton2011,groendyke2012,becker2016}.
-
-{[}{[}OLD TEXT
+\cite{surveillance2017}. This data set includes a rich collection of
+features and is an ideal testing ground for methodology in infectious
+disease epidemiology
+\cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\textcolor{orange}{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's goingon.}}
+
+{[}{[}\textcolor{orange}{OLD TEXT}
 
@@
 
-With \pkg{EpiCompare}, we can easily obtain the empirical cumulative
-incidence function with respect to the measles rash appearance (variable
-\code{ERU}) with the following tidy-style function,
+The Hagelloch data include a rich set of features at the individual
+level. Recorded features include household members, school level,
+household locations, date of first symptoms (prodromes), date of measles
+rash, and even the alleged infector. A subset of the data is shown in
+Table \ref{tab:hags-people}. With \pkg{EpiCompare}, we can easily
+\textcolor{orange}{pre-process the data to} obtain the empirical
+cumulative incidence function with respect to the measles rash
+appearance (variable \code{ERU}) with the following tidy-style function,
 \code{agents_to_aggregate()}. The function \code{agents_to_aggregate()}
 is a key component of \pkg{EpiCompare}, allowing the user to easily
-switch from an individual-level (i.e.~an agent) view of a disease to an
-aggregate level. For example, the below code shows how we can convert
-the agent data to a cumulative incidence of the measles rash, in order
-to see how the disease spread through the population over time. We can
-then compare the cumulative incidence of the rash to the cumulative
-incidence of the prodromes, i.e.~the initial symptoms. We do this with
-the below code, and a part of the cumulative incidence data output is
-shown in Table \ref{tab:cif-rash}. The argument
+switch from an individual-level (i.e.~an agent) \sout{view} lens of a
+disease to an aggregate \sout{level} lens. For example, the below code
+shows how we can convert the agent data to a cumulative incidence
+\textcolor{orange}{plot} of the measles
+rash\sout{, in order to see how the disease spread through the population over time.}
+We can then compare the cumulative incidence of the rash to the
+cumulative incidence of the prodromes, i.e.~the initial symptoms. We do
+this with the below code, and a part of the cumulative incidence data
+output is shown in Table \ref{tab:cif-rash}. The argument
 \code{integer_time_expansion} indicates whether we should include all
@@
 of prodromes and the appearance of the measles rash. Since
-\code{agent_to_aggregate()} outputs a tidy-style data frame, it is a
+\code{agents_to_aggregate()} outputs a tidy-style data frame, it is a
 simple task to plot the two sets of incidence curves on the same graph
@@
 Along with multiple epidemic states, the function
-\code{agents_to_aggregate} can also be extended to populations with
+\code{agents_to_aggregate()} can also be extended to populations with
 vital dynamics (e.g.~birth and death) and examples of this are shown in
@@
 
+\hypertarget{modeling-and-simulation}{%
+\subsection{Modeling and Simulation}\label{modeling-and-simulation}}
+
+\footnote{\textcolor{orange}{section headings to align with our pipeline}}
+
 Up to this point, we have used \pkg{EpiCompare} in the context of
@@
 \pkg{EpiCompare} aids in that process via a simple yet flexible
-individual-level simulator, conversion tools for popular epidemic model
-packages, and model assessments. We demonstrate an example here.
+individual-level
+simulator\sout{, conversion tools for popular epidemic model packages, and model assessments.}
+We demonstrate an example here.
 
 We first try to model the Hagelloch data with a stochastic SIR model,
-which we refer to as the `simple SIR.' In our vignette, we show how to
-fit this simple SIR model via maximum likelihood and simulate from the
-model with those best fit parameters. Our function
-\code{simulate_agents()} generates individual level data according to
-discrete time multinomial draws, which depend on the number of
-individuals in each state at the previous time step and a matrix of
-transition probabilities. For example, the below code generates 100
-simulations of an outbreak of a disease with one initial infector in a
-population of \(n= 188\) individuals.
+which we refer to as the `simple SIR.' In our full vignette
+\textcolor{orange}{(available online),} we show how to fit this simple
+SIR model via maximum likelihood and simulate from the model with those
+best fit parameters. Our function \code{simulate_agents()} generates
+individual level data according to discrete time multinomial draws,
+which depend on the number of individuals in each state at the previous
+time step and a matrix of transition probabilities. For example, the
+below code generates 100 simulations of an outbreak of a disease with
+one initial infector in a population of \(n= 188\) individuals.
 
@@
 the results of this simple SIR model, we will also examine another, more
-sophisticated SIR model, this time from the package \pkg{EpiModel}.
-Briefly, this model first fits a contact network to the set of
-indivduals, where the class of the student is a covariate. The model
-then simulates a SIR-epidemic on that network.
+sophisticated SIR model, this time from the package \pkg{EpiModel}
+\citep{Jenness2018}. Briefly, this model first fits a contact network to
+the set of individuals, where the class of the child is a covariate. The
+model then simulates a SIR-epidemic on that network.
 
@@
 The output of this model is \code{epimodel_sir}, an object of class
-\code{netsim}, which contains a plethora of modeling information. We
-provide the function \code{fortify_aggregate()}, which can take objects
-from specialized classes of modeling output and transform it into a
-tidy-style data frame.
+\code{netsim}, which contains a plethora of modeling information.
+
+\hypertarget{post-processing-and-comparison}{%
+\subsection{Post-processing and
+comparison}\label{post-processing-and-comparison}}
+
+The next step is to compare the Simple SIR model to the EpiModel SIR
+model. We provide the function \code{fortify_aggregate()}, which can
+take objects from specialized classes of modeling output
+\textcolor{orange}{(like those made by \code{netsim()})} and transform
+it into a tidy-style data frame.
 
@@
 prediction band is very large, covering up a large area of the ternary
-plot. On the other hand, for the \pkg{EpiModel} model, we see that the
+plot. On the other hand, for the EpiModel network model, we see that the
 prediction band covers the data quite well and takes up less area.
@@
 much less likely then \textbf{any} of the simulations (reported in Table
-\ref{tab:hags-extreme}). In conclusion, \pkg{EpiCompare} makes it clear
-that, at a glance, 1) the \pkg{EpiModel} network model is a better fit
-than the Simple SIR model, and 2) the fit is only good at the geometric
-filamental level as opposed to the epidemic trajectory filamental level.
+\ref{tab:hags-extreme})\footnote{\textcolor{orange}{Ben, do we want to add another sentence or two explaining the two columns in the table?
+The second one I think makes sense to me but not the first.}}
+In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the
+EpiModel network model is a better fit than the Simple SIR model, and 2)
+the fit is only good at the
+\sout{geometric filamental level as opposed to the epidemic trajectory filamental level.}
+\textcolor{orange}{individual point level as opposed to the geometric filamental level.}
 
@@
 
-\caption{\label{tab:cif-all-together-df}Top and bottom 2 rows of \tt{all\_together\_df}\text{, combining both simulated epidemics and the true epidemic.}}
+\caption{\label{tab:cif-all-together-df}Top and bottom 2 rows of \tt{all\_together\_df}\textnormal{, combining both simulated epidemics and the true epidemic.}}
 \centering
@@
 
-\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing psuedo-density estimates between true vs simulated curves}
+\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing pseudo-density estimates between true vs simulated curves}
 \centering
@@
 \toprule
-Type & simulations-based estimated psuedo-density & proportion of simulations with lower estimated psuedo-density\\
+Type & simulations-based estimated pseudo-density & proportion of simulations with lower estimated pseudo-density\\
 \midrule

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Section 3 highlighting

@@
 In Fig.
-\ref{fig:pipeline2}, we show\footnote{\ben{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
-I'm happy with other synonyms as well.} \shannon{accept}} how our package's functions fit into the data analysis pipeline introduced in Fig.
+\ref{fig:pipeline2}, we show how our package's functions fit into the data analysis pipeline introduced in Fig.
 \ref{fig:pipeline}.
@@
 Before data can be explored, they must be collected.
-\sout{Raw data can appear in many different formats.}\footnote{\ben{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\shannon{I think we can get rid of the 'raw data can appear...' sentence.} RESOLVED 5/17} Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
+Sometimes individual records are collected, with times of different states of the epidemic (infection, recovery, etc.) as well as individual information like network structure, location, and sub-population information.
 Other data collections focus on aggregate counts of individuals in each epidemic state.
@@
 start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
-There are often situations where \ben{\sout{aggregating} grouping}\footnote{RESOLVED 5/17 (aggregating)} agents into subpopulations (e.g.
+There are often situations where grouping agents into subpopulations (e.g.
 subpopulations defined by age or sex) can highlight different aggregate level trends.
 For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`, the tidyverse approach to the split-apply-combine paradigm for data analysis \citep{wickham2011}.\footnote{Changed to try to clarify how how agents_to_aggregate works with group_by and why (split-apply-combine)}
-
-\shannon{old text below}
-[by combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
+In \pkg{EpiCompare}, we facilitate subpopulation analysis\footnote{\shannon{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}} by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`, the tidyverse approach to the split-apply-combine paradigm for data analysis \citep{wickham2011}.
+
+[[\textcolor{red}{OLD SENTENCE}
+by combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
 ]}\shannon{agreed.
-aggregation at a group level}}.]
-
-Besides aiding subpopulation analysis, `agents_to_aggregate()` can  accommodate\footnote{\ben{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
-Again, I'm open to other synonyms.]}\shannon{I like accomodate}  RESOLVED 5/17} a wide range of information about each individual.
+aggregation at a group level}}.
+]]
+
+Besides aiding subpopulation analysis, `agents_to_aggregate()` can  accommodate a wide range of information about each individual.
 In fact, this function can account for infinitely many states.
@@
 
-**Exploratory Data Analysis (EDA)*
+**Exploratory Data Analysis (EDA)**
 
@@
 An expert coder can start with `agents_to_aggregate()` to successfully accomplish exploratory data analysis (EDA) in many ways.
-\pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there \ben{are} three unique epidemiological states (like in the SIR model).
-Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a \footnote{\ben{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \shannon{accepted}.
-RESOLVED 5/17} way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
+\pkg{EpiCompare} also includes tools that allow a novice coder to rapidly explore data, provided there are three unique epidemiological states (like in the SIR model).
+Building on \pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s `geom_aggregate()` provides a way to explore how different subpopulations experience of an epidemic \citep{Wickham2016, Hamilton2018}.
 The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.
-Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a time-invariant\footnote{\ben{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \shannon{agreed} RESOLVED 5/17} representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 
-After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often\footnote{\ben{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence.
-Maybe it's just me.]}RESOLVED 5/17} model fitting and/or simulation.
+After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often model fitting and/or simulation.
 In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.
@@
 
+\shannon{[[NEW TEXT}
 If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format.
 In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.
-In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{new signpost.
+In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.
 There are two paragraphs below corresponding to the two points.
 The second paragraph could use another sentence maybe.
-I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}
+I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}
 
@@
 While incredibly useful, these special classes can be difficult for novice coders to handle.
-\shannon{motivates for fortify but not filamental compression}.
-To this end, \pkg{EpiCompare} provides a series of `fortify`-style methods, called `fortify_aggregate()` which transform output from infectious disease modeling and simulation packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
-These fortify functions have output that is consistent with that of `agents_to_aggregate()`.
+To this end, \pkg{EpiCompare} provides a series of fortify-style methods, called \code{fortify_aggregate()} which transform output from infectious disease modeling and simulation packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
+These fortify functions have output that is consistent with that of \code{agents_to_aggregate()}.
 These standardized outputs can then be piped to summaries, tables, and plots.
 
-Because the fortified output contains temporal information and we provide the function, `filament_compression()`, to transform temporally defined epidemics to their filamental representations.
+Because the fortified output contains temporal information and we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations.
 These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.
-
-[[\shannon{OLD TEXT BELOW}
+\shannon{]]}
+
+[[\textcolor{red}{OLD TEXT BELOW}
 
@@
 **Comparisons and Assessment**
+\shannon{[[ NEW TEXT}
 
@@
 In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
-mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{a bit of an awk transition but I wanted to use past wording about TIA.}  We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of an epidemic has occurred.
+mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of an epidemic has occurred.
 
@@
 
-[[\shannon{OLD TEXT}
+\shannon{]]}
+
+[[\textcolor{red}{OLD TEXT}
 
@@
 
+\shannon{[[NEW TEXT}
+
 Finally, in this section we show how tools from \pkg{EpiCompare} can be used in each step of the data analysis pipeline shown in Fig.
@@
 The data was later made visible by \cite{oesterle1992} and made available in an \proglang{R} by \cite{surveillance2017}.
-This data set includes a rich collection of features and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\shannon{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's goingon.}}
-
-[[\shannon{OLD TEXT}
+This data set includes a rich collection of features and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\shannon{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's going on.}}
+
+\shannon{]]}
+
+[[\textcolor{red}{OLD TEXT}
 

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.tex
Section 3 highlighting

@@
 and explain how they aid in the data analysis pipeline. In Fig.
-\ref{fig:pipeline2}, we
-show\footnote{\textcolor{violet}{[Ben says: I think illustrate isn't the best adjective here (as it suggests a little more like a guided demonstration, which is a little extreme).
-I'm happy with other synonyms as well.} \textcolor{orange}{accept}}
-how our package's functions fit into the data analysis pipeline
-introduced in Fig. \ref{fig:pipeline}. All front-facing functions in
-\pkg{EpiCompare} are aimed to be as user-friendly as possible. We also
-focus on providing the user ``tidyverse'' style functions, that
-encourage piping objects from one function to the next and follow clear
-``verb'' naming schemes \citep{Wickham2019}. Although users can
-incorporate \pkg{EpiCompare} into any step in the data analysis
-pipeline, there are two primary points of entry. The first point of
-entry is the very beginning with pre-processing and visualizing raw
-data, and the second point of entry is after modeling and simulation.
-Figure \ref{fig:pipeline2} captures these different paths, and we
-highlight how to leverage \pkg{EpiCompare} functionalities in the
-subsections below.
+\ref{fig:pipeline2}, we show how our package's functions fit into the
+data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
+front-facing functions in \pkg{EpiCompare} are aimed to be as
+user-friendly as possible. We also focus on providing the user
+``tidyverse'' style functions, that encourage piping objects from one
+function to the next and follow clear ``verb'' naming schemes
+\citep{Wickham2019}. Although users can incorporate \pkg{EpiCompare}
+into any step in the data analysis pipeline, there are two primary
+points of entry. The first point of entry is the very beginning with
+pre-processing and visualizing raw data, and the second point of entry
+is after modeling and simulation. Figure \ref{fig:pipeline2} captures
+these different paths, and we highlight how to leverage \pkg{EpiCompare}
+functionalities in the subsections below.
 
@@
 can be explored. Before data can be explored, they must be collected.
-\sout{Raw data can appear in many different formats.}\footnote{\textcolor{violet}{[Ben says: the previous too sentences are very short and seem choppy when I read them - do they for you as well?]}\textcolor{orange}{I think we can get rid of the 'raw data can appear...' sentence.} RESOLVED 5/17}
 Sometimes individual records are collected, with times of different
@@
 aggregate view of how many individuals were in a state at a given time.
-There are often situations where
-\textcolor{violet}{\sout{aggregating} grouping}\footnote{RESOLVED 5/17 (aggregating)}
-agents into subpopulations (e.g.~subpopulations defined by age or sex)
-can highlight different aggregate level trends. For example, research by
+There are often situations where grouping agents into subpopulations
+(e.g.~subpopulations defined by age or sex) can highlight different
+aggregate level trends. For example, research by
 \citet{rvachev1985,anderson1992,worby2015} develop state-based models
 that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis by extending
-\texttt{agents\_to\_aggregate()} to work with
+In \pkg{EpiCompare}, we facilitate subpopulation
+analysis\footnote{\textcolor{orange}{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}}
+by extending \texttt{agents\_to\_aggregate()} to work with
 \pkg{dplyr}::\texttt{group\_by()}, the tidyverse approach to the
-split-apply-combine paradigm for data analysis
-\citep{wickham2011}.\textbackslash footnote\{Changed to try to clarify
-how how agents\_to\_aggregate works with group\_by and why
-(split-apply-combine)\}
-
-\textcolor{orange}{old text below}
-
-{[}by combining the function \pkg{dplyr}::\texttt{group\_by()} and
-\texttt{agents\_to\_aggregate()} to provide aggregation at a
-subpopulation
-level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed. aggregation at a group level}}.{]}
+split-apply-combine paradigm for data analysis \citep{wickham2011}.
+
+{[}{[}\textcolor{red}{OLD SENTENCE} by combining the function
+\pkg{dplyr}::\texttt{group\_by()} and \texttt{agents\_to\_aggregate()}
+to provide aggregation at a subpopulation
+level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed. aggregation at a group level}}.{]}{]}
 
 Besides aiding subpopulation analysis, \texttt{agents\_to\_aggregate()}
-can
-accommodate\footnote{\textcolor{violet}{[Ben says: I think the original word is a bit vague it as "deal with" could mean "accommodate" or "avoid traps associated with", etc.
-Again, I'm open to other synonyms.]}\textcolor{orange}{I like accomodate}  RESOLVED 5/17}
-a wide range of information about each individual. In fact, this
-function can account for infinitely many states. This functionality
-allows the practitioner to aggregate information relative to common
-states (e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well
-as more complex states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'').
-Additionally, \texttt{agents\_to\_aggregate()} permits indicators for
-death/exit and birth/entry dates. Overall, this function is a powerful
-tool for pre-processing data.
-
-**Exploratory Data Analysis (EDA)*
+can accommodate a wide range of information about each individual. In
+fact, this function can account for infinitely many states. This
+functionality allows the practitioner to aggregate information relative
+to common states (e.g.~``Susceptible'', ``Infectious'', and
+``Recovered'') as well as more complex states (e.g.~``Exposed'',
+``iMmune'', ``Hospitalized''). Additionally,
+\texttt{agents\_to\_aggregate()} permits indicators for death/exit and
+birth/entry dates. Overall, this function is a powerful tool for
+pre-processing data.
+
+\textbf{Exploratory Data Analysis (EDA)}
 
@@
 \pkg{EpiCompare} also includes tools that allow a novice coder to
-rapidly explore data, provided there \textcolor{violet}{are} three
-unique epidemiological states (like in the SIR model). Building on
-\pkg{ggplot2} and \pkg{ggtern} packages, \pkg{EpiCompare}'s
-\texttt{geom\_aggregate()} provides a
-\footnote{\textcolor{violet}{[Ben says: this word is used in the sentence before, and I don't think it adds too much to this sentence.]} \textcolor{orange}{accepted}. RESOLVED 5/17}
-way to explore how different subpopulations experience of an epidemic
-\citep{Wickham2016, Hamilton2018}. The function
+rapidly explore data, provided there are three unique epidemiological
+states (like in the SIR model). Building on \pkg{ggplot2} and
+\pkg{ggtern} packages, \pkg{EpiCompare}'s \texttt{geom\_aggregate()}
+provides a way to explore how different subpopulations experience of an
+epidemic \citep{Wickham2016, Hamilton2018}. The function
 \texttt{geom\_aggregate()} provides a visualization tool to holistically
@@
 and basic epidemic state-based models and (2) our three-dimensional
-simplex representation of these epidemics emphasizes a
-time-invariant\footnote{\textcolor{violet}{[Ben says: I don't think we really need quotation marks here as we've discussed the idea in the previous sections.]} \textcolor{orange}{agreed} RESOLVED 5/17}
+simplex representation of these epidemics emphasizes a time-invariant
 representation of the data (for a refresher see Section
@@
 After getting a sense of what a past or current epidemic looks like with
-EDA, the next step in the data analysis pipeline is
-often\footnote{\textcolor{violet}{[Ben says: By moving this word here I found, as a reader, I didn't have as many pauses while reading this sentence. Maybe it's just me.]}RESOLVED 5/17}
-model fitting and/or simulation. In this step and the next step
-(post-processing), we highlight how practitioners can pair models and
-simulations of epidemics from outside of \pkg{EpiCompare} with analysis
-and simulation tools in \pkg{EpiCompare}. While \pkg{EpiCompare} does
-not focus on fitting models to data, we do provide some flexible
-functions for simulation of basic discrete-time epidemic-state models.
-These functions simulate individual-level information based on
-practitioner estimated transition rates between states and can be
-combined with \texttt{agents\_to\_aggregate()} to view these simulations
-through an aggregate lens. The function \texttt{simulate\_SIR\_agents()}
-simulates a basic SIR epidemic with user inputs for the number of
-simulations, the initial number in each state, the infection and
-recovery parameters \((\beta, \gamma\)), and the total number of
-discrete time steps. This function allows for easy access to SIR model
-analysis and comparison. Beyond SIR models, the function
-\texttt{simulate\_agents()} takes as input a user-specified
-state-transition matrix and other epidemic parameters to allow the user
-to create simulations for an outbreak with \textit{any} number of states
-and any number of transitions among them. This flexibility in states can
-be used to also reflect group-based dynamics. In turn, this allows for
-users to explore the space of models in an intuitive way without getting
-bogged down by too much mathematical detail. For consistency, we have
-made output from \texttt{simulate\_agents()} and
-\texttt{simulate\_SIR\_agents()} compatible with
-\texttt{agents\_to\_aggregate()} so aggregate information may easily be
-accessed.
+EDA, the next step in the data analysis pipeline is often model fitting
+and/or simulation. In this step and the next step (post-processing), we
+highlight how practitioners can pair models and simulations of epidemics
+from outside of \pkg{EpiCompare} with analysis and simulation tools in
+\pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting
+models to data, we do provide some flexible functions for simulation of
+basic discrete-time epidemic-state models. These functions simulate
+individual-level information based on practitioner estimated transition
+rates between states and can be combined with
+\texttt{agents\_to\_aggregate()} to view these simulations through an
+aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
+a basic SIR epidemic with user inputs for the number of simulations, the
+initial number in each state, the infection and recovery parameters
+\((\beta, \gamma\)), and the total number of discrete time steps. This
+function allows for easy access to SIR model analysis and comparison.
+Beyond SIR models, the function \texttt{simulate\_agents()} takes as
+input a user-specified state-transition matrix and other epidemic
+parameters to allow the user to create simulations for an outbreak with
+\textit{any} number of states and any number of transitions among them.
+This flexibility in states can be used to also reflect group-based
+dynamics. In turn, this allows for users to explore the space of models
+in an intuitive way without getting bogged down by too much mathematical
+detail. For consistency, we have made output from
+\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
+compatible with \texttt{agents\_to\_aggregate()} so aggregate
+information may easily be accessed.
 
 \textbf{Post-processing}
+
+\textcolor{orange}{[[NEW TEXT}
 
@@
 model output into a format amenable to time-invariant
-analysis.\footnote{new signpost.
+analysis.\footnote{\textcolor{orange}{new signpost.
 There are two paragraphs below corresponding to the two points.
 The second paragraph could use another sentence maybe.
-I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}
+I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}
 
@@
 more. While incredibly useful, these special classes can be difficult
-for novice coders to handle.
-\textcolor{orange}{motivates for fortify but not filamental compression}.
-To this end, \pkg{EpiCompare} provides a series of
-\texttt{fortify}-style methods, called \texttt{fortify\_aggregate()}
-which transform output from infectious disease modeling and simulation
+for novice coders to handle. To this end, \pkg{EpiCompare} provides a
+series of fortify-style methods, called \code{fortify_aggregate()} which
+transform output from infectious disease modeling and simulation
 packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
@@
 have output that is consistent with that of
-\texttt{agents\_to\_aggregate()}. These standardized outputs can then be
+\code{agents_to_aggregate()}. These standardized outputs can then be
 piped to summaries, tables, and plots.
@@
 Because the fortified output contains temporal information and we
-provide the function, \texttt{filament\_compression()}, to transform
+provide the function, \code{filament_compression()}, to transform
 temporally defined epidemics to their filamental representations. These
@@
 further time-invariant analysis tools described below.
-
-{[}{[}\textcolor{orange}{OLD TEXT BELOW}
+\textcolor{orange}{]]}
+
+{[}{[}\textcolor{red}{OLD TEXT BELOW}
 
@@
 
-\textbf{Comparisons and Assessment}
+\textbf{Comparisons and Assessment} \textcolor{orange}{[[ NEW TEXT}
 
@@
 or AIC) and into the lens of time-invariant
-analysis.\footnote{a bit of an awk transition but I wanted to use past wording about TIA.}
+analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}
 We have found these tools to be specifically applicable for situations
@@
 
-{[}{[}\textcolor{orange}{OLD TEXT}
+\textcolor{orange}{]]}
+
+{[}{[}\textcolor{red}{OLD TEXT}
 
@@
 
+\textcolor{orange}{[[NEW TEXT}
+
 Finally, in this section we show how tools from \pkg{EpiCompare} can be
@@
 disease epidemiology
-\cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\textcolor{orange}{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's goingon.}}
-
-{[}{[}\textcolor{orange}{OLD TEXT}
+\cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\textcolor{orange}{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's going on.}}
+
+\textcolor{orange}{]]}
+
+{[}{[}\textcolor{red}{OLD TEXT}
 

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper2.tex
Section 3 highlighting

@@
 time-invariant, and can be plugged into multiple distance-based
-assessment tools to examine the overall ``extremeness'' of points,
+assessment tools to examine the overall ``extremeness'\,' of points,
 including pseudo-density estimators and depth/local depth functions

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
Pushing.  Edits and edits accepted for Sec 1.
Couldn't get two footnotes to compile.

1.  Changed 'aid' to help.  Does 'to aid an individual decide the correct approach?' make grammatical sense?  I may be overthinking.

2.  'extend the user toolkit' - wanted to remove the 'give' part.

@@
 historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}.
-At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process \ben{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process, as has been seen in the current COVID-19 \shannon{\sout{epi}}\shannon{pan}demic \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
 \footnote{\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.
-*This is less so the case now.]}}
+*This is less so the case now.]}\shannon{Seems fine to me now}}
 
-With the many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021}, this field has a lot of devices to aid an individual practitioner decide the correct approach.
+With \shannon{\sout{the}} many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} \shannon{available}, this field has a \shannon{\sout{lot of}} \shannon{number of} devices to \shannon{\sout{aid}} \shannon{help}  an individual practitioner decide the correct approach.
 For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
 
-At the same time, analysis packages often only address a specific portion of the analysis pipeline\ben{\sout{, for instance focusing on certain types of models.}} \ben{These m}odeling tools\ben{\sout{, which}} usually require learning package-specific syntax\ben{\sout{,}} and often don't provide easy ways to compare and assess their models on new data.
-Moreover, exploring\ben{, \sout{and}} modeling \ben{and comparing} epidemics require transforming and \textit{tidying} data in different ways.
+At the same time, analysis packages often only address a specific portion of the analysis pipeline.
+These modeling tools usually require learning package-specific syntax and often don't provide easy ways to compare and assess their models on new data.
+Moreover, exploring modeling and comparing epidemics require transforming and \textit{tidying} data in different ways.
 To fill these gaps, we present our \proglang{R} package \pkg{EpiCompare}.
@@
 \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.
-
-<!-- # old draft
-The recent (and currently on-going) COVID-19 global pandemic has galvanized public interest in understanding more about infectious disease modeling and has highlighted the usefulness of research in the area of infectious disease epidemiology.
-Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g.
-flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g.
-historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.
-...][]{}.
-The COVID-19 pandemic highlights how all three goals are important both separately and taken as a whole.
-Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.
-Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks.
-\shannon{really like this paragraph}
-
-The current epidemic of COVID-19 also highlights that infectious disease  models are only one piece of the overall analysis pipeline.
-University based resources like John Hopkins' and government numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and visualization can be a important first step in the process (multiple \cite{}?).} \shannon{add comment about NYT or something} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individual / agent level, network structure or just aggregate flows (review paper \cite{}?).
-All placing individuals / proportions of the populations into different states (e.g.
-suspectible, exposed, infected, recovered, etc.).
-With all these models, review and comparison papers in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center helps the individual practitioner decide the correct approach.
-\shannon{along with expertise from healthcare professionals...}
-
-At the same time, analysis packages often only address a portion of the analysis pipeline.
-Modeling tools often don't provide easy ways to compare and assess their models on new data.
-Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways.
-To fill these gaps, we present our our \proglang{R} package \pkg{EpiCompare}.
-Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models.
-In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next.
-\pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.
--->
 
@@
 
-[One of \pkg{EpiCompare}’s main contribution to] comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments.
-Epidemics, despite being defined a a process that evolves overt time, often needs to be compared in a way not constrained to initial times or time scales to understand the processes at play.] [With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible.]\footnote{\ben{[Ben says: brought this sentence from a later paragraph as it’s much more introductory.]}} Compared to time-dependent comparison tools for state-space modeling,  [time-invariant analysis can make it easier to compare state-space \ben{\sout{models} epidemic representations} in a more global, holistic fashion.] Many time-dependent comparison tools only [examine the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.] [These \ben{time-dependent approaches can} reduce the amount of \ben{\sout{connections that can be seen} insights that can be drawn}, similar to \ben{examining} projections of a multidimensional distribution onto a single axis \ben{one} at a time.
-Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens.
-\ben{At the same time, t}he goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.]
+One of \pkg{EpiCompare}’s main contribution to comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments.
+Epidemics, despite being defined as a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales \shannon{in order}\footnote{\shannon{wanted a clearer signal that this is a bridge to a new thought}}to understand the processes at play.
+With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible.
+Compared to time-dependent comparison tools for state-space modeling,  time-invariant analysis can make it easier to compare state-space epidemic representations in a more global, holistic fashion.
+Many time-dependent comparison tools only examine the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
+These time-dependent approaches can reduce the amount of connections that can be seen and insights that can be drawn}, similar to examining projections of a multidimensional distribution onto a single axis, one at a time.
+Tools in \pkg{EpiCompare} \shannon{\sout{give the user the ability to extend their} \shannon{extend the user}} toolkit to evaluate epidemics within a time-invariant lens.
+At the same time, the goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
 
-This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a \ben{\sout{thorough}} demonstrating of the tools through a detailed example of a full data analysis pipeline.
+This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a  demonstration of the tools through a detailed example\shannon{\sout{ of a full data analysis pipeline.}} \shannon{from start to finish of the data analysis pipeline}\footnote{\shannon{maybe clearer?
+up to you}}
 
-# Time-invariant analysis
+# Time-invariant analysis {#sec:time-invariant}
 

### 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper2.tex
Pushing.  Edits and edits accepted for Sec 1.
Couldn't get two footnotes to compile.

1.  Changed 'aid' to help.  Does 'to aid an individual decide the correct approach?' make grammatical sense?  I may be overthinking.

2.  'extend the user toolkit' - wanted to remove the 'give' part.

@@
 government, and news organizations are an important first step of the
-process
-\textcolor{violet}{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
-\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}}
-
-With the many visualization and exploratory tools, models and modeling
-paradigms, and reviews and comparisons in the literature and through the
-MIDAS (Models of Infectious Disease Agent Study) network
-\citep{midasnetwork2021}, this field has a lot of devices to aid an
-individual practitioner decide the correct approach. For example,
-\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
-\pkg{pomp} have all made significant steps in standardizing the flow of
-the data analysis pipeline for epidemic modeling through digitizing data
-sets, making accessible statistical models, and providing a plethora of
-educational material for both coding novices and experts alike
-\citep{surveillance2017,Jenness2018,King2016}.
+process, as has been seen in the current COVID-19
+\textcolor{orange}{\sout{epi}}\textcolor{orange}{pan}demic
+\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}\textcolor{orange}{Seems fine to me now}}
+
+With \textcolor{orange}{\sout{the}} many visualization and exploratory
+tools, models and modeling paradigms, and reviews and comparisons in the
+literature and through the MIDAS (Models of Infectious Disease Agent
+Study) network \citep{midasnetwork2021} \textcolor{orange}{available},
+this field has a \textcolor{orange}{\sout{lot of}}
+\textcolor{orange}{number of} devices to \textcolor{orange}{\sout{aid}}
+\textcolor{orange}{help} an individual practitioner decide the correct
+approach. For example, \proglang{R} packages such as \pkg{surveillance},
+\pkg{EpiModel}, and \pkg{pomp} have all made significant steps in
+standardizing the flow of the data analysis pipeline for epidemic
+modeling through digitizing data sets, making accessible statistical
+models, and providing a plethora of educational material for both coding
+novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
 
 At the same time, analysis packages often only address a specific
-portion of the analysis
-pipeline\textcolor{violet}{\sout{, for instance focusing on certain types of models.}}
-\textcolor{violet}{These m}odeling
-tools\textcolor{violet}{\sout{, which}} usually require learning
-package-specific syntax\textcolor{violet}{\sout{,}} and often don't
-provide easy ways to compare and assess their models on new data.
-Moreover, exploring\textcolor{violet}{, \sout{and}} modeling
-\textcolor{violet}{and comparing} epidemics require transforming and
+portion of the analysis pipeline. These modeling tools usually require
+learning package-specific syntax and often don't provide easy ways to
+compare and assess their models on new data. Moreover, exploring
+modeling and comparing epidemics require transforming and
 \textit{tidying} data in different ways. To fill these gaps, we present
@@
 
-{[}One of \pkg{EpiCompare}'s main contribution to{]} comparison and
-assessment of epidemics is through tools that provide
-\textit{time-invariant} assessments. Epidemics, despite being defined a
-a process that evolves overt time, often needs to be compared in a way
-not constrained to initial times or time scales to understand the
-processes at play.{]} {[}With time-invariant analysis, comparing
-decades-long outbreaks of HIV in the US to a 10 day outbreak of
-norovirus on a cruise ship is
-possible.{]}\footnote{\textcolor{violet}{[Ben says: brought this sentence from a later paragraph as it’s much more introductory.]}}
-Compared to time-dependent comparison tools for state-space modeling,
-{[}time-invariant analysis can make it easier to compare state-space
-\textcolor{violet}{\sout{models} epidemic representations} in a more
-global, holistic fashion.{]} Many time-dependent comparison tools only
-{[}examine the proportion of individuals in each state (at a given time)
-in a piece-wise / marginal fashion.{]} {[}These
-\textcolor{violet}{time-dependent approaches can} reduce the amount of
-\textcolor{violet}{\sout{connections that can be seen} insights that can be drawn},
-similar to \textcolor{violet}{examining} projections of a
-multidimensional distribution onto a single axis \textcolor{violet}{one}
-at a time. Tools in \pkg{EpiCompare} give the user the ability to extend
-their toolkit to evaluate epidemics within a time-invariant lens.
-\textcolor{violet}{At the same time, t}he goal of \pkg{EpiCompare} is
-not to supplant existing infectious disease modeling tools and software
-but, rather, is a concerted effort to create standard and fair
-comparisons among models developed for disease outbreaks and outbreak
-data.{]}
+One of \pkg{EpiCompare}'s main contribution to comparison and assessment
+of epidemics is through tools that provide \textit{time-invariant}
+assessments. Epidemics, despite being defined as a process that evolves
+over time, often need to be compared in a way not constrained to initial
+times or time scales
+\textcolor{orange}{in order}\footnote{\textcolor{orange}{wanted a clearer signal that this is a bridge to a new thought}}to
+understand the processes at play. With time-invariant analysis,
+comparing decades-long outbreaks of HIV in the US to a 10 day outbreak
+of norovirus on a cruise ship is possible. Compared to time-dependent
+comparison tools for state-space modeling, time-invariant analysis can
+make it easier to compare state-space epidemic representations in a more
+global, holistic fashion. Many time-dependent comparison tools only
+examine the proportion of individuals in each state (at a given time) in
+a piece-wise / marginal fashion. These time-dependent approaches can
+reduce the amount of connections that can be seen and insights that can
+be drawn\}, similar to examining projections of a multidimensional
+distribution onto a single axis, one at a time. Tools in
+\pkg{EpiCompare}
+\textcolor{orange}{\sout{give the user the ability to extend their} \textcolor{orange}{extend the user}}
+toolkit to evaluate epidemics within a time-invariant lens. At the same
+time, the goal of \pkg{EpiCompare} is not to supplant existing
+infectious disease modeling tools and software but, rather, is a
+concerted effort to create standard and fair comparisons among models
+developed for disease outbreaks and outbreak data.
 
@@
 \pkg{EpiCompare} aids a practitioner in every step of the pipeline and
-section \ref{sec:tour} provides a \textcolor{violet}{\sout{thorough}}
-demonstrating of the tools through a detailed example of a full data
-analysis pipeline.
-
-\hypertarget{time-invariant-analysis}{%
-\section{Time-invariant analysis}\label{time-invariant-analysis}}
+section \ref{sec:tour} provides a demonstration of the tools through a
+detailed
+example\textcolor{orange}{\sout{ of a full data analysis pipeline.}}
+\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer? up to you}}
+
+\hypertarget{sec:time-invariant}{%
+\section{Time-invariant analysis}\label{sec:time-invariant}}
 

### 2021-05-18 jss/EpiCompare-paper/EpiCompare-paper.Rmd
comments and updates to section 3

@@
 In this section, we present the tools implemented in \pkg{EpiCompare} and explain how they aid in the  data analysis pipeline.
-In Fig.
-\ref{fig:pipeline2}, we show how our package's functions fit into the data analysis pipeline introduced in Fig.
-\ref{fig:pipeline}.
+In Figure \ref{fig:pipeline2}, we show how our package's functions fit into the data analysis pipeline introduced in Figure \ref{fig:pipeline}.
 All front-facing functions in \pkg{EpiCompare} are aimed to be as user-friendly as possible.
@@
 
-**Data Pre-processing**
+**Data pre-processing**
 
@@
 weekly case counts) is observed.
-Compartment totals (amounts of individuals in each state) are then imputed from those case counts other information about the disease and the population of interest.
+\ben{In this setting, c}ompartment totals (amounts of individuals in each state) are then imputed from those case counts \ben{and using} other information about the disease and the population of interest\footnote{\ben{[Ben says: updates to connect to the previous sentence more clearly and the second was grammar.]}}.
 In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.
@@
 start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
-There are often situations where grouping agents into subpopulations (e.g.
+\footnote{\ben{[Ben says: I'd like to replace these two sentences will a new one.
+The original goal was to make a strong connection between the first sentence "There are..." to the previous sentence, but this lead to a rewrite (which I know isn't best).
+Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregating conditional on subpopulations (e.g.
+subpopulations defined by age or sex).]}}There are often situations where grouping agents into subpopulations (e.g.
 subpopulations defined by age or sex) can highlight different aggregate level trends.
@@
 [[\textcolor{red}{OLD SENTENCE}
-by combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
+by combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()` to provide \ben{group level aggregation\sout{aggregation at a subpopulation level}}\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
 ]}\shannon{agreed.
-aggregation at a group level}}.
+aggregation at a group level} \ben{[Ben says: I think the replacement doesn't clean things up - I probably wasn't clear that I actually don't think split-apply-combine is well known.
+My edit to the old sentence was aimed to making it clear the process through our words "group-then-aggregate".]}}.
 ]]
@@
 
-**Exploratory Data Analysis (EDA)**
+**Exploratory data analysis (EDA)**
 
@@
 The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.
-Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).
-
-**Model Fitting and Simulations**
+Visualization tools for three-state models were developed because \ben{\sout{(1)}} SIR models are some of the most common and basic epidemic state-based models and \ben{\sout{(2)}}\footnote{\ben{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}} our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+
+**Model fitting and simulations**
 
 After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often model fitting and/or simulation.
-In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.
-While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
+\ben{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\ben{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}} While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
 These functions simulate individual-level information based on practitioner estimated transition rates between states and can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
 The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
-This function allows for easy access to SIR model analysis and comparison.
-Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.
+\ben{\sout{This function allows for easy access to SIR model analysis and comparison.}}\footnote{\ben{[Ben says: see edit below as well.
+This sentence seemed misplaced, and we have a very similar sentence a few lines later.
+I'm happy to be overruled here.]}}  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.
 This flexibility in states can be used to also reflect group-based dynamics.
-In turn, this allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
+\ben{\sout{In turn, this allows for} Both of these functions allow} users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
 For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.
@@
 If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format.
-In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.
-In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.
+\ben{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\ben{[Ben says: this really isn't true anymore - that's more the next section.]}} In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output\footnote{\ben{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}} from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.
 There are two paragraphs below corresponding to the two points.
@@
 
-Because the fortified output contains temporal information and we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations.
+Because \ben{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\ben{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]}} we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations.
 These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.
@@
 
-**Comparisons and Assessment**
+**Comparisons and assessment**
 \shannon{[[ NEW TEXT}
 
-Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare different models, and simulations to one another.
+Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare different models, and simulations to one another\footnote{\ben{[Ben says: I'm confused why this is just about comparison and not assessment.
+Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
 In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.

### 2021-05-18 jss/EpiCompare-paper/EpiCompare-paper.tex
comments and updates to section 3

@@
 In this section, we present the tools implemented in \pkg{EpiCompare}
-and explain how they aid in the data analysis pipeline. In Fig.
+and explain how they aid in the data analysis pipeline. In Figure
 \ref{fig:pipeline2}, we show how our package's functions fit into the
-data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
+data analysis pipeline introduced in Figure \ref{fig:pipeline}. All
 front-facing functions in \pkg{EpiCompare} are aimed to be as
@@
 
-\textbf{Data Pre-processing}
+\textbf{Data pre-processing}
 
@@
 of new infections at each time step (e.g.~weekly case counts) is
-observed. Compartment totals (amounts of individuals in each state) are
-then imputed from those case counts other information about the disease
-and the population of interest. In \pkg{EpiCompare}, we focus on
-understanding the overall impact of an outbreak at the
-aggregate/population level, which allows for streamlined examination of
-overall trends of an epidemic.
+observed. \textcolor{violet}{In this setting, c}ompartment totals
+(amounts of individuals in each state) are then imputed from those case
+counts \textcolor{violet}{and using} other information about the disease
+and the population of
+interest\footnote{\textcolor{violet}{[Ben says: updates to connect to the previous sentence more clearly and the second was grammar.]}}.
+In \pkg{EpiCompare}, we focus on understanding the overall impact of an
+outbreak at the aggregate/population level, which allows for streamlined
+examination of overall trends of an epidemic.
 
@@
 aggregate view of how many individuals were in a state at a given time.
-There are often situations where grouping agents into subpopulations
+\footnote{\textcolor{violet}{[Ben says: I'd like to replace these two sentences will a new one.
+The original goal was to make a strong connection between the first sentence "There are..." to the previous sentence, but this lead to a rewrite (which I know isn't best).
+Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregating conditional on subpopulations (e.g.
+subpopulations defined by age or sex).]}}There
+are often situations where grouping agents into subpopulations
 (e.g.~subpopulations defined by age or sex) can highlight different
@@
 \pkg{dplyr}::\texttt{group\_by()} and \texttt{agents\_to\_aggregate()}
-to provide aggregation at a subpopulation
-level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed. aggregation at a group level}}.{]}{]}
+to provide
+\textcolor{violet}{group level aggregation\sout{aggregation at a subpopulation level}}\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation".
+]}\textcolor{orange}{agreed.
+aggregation at a group level} \textcolor{violet}{[Ben says: I think the replacement doesn't clean things up - I probably wasn't clear that I actually don't think split-apply-combine is well known.
+My edit to the old sentence was aimed to making it clear the process through our words "group-then-aggregate".]}}.{]}{]}
 
@@
 
-\textbf{Exploratory Data Analysis (EDA)}
+\textbf{Exploratory data analysis (EDA)}
 
@@
 three-dimensional state space. Visualization tools for three-state
-models were developed because (1) SIR models are some of the most common
-and basic epidemic state-based models and (2) our three-dimensional
-simplex representation of these epidemics emphasizes a time-invariant
-representation of the data (for a refresher see Section
-\ref{sec:time-invariant}).
-
-\textbf{Model Fitting and Simulations}
+models were developed because \textcolor{violet}{\sout{(1)}} SIR models
+are some of the most common and basic epidemic state-based models and
+\textcolor{violet}{\sout{(2)}}\footnote{\textcolor{violet}{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}}
+our three-dimensional simplex representation of these epidemics
+emphasizes a time-invariant representation of the data (for a refresher
+see Section \ref{sec:time-invariant}).
+
+\textbf{Model fitting and simulations}
 
@@
 EDA, the next step in the data analysis pipeline is often model fitting
-and/or simulation. In this step and the next step (post-processing), we
-highlight how practitioners can pair models and simulations of epidemics
-from outside of \pkg{EpiCompare} with analysis and simulation tools in
-\pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting
-models to data, we do provide some flexible functions for simulation of
-basic discrete-time epidemic-state models. These functions simulate
-individual-level information based on practitioner estimated transition
-rates between states and can be combined with
-\texttt{agents\_to\_aggregate()} to view these simulations through an
-aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
-a basic SIR epidemic with user inputs for the number of simulations, the
-initial number in each state, the infection and recovery parameters
-\((\beta, \gamma\)), and the total number of discrete time steps. This
-function allows for easy access to SIR model analysis and comparison.
+and/or simulation.
+\textcolor{violet}{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\textcolor{violet}{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}}
+While \pkg{EpiCompare} does not focus on fitting models to data, we do
+provide some flexible functions for simulation of basic discrete-time
+epidemic-state models. These functions simulate individual-level
+information based on practitioner estimated transition rates between
+states and can be combined with \texttt{agents\_to\_aggregate()} to view
+these simulations through an aggregate lens. The function
+\texttt{simulate\_SIR\_agents()} simulates a basic SIR epidemic with
+user inputs for the number of simulations, the initial number in each
+state, the infection and recovery parameters \((\beta, \gamma\)), and
+the total number of discrete time steps.
+\textcolor{violet}{\sout{This function allows for easy access to SIR model analysis and comparison.}}\footnote{\textcolor{violet}{[Ben says: see edit below as well.
+This sentence seemed misplaced, and we have a very similar sentence a few lines later.
+I'm happy to be overruled here.]}}
 Beyond SIR models, the function \texttt{simulate\_agents()} takes as
@@
 This flexibility in states can be used to also reflect group-based
-dynamics. In turn, this allows for users to explore the space of models
-in an intuitive way without getting bogged down by too much mathematical
-detail. For consistency, we have made output from
-\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
-compatible with \texttt{agents\_to\_aggregate()} so aggregate
-information may easily be accessed.
+dynamics.
+\textcolor{violet}{\sout{In turn, this allows for} Both of these functions allow}
+users to explore the space of models in an intuitive way without getting
+bogged down by too much mathematical detail. For consistency, we have
+made output from \texttt{simulate\_agents()} and
+\texttt{simulate\_SIR\_agents()} compatible with
+\texttt{agents\_to\_aggregate()} so aggregate information may easily be
+accessed.
 
@@
 models-to-models, they need to post-process their models and simulations
-to disseminate the results in an easily digestible format. In general,
-post-processing of modeling and simulation consists of making summary
-statistics, plots, and tables. In \pkg{EpiCompare}, we provide (1)
-functions to standardize simulation and model output from external
-packages and (2) a function to transform standardized simulation and
-model output into a format amenable to time-invariant
+to disseminate the results in an easily digestible format.
+\textcolor{violet}{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\textcolor{violet}{[Ben says: this really isn't true anymore - that's more the next section.]}}
+In \pkg{EpiCompare}, we provide (1) functions to standardize simulation
+and model
+output\footnote{\textcolor{violet}{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}}
+from external packages and (2) a function to transform standardized
+simulation and model output into a format amenable to time-invariant
 analysis.\footnote{\textcolor{orange}{new signpost.
@@
 
-Because the fortified output contains temporal information and we
-provide the function, \code{filament_compression()}, to transform
+Because
+\textcolor{violet}{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\textcolor{violet}{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]}}
+we provide the function, \code{filament_compression()}, to transform
 temporally defined epidemics to their filamental representations. These
@@
 
-\textbf{Comparisons and Assessment} \textcolor{orange}{[[ NEW TEXT}
+\textbf{Comparisons and assessment} \textcolor{orange}{[[ NEW TEXT}
 
@@
 summary statistics that are used to compare different models, and
-simulations to one another. In \pkg{EpiCompare} we provide a set of
-comparison and assessment tools for model and simulation results that
-extend beyond the standard performance metrics (e.g.~mean squared error
-or AIC) and into the lens of time-invariant
+simulations to one
+another\footnote{\textcolor{violet}{[Ben says: I'm confused why this is just about comparison and not assessment.
+Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
+In \pkg{EpiCompare} we provide a set of comparison and assessment tools
+for model and simulation results that extend beyond the standard
+performance metrics (e.g.~mean squared error or AIC) and into the lens
+of time-invariant
 analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}

### 2021-05-20 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Responses to Ben's Sec 3 edits

@@
 weekly case counts) is observed.
-\ben{In this setting, c}ompartment totals (amounts of individuals in each state) are then imputed from those case counts \ben{and using} other information about the disease and the population of interest\footnote{\ben{[Ben says: updates to connect to the previous sentence more clearly and the second was grammar.]}}.
+In this setting, compartment totals (amounts of individuals in each state) are then imputed from those case counts and using other information about the disease and the population of interest.
 In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.
@@
 The original goal was to make a strong connection between the first sentence "There are..." to the previous sentence, but this lead to a rewrite (which I know isn't best).
-Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregating conditional on subpopulations (e.g.
-subpopulations defined by age or sex).]}}There are often situations where grouping agents into subpopulations (e.g.
+Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregati\shannon{on}, conditional on subpopulations (e.g.
+subpopulations defined by age or sex).]}\shannon{I like your edits}}There are often situations where grouping agents into subpopulations (e.g.
 subpopulations defined by age or sex) can highlight different aggregate level trends.
 For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis\footnote{\shannon{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}} by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`, the tidyverse approach to the split-apply-combine paradigm for data analysis \citep{wickham2011}.
+In \pkg{EpiCompare}, we facilitate subpopulation analysis\footnote{\shannon{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}} by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`.\sout{, the tidyverse approach to the split-apply-combine paradigm for data analysis} \citep{wickham2011}.
 
@@
 The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.
-Visualization tools for three-state models were developed because \ben{\sout{(1)}} SIR models are some of the most common and basic epidemic state-based models and \ben{\sout{(2)}}\footnote{\ben{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}} our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+Visualization tools for three-state models were developed because \ben{\sout{(1)}} SIR models are some of the most common and basic epidemic state-based models and \ben{\sout{(2)}}\footnote{\ben{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}\shannon{accepted}} our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).
 
@@
 After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often model fitting and/or simulation.
-\ben{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\ben{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}} While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
+\ben{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\ben{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}\shannon{accept}} While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
 These functions simulate individual-level information based on practitioner estimated transition rates between states and can be combined with `agents_to_aggregate()` to view these simulations through an aggregate lens.
@@
 This sentence seemed misplaced, and we have a very similar sentence a few lines later.
-I'm happy to be overruled here.]}}  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.
+I'm happy to be overruled here.]}\shannon{accept, yeah it's unnecessary}}  Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.
 This flexibility in states can be used to also reflect group-based dynamics.
-\ben{\sout{In turn, this allows for} Both of these functions allow} users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
+Both of these functions allow users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
 For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.
@@
 If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format.
-\ben{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\ben{[Ben says: this really isn't true anymore - that's more the next section.]}} In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output\footnote{\ben{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}} from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.
+\ben{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\ben{[Ben says: this really isn't true anymore - that's more the next section.]}\shannon{accept}} In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model \shannon{objects}\ sout{output} \footnote{\ben{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}\shannon{trying out objects..}} from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.
 There are two paragraphs below corresponding to the two points.
@@
 
-Modeling and simulation output can be very complex objects, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.
+Modeling and simulation \shannon{objects} \sout{output} can be very complex objects, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.
 The special classes often contain a plethora of information about residuals, model diagnostics, input parameters, and more.
@@
 
-Because \ben{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\ben{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]}} we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations.
+Because \ben{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\ben{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]} \shannon{accept}} we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations.
 These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.
@@
 
-Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare different models, and simulations to one another\footnote{\ben{[Ben says: I'm confused why this is just about comparison and not assessment.
+\shannon{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.
+We first talk about the step in the pipeline and then how epicompare fits into that pipeline.
+The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.
+I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}
+
+Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare \shannon{and assess} different models, and simulations to one another\footnote{\ben{[Ben says: I'm confused why this is just about comparison and not assessment.
 Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
 In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
-mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of an epidemic has occurred.
+mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one season or cycle of an epidemic has occurred.
 

### 2021-05-20 jss/EpiCompare-paper/EpiCompare-paper.tex
Responses to Ben's Sec 3 edits

@@
 of new infections at each time step (e.g.~weekly case counts) is
-observed. \textcolor{violet}{In this setting, c}ompartment totals
-(amounts of individuals in each state) are then imputed from those case
-counts \textcolor{violet}{and using} other information about the disease
-and the population of
-interest\footnote{\textcolor{violet}{[Ben says: updates to connect to the previous sentence more clearly and the second was grammar.]}}.
-In \pkg{EpiCompare}, we focus on understanding the overall impact of an
+observed. In this setting, compartment totals (amounts of individuals in
+each state) are then imputed from those case counts and using other
+information about the disease and the population of interest. In
+\pkg{EpiCompare}, we focus on understanding the overall impact of an
 outbreak at the aggregate/population level, which allows for streamlined
@@
 The original goal was to make a strong connection between the first sentence "There are..." to the previous sentence, but this lead to a rewrite (which I know isn't best).
-Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregating conditional on subpopulations (e.g.
-subpopulations defined by age or sex).]}}There
+Suggestion: "Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregati\textcolor{orange}{on}, conditional on subpopulations (e.g.
+subpopulations defined by age or sex).]}\textcolor{orange}{I like your edits}}There
 are often situations where grouping agents into subpopulations
@@
 by extending \texttt{agents\_to\_aggregate()} to work with
-\pkg{dplyr}::\texttt{group\_by()}, the tidyverse approach to the
-split-apply-combine paradigm for data analysis \citep{wickham2011}.
+\pkg{dplyr}::\texttt{group\_by()}.\sout{, the tidyverse approach to the split-apply-combine paradigm for data analysis}
+\citep{wickham2011}.
 
@@
 are some of the most common and basic epidemic state-based models and
-\textcolor{violet}{\sout{(2)}}\footnote{\textcolor{violet}{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}}
+\textcolor{violet}{\sout{(2)}}\footnote{\textcolor{violet}{[Ben says: I don't think this is necessary anymore, reads more clearly without these.]}\textcolor{orange}{accepted}}
 our three-dimensional simplex representation of these epidemics
@@
 and/or simulation.
-\textcolor{violet}{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\textcolor{violet}{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}}
+\textcolor{violet}{\sout{In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.}}\footnote{\textcolor{violet}{[Ben says: I think we should drop this as it's not really true about this section and tries to weirdly introduce the next section.]}\textcolor{orange}{accept}}
 While \pkg{EpiCompare} does not focus on fitting models to data, we do
@@
 This sentence seemed misplaced, and we have a very similar sentence a few lines later.
-I'm happy to be overruled here.]}}
+I'm happy to be overruled here.]}\textcolor{orange}{accept, yeah it's unnecessary}}
 Beyond SIR models, the function \texttt{simulate\_agents()} takes as
@@
 This flexibility in states can be used to also reflect group-based
-dynamics.
-\textcolor{violet}{\sout{In turn, this allows for} Both of these functions allow}
-users to explore the space of models in an intuitive way without getting
-bogged down by too much mathematical detail. For consistency, we have
-made output from \texttt{simulate\_agents()} and
-\texttt{simulate\_SIR\_agents()} compatible with
-\texttt{agents\_to\_aggregate()} so aggregate information may easily be
-accessed.
+dynamics. Both of these functions allow users to explore the space of
+models in an intuitive way without getting bogged down by too much
+mathematical detail. For consistency, we have made output from
+\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
+compatible with \texttt{agents\_to\_aggregate()} so aggregate
+information may easily be accessed.
 
@@
 to disseminate the results in an easily digestible format.
-\textcolor{violet}{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\textcolor{violet}{[Ben says: this really isn't true anymore - that's more the next section.]}}
+\textcolor{violet}{\sout{In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.}}\footnote{\textcolor{violet}{[Ben says: this really isn't true anymore - that's more the next section.]}\textcolor{orange}{accept}}
 In \pkg{EpiCompare}, we provide (1) functions to standardize simulation
-and model
-output\footnote{\textcolor{violet}{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}}
+and model \textcolor{orange}{objects}~sout\{output\}
+\footnote{\textcolor{violet}{[Ben says: I find "output" a weird word - but my only suggested replacment is "objects", which isn't much better and is used in the next paragraph...]}\textcolor{orange}{trying out objects..}}
 from external packages and (2) a function to transform standardized
@@
 
-Modeling and simulation output can be very complex objects, and as a
-result, a number of epidemic modeling \proglang{R} packages return a
-special class. The special classes often contain a plethora of
-information about residuals, model diagnostics, input parameters, and
-more. While incredibly useful, these special classes can be difficult
-for novice coders to handle. To this end, \pkg{EpiCompare} provides a
-series of fortify-style methods, called \code{fortify_aggregate()} which
-transform output from infectious disease modeling and simulation
-packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
-which contain information about the total number of individuals in each
-state at a given time, for a given simulation. These fortify functions
-have output that is consistent with that of
-\code{agents_to_aggregate()}. These standardized outputs can then be
-piped to summaries, tables, and plots.
+Modeling and simulation \textcolor{orange}{objects} \sout{output} can be
+very complex objects, and as a result, a number of epidemic modeling
+\proglang{R} packages return a special class. The special classes often
+contain a plethora of information about residuals, model diagnostics,
+input parameters, and more. While incredibly useful, these special
+classes can be difficult for novice coders to handle. To this end,
+\pkg{EpiCompare} provides a series of fortify-style methods, called
+\code{fortify_aggregate()} which transform output from infectious
+disease modeling and simulation packages like \pkg{pomp} and
+\pkg{EpiModel} into tidy-styled data frames which contain information
+about the total number of individuals in each state at a given time, for
+a given simulation. These fortify functions have output that is
+consistent with that of \code{agents_to_aggregate()}. These standardized
+outputs can then be piped to summaries, tables, and plots.
 
 Because
-\textcolor{violet}{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\textcolor{violet}{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]}}
+\textcolor{violet}{epidemic data is stored in a temporal way, \sout{the fortified output contains temporal information and}}\footnote{\textcolor{violet}{[Ben says: this isn't my best suggestion, but I think we should avoid suggesting filament\_compression is only for the fortified data - it can be used on processed raw data as well.]} \textcolor{orange}{accept}}
 we provide the function, \code{filament_compression()}, to transform
@@
 
+\textcolor{orange}{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.
+We first talk about the step in the pipeline and then how epicompare fits into that pipeline.
+The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.
+I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}
+
 Finally, the data analysis pipeline often ends with plots, tables, and
-summary statistics that are used to compare different models, and
-simulations to one
+summary statistics that are used to compare
+\textcolor{orange}{and assess} different models, and simulations to one
 another\footnote{\textcolor{violet}{[Ben says: I'm confused why this is just about comparison and not assessment.
@@
 We have found these tools to be specifically applicable for situations
-where only one ``season'' or ``cycle'' of an epidemic has occurred.
+where only one season or cycle of an epidemic has occurred.
 

### 2021-05-20 jss/EpiCompare-paper/EpiCompare-paper.Rmd
section 3 done except for final subsection

--- 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
 weekly case counts) is observed.
-Compartment totals (amounts of individuals in each state) are then imputed from those case counts other information about the disease and the population of interest.
+In this setting, compartment totals (amounts of individuals in each state) are then imputed from those case counts and using other information about the disease and the population of interest.
 In \pkg{EpiCompare}, we focus on understanding the overall impact of an outbreak at the aggregate/population level, which allows for streamlined examination of overall trends of an epidemic.
@@
 start of infection, start of recovery, etc.) to an aggregate view of how many individuals were in a state at a given time.
-There are often situations where grouping agents into subpopulations (e.g.
-subpopulations defined by age or sex) can highlight different aggregate level trends.
-For example, research by \citet{rvachev1985,anderson1992,worby2015} develop state-based models that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation analysis\footnote{\shannon{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}} by extending `agents_to_aggregate()` to work with \pkg{dplyr}::`group_by()`, the tidyverse approach to the split-apply-combine paradigm for data analysis \citep{wickham2011}.
-
-[[\textcolor{red}{OLD SENTENCE}
-by combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()` to provide aggregation at a subpopulation level\footnote{\ben{[Ben says: This sentence seems a bit repetitive with "supopulation".
-]}\shannon{agreed.
-aggregation at a group level}}.
-]]
+Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often are interesting in more granular trends that can be detected by aggregation, conditional on subpopulations (e.g.
+subpopulations defined by age or sex).
+By combining the function \pkg{dplyr}::`group_by()` and `agents_to_aggregate()`, \pkg{EpiCompare} provides group level aggregation.
 
@@
 The function `geom_aggregate()` provides a visualization tool to holistically examine aggregate level information across different subpopulations by visualizing each subpopulation's epidemic trajectory in the three-dimensional state space.
-Visualization tools for three-state models were developed because (1) SIR models are some of the most common and basic epidemic state-based models and (2) our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).
-
-**Model Fitting and Simulations**
+Visualization tools for three-state models were developed because SIR models are some of the most common and basic epidemic state-based models and our three-dimensional simplex representation of these epidemics emphasizes a time-invariant representation of the data (for a refresher see Section \ref{sec:time-invariant}).
+
+**Model fitting and simulations**
 
 After getting a sense of what a past or current epidemic looks like with EDA, the next step in the data analysis pipeline is often model fitting and/or simulation.
-In this step and the next step (post-processing), we highlight how practitioners can pair models and simulations of epidemics from outside of \pkg{EpiCompare} with analysis and simulation tools in \pkg{EpiCompare}.
 While \pkg{EpiCompare} does not focus on fitting models to data, we do provide some flexible functions for simulation of basic discrete-time epidemic-state models.
@@
 The function `simulate_SIR_agents()` simulates a basic SIR epidemic with user inputs for the number of simulations, the initial number in each state, the infection and recovery parameters $(\beta, \gamma$), and the total number of discrete time steps.
-This function allows for easy access to SIR model analysis and comparison.
 Beyond SIR models, the function `simulate_agents()` takes as input a user-specified state-transition matrix and other epidemic parameters to allow the user to create simulations for an outbreak with \textit{any} number of states and any number of transitions among them.
 This flexibility in states can be used to also reflect group-based dynamics.
-In turn, this allows for users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
+Both of these functions allow users to explore the space of models in an intuitive way without getting bogged down by too much mathematical detail.
 For consistency, we have made output from `simulate_agents()` and `simulate_SIR_agents()` compatible with `agents_to_aggregate()` so aggregate information may easily be accessed.
@@
 
-\shannon{[[NEW TEXT}
 If practitioners wish to compare models-to-observations or even models-to-models, they need to post-process their models and simulations to disseminate the results in an easily digestible format.
-In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.
-In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.\footnote{\shannon{new signpost.
-There are two paragraphs below corresponding to the two points.
-The second paragraph could use another sentence maybe.
-I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}
+In \pkg{EpiCompare}, we provide (1) functions to standardize simulation and model output from external packages and (2) a function to transform standardized simulation and model output into a format amenable to time-invariant analysis.
 
@@
 
-Because the fortified output contains temporal information and we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations.
+Because epidemic data is stored in a temporal way, we provide the function, \code{filament_compression()}, to transform temporally defined epidemics to their filamental representations.
 These filaments can then be fairly compared to one another or passed to further time-invariant analysis tools described below.
-\shannon{]]}
-
-[[\textcolor{red}{OLD TEXT BELOW}
-
-\footnote{\ben{[Ben says: I found the structure of this section slightly confusing.
-It appears that the 2 subsections are (1) model based transformation and (2) simulation based transformations.
-The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}\shannon{agreed that this section needs some work.
-but to me those aren't the clear points, it seems like (1) put external model results in the same format and (2) transform model output to filaments.
-that said i don't see how we're over-selling the package.
-would 'transform and disseminate' be better?}}
-If practitioners wish to compare models-to-observations or even models-to-models, they needs to post-process their models and simulations to disseminate the results in an easily digestible format.
-In general, post-processing of modeling and simulation consists of making summary statistics, plots, and tables.
-
-Model output summaries can be very complex, and as a result, a number of epidemic modeling \proglang{R} packages return a special class.
-The special classes often contain a plethora of information about residuals, model diagnostics, input parameters, and more.
-While incredibly useful, these special classes can be difficult for novice coders to handle.
-\shannon{motivates for fortify but not filamental compression}
-
-To this end, \pkg{EpiCompare} provides a series of `fortify`-style methods, called `fortify_aggregate()` which transform output from infectious disease modeling and simulation packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which contain information about the total number of individuals in each state at a given time, for a given simulation.
-These fortify functions have output that is consistent with that of `agents_to_aggregate()`.
-
-To utilize simulations of epidemics in  time-invariant analysis, we also provide a function to convert temporally defined epidemics to filamental representations.
-Specifically, we provide the function `filament_compression()` to convert simulation(s) to filamental trajectories as expressed by presenting the epidemic as an ordering of some common fixed number of points so that they are equally spaced along the original path in the proportional state space\footnote{\ben{[Ben says: there's a lot going on in this sentence, and the phrase "as expressed by presenting the epidemic as an ordering of some common fixed number of points" is between the subject and our verb "so that".]}}.
-These tools were developed to provide another natural entry point into the \pkg{EpiCompare} data analysis pipeline for situations where modeling and analysis are already completed and practitioners are looking for simple and transparent tools to help understand and disseminate results.\footnote{\ben{[Ben says: This sentence still doesn't seem to fit into the paragraph or section too well.
-First, given it's in this paragraph: is it just the simulation based "tools"?
-I might suggest moving it ot the first introductory sentence - but probably needs some word-smithing as well - I feel like it might be overselling the package a bit.]}}
-
-]]
-
-**Comparisons and Assessment**
+
+**Comparisons and assessment**
 \shannon{[[ NEW TEXT}
 
-Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare different models, and simulations to one another.
+\shannon{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.
+We first talk about the step in the pipeline and then how epicompare fits into that pipeline.
+The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.
+I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}
+
+Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare \shannon{and assess} different models, and simulations to one another\footnote{\ben{[Ben says: I'm confused why this is just about comparison and not assessment.
+Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
 In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
-mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one "season" or "cycle" of an epidemic has occurred.
+mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one season or cycle of an epidemic has occurred.
 

### 2021-05-20 jss/EpiCompare-paper/EpiCompare-paper.tex
section 3 done except for final subsection

--- 2021-05-17 jss/EpiCompare-paper/EpiCompare-paper.tex
@@
 of new infections at each time step (e.g.~weekly case counts) is
-observed. Compartment totals (amounts of individuals in each state) are
-then imputed from those case counts other information about the disease
-and the population of interest. In \pkg{EpiCompare}, we focus on
-understanding the overall impact of an outbreak at the
-aggregate/population level, which allows for streamlined examination of
-overall trends of an epidemic.
+observed. In this setting, compartment totals (amounts of individuals in
+each state) are then imputed from those case counts and using other
+information about the disease and the population of interest. In
+\pkg{EpiCompare}, we focus on understanding the overall impact of an
+outbreak at the aggregate/population level, which allows for streamlined
+examination of overall trends of an epidemic.
 
@@
 aggregate view of how many individuals were in a state at a given time.
-There are often situations where grouping agents into subpopulations
-(e.g.~subpopulations defined by age or sex) can highlight different
-aggregate level trends. For example, research by
-\citet{rvachev1985,anderson1992,worby2015} develop state-based models
-that account for differing disease dynamics in different subpopulations.
-In \pkg{EpiCompare}, we facilitate subpopulation
-analysis\footnote{\textcolor{orange}{Changed to try to clarify how how agents to aggregate works with group by and why: (split-apply-combine)}}
-by extending \texttt{agents\_to\_aggregate()} to work with
-\pkg{dplyr}::\texttt{group\_by()}, the tidyverse approach to the
-split-apply-combine paradigm for data analysis \citep{wickham2011}.
-
-{[}{[}\textcolor{red}{OLD SENTENCE} by combining the function
-\pkg{dplyr}::\texttt{group\_by()} and \texttt{agents\_to\_aggregate()}
-to provide aggregation at a subpopulation
-level\footnote{\textcolor{violet}{[Ben says: This sentence seems a bit repetitive with "supopulation". ]}\textcolor{orange}{agreed. aggregation at a group level}}.{]}{]}
+Researchers, including \citet{rvachev1985,anderson1992,worby2015}, often
+are interesting in more granular trends that can be detected by
+aggregation, conditional on subpopulations (e.g.~subpopulations defined
+by age or sex). By combining the function
+\pkg{dplyr}::\texttt{group\_by()} and \texttt{agents\_to\_aggregate()},
+\pkg{EpiCompare} provides group level aggregation.
 
@@
 three-dimensional state space. Visualization tools for three-state
-models were developed because (1) SIR models are some of the most common
-and basic epidemic state-based models and (2) our three-dimensional
-simplex representation of these epidemics emphasizes a time-invariant
+models were developed because SIR models are some of the most common and
+basic epidemic state-based models and our three-dimensional simplex
+representation of these epidemics emphasizes a time-invariant
 representation of the data (for a refresher see Section
@@
 
-\textbf{Model Fitting and Simulations}
+\textbf{Model fitting and simulations}
 
@@
 EDA, the next step in the data analysis pipeline is often model fitting
-and/or simulation. In this step and the next step (post-processing), we
-highlight how practitioners can pair models and simulations of epidemics
-from outside of \pkg{EpiCompare} with analysis and simulation tools in
-\pkg{EpiCompare}. While \pkg{EpiCompare} does not focus on fitting
+and/or simulation. While \pkg{EpiCompare} does not focus on fitting
 models to data, we do provide some flexible functions for simulation of
@@
 initial number in each state, the infection and recovery parameters
-\((\beta, \gamma\)), and the total number of discrete time steps. This
-function allows for easy access to SIR model analysis and comparison.
-Beyond SIR models, the function \texttt{simulate\_agents()} takes as
-input a user-specified state-transition matrix and other epidemic
-parameters to allow the user to create simulations for an outbreak with
-\textit{any} number of states and any number of transitions among them.
-This flexibility in states can be used to also reflect group-based
-dynamics. In turn, this allows for users to explore the space of models
-in an intuitive way without getting bogged down by too much mathematical
+\((\beta, \gamma\)), and the total number of discrete time steps. Beyond
+SIR models, the function \texttt{simulate\_agents()} takes as input a
+user-specified state-transition matrix and other epidemic parameters to
+allow the user to create simulations for an outbreak with \textit{any}
+number of states and any number of transitions among them. This
+flexibility in states can be used to also reflect group-based dynamics.
+Both of these functions allow users to explore the space of models in an
+intuitive way without getting bogged down by too much mathematical
 detail. For consistency, we have made output from
@@
 
-\textcolor{orange}{[[NEW TEXT}
-
 If practitioners wish to compare models-to-observations or even
 models-to-models, they need to post-process their models and simulations
-to disseminate the results in an easily digestible format. In general,
-post-processing of modeling and simulation consists of making summary
-statistics, plots, and tables. In \pkg{EpiCompare}, we provide (1)
-functions to standardize simulation and model output from external
-packages and (2) a function to transform standardized simulation and
-model output into a format amenable to time-invariant
-analysis.\footnote{\textcolor{orange}{new signpost.
-There are two paragraphs below corresponding to the two points.
-The second paragraph could use another sentence maybe.
-I tried to emphasize the process of output -> standard format -> transformation to time invariant analysis.}}
+to disseminate the results in an easily digestible format. In
+\pkg{EpiCompare}, we provide (1) functions to standardize simulation and
+model output from external packages and (2) a function to transform
+standardized simulation and model output into a format amenable to
+time-invariant analysis.
 
@@
 
-Because the fortified output contains temporal information and we
-provide the function, \code{filament_compression()}, to transform
-temporally defined epidemics to their filamental representations. These
-filaments can then be fairly compared to one another or passed to
-further time-invariant analysis tools described below.
-\textcolor{orange}{]]}
-
-{[}{[}\textcolor{red}{OLD TEXT BELOW}
-
-\footnote{\textcolor{violet}{[Ben says: I found the structure of this section slightly confusing.
-It appears that the 2 subsections are (1) model based transformation and (2) simulation based transformations.
-The fact that there are 2 points isn't quickly apparent when reading it, and I found it odd that the 3rd sentence of the first paragraph introduces model processing (which the second paragraph also discusses).]}\textcolor{orange}{agreed that this section needs some work.
-but to me those aren't the clear points, it seems like (1) put external model results in the same format and (2) transform model output to filaments.
-that said i don't see how we're over-selling the package.
-would 'transform and disseminate' be better?}}
-If practitioners wish to compare models-to-observations or even
-models-to-models, they needs to post-process their models and
-simulations to disseminate the results in an easily digestible format.
-In general, post-processing of modeling and simulation consists of
-making summary statistics, plots, and tables.
-
-Model output summaries can be very complex, and as a result, a number of
-epidemic modeling \proglang{R} packages return a special class. The
-special classes often contain a plethora of information about residuals,
-model diagnostics, input parameters, and more. While incredibly useful,
-these special classes can be difficult for novice coders to handle.
-\textcolor{orange}{motivates for fortify but not filamental compression}
-
-To this end, \pkg{EpiCompare} provides a series of
-\texttt{fortify}-style methods, called \texttt{fortify\_aggregate()}
-which transform output from infectious disease modeling and simulation
-packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
-which contain information about the total number of individuals in each
-state at a given time, for a given simulation. These fortify functions
-have output that is consistent with that of
-\texttt{agents\_to\_aggregate()}.
-
-To utilize simulations of epidemics in time-invariant analysis, we also
-provide a function to convert temporally defined epidemics to filamental
-representations. Specifically, we provide the function
-\texttt{filament\_compression()} to convert simulation(s) to filamental
-trajectories as expressed by presenting the epidemic as an ordering of
-some common fixed number of points so that they are equally spaced along
-the original path in the proportional state
-space\footnote{\textcolor{violet}{[Ben says: there's a lot going on in this sentence, and the phrase "as expressed by presenting the epidemic as an ordering of some common fixed number of points" is between the subject and our verb "so that".]}}.
-These tools were developed to provide another natural entry point into
-the \pkg{EpiCompare} data analysis pipeline for situations where
-modeling and analysis are already completed and practitioners are
-looking for simple and transparent tools to help understand and
-disseminate
-results.\footnote{\textcolor{violet}{[Ben says: This sentence still doesn't seem to fit into the paragraph or section too well.
-First, given it's in this paragraph: is it just the simulation based "tools"?
-I might suggest moving it ot the first introductory sentence - but probably needs some word-smithing as well - I feel like it might be overselling the package a bit.]}}
-
-{]}{]}
-
-\textbf{Comparisons and Assessment} \textcolor{orange}{[[ NEW TEXT}
+Because epidemic data is stored in a temporal way, we provide the
+function, \code{filament_compression()}, to transform temporally defined
+epidemics to their filamental representations. These filaments can then
+be fairly compared to one another or passed to further time-invariant
+analysis tools described below.
+
+\textbf{Comparisons and assessment} \textcolor{orange}{[[ NEW TEXT}
+
+\textcolor{orange}{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.
+We first talk about the step in the pipeline and then how epicompare fits into that pipeline.
+The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.
+I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}
 
 Finally, the data analysis pipeline often ends with plots, tables, and
-summary statistics that are used to compare different models, and
-simulations to one another. In \pkg{EpiCompare} we provide a set of
-comparison and assessment tools for model and simulation results that
-extend beyond the standard performance metrics (e.g.~mean squared error
-or AIC) and into the lens of time-invariant
+summary statistics that are used to compare
+\textcolor{orange}{and assess} different models, and simulations to one
+another\footnote{\textcolor{violet}{[Ben says: I'm confused why this is just about comparison and not assessment.
+Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
+In \pkg{EpiCompare} we provide a set of comparison and assessment tools
+for model and simulation results that extend beyond the standard
+performance metrics (e.g.~mean squared error or AIC) and into the lens
+of time-invariant
 analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}
 We have found these tools to be specifically applicable for situations
-where only one ``season'' or ``cycle'' of an epidemic has occurred.
+where only one season or cycle of an epidemic has occurred.
 

### 2021-05-21 jss/EpiCompare-paper/EpiCompare-paper2.Rmd
minor updates to section 1 and 2 - basically clean up after chat and Shannon's edits

@@
 historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}.
-At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process, as has been seen in the current COVID-19 \shannon{\sout{epi}}\shannon{pan}demic \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
-\footnote{\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.
-*This is less so the case now.]}\shannon{Seems fine to me now}}
+At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process, as has been seen in the current COVID-19 pandemic \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
 
-With \shannon{\sout{the}} many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} \shannon{available}, this field has a \shannon{\sout{lot of}} \shannon{number of} devices to \shannon{\sout{aid}} \shannon{help}  an individual practitioner decide the correct approach.
+With many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} available, this field has a number of devices to \shannon{\sout{aid}} \shannon{help}\footnote{\ben{[Ben says: I'm not sure this an improvement]}}  an individual practitioner decide the correct approach.
 For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
@@
 One of \pkg{EpiCompare}’s main contribution to comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments.
-Epidemics, despite being defined as a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales \shannon{in order}\footnote{\shannon{wanted a clearer signal that this is a bridge to a new thought}}to understand the processes at play.
+Epidemics, despite being defined as a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales in order to understand the processes at play.
 With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible.
@@
 These time-dependent approaches can reduce the amount of connections that can be seen and insights that can be drawn}, similar to examining projections of a multidimensional distribution onto a single axis, one at a time.
-Tools in \pkg{EpiCompare} \shannon{\sout{give the user the ability to extend their} \shannon{extend the user}} toolkit to evaluate epidemics within a time-invariant lens.
+Tools in \pkg{EpiCompare} extend the user toolkit to evaluate epidemics within a time-invariant lens.
 At the same time, the goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
@@
 This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a  demonstration of the tools through a detailed example\shannon{\sout{ of a full data analysis pipeline.}} \shannon{from start to finish of the data analysis pipeline}\footnote{\shannon{maybe clearer?
-up to you}}
+up to you} \ben{The current suggestion isn't clearer - feel free to try again.}}
 
@@
 In this section we highlight some weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take.
-To accomplish this goal we  first demonstrate the inability of time-dependent tools to \ben{adequately} quantify\footnote{\ben{[Ben says: original suggestion was "simply capture", and original phrase was "well quantify".]}} a classic epidemic parameter, the reproduction number $R_0$.
+To accomplish this goal we  first demonstrate the inability of time-dependent tools to adequately quantify a classic epidemic parameter, the reproduction number $R_0$. 
 Then we motivate new time-invariant approaches for more complex situations where $R_0$ does not capture the complexities of outbreaks.
@@
 One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.
-This \ben{numerical summary} \sout{intrinsic property} is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}.
+This numerical summary is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}.
 \citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic.
@@
 
-Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space \sout{(i.e.
-through a time-invariant lens)}\footnote{\ben{[Ben says: the next sentence mentions time-invariance - where you thinking something more global?]}} allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted.
+Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted.
 In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states.
@@
 \]
-where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$ (where $|\psi_i|$ is the length of $\psi_i$).
+where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$ (with $|\psi_i|$ is the length of $\psi_i$).
 This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.

### 2021-05-21 jss/EpiCompare-paper/EpiCompare-paper2.tex
minor updates to section 1 and 2 - basically clean up after chat and Shannon's edits

@@
 government, and news organizations are an important first step of the
-process, as has been seen in the current COVID-19
-\textcolor{orange}{\sout{epi}}\textcolor{orange}{pan}demic
+process, as has been seen in the current COVID-19 pandemic
 \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
-\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}\textcolor{orange}{Seems fine to me now}}
-
-With \textcolor{orange}{\sout{the}} many visualization and exploratory
-tools, models and modeling paradigms, and reviews and comparisons in the
-literature and through the MIDAS (Models of Infectious Disease Agent
-Study) network \citep{midasnetwork2021} \textcolor{orange}{available},
-this field has a \textcolor{orange}{\sout{lot of}}
-\textcolor{orange}{number of} devices to \textcolor{orange}{\sout{aid}}
-\textcolor{orange}{help} an individual practitioner decide the correct
-approach. For example, \proglang{R} packages such as \pkg{surveillance},
-\pkg{EpiModel}, and \pkg{pomp} have all made significant steps in
-standardizing the flow of the data analysis pipeline for epidemic
-modeling through digitizing data sets, making accessible statistical
-models, and providing a plethora of educational material for both coding
-novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
+
+With many visualization and exploratory tools, models and modeling
+paradigms, and reviews and comparisons in the literature and through the
+MIDAS (Models of Infectious Disease Agent Study) network
+\citep{midasnetwork2021} available, this field has a number of devices
+to \textcolor{orange}{\sout{aid}}
+\textcolor{orange}{help}\footnote{\textcolor{violet}{[Ben says: I'm not sure this an improvement]}}
+an individual practitioner decide the correct approach. For example,
+\proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
+\pkg{pomp} have all made significant steps in standardizing the flow of
+the data analysis pipeline for epidemic modeling through digitizing data
+sets, making accessible statistical models, and providing a plethora of
+educational material for both coding novices and experts alike
+\citep{surveillance2017,Jenness2018,King2016}.
 
@@
 over time, often need to be compared in a way not constrained to initial
-times or time scales
-\textcolor{orange}{in order}\footnote{\textcolor{orange}{wanted a clearer signal that this is a bridge to a new thought}}to
-understand the processes at play. With time-invariant analysis,
-comparing decades-long outbreaks of HIV in the US to a 10 day outbreak
-of norovirus on a cruise ship is possible. Compared to time-dependent
-comparison tools for state-space modeling, time-invariant analysis can
-make it easier to compare state-space epidemic representations in a more
-global, holistic fashion. Many time-dependent comparison tools only
-examine the proportion of individuals in each state (at a given time) in
-a piece-wise / marginal fashion. These time-dependent approaches can
-reduce the amount of connections that can be seen and insights that can
-be drawn\}, similar to examining projections of a multidimensional
-distribution onto a single axis, one at a time. Tools in
-\pkg{EpiCompare}
-\textcolor{orange}{\sout{give the user the ability to extend their} \textcolor{orange}{extend the user}}
+times or time scales in order to understand the processes at play. With
+time-invariant analysis, comparing decades-long outbreaks of HIV in the
+US to a 10 day outbreak of norovirus on a cruise ship is possible.
+Compared to time-dependent comparison tools for state-space modeling,
+time-invariant analysis can make it easier to compare state-space
+epidemic representations in a more global, holistic fashion. Many
+time-dependent comparison tools only examine the proportion of
+individuals in each state (at a given time) in a piece-wise / marginal
+fashion. These time-dependent approaches can reduce the amount of
+connections that can be seen and insights that can be drawn\}, similar
+to examining projections of a multidimensional distribution onto a
+single axis, one at a time. Tools in \pkg{EpiCompare} extend the user
 toolkit to evaluate epidemics within a time-invariant lens. At the same
@@
 example\textcolor{orange}{\sout{ of a full data analysis pipeline.}}
-\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer? up to you}}
+\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer?
+up to you} \textcolor{violet}{The current suggestion isn't clearer - feel free to try again.}}
 
@@
 goal we first demonstrate the inability of time-dependent tools to
-\textcolor{violet}{adequately}
-quantify\footnote{\textcolor{violet}{[Ben says: original suggestion was "simply capture", and original phrase was "well quantify".]}}
-a classic epidemic parameter, the reproduction number \(R_0\). Then we
-motivate new time-invariant approaches for more complex situations where
-\(R_0\) does not capture the complexities of outbreaks. This leads to
-the final subsection that defines ways to view epidemics in a
-time-invariant lens and discusses natural properties of these
-representations that allow for clearer ways to compare models and
+adequately quantify a classic epidemic parameter, the reproduction
+number \(R_0\). Then we motivate new time-invariant approaches for more
+complex situations where \(R_0\) does not capture the complexities of
+outbreaks. This leads to the final subsection that defines ways to view
+epidemics in a time-invariant lens and discusses natural properties of
+these representations that allow for clearer ways to compare models and
 epidemics.
@@
 summaries of an epidemic is the reproduction number - \(R_0\). This
-\textcolor{violet}{numerical summary} \sout{intrinsic property} is a
-time-invariant value that is defined as the expected number of
-infections caused by a single infect or who is added to a completely
-susceptible population \citep{anderson1992}. \citet{Gallagher2020}
-showed that the estimation of \(R_0\) can be sensitive to time-dependent
-parameters like estimation of the beginning and end of an epidemic. This
-time-dependent sensitivity in estimating \(R_0\) reflects a general
-problem in epidemiology surrounding how to transform time-dependent
-information into insights into desirable, intrinsic properties of an
-epidemic.
+numerical summary is a time-invariant value that is defined as the
+expected number of infections caused by a single infect or who is added
+to a completely susceptible population \citep{anderson1992}.
+\citet{Gallagher2020} showed that the estimation of \(R_0\) can be
+sensitive to time-dependent parameters like estimation of the beginning
+and end of an epidemic. This time-dependent sensitivity in estimating
+\(R_0\) reflects a general problem in epidemiology surrounding how to
+transform time-dependent information into insights into desirable,
+intrinsic properties of an epidemic.
 
@@
 Through the \(R_0\) example, we see that treating epidemics like
-filamental trajectories embedded in a lower dimensional space
-\sout{(i.e. through a time-invariant lens)}\footnote{\textcolor{violet}{[Ben says: the next sentence mentions time-invariance - where you thinking something more global?]}}
-allows us to better compare the overall structure of the epidemic and
-see how the population is directly impacted. In this section we present
+filamental trajectories embedded in a lower dimensional space allows us
+to better compare the overall structure of the epidemic and see how the
+population is directly impacted. In this section we present
 time-invariant tools that can be applied to complex epidemics where the
@@
 \] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
-\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (where
+\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (with
 \(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
 time-invariant, and can be plugged into multiple distance-based
-assessment tools to examine the overall ``extremeness'\,' of points,
+assessment tools to examine the overall ``extremeness'' of points,
 including pseudo-density estimators and depth/local depth functions

### 2021-05-22 jss/EpiCompare-paper/EpiCompare-paper.Rmd
step 1 of edits to section 3

@@
 
-\shannon{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.
-We first talk about the step in the pipeline and then how epicompare fits into that pipeline.
-The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.
-I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}
-
-Finally, the data analysis pipeline often ends with plots, tables, and summary statistics that are used to compare \shannon{and assess} different models, and simulations to one another\footnote{\ben{[Ben says: I'm confused why this is just about comparison and not assessment.
-Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
-In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
-mean squared error or AIC) and into the lens of time-invariant analysis.\footnote{\shannon{a bit of an awk transition but I wanted to use past wording about TIA.}}  We have found these tools to be specifically applicable for situations where only one season or cycle of an epidemic has occurred.
+\ben{\sout{Finally,} Often}\footnote{\ben{[Ben says: I feel like starting with "finally" causes a weird pause - but maybe that's just me.]}} the data analysis pipeline \ben{\sout{often}} ends with plots, tables, and summary statistics \ben{\sout{that} which} are used to \sout{\ben{compare} \shannon{and assess}} \ben{\sout{different models, and simulations to one another} assess model preformance and compare across models or simulations.} In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
+mean squared error or AIC) and into the lens of time-invariant analysis.
+We have found \ben{that} these tools \ben{\sout{to be} are} specifically applicable for situations where only one season or cycle of an epidemic has occurred, \ben{or is the object of interest}.
+
+\ben{The first set of of tools surround the creation of prediction regions.
+We can create a prediction regions from model simulations to examine if our model simulations capture the true epidemic trajectory.
+We do so in a time-invariant way and utlize a filamental representation of the model simulations and the true epidemic.} [For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
+In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done by eye.
+In \pkg{EpiCompare}, we also provide these prediction regions for epidemic models with with more than three states.
+The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model.
+For both of these geometric structures, we provide functions to check if a path is contained (`contained()`)].
+\ben{We can also use these prediction regions to visually or mathematically compare how similar two sets of simulations are.
+In \pkg{EpiCompare} we provide the} `hausdorff_dist()` \ben{function to calculate the Hausdorff distance between multiple prediction regions.}
+
+[We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.] \ben{We provide implementations of a few score functions that capture extremeness of epidemics relative to other epidemics, and these scores can be turned into an extremeness measure with} `mean(sim_scores > truth_score)`.
+[Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
+Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.]
+
+]]
+
+[[\textcolor{red}{ OLD TEXT}
 
@@
 
+We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
+Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
+Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
+
 \shannon{]]}
 
-[[\textcolor{red}{OLD TEXT}
+[[\textcolor{red}{Really OLD TEXT}
 

### 2021-05-22 jss/EpiCompare-paper/EpiCompare-paper.tex
step 1 of edits to section 3

@@
 
-\textcolor{orange}{paragraph re-write thoughts: I changed this paragraph to align with the other sub-sections.
-We first talk about the step in the pipeline and then how epicompare fits into that pipeline.
-The old paragraph skipped the general plots, tables, and 'standard' tools and went straight into filaments.
-I re-wrote the paragraph to more clearly indicate that we don't think filaments are the only tools for comparison and assessment.}
-
-Finally, the data analysis pipeline often ends with plots, tables, and
-summary statistics that are used to compare
-\textcolor{orange}{and assess} different models, and simulations to one
-another\footnote{\textcolor{violet}{[Ben says: I'm confused why this is just about comparison and not assessment.
-Also, I didn't really want to make tons of comments on this section, so could you add a footnote on why you rewrote it and what you were going for?]}}.
+\textcolor{violet}{\sout{Finally,} Often}\footnote{\textcolor{violet}{[Ben says: I feel like starting with "finally" causes a weird pause - but maybe that's just me.]}}
+the data analysis pipeline \textcolor{violet}{\sout{often}} ends with
+plots, tables, and summary statistics
+\textcolor{violet}{\sout{that} which} are used to
+\sout{\textcolor{violet}{compare} \textcolor{orange}{and assess}}
+\textcolor{violet}{\sout{different models, and simulations to one another} assess model preformance and compare across models or simulations.}
 In \pkg{EpiCompare} we provide a set of comparison and assessment tools
@@
 performance metrics (e.g.~mean squared error or AIC) and into the lens
-of time-invariant
-analysis.\footnote{\textcolor{orange}{a bit of an awk transition but I wanted to use past wording about TIA.}}
-We have found these tools to be specifically applicable for situations
-where only one season or cycle of an epidemic has occurred.
+of time-invariant analysis. We have found \textcolor{violet}{that} these
+tools \textcolor{violet}{\sout{to be} are} specifically applicable for
+situations where only one season or cycle of an epidemic has occurred,
+\textcolor{violet}{or is the object of interest}.
+
+\textcolor{violet}{The first set of of tools surround the creation of prediction regions.
+We can create a prediction regions from model simulations to examine if our model simulations capture the true epidemic trajectory.
+We do so in a time-invariant way and utlize a filamental representation of the model simulations and the true epidemic.}
+{[}For three-state epidemic models, we provide the
+\texttt{ggplot}/\texttt{ggtern} extension
+\texttt{geom\_prediction\_band()} which creates a prediction region
+around the top \(1-\alpha\) proportion of the simulations. In this
+visual setting, comparing this prediction to the true epidemic
+trajectory or comparing the prediction regions defined by two different
+models' simulations can be done by eye. In \pkg{EpiCompare}, we also
+provide these prediction regions for epidemic models with with more than
+three states. The functions \texttt{create\_convex\_hull\_structure()}
+and \texttt{create\_delta\_ball\_structure()} create different geometric
+representations of prediction regions for any dimensional state-based
+model. For both of these geometric structures, we provide functions to
+check if a path is contained (\texttt{contained()}){]}.
+\textcolor{violet}{We can also use these prediction regions to visually or mathematically compare how similar two sets of simulations are. In \pkg{EpiCompare} we provide the}
+\texttt{hausdorff\_dist()}
+\textcolor{violet}{function to calculate the Hausdorff distance between multiple prediction regions.}
+
+{[}We also provide functions to calculate the ``extremeness'' of a true
+epidemic trajectory compared to simulated epidemics via the
+equi-distance filamental trajectory representation as mentioned in
+Section \ref{sec:beyond-r0-sir}.{]}
+\textcolor{violet}{We provide implementations of a few score functions that capture extremeness of epidemics relative to other epidemics, and these scores can be turned into an extremeness measure with}
+\texttt{mean(sim\_scores\ \textgreater{}\ truth\_score)}.
+{[}Specifically, functions like
+\texttt{distance\_pseudo\_density\_function()} can calculate a
+pseudo-density estimate of the true epidemic relative to simulated ones.
+Functions \texttt{distance\_depth\_function()} and
+\texttt{local\_distance\_depth\_function()} provide depth scores that
+suggest how geometrically central an epidemic is to simulations.{]}
+
+{]}{]}
+
+{[}{[}\textcolor{red}{ OLD TEXT}
 
@@
 
+We also provide functions to calculate the ``extremeness'' of a true
+epidemic trajectory compared to simulated epidemics via the
+equi-distance filamental trajectory representation as mentioned in
+Section \ref{sec:beyond-r0-sir}. Specifically, functions like
+\texttt{distance\_pseudo\_density\_function()} can calculate a
+pseudo-density estimate of the true epidemic relative to simulated ones.
+Functions \texttt{distance\_depth\_function()} and
+\texttt{local\_distance\_depth\_function()} provide depth scores that
+suggest how geometrically central an epidemic is to simulations.
+
 \textcolor{orange}{]]}
 
-{[}{[}\textcolor{red}{OLD TEXT}
+{[}{[}\textcolor{red}{Really OLD TEXT}
 

### 2021-05-22 jss/EpiCompare-paper/EpiCompare-paper.Rmd
second round of minor edits to 3's final subpart

@@
 
-\ben{\sout{Finally,} Often}\footnote{\ben{[Ben says: I feel like starting with "finally" causes a weird pause - but maybe that's just me.]}} the data analysis pipeline \ben{\sout{often}} ends with plots, tables, and summary statistics \ben{\sout{that} which} are used to \sout{\ben{compare} \shannon{and assess}} \ben{\sout{different models, and simulations to one another} assess model preformance and compare across models or simulations.} In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
+\ben{\sout{Finally,} Often}\footnote{\ben{[Ben says: I feel like starting with "finally" causes a weird pause - but maybe that's just me.]}} the data analysis pipeline \ben{\sout{often}} ends with plots, tables, and summary statistics that are used to \sout{\ben{compare} \shannon{and assess}} \ben{\sout{different models, and simulations to one another} assess model preformance and compare across models or simulations.} In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
 mean squared error or AIC) and into the lens of time-invariant analysis.
@@
 
-\ben{The first set of of tools surround the creation of prediction regions.
+\footnote{\ben{[Ben says: This changed introduction was done to highlight that we can use prediction regions for mutiple types of comparisons.]}}\ben{The first set of of tools surround the creation of prediction regions.
 We can create a prediction regions from model simulations to examine if our model simulations capture the true epidemic trajectory.
-We do so in a time-invariant way and utlize a filamental representation of the model simulations and the true epidemic.} [For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
-In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done by eye.
+We do so in a time-invariant way and utilizing filamental representations of the model simulations and the true epidemic.} [For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
+In this visual setting, comparing this prediction \ben{region} to the true epidemic trajectory \ben{\sout{or comparing the prediction regions defined by two different models' simulations}}\footnote{\ben{[Ben says: this part will be highlighted in later sentences.]}} can be done by eye.
 In \pkg{EpiCompare}, we also provide these prediction regions for epidemic models with with more than three states.
@@
 \ben{We can also use these prediction regions to visually or mathematically compare how similar two sets of simulations are.
-In \pkg{EpiCompare} we provide the} `hausdorff_dist()` \ben{function to calculate the Hausdorff distance between multiple prediction regions.}
-
-[We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.] \ben{We provide implementations of a few score functions that capture extremeness of epidemics relative to other epidemics, and these scores can be turned into an extremeness measure with} `mean(sim_scores > truth_score)`.
+In \pkg{EpiCompare} we provide the} `hausdorff_dist()` \ben{function to calculate the Hausdorff distance between multiple prediction regions, when visual comparison is not possible.}
+
+[We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.] \ben{We provide implementations of a few distance-based score functions that capture how "reasonable" an epidemic is relative to other epidemics, and these scores can be turned into an extremeness measure with} `mean(sim_scores > truth_score)`.
 [Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.

### 2021-05-22 jss/EpiCompare-paper/EpiCompare-paper.tex
second round of minor edits to 3's final subpart

@@
 the data analysis pipeline \textcolor{violet}{\sout{often}} ends with
-plots, tables, and summary statistics
-\textcolor{violet}{\sout{that} which} are used to
+plots, tables, and summary statistics that are used to
 \sout{\textcolor{violet}{compare} \textcolor{orange}{and assess}}
@@
 
-\textcolor{violet}{The first set of of tools surround the creation of prediction regions.
+\footnote{\textcolor{violet}{[Ben says: This changed introduction was done to highlight that we can use prediction regions for mutiple types of comparisons.]}}\textcolor{violet}{The first set of of tools surround the creation of prediction regions.
 We can create a prediction regions from model simulations to examine if our model simulations capture the true epidemic trajectory.
-We do so in a time-invariant way and utlize a filamental representation of the model simulations and the true epidemic.}
+We do so in a time-invariant way and utilizing filamental representations of the model simulations and the true epidemic.}
 {[}For three-state epidemic models, we provide the
@@
 around the top \(1-\alpha\) proportion of the simulations. In this
-visual setting, comparing this prediction to the true epidemic
-trajectory or comparing the prediction regions defined by two different
-models' simulations can be done by eye. In \pkg{EpiCompare}, we also
-provide these prediction regions for epidemic models with with more than
-three states. The functions \texttt{create\_convex\_hull\_structure()}
-and \texttt{create\_delta\_ball\_structure()} create different geometric
+visual setting, comparing this prediction \textcolor{violet}{region} to
+the true epidemic trajectory
+\textcolor{violet}{\sout{or comparing the prediction regions defined by two different models' simulations}}\footnote{\textcolor{violet}{[Ben says: this part will be highlighted in later sentences.]}}
+can be done by eye. In \pkg{EpiCompare}, we also provide these
+prediction regions for epidemic models with with more than three states.
+The functions \texttt{create\_convex\_hull\_structure()} and
+\texttt{create\_delta\_ball\_structure()} create different geometric
 representations of prediction regions for any dimensional state-based
@@
 \texttt{hausdorff\_dist()}
-\textcolor{violet}{function to calculate the Hausdorff distance between multiple prediction regions.}
+\textcolor{violet}{function to calculate the Hausdorff distance between multiple prediction regions, when visual comparison is not possible.}
 
@@
 Section \ref{sec:beyond-r0-sir}.{]}
-\textcolor{violet}{We provide implementations of a few score functions that capture extremeness of epidemics relative to other epidemics, and these scores can be turned into an extremeness measure with}
+\textcolor{violet}{We provide implementations of a few distance-based score functions that capture how "reasonable" an epidemic is relative to other epidemics, and these scores can be turned into an extremeness measure with}
 \texttt{mean(sim\_scores\ \textgreater{}\ truth\_score)}.

### 2021-05-24 jss/EpiCompare-paper/EpiCompare-paper.Rmd
Tried again on intro to 4

@@
 
-\shannon{[[NEW TEXT}
+\shannon{[[NEWER TEXT}
+
+To conclude our paper, we demonstrate the capabilities of \pkg{EpiCompare} with a complete data analysis of a measles outbreak in 1861-1862 Germany.
+Specifically, we demonstrate how tools in \pkg{EpiCompare} can be used in each step of the data analysis pipeline (see Fig.
+\ref{fig:pipeline}).
+Additionally, we highlight how time-invariant analysis (see Section \ref{sec:time-invariant}) can be used to enhance understanding of an outbreak.
+
+Before demonstrating \pkg{EpiCompare}, we provide some context for the measles outbreak presented here.
+The data was originally organized by \cite{pfeilsticker1863}, later made visible by \cite{oesterle1992}, and made available in an \proglang{R} by \cite{surveillance2017}.
+This data set includes a rich collection of features including household location, class level, and allged infector ID, and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}.
+In this data set, there are 188 children who became infected with the measles over the course approximately 90 days.
+
+\shannon{]]}
+
+\shannon{[[LESS NEW TEXT}
 
@@
 
-The Hagelloch data include a rich set of features at the individual level.
+The Hagelloch data include a rich set of features at the individual level, \shannon{and the tools in \pkg{EpiCompare} help with pre-processsing and EDA}.
 Recorded features include household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.
 A subset of the data is shown in Table \ref{tab:hags-people}.
-With \pkg{EpiCompare}, we can easily \shannon{pre-process the data to} obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate()}.
+\shannon{For example,} with \pkg{EpiCompare}, we can easily \shannon{pre-process the data to} obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate()}.
 The function \code{agents_to_aggregate()} is a key component of \pkg{EpiCompare}, allowing the user to easily switch from an individual-level (i.e.

### 2021-05-24 jss/EpiCompare-paper/EpiCompare-paper.tex
Tried again on intro to 4

@@
 
-\textcolor{orange}{[[NEW TEXT}
+\textcolor{orange}{[[NEWER TEXT}
+
+To conclude our paper, we demonstrate the capabilities of
+\pkg{EpiCompare} with a complete data analysis of a measles outbreak in
+1861-1862 Germany. Specifically, we demonstrate how tools in
+\pkg{EpiCompare} can be used in each step of the data analysis pipeline
+(see Fig. \ref{fig:pipeline}). Additionally, we highlight how
+time-invariant analysis (see Section \ref{sec:time-invariant}) can be
+used to enhance understanding of an outbreak.
+
+Before demonstrating \pkg{EpiCompare}, we provide some context for the
+measles outbreak presented here. The data was originally organized by
+\cite{pfeilsticker1863}, later made visible by \cite{oesterle1992}, and
+made available in an \proglang{R} by \cite{surveillance2017}. This data
+set includes a rich collection of features including household location,
+class level, and allged infector ID, and is an ideal testing ground for
+methodology in infectious disease epidemiology
+\cite{Neal2004,britton2011,groendyke2012,becker2016}. In this data set,
+there are 188 children who became infected with the measles over the
+course approximately 90 days.
+
+\textcolor{orange}{]]}
+
+\textcolor{orange}{[[LESS NEW TEXT}
 
@@
 The Hagelloch data include a rich set of features at the individual
-level. Recorded features include household members, school level,
-household locations, date of first symptoms (prodromes), date of measles
-rash, and even the alleged infector. A subset of the data is shown in
-Table \ref{tab:hags-people}. With \pkg{EpiCompare}, we can easily
+level,
+\textcolor{orange}{and the tools in \pkg{EpiCompare} help with pre-processsing and EDA}.
+Recorded features include household members, school level, household
+locations, date of first symptoms (prodromes), date of measles rash, and
+even the alleged infector. A subset of the data is shown in Table
+\ref{tab:hags-people}. \textcolor{orange}{For example,} with
+\pkg{EpiCompare}, we can easily
 \textcolor{orange}{pre-process the data to} obtain the empirical

### 2021-05-24 jss/EpiCompare-paper/EpiCompare-paper.Rmd
early comments on section 4 - more global thoughts

--- 2021-05-22 jss/EpiCompare-paper/EpiCompare-paper.Rmd
@@
 
-\shannon{[[NEW TEXT}
+\shannon{[[NEWER TEXT}
+
+\footnote{\ben{[Ben says: Generally I found this whole section to be pretty passive and not well motivated on why we'd actually do the analysis.]}}
+To conclude our paper, we demonstrate the capabilities of \pkg{EpiCompare} with a complete data analysis of a measles outbreak in 1861-1862 Germany.
+Specifically, we demonstrate how tools in \pkg{EpiCompare} can be used in each step of the data analysis pipeline \footnote{\ben{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]}} \ben{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\ben{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that?
+/ what does it add to the conv/ why should they reference it?]}}.
+Additionally, we highlight how time-invariant analysis (see Section \ref{sec:time-invariant}) can be used to enhance understanding of an outbreak\footnote{\ben{[Ben says: I'm not sure this is necessary to state.
+Additionally - is it really true do we highlight this?]}}.
+
+\footnote{\ben{[Ben says:The data introduction could be another subsection.
+Why does it fit better here?.
+Other comment: This paragraph is pretty rambly.
+It is unclear what you want someone to take away from it.
+I might suggest highlight the fact that the data is "an ideal testing group for methodology".]}} Before demonstrating \pkg{EpiCompare}, we provide some context for the measles outbreak presented here.\footnote{\ben{[Ben says: this first sentence is pretty sign-post-y yet it only relates to the next few sentences.
+Update so it's less sign-post-y.]}}  The data was originally organized by \cite{pfeilsticker1863}, later made visible by \cite{oesterle1992}, and made available in an \proglang{R} by \cite{surveillance2017}.
+This data set includes a rich collection of features including household location, class level, and allged infector ID, and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}\footnote{\ben{[Ben says: This citation doesn't make sense - should it be "citep"?]}}.
+In this data set, there are 188 children who became infected with the measles over the course approximately 90 days.
+
+\shannon{]]}
+
+\shannon{[[LESS NEW TEXT}
 
@@
 
-The Hagelloch data include a rich set of features at the individual level.
+\footnote{\ben{[Ben says: The first two sentences is very similar to the data paragraph in the section above.
+Given it's not really connecting the 2 sections I suggest a rewrite - could move some of this stuff above.]}}The Hagelloch data include a rich set of features at the individual level, \shannon{and the tools in \pkg{EpiCompare} help with pre-processsing and EDA}.
 Recorded features include household members, school level, household locations, date of first symptoms (prodromes), date of measles rash, and even the alleged infector.
 A subset of the data is shown in Table \ref{tab:hags-people}.
-With \pkg{EpiCompare}, we can easily \shannon{pre-process the data to} obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate()}.
+\shannon{For example,}\footnote{\ben{[Ben says: I'm unclear of what this is actually an example of.]}} with \pkg{EpiCompare}, we can easily \shannon{pre-process the data to} obtain the empirical cumulative incidence function with respect to the measles rash appearance (variable \code{ERU}) with the following tidy-style function, \code{agents_to_aggregate()}.
 The function \code{agents_to_aggregate()} is a key component of \pkg{EpiCompare}, allowing the user to easily switch from an individual-level (i.e.
 an agent) \sout{view} lens of a disease to an aggregate \sout{level} lens.
-For example, the below code shows how we can convert the agent data to a cumulative incidence \shannon{plot} of the measles rash\sout{, in order to see how the disease spread through the population over time.} We can then compare the cumulative incidence of the rash to the cumulative incidence of the prodromes, i.e.
-the initial symptoms.
+For example\footnote{\ben{[Ben says: Same comment as before.]}}, the below code shows how we can convert the agent data to a cumulative incidence \shannon{plot} of the measles rash\sout{, in order to see how the disease spread through the population over time.} We can then compare the cumulative incidence of the rash to the cumulative incidence of the prodromes, i.e.
+the initial\footnote{\ben{[Ben says: please be clearer on what these could be given you comparing them to start of the rash - which seems like an early symptom to me...]}} symptoms\footnote{\ben{[Ben says: This action in the analysis pipeline is unmotivated - which naturally makes me want to ask "why would I do this?"]}}.
 We do this with the below code, and a part of the cumulative incidence data output is shown in Table \ref{tab:cif-rash}.
@@
 
-One possible question of interest is the duration between initial onset of prodromes  and the appearance of the measles rash.
+One possible question of interest is the duration between initial onset of prodromes and the appearance of the measles rash\footnote{\ben{[Ben says: You don't give a good definition of prodromes above, and you only use the name twice.
+Is this a super common term in Epi?
+I find it a bit taxing on the reader to remember what this is referring to.]}}.
 Since \code{agents_to_aggregate()} outputs a tidy-style data frame, it is a simple task to plot the two sets of incidence curves on the same graph (Fig.
@@
 ```
+
+\footnote{\ben{I'm confused why Figure 5 is included. What is the conclusion you'd like to take away? / Why do people create plots like this?}}
 
@@
 For example, the Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the time of infection and recovery, respectively of each individual.
-We can then plot the SIR values through a time-invariant lens using \pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
+We can then\footnote{\ben{[Ben says: using "then" here captures a very progression step of analysis but I stopped here and asked "what is this following" - and the previous "step" occurred a paragraph back and wasn't described as a direct progression but just a possible thing to do.]}} plot the SIR values through a time-invariant lens using \pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
 \ref{fig:hag-tern-raw}) or with our custom \code{geom}, \code{geom_aggregate}, which takes the raw agent data as input.
@@
 
-Moreover, we can look at the outbreaks of the disease by group within \code{agent_to_aggregate()} or \code{geom_aggregate()}.
+\footnote{\ben{I found this pargraph very unmotivated.
+I recommend first arguing why we might a care to look into the class subpopulation grouping.
+And maybe comment that is is a common desire for practioners.}}Moreover, we can look at the outbreaks of the disease by group within \code{agent_to_aggregate()} or \code{geom_aggregate()}.
 This allows us to examine differences among the different groups of individuals.
@@
 
-Along with multiple epidemic states, the function \code{agents_to_aggregate()} can also be extended to populations with vital dynamics (e.g.
+\ben{\sout{Along with multiple epidemic states, the function}} \code{agents_to_aggregate()} \ben{\sout{can also be extended to populations with vital dynamics (e.g.
 birth and death) and examples of this are shown in the package vignette.
-In summary, \code{agents_to_aggregate()} is a multi-purpose workhorse that may be leveraged to convert individual level records into aggregate information that may be more useful for some forms of epidemic modeling such as compartment modeling.
-
-## Modeling and Simulation
+In summary, }} \code{agents_to_aggregate()} \ben{\sout{is a multi-purpose workhorse that may be leveraged to convert individual level records into aggregate information that may be more useful for some forms of epidemic modeling such as compartment modeling.}}\footnote{\ben{Is this not just a repeat of section 3?}}
+
+## Modeling and simulation
 
@@
 Up to this point, we have used \pkg{EpiCompare} in the context of observed data.
-We also want to compare statistical models, and \pkg{EpiCompare} aids in that process via a simple yet flexible individual-level simulator\sout{, conversion tools for popular epidemic model packages, and model assessments.}  We demonstrate an example here.
-
-We first try to model the Hagelloch data with a stochastic SIR model, which we refer to as the 'simple SIR.'  In our full vignette \shannon{(available online),} we show how to fit this simple SIR model via maximum likelihood and simulate from the model with those best fit parameters.
-Our function \code{simulate_agents()} generates individual level data according to discrete time multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.
+\footnote{\ben{[Ben says: Why?]}}We also want to compare statistical models, and \pkg{EpiCompare} aids in that process via a simple yet flexible individual-level simulator\sout{, conversion tools for popular epidemic model packages, and model assessments.}  We demonstrate an example\footnote{\ben{[Ben says: to me this whole section is 1 example - as such this wording is confusing to me.]}} here.
+
+We first try to model the Hagelloch data with a stochastic SIR model, which we refer to as the 'simple SIR.'\footnote{\ben{[Ben says: could / should this be thought of as a "base" model?]}}  In our full vignette \shannon{(available online),} we show how to fit this simple SIR model via maximum likelihood and simulate from the model with those best fit parameters\footnote{\ben{[Ben says: should you highlight that this is a common approach?]}}.
+Our function \code{simulate_agents()}\footnote{\ben{[Ben Should somethings be said about the } `simulation_SIR_agents()` \ben{function?]}} generates individual level data according to discrete time multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.
 For example, the below code generates 100 simulations of an outbreak of a disease with one initial infector in a population of $n= 188$ individuals.
@@
 The result of our simulation is the object \code{agents} which is a `r nrow(agents)` $\times$ `r ncol(agents)` data frame, which details the time of entry into the $S$, $I$, and $R$ states for a given simulation.
-Before we examine the results of this simple SIR model, we will also examine another, more sophisticated SIR model, this time from the package \pkg{EpiModel} \citep{Jenness2018}.
-Briefly, this model first fits a contact network to the set of individuals, where the class of the child is a covariate.
+\footnote{\ben{[Ben says: please motivate - through model comparison?]}}Before we examine the results of this simple SIR model, we will also examine another, more sophisticated SIR model, this time from the package \pkg{EpiModel} \citep{Jenness2018}.
+Briefly, this model first fits a contact network to the set of individuals, where the class of the child is a covariate\footnote{\ben{[Ben says: Make this connect better to the problem at hand - why do you think you should build this bigger model?]}}.
 The model then simulates a SIR-epidemic on that network.
@@
 The output of this model is \code{epimodel_sir}, an object of class \code{`r class(epimodel_sir)`}, which contains a plethora of modeling information.
+\footnote{\ben{[Ben says: what's the point for this sentence.
+It also doesn't flow/ connect to previous and later text.]}}
 
@@
 
-The next step is to compare the Simple SIR model to the EpiModel SIR model.
-We provide the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transform it into a tidy-style data frame.
+The next step is to compare the simple SIR model to the EpiModel SIR model.
+We provide\footnote{\ben{[Ben says: the phrase "We provide" is very passive / distance from the current demonstration at hand.
+Moreover section 3 already phrases things this way.]}} the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transform it into a tidy-style data frame.
 
@@
 
-We can then analyze the results of the two models side by side as time-invariant epidemic curves.
-The results are shown in Figure \ref{fig:hag-simple-sir}, where a 90\% prediction band is estimated from the delta ball method for each of the two models.
-For the Simple SIR model, we see that the data generally covers the data fairly well but clearly misses the second peak of infection.
+We can then analyze the results of the two models side by side as time-invariant\footnote{\ben{[Ben says: this isn't a clear phrrase here - what are you trying to say?]}} epidemic curves.
+The results are shown in Figure \ref{fig:hag-simple-sir}, where a 90\% prediction band is estimated from the delta ball\footnote{\ben{[Ben says: this has never been discussed anyway.]}} method for each of the two models.
+For the Simple SIR model, we see that the data generally covers the data fairly well but clearly misses the second peak of infection\footnote{\ben{[Ben says: This could be better movitated with talk of model fit...]}}.
 We also see that the prediction band is very large, covering up a large area of the ternary plot.
@@
 
-g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0),
+g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0) %>%
+                  mutate(Type = factor(Type, levels = c("Simple SIR", 
+                                                        "EpiModel SIR"))),
          aes(x = X0, y = X1, z = X2,
@@
 
+\ben{[Ben says: In figure 8 I changed the order of the facets given we talk about the simple model first and its more like the "base" model. I think the title should be changed?]}
+
 ```{r , fig.cap = "\\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models."}
-g +   geom_path(data = both_models %>% filter(t !=0),
+g +   geom_path(data = both_models %>% filter(t != 0) %>%
+                  mutate(Type = factor(Type, levels = c("Simple SIR", 
+                                                        "EpiModel SIR"))),
             aes(x = X0, y = X1, z = X2, group = paste(Type, sim)),
@@
 However, both models are not a good fit to the filamental path as opposed to the individual points in $(S, I, R)$-space.
-This can be captured with the set of simulations both models predict (gray lines), which all generally have a single defined peak of infection whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
-This observation is backed up by the below analysis that demonstrates that the estimated pseudo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme})\footnote{\shannon{Ben, do we want to add another sentence or two explaining the two columns in the table?
-The second one I think makes sense to me but not the first.}} In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the EpiModel network model is a better fit than the Simple SIR model, and 2) the fit is only good at the \sout{geometric filamental level as opposed to the epidemic trajectory filamental level.} \shannon{individual point level as opposed to the geometric filamental level.}
+This can be\footnote{\ben{[Ben says: this is a very passive way to say such things.
+Try being more direct.]}} captured with the set of simulations both models predict (gray lines), which all generally have a single defined peak of infection whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
+This observation is backed up\footnote{\ben{[Ben says: describe this?]}} by the below analysis that demonstrates that the estimated pseudo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme})\footnote{\shannon{Ben, do we want to add another sentence or two explaining the two columns in the table?
+The second one I think makes sense to me but not the first.}} In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the EpiModel network model is a better fit than the Simple SIR model, and 2) the fit is only good at the \sout{geometric filamental level as opposed to the epidemic trajectory filamental level.} \shannon{individual point level as opposed to the geometric filamental level.}\footnote{\ben{[Ben says: how would this look with the time plots?
+Do we add value here?]}}
 
@@
 
-Overall, \pkg{EpiCompare} aids in the data analysis pipeline for both novice and expert practitioners and coders alike.
+\footnote{\ben{I think this paragraph captures some good goals, but I don't think we've done some of this.
+For example - we don't really highlight novice/expert usage, and we don't highlight side-by-side comparisons of models.}}Overall, \pkg{EpiCompare} aids in the data analysis pipeline for both novice and expert practitioners and coders alike.
 These tools encourage model and simulation exploration of many of the existing and well-supported packages that already exist, and side-by-side comparison thereof.

### 2021-05-24 jss/EpiCompare-paper/EpiCompare-paper.tex
early comments on section 4 - more global thoughts

--- 2021-05-22 jss/EpiCompare-paper/EpiCompare-paper.tex
@@
 
-\textcolor{orange}{[[NEW TEXT}
+\textcolor{orange}{[[NEWER TEXT}
+
+\footnote{\textcolor{violet}{[Ben says: Generally I found this whole section to be pretty passive and not well motivated on why we'd actually do the analysis.]}}
+To conclude our paper, we demonstrate the capabilities of
+\pkg{EpiCompare} with a complete data analysis of a measles outbreak in
+1861-1862 Germany. Specifically, we demonstrate how tools in
+\pkg{EpiCompare} can be used in each step of the data analysis pipeline
+\footnote{\textcolor{violet}{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]}}
+\textcolor{violet}{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\textcolor{violet}{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that?
+/ what does it add to the conv/ why should they reference it?]}}.
+Additionally, we highlight how time-invariant analysis (see Section
+\ref{sec:time-invariant}) can be used to enhance understanding of an
+outbreak\footnote{\textcolor{violet}{[Ben says: I'm not sure this is necessary to state. Additionally - is it really true do we highlight this?]}}.
+
+\footnote{\textcolor{violet}{[Ben says:The data introduction could be another subsection.
+Why does it fit better here?.
+Other comment: This paragraph is pretty rambly.
+It is unclear what you want someone to take away from it.
+I might suggest highlight the fact that the data is "an ideal testing group for methodology".]}}
+Before demonstrating \pkg{EpiCompare}, we provide some context for the
+measles outbreak presented
+here.\footnote{\textcolor{violet}{[Ben says: this first sentence is pretty sign-post-y yet it only relates to the next few sentences. Update so it's less sign-post-y.]}}
+The data was originally organized by \cite{pfeilsticker1863}, later made
+visible by \cite{oesterle1992}, and made available in an \proglang{R} by
+\cite{surveillance2017}. This data set includes a rich collection of
+features including household location, class level, and allged infector
+ID, and is an ideal testing ground for methodology in infectious disease
+epidemiology
+\cite{Neal2004,britton2011,groendyke2012,becker2016}\footnote{\textcolor{violet}{[Ben says: This citation doesn't make sense - should it be "citep"?]}}.
+In this data set, there are 188 children who became infected with the
+measles over the course approximately 90 days.
+
+\textcolor{orange}{]]}
+
+\textcolor{orange}{[[LESS NEW TEXT}
 
@@
 
-The Hagelloch data include a rich set of features at the individual
-level. Recorded features include household members, school level,
-household locations, date of first symptoms (prodromes), date of measles
-rash, and even the alleged infector. A subset of the data is shown in
-Table \ref{tab:hags-people}. With \pkg{EpiCompare}, we can easily
+\footnote{\textcolor{violet}{[Ben says: The first two sentences is very similar to the data paragraph in the section above.
+Given it's not really connecting the 2 sections I suggest a rewrite - could move some of this stuff above.]}}The
+Hagelloch data include a rich set of features at the individual level,
+\textcolor{orange}{and the tools in \pkg{EpiCompare} help with pre-processsing and EDA}.
+Recorded features include household members, school level, household
+locations, date of first symptoms (prodromes), date of measles rash, and
+even the alleged infector. A subset of the data is shown in Table
+\ref{tab:hags-people}.
+\textcolor{orange}{For example,}\footnote{\textcolor{violet}{[Ben says: I'm unclear of what this is actually an example of.]}}
+with \pkg{EpiCompare}, we can easily
 \textcolor{orange}{pre-process the data to} obtain the empirical
@@
 switch from an individual-level (i.e.~an agent) \sout{view} lens of a
-disease to an aggregate \sout{level} lens. For example, the below code
-shows how we can convert the agent data to a cumulative incidence
-\textcolor{orange}{plot} of the measles
+disease to an aggregate \sout{level} lens. For
+example\footnote{\textcolor{violet}{[Ben says: Same comment as before.]}},
+the below code shows how we can convert the agent data to a cumulative
+incidence \textcolor{orange}{plot} of the measles
 rash\sout{, in order to see how the disease spread through the population over time.}
 We can then compare the cumulative incidence of the rash to the
-cumulative incidence of the prodromes, i.e.~the initial symptoms. We do
-this with the below code, and a part of the cumulative incidence data
-output is shown in Table \ref{tab:cif-rash}. The argument
+cumulative incidence of the prodromes, i.e.~the
+initial\footnote{\textcolor{violet}{[Ben says: please be clearer on what these could be given you comparing them to start of the rash - which seems like an early symptom to me...]}}
+symptoms\footnote{\textcolor{violet}{[Ben says: This action in the analysis pipeline is unmotivated - which naturally makes me want to ask "why would I do this?"]}}.
+We do this with the below code, and a part of the cumulative incidence
+data output is shown in Table \ref{tab:cif-rash}. The argument
 \code{integer_time_expansion} indicates whether we should include all
@@
 One possible question of interest is the duration between initial onset
-of prodromes and the appearance of the measles rash. Since
-\code{agents_to_aggregate()} outputs a tidy-style data frame, it is a
-simple task to plot the two sets of incidence curves on the same graph
-(Fig. \ref{fig:cifs}).
+of prodromes and the appearance of the measles
+rash\footnote{\textcolor{violet}{[Ben says: You don't give a good definition of prodromes above, and you only use the name twice.
+Is this a super common term in Epi?
+I find it a bit taxing on the reader to remember what this is referring to.]}}.
+Since \code{agents_to_aggregate()} outputs a tidy-style data frame, it
+is a simple task to plot the two sets of incidence curves on the same
+graph (Fig. \ref{fig:cifs}).
 
@@
 \end{CodeChunk}
+
+\footnote{\textcolor{violet}{I'm confused why Figure 5 is included. What is the conclusion you'd like to take away? / Why do people create plots like this?}}
 
@@
 time of infection and recovery, respectively of each individual. We can
-then plot the SIR values through a time-invariant lens using
-\pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
-\ref{fig:hag-tern-raw}) or with our custom \code{geom},
-\code{geom_aggregate}, which takes the raw agent data as input.
+then\footnote{\textcolor{violet}{[Ben says: using "then" here captures a very progression step of analysis but I stopped here and asked "what is this following" - and the previous "step" occurred a paragraph back and wasn't described as a direct progression but just a possible thing to do.]}}
+plot the SIR values through a time-invariant lens using \pkg{ggplot2}
+and \pkg{ggtern} functions (as shown in Fig. \ref{fig:hag-tern-raw}) or
+with our custom \code{geom}, \code{geom_aggregate}, which takes the raw
+agent data as input.
 
@@
 
-Moreover, we can look at the outbreaks of the disease by group within
+\footnote{\textcolor{violet}{I found this pargraph very unmotivated.
+I recommend first arguing why we might a care to look into the class subpopulation grouping.
+And maybe comment that is is a common desire for practioners.}}Moreover,
+we can look at the outbreaks of the disease by group within
 \code{agent_to_aggregate()} or \code{geom_aggregate()}. This allows us
@@
 
-Along with multiple epidemic states, the function
-\code{agents_to_aggregate()} can also be extended to populations with
-vital dynamics (e.g.~birth and death) and examples of this are shown in
-the package vignette. In summary, \code{agents_to_aggregate()} is a
-multi-purpose workhorse that may be leveraged to convert individual
-level records into aggregate information that may be more useful for
-some forms of epidemic modeling such as compartment modeling.
+\textcolor{violet}{\sout{Along with multiple epidemic states, the function}}
+\code{agents_to_aggregate()}
+\textcolor{violet}{\sout{can also be extended to populations with vital dynamics (e.g. birth and death) and examples of this are shown in the package vignette.  In summary, }}
+\code{agents_to_aggregate()}
+\textcolor{violet}{\sout{is a multi-purpose workhorse that may be leveraged to convert individual level records into aggregate information that may be more useful for some forms of epidemic modeling such as compartment modeling.}}\footnote{\textcolor{violet}{Is this not just a repeat of section 3?}}
 
 \hypertarget{modeling-and-simulation}{%
-\subsection{Modeling and Simulation}\label{modeling-and-simulation}}
+\subsection{Modeling and simulation}\label{modeling-and-simulation}}
 
@@
 Up to this point, we have used \pkg{EpiCompare} in the context of
-observed data. We also want to compare statistical models, and
-\pkg{EpiCompare} aids in that process via a simple yet flexible
-individual-level
+observed data. \footnote{\textcolor{violet}{[Ben says: Why?]}}We also
+want to compare statistical models, and \pkg{EpiCompare} aids in that
+process via a simple yet flexible individual-level
 simulator\sout{, conversion tools for popular epidemic model packages, and model assessments.}
-We demonstrate an example here.
+We demonstrate an
+example\footnote{\textcolor{violet}{[Ben says: to me this whole section is 1 example - as such this wording is confusing to me.]}}
+here.
 
 We first try to model the Hagelloch data with a stochastic SIR model,
-which we refer to as the `simple SIR.' In our full vignette
-\textcolor{orange}{(available online),} we show how to fit this simple
-SIR model via maximum likelihood and simulate from the model with those
-best fit parameters. Our function \code{simulate_agents()} generates
-individual level data according to discrete time multinomial draws,
-which depend on the number of individuals in each state at the previous
-time step and a matrix of transition probabilities. For example, the
-below code generates 100 simulations of an outbreak of a disease with
-one initial infector in a population of \(n= 188\) individuals.
+which we refer to as the `simple
+SIR.'\footnote{\textcolor{violet}{[Ben says: could / should this be thought of as a "base" model?]}}
+In our full vignette \textcolor{orange}{(available online),} we show how
+to fit this simple SIR model via maximum likelihood and simulate from
+the model with those best fit
+parameters\footnote{\textcolor{violet}{[Ben says: should you highlight that this is a common approach?]}}.
+Our function
+\code{simulate_agents()}\textbackslash footnote\{\textcolor{violet}{[Ben Should somethings be said about the }
+\texttt{simulation\_SIR\_agents()} \textcolor{violet}{function?]}\}
+generates individual level data according to discrete time multinomial
+draws, which depend on the number of individuals in each state at the
+previous time step and a matrix of transition probabilities. For
+example, the below code generates 100 simulations of an outbreak of a
+disease with one initial infector in a population of \(n= 188\)
+individuals.
 
@@
 18800 \(\times\) 5 data frame, which details the time of entry into the
-\(S\), \(I\), and \(R\) states for a given simulation. Before we examine
-the results of this simple SIR model, we will also examine another, more
-sophisticated SIR model, this time from the package \pkg{EpiModel}
-\citep{Jenness2018}. Briefly, this model first fits a contact network to
-the set of individuals, where the class of the child is a covariate. The
-model then simulates a SIR-epidemic on that network.
+\(S\), \(I\), and \(R\) states for a given simulation.
+\footnote{\textcolor{violet}{[Ben says: please motivate - through model comparison?]}}Before
+we examine the results of this simple SIR model, we will also examine
+another, more sophisticated SIR model, this time from the package
+\pkg{EpiModel} \citep{Jenness2018}. Briefly, this model first fits a
+contact network to the set of individuals, where the class of the child
+is a
+covariate\footnote{\textcolor{violet}{[Ben says: Make this connect better to the problem at hand - why do you think you should build this bigger model?]}}.
+The model then simulates a SIR-epidemic on that network.
 
@@
 \code{netsim}, which contains a plethora of modeling information.
+\footnote{\textcolor{violet}{[Ben says: what's the point for this sentence. It also doesn't flow/ connect to previous and later text.]}}
 
@@
 
-The next step is to compare the Simple SIR model to the EpiModel SIR
-model. We provide the function \code{fortify_aggregate()}, which can
-take objects from specialized classes of modeling output
+The next step is to compare the simple SIR model to the EpiModel SIR
+model. We
+provide\footnote{\textcolor{violet}{[Ben says: the phrase "We provide" is very passive / distance from the current demonstration at hand. Moreover section 3 already phrases things this way.]}}
+the function \code{fortify_aggregate()}, which can take objects from
+specialized classes of modeling output
 \textcolor{orange}{(like those made by \code{netsim()})} and transform
@@
 We can then analyze the results of the two models side by side as
-time-invariant epidemic curves. The results are shown in Figure
+time-invariant\footnote{\textcolor{violet}{[Ben says: this isn't a clear phrrase here - what are you trying to say?]}}
+epidemic curves. The results are shown in Figure
 \ref{fig:hag-simple-sir}, where a 90\% prediction band is estimated from
-the delta ball method for each of the two models. For the Simple SIR
-model, we see that the data generally covers the data fairly well but
-clearly misses the second peak of infection. We also see that the
-prediction band is very large, covering up a large area of the ternary
-plot. On the other hand, for the EpiModel network model, we see that the
-prediction band covers the data quite well and takes up less area.
+the delta
+ball\footnote{\textcolor{violet}{[Ben says: this has never been discussed anyway.]}}
+method for each of the two models. For the Simple SIR model, we see that
+the data generally covers the data fairly well but clearly misses the
+second peak of
+infection\footnote{\textcolor{violet}{[Ben says: This could be better movitated with talk of model fit...]}}.
+We also see that the prediction band is very large, covering up a large
+area of the ternary plot. On the other hand, for the EpiModel network
+model, we see that the prediction band covers the data quite well and
+takes up less area.
 
@@
 R> 
-R> g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0),
+R> g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0) %>%
++                   mutate(Type = factor(Type, levels = c("Simple SIR", 
++                                                         "EpiModel SIR"))),
 +          aes(x = X0, y = X1, z = X2,
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-R> g +   geom_path(data = both_models %>% filter(t !=0),
+\textcolor{violet}{[Ben says: In figure 8 I changed the order of the facets given we talk about the simple model first and its more like the "base" model. I think the title should be changed?]}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> g +   geom_path(data = both_models %>% filter(t != 0) %>%
++                   mutate(Type = factor(Type, levels = c("Simple SIR", 
++                                                         "EpiModel SIR"))),
 +             aes(x = X0, y = X1, z = X2, group = paste(Type, sim)),
@@
 However, both models are not a good fit to the filamental path as
-opposed to the individual points in \((S, I, R)\)-space. This can be
+opposed to the individual points in \((S, I, R)\)-space. This can
+be\footnote{\textcolor{violet}{[Ben says: this is a very passive way to say such things. Try being more direct.]}}
 captured with the set of simulations both models predict (gray lines),
@@
 data certainly looks like it has two distinct peaks, likely caused by
-our assumed super-spreader event. This observation is backed up by the
-below analysis that demonstrates that the estimated pseudo-density of
-the observed epidemic (relative to the simulations from either model) is
+our assumed super-spreader event. This observation is backed
+up\footnote{\textcolor{violet}{[Ben says: describe this?]}} by the below
+analysis that demonstrates that the estimated pseudo-density of the
+observed epidemic (relative to the simulations from either model) is
 much less likely then \textbf{any} of the simulations (reported in Table
@@
 \sout{geometric filamental level as opposed to the epidemic trajectory filamental level.}
-\textcolor{orange}{individual point level as opposed to the geometric filamental level.}
+\textcolor{orange}{individual point level as opposed to the geometric filamental level.}\footnote{\textcolor{violet}{[Ben says: how would this look with the time plots? Do we add value here?]}}
 
@@
 
-Overall, \pkg{EpiCompare} aids in the data analysis pipeline for both
-novice and expert practitioners and coders alike. These tools encourage
-model and simulation exploration of many of the existing and
-well-supported packages that already exist, and side-by-side comparison
-thereof. Finally, we hope that practicioners will consider using
-time-invariant analysis when trying to assess and compare epidemics and
-epidemic models.
+\footnote{\textcolor{violet}{I think this paragraph captures some good goals, but I don't think we've done some of this.
+For example - we don't really highlight novice/expert usage, and we don't highlight side-by-side comparisons of models.}}Overall,
+\pkg{EpiCompare} aids in the data analysis pipeline for both novice and
+expert practitioners and coders alike. These tools encourage model and
+simulation exploration of many of the existing and well-supported
+packages that already exist, and side-by-side comparison thereof.
+Finally, we hope that practicioners will consider using time-invariant
+analysis when trying to assess and compare epidemics and epidemic
+models.
 

### 2021-06-07 jss/EpiCompare-paper/EpiCompare-paper.Rmd
major section 4 edits
also cleaned up end of section 3

@@
 **Comparisons and assessment**
-\shannon{[[ NEW TEXT}
-
-\ben{\sout{Finally,} Often}\footnote{\ben{[Ben says: I feel like starting with "finally" causes a weird pause - but maybe that's just me.]}} the data analysis pipeline \ben{\sout{often}} ends with plots, tables, and summary statistics that are used to \sout{\ben{compare} \shannon{and assess}} \ben{\sout{different models, and simulations to one another} assess model preformance and compare across models or simulations.} In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
+
+The last step of the data analysis pipeline often ends with plots, tables, and summary statistics that are used to assess model performance and compare across models or simulations.
+In \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
 mean squared error or AIC) and into the lens of time-invariant analysis.
-We have found \ben{that} these tools \ben{\sout{to be} are} specifically applicable for situations where only one season or cycle of an epidemic has occurred, \ben{or is the object of interest}.
-
-\footnote{\ben{[Ben says: This changed introduction was done to highlight that we can use prediction regions for mutiple types of comparisons.]}}\ben{The first set of of tools surround the creation of prediction regions.
+We have found that these tools are specifically applicable for situations where only one season or cycle of an epidemic has occurred or is the object of interest.
+
+The first set of of tools surround the creation of prediction regions.
 We can create a prediction regions from model simulations to examine if our model simulations capture the true epidemic trajectory.
-We do so in a time-invariant way and utilizing filamental representations of the model simulations and the true epidemic.} [For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
-In this visual setting, comparing this prediction \ben{region} to the true epidemic trajectory \ben{\sout{or comparing the prediction regions defined by two different models' simulations}}\footnote{\ben{[Ben says: this part will be highlighted in later sentences.]}} can be done by eye.
+We do so in a time-invariant way and utilizing filamental representations of the model simulations and the true epidemic.
+For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
+In this visual setting, comparing this prediction region to the true epidemic trajectory can be done by eye.
 In \pkg{EpiCompare}, we also provide these prediction regions for epidemic models with with more than three states.
@@
 For both of these geometric structures, we provide functions to check if a path is contained (`contained()`)].
-\ben{We can also use these prediction regions to visually or mathematically compare how similar two sets of simulations are.
-In \pkg{EpiCompare} we provide the} `hausdorff_dist()` \ben{function to calculate the Hausdorff distance between multiple prediction regions, when visual comparison is not possible.}
-
-[We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.] \ben{We provide implementations of a few distance-based score functions that capture how "reasonable" an epidemic is relative to other epidemics, and these scores can be turned into an extremeness measure with} `mean(sim_scores > truth_score)`.
+We can also use these prediction regions to visually or mathematically compare how similar two sets of simulations are.
+In \pkg{EpiCompare} we provide the `hausdorff_dist()` function to calculate the Hausdorff distance between multiple prediction regions, when visual comparison is not possible.
+
+ We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.] We provide implementations of a few distance-based score functions that capture how "reasonable" an epidemic is relative to other epidemics, and these scores can be turned into an extremeness measure with `mean(sim_scores > truth_score)`.
 [Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
-Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.]
-
-]]
-
-[[\textcolor{red}{ OLD TEXT}
-
-One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
-If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space.
-For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
-In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done by eye.
-In \pkg{EpiCompare}, we also provide these prediction regions for epidemic models with with more than three states.
-The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model.
-For both of these geometric structures, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hausdorff_dist()`).
-The vignettes for `EpiCompare` go into much more mathematical detail about how these functions can be used in conjunction with time-invariant analysis.
-
-We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
-Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
 Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
 
-\shannon{]]}
-
-[[\textcolor{red}{Really OLD TEXT}
-
-Finally, in \pkg{EpiCompare} we provide a set of comparison and assessment tools for model and simulation results that extend beyond the standard performance metrics (e.g.
-mean squared error or AIC).
-Corresponding\footnote{\ben{[Ben says: "corresponding to" doesn't seem to be the correct phrase here.
-If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect word choice.]}\shannon{"To support the ideas presented in Sec."?}} to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of time-invariant tools to compare and evaluate epidemic models and simulations.
-We have found these tools to be specifically applicable for situations where only one "season" or "cycle"\footnote{\ben{[Ben says: I think we can remove the paratheses here.]} \shannon{accepted}} of the epidemic has occurred.
-
-\footnote{\ben{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\shannon{motivate better}}One tool we provide to assess models is through the creation of geometric prediction regions, which are useful when we treat epidemics like filaments.
-If there is a set of simulated epidemics from a model, we can create a geometric prediction region for the expected trajectory of the epidemic in the state space.
-For three-state epidemic models, we provide the `ggplot`/`ggtern` extension `geom_prediction_band()` which creates a prediction region around the top $1-\alpha$ proportion of the simulations.
-In this visual setting, comparing this prediction to the true epidemic trajectory or comparing the prediction regions defined by two different models' simulations can be done \ben{by eye.\sout{with the 'eye test.'}\shannon{I'm not saying 'no' but I'd do it in an appendix and not the main text.
-I don't think equations belong in Section 3.}} In \pkg{EpiCompare} we also provide these prediction regions for epidemic models with with more than three states.
-The functions `create_convex_hull_structure()` and `create_delta_ball_structure()` create different geometric representations of prediction regions for any dimensional state-based model.
-For both of these geometric structures, we provide functions to check if a path is contained (`contained()`) and the ability to assess the Hausdorff distance between prediction regions based on simulations from different model (`hausdorff_dist()`).
-
-\footnote{\ben{[Ben says: random thought (not needed) - maybe we do provide some equations for these measures and also maybe mathematically express how we'd examine extremeness?]}}We also provide functions to calculate the "extremeness" of a true epidemic trajectory compared to simulated epidemics via the equi-distance filamental trajectory representation as mentioned in Section \ref{sec:beyond-r0-sir}.
-Specifically, functions like  `distance_pseudo_density_function()` can calculate a pseudo-density estimate of the true epidemic relative to simulated ones.
-Functions `distance_depth_function()` and `local_distance_depth_function()` provide depth scores that suggest how geometrically central an epidemic is to simulations.
-]]
-
 # A tour of \pkg{EpiCompare} {short-title="Tour" #sec:tour}
-
-\shannon{[[NEWER TEXT}
-
-\footnote{\ben{[Ben says: Generally I found this whole section to be pretty passive and not well motivated on why we'd actually do the analysis.]}}
-To conclude our paper, we demonstrate the capabilities of \pkg{EpiCompare} with a complete data analysis of a measles outbreak in 1861-1862 Germany.
-Specifically, we demonstrate how tools in \pkg{EpiCompare} can be used in each step of the data analysis pipeline \footnote{\ben{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]}} \ben{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\ben{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that?
-/ what does it add to the conv/ why should they reference it?]}}.
-Additionally, we highlight how time-invariant analysis (see Section \ref{sec:time-invariant}) can be used to enhance understanding of an outbreak\footnote{\ben{[Ben says: I'm not sure this is necessary to state.
-Additionally - is it really true do we highlight this?]}}.
-
-\footnote{\ben{[Ben says:The data introduction could be another subsection.
-Why does it fit better here?.
-Other comment: This paragraph is pretty rambly.
-It is unclear what you want someone to take away from it.
-I might suggest highlight the fact that the data is "an ideal testing group for methodology".]}} Before demonstrating \pkg{EpiCompare}, we provide some context for the measles outbreak presented here.\footnote{\ben{[Ben says: this first sentence is pretty sign-post-y yet it only relates to the next few sentences.
-Update so it's less sign-post-y.]}}  The data was originally organized by \cite{pfeilsticker1863}, later made visible by \cite{oesterle1992}, and made available in an \proglang{R} by \cite{surveillance2017}.
-This data set includes a rich collection of features including household location, class level, and allged infector ID, and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}\footnote{\ben{[Ben says: This citation doesn't make sense - should it be "citep"?]}}.
-In this data set, there are 188 children who became infected with the measles over the course approximately 90 days.
-
-\shannon{]]}
-
-\shannon{[[LESS NEW TEXT}
-
-Finally, in this section we show how tools from \pkg{EpiCompare} can be used in each step of the data analysis pipeline shown in Fig.
-\ref{fig:pipeline}.
+\shannon{I have substantially overhauled section 4.
+The older versions are kept after the entirety of this section but note that much of the text (but not the structure) of this section is different.
+I've tried to better motivate why we use the epicompare functions to answer our questions of interest.}
+
+\shannon{[[NEWEST TEXT}
+To conclude our paper, we demonstrate how \pkg{EpiCompare}  can streamline the data analysis process with a case study of a measles outbreak in 1861-1862 Germany.
+With the help of \pkg{EpiCompare}, we can answer questions such as is a SIR model a good fit for the data, does the outbreak spread differently within the different school classes of the children, and can incorporating an underlying network structure enhance model fit.
+We first begin with some context for the epidemic.
+
+## Background for 1861-1862 measles epidemic
+
  We analyze an outbreak of measles in the town of Hagelloch, Germany from 1861-1862, a data set organized by \cite{pfeilsticker1863}.
 The data was later made visible by \cite{oesterle1992} and made available in an \proglang{R} by \cite{surveillance2017}.
-This data set includes a rich collection of features and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\shannon{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's going on.}}
-
-\shannon{]]}
-
-[[\textcolor{red}{OLD TEXT}
-
-\footnote{\ben{[Ben says: Shannon, would you mind reading this whole section over again once we've finished edits for section 2 and 3?
-This initial paragraph seems to be stating section 3's story.]}}In this section, we highlight many of the tools available in \pkg{EpiCompare}.
-As previously discussed, these tools include data cleaning; visualization; modeling and simulation; post-processing; and comparison and model assessment, in accordance with the data analysis pipeline (Fig.
-\ref{fig:pipeline}).
-We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner via \pkg{EpiCompare}.
-]]
-
-\subsection{Pre-processing and EDA}
+In this outbreak, 188 children were infected with measles over the course of three months.
+This data set includes a rich collection of features including day of measles rash, age, sex, school class, household and household location, and alleged infector of each child.
+We show a subset of the data in Table \ref{tab:hags-people}.
+We are particularly interested if and how group structure plays a role in the spread of measles.
 
@@
 ```
+
+## Pre-processing and EDA
+
+We begin our analysis by first examining and transforming the raw data (\code{hagelloch_raw}), which is  individual-level data.
+With \pkg{EpiCompare}, we can quickly determine how the epidemic progresses through the  population at an aggregate level.
+By specifying the time of infection (\code{tI}) and the time of recovery (\code{tR}), the function \code{agents_to_aggregate()} calculates the number susceptible, infectious, and recovered individuals at each time step.
+Once aggregated, we can plot the SIR values through a time-invariant lens using \pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
+\ref{fig:hag-tern-raw}) or with our custom \code{geom}, \code{geom_aggregate()}, which takes the raw agent data as input.
+
+```{r fig.cap = "\\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.
+We see there are two peaks of infection (the vertical axis)."}
+hagelloch_sir <- hagelloch_raw %>%
+  agents_to_aggregate(states = c(tI, tR),
+                      min_max_time = c(0, 55)) %>%
+  rename(time = t, S = X0, I = X1, R = X2)
+
+ggplot(hagelloch_sir, aes(x = S, y = I, z = R))+
+  coord_tern() +
+  geom_path() +
+  labs(x = "S", y = "I", z = "R",
+       title = "Time invariant view of Hagelloch measles outbreak") + 
+  theme_sir(base_size = 24)
+
+```
+
+In Figure \ref{fig:hag-tern-raw}, we can focus on the infections over time by analyzing the vertical axis.
+Specifically, we see two peaks of infection.
+This is interesting because the SIR model, which is sometimes used to model the spread of measles, generally has one defined peak of infection.
+We may wonder if the two peaks in the observed data may be due to random noise or if a model more complex than the simple SIR is needed to adequately capture these two peaks.
+
+Previous study tells us that measles outbreaks are often associated with children within the same grade level, and we examine if this is the case here.
+By combining the function \code{facet_wrap()} with \code{geom_aggregate()} we can easily analyze this scenario,
+
+```{r fig.cap = "\\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.
+The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined."}
+hagelloch_raw %>%
+  ggplot(aes(y = tI, z = tR, color = CL)) +
+  geom_aggregate(size = 2) + coord_tern() +
+  labs(x = "S", y = "I", z = "R",
+       color = "Class") +
+  scale_color_brewer(palette = "Dark2") +
+  facet_wrap(~CL)
+```
+
+Immediately in Fig.
+\ref{fig:tern-class-data}, we see that time invariant infection curve is different for the pre-school class compared to the 1st class.
+In the 1st class, we see about 95\% of the class become infected and less than 10\% of them having recovered, which may be indicative of a super-spreading event.
+This suspicion is further supported in that `r max(table(hagelloch_raw$IFTO[hagelloch_raw$CL == "1st class"]))` of the `r nrow(hagelloch_raw[hagelloch_raw$CL == "1st class",])` 1st class students have been reportedly infected by the same individual.
+We now have some evidence that class structure may play a role in the spread of infection.
+We can further analyze this claim with modeling and simulation.
+
+## Modeling and Simulation
+
+We first try to model the Hagelloch data with a baseline stochastic SIR model, which we refer to as the 'simple SIR.' In our full vignette \shannon{(available online),} we show how to fit this simple SIR model via maximum likelihood, a common approach used to fit parameters, and simulate from the model with those best fit parameters.
+Our function \code{simulate_agents()} (or \code{simulate_SIR_agents()})  generates individual level data according to discrete-time multinomial draws, which depend on the number of individuals in each state at the previous time step and a matrix of transition probabilities.
+For example, the below code generates 100 simulations of an outbreak of a disease with one initial infector in a population of $n= 188$ individuals, a scenario analogous to the actual outbreak.
+
+```{r}
+trans_mat <- matrix(c("X0 * (1 - X1 * par1 / N)", "X0 * X1  * par1 / N", "0",
+                  "0", "X1 * (1 - par2)", "par2 * X1",
+                  "0", "0", "X2"), byrow = TRUE, nrow = 3)
+```
+
+```{r cache = TRUE}
+set.seed(2020)
+
+best_params <- c("beta" = .36, "gamma" = .13)
+## This is the SIR representation
+
+rownames(trans_mat) <- c("S", "I", "R")
+init_vals <- c(187, 1, 0)
+par_vals <- c(par1 = best_params[1], par2 = best_params[2])
+max_T <- 55
+n_sims <- 100
+
+agents <- simulate_agents(trans_mat,
+                       init_vals,
+                       par_vals,
+                       max_T,
+                       n_sims,
+                       verbose = FALSE)
+
+```
+```{r}
+agg_model <- agents %>% group_by(sim) %>%
+  agents_to_aggregate(states = c(I, R)) %>%
+  mutate(Type = "Simple SIR")
+```
+The result of our simulation is the object \code{agents} which is a `r nrow(agents)` $\times$ `r ncol(agents)` data frame, which details the time of entry into the $S$, $I$, and $R$ states for a given simulation.
+Ultimately, we would like to know if this simple SIR model is a good fit to the data, especially in comparison to a SIR model where we can incorporate the class structure.
+
+To fit a more complex SIR model with a network structure, we use package \pkg{EpiModel} \citep{Jenness2018}.
+The below code sets up a network of individuals (which includes class as a variable) and then simulates infections over this network.
+
+```{r cache = TRUE, results = 'hide'}
+library(EpiModel)
+## WARNING:  Will take a minute or two
+
+set.seed(42)
+nw <- network.initialize(n = 188, directed = FALSE)
+nw <- set.vertex.attribute(nw, "group", rep(0:2, each = 90, 30, 68))
+formation <- ~edges + nodematch("group") + concurrent
+target.stats <- c(200, 300, 200)
+coef.diss <- dissolution_coefs(dissolution = ~offset(edges),  duration = 5)
+est1 <- netest(nw, formation, target.stats, coef.diss, edapprox = TRUE)
+
+param <- param.net(inf.prob = 0.1, act.rate = 5, rec.rate = 0.1)
+status.vector <- c(rep(0, 90), rep(0, 30), rep(0, 67), 1)
+status.vector <- ifelse(status.vector == 1, "i", "s")
+init <- init.net(status.vector = status.vector)
+control <- control.net(type = "SIR", nsteps = 55,
+                       nsims = 100, epi.by = "group")
+epimodel_sir <- netsim(est1, param, init, control)
+
+```
+
+The output of this network model is \code{epimodel_sir}, an object of class \code{`r class(epimodel_sir)`}, which contains a plethora of modeling information.
+In the following section, we will use the capabilities of \pkg{EpiCompare} to streamline the process of comparing the two models contained in the objects \code{agg_model} (the simple SIR) and \code{epimodel_sir} (the network SIR model).
+
+## Post-processing and comparison
+
+The next step is to compare the simple SIR model to the EpiModel SIR model.
+The \pkg{EpiCompare} function \code{fortify_aggregate()}, takes in an object from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transforms it into a tidy-style data frame.
+
+```{r}
+fortified_net <- fortify_aggregate(epimodel_sir, 
+                                   states = c("s.num", "i.num", "r.num")) %>%
+  mutate(Type = "EpiModel SIR",
+         sim = as.numeric(gsub("sim", "", sim)))
+```
+
+With the two modeling objects both in the same format, we can then compare the models side-by-side.
+The results are shown in Figure \ref{fig:hag-simple-sir}, where a 90\% prediction region is estimated for the two models.
+For the Simple SIR model, we see that while the prediciton region covers the data fairly well, the prediction region clearly misses the second peak of infection.
+This indicates that the simple SIR is not a good fit to our data.
+We also see that the prediction region is very large, covering up a large area of the ternary plot.
+Together, this indicates that the simple SIR model produces a biased model with a large amount of variance.
+On the other hand, for the EpiModel network model, we see that the prediction region covers the data quite well and takes up less area.
+With \pkg{EpiCompare}, we can see that the model using the class structure is a better fit to the outbreak than the simple SIR model.
+
+```{r cache = TRUE}
+
+both_models <- bind_rows(agg_model, fortified_net)
+
+g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0) %>%
+                  mutate(Type = factor(Type, levels = c("Simple SIR", 
+                                                        "EpiModel SIR"))),
+         aes(x = X0, y = X1, z = X2,
+              sim_group = sim, fill = Type),
+         alpha = .5,
+         conf_level = .90) 
+```
+
+```{r , fig.cap = "\\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models."}
+g +   geom_path(data = both_models %>% filter(t != 0) %>%
+                  mutate(Type = factor(Type, levels = c("Simple SIR", 
+                                                        "EpiModel SIR"))),
+            aes(x = X0, y = X1, z = X2, group = paste(Type, sim)),
+            alpha = .3, col = "gray40") + 
+    coord_tern() + theme_sir(base_size = 24) +
+  geom_point(data = hagelloch_sir,
+             aes(x = S, y = I, z =R), col = "black") +
+  labs(title = "Simple SIR model",
+       subtitle = "90% Prediction band and original data",
+       x = "S", y = "I", z = "R") +
+  scale_fill_manual(values = c("#006677", "#AA6600")) + 
+  facet_wrap(~Type) +
+  theme(legend.position = "bottom")
+
+```
+
+The prediction region generated by the network model covers the observed data well, but that does not mean that individual filaments generated from the network model are good fits to the observed data.
+We can further examine the model fits by incorporating these filaments into our visualization.
+In Fig.
+\ref{fig:hag-simple-sir} we show the indivdual filaments generated from the two sets of models as gray lines.
+Examining these, we see that the individual lines typically only have one defined peak, whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
+
+We can also examine these filaments quantitatively by using functions that take the distance between the filaments with the observed data.
+In the below code, we  transform the simulations to a more computationally-friendly format with the function \code{filament_compression()}.
+Following that, we calculate the distance between the  simulated filaments and the observed filament with the function \code{dist_matrix_innersq_direction()} and calculate the probability of the truth with respect to those simulations with the function \code{compare_new_to_rest_via_distance()}.
+The estimated pseudo-density of the observed epidemic (relative to the simulations from either model) is much less likely then **any** of the simulations (reported in Table \ref{tab:hags-extreme}).
+This indicates that neither of two SIR models are good fits to the data at the filament level.
+
+```{r}
+simple_sir <- both_models %>% filter(Type == "Simple SIR") %>%
+  rename(S = "X0", I = "X1", R = "X2") %>%
+  select(Type, sim, t, S, I, R)
+
+hagelloch_sir2 <- hagelloch_sir %>%
+  rename(t = "time") %>%
+  mutate(Type = "true observation",
+         sim = 0) %>%
+  select(Type, sim, t, S, I, R)
+```
+
+```{r}
+#-- after cleaning up and combining --
+all_together_df <- rbind(simple_sir,
+                         hagelloch_sir2)
+```
+
+```{r echo = F}
+all_together_df[c(1:2, nrow(all_together_df) - c(1:0)),] %>% 
+  kable(booktabs = TRUE,
+        caption = paste("Top and bottom 2 rows of \\tt{all\\_together\\_df}\\textnormal{,",
+                        "combining both simulated epidemics and the true",
+                        "epidemic.}"), label = "cif-all-together-df") %>%
+  kable_styling(position = "center", 
+                latex_options = "hold_position"
+                      )
+```
+
+```{r}
+compression_df <- all_together_df %>% group_by(Type, sim) %>% 
+  filament_compression(data_columns = c("S","I","R"), 
+                       number_points = 20)
+```
+
+```{r}
+tdmat <- compression_df %>% 
+  dist_matrix_innersq_direction(
+    position = c(1:length(compression_df))[
+      names(compression_df) %in% c("S","I", "R")],
+    tdm_out = T)
+
+simple_sir_true_obs_info <- tdmat %>% 
+  compare_new_to_rest_via_distance(
+    new_name_id = data.frame(Type = "true observation", sim = 0),
+    distance_func = distance_psuedo_density_function, 
+    sigma = "20%")
+
+```
+
+```{r echo = F}
+# EpiModel simulations:
+epimodel_sir <- both_models %>% filter(Type == "EpiModel SIR") %>%
+  rename(S = "X0", I = "X1", R = "X2") %>%
+  select(Type, sim, t, S, I, R)
+
+hagelloch_sir2 <- hagelloch_sir %>%
+  rename(t = "time") %>%
+  mutate(Type = "true observation",
+         sim = 0) %>%
+  select(Type, sim, t, S, I, R)
+
+all_together_df <- rbind(epimodel_sir,
+                         hagelloch_sir2)
+
+compression_df <- all_together_df %>% group_by(Type, sim) %>% 
+  filament_compression(data_columns = c("S","I","R"), 
+                       number_points = 20)
+
+tdmat <- compression_df %>% 
+  dist_matrix_innersq_direction(
+    position = c(1:length(compression_df))[
+      names(compression_df) %in% c("S","I", "R")],
+    tdm_out = T)
+
+epimodel_sir_true_obs_info <- tdmat %>% 
+  compare_new_to_rest_via_distance(
+    new_name_id = data.frame(Type = "true observation", sim = 0),
+    distance_func = distance_psuedo_density_function, 
+    sigma = "20%") 
+```
+
+```{r hagelloch-extremeness2, echo = FALSE}
+simple_sir_info <- simple_sir_true_obs_info %>%
+  select(-sim) %>%
+  mutate(Type = "Simple SIR")
+eimodel_sir_info <- epimodel_sir_true_obs_info %>%
+  select(-sim) %>%
+  mutate(Type = "EpiModel SIR")
+
+rbind(simple_sir_info, eimodel_sir_info)  %>%
+  kable(format = "latex", booktabs = TRUE, 
+        col.names = 
+          linebreak(c("Type",
+                      "simulations-based estimated pseudo-density",
+              "proportion of simulations with lower estimated pseudo-density"),
+              align = c("l")),
+        caption = paste("The extremeness of the true simulations based on",
+                        "comparing pseudo-density estimates between true",
+                        "vs simulated curves"), label = "hags-extreme") %>%
+  kable_styling(position = "center", latex_options = "hold_position") %>%
+  column_spec(2:3, width = "6cm")
+```
+
+In conclusion, \pkg{EpiCompare} allows us to fully examine this outbreak at every step in the data analysis pipeline (see Fig.
+\ref{fig:pipeline2}) in a streamlined fashion.
+With EDA, we saw evidence that class structure may be important in the spread of of measles.
+We then compared a baseline simple SIR model to a more complicated SIR model which incorporated a network structure which included the class structure.
+Based on the prediction regions generated from these models, we saw that the network model fit the data much better than the simple SIR model.
+However, when we examined the individual filaments generated by the network model, we found that the data is highly unlikely to be generated from such a model.
+For further analysis, we would recommend looking into models that can more accurately capture super-spreading events based on the observation that one child was allegedly responsible for nearly all of his classmates' infections.
+Overall,  this analysis demonstrates how \pkg{EpiCompare} aids in the data analysis pipeline for both novice and expert practitioners and coders alike.
+
+]]
+
+\shannon{[[NEWER TEXT}
+
+\footnote{\ben{[Ben says: Generally I found this whole section to be pretty passive and not well motivated on why we'd actually do the analysis.]}}
+To conclude our paper, we demonstrate the capabilities of \pkg{EpiCompare} with a complete data analysis of a measles outbreak in 1861-1862 Germany.
+Specifically, we demonstrate how tools in \pkg{EpiCompare} can be used in each step of the data analysis pipeline \footnote{\ben{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]} \shannon{check out new re-write.
+new goal is to motivate question of hagelloch and then show how epicompare helps}} \ben{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\ben{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that?
+/ what does it add to the conv/ why should they reference it?]}}.
+Additionally, we highlight how time-invariant analysis (see Section \ref{sec:time-invariant}) can be used to enhance understanding of an outbreak\footnote{\ben{[Ben says: I'm not sure this is necessary to state.
+Additionally - is it really true do we highlight this?]}}.
+
+\footnote{\ben{[Ben says:The data introduction could be another subsection.
+Why does it fit better here?.
+Other comment: This paragraph is pretty rambly.
+It is unclear what you want someone to take away from it.
+I might suggest highlight the fact that the data is "an ideal testing group for methodology".]}} Before demonstrating \pkg{EpiCompare}, we provide some context for the measles outbreak presented here.\footnote{\ben{[Ben says: this first sentence is pretty sign-post-y yet it only relates to the next few sentences.
+Update so it's less sign-post-y.]}}  The data was originally organized by \cite{pfeilsticker1863}, later made visible by \cite{oesterle1992}, and made available in an \proglang{R} by \cite{surveillance2017}.
+This data set includes a rich collection of features including household location, class level, and allged infector ID, and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}\footnote{\ben{[Ben says: This citation doesn't make sense - should it be "citep"?]}}.
+In this data set, there are 188 children who became infected with the measles over the course approximately 90 days.
+
+\shannon{]]}
+
+\shannon{[[LESS NEW TEXT}
+
+Finally, in this section we show how tools from \pkg{EpiCompare} can be used in each step of the data analysis pipeline shown in Fig.
+\ref{fig:pipeline}.
+We analyze an outbreak of measles in the town of Hagelloch, Germany from 1861-1862, a data set organized by \cite{pfeilsticker1863}.
+The data was later made visible by \cite{oesterle1992} and made available in an \proglang{R} by \cite{surveillance2017}.
+This data set includes a rich collection of features and  is an ideal testing ground for methodology in infectious disease epidemiology \cite{Neal2004,britton2011,groendyke2012,becker2016}.\footnote{\shannon{The old first paragraph from data and exploratory analysis paragraph was combined with the intro as a better lead-in to what's going on.}}
+
+\shannon{]]}
+
+[[\textcolor{red}{OLD TEXT}
+
+\footnote{\ben{[Ben says: Shannon, would you mind reading this whole section over again once we've finished edits for section 2 and 3?
+This initial paragraph seems to be stating section 3's story.]}}In this section, we highlight many of the tools available in \pkg{EpiCompare}.
+As previously discussed, these tools include data cleaning; visualization; modeling and simulation; post-processing; and comparison and model assessment, in accordance with the data analysis pipeline (Fig.
+\ref{fig:pipeline}).
+We show a full data analysis from beginning to end that can be accomplished in a streamlined and standardized manner via \pkg{EpiCompare}.
+]]
+
+\subsection{Pre-processing and EDA}
 
@@
 We provide\footnote{\ben{[Ben says: the phrase "We provide" is very passive / distance from the current demonstration at hand.
-Moreover section 3 already phrases things this way.]}} the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transform it into a tidy-style data frame.
+Moreover section 3 already phrases things this way.]}\shannon{more of sentence 2 of ben than 1.}} the function \code{fortify_aggregate()}, which can take objects from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transform it into a tidy-style data frame.
 

### 2021-06-07 jss/EpiCompare-paper/EpiCompare-paper.tex
major section 4 edits
also cleaned up end of section 3

@@
 
-\textbf{Comparisons and assessment} \textcolor{orange}{[[ NEW TEXT}
-
-\textcolor{violet}{\sout{Finally,} Often}\footnote{\textcolor{violet}{[Ben says: I feel like starting with "finally" causes a weird pause - but maybe that's just me.]}}
-the data analysis pipeline \textcolor{violet}{\sout{often}} ends with
-plots, tables, and summary statistics that are used to
-\sout{\textcolor{violet}{compare} \textcolor{orange}{and assess}}
-\textcolor{violet}{\sout{different models, and simulations to one another} assess model preformance and compare across models or simulations.}
-In \pkg{EpiCompare} we provide a set of comparison and assessment tools
-for model and simulation results that extend beyond the standard
-performance metrics (e.g.~mean squared error or AIC) and into the lens
-of time-invariant analysis. We have found \textcolor{violet}{that} these
-tools \textcolor{violet}{\sout{to be} are} specifically applicable for
-situations where only one season or cycle of an epidemic has occurred,
-\textcolor{violet}{or is the object of interest}.
-
-\footnote{\textcolor{violet}{[Ben says: This changed introduction was done to highlight that we can use prediction regions for mutiple types of comparisons.]}}\textcolor{violet}{The first set of of tools surround the creation of prediction regions.
-We can create a prediction regions from model simulations to examine if our model simulations capture the true epidemic trajectory.
-We do so in a time-invariant way and utilizing filamental representations of the model simulations and the true epidemic.}
-{[}For three-state epidemic models, we provide the
-\texttt{ggplot}/\texttt{ggtern} extension
+\textbf{Comparisons and assessment}
+
+The last step of the data analysis pipeline often ends with plots,
+tables, and summary statistics that are used to assess model performance
+and compare across models or simulations. In \pkg{EpiCompare} we provide
+a set of comparison and assessment tools for model and simulation
+results that extend beyond the standard performance metrics (e.g.~mean
+squared error or AIC) and into the lens of time-invariant analysis. We
+have found that these tools are specifically applicable for situations
+where only one season or cycle of an epidemic has occurred or is the
+object of interest.
+
+The first set of of tools surround the creation of prediction regions.
+We can create a prediction regions from model simulations to examine if
+our model simulations capture the true epidemic trajectory. We do so in
+a time-invariant way and utilizing filamental representations of the
+model simulations and the true epidemic. For three-state epidemic
+models, we provide the \texttt{ggplot}/\texttt{ggtern} extension
 \texttt{geom\_prediction\_band()} which creates a prediction region
 around the top \(1-\alpha\) proportion of the simulations. In this
-visual setting, comparing this prediction \textcolor{violet}{region} to
-the true epidemic trajectory
-\textcolor{violet}{\sout{or comparing the prediction regions defined by two different models' simulations}}\footnote{\textcolor{violet}{[Ben says: this part will be highlighted in later sentences.]}}
-can be done by eye. In \pkg{EpiCompare}, we also provide these
-prediction regions for epidemic models with with more than three states.
-The functions \texttt{create\_convex\_hull\_structure()} and
+visual setting, comparing this prediction region to the true epidemic
+trajectory can be done by eye. In \pkg{EpiCompare}, we also provide
+these prediction regions for epidemic models with with more than three
+states. The functions \texttt{create\_convex\_hull\_structure()} and
 \texttt{create\_delta\_ball\_structure()} create different geometric
@@
 model. For both of these geometric structures, we provide functions to
-check if a path is contained (\texttt{contained()}){]}.
-\textcolor{violet}{We can also use these prediction regions to visually or mathematically compare how similar two sets of simulations are. In \pkg{EpiCompare} we provide the}
-\texttt{hausdorff\_dist()}
-\textcolor{violet}{function to calculate the Hausdorff distance between multiple prediction regions, when visual comparison is not possible.}
-
-{[}We also provide functions to calculate the ``extremeness'' of a true
+check if a path is contained (\texttt{contained()}){]}. We can also use
+these prediction regions to visually or mathematically compare how
+similar two sets of simulations are. In \pkg{EpiCompare} we provide the
+\texttt{hausdorff\_dist()} function to calculate the Hausdorff distance
+between multiple prediction regions, when visual comparison is not
+possible.
+
+We also provide functions to calculate the ``extremeness'' of a true
 epidemic trajectory compared to simulated epidemics via the
 equi-distance filamental trajectory representation as mentioned in
-Section \ref{sec:beyond-r0-sir}.{]}
-\textcolor{violet}{We provide implementations of a few distance-based score functions that capture how "reasonable" an epidemic is relative to other epidemics, and these scores can be turned into an extremeness measure with}
+Section \ref{sec:beyond-r0-sir}.{]} We provide implementations of a few
+distance-based score functions that capture how ``reasonable'' an
+epidemic is relative to other epidemics, and these scores can be turned
+into an extremeness measure with
 \texttt{mean(sim\_scores\ \textgreater{}\ truth\_score)}.
@@
 \texttt{local\_distance\_depth\_function()} provide depth scores that
-suggest how geometrically central an epidemic is to simulations.{]}
+suggest how geometrically central an epidemic is to simulations.
+
+\section[Tour]{A tour of \pkg{EpiCompare}}\label{sec:tour}
+
+\textcolor{orange}{I have substantially overhauled section 4.
+The older versions are kept after the entirety of this section but note that much of the text (but not the structure) of this section is different.
+I've tried to better motivate why we use the epicompare functions to answer our questions of interest.}
+
+\textcolor{orange}{[[NEWEST TEXT}
+
+To conclude our paper, we demonstrate how \pkg{EpiCompare} can
+streamline the data analysis process with a case study of a measles
+outbreak in 1861-1862 Germany. With the help of \pkg{EpiCompare}, we can
+answer questions such as is a SIR model a good fit for the data, does
+the outbreak spread differently within the different school classes of
+the children, and can incorporating an underlying network structure
+enhance model fit. We first begin with some context for the epidemic.
+
+\hypertarget{background-for-1861-1862-measles-epidemic}{%
+\subsection{Background for 1861-1862 measles
+epidemic}\label{background-for-1861-1862-measles-epidemic}}
+
+We analyze an outbreak of measles in the town of Hagelloch, Germany from
+1861-1862, a data set organized by \cite{pfeilsticker1863}. The data was
+later made visible by \cite{oesterle1992} and made available in an
+\proglang{R} by \cite{surveillance2017}. In this outbreak, 188 children
+were infected with measles over the course of three months. This data
+set includes a rich collection of features including day of measles
+rash, age, sex, school class, household and household location, and
+alleged infector of each child. We show a subset of the data in Table
+\ref{tab:hags-people}. We are particularly interested if and how group
+structure plays a role in the spread of measles.
+
+\begin{CodeChunk}
+\begin{table}[!h]
+
+\caption{\label{tab:hags-people}Subset of Hagelloch infection data.
+Features include the person ID, household ID (HH ID), age, sex, class level (Pre-K/1st/2nd), date of first symptoms, date of the appearance of the measles rash, and the alleged infector ID of the individual.}
+\centering
+\begin{tabular}[t]{rrlrllllr}
+\toprule
+ID & HH ID & Name & Age & Sex & Class & Symp. Start & Rash Date & Infector ID\\
+\midrule
+1 & 61 & Mueller & 7 & female & 1st class & 1861-11-21 & 1861-11-25 & 45\\
+2 & 61 & Mueller & 6 & female & 1st class & 1861-11-23 & 1861-11-27 & 45\\
+3 & 61 & Mueller & 4 & female & preschool & 1861-11-28 & 1861-12-02 & 172\\
+4 & 62 & Seibold & 13 & male & 2nd class & 1861-11-27 & 1861-11-28 & 180\\
+5 & 63 & Motzer & 8 & female & 1st class & 1861-11-22 & 1861-11-27 & 45\\
+45 & 51 & Goehring & 7 & male & 1st class & 1861-11-11 & 1861-11-13 & 184\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
+\hypertarget{pre-processing-and-eda}{%
+\subsection{Pre-processing and EDA}\label{pre-processing-and-eda}}
+
+We begin our analysis by first examining and transforming the raw data
+(\code{hagelloch_raw}), which is individual-level data. With
+\pkg{EpiCompare}, we can quickly determine how the epidemic progresses
+through the population at an aggregate level. By specifying the time of
+infection (\code{tI}) and the time of recovery (\code{tR}), the function
+\code{agents_to_aggregate()} calculates the number susceptible,
+infectious, and recovered individuals at each time step. Once
+aggregated, we can plot the SIR values through a time-invariant lens
+using \pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
+\ref{fig:hag-tern-raw}) or with our custom \code{geom},
+\code{geom_aggregate()}, which takes the raw agent data as input.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> hagelloch_sir <- hagelloch_raw %>%
++   agents_to_aggregate(states = c(tI, tR),
++                       min_max_time = c(0, 55)) %>%
++   rename(time = t, S = X0, I = X1, R = X2)
+R> 
+R> 
+R> ggplot(hagelloch_sir, aes(x = S, y = I, z = R))+
++   coord_tern() +
++   geom_path() +
++   labs(x = "S", y = "I", z = "R",
++        title = "Time invariant view of Hagelloch measles outbreak") + 
++   theme_sir(base_size = 24)
+\end{CodeInput}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-4-1}
+
+}
+
+\caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.
+We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-4}
+\end{figure}
+\end{CodeChunk}
+
+In Figure \ref{fig:hag-tern-raw}, we can focus on the infections over
+time by analyzing the vertical axis. Specifically, we see two peaks of
+infection. This is interesting because the SIR model, which is sometimes
+used to model the spread of measles, generally has one defined peak of
+infection. We may wonder if the two peaks in the observed data may be
+due to random noise or if a model more complex than the simple SIR is
+needed to adequately capture these two peaks.
+
+Previous study tells us that measles outbreaks are often associated with
+children within the same grade level, and we examine if this is the case
+here. By combining the function \code{facet_wrap()} with
+\code{geom_aggregate()} we can easily analyze this scenario,
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> hagelloch_raw %>%
++   ggplot(aes(y = tI, z = tR, color = CL)) +
++   geom_aggregate(size = 2) + coord_tern() +
++   labs(x = "S", y = "I", z = "R",
++        color = "Class") +
++   scale_color_brewer(palette = "Dark2") +
++   facet_wrap(~CL)
+\end{CodeInput}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-5-1}
+
+}
+
+\caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.
+The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-5}
+\end{figure}
+\end{CodeChunk}
+
+Immediately in Fig. \ref{fig:tern-class-data}, we see that time
+invariant infection curve is different for the pre-school class compared
+to the 1st class. In the 1st class, we see about 95\% of the class
+become infected and less than 10\% of them having recovered, which may
+be indicative of a super-spreading event. This suspicion is further
+supported in that 26 of the 30 1st class students have been reportedly
+infected by the same individual. We now have some evidence that class
+structure may play a role in the spread of infection. We can further
+analyze this claim with modeling and simulation.
+
+\hypertarget{modeling-and-simulation}{%
+\subsection{Modeling and Simulation}\label{modeling-and-simulation}}
+
+We first try to model the Hagelloch data with a baseline stochastic SIR
+model, which we refer to as the `simple SIR.' In our full vignette
+\textcolor{orange}{(available online),} we show how to fit this simple
+SIR model via maximum likelihood, a common approach used to fit
+parameters, and simulate from the model with those best fit parameters.
+Our function \code{simulate_agents()} (or \code{simulate_SIR_agents()})
+generates individual level data according to discrete-time multinomial
+draws, which depend on the number of individuals in each state at the
+previous time step and a matrix of transition probabilities. For
+example, the below code generates 100 simulations of an outbreak of a
+disease with one initial infector in a population of \(n= 188\)
+individuals, a scenario analogous to the actual outbreak.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> trans_mat <- matrix(c("X0 * (1 - X1 * par1 / N)", "X0 * X1  * par1 / N", "0",
++                   "0", "X1 * (1 - par2)", "par2 * X1",
++                   "0", "0", "X2"), byrow = TRUE, nrow = 3)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> set.seed(2020)
+R> 
+R> best_params <- c("beta" = .36, "gamma" = .13)
+R> ## This is the SIR representation
+R> 
+R> rownames(trans_mat) <- c("S", "I", "R")
+R> init_vals <- c(187, 1, 0)
+R> par_vals <- c(par1 = best_params[1], par2 = best_params[2])
+R> max_T <- 55
+R> n_sims <- 100
+R> 
+R> agents <- simulate_agents(trans_mat,
++                        init_vals,
++                        par_vals,
++                        max_T,
++                        n_sims,
++                        verbose = FALSE)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> agg_model <- agents %>% group_by(sim) %>%
++   agents_to_aggregate(states = c(I, R)) %>%
++   mutate(Type = "Simple SIR")
+\end{CodeInput}
+\end{CodeChunk}
+
+The result of our simulation is the object \code{agents} which is a
+18800 \(\times\) 5 data frame, which details the time of entry into the
+\(S\), \(I\), and \(R\) states for a given simulation. Ultimately, we
+would like to know if this simple SIR model is a good fit to the data,
+especially in comparison to a SIR model where we can incorporate the
+class structure.
+
+To fit a more complex SIR model with a network structure, we use package
+\pkg{EpiModel} \citep{Jenness2018}. The below code sets up a network of
+individuals (which includes class as a variable) and then simulates
+infections over this network.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> library(EpiModel)
+R> ## WARNING:  Will take a minute or two
+R> 
+R> set.seed(42)
+R> nw <- network.initialize(n = 188, directed = FALSE)
+R> nw <- set.vertex.attribute(nw, "group", rep(0:2, each = 90, 30, 68))
+R> formation <- ~edges + nodematch("group") + concurrent
+R> target.stats <- c(200, 300, 200)
+R> coef.diss <- dissolution_coefs(dissolution = ~offset(edges),  duration = 5)
+R> est1 <- netest(nw, formation, target.stats, coef.diss, edapprox = TRUE)
+R> 
+R> param <- param.net(inf.prob = 0.1, act.rate = 5, rec.rate = 0.1)
+R> status.vector <- c(rep(0, 90), rep(0, 30), rep(0, 67), 1)
+R> status.vector <- ifelse(status.vector == 1, "i", "s")
+R> init <- init.net(status.vector = status.vector)
+R> control <- control.net(type = "SIR", nsteps = 55,
++                        nsims = 100, epi.by = "group")
+R> epimodel_sir <- netsim(est1, param, init, control)
+\end{CodeInput}
+\end{CodeChunk}
+
+The output of this network model is \code{epimodel_sir}, an object of
+class \code{netsim}, which contains a plethora of modeling information.
+In the following section, we will use the capabilities of
+\pkg{EpiCompare} to streamline the process of comparing the two models
+contained in the objects \code{agg_model} (the simple SIR) and
+\code{epimodel_sir} (the network SIR model).
+
+\hypertarget{post-processing-and-comparison}{%
+\subsection{Post-processing and
+comparison}\label{post-processing-and-comparison}}
+
+The next step is to compare the simple SIR model to the EpiModel SIR
+model. The \pkg{EpiCompare} function \code{fortify_aggregate()}, takes
+in an object from specialized classes of modeling output
+\textcolor{orange}{(like those made by \code{netsim()})} and transforms
+it into a tidy-style data frame.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> fortified_net <- fortify_aggregate(epimodel_sir, 
++                                    states = c("s.num", "i.num", "r.num")) %>%
++   mutate(Type = "EpiModel SIR",
++          sim = as.numeric(gsub("sim", "", sim)))
+\end{CodeInput}
+\end{CodeChunk}
+
+With the two modeling objects both in the same format, we can then
+compare the models side-by-side. The results are shown in Figure
+\ref{fig:hag-simple-sir}, where a 90\% prediction region is estimated
+for the two models. For the Simple SIR model, we see that while the
+prediciton region covers the data fairly well, the prediction region
+clearly misses the second peak of infection. This indicates that the
+simple SIR is not a good fit to our data. We also see that the
+prediction region is very large, covering up a large area of the ternary
+plot. Together, this indicates that the simple SIR model produces a
+biased model with a large amount of variance. On the other hand, for the
+EpiModel network model, we see that the prediction region covers the
+data quite well and takes up less area. With \pkg{EpiCompare}, we can
+see that the model using the class structure is a better fit to the
+outbreak than the simple SIR model.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> both_models <- bind_rows(agg_model, fortified_net)
+R> 
+R> 
+R> g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0) %>%
++                   mutate(Type = factor(Type, levels = c("Simple SIR", 
++                                                         "EpiModel SIR"))),
++          aes(x = X0, y = X1, z = X2,
++               sim_group = sim, fill = Type),
++          alpha = .5,
++          conf_level = .90) 
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> g +   geom_path(data = both_models %>% filter(t != 0) %>%
++                   mutate(Type = factor(Type, levels = c("Simple SIR", 
++                                                         "EpiModel SIR"))),
++             aes(x = X0, y = X1, z = X2, group = paste(Type, sim)),
++             alpha = .3, col = "gray40") + 
++     coord_tern() + theme_sir(base_size = 24) +
++   geom_point(data = hagelloch_sir,
++              aes(x = S, y = I, z =R), col = "black") +
++   labs(title = "Simple SIR model",
++        subtitle = "90% Prediction band and original data",
++        x = "S", y = "I", z = "R") +
++   scale_fill_manual(values = c("#006677", "#AA6600")) + 
++   facet_wrap(~Type) +
++   theme(legend.position = "bottom")
+\end{CodeInput}
+\begin{figure}[H]
+
+{\centering \includegraphics{Figs/unnamed-chunk-12-1}
+
+}
+
+\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-12}
+\end{figure}
+\end{CodeChunk}
+
+The prediction region generated by the network model covers the observed
+data well, but that does not mean that individual filaments generated
+from the network model are good fits to the observed data. We can
+further examine the model fits by incorporating these filaments into our
+visualization. In Fig. \ref{fig:hag-simple-sir} we show the indivdual
+filaments generated from the two sets of models as gray lines. Examining
+these, we see that the individual lines typically only have one defined
+peak, whereas the data certainly looks like it has two distinct peaks,
+likely caused by our assumed super-spreader event.
+
+We can also examine these filaments quantitatively by using functions
+that take the distance between the filaments with the observed data. In
+the below code, we transform the simulations to a more
+computationally-friendly format with the function
+\code{filament_compression()}. Following that, we calculate the distance
+between the simulated filaments and the observed filament with the
+function \code{dist_matrix_innersq_direction()} and calculate the
+probability of the truth with respect to those simulations with the
+function \code{compare_new_to_rest_via_distance()}. The estimated
+pseudo-density of the observed epidemic (relative to the simulations
+from either model) is much less likely then \textbf{any} of the
+simulations (reported in Table \ref{tab:hags-extreme}). This indicates
+that neither of two SIR models are good fits to the data at the filament
+level.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> simple_sir <- both_models %>% filter(Type == "Simple SIR") %>%
++   rename(S = "X0", I = "X1", R = "X2") %>%
++   select(Type, sim, t, S, I, R)
+R> 
+R> hagelloch_sir2 <- hagelloch_sir %>%
++   rename(t = "time") %>%
++   mutate(Type = "true observation",
++          sim = 0) %>%
++   select(Type, sim, t, S, I, R)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> #-- after cleaning up and combining --
+R> all_together_df <- rbind(simple_sir,
++                          hagelloch_sir2)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}[!h]
+
+\caption{\label{tab:cif-all-together-df}Top and bottom 2 rows of \tt{all\_together\_df}\textnormal{, combining both simulated epidemics and the true epidemic.}}
+\centering
+\begin{tabular}[t]{lrrrrr}
+\toprule
+Type & sim & t & S & I & R\\
+\midrule
+Simple SIR & 1 & 0 & 188 & 0 & 0\\
+Simple SIR & 1 & 1 & 187 & 1 & 0\\
+true observation & 0 & 54 & 1 & 0 & 187\\
+true observation & 0 & 55 & 1 & 0 & 187\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> compression_df <- all_together_df %>% group_by(Type, sim) %>% 
++   filament_compression(data_columns = c("S","I","R"), 
++                        number_points = 20)
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{CodeInput}
+R> tdmat <- compression_df %>% 
++   dist_matrix_innersq_direction(
++     position = c(1:length(compression_df))[
++       names(compression_df) %in% c("S","I", "R")],
++     tdm_out = T)
+R> 
+R> simple_sir_true_obs_info <- tdmat %>% 
++   compare_new_to_rest_via_distance(
++     new_name_id = data.frame(Type = "true observation", sim = 0),
++     distance_func = distance_psuedo_density_function, 
++     sigma = "20%") 
+\end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
+\begin{table}[!h]
+
+\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing pseudo-density estimates between true vs simulated curves}
+\centering
+\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
+\toprule
+Type & simulations-based estimated pseudo-density & proportion of simulations with lower estimated pseudo-density\\
+\midrule
+Simple SIR & 0.0036733 & 0.00\\
+EpiModel SIR & 0.0118283 & 0.03\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\end{CodeChunk}
+
+In conclusion, \pkg{EpiCompare} allows us to fully examine this outbreak
+at every step in the data analysis pipeline (see Fig.
+\ref{fig:pipeline2}) in a streamlined fashion. With EDA, we saw evidence
+that class structure may be important in the spread of of measles. We
+then compared a baseline simple SIR model to a more complicated SIR
+model which incorporated a network structure which included the class
+structure. Based on the prediction regions generated from these models,
+we saw that the network model fit the data much better than the simple
+SIR model. However, when we examined the individual filaments generated
+by the network model, we found that the data is highly unlikely to be
+generated from such a model. For further analysis, we would recommend
+looking into models that can more accurately capture super-spreading
+events based on the observation that one child was allegedly responsible
+for nearly all of his classmates' infections. Overall, this analysis
+demonstrates how \pkg{EpiCompare} aids in the data analysis pipeline for
+both novice and expert practitioners and coders alike.
 
 {]}{]}
-
-{[}{[}\textcolor{red}{ OLD TEXT}
-
-One tool we provide to assess models is through the creation of
-geometric prediction regions, which are useful when we treat epidemics
-like filaments. If there is a set of simulated epidemics from a model,
-we can create a geometric prediction region for the expected trajectory
-of the epidemic in the state space. For three-state epidemic models, we
-provide the \texttt{ggplot}/\texttt{ggtern} extension
-\texttt{geom\_prediction\_band()} which creates a prediction region
-around the top \(1-\alpha\) proportion of the simulations. In this
-visual setting, comparing this prediction to the true epidemic
-trajectory or comparing the prediction regions defined by two different
-models' simulations can be done by eye. In \pkg{EpiCompare}, we also
-provide these prediction regions for epidemic models with with more than
-three states. The functions \texttt{create\_convex\_hull\_structure()}
-and \texttt{create\_delta\_ball\_structure()} create different geometric
-representations of prediction regions for any dimensional state-based
-model. For both of these geometric structures, we provide functions to
-check if a path is contained (\texttt{contained()}) and the ability to
-assess the Hausdorff distance between prediction regions based on
-simulations from different model (\texttt{hausdorff\_dist()}). The
-vignettes for \texttt{EpiCompare} go into much more mathematical detail
-about how these functions can be used in conjunction with time-invariant
-analysis.
-
-We also provide functions to calculate the ``extremeness'' of a true
-epidemic trajectory compared to simulated epidemics via the
-equi-distance filamental trajectory representation as mentioned in
-Section \ref{sec:beyond-r0-sir}. Specifically, functions like
-\texttt{distance\_pseudo\_density\_function()} can calculate a
-pseudo-density estimate of the true epidemic relative to simulated ones.
-Functions \texttt{distance\_depth\_function()} and
-\texttt{local\_distance\_depth\_function()} provide depth scores that
-suggest how geometrically central an epidemic is to simulations.
-
-\textcolor{orange}{]]}
-
-{[}{[}\textcolor{red}{Really OLD TEXT}
-
-Finally, in \pkg{EpiCompare} we provide a set of comparison and
-assessment tools for model and simulation results that extend beyond the
-standard performance metrics (e.g.~mean squared error or AIC).
-Corresponding\footnote{\textcolor{violet}{[Ben says: "corresponding to" doesn't seem to be the correct phrase here.
-If we wanted to use "corresponding" I might imagine saying "corresponding to the araguments for time-invariant analysis in Section 2.2, EpiCompare provides such tools." I'm open to keeping it if you think it's the perfect word choice.]}\textcolor{orange}{"To support the ideas presented in Sec."?}}
-to Section \ref{sec:beyond-r0-sir}, \pkg{EpiCompare} provides a set of
-time-invariant tools to compare and evaluate epidemic models and
-simulations. We have found these tools to be specifically applicable for
-situations where only one ``season'' or
-``cycle''\footnote{\textcolor{violet}{[Ben says: I think we can remove the paratheses here.]} \textcolor{orange}{accepted}}
-of the epidemic has occurred.
-
-\footnote{\textcolor{violet}{[Ben says: multiple parts of this paragraph might benefit with capturing a bit more of the motivation in the rewrite of Section 2 for the tools we're presenting.]}\textcolor{orange}{motivate better}}One
-tool we provide to assess models is through the creation of geometric
-prediction regions, which are useful when we treat epidemics like
-filaments. If there is a set of simulated epidemics from a model, we can
-create a geometric prediction region for the expected trajectory of the
-epidemic in the state space. For three-state epidemic models, we provide
-the \texttt{ggplot}/\texttt{ggtern} extension
-\texttt{geom\_prediction\_band()} which creates a prediction region
-around the top \(1-\alpha\) proportion of the simulations. In this
-visual setting, comparing this prediction to the true epidemic
-trajectory or comparing the prediction regions defined by two different
-models' simulations can be done
-\textcolor{violet}{by eye.\sout{with the 'eye test.'}\textcolor{orange}{I'm not saying 'no' but I'd do it in an appendix and not the main text.  I don't think equations belong in Section 3.}}
-In \pkg{EpiCompare} we also provide these prediction regions for
-epidemic models with with more than three states. The functions
-\texttt{create\_convex\_hull\_structure()} and
-\texttt{create\_delta\_ball\_structure()} create different geometric
-representations of prediction regions for any dimensional state-based
-model. For both of these geometric structures, we provide functions to
-check if a path is contained (\texttt{contained()}) and the ability to
-assess the Hausdorff distance between prediction regions based on
-simulations from different model (\texttt{hausdorff\_dist()}).
-
-\footnote{\textcolor{violet}{[Ben says: random thought (not needed) - maybe we do provide some equations for these measures and also maybe mathematically express how we'd examine extremeness?]}}We
-also provide functions to calculate the ``extremeness'' of a true
-epidemic trajectory compared to simulated epidemics via the
-equi-distance filamental trajectory representation as mentioned in
-Section \ref{sec:beyond-r0-sir}. Specifically, functions like
-\texttt{distance\_pseudo\_density\_function()} can calculate a
-pseudo-density estimate of the true epidemic relative to simulated ones.
-Functions \texttt{distance\_depth\_function()} and
-\texttt{local\_distance\_depth\_function()} provide depth scores that
-suggest how geometrically central an epidemic is to simulations. {]}{]}
-
-\section[Tour]{A tour of \pkg{EpiCompare}}\label{sec:tour}
 
@@
 \pkg{EpiCompare} can be used in each step of the data analysis pipeline
-\footnote{\textcolor{violet}{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]}}
+\footnote{\textcolor{violet}{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]} \textcolor{orange}{check out new re-write.
+new goal is to motivate question of hagelloch and then show how epicompare helps}}
 \textcolor{violet}{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\textcolor{violet}{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that?
@@
 \subsection{Pre-processing and EDA}
-
-\begin{CodeChunk}
-\begin{table}[!h]
-
-\caption{\label{tab:hags-people}Subset of Hagelloch infection data.
-Features include the person ID, household ID (HH ID), age, sex, class level (Pre-K/1st/2nd), date of first symptoms, date of the appearance of the measles rash, and the alleged infector ID of the individual.}
-\centering
-\begin{tabular}[t]{rrlrllllr}
-\toprule
-ID & HH ID & Name & Age & Sex & Class & Symp. Start & Rash Date & Infector ID\\
-\midrule
-1 & 61 & Mueller & 7 & female & 1st class & 1861-11-21 & 1861-11-25 & 45\\
-2 & 61 & Mueller & 6 & female & 1st class & 1861-11-23 & 1861-11-27 & 45\\
-3 & 61 & Mueller & 4 & female & preschool & 1861-11-28 & 1861-12-02 & 172\\
-4 & 62 & Seibold & 13 & male & 2nd class & 1861-11-27 & 1861-11-28 & 180\\
-5 & 63 & Motzer & 8 & female & 1st class & 1861-11-22 & 1861-11-27 & 45\\
-45 & 51 & Goehring & 7 & male & 1st class & 1861-11-11 & 1861-11-13 & 184\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
 
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-7-1}
+{\centering \includegraphics{Figs/unnamed-chunk-22-1}
 
@@
 \caption{\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.
-We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-7}
+We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-22}
 \end{figure}
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-9-1}
+{\centering \includegraphics{Figs/unnamed-chunk-24-1}
 
@@
 \caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.
-We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-9}
+We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-24}
 \end{figure}
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-10-1}
+{\centering \includegraphics{Figs/unnamed-chunk-25-1}
 
@@
 \caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.
-The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-10}
+The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-25}
 \end{figure}
@@
 
-\hypertarget{modeling-and-simulation}{%
-\subsection{Modeling and simulation}\label{modeling-and-simulation}}
+\hypertarget{modeling-and-simulation-1}{%
+\subsection{Modeling and simulation}\label{modeling-and-simulation-1}}
 
@@
 
-\hypertarget{post-processing-and-comparison}{%
+\hypertarget{post-processing-and-comparison-1}{%
 \subsection{Post-processing and
-comparison}\label{post-processing-and-comparison}}
+comparison}\label{post-processing-and-comparison-1}}
 
@@
 model. We
-provide\footnote{\textcolor{violet}{[Ben says: the phrase "We provide" is very passive / distance from the current demonstration at hand. Moreover section 3 already phrases things this way.]}}
+provide\footnote{\textcolor{violet}{[Ben says: the phrase "We provide" is very passive / distance from the current demonstration at hand.
+Moreover section 3 already phrases things this way.]}\textcolor{orange}{more of sentence 2 of ben than 1.}}
 the function \code{fortify_aggregate()}, which can take objects from
@@
 
-{\centering \includegraphics{Figs/unnamed-chunk-17-1}
+{\centering \includegraphics{Figs/unnamed-chunk-32-1}
 
@@
 
-\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-17}
+\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-32}
 \end{figure}

### 2021-06-11 jss/EpiCompare-paper/EpiCompare-paper.Rmd
More section 4 edits

@@
 
+UPDATE 6/11/2021.  I am now adding overall goals of section and each subheading.  Additionally there are a few updates from last draft (unmarked since you didn't comment on last draft).
+
+SECTION GOALS:
+
+The main goal of this section is to show how a person can use EpiCompare in a real data analysis.
+With that said my primary focus is on EpiCompare and not the data analysis -- a point I think you may be having an issue with.
+I've given this some thought and believe that the data analysis is a means for an end to us -- showing off EpiCompare.
+That said again, I've made some changes in the first paragraph below.
+
+To accomplish this main goal, I shadow the data analysis pipeline we have introduced and show how epicompare can be used in the steps.
+
+The secondary goals are figuring out useful information about this measles outbreak.
+
 \shannon{[[NEWEST TEXT}
-To conclude our paper, we demonstrate how \pkg{EpiCompare}  can streamline the data analysis process with a case study of a measles outbreak in 1861-1862 Germany.
-With the help of \pkg{EpiCompare}, we can answer questions such as is a SIR model a good fit for the data, does the outbreak spread differently within the different school classes of the children, and can incorporating an underlying network structure enhance model fit.
-We first begin with some context for the epidemic.
-
-## Background for 1861-1862 measles epidemic
+[[sub-section goal: outline goal of section -- how to use epicompare in a real situation
+
+To conclude our paper, we demonstrate how \pkg{EpiCompare} streamlines the data analysis process with a case study of a measles outbreak in 1861-1862 Germany.
+With the help of \pkg{EpiCompare}, we can answer important questions such as is a SIR model a good fit for the data, does the outbreak spread differently within the different school classes of the children, and can incorporating an underlying network structure enhance model fit?
+After introducing the data available in this outbreak, we show how \pkg{EpiCompare} can aid in each step of the data analysis pipeline (see Fig.
+\ref{fig:pipeline2}) to answer these questions motivated by this outbreak.]]
+
+## Background for 1861-1862 measles outbreak
+[[Goal of this section isto provide context for our case study.]]
 
@@
 We show a subset of the data in Table \ref{tab:hags-people}.
-We are particularly interested if and how group structure plays a role in the spread of measles.
+We are particularly interested if the SIR model is a good fit to our data, whether we can identify subgroups with differing infection behavior, and whether incorporating those subgroups into a network-based SIR model enhances model fit compared to the baseline SIR.
 
@@
 
-We begin our analysis by first examining and transforming the raw data (\code{hagelloch_raw}), which is  individual-level data.
-With \pkg{EpiCompare}, we can quickly determine how the epidemic progresses through the  population at an aggregate level.
+[[The goal of this is to show how EpiCompare can be used to transform the data for basic tablaes and plotting, especially with regards to aggregating the data for the SIR model.]]
+
+We begin our analysis by first examining and transforming the raw data (\code{hagelloch_raw}) to explore whether the SIR model is a good fit to the data.
+The raw data is in the format of individual-level data and first needs to be aggregated.
 By specifying the time of infection (\code{tI}) and the time of recovery (\code{tR}), the function \code{agents_to_aggregate()} calculates the number susceptible, infectious, and recovered individuals at each time step.
@@
 \ref{fig:hag-tern-raw}) or with our custom \code{geom}, \code{geom_aggregate()}, which takes the raw agent data as input.
+This is show in the below code.
 
@@
 
-Previous study tells us that measles outbreaks are often associated with children within the same grade level, and we examine if this is the case here.
-By combining the function \code{facet_wrap()} with \code{geom_aggregate()} we can easily analyze this scenario,
+With regards to our goal of identifying interesting subgroups with different infection behavior, previous study tells us that measles outbreaks are often associated with children within the same grade level, and we examine if this is the case here.
+By combining the \pkg{dplyr} function \code{facet_wrap()} with \code{geom_aggregate()} we can easily analyze this scenario,
 
@@
 
+[[The goal is to 'test' (although not formally) whether our observations from EDA are correct.
+We want to see whether SIR model is a good fit and whether including these found sub-groups helps model fit]]
+
+We now use modeling and simulation to informally test whether the baseline SIR model is a good fit for the data and whether incorporating a network-based structure dependent on the class of the children improves model fit.
+
 We first try to model the Hagelloch data with a baseline stochastic SIR model, which we refer to as the 'simple SIR.' In our full vignette \shannon{(available online),} we show how to fit this simple SIR model via maximum likelihood, a common approach used to fit parameters, and simulate from the model with those best fit parameters.
@@
 The result of our simulation is the object \code{agents} which is a `r nrow(agents)` $\times$ `r ncol(agents)` data frame, which details the time of entry into the $S$, $I$, and $R$ states for a given simulation.
-Ultimately, we would like to know if this simple SIR model is a good fit to the data, especially in comparison to a SIR model where we can incorporate the class structure.
-
-To fit a more complex SIR model with a network structure, we use package \pkg{EpiModel} \citep{Jenness2018}.
-The below code sets up a network of individuals (which includes class as a variable) and then simulates infections over this network.
+
+To fit a more complex SIR model with a network structure, we use the package \pkg{EpiModel} \citep{Jenness2018}.
+The below code sets up a network of individuals (which includes class as a variable) and then simulates infection and recovery over this network.
 
@@
 
-The next step is to compare the simple SIR model to the EpiModel SIR model.
-The \pkg{EpiCompare} function \code{fortify_aggregate()}, takes in an object from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transforms it into a tidy-style data frame.
+[[Finally we want to answer the questions in our analysis by comparing the models to the data and the models to each other]].
+
+Ultimately we want to compare the models to the data and the models to one another to answer our questions of interest.
+The \pkg{EpiCompare} function \code{fortify_aggregate()} takes in an object from specialized classes of modeling output \shannon{(like those made by \code{netsim()})} and transforms it into a tidy-style data frame.
 
@@
 With the two modeling objects both in the same format, we can then compare the models side-by-side.
-The results are shown in Figure \ref{fig:hag-simple-sir}, where a 90\% prediction region is estimated for the two models.
-For the Simple SIR model, we see that while the prediciton region covers the data fairly well, the prediction region clearly misses the second peak of infection.
-This indicates that the simple SIR is not a good fit to our data.
+The results are shown in Figure \ref{fig:hag-simple-sir}, where a 90\% prediction region is estimated for the two models and the actual data are plotted as black dots.
+For the Simple SIR model, we see that while the prediction region covers the data fairly well, the prediction region clearly misses the second peak of infection.
+This indicates that the simple SIR is not a good fit to our data, even after incorporating noise.
 We also see that the prediction region is very large, covering up a large area of the ternary plot.
 Together, this indicates that the simple SIR model produces a biased model with a large amount of variance.
-On the other hand, for the EpiModel network model, we see that the prediction region covers the data quite well and takes up less area.
-With \pkg{EpiCompare}, we can see that the model using the class structure is a better fit to the outbreak than the simple SIR model.
+On the other hand, for the EpiModel network model, we see that the prediction region covers the data quite well and takes up less area compared to the simple SIR.
+With \pkg{EpiCompare}, we see that the model using the class structure is a better fit to the outbreak than the simple SIR model.
 
@@
 
-The prediction region generated by the network model covers the observed data well, but that does not mean that individual filaments generated from the network model are good fits to the observed data.
+Although the prediction region generated by the network model covers the observed data well visually, that does not mean that individual filaments generated from the network model are good fits to the observed data.
 We can further examine the model fits by incorporating these filaments into our visualization.
 In Fig.
-\ref{fig:hag-simple-sir} we show the indivdual filaments generated from the two sets of models as gray lines.
-Examining these, we see that the individual lines typically only have one defined peak, whereas the data certainly looks like it has two distinct peaks, likely caused by our assumed super-spreader event.
+\ref{fig:hag-simple-sir} we show the individual filaments generated from the two sets of models as gray lines.
+Examining these, we see that the individual lines typically only have one defined peak, whereas the data certainly looks like it has two distinct peaks, a feature possibly caused by our speculated super-spreader event.
 
@@
 We then compared a baseline simple SIR model to a more complicated SIR model which incorporated a network structure which included the class structure.
-Based on the prediction regions generated from these models, we saw that the network model fit the data much better than the simple SIR model.
-However, when we examined the individual filaments generated by the network model, we found that the data is highly unlikely to be generated from such a model.
+Based on the prediction regions generated from these models, we saw that the network model fit the data better than the simple SIR model.
+However, when we examined the individual filaments generated by the network model, we found that the data are unlikely to be generated from such a model.
 For further analysis, we would recommend looking into models that can more accurately capture super-spreading events based on the observation that one child was allegedly responsible for nearly all of his classmates' infections.

### 2021-07-01 jss/EpiCompare-paper/EpiCompare-paper.Rmd
merge of EpiCompare2 and EPiCompare papers and addition of .trackdown

@@
 historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g.][]{ferguson2020}.
-At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process \ben{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
-\footnote{\ben{[Ben says: probably should have a conclusion sentence here - seems to end abruptly.
-*This is less so the case now.]}}
-
-With the many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021}, this field has a lot of devices to aid an individual practitioner decide the correct approach.
+At the same time, descriptive statistics and visualizations from universities, many branches and levels of government, and news organizations  are an important first step of the process, as has been seen in the current COVID-19 pandemic \citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+
+With many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} available, this field has a number of devices to \shannon{\sout{aid}} \shannon{help}\footnote{\ben{[Ben says: I'm not sure this an improvement]}}  an individual practitioner decide the correct approach.
 For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
 
-At the same time, analysis packages often only address a specific portion of the analysis pipeline\ben{\sout{, for instance focusing on certain types of models.}} \ben{These m}odeling tools\ben{\sout{, which}} usually require learning package-specific syntax\ben{\sout{,}} and often don't provide easy ways to compare and assess their models on new data.
-Moreover, exploring\ben{, \sout{and}} modeling \ben{and comparing} epidemics require transforming and \textit{tidying} data in different ways.
+At the same time, analysis packages often only address a specific portion of the analysis pipeline.
+These modeling tools usually require learning package-specific syntax and often don't provide easy ways to compare and assess their models on new data.
+Moreover, exploring modeling and comparing epidemics require transforming and \textit{tidying} data in different ways.
 To fill these gaps, we present our \proglang{R} package \pkg{EpiCompare}.
@@
 
-\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a \textit{time-invariant} way.
-Epidemics, despite by definition being a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales to understand the processes at play.
-<!-- Additionally, many tools designed to examine the quantity of the population in each epidemic state (for example: quantity of Susceptible vs Infectious vs Recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states.
-\shannon{I don't quite understand this.
-Is this because of underreporting?
-If so I'm not sure how EpiCompare helps us here.} -->
-\ben{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion.
-\sout{Moreover, m} M}any current \ben{time-dependent} comparison tools for state-space models (e.g.
-SIR models) \ben{\sout{highlight} examine} the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
-\ben{These \sout{This}} approach\ben{es} may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.
-Tools in \pkg{EpiCompare} give the user the ability to extend their toolkit to evaluate epidemics within a time-invariant lens.
-The goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
-
-This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a \ben{\sout{thorough}} demonstrating of the tools through a detailed example of a full data analysis pipeline.
-
-# Motivation and tools for time-invariant analysis {short-title="Time-invariant" #sec:time-invariant}
-
-<!-- \shannon{time invariance vs. time invariant analysis??} -->
-
-<!-- \shannon{I again changed this section a lot, but hopefully it's more in line with what you are thinking.
-The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot\footnote{\ben{[Ben says: defend how this fits into the narrative?]}} and 4) talk }\ben{Time invariance in} \sout{about}\shannon{ models beyond SIR.} -->
-
-<!-- \shannon{I noticed the ternary plot kinda fell out of this section and it's not really detailed anywhere else in the paper.
-So I put it back in.
-It's a non-standard plot and a feature of EpiCompare so I think it's worth talking about, and I think it's appropriate to introduce in this section.} -->
-
-<!--Epidemics can be difficult to compare to one another due to differences in diseases, locations, or population behaviors, or times.
-In \pkg{EpiCompare}, we emphasize comparisons between epidemics adjusting for the last component, time.
-Time-invariant analysis is beneficial because by adjusting for the unit of infection rate, we can focus on the "lifetime" of an epidemic, a view that is concerned more with the number of lives affected than as opposed to any specific time constraint.
-Time-invariant analysis is also beneficial when there are gaps in time between occurrences of outbreaks of a similar nature in different geographic regions.
-Finally, time-invariant analysis is beneficial because many studies talk about a period of ``exponential growth'' of the number of outbreaks \citep{chowell2007,wallinga2007generation,forsberg2008}.
-
-Time-invariant analysis, as it appears in \pkg{EpiCompare} solves the above issues by observing functionals which attempt to capture the general shape of the epidemic with respect to the proportion of the population in each epidemic state.
-This allows us to compare scenarios as different as, for instance as a decades-long outbreak HIV in the US compared to a 10 day outbreak of norovirus on a cruise ship.
-Moreover, this tool avoids the need for choosing a 'beginning' $t_0$ or 'end' $t_F$ of an epidemic, choices \cite{gallagher2020} show that can heavily influence, for example,  estimates of peak infection height or the reproduction number $R_0$.
--->
-
-\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking a global, not marginal view of how epidemics move through populations and (2) by treating full epidemics as filamental trajectories.
-The following section aims to highlight the strengths of \textit{time-invariant} \shannon{analysis} and define the mathematical foundations that \pkg{EpiCompare}’s tools stand upon.
-
-Mathematically, epidemics are complex objects.
-They can be hard to assess and compare to one another due to the differences in the diseases, the location where the outbreak occurs,  how the affected population reacts, and the time \sout{related}\shannon{related} features (including start of the epidemic, speed of infection and more).
-Time-invariant analysis makes different epidemics easier to compare by removing many time dependent aspects of an epidemic.
-\sout{Instead,} \shannon{Time-invariant analysis} focuses \shannon{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.} \ben{[Ben wants to try this sentence again.]}
-
-## Motivating time-invariant analysis through the reproduction number $R_0$ {short-title="motivating through R0" #sec:r0}
-
-Time-invariant analysis, as it appears in \pkg{EpiCompare}, \shannon{bypasses} many difficulties \shannon{in} comparing different epidemics.
-With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible.
-Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$.
-\sout{\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}^[\shannon{I don't think this is a necessary sentence.} \ben{I still think it adds value to the story and I'm not sure people really read section titles that are long.}]
-
-<!-- \ben{ -->
-<!-- Story:  -->
-<!-- (1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic.
-Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020).
--->
-<!-- } -->
-
-$R_0$ is probably the most famous \sout{time-invariant} numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
-$R_0$ is \sout{a one-number summary of a disease and is }defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.
-\ben{This \shannon{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\shannon{.
-Yet $R_0$ is} estimated with time-\sout{based}\shannon{dependent} data, which can make it a difficult quantity to estimate.} \shannon{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\shannon{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\ben{hard to define precisely.\sout{do not have precise definitions}}.
-To demonstrate the difficulty of discerning $R_0$ in \ben{\sout{a}other}\footnote{\ben{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}} time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model.
-This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,
+One of \pkg{EpiCompare}’s main contribution to comparison and assessment of epidemics is through tools that provide \textit{time-invariant} assessments.
+Epidemics, despite being defined as a process that evolves over time, often need to be compared in a way not constrained to initial times or time scales in order to understand the processes at play.
+With time-invariant analysis, comparing decades-long outbreaks of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is possible.
+Compared to time-dependent comparison tools for state-space modeling,  time-invariant analysis can make it easier to compare state-space epidemic representations in a more global, holistic fashion.
+Many time-dependent comparison tools only examine the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion.
+These time-dependent approaches can reduce the amount of connections that can be seen and insights that can be drawn}, similar to examining projections of a multidimensional distribution onto a single axis, one at a time.
+Tools in \pkg{EpiCompare} extend the user toolkit to evaluate epidemics within a time-invariant lens.
+At the same time, the goal of \pkg{EpiCompare} is not to supplant existing infectious disease modeling tools and software but, rather, is a concerted effort to create standard and fair comparisons among models developed for disease outbreaks and outbreak data.
+
+This paper is broken up into the following sections; section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a  demonstration of the tools through a detailed example\shannon{\sout{ of a full data analysis pipeline.}} \shannon{from start to finish of the data analysis pipeline}\footnote{\shannon{maybe clearer?
+up to you} \ben{The current suggestion isn't clearer - feel free to try again.}}
+
+# Time-invariant analysis {#sec:time-invariant}
+
+\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a \textit{time-invariant} way -  approaches that remove some or all of the impact of start/end times and recording time scales when preforming the analysis.
+In this section we highlight some weaknesses of time-\textit{de}pendent analysis and define the mathematical underpinning of the time-invariant approach we take.
+To accomplish this goal we  first demonstrate the inability of time-dependent tools to adequately quantify a classic epidemic parameter, the reproduction number $R_0$. 
+Then we motivate new time-invariant approaches for more complex situations where $R_0$ does not capture the complexities of outbreaks.
+This leads to the final subsection that defines ways to view epidemics in a time-invariant lens and discusses natural properties of these representations that allow for clearer ways to compare models and epidemics.
+
+## Motivation of time-invariant analysis through the reproduction number $R_0$ {short-title="Motivation through the reproduction number $R_0$" #sec:r0_subsection}
+
+Epidemiological research is often interested in understanding intrinsic properties of the epidemic, independent from when the epidemic occurred or how frequently data was collected.
+One of the most famous numerical summaries of an epidemic is the reproduction number - $R_0$.
+This numerical summary is a time-invariant value that is defined as the expected number of infections caused by a single infect or who is added to a completely susceptible population \citep{anderson1992}.
+\citet{Gallagher2020} showed that the estimation of $R_0$  can be sensitive to time-dependent parameters like estimation of the beginning and end of an epidemic.
+This time-dependent sensitivity in estimating $R_0$ reflects a general problem in epidemiology surrounding how to transform time-dependent information into insights into desirable, intrinsic properties of an epidemic.
+
+In many situations, epidemiologists want a much deeper understand of an epidemic than just the reproduction number.
+One common time-dependent tool used to better understand epidemics is a series of time series line plots that track the proportion of the population in a given state (e.g.
+infected) at a given time.
+Figure \ref{fig:different-scales-standard} visualizes two different simulated epidemics with population states (S)usceptible, (I)nfectious and (R)ecovered.
+These simulations have been generated under a discrete approximation of  \citet{Kermack1927}’s SIR model.
+The Kermack and McKendrick model captures the transitions from one state to the next as a system of ordinary differential equations
 
@@
     I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
-    R^\prime(t) &= \gamma I(t) \nonumber.
+    R^\prime(t) &= \gamma I(t) \nonumber,
 \end{align}
 
-From this model,\shannon{the reproduction number is the ratio of the infection rate to the recovery rate,} $R_0 = \beta/\gamma$\sout{, aka the ratio of the infection rate compared to the recovery rate.} \ben{From this definition, given
-\sout{Since}} $\beta$ and $\gamma$ are both rates, \ben{it should be clear that} the ratio of the two, $R_0$, is a time-invariant quantity.\footnote{\ben{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}} \sout{ Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
-Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
-}^[cool facts about r0, but not the central point]
-
-\ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
-Also, I feel like the story is weak here.
-The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$.
-In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
-\shannon{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
-also highlighted tie to time-invariant analysis and $R_0$.
-I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}\shannon{ Shannon tries again in \textcolor{blue}{blue}}
-
-\sout{\ben{[Ben says: this paragraph needs to still be looked at.
-Also I'm not sure why this particular paragraph was changed so much at all.
-Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \shannon{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
-The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
-Both epidemics have populations of 1000 people with 10 individuals initially infected.
-Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.
-We plot the epidemics with traditional $state$ vs.
-$time$ plots}\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
-\sout{In Fig.
-\ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.
-In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.
-On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \ben{epidemic} has not yet peaked at time 15.
-In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$}\footnote{\ben{This sentnce doesn't connect wit previous examples.}}.
-
-\textcolor{blue}{Since $R_0$ is an important value, it would be helpful to have more intuitive ways of comparing one $R_0$ to another.
-Usually numerical summaries of $R_0$ are presented, which while overall very helpful, may be confusing when presented along side epidemic data that are visualized in a traditional, time-dependent manner.}
-
-\textcolor{blue}{For example, consider two epidemics generated from the Kermack and McKendrick SIR equations where both have the same value of $R_0$.
-The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
-Both epidemics have populations of 1000 people with 10 individuals initially infected.
-An analysis may present an estimate of $\hat{R}_0 = 2$ alongside state vs.
-time plots like those shown in Figure \ref{fig:different-scales-standard}.
-The paths of the epidemics in the state vs.
-time view seem to differ from one another including having different infection peaks.
-From these traditional time-based plots, there is no intuitive way to conclude that these two epidemics have the same value of $R_0$.}
+where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery.
+
+From this model, the reproduction number is the ratio of the infection rate to the recovery rate, $R_0 = \beta/\gamma$.
 
@@
 
-\textcolor{blue}{\pkg{EpiCompare} provides a time-invariant tool to visualize these epidemics in a more intuitive manner, at least in regards to comparing values of $R_0$.} \sout{A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.}  \ben{For every time \sout{point} $t$ we have a \shannon{point} $(S(t),I(t), R(t))$\sout{, so we can treat epidemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.} \shannon{so we can visualize the trajectory of the epidemic in three-dimensional space (see Fig.
-\ref{fig:different-scales-tern} (left)).} For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern} (right).
-\sout{In Fig.
-\ref{fig:different-scales-tern}
-we plot the filamental trajectories of the two epidemics in a time-invariant view.
-We will explain how and why this works shortly.
-The important takeaway is that in this time-invariant view, i}I}t is apparent that these epidemics are on "the same path." In this case, this indicates that two epidemics have the same value of $R_0$.
+Even though Figure \ref{fig:different-scales-standard}’s visual analysis might be able to present more complex properties of the epidemic, one might wish to understand how these two simulated epidemics’ $R_0$ values compare.
+Even though these two simulated epidemics appear different in these figures (including having different infection peaks), there is no real intuitive way to compare the epidemics' $R_0$ values.
+In fact, both of these simulated epidemics started with the same population (1000 total people with 10 infected) and have the same $R_0$ value, just with slightly different infection and recovery rates ($\beta_1, \gamma_1 = 0.8, 0.4$, and $\beta_2, \gamma_2 = .64, .32$).
+Even for simple generative models, this time-dependent visualization cannot be used to compare a very simple but powerful numerical summary, $R_0$.
+
+## Ternary plots: a time-invariant visualization tool
+
+The faceted time series plot, like that seen in Figure \ref{fig:different-scales-standard},  not only fails to allow the practitioner to compare simulated epidemics’ $R_0$ values but also presented epidemics’ trajectory data so that only the marginal information for each state can be examined at a given time.
+A more holistic approach to visualizing the overall path of our simulated epidemics is to examine how each epidemic traverses the three-dimensional $(S(t),I(t),R(t))$ space.
+Figure 3’s left plot does just that, presenting the trajectories of the simulated epidemics in this three-dimensional space.
+For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size) we can visual these point in a a two-dimensional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern}’s right plot.
+In both of Figure  \ref{fig:different-scales-tern}’s plots we observe that the two epidemics are on the same trajectory.
+In this simple generative example setting, being on the same trajectory indicates that the two epidemics have the same $R_0$ value.
+This can be proven mathematically.
 
 ```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Left: trajectory of epidemic in three-dimensional space, plotting $(S(t), I(t), R(t))$.
-Right: the gray-shaded region and epidemic trajectory shown from (left) now shown in two-dimensional space.
+Right: the gray-shaded region and epidemic trajectories shown from (left) now shown in two-dimensional space.
 This is more commonly known as a ternary plot.", fig.show="hold", out.width="49%"}
@@
 
-\ben{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig.
-\ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space.
-\sout{The reason why we can visually compare $R_0$ in Fig.
-\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}} A filamental trajectory can be mathematically viewed as a set of points in space that have an ordering, and that all points on the line between these ordered points are also contained in the geometric object.
-For a SIR epidemic, we can represent the associated filamental trajectory $\psi$ as
-
-$$
-\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
-$$
-where a mapping $\xi : s \to \mathbb{R}$ that is strictly monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi \equiv \psi$ where :
-
-$$
-\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
-$$
-\ben{[Ben says: removed this paragraph now]} Since the number \shannon{of individuals} in each state is non-negative and the sum over the three states for a given time point \sout{sums to}\shannon{is} $N$, then \ben{all points in} $\psi$ will lay in a  \ben{two-dimensional triangular plane in three-dimensional space.
-\sout{We can then which can be visualize the \shannon{full filamental trajectory} in a two-dimensional ternary plot}  As a result, we can visualize \shannon{the full filamental trajectory} of an epidemic in a single \ben{ternary plot \sout{2d plot and ultimately $R_0$}}.
-\shannon{[Shannon says: Show pic here?]}]\footnote{Which?
-\shannon{3d space one?
-But I'm leaning against it now.
-3d never looks good in a paper.}}
-
-\sout{\ben{[This section could be a bit less wordy.
-But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
-Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.
-Now, it is much clearer that \ben{\sout{Model} epidemic} 2 is following the same trajectory as \ben{\sout{Model} epidemic} 1 but is not as far along in the infection process.
-}
-
-\sout{\ben{As suggested a few paragraphs back, t}}The filamental trajectories in Fig.
-\ref{fig:different-scales-tern} \sout{seem to overlap}, and we may suspect that something is fundamentally linking these two different epidemics together.
-Mathematically, we can show this fundamental link \sout{turns} is $R_0$.
-Let our two epidemics be presented as $\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}$, $\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}$ respectively.
-As with the example, assume both models have the same initial values $(S(0), I(0), R(0))$, and let $R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
-And define $a>0$ to be the relative scalar such that $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$.
-
 \begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+ Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (i.e.
+$R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$) and define $a > 0$ such that $\beta_2 = a\beta_1$.
+Then for all $t > 0$ there exists $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
+Moreover, $s = \frac{1}{a}t$.
 \end{theorem}
@@
 The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model.
-\shannon{In other words, the two epidemics form the same filamental trajectory.
-For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}
-
-<!-- ## $R_0$  and time-invariant analysis {short-title="r0" #r0}
-
- By definition, $R_0$ is the number of expected secondary infections when a primary infection is introduced to a susceptible population.
-$R_0$ is also, maybe, the most famous \textit{time-invariant} numerical summary of an epidemic, which allows epidemics to be compared to one another in both different time and geographic scales.
-For example, $R_0$ for Covid-19 is estimated to be between 2-3, seasonal influenza between 1.2-2, and modern measles outbreaks as large as 12 \citep{midas2020,biggerstaff2014,namee2018}.
-
-Estimators for $R_0$ are dependent on the epidemic modeling framework, which consists of which states an individual can occupy (e.g.
-susceptible, infectious, recovered) and a description of how individuals move from one state to the next over time (see \cite{hethcote1994}).
-A common epidemic modeling framework is the SIR model, originally introduced by @kermack1927.
-Transitions from one state to the next are defined by a series of ordinary differential equations, where $N$ is the (fixed) total number individuals, $\beta$ is the rate of infection, and $\gamma$ is rate of recovery,
-\begin{align}\label{eq:sir-ode}
-      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
-      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
-      R^\prime(t) &= \gamma I(t) \nonumber.
-  \end{align}
-From this, $\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}$, the ratio of the estimated infection rate compared to the estimated recovery rate.
-
-With regards to traditional epidemic $state$ vs.
-$time$ plots, $R_0$ is difficult to visualize, especially with respect from one epidemic to another.
-For example, consider the scenarios where the first epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_1 = 0.3$ and $\gamma_1 = 0.15$, and the second epidemic is generated from a SIR model with $(S(0) = 990, I(0) = 10)$, $\beta_2 = 0.24$ and $\gamma_1 = 0.12$ over the first 15 days.
-Both epidemics have the same value of $R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2$.
-The epidemic trajectories are shown in the $state$ vs.
-time plots in Figure \ref{fig:different-scales-standard}.
-At a glance, we may assume that Model 1 has a larger $R_0$ than Model 2 because the peak of infection occurs more quickly than in Model 2.
-On the other hand, we may think Model 2 has a larger $R_0$ because we may think the number of infections in that model has not yet peaked at time 15.
--->
-
-<!--
-However, when we plot the trajectories as a single curve using the ternary plot in the time-invariant view, we immediately see a different story.
-In this time-invariant view in Fig.
-\ref{fig:different-scales-tern}, the points seem to overlap and form the same trajectory.
-Now it seems to be that Model 2 is following the same trajectory as Model 1 but is not as far along in the infection process.
-We can see there is something fundamentally linking these two different epidemics, and this fundamental link turns out to be $R_0$.
-
-More formally, let two Kermack and McKendrick  SIR models(see Eq.\eqref{eq:sir-ode} )  be denoted $(S_1(t), I_1(t), R_1(t))$ and $(S_2(t), I_2(t), R_2(t))$, respectively, for $t > 0$.
-Assume both models have initial values $(S(0), I(0), R(0))$.
-Let $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ where $\beta_i$ and $\gamma_i$ are the average infection rate and recovery rate, respectively, for SIR model $i=1, 2$.
-Equivalently, $\beta_2 = a \beta_1$ if and only if $\gamma_2 = a \gamma_1$ for some $a > 0$.
-
-\begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
-\end{theorem}
-
-The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.
-The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model is also a point on the epidemic path of the second SIR model.
-Taking the sample simulations from Fig.
-\ref{fig:different-scales-standard}, Fig.
-\ref{fig:different-scales-tern} presents these two models in a ternary plot.
-This means with our time-invariant ternary plot, that at a glance, we can tell if two epidemics have different values of $R_0$.
--->
-
-##  Time-invariant analysis beyond $R_0$ and \sout{Kermack’s and McKendrick} SIR Models\footnote{Probably will need to change this title...} {short-title="Beyond R0 and SIR" #sec:beyond-r0-sir}
-
-Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us to \sout{better}\shannon{more fully} compare the overall structure of the epidemic and see how the population is directly impacted.
-Time-invariant tools \sout{that} can be useful even when the underlying generative model for the epidemic is unknown or have more than three epidemic states.
-
-\ben{New paragraph} Viewing epidemics as filamental trajectories provides \sout{a lot }new ways to compare and examine epidemics in a time-invariant manner.
-\ben{For \shannon{completed?}epidemics that have ended, one way to examine their filamental trajectories is to \sout{redefine}\shannon{represent} the \shannon{filamental} trajectory as a \shannon{finite sequence of equally spaced points.} \sout{finite sequence of points on the filamental trajectory that are equally spaced (i.e.
-equa-distance between pairs of ordered points).} \sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \shannon{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \shannon{[Shannon says: are we missing a few words in this definition?]}}.
-This representation induces a natural distance between this type of representation between epidemics, specifically:
+The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$, then every point on the epidemic path of the first SIR model can be mapped to a point on the epidemic path of the second SIR model.
+In other words, the two epidemics form the same filamental trajectory.
+
+## Time-invariance beyond SIR models: Trajectories and Filaments
+
+Through the $R_0$ example, we see that treating epidemics like filamental trajectories embedded in a lower dimensional space allows us  to better compare the overall structure of the epidemic and see how the population is directly impacted.
+In this section we present time-invariant tools that can be applied to complex epidemics  where the epidemic’s generative process is unknown and can have more than three states.
+These tools leverage the idea that an epidemic can be viewed as a trajectory and that many properties of the epidemic are well captured when we do so.
+This approach is useful when the epidemic of interest has only gone through a single realization of its outbreak (before the population of individuals become susceptible again).
+
+The first set of tools allows a practitioner to define distances between epidemics whose time features do not align.
+For completed epidemics, one way to better examine their properties is to represent their filamental trajectories as a finite sequence of equally spaced points.
+This representation induces a natural distance between epidemics, specifically:
+
 \[
-d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
 \]
-where $\psi_i'(s)$ the point along $\psi_1$ that is $s$ fraction of $|\psi_1|$ distance away from the start of $\psi_i$.\footnote{\shannon{I think you're trying to say something about a distance based on the equally space points.
-Some clarifying questions:  is $\psi^\prime$ the derivative?
-Does proportion make more sense than fraction?
-or simply $\frac{|\psi|}{s}$?
-It's only naturally time-invariant if we have a well defined ending point, right?}} This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
-These extremeness estimators can be \sout{very} useful when comparing \sout{between a setof simulation} \shannon{a set of simulated} epidemics and the true epidemic\shannon{.
-Moreover these extremness estimators,} \sout{and does} \shannon{} not constrain the number of states of the model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\ben{I'm not sure we've talked about this before...
-\shannon{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
-
-\ben{New paragraph:} \ben{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If \sout{one} \shannon{a practicioner} is interested in understanding \shannon{an} epidemic\sout{s} through a single \sout{cycle}\shannon{realization} of \sout{their}\shannon{its} outbreak} (before the population of individuals have become susceptible again), then additional time-invariant tools, including prediction regions can be leveraged \shannon{awk sentence}^[{\shannon{What are we predicting if the epidemic is done?
-Update 4/6: I'm satisfied.}}].
-In these settings, \shannon{\pkg{EpiCompare} goes} \sout{we go }a step further and treats epidemics more like geometric filaments \shannon{(i.e.
-filamental trajectories without an ordering of points)} than filamental trajectories.
-In \pkg{EpiCompare}, we create prediction regions that  contain a the top ($1-\alpha$) proportion of simulated curves by defining geometric regions defined by the union of small \shannon{geometric?} filaments around the subset of simulations (\sout{subset} \shannon{grouped} by measures like the above pseudo-density estimates or depth estimates).
-These regions \sout{look at}\shannon{show} where in the state-space we expect the epidemic to traverse\sout{, and }.
-\shannon{Additionally, }we can compare prediction regions defined by different models using \ben{many set difference distances \sout{the Haussdorf \shannon{why Haussdorf specifically?} distance}} as well as examining how well the truth epidemic matches the simulations by examining if the epidemic's trajectory lies within the prediction region.
-All these \shannon{mentioned?} geometric structures and distance notations apply to epidemics with any number of states, and at the end of  Section \ref{sec:overview} we \sout{also} highlight how these prediction regions can aid in visual comparisons for epidemics with 3 states (like the SIR models).
-
-<!--
-## Beyond the Kermack and McKendrick SIR models
-
-Although the result of Theorem \ref{thm:sir-scale} allows for easy visual comparison of $R_0$ in Kermack SIR models, it does require stringent assumptions of homogeneity of behavior in populations.
-The use of visualizing epidemics in a time-invariant lens via ternary plots extends beyond those of models that follow the ODEs in the Kermack-McKendrick equations.
-Any model with S, I, and R states can be visualized with ternary plots, regardless of birth and death dynamics and regardless of homogeneity of individual behavior.
-We can use ternary plots to compare the spread of a disease for groups within a population without time as a confounding factor.
-We show an example of this in a later section.
-
-Moreover, time-invariant analysis is also useful for epidemic models with more than three epidemic states.
-The constraints in three dimensions that are met with the SIR model (there are three epidemic states and for each time step, the sum of total number of individuals in each state is the population size at that time) represents a space of 3d simplices, and the ternary plot specifically represents these simplices, after scaling the values in each state as proportions of the total population at that time (ternary plots are known as a 3d unit simplex due to it's scaling).
-This same scaling for larger models (i.e.
-with more states) can be done onto different simplexes.
-In this package we present tools to help compare models (mostly through simulations).
-\pkg{EpiCompare}'s tools permits comparisons of $d$-dimensional objects by projecting them into ($d-1$)-dimensional space.
-
- Even though higher dimensional models may not be able to visualized simply, we provide a number of tools to aid in the comparison of models and epidemics.
-The first of which uses multiple simulations under specific model parameters to assess the \shannon{variability?} bairabilty of the model fit.
-In \pkg{EpiCompare}, we provide ways to create prediction regions for a `true' epidemic under a fully specified model.
-These regions require representing multi-dimensional structures for functions to completely contain epidemics.
-\pkg{EpiCompare} treats these simulations and epidemics as \textit{filamental} objects -- \shannon{need a description for filamental, Ben feel free to change} where not only the outbreak trajectory points are important but also their form which can be considered a function.
-\shannon{could be useful to have a picture/example here?
-like  three curves and 2 metrics one where closer under a pointwise distance and the other closer in filamental distance?
-I'll sketch up an idea unless you have something in your proposal already}
-
- We extend off of papers like \citet{Dalmasso2019a} to create these bands.
-These high dimensional bands allow the user to assess if the true epidemic is within the band (thereby assessing the model's representation of the epidemic).
-With these bands, we can compare \textit{models} to \textit{models} by assessing the distance between the two bands of model representation.
-We recommend using the Hausdorff distance to compare two bands to one another, as this distance captures how much bigger the sets would have to expand to cover each other.
-The Hausdorff distance is defined mathematically as
-\[
-d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y),\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
-\]
-
--->
+where $\tilde{\psi}_i(s)$ is the point along $\psi_i$ that is $s\cdot|\psi_i|$ distance away from the start of $\psi_i$ (with $|\psi_i|$ is the length of $\psi_i$).
+This distance is naturally time-invariant, and can be plugged into multiple distance-based assessment tools to examine the overall ``extremeness'' of points, including pseudo-density estimators and depth/local depth functions \citep[for examples see][]{Ciollaro2016, Geenens2017}.
+These extremeness estimators can be useful when comparing the true epidemic to a set of simulated epidemics, and practitioners can interpret the epidemic's extremeness score relative to the extremeness scores of the simulations very easily.
+
+In addition to being used to better understand completed epidemics, time-invariant tools can be used to aid prediction of future epidemics by providing a state-space regions in which we expect the true epidemic to traverse.
+In settings where the epidemic only generates a single outbreak, these regions can be very telling if simulation models capture the epidemic’s structure.
+In \pkg{EpiCompare} we create geometric prediction regions around all but the $\alpha$ proportion of most extreme simulated trajectories.
+These geometric regions can also be used to compared simulation models that have different time scales, parameters and even different statistical philosophies, through set different distances.
+Although visualization is the easiest when the epidemic has three states, prediction regions can be useful to assess and compare simulation models where the epidemics have multiple states.
+
+Overall, there are many tools to aid in the assessment and comparison of epidemics and models that avoid being affected by time-based parameters.
+We believe the time-invariant analysis provides many insights and should be in the toolkit of many epidemiologists.
+\pkg{EpiCompare} provides a strong starting point to do just that.
 

### 2021-07-01 jss/EpiCompare-paper/EpiCompare-paper.tex
merge of EpiCompare2 and EPiCompare papers and addition of .trackdown

@@
 government, and news organizations are an important first step of the
-process
-\textcolor{violet}{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
-\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}}
-
-With the many visualization and exploratory tools, models and modeling
+process, as has been seen in the current COVID-19 pandemic
+\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+
+With many visualization and exploratory tools, models and modeling
 paradigms, and reviews and comparisons in the literature and through the
 MIDAS (Models of Infectious Disease Agent Study) network
-\citep{midasnetwork2021}, this field has a lot of devices to aid an
-individual practitioner decide the correct approach. For example,
+\citep{midasnetwork2021} available, this field has a number of devices
+to \textcolor{orange}{\sout{aid}}
+\textcolor{orange}{help}\footnote{\textcolor{violet}{[Ben says: I'm not sure this an improvement]}}
+an individual practitioner decide the correct approach. For example,
 \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
@@
 At the same time, analysis packages often only address a specific
-portion of the analysis
-pipeline\textcolor{violet}{\sout{, for instance focusing on certain types of models.}}
-\textcolor{violet}{These m}odeling
-tools\textcolor{violet}{\sout{, which}} usually require learning
-package-specific syntax\textcolor{violet}{\sout{,}} and often don't
-provide easy ways to compare and assess their models on new data.
-Moreover, exploring\textcolor{violet}{, \sout{and}} modeling
-\textcolor{violet}{and comparing} epidemics require transforming and
+portion of the analysis pipeline. These modeling tools usually require
+learning package-specific syntax and often don't provide easy ways to
+compare and assess their models on new data. Moreover, exploring
+modeling and comparing epidemics require transforming and
 \textit{tidying} data in different ways. To fill these gaps, we present
@@
 
-\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
-\textit{time-invariant} way. Epidemics, despite by definition being a
-process that evolves over time, often need to be compared in a way not
-constrained to initial times or time scales to understand the processes
-at play.
-\textcolor{violet}{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion. \sout{Moreover, m} M}any
-current \textcolor{violet}{time-dependent} comparison tools for
-state-space models (e.g.~SIR models)
-\textcolor{violet}{\sout{highlight} examine} the proportion of
+One of \pkg{EpiCompare}'s main contribution to comparison and assessment
+of epidemics is through tools that provide \textit{time-invariant}
+assessments. Epidemics, despite being defined as a process that evolves
+over time, often need to be compared in a way not constrained to initial
+times or time scales in order to understand the processes at play. With
+time-invariant analysis, comparing decades-long outbreaks of HIV in the
+US to a 10 day outbreak of norovirus on a cruise ship is possible.
+Compared to time-dependent comparison tools for state-space modeling,
+time-invariant analysis can make it easier to compare state-space
+epidemic representations in a more global, holistic fashion. Many
+time-dependent comparison tools only examine the proportion of
 individuals in each state (at a given time) in a piece-wise / marginal
-fashion. \textcolor{violet}{These \sout{This}}
-approach\textcolor{violet}{es} may reduce the amount of connections that
-can be seen, similar to projections of a multidimensional distribution
-onto a single axis at a time. Tools in \pkg{EpiCompare} give the user
-the ability to extend their toolkit to evaluate epidemics within a
-time-invariant lens. The goal of \pkg{EpiCompare} is not to supplant
-existing infectious disease modeling tools and software but, rather, is
-a concerted effort to create standard and fair comparisons among models
+fashion. These time-dependent approaches can reduce the amount of
+connections that can be seen and insights that can be drawn\}, similar
+to examining projections of a multidimensional distribution onto a
+single axis, one at a time. Tools in \pkg{EpiCompare} extend the user
+toolkit to evaluate epidemics within a time-invariant lens. At the same
+time, the goal of \pkg{EpiCompare} is not to supplant existing
+infectious disease modeling tools and software but, rather, is a
+concerted effort to create standard and fair comparisons among models
 developed for disease outbreaks and outbreak data.
@@
 \pkg{EpiCompare} aids a practitioner in every step of the pipeline and
-section \ref{sec:tour} provides a \textcolor{violet}{\sout{thorough}}
-demonstrating of the tools through a detailed example of a full data
-analysis pipeline.
-
-\section[Time-invariant]{Motivation and tools for time-invariant
-analysis}\label{sec:time-invariant}
-
-\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking
-a global, not marginal view of how epidemics move through populations
-and (2) by treating full epidemics as filamental trajectories. The
-following section aims to highlight the strengths of
-\textit{time-invariant} \textcolor{orange}{analysis} and define the
-mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
-
-Mathematically, epidemics are complex objects. They can be hard to
-assess and compare to one another due to the differences in the
-diseases, the location where the outbreak occurs, how the affected
-population reacts, and the time
-\sout{related}\textcolor{orange}{related} features (including start of
-the epidemic, speed of infection and more). Time-invariant analysis
-makes different epidemics easier to compare by removing many time
-dependent aspects of an epidemic. \sout{Instead,}
-\textcolor{orange}{Time-invariant analysis} focuses
-\textcolor{orange}{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.}
-\textcolor{violet}{[Ben wants to try this sentence again.]}
-
-\subsection[motivating through R0]{Motivating time-invariant analysis
-through the reproduction number \(R_0\)}\label{sec:r0}
-
-Time-invariant analysis, as it appears in \pkg{EpiCompare},
-\textcolor{orange}{bypasses} many difficulties \textcolor{orange}{in}
-comparing different epidemics. With time-invariant analysis, comparing
-the decades-long outbreak of HIV in the US to a 10 day outbreak of
-norovirus on a cruise ship is \sout{still} possible. Time-dependent
-problems can arise when estimating epidemiological parameters, including
-the reproduction number \(R_0\).
-\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence.}
-  \textcolor{violet}{I still think it adds value to the story and I'm not sure people really read section titles that are long.}}
-
-\(R_0\) is probably the most famous \sout{time-invariant} numerical
-summary of an epidemic and is often associated with the
-Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
-\(R_0\) is \sout{a one-number summary of a disease and is }defined as
-the expected number of infections caused by a single infector who is
-added to a completely susceptible population \citep{anderson1992}.
-\textcolor{violet}{This \textcolor{orange}{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\textcolor{orange}{.
-Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a difficult quantity to estimate.}
-\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
-To demonstrate the difficulty of discerning \(R_0\) in
-\textcolor{violet}{\sout{a}other}\footnote{\textcolor{violet}{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}}
-time-dependent analysis, we first introduce \citet{Kermack1927}'s SIR
-model. This model captures the transitions from one state to the next as
-a system of ordinary differential equations, where \(N\) is the total
-number of individuals, \(\beta\) is the rate of infection, and
-\(\gamma\) is the rate of recovery,
+section \ref{sec:tour} provides a demonstration of the tools through a
+detailed
+example\textcolor{orange}{\sout{ of a full data analysis pipeline.}}
+\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer?
+up to you} \textcolor{violet}{The current suggestion isn't clearer - feel free to try again.}}
+
+\hypertarget{sec:time-invariant}{%
+\section{Time-invariant analysis}\label{sec:time-invariant}}
+
+\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a
+\textit{time-invariant} way - approaches that remove some or all of the
+impact of start/end times and recording time scales when preforming the
+analysis. In this section we highlight some weaknesses of
+time-\textit{de}pendent analysis and define the mathematical
+underpinning of the time-invariant approach we take. To accomplish this
+goal we first demonstrate the inability of time-dependent tools to
+adequately quantify a classic epidemic parameter, the reproduction
+number \(R_0\). Then we motivate new time-invariant approaches for more
+complex situations where \(R_0\) does not capture the complexities of
+outbreaks. This leads to the final subsection that defines ways to view
+epidemics in a time-invariant lens and discusses natural properties of
+these representations that allow for clearer ways to compare models and
+epidemics.
+
+\subsection[Motivation through the reproduction number $R_0$]{Motivation
+of time-invariant analysis through the reproduction number
+\(R_0\)}\label{sec:r0_subsection}
+
+Epidemiological research is often interested in understanding intrinsic
+properties of the epidemic, independent from when the epidemic occurred
+or how frequently data was collected. One of the most famous numerical
+summaries of an epidemic is the reproduction number - \(R_0\). This
+numerical summary is a time-invariant value that is defined as the
+expected number of infections caused by a single infect or who is added
+to a completely susceptible population \citep{anderson1992}.
+\citet{Gallagher2020} showed that the estimation of \(R_0\) can be
+sensitive to time-dependent parameters like estimation of the beginning
+and end of an epidemic. This time-dependent sensitivity in estimating
+\(R_0\) reflects a general problem in epidemiology surrounding how to
+transform time-dependent information into insights into desirable,
+intrinsic properties of an epidemic.
+
+In many situations, epidemiologists want a much deeper understand of an
+epidemic than just the reproduction number. One common time-dependent
+tool used to better understand epidemics is a series of time series line
+plots that track the proportion of the population in a given state
+(e.g.~infected) at a given time. Figure
+\ref{fig:different-scales-standard} visualizes two different simulated
+epidemics with population states (S)usceptible, (I)nfectious and
+(R)ecovered. These simulations have been generated under a discrete
+approximation of \citet{Kermack1927}'s SIR model. The Kermack and
+McKendrick model captures the transitions from one state to the next as
+a system of ordinary differential equations
 
@@
     I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
-    R^\prime(t) &= \gamma I(t) \nonumber.
+    R^\prime(t) &= \gamma I(t) \nonumber,
 \end{align}
 
-From this
-model,\textcolor{orange}{the reproduction number is the ratio of the infection rate to the recovery rate,}
-\(R_0 = \beta/\gamma\)\sout{, aka the ratio of the infection rate compared to the recovery rate.}
-\textcolor{violet}{From this definition, given 
-\sout{Since}} \(\beta\) and \(\gamma\) are both rates,
-\textcolor{violet}{it should be clear that} the ratio of the two,
-\(R_0\), is a time-invariant
-quantity.\footnote{\textcolor{violet}{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}
-\sout{ Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
-Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
-}\footnote{cool
-  facts about r0, but not the central point}
-
-\textcolor{violet}{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
-Also, I feel like the story is weak here.
-The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$.
-In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
-\textcolor{orange}{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
-also highlighted tie to time-invariant analysis and $R_0$.
-I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}\textcolor{orange}{ Shannon tries again in \textcolor{blue}{blue}}
-
-\sout{\textcolor{violet}{[Ben says: this paragraph needs to still be looked at.
-Also I'm not sure why this particular paragraph was changed so much at all.
-Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
-The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
-Both epidemics have populations of 1000 people with 10 individuals initially infected.
-Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.
-We plot the epidemics with traditional $state$ vs.
-$time$ plots}\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
-\sout{In Fig.
-\ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.
-In this time-variant view, we may believe that \textcolor{violet}{epidemic} 1 has a larger $R_0$ than \textcolor{violet}{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.
-On the other hand, we may believe \textcolor{violet}{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \textcolor{violet}{epidemic} has not yet peaked at time 15.
-In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$}\footnote{\textcolor{violet}{This sentnce doesn't connect wit previous examples.}}.
-
-\textcolor{blue}{Since $R_0$ is an important value, it would be helpful to have more intuitive ways of comparing one $R_0$ to another.
-Usually numerical summaries of $R_0$ are presented, which while overall very helpful, may be confusing when presented along side epidemic data that are visualized in a traditional, time-dependent manner.}
-
-\textcolor{blue}{For example, consider two epidemics generated from the Kermack and McKendrick SIR equations where both have the same value of $R_0$.
-The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
-Both epidemics have populations of 1000 people with 10 individuals initially infected.
-An analysis may present an estimate of $\hat{R}_0 = 2$ alongside state vs.
-time plots like those shown in Figure \ref{fig:different-scales-standard}.
-The paths of the epidemics in the state vs.
-time view seem to differ from one another including having different infection peaks.
-From these traditional time-based plots, there is no intuitive way to conclude that these two epidemics have the same value of $R_0$.}
+where \(N\) is the total number of individuals, \(\beta\) is the rate of
+infection, and \(\gamma\) is the rate of recovery.
+
+From this model, the reproduction number is the ratio of the infection
+rate to the recovery rate, \(R_0 = \beta/\gamma\).
 
@@
 
-\textcolor{blue}{\pkg{EpiCompare} provides a time-invariant tool to visualize these epidemics in a more intuitive manner, at least in regards to comparing values of $R_0$.}
-\sout{A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.}
-\textcolor{violet}{For every time \sout{point} $t$ we have a \textcolor{orange}{point} $(S(t),I(t), R(t))$\sout{, so we can treat epidemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.} \textcolor{orange}{so we can visualize the trajectory of the epidemic in three-dimensional space (see Fig.
-\ref{fig:different-scales-tern} (left)).} For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern} (right).
-\sout{In Fig.
-\ref{fig:different-scales-tern}
-we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, i}I}t
-is apparent that these epidemics are on ``the same path.'' In this case,
-this indicates that two epidemics have the same value of \(R_0\).
+Even though Figure \ref{fig:different-scales-standard}'s visual analysis
+might be able to present more complex properties of the epidemic, one
+might wish to understand how these two simulated epidemics' \(R_0\)
+values compare. Even though these two simulated epidemics appear
+different in these figures (including having different infection peaks),
+there is no real intuitive way to compare the epidemics' \(R_0\) values.
+In fact, both of these simulated epidemics started with the same
+population (1000 total people with 10 infected) and have the same
+\(R_0\) value, just with slightly different infection and recovery rates
+(\(\beta_1, \gamma_1 = 0.8, 0.4\), and
+\(\beta_2, \gamma_2 = .64, .32\)). Even for simple generative models,
+this time-dependent visualization cannot be used to compare a very
+simple but powerful numerical summary, \(R_0\).
+
+\hypertarget{ternary-plots-a-time-invariant-visualization-tool}{%
+\subsection{Ternary plots: a time-invariant visualization
+tool}\label{ternary-plots-a-time-invariant-visualization-tool}}
+
+The faceted time series plot, like that seen in Figure
+\ref{fig:different-scales-standard}, not only fails to allow the
+practitioner to compare simulated epidemics' \(R_0\) values but also
+presented epidemics' trajectory data so that only the marginal
+information for each state can be examined at a given time. A more
+holistic approach to visualizing the overall path of our simulated
+epidemics is to examine how each epidemic traverses the
+three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
+just that, presenting the trajectories of the simulated epidemics in
+this three-dimensional space. For state space models like in our
+example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
+to \(N\) (the total population size) we can visual these point in a a
+two-dimensional \textit{ternary} plot, as seen in Figure
+\ref{fig:different-scales-tern}'s right plot. In both of Figure
+\ref{fig:different-scales-tern}'s plots we observe that the two
+epidemics are on the same trajectory. In this simple generative example
+setting, being on the same trajectory indicates that the two epidemics
+have the same \(R_0\) value. This can be proven mathematically.
 
@@
 
-\textcolor{violet}{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig.
-\ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space.
-\sout{The reason why we can visually compare $R_0$ in Fig.
-\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}}
-A filamental trajectory can be mathematically viewed as a set of points
-in space that have an ordering, and that all points on the line between
-these ordered points are also contained in the geometric object. For a
-SIR epidemic, we can represent the associated filamental trajectory
-\(\psi\) as
-
-\[
-\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
-\] where a mapping \(\xi : s \to \mathbb{R}\) that is strictly
-monotonically increasing would not change the definition of \(\psi\),
-i.e.~\(\psi_\xi \equiv \psi\) where :
-
-\[
-\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
-\] \textcolor{violet}{[Ben says: removed this paragraph now]} Since the
-number \textcolor{orange}{of individuals} in each state is non-negative
-and the sum over the three states for a given time point
-\sout{sums to}\textcolor{orange}{is} \(N\), then
-\textcolor{violet}{all points in} \(\psi\) will lay in a
-\textbackslash ben\{two-dimensional triangular plane in
-three-dimensional space.
-\sout{We can then which can be visualize the \textcolor{orange}{full filamental trajectory} in a two-dimensional ternary plot}
-As a result, we can visualize
-\textcolor{orange}{the full filamental trajectory} of an epidemic in a
-single
-\textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
-\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?  \textcolor{orange}{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}
-
-\sout{\textcolor{violet}{[This section could be a bit less wordy.
-But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
-Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.
-Now, it is much clearer that \textcolor{violet}{\sout{Model} epidemic} 2 is following the same trajectory as \textcolor{violet}{\sout{Model} epidemic} 1 but is not as far along in the infection process.
-}
-
-\sout{\textcolor{violet}{As suggested a few paragraphs back, t}}The
-filamental trajectories in Fig. \ref{fig:different-scales-tern}
-\sout{seem to overlap}, and we may suspect that something is
-fundamentally linking these two different epidemics together.
-Mathematically, we can show this fundamental link \sout{turns} is
-\(R_0\). Let our two epidemics be presented as
-\(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
-\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
-example, assume both models have the same initial values
-\((S(0), I(0), R(0))\), and let
-\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
-\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
-rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
-the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
-\(\gamma_2 = a \gamma_1\).
-
 \begin{theorem}\label{thm:sir-scale}
-Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
+ Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (i.e.
+$R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$) and define $a > 0$ such that $\beta_2 = a\beta_1$.
+Then for all $t > 0$ there exists $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
+Moreover, $s = \frac{1}{a}t$.
 \end{theorem}
@@
 for two SIR models that have the same initial percent of individuals in
-each state and \(R_0\) then for every point on the epidemic path of the
-first SIR model \textcolor{violet}{\sout{is also} can be mapped to} a
-point on the epidemic path of the second SIR model.
-\textcolor{orange}{In other words, the two epidemics form the same filamental trajectory.
-For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}
-
-\subsection[Beyond R0 and SIR]{Time-invariant analysis beyond \(R_0\)
-and \sout{Kermack’s and McKendrick} SIR
-Models\footnote{Probably will need to change this title...}}\label{sec:beyond-r0-sir}
+each state and \(R_0\), then every point on the epidemic path of the
+first SIR model can be mapped to a point on the epidemic path of the
+second SIR model. In other words, the two epidemics form the same
+filamental trajectory.
+
+\hypertarget{time-invariance-beyond-sir-models-trajectories-and-filaments}{%
+\subsection{Time-invariance beyond SIR models: Trajectories and
+Filaments}\label{time-invariance-beyond-sir-models-trajectories-and-filaments}}
 
@@
 filamental trajectories embedded in a lower dimensional space allows us
-to \sout{better}\textcolor{orange}{more fully} compare the overall
-structure of the epidemic and see how the population is directly
-impacted. Time-invariant tools \sout{that} can be useful even when the
-underlying generative model for the epidemic is unknown or have more
-than three epidemic states.
-
-\textcolor{violet}{New paragraph} Viewing epidemics as filamental
-trajectories provides \sout{a lot }new ways to compare and examine
-epidemics in a time-invariant manner. \textbackslash ben\{For
-\textcolor{orange}{completed?}epidemics that have ended, one way to
-examine their filamental trajectories is to
-\sout{redefine}\textcolor{orange}{represent} the
-\textcolor{orange}{filamental} trajectory as a
-\textcolor{orange}{finite sequence of equally spaced points.}
-\sout{finite sequence of points on the filamental trajectory that are equally spaced (i.e. equa-distance between pairs of ordered points).}
-\sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \textcolor{orange}{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \textcolor{orange}{[Shannon says: are we missing a few words in this definition?]}}.
-This representation induces a natural distance between this type of
-representation between epidemics, specifically: \[
-d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
-\] where \(\psi_i'(s)\) the point along \(\psi_1\) that is \(s\)
-fraction of \(|\psi_1|\) distance away from the start of
-\(\psi_i\).\footnote{\textcolor{orange}{I think you're trying to say something about a distance based on the equally space points.
-Some clarifying questions:  is $\psi^\prime$ the derivative?
-Does proportion make more sense than fraction?
-or simply $\frac{|\psi|}{s}$?
-It's only naturally time-invariant if we have a well defined ending point, right?}}
-This distance is naturally time-invariant, and can be plugged into
-multiple distance-based assessment tools to examine the overall
-``extremeness'\,' of points, including pseudo-density estimators and
-depth/local depth functions
+to better compare the overall structure of the epidemic and see how the
+population is directly impacted. In this section we present
+time-invariant tools that can be applied to complex epidemics where the
+epidemic's generative process is unknown and can have more than three
+states. These tools leverage the idea that an epidemic can be viewed as
+a trajectory and that many properties of the epidemic are well captured
+when we do so. This approach is useful when the epidemic of interest has
+only gone through a single realization of its outbreak (before the
+population of individuals become susceptible again).
+
+The first set of tools allows a practitioner to define distances between
+epidemics whose time features do not align. For completed epidemics, one
+way to better examine their properties is to represent their filamental
+trajectories as a finite sequence of equally spaced points. This
+representation induces a natural distance between epidemics,
+specifically:
+
+\[
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
+\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
+\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (with
+\(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
+time-invariant, and can be plugged into multiple distance-based
+assessment tools to examine the overall ``extremeness'' of points,
+including pseudo-density estimators and depth/local depth functions
 \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
-estimators can be \sout{very} useful when comparing
-\sout{between a setof simulation} \textcolor{orange}{a set of simulated}
-epidemics and the true
-epidemic\textcolor{orange}{. Moreover these extremness estimators,}
-\sout{and does} \textcolor{orange}{} not constrain the number of states
-of the
-model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\textcolor{violet}{I'm not sure we've talked about this before...
-\textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
-
-\textcolor{violet}{New paragraph:}
-\textcolor{violet}{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If \sout{one} \textcolor{orange}{a practicioner} is interested in understanding \textcolor{orange}{an} epidemic\sout{s} through a single \sout{cycle}\textcolor{orange}{realization} of \sout{their}\textcolor{orange}{its} outbreak}
-(before the population of individuals have become susceptible again),
-then additional time-invariant tools, including prediction regions can
-be leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done? Update 4/6: I'm satisfied.}\}}.
-In these settings, \textcolor{orange}{\pkg{EpiCompare} goes}
-\sout{we go }a step further and treats epidemics more like geometric
-filaments
-\textcolor{orange}{(i.e. filamental trajectories without an ordering of points)}
-than filamental trajectories. In \pkg{EpiCompare}, we create prediction
-regions that contain a the top (\(1-\alpha\)) proportion of simulated
-curves by defining geometric regions defined by the union of small
-\textcolor{orange}{geometric?} filaments around the subset of
-simulations (\sout{subset} \textcolor{orange}{grouped} by measures like
-the above pseudo-density estimates or depth estimates). These regions
-\sout{look at}\textcolor{orange}{show} where in the state-space we
-expect the epidemic to traverse\sout{, and }.
-\textcolor{orange}{Additionally, }we can compare prediction regions
-defined by different models using
-\textcolor{violet}{many set difference distances \sout{the Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance}}
-as well as examining how well the truth epidemic matches the simulations
-by examining if the epidemic's trajectory lies within the prediction
-region. All these \textcolor{orange}{mentioned?} geometric structures
-and distance notations apply to epidemics with any number of states, and
-at the end of Section \ref{sec:overview} we \sout{also} highlight how
-these prediction regions can aid in visual comparisons for epidemics
-with 3 states (like the SIR models).
+estimators can be useful when comparing the true epidemic to a set of
+simulated epidemics, and practitioners can interpret the epidemic's
+extremeness score relative to the extremeness scores of the simulations
+very easily.
+
+In addition to being used to better understand completed epidemics,
+time-invariant tools can be used to aid prediction of future epidemics
+by providing a state-space regions in which we expect the true epidemic
+to traverse. In settings where the epidemic only generates a single
+outbreak, these regions can be very telling if simulation models capture
+the epidemic's structure. In \pkg{EpiCompare} we create geometric
+prediction regions around all but the \(\alpha\) proportion of most
+extreme simulated trajectories. These geometric regions can also be used
+to compared simulation models that have different time scales,
+parameters and even different statistical philosophies, through set
+different distances. Although visualization is the easiest when the
+epidemic has three states, prediction regions can be useful to assess
+and compare simulation models where the epidemics have multiple states.
+
+Overall, there are many tools to aid in the assessment and comparison of
+epidemics and models that avoid being affected by time-based parameters.
+We believe the time-invariant analysis provides many insights and should
+be in the toolkit of many epidemiologists. \pkg{EpiCompare} provides a
+strong starting point to do just that.
 
@@
 
+UPDATE 6/11/2021. I am now adding overall goals of section and each
+subheading. Additionally there are a few updates from last draft
+(unmarked since you didn't comment on last draft).
+
+SECTION GOALS:
+
+The main goal of this section is to show how a person can use EpiCompare
+in a real data analysis. With that said my primary focus is on
+EpiCompare and not the data analysis -- a point I think you may be
+having an issue with. I've given this some thought and believe that the
+data analysis is a means for an end to us -- showing off EpiCompare.
+That said again, I've made some changes in the first paragraph below.
+
+To accomplish this main goal, I shadow the data analysis pipeline we
+have introduced and show how epicompare can be used in the steps.
+
+The secondary goals are figuring out useful information about this
+measles outbreak.
+
 \textcolor{orange}{[[NEWEST TEXT}
 
-To conclude our paper, we demonstrate how \pkg{EpiCompare} can
-streamline the data analysis process with a case study of a measles
-outbreak in 1861-1862 Germany. With the help of \pkg{EpiCompare}, we can
-answer questions such as is a SIR model a good fit for the data, does
+{[}{[}sub-section goal: outline goal of section -- how to use epicompare
+in a real situation
+
+To conclude our paper, we demonstrate how \pkg{EpiCompare} streamlines
+the data analysis process with a case study of a measles outbreak in
+1861-1862 Germany. With the help of \pkg{EpiCompare}, we can answer
+important questions such as is a SIR model a good fit for the data, does
 the outbreak spread differently within the different school classes of
 the children, and can incorporating an underlying network structure
-enhance model fit. We first begin with some context for the epidemic.
-
-\hypertarget{background-for-1861-1862-measles-epidemic}{%
+enhance model fit? After introducing the data available in this
+outbreak, we show how \pkg{EpiCompare} can aid in each step of the data
+analysis pipeline (see Fig. \ref{fig:pipeline2}) to answer these
+questions motivated by this outbreak.{]}{]}
+
+\hypertarget{background-for-1861-1862-measles-outbreak}{%
 \subsection{Background for 1861-1862 measles
-epidemic}\label{background-for-1861-1862-measles-epidemic}}
+outbreak}\label{background-for-1861-1862-measles-outbreak}}
+
+{[}{[}Goal of this section isto provide context for our case
+study.{]}{]}
 
@@
 alleged infector of each child. We show a subset of the data in Table
-\ref{tab:hags-people}. We are particularly interested if and how group
-structure plays a role in the spread of measles.
+\ref{tab:hags-people}. We are particularly interested if the SIR model
+is a good fit to our data, whether we can identify subgroups with
+differing infection behavior, and whether incorporating those subgroups
+into a network-based SIR model enhances model fit compared to the
+baseline SIR.
 
@@
 
+{[}{[}The goal of this is to show how EpiCompare can be used to
+transform the data for basic tablaes and plotting, especially with
+regards to aggregating the data for the SIR model.{]}{]}
+
 We begin our analysis by first examining and transforming the raw data
-(\code{hagelloch_raw}), which is individual-level data. With
-\pkg{EpiCompare}, we can quickly determine how the epidemic progresses
-through the population at an aggregate level. By specifying the time of
-infection (\code{tI}) and the time of recovery (\code{tR}), the function
+(\code{hagelloch_raw}) to explore whether the SIR model is a good fit to
+the data. The raw data is in the format of individual-level data and
+first needs to be aggregated. By specifying the time of infection
+(\code{tI}) and the time of recovery (\code{tR}), the function
 \code{agents_to_aggregate()} calculates the number susceptible,
@@
 \ref{fig:hag-tern-raw}) or with our custom \code{geom},
-\code{geom_aggregate()}, which takes the raw agent data as input.
+\code{geom_aggregate()}, which takes the raw agent data as input. This
+is show in the below code.
 
@@
 
-Previous study tells us that measles outbreaks are often associated with
-children within the same grade level, and we examine if this is the case
-here. By combining the function \code{facet_wrap()} with
-\code{geom_aggregate()} we can easily analyze this scenario,
+With regards to our goal of identifying interesting subgroups with
+different infection behavior, previous study tells us that measles
+outbreaks are often associated with children within the same grade
+level, and we examine if this is the case here. By combining the
+\pkg{dplyr} function \code{facet_wrap()} with \code{geom_aggregate()} we
+can easily analyze this scenario,
 
@@
 \subsection{Modeling and Simulation}\label{modeling-and-simulation}}
+
+{[}{[}The goal is to `test' (although not formally) whether our
+observations from EDA are correct. We want to see whether SIR model is a
+good fit and whether including these found sub-groups helps model
+fit{]}{]}
+
+We now use modeling and simulation to informally test whether the
+baseline SIR model is a good fit for the data and whether incorporating
+a network-based structure dependent on the class of the children
+improves model fit.
 
@@
 18800 \(\times\) 5 data frame, which details the time of entry into the
-\(S\), \(I\), and \(R\) states for a given simulation. Ultimately, we
-would like to know if this simple SIR model is a good fit to the data,
-especially in comparison to a SIR model where we can incorporate the
-class structure.
-
-To fit a more complex SIR model with a network structure, we use package
-\pkg{EpiModel} \citep{Jenness2018}. The below code sets up a network of
-individuals (which includes class as a variable) and then simulates
-infections over this network.
+\(S\), \(I\), and \(R\) states for a given simulation.
+
+To fit a more complex SIR model with a network structure, we use the
+package \pkg{EpiModel} \citep{Jenness2018}. The below code sets up a
+network of individuals (which includes class as a variable) and then
+simulates infection and recovery over this network.
 
@@
 
-The next step is to compare the simple SIR model to the EpiModel SIR
-model. The \pkg{EpiCompare} function \code{fortify_aggregate()}, takes
-in an object from specialized classes of modeling output
+{[}{[}Finally we want to answer the questions in our analysis by
+comparing the models to the data and the models to each other{]}{]}.
+
+Ultimately we want to compare the models to the data and the models to
+one another to answer our questions of interest. The \pkg{EpiCompare}
+function \code{fortify_aggregate()} takes in an object from specialized
+classes of modeling output
 \textcolor{orange}{(like those made by \code{netsim()})} and transforms
@@
 \ref{fig:hag-simple-sir}, where a 90\% prediction region is estimated
-for the two models. For the Simple SIR model, we see that while the
-prediciton region covers the data fairly well, the prediction region
-clearly misses the second peak of infection. This indicates that the
-simple SIR is not a good fit to our data. We also see that the
+for the two models and the actual data are plotted as black dots. For
+the Simple SIR model, we see that while the prediction region covers the
+data fairly well, the prediction region clearly misses the second peak
+of infection. This indicates that the simple SIR is not a good fit to
+our data, even after incorporating noise. We also see that the
 prediction region is very large, covering up a large area of the ternary
@@
 EpiModel network model, we see that the prediction region covers the
-data quite well and takes up less area. With \pkg{EpiCompare}, we can
-see that the model using the class structure is a better fit to the
-outbreak than the simple SIR model.
+data quite well and takes up less area compared to the simple SIR. With
+\pkg{EpiCompare}, we see that the model using the class structure is a
+better fit to the outbreak than the simple SIR model.
 
@@
 
-The prediction region generated by the network model covers the observed
-data well, but that does not mean that individual filaments generated
-from the network model are good fits to the observed data. We can
-further examine the model fits by incorporating these filaments into our
-visualization. In Fig. \ref{fig:hag-simple-sir} we show the indivdual
-filaments generated from the two sets of models as gray lines. Examining
-these, we see that the individual lines typically only have one defined
-peak, whereas the data certainly looks like it has two distinct peaks,
-likely caused by our assumed super-spreader event.
+Although the prediction region generated by the network model covers the
+observed data well visually, that does not mean that individual
+filaments generated from the network model are good fits to the observed
+data. We can further examine the model fits by incorporating these
+filaments into our visualization. In Fig. \ref{fig:hag-simple-sir} we
+show the individual filaments generated from the two sets of models as
+gray lines. Examining these, we see that the individual lines typically
+only have one defined peak, whereas the data certainly looks like it has
+two distinct peaks, a feature possibly caused by our speculated
+super-spreader event.
 
@@
 \midrule
-Simple SIR & 0.0036733 & 0.00\\
-EpiModel SIR & 0.0118283 & 0.03\\
+Simple SIR & 0.0036733 & 0\\
+EpiModel SIR & 0.0028813 & 0\\
 \bottomrule
@@
 structure. Based on the prediction regions generated from these models,
-we saw that the network model fit the data much better than the simple
-SIR model. However, when we examined the individual filaments generated
-by the network model, we found that the data is highly unlikely to be
-generated from such a model. For further analysis, we would recommend
-looking into models that can more accurately capture super-spreading
-events based on the observation that one child was allegedly responsible
-for nearly all of his classmates' infections. Overall, this analysis
-demonstrates how \pkg{EpiCompare} aids in the data analysis pipeline for
-both novice and expert practitioners and coders alike.
+we saw that the network model fit the data better than the simple SIR
+model. However, when we examined the individual filaments generated by
+the network model, we found that the data are unlikely to be generated
+from such a model. For further analysis, we would recommend looking into
+models that can more accurately capture super-spreading events based on
+the observation that one child was allegedly responsible for nearly all
+of his classmates' infections. Overall, this analysis demonstrates how
+\pkg{EpiCompare} aids in the data analysis pipeline for both novice and
+expert practitioners and coders alike.
 

### 2021-07-15 jss/EpiCompare-paper/EpiCompare-paper.Rmd
minor commit for EpiComparepaper

@@
 
-With many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} available, this field has a number of devices to \shannon{\sout{aid}} \shannon{help}\footnote{\ben{[Ben says: I'm not sure this an improvement]}}  an individual practitioner decide the correct approach.
+With many visualization and exploratory tools, models and modeling paradigms, and reviews and comparisons in the literature and through the MIDAS (Models of Infectious Disease Agent Study) network \citep{midasnetwork2021} available, this field has a number of devices to \shannon{\sout{aid}} \shannon{help}[b] an individual practitioner decide the correct approach.
 For example, \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and \pkg{pomp} have all made significant steps in standardizing the flow of the data analysis pipeline for epidemic modeling through digitizing data sets, making accessible statistical models, and providing a plethora of educational material for both coding novices and experts alike \citep{surveillance2017,Jenness2018,King2016}.
@@
 To conclude our paper, we demonstrate the capabilities of \pkg{EpiCompare} with a complete data analysis of a measles outbreak in 1861-1862 Germany.
-Specifically, we demonstrate how tools in \pkg{EpiCompare} can be used in each step of the data analysis pipeline \footnote{\ben{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]} \shannon{check out new re-write.
-new goal is to motivate question of hagelloch and then show how epicompare helps}} \ben{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\ben{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that?
-/ what does it add to the conv/ why should they reference it?]}}.
+Specifically, we demonstrate how tools in \pkg{EpiCompare} can be used in each step of the data analysis pipeline [c]\shannon{check out new re-write.
+new goal is to motivate question of hagelloch and then show how epicompare helps}} (see Figure \ref{fig:pipeline})[d].
 Additionally, we highlight how time-invariant analysis (see Section \ref{sec:time-invariant}) can be used to enhance understanding of an outbreak\footnote{\ben{[Ben says: I'm not sure this is necessary to state.
@@
 \end{proof}
+
+[a]Welcome to "trackdown" version of our Rmd document. Hopefully this is useful for discussions. Let's try to convert \sout{} and colors to make it cleaner.
+[b]Ben says: I'm not sure this an improvement
+[c]based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?
+[d]Ben says: It's unclear why you want to highlight the figure again - could you be clearer on that? / what does it add to the conv/ why should they reference it?

### 2021-07-15 jss/EpiCompare-paper/EpiCompare-paper.tex
minor commit for EpiComparepaper

@@
 \citep{midasnetwork2021} available, this field has a number of devices
-to \textcolor{orange}{\sout{aid}}
-\textcolor{orange}{help}\footnote{\textcolor{violet}{[Ben says: I'm not sure this an improvement]}}
-an individual practitioner decide the correct approach. For example,
+to \textcolor{orange}{\sout{aid}} \textcolor{orange}{help}{[}b{]} an
+individual practitioner decide the correct approach. For example,
 \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
@@
 R> 
-R> 
 R> ggplot(hagelloch_sir, aes(x = S, y = I, z = R))+
@@
 R> both_models <- bind_rows(agg_model, fortified_net)
-R> 
 R> 
@@
 \pkg{EpiCompare} can be used in each step of the data analysis pipeline
-\footnote{\textcolor{violet}{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]} \textcolor{orange}{check out new re-write.
-new goal is to motivate question of hagelloch and then show how epicompare helps}}
-\textcolor{violet}{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\textcolor{violet}{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that?
-/ what does it add to the conv/ why should they reference it?]}}.
-Additionally, we highlight how time-invariant analysis (see Section
-\ref{sec:time-invariant}) can be used to enhance understanding of an
+{[}c{]}\textcolor{orange}{check out new re-write.  new goal is to motivate question of hagelloch and then show how epicompare helps}\}
+(see Figure \ref{fig:pipeline}){[}d{]}. Additionally, we highlight how
+time-invariant analysis (see Section \ref{sec:time-invariant}) can be
+used to enhance understanding of an
 outbreak\footnote{\textcolor{violet}{[Ben says: I'm not sure this is necessary to state. Additionally - is it really true do we highlight this?]}}.
@@
 
+{[}a{]}Welcome to ``trackdown'' version of our Rmd document. Hopefully
+this is useful for discussions. Let's try to convert \sout{} and colors
+to make it cleaner. {[}b{]}Ben says: I'm not sure this an improvement
+{[}c{]}based on an earlier read of this section, I might suggest
+something like ``and streamline the analysis process'' - is that a
+selling point you want to highlight? {[}d{]}Ben says: It's unclear why
+you want to highlight the figure again - could you be clearer on that? /
+what does it add to the conv/ why should they reference it?
+
 \bibliography{EpiCompare.bib}

### 2021-08-21 jss/EpiCompare-paper/EpiCompare-paper.tex
three group likelhood for SIR and hagelloch vignettte

--- 2021-07-01 jss/EpiCompare-paper/EpiCompare-paper.tex
@@
 \midrule
-Simple SIR & 0.0036733 & 0\\
-EpiModel SIR & 0.0028813 & 0\\
+Simple SIR & 0.0036733 & 0.00\\
+EpiModel SIR & 0.0118283 & 0.03\\
 \bottomrule

### 2021-09-11 jss/EpiCompare-paper/EpiCompare-paper.tex

@@
 government, and news organizations are an important first step of the
-process, as has been seen in the current COVID-19 pandemic
-\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
-
-With many visualization and exploratory tools, models and modeling
+process
+\textcolor{violet}{as has been seen in the current COVID-19 epidemic}\citep{dong2020,cdc-covid-tracker2021,wp-covid-tracker2021}.
+\footnote{\textcolor{violet}{[Ben says: probably should have a conclusion sentence here - seems to end abruptly. *This is less so the case now.]}}
+
+With the many visualization and exploratory tools, models and modeling
 paradigms, and reviews and comparisons in the literature and through the
 MIDAS (Models of Infectious Disease Agent Study) network
-\citep{midasnetwork2021} available, this field has a number of devices
-to \textcolor{orange}{\sout{aid}}
-\textcolor{orange}{help}\footnote{\textcolor{violet}{[Ben says: I'm not sure this an improvement]}}
-an individual practitioner decide the correct approach. For example,
+\citep{midasnetwork2021}, this field has a lot of devices to aid an
+individual practitioner decide the correct approach. For example,
 \proglang{R} packages such as \pkg{surveillance}, \pkg{EpiModel}, and
@@
 At the same time, analysis packages often only address a specific
-portion of the analysis pipeline. These modeling tools usually require
-learning package-specific syntax and often don't provide easy ways to
-compare and assess their models on new data. Moreover, exploring
-modeling and comparing epidemics require transforming and
+portion of the analysis
+pipeline\textcolor{violet}{\sout{, for instance focusing on certain types of models.}}
+\textcolor{violet}{These m}odeling
+tools\textcolor{violet}{\sout{, which}} usually require learning
+package-specific syntax\textcolor{violet}{\sout{,}} and often don't
+provide easy ways to compare and assess their models on new data.
+Moreover, exploring\textcolor{violet}{, \sout{and}} modeling
+\textcolor{violet}{and comparing} epidemics require transforming and
 \textit{tidying} data in different ways. To fill these gaps, we present
@@
 
-One of \pkg{EpiCompare}'s main contribution to comparison and assessment
-of epidemics is through tools that provide \textit{time-invariant}
-assessments. Epidemics, despite being defined as a process that evolves
-over time, often need to be compared in a way not constrained to initial
-times or time scales in order to understand the processes at play. With
-time-invariant analysis, comparing decades-long outbreaks of HIV in the
-US to a 10 day outbreak of norovirus on a cruise ship is possible.
-Compared to time-dependent comparison tools for state-space modeling,
-time-invariant analysis can make it easier to compare state-space
-epidemic representations in a more global, holistic fashion. Many
-time-dependent comparison tools only examine the proportion of
+\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
+\textit{time-invariant} way. Epidemics, despite by definition being a
+process that evolves over time, often need to be compared in a way not
+constrained to initial times or time scales to understand the processes
+at play.
+\textcolor{violet}{Time-invariant analysis can also make it easier to compare state-space models in a more global, holistic fashion. \sout{Moreover, m} M}any
+current \textcolor{violet}{time-dependent} comparison tools for
+state-space models (e.g.~SIR models)
+\textcolor{violet}{\sout{highlight} examine} the proportion of
 individuals in each state (at a given time) in a piece-wise / marginal
-fashion. These time-dependent approaches can reduce the amount of
-connections that can be seen and insights that can be drawn\}, similar
-to examining projections of a multidimensional distribution onto a
-single axis, one at a time. Tools in \pkg{EpiCompare} extend the user
-toolkit to evaluate epidemics within a time-invariant lens. At the same
-time, the goal of \pkg{EpiCompare} is not to supplant existing
-infectious disease modeling tools and software but, rather, is a
-concerted effort to create standard and fair comparisons among models
+fashion. \textcolor{violet}{These \sout{This}}
+approach\textcolor{violet}{es} may reduce the amount of connections that
+can be seen, similar to projections of a multidimensional distribution
+onto a single axis at a time. Tools in \pkg{EpiCompare} give the user
+the ability to extend their toolkit to evaluate epidemics within a
+time-invariant lens. The goal of \pkg{EpiCompare} is not to supplant
+existing infectious disease modeling tools and software but, rather, is
+a concerted effort to create standard and fair comparisons among models
 developed for disease outbreaks and outbreak data.
@@
 \pkg{EpiCompare} aids a practitioner in every step of the pipeline and
-section \ref{sec:tour} provides a demonstration of the tools through a
-detailed
-example\textcolor{orange}{\sout{ of a full data analysis pipeline.}}
-\textcolor{orange}{from start to finish of the data analysis pipeline}\footnote{\textcolor{orange}{maybe clearer?
-up to you} \textcolor{violet}{The current suggestion isn't clearer - feel free to try again.}}
-
-\hypertarget{sec:time-invariant}{%
-\section{Time-invariant analysis}\label{sec:time-invariant}}
-
-\pkg{EpiCompare} emphasizes the value of analyzing epidemics in a
-\textit{time-invariant} way - approaches that remove some or all of the
-impact of start/end times and recording time scales when preforming the
-analysis. In this section we highlight some weaknesses of
-time-\textit{de}pendent analysis and define the mathematical
-underpinning of the time-invariant approach we take. To accomplish this
-goal we first demonstrate the inability of time-dependent tools to
-adequately quantify a classic epidemic parameter, the reproduction
-number \(R_0\). Then we motivate new time-invariant approaches for more
-complex situations where \(R_0\) does not capture the complexities of
-outbreaks. This leads to the final subsection that defines ways to view
-epidemics in a time-invariant lens and discusses natural properties of
-these representations that allow for clearer ways to compare models and
-epidemics.
-
-\subsection[Motivation through the reproduction number $R_0$]{Motivation
-of time-invariant analysis through the reproduction number
-\(R_0\)}\label{sec:r0_subsection}
-
-Epidemiological research is often interested in understanding intrinsic
-properties of the epidemic, independent from when the epidemic occurred
-or how frequently data was collected. One of the most famous numerical
-summaries of an epidemic is the reproduction number - \(R_0\). This
-numerical summary is a time-invariant value that is defined as the
-expected number of infections caused by a single infect or who is added
-to a completely susceptible population \citep{anderson1992}.
-\citet{Gallagher2020} showed that the estimation of \(R_0\) can be
-sensitive to time-dependent parameters like estimation of the beginning
-and end of an epidemic. This time-dependent sensitivity in estimating
-\(R_0\) reflects a general problem in epidemiology surrounding how to
-transform time-dependent information into insights into desirable,
-intrinsic properties of an epidemic.
-
-In many situations, epidemiologists want a much deeper understand of an
-epidemic than just the reproduction number. One common time-dependent
-tool used to better understand epidemics is a series of time series line
-plots that track the proportion of the population in a given state
-(e.g.~infected) at a given time. Figure
-\ref{fig:different-scales-standard} visualizes two different simulated
-epidemics with population states (S)usceptible, (I)nfectious and
-(R)ecovered. These simulations have been generated under a discrete
-approximation of \citet{Kermack1927}'s SIR model. The Kermack and
-McKendrick model captures the transitions from one state to the next as
-a system of ordinary differential equations
+section \ref{sec:tour} provides a \textcolor{violet}{\sout{thorough}}
+demonstrating of the tools through a detailed example of a full data
+analysis pipeline.
+
+\section[Time-invariant]{Motivation and tools for time-invariant
+analysis}\label{sec:time-invariant}
+
+\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking
+a global, not marginal view of how epidemics move through populations
+and (2) by treating full epidemics as filamental trajectories. The
+following section aims to highlight the strengths of
+\textit{time-invariant} \textcolor{orange}{analysis} and define the
+mathematical foundations that \pkg{EpiCompare}'s tools stand upon.
+
+Mathematically, epidemics are complex objects. They can be hard to
+assess and compare to one another due to the differences in the
+diseases, the location where the outbreak occurs, how the affected
+population reacts, and the time
+\sout{related}\textcolor{orange}{related} features (including start of
+the epidemic, speed of infection and more). Time-invariant analysis
+makes different epidemics easier to compare by removing many time
+dependent aspects of an epidemic. \sout{Instead,}
+\textcolor{orange}{Time-invariant analysis} focuses
+\textcolor{orange}{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.}
+\textcolor{violet}{[Ben wants to try this sentence again.]}
+
+\subsection[motivating through R0]{Motivating time-invariant analysis
+through the reproduction number \(R_0\)}\label{sec:r0}
+
+Time-invariant analysis, as it appears in \pkg{EpiCompare},
+\textcolor{orange}{bypasses} many difficulties \textcolor{orange}{in}
+comparing different epidemics. With time-invariant analysis, comparing
+the decades-long outbreak of HIV in the US to a 10 day outbreak of
+norovirus on a cruise ship is \sout{still} possible. Time-dependent
+problems can arise when estimating epidemiological parameters, including
+the reproduction number \(R_0\).
+\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence.}
+  \textcolor{violet}{I still think it adds value to the story and I'm not sure people really read section titles that are long.}}
+
+\(R_0\) is probably the most famous \sout{time-invariant} numerical
+summary of an epidemic and is often associated with the
+Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
+\(R_0\) is \sout{a one-number summary of a disease and is }defined as
+the expected number of infections caused by a single infector who is
+added to a completely susceptible population \citep{anderson1992}.
+\textcolor{violet}{This \textcolor{orange}{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\textcolor{orange}{.
+Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a difficult quantity to estimate.}
+\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
+To demonstrate the difficulty of discerning \(R_0\) in
+\textcolor{violet}{\sout{a}other}\footnote{\textcolor{violet}{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}}
+time-dependent analysis, we first introduce \citet{Kermack1927}'s SIR
+model. This model captures the transitions from one state to the next as
+a system of ordinary differential equations, where \(N\) is the total
+number of individuals, \(\beta\) is the rate of infection, and
+\(\gamma\) is the rate of recovery,
 
@@
     I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
-    R^\prime(t) &= \gamma I(t) \nonumber,
+    R^\prime(t) &= \gamma I(t) \nonumber.
 \end{align}
 
-where \(N\) is the total number of individuals, \(\beta\) is the rate of
-infection, and \(\gamma\) is the rate of recovery.
-
-From this model, the reproduction number is the ratio of the infection
-rate to the recovery rate, \(R_0 = \beta/\gamma\).
+From this
+model,\textcolor{orange}{the reproduction number is the ratio of the infection rate to the recovery rate,}
+\(R_0 = \beta/\gamma\)\sout{, aka the ratio of the infection rate compared to the recovery rate.}
+\textcolor{violet}{From this definition, given 
+\sout{Since}} \(\beta\) and \(\gamma\) are both rates,
+\textcolor{violet}{it should be clear that} the ratio of the two,
+\(R_0\), is a time-invariant
+quantity.\footnote{\textcolor{violet}{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}
+\sout{ Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.
+Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.
+}\footnote{cool
+  facts about r0, but not the central point}
+
+\textcolor{violet}{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$?
+Also, I feel like the story is weak here.
+The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$.
+In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
+\textcolor{orange}{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.
+also highlighted tie to time-invariant analysis and $R_0$.
+I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}\textcolor{orange}{ Shannon tries again in \textcolor{blue}{blue}}
+
+\sout{\textcolor{violet}{[Ben says: this paragraph needs to still be looked at.
+Also I'm not sure why this particular paragraph was changed so much at all.
+Currentlly, the way we present comparing these epidemic's $R_0$s isn't well grouped/motivated.]} \textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations.
+The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
+Both epidemics have populations of 1000 people with 10 individuals initially infected.
+Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.
+We plot the epidemics with traditional $state$ vs.
+$time$ plots}\footnote{This sentecne is out of place / doesn't connect with the other sentences.}.
+\sout{In Fig.
+\ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.
+In this time-variant view, we may believe that \textcolor{violet}{epidemic} 1 has a larger $R_0$ than \textcolor{violet}{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.
+On the other hand, we may believe \textcolor{violet}{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \textcolor{violet}{epidemic} has not yet peaked at time 15.
+In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$}\footnote{\textcolor{violet}{This sentnce doesn't connect wit previous examples.}}.
+
+\textcolor{blue}{Since $R_0$ is an important value, it would be helpful to have more intuitive ways of comparing one $R_0$ to another.
+Usually numerical summaries of $R_0$ are presented, which while overall very helpful, may be confusing when presented along side epidemic data that are visualized in a traditional, time-dependent manner.}
+
+\textcolor{blue}{For example, consider two epidemics generated from the Kermack and McKendrick SIR equations where both have the same value of $R_0$.
+The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.
+Both epidemics have populations of 1000 people with 10 individuals initially infected.
+An analysis may present an estimate of $\hat{R}_0 = 2$ alongside state vs.
+time plots like those shown in Figure \ref{fig:different-scales-standard}.
+The paths of the epidemics in the state vs.
+time view seem to differ from one another including having different infection peaks.
+From these traditional time-based plots, there is no intuitive way to conclude that these two epidemics have the same value of $R_0$.}
 
@@
 
-\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ parameters but the same initial reproduction number $R_0$ = 2.
+\caption{\label{fig:different-scales-standard}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.
 Both epidemics are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
@@
 
-Even though Figure \ref{fig:different-scales-standard}'s visual analysis
-might be able to present more complex properties of the epidemic, one
-might wish to understand how these two simulated epidemics' \(R_0\)
-values compare. Even though these two simulated epidemics appear
-different in these figures (including having different infection peaks),
-there is no real intuitive way to compare the epidemics' \(R_0\) values.
-In fact, both of these simulated epidemics started with the same
-population (1000 total people with 10 infected) and have the same
-\(R_0\) value, just with slightly different infection and recovery rates
-(\(\beta_1, \gamma_1 = 0.8, 0.4\), and
-\(\beta_2, \gamma_2 = .64, .32\)). Even for simple generative models,
-this time-dependent visualization cannot be used to compare a very
-simple but powerful numerical summary, \(R_0\).
-
-\hypertarget{ternary-plots-a-time-invariant-visualization-tool}{%
-\subsection{Ternary plots: a time-invariant visualization
-tool}\label{ternary-plots-a-time-invariant-visualization-tool}}
-
-The faceted time series plot, like that seen in Figure
-\ref{fig:different-scales-standard}, not only fails to allow the
-practitioner to compare simulated epidemics' \(R_0\) values but also
-presented epidemics' trajectory data so that only the marginal
-information for each state can be examined at a given time. A more
-holistic approach to visualizing the overall path of our simulated
-epidemics is to examine how each epidemic traverses the
-three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
-just that, presenting the trajectories of the simulated epidemics in
-this three-dimensional space. For state space models like in our
-example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
-to \(N\) (the total population size) we can visual these point in a a
-two-dimensional \textit{ternary} plot, as seen in Figure
-\ref{fig:different-scales-tern}'s right plot. In both of Figure
-\ref{fig:different-scales-tern}'s plots we observe that the two
-epidemics are on the same trajectory. In this simple generative example
-setting, being on the same trajectory indicates that the two epidemics
-have the same \(R_0\) value. This can be proven mathematically.
+\textcolor{blue}{\pkg{EpiCompare} provides a time-invariant tool to visualize these epidemics in a more intuitive manner, at least in regards to comparing values of $R_0$.}
+\sout{A time-invariant approach to visualizing epidemics, in comparison, allows us to directly compare $R_0$ from a single plot.}
+\textcolor{violet}{For every time \sout{point} $t$ we have a \textcolor{orange}{point} $(S(t),I(t), R(t))$\sout{, so we can treat epidemics as a trajectory in this three-dimensional space, as we visual in the left subplot of Figure \ref{fig:different-sacles-tern}.} \textcolor{orange}{so we can visualize the trajectory of the epidemic in three-dimensional space (see Fig.
+\ref{fig:different-scales-tern} (left)).} For state space models like in our example, given the constraint that $S(t) + I(t)+R(t)$ is always equal to $N$ (the total population size), we can visual these point in a a two-dimesional \textit{ternary} plot, as seen in Figure \ref{fig:different-scales-tern} (right).
+\sout{In Fig.
+\ref{fig:different-scales-tern}
+we plot the filamental trajectories of the two epidemics in a time-invariant view. We will explain how and why this works shortly. The important takeaway is that in this time-invariant view, i}I}t
+is apparent that these epidemics are on ``the same path.'' In this case,
+this indicates that two epidemics have the same value of \(R_0\).
 
@@
 \caption{\label{fig:different-scales-tern}Left: trajectory of epidemic in three-dimensional space, plotting $(S(t), I(t), R(t))$.
-Right: the gray-shaded region and epidemic trajectories shown from (left) now shown in two-dimensional space.
+Right: the gray-shaded region and epidemic trajectory shown from (left) now shown in two-dimensional space.
 This is more commonly known as a ternary plot.}\label{fig:unnamed-chunk-3}
@@
 
+\textcolor{violet}{Underlying our time-invariant visualization that allowed for the comparison of $R_0$ in Fig.
+\ref{fig:different-scales-tern} is the treatment of the epidemic as a single filamental trajectory in the state space.
+\sout{The reason why we can visually compare $R_0$ in Fig.
+\ref{fig:different-scales-tern} is because of the time-invariant nature of the filamental trajectory associated with an epidemic.}}
+A filamental trajectory can be mathematically viewed as a set of points
+in space that have an ordering, and that all points on the line between
+these ordered points are also contained in the geometric object. For a
+SIR epidemic, we can represent the associated filamental trajectory
+\(\psi\) as
+
+\[
+\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
+\] where a mapping \(\xi : s \to \mathbb{R}\) that is strictly
+monotonically increasing would not change the definition of \(\psi\),
+i.e.~\(\psi_\xi \equiv \psi\) where :
+
+\[
+\psi_\xi = \left \{(S(\xi(s)), I(\xi(s)), R(\xi(s))): S, I, R \ge 0, S + I + R = N \right \}_{s\in[0,T]}.
+\] \textcolor{violet}{[Ben says: removed this paragraph now]} Since the
+number \textcolor{orange}{of individuals} in each state is non-negative
+and the sum over the three states for a given time point
+\sout{sums to}\textcolor{orange}{is} \(N\), then
+\textcolor{violet}{all points in} \(\psi\) will lay in a
+\textbackslash ben\{two-dimensional triangular plane in
+three-dimensional space.
+\sout{We can then which can be visualize the \textcolor{orange}{full filamental trajectory} in a two-dimensional ternary plot}
+As a result, we can visualize
+\textcolor{orange}{the full filamental trajectory} of an epidemic in a
+single
+\textcolor{violet}{ternary plot \sout{2d plot and ultimately $R_0$}}.
+\textcolor{orange}{[Shannon says: Show pic here?]}{]}\footnote{Which?  \textcolor{orange}{3d space one? But I'm leaning against it now.   3d never looks good in a paper.}}
+
+\sout{\textcolor{violet}{[This section could be a bit less wordy.
+But generally good.]} We visualize the two epidemics in a global, ternary view in Figure \ref{fig:different-scales-tern}.
+Without getting into too much detail of the intricacies of this plot, we immediately see the points of the two filaments $\psi$ seem to form the same trajectory.
+Now, it is much clearer that \textcolor{violet}{\sout{Model} epidemic} 2 is following the same trajectory as \textcolor{violet}{\sout{Model} epidemic} 1 but is not as far along in the infection process.
+}
+
+\sout{\textcolor{violet}{As suggested a few paragraphs back, t}}The
+filamental trajectories in Fig. \ref{fig:different-scales-tern}
+\sout{seem to overlap}, and we may suspect that something is
+fundamentally linking these two different epidemics together.
+Mathematically, we can show this fundamental link \sout{turns} is
+\(R_0\). Let our two epidemics be presented as
+\(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
+\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
+example, assume both models have the same initial values
+\((S(0), I(0), R(0))\), and let
+\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
+\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
+rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
+the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
+\(\gamma_2 = a \gamma_1\).
+
 \begin{theorem}\label{thm:sir-scale}
- Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (i.e.
-$R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$) and define $a > 0$ such that $\beta_2 = a\beta_1$.
-Then for all $t > 0$ there exists $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.
-Moreover, $s = \frac{1}{a}t$.
+Let there be two SIR models as described above.  Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
 \end{theorem}
@@
 for two SIR models that have the same initial percent of individuals in
-each state and \(R_0\), then every point on the epidemic path of the
-first SIR model can be mapped to a point on the epidemic path of the
-second SIR model. In other words, the two epidemics form the same
-filamental trajectory.
-
-\hypertarget{time-invariance-beyond-sir-models-trajectories-and-filaments}{%
-\subsection{Time-invariance beyond SIR models: Trajectories and
-Filaments}\label{time-invariance-beyond-sir-models-trajectories-and-filaments}}
+each state and \(R_0\) then for every point on the epidemic path of the
+first SIR model \textcolor{violet}{\sout{is also} can be mapped to} a
+point on the epidemic path of the second SIR model.
+\textcolor{orange}{In other words, the two epidemics form the same filamental trajectory.
+For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}
+
+\subsection[Beyond R0 and SIR]{Time-invariant analysis beyond \(R_0\)
+and \sout{Kermack’s and McKendrick} SIR
+Models\footnote{Probably will need to change this title...}}\label{sec:beyond-r0-sir}
 
@@
 filamental trajectories embedded in a lower dimensional space allows us
-to better compare the overall structure of the epidemic and see how the
-population is directly impacted. In this section we present
-time-invariant tools that can be applied to complex epidemics where the
-epidemic's generative process is unknown and can have more than three
-states. These tools leverage the idea that an epidemic can be viewed as
-a trajectory and that many properties of the epidemic are well captured
-when we do so. This approach is useful when the epidemic of interest has
-only gone through a single realization of its outbreak (before the
-population of individuals become susceptible again).
-
-The first set of tools allows a practitioner to define distances between
-epidemics whose time features do not align. For completed epidemics, one
-way to better examine their properties is to represent their filamental
-trajectories as a finite sequence of equally spaced points. This
-representation induces a natural distance between epidemics,
-specifically:
-
-\[
-d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds \nonumber,
-\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
-\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\) (with
-\(|\psi_i|\) is the length of \(\psi_i\)). This distance is naturally
-time-invariant, and can be plugged into multiple distance-based
-assessment tools to examine the overall ``extremeness'\,' of points,
-including pseudo-density estimators and depth/local depth functions
+to \sout{better}\textcolor{orange}{more fully} compare the overall
+structure of the epidemic and see how the population is directly
+impacted. Time-invariant tools \sout{that} can be useful even when the
+underlying generative model for the epidemic is unknown or have more
+than three epidemic states.
+
+\textcolor{violet}{New paragraph} Viewing epidemics as filamental
+trajectories provides \sout{a lot }new ways to compare and examine
+epidemics in a time-invariant manner. \textbackslash ben\{For
+\textcolor{orange}{completed?}epidemics that have ended, one way to
+examine their filamental trajectories is to
+\sout{redefine}\textcolor{orange}{represent} the
+\textcolor{orange}{filamental} trajectory as a
+\textcolor{orange}{finite sequence of equally spaced points.}
+\sout{finite sequence of points on the filamental trajectory that are equally spaced (i.e. equa-distance between pairs of ordered points).}
+\sout{For epidemics that have "played" themselves out, one way to represent their filamental trajectories to avoid \textcolor{orange}{confusion stemming from}\sout{impacts of} temporal structure is to define them as a sequence of points their trajectory with equi-distance between each point \textcolor{orange}{[Shannon says: are we missing a few words in this definition?]}}.
+This representation induces a natural distance between this type of
+representation between epidemics, specifically: \[
+d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\psi_1'(s) - \psi_2'(s))^2 ds
+\] where \(\psi_i'(s)\) the point along \(\psi_1\) that is \(s\)
+fraction of \(|\psi_1|\) distance away from the start of
+\(\psi_i\).\footnote{\textcolor{orange}{I think you're trying to say something about a distance based on the equally space points.
+Some clarifying questions:  is $\psi^\prime$ the derivative?
+Does proportion make more sense than fraction?
+or simply $\frac{|\psi|}{s}$?
+It's only naturally time-invariant if we have a well defined ending point, right?}}
+This distance is naturally time-invariant, and can be plugged into
+multiple distance-based assessment tools to examine the overall
+``extremeness'\,' of points, including pseudo-density estimators and
+depth/local depth functions
 \citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
-estimators can be useful when comparing the true epidemic to a set of
-simulated epidemics, and practitioners can interpret the epidemic's
-extremeness score relative to the extremeness scores of the simulations
-very easily.
-
-In addition to being used to better understand completed epidemics,
-time-invariant tools can be used to aid prediction of future epidemics
-by providing a state-space regions in which we expect the true epidemic
-to traverse. In settings where the epidemic only generates a single
-outbreak, these regions can be very telling if simulation models capture
-the epidemic's structure. In \pkg{EpiCompare} we create geometric
-prediction regions around all but the \(\alpha\) proportion of most
-extreme simulated trajectories. These geometric regions can also be used
-to compared simulation models that have different time scales,
-parameters and even different statistical philosophies, through set
-different distances. Although visualization is the easiest when the
-epidemic has three states, prediction regions can be useful to assess
-and compare simulation models where the epidemics have multiple states.
-
-Overall, there are many tools to aid in the assessment and comparison of
-epidemics and models that avoid being affected by time-based parameters.
-We believe the time-invariant analysis provides many insights and should
-be in the toolkit of many epidemiologists. \pkg{EpiCompare} provides a
-strong starting point to do just that.
+estimators can be \sout{very} useful when comparing
+\sout{between a setof simulation} \textcolor{orange}{a set of simulated}
+epidemics and the true
+epidemic\textcolor{orange}{. Moreover these extremness estimators,}
+\sout{and does} \textcolor{orange}{} not constrain the number of states
+of the
+model\sout{, though we recommend projecting the points into the unit simplex}\footnote{\textcolor{violet}{I'm not sure we've talked about this before...
+\textcolor{orange}{I don't think we have but am wondering if we're getting in the weeds}}}\sout{(by making all values the proportion of the population in the given state)}.
+
+\textcolor{violet}{New paragraph:}
+\textcolor{violet}{\sout{If the set of epidemics that one is examining have only gone through a single cycle of the outbreak} If \sout{one} \textcolor{orange}{a practicioner} is interested in understanding \textcolor{orange}{an} epidemic\sout{s} through a single \sout{cycle}\textcolor{orange}{realization} of \sout{their}\textcolor{orange}{its} outbreak}
+(before the population of individuals have become susceptible again),
+then additional time-invariant tools, including prediction regions can
+be leveraged \textcolor{orange}{awk sentence}\footnote{\{\textcolor{orange}{What are we predicting if the epidemic is done? Update 4/6: I'm satisfied.}\}}.
+In these settings, \textcolor{orange}{\pkg{EpiCompare} goes}
+\sout{we go }a step further and treats epidemics more like geometric
+filaments
+\textcolor{orange}{(i.e. filamental trajectories without an ordering of points)}
+than filamental trajectories. In \pkg{EpiCompare}, we create prediction
+regions that contain a the top (\(1-\alpha\)) proportion of simulated
+curves by defining geometric regions defined by the union of small
+\textcolor{orange}{geometric?} filaments around the subset of
+simulations (\sout{subset} \textcolor{orange}{grouped} by measures like
+the above pseudo-density estimates or depth estimates). These regions
+\sout{look at}\textcolor{orange}{show} where in the state-space we
+expect the epidemic to traverse\sout{, and }.
+\textcolor{orange}{Additionally, }we can compare prediction regions
+defined by different models using
+\textcolor{violet}{many set difference distances \sout{the Haussdorf \textcolor{orange}{why Haussdorf specifically?} distance}}
+as well as examining how well the truth epidemic matches the simulations
+by examining if the epidemic's trajectory lies within the prediction
+region. All these \textcolor{orange}{mentioned?} geometric structures
+and distance notations apply to epidemics with any number of states, and
+at the end of Section \ref{sec:overview} we \sout{also} highlight how
+these prediction regions can aid in visual comparisons for epidemics
+with 3 states (like the SIR models).
 
